<!-- \mainmatter -->

# Return Calculations {#Return-Calculations}

Updated: August 30, 2026

Copyright © Eric Zivot 2022, 2023, 2025, 2026

In this chapter I cover asset return calculations with an emphasis on
equity returns. It is important to understand the different ways in
which asset returns are calculated because most of the models presented
in the book are for asset returns. Simple returns are most commonly used
for money and portfolio calculations in practice. However, simple
returns have some undesirable properties that make mathematical and
statistical modeling difficult. Continuously compounded returns, in
contrast, have nicer properties that make mathematical and statistical
modeling easier.

This chapter is organized as follows. Section @sec-TimeValueMoney covers
basic time value of money calculations. Section
@sec-AssetReturnCalculations covers asset return calculations, including
both simple and continuously compounded returns. Section
@sec-portfolios-and-portfolio-returns defines portfolios and details
portfolio return calculations. Section @sec-ReturnCalculationsR
illustrates asset return calculations using R.

In this chapter I use the following R packages: **ggplot2**, **introcompfinr**,
**methods**, **PerformanceAnalytics**, **quantmod**, **xts**, and **zoo**. 
Make sure these packages are installed and loaded before replicating the chapter
examples.

```{r}
#| include: false
options(digits=3)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(PerformanceAnalytics))
suppressPackageStartupMessages(library(xts))
suppressPackageStartupMessages(library(methods))
```

## The Time Value of Money {#sec-TimeValueMoney}

This section reviews basic time value of money calculations. The
concepts of future value, present value and the compounding of interest
are defined and discussed.

### Future value, present value and simple interest

Consider an amount $\$V$ invested for $n$ years at a *simple interest
rate* \index{Simple interest rate} of $R$ per annum (where $R$ is
expressed as a decimal). If compounding takes place only at the end of
the year, the *future value* \index{Future value} after $n$ years is:
$$
FV_{n}=\$V(1+R)\times\cdots\times(1+R)=\$V(1+R)^{n}.
$$ {#eq-futureValue} Over the first year, $\$V$ grows to
$\$V(1+R)=\$V+\$V\times R$ which represents the initial principal $\$V$
plus the payment of simple interest $\$V\times R$ for the year. Over the
second year, the new principal $\$V(1+R)$ grows to
$\$V(1+R)(1+R)=\$V(1+R)^{2},$ and so on.

::: {#exm-example0101 .example name="Future value with simple interest"}
Consider putting \$1000 in an interest checking account that pays a
simple annual percentage rate of 3%. The future value after $n=1,5$ and
10 years is, respectively, \begin{align*}
FV_{1} & =\$1000\cdot(1.03)^{1}=\$1030,\\
FV_{5} & =\$1000\cdot(1.03)^{5}=\$1159.27,\\
FV_{10} & =\$1000\cdot(1.03)^{10}=\$1343.92.
\end{align*} Over the first year, \$30 in interest is accrued; over
three years, \$159.27 in interest is accrued; over five years, \$343.92
in interest is accrued. These calculations are easy to perform in R.
:::

```{r}
#| echo: true

V = 1000 
R = 0.03 
FV1 = V*(1 + R) 
FV5 = V*(1 + R)^5 
FV10 = V*(1 + R)^10
c(FV1, FV5, FV10)
```

\hfill $\blacksquare$

The future value formula @eq-futureValue defines a relationship between
four variables: $FV_{n},$ $V$, $R$ and $n$. Given three variables, the
fourth variable can be determined. For example, given $FV_{n}$, $R$ and
$n$ and solving for $V$ gives the *present value* \index{Present value}
formula: $$
V=\frac{FV_{n}}{(1+R)^{n}}.
$$ {#eq-presentValue} Here, $V$ represents the present value (or current value)
of the future value $FV_{n}$ that is to be received $n$ years from
today. The future value is discounted by the factor $(1+R)^{-n}$ to
reflect the fact that dollars to be received in $n$ years are worth less
than current dollars because current dollars can be invested today
returning $R$ per year. The discount factor $(1+R)^{-n}$ is less than
$1$, and represents the price today of $\$1$ to be received in $n$ years
when the interest rate is $R$ per year.

Taking $FV_{n}$, $n$ and $V$ as given, the *annual interest rate* on the
investment is defined as: $$
R=\left(\frac{FV_{n}}{V}\right)^{1/n}-1.
$$ {#eq-annualReturn} Here, $R$ represents a type of average rate of return on
an investment known as the *compound annual return*.

Finally, given $FV_{n}$, $V$ and $R$ we can solve for $n$:
$$
n=\frac{ln(FV_{n}/V)}{ln(1+R)}.
$$ {#eq-horizon} The expression @eq-horizon is called the *horizon period*
and can be used to determine the number of years it takes for an
investment of $\$V$ to double. Setting $FV_{n}=2V$ in @eq-horizon gives:
$$n=\frac{\ln(2)}{\ln(1+R)}\approx\frac{0.7}{R},$$
which uses the approximations $\ln(2)=0.6931\approx0.7$ and
$\ln(1+R)\approx R$ for $R$ close to zero. The approximation
$n\approx0.7/R$ is called the *rule of 70*.

::: {#exm-example0102 .example name="Using the rule of 70"}
\index{Rule of 70} The rule of 70 gives a good approximation as long as
the interest rate is not too high.
:::

```{r}
#| echo: true

R = seq(0.01, 0.10, by=0.01) 
nExact = log(2)/log(1 + R) 
nRule70 = 0.7/R 
cbind(R, nExact, nRule70)
```

For $R$ values between 1% and 10% the approximation error in the rule of
70 is never more than half a year.

\hfill{}$\blacksquare$

### Multiple compounding periods

\index{Compound interest} If interest is paid $m$ times per year, then
the future value after $n$ years is:
$$FV_{n}^{m}=\$V\cdot\left(1+\frac{R}{m}\right)^{m\cdot n}.$$
$R/m$ is often referred to as the *periodic interest rate*
\index{Periodic interest rate}. As $m$, the frequency of compounding,
increases, the rate becomes continuously compounded and it can be shown
that the future value becomes:
$$FV_{n}^{c}=\lim_{m\rightarrow\infty}\$V\cdot\left(1+\frac{R}{m}\right)^{m\cdot n}=\$V\cdot e^{R\cdot n},$$
where $e^{(\cdot)}$ is the exponential function and $e^{1}=2.71828$.

::: {#exm-example0103 .example name="Future value with different compounding frequencies"}
For a simple annual percentage rate of 10%, the value of \$1000 at the
end of one year ($n=1)$ for different values of $m$ is calculated below.
:::

```{r}
#| echo: true

V = 1000 
R = 0.1 
m = c(1, 2, 4, 356, 10000) 
FV = V*(1 + R/m)^(m) 
print(cbind(m, FV), digits=7)
```

The result with continuous compounding is

```{r}
#| echo: true

print(V*exp(R), digits=7)
```

\hfill{}$\blacksquare$

The continuously compounded return analogues to the present value,
annual return and horizon period formulas @eq-presentValue,
@eq-annualReturn and @eq-horizon are: \begin{align*}
V & =e^{-Rn}FV_{n},\\
R & =\frac{1}{n}\ln\left(\frac{FV_{n}}{V}\right),\\
n & =\frac{1}{R}\ln\left(\frac{FV_{n}}{V}\right).
\end{align*}

### Effective annual rate

We now consider the relationship between simple interest rates, periodic
rates, effective annual rates and continuously compounded rates. Suppose
an investment pays a periodic interest rate of 2% each quarter. This
gives rise to a simple annual rate of 8% (2% $\times$ 4 quarters). At
the end of the year, \$1000 invested accrues to:
$$\$1000\cdot\left(1+\frac{0.08}{4}\right)^{4\cdot1}=\$1082.40.$$
The *effective annual rate*,\index{Effective annual rate} $R_{A}$, on
the investment is determined by the relationship:
$$\$1000\cdot(1+R_{A})=\$1082.40.$$
Solving for $R_{A}$ gives:
$$R_{A}=\frac{\$1082.40}{\$1000}-1=0.0824,$$
or $R_{A}=8.24\%.$ Here, the effective annual rate is the simple
interest rate with annual compounding that gives the same future value
that occurs with simple interest compounded four times per year. The
effective annual rate is greater than the simple annual rate due to the
payment of interest on interest.

The general relationship between the simple annual rate $R$ with
payments $m$ times per year and the effective annual rate, $R_{A}$, is:

$$(1+R_{A})=\left(1+\frac{R}{m}\right)^{m}.$$
Given the simple rate $R$, we can solve for the effective annual rate
using: $$
R_{A}=\left(1+\frac{R}{m}\right)^{m}-1.
$$ {#eq-effectiveAnnualRate} Given the effective annual rate $R_{A}$, we can solve for
the simple rate using:
$$R=m\left[(1+R_{A})^{1/m}-1\right].$$
The relationship between the effective annual rate and the simple rate
that is compounded continuously is:
$$(1+R_{A})=e^{R}.$$
Hence, \begin{align*}
R_{A} & =e^{R}-1,\\
R & =\ln(1+R_{A}).
\end{align*}

::: {#exm-example0104 .example name="Determine effective annual rates"}
The effective annual rates associated with the investments in example
@exm-example0102 are calculated below.
:::

```{r}
#| echo: true

RA = FV/V - 1  
print(cbind(m, FV, RA), digits=7)
```

The effective annual rate with continuous compounding is

```{r}
#| echo: true
print(exp(R) - 1, digits=7)
```

\hfill{}$\blacksquare$

::: {#exm-example0105 .example name="Determine continuously compounded rate from effective annual rate"}
Suppose an investment pays a periodic interest rate of 5% every six
months ($m=2,R/2=0.05$). In the market this would be quoted as having an
annual percentage rate, $R_{A}$, of 10%. An investment of \$100 yields
$\$100\cdot(1.05)^{2}=\$110.25$ after one year. The effective annual
rate, $R_{A}$, is then 10.25%. To find the continuously compounded
simple rate that gives the same future value as investing at the
effective annual rate we solve: $R=\ln(1.1025)=0.09758.$ That is, if
interest is compounded continuously at a simple annual rate of 9.758%
then \$100 invested today would grow to
$\$100\cdot e^{0.09758}=\$110.25$
:::

\hfill{}$\blacksquare$

## Asset Return Calculations {#sec-AssetReturnCalculations}

In this section, we review asset return calculations given initial and
future prices associated with an investment. We first summarize the main
type of assets used for investment purposes. Next, we cover simple
return calculations, which are typically reported in practice but are
often not convenient for statistical modeling purposes. We then describe
continuously compounded return calculations, which are more convenient
for statistical modeling purposes. We finish the section with a
discussion of portfolio return calculations.

### Assets

When we talk about assets in this book, we mean the typical investment
assets available for purchase to individuals through brokerage and
retirement accounts. These include bonds, stocks, mutual funds, exchange
traded funds and notes, and derivative securities. Investment assets can
be purchased through online brokerage accounts at firms like Ally
Invest, Charles Schwab, E\*TRADE, Fidelity, Interactive Brokers, Merrill
Lynch, Robinhood, TD Ameritrade, and Vanguard. Due to the growth and
competition of online investment accounts, the cost of purchasing
investment assets has dropped significantly over time.

*Bonds*

Bonds are assets that represent debt obligations from corporations or
the government used to raise capital (money) to finance operations. A
typical bond is characterized by (1) a principal amount, representing
initial capital borrowed; (2) maturity date, representing when the
principal must be repaid, (3) periodic coupon payments, representing
interest paid on specific dates. Bonds issued by the U.S. government are
called *Treasury bonds*, and are generally thought to be the safest
bonds as it is very unlikely that the U.S. government will default on
its debt obligations. Other government agencies as well as states and
municipalities issue bonds to raise money. Corporate bonds are bonds
issued by private and public corporations. Bonds other than Treasury
bonds have the risk, called *default risk*, that the principal or coupon
payments many not be made.

*Stocks*

A *stock* represents an ownership share in corporation. Public stock
represents ownership in a publicly traded corporation, where the shares
can be purchased or sold (traded) by the public in organized stock
exchanges (markets). In the U.S. the main stock exchanges are the New
York Stock Exchange (NYSE) and the National Association of Securities
Dealers Automated Quotations (NASDAQ) exchange. Private stock represents
ownership in a private corporation whose stock is not traded in an
exchange. Private corporations become public companies when they decide
to raise money (capital) by issuing new stock in a public exchange
through an *initial public offering* (IPO). This new stock purchased by
individuals can then be sold to others in the exchange. Ownership of
stock often provides voting rights which allows stock holders to have a
say in the management of the company. Owners of stock have a right to
receive dividends, which are cash payments representing corporate
profits. Corporations may or many not pay dividends. Typically, young
growing companies reinvest their profits instead of paying dividends.
The stocks of these companies are called *growth stocks*. More
established companies more often pay dividends instead of reinvesting
all profits.

*Mutual Funds*

A *mutual fund* is a professionally managed investment fund that pools
money from many investors to purchase a collection of assets called a
*portfolio*. Mutual funds typically invest in bonds, stocks, and money
market instruments. Mutual funds are commonly used in *individual
retirement accounts* (IRAs) and employee sponsored retirement accounts.
Because mutual funds are managed portfolios they charge fees for this
service. The amount of the fee depends on the activity of the portfolio
manager. Actively managed funds, where the portfolio manager actively
picks the assets in the portfolio, have the highest fees. Actively
managed mutual funds are offered by many firms such as the Capital
Group, Fidelity, Janice, Putnam, and T. Rowe Price. Passively managed
funds have lower fees because the managers follow a rule to match a
particular asset index such as the S&P 500 index. Passive mutual funds
were pioneered by Vandguard. Mutual funds are traded on exchanges but
are only traded at the end of each day.

*Exchange Traded Funds and Notes*

An *exchange traded fund* (ETF) is similar in many ways to a mutual
fund, except that an ETF can be bought and sold throughout the day on
stock exchanges. Most ETFs are index funds, that is, they hold the same
securities in the same proportions as a certain stock market index or
bond market index. The most popular ETFs in the U.S. replicate the S&P
500 Index, the total market index, the NASDAQ-100 index, the price of
gold, the "growth" stocks in the Russell 1000 Index, or the index of the
largest technology companies. ETNs are similar to ETFs. The main
companies issuing ETFs and ETNs are Barclays Capital Inc. (iPath),
BlackRock (iShares), Credit Swisse, Direxion, StateStreet (SPDR),
Invesco (PowerShares), and ProShares. Each of these companies provide
online information on their ETFs and ETNs. You can easily find ETFs and
ETNs for particular asset categories and investment strategies at
www.etf.com.

*Derivative Securities*

A *derivative security* is an asset whose value is derived from the
behavior of other underlying assets such as bonds, stocks, ETFs, etc.
Common derivative securities are forwards, futures, options and swaps.
We will not consider derivative securities is this book.

Information and historical price data on many publicly traded assets is
freely available at https://finance.yahoo.com/.

### Simple returns

\index{Simple returns} Consider purchasing one share of an asset (e.g.,
stock, bond, ETF, mutual fund, option, etc.) at time $t_{0}$ for the
price $P_{t_{0}}$, and then selling that share at time $t_{1}$ for the
price $P_{t_{1}}.$ If there are no intermediate cash flows (e.g.,
dividends) between $t_{0}$ and $t_{1}$, the rate of return over the
period $t_{0}$ to $t_{1}$ is the percentage change in price:
$$
R(t_{0},t_{1})=\frac{P_{t_{1}}-P_{t_{0}}}{P_{t_{0}}}.
$$ {#eq-holding-period-return} The time between $t_{0}$ and $t_{1}$ is called the
*holding period* \index{Holding period} and equation
@eq-holding-period-return is called the *holding period
return*\index{Holding period return}. In principle, the holding period
can be any amount of time: one second, five minutes, eight hours, two
days six minutes and two seconds, fifteen years, etc.. To simplify
matters, in this chapter we will assume that the holding period is some
increment of calendar time; e.g., one day, one month or one year. In
particular, we will assume a default holding period of one month in what
follows.

Let $P_{t}$ denote the price at the end of month $t$ of an asset that
pays no dividends and let $P_{t-1}$ denote the price at the end of month
$t-1$. Then the one-month *simple net return*\index{Simple net return}
on an investment in the asset between months $t-1$ and $t$ is defined
as: $$
R_{t}=\frac{P_{t}-P_{t-1}}{P_{t-1}}=\Delta P_{t}.
$$ {#eq-simple-net-return} Writing $(P_{t}-P_{t-1})/P_{t-1}=P_{t}/P_{t-1}-1$, we can
define the *simple gross return*\index{Simple gross return} as:
$$
1+R_{t}=\frac{P_{t}}{P_{t-1}}.
$$ {#eq-simple-gross-return} The one-month gross return has the interpretation of the
future value of \$1 invested in the asset for one-month with simple
interest rate $R_{t}$. Solving @eq-simple-gross-return for $P_{t}$ gives
$$P_{t}=P_{t-1}(1+R_{t}),$$
which shows $P_{t}$ can be thought of as the future value of $P_{t-1}$
invested for one month with simple return $R_{t}.$ Unless otherwise
stated, when we refer to returns we mean net returns. Since asset prices
must always be non-negative (a long position in an asset is a limited
liability investment), the smallest value for $R_{t}$ is -1 or -100%.

::: {#exm-example0106 .example name="Simple return calculation"}
Consider a one-month investment in Microsoft stock. Suppose you buy the
stock in month $t-1$ (e.g., end of February) at $P_{t-1}=\$85$ and sell
the stock the next month (e.g., end of March) for $P_{t}=\$90.$ Further
assume that Microsoft does not pay a dividend between months $t-1$ and
$t$. The one-month simple net and gross returns (over the month of
March) are then: \begin{align*}
R_{t} & =\frac{\$90-\$85}{\$85}=\frac{\$90}{\$85}-1=1.0588-1=0.0588,\\
1+R_{t} & =1.0588.
\end{align*} The one-month investment in Microsoft yielded a 5.88% per
month return. Alternatively, \$1 invested in Microsoft stock in month
$t-1$ grew to \$1.0588 in month $t$. The calculations using R are:
:::

```{r}
#| echo: true

P0 = 90 
Pm1 = 85 
R0 = (P0 - Pm1)/Pm1 
c(R0,1 + R0)
```

\hfill $\blacksquare$

The return calculation works the same way if multiple shares of an asset
are initially purchased. Let $s \ge 1$ denote the number of shares of an
asset purchased at time $t-1$. The initial value of the investment is
then $V_{t-1}=s \times P_{t-1}$. The value of the investment at time $t$
is $V_{t} = s \times P_{t}.$ The rate of return on the investment is the
percentage change in value over the holding period: 
\begin{align}
R_{t} &= \frac{V_{t} - V_{t-1}}{V_{t-1}} = \frac{s \times P_{t} - s \times P_{t-1}}{s \times P_{t-1}} \\
&= \frac{s \times(P_{t} - P_{t-1})}{s \times P_{t-1}} = \frac{P_{t} - P_{t-1}}{P_{t-1}}.
\end{align}

As a result, the future value of the investment is equal to the current
value times the gross return:
$$V_{t} = V_{t-1} \times (1 + R_{t})$$

#### Multi-period returns

\index{Multi-period returns} The simple two-month return on an
investment in an asset between months $t-2$ and $t$ is defined as:
$$R_{t}(2)=\frac{P_{t}-P_{t-2}}{P_{t-2}}=\frac{P_{t}}{P_{t-2}}-1.$$
Writing $P_{t}/P_{t-2}=(P_{t}/P_{t-1})\cdot (P_{t-1}/P_{t-2})$ the
two-month return can be expressed as: \begin{align*}
R_{t}(2) & =\frac{P_{t}}{P_{t-1}}\cdot\frac{P_{t-1}}{P_{t-2}}-1\\
 & =(1+R_{t})(1+R_{t-1})-1.
\end{align*} Then the simple two-month gross return becomes:
$$1+R_{t}(2)=(1+R_{t})(1+R_{t-1})=1+R_{t-1}+R_{t}+R_{t-1}R_{t},$$
which is the product of the two simple one-month gross returns, called
the *geometric average*, and not one plus the sum of the two one-month
returns. Hence, the two-month net return is
$$R_{t}(2)=R_{t-1}+R_{t}+R_{t-1}R_{t}.$$
If, however, $R_{t-1}$ and $R_{t}$ are close to zero then
$R_{t-1}R_{t}\approx0$ and $1+R_{t}(2)\approx1+R_{t-1}+R_{t}$ so that
$R_{t}(2)\approx R_{t-1}+R_{t}$.

::: {#exm-example0107 .example name="Computing two-period returns"}
Continuing with the previous example, suppose that the price of
Microsoft stock in month $t-2$ is $\$80$ and no dividend is paid between
months $t-2$ and $t.$ The two-month net return is: 
$$
R_{t}(2)=\frac{\$90-\$80}{\$80}=\frac{\$90}{\$80}-1=1.1250-1=0.1250, 
$$
or 12.50% per two months. The two one-month returns are: 
\begin{align*}
R_{t-1} & =\frac{\$85-\$80}{\$80}=1.0625-1=0.0625,\\
R_{t} & =\frac{\$90-85}{\$85}=1.0588-1=0.0588,
\end{align*} 
and the geometric average of the two one-month gross returns is: $1+R_{t}(2)=1.0625 \times 1.0588=1.1250.$

The calculations can be *vectorized* in R as follows:
:::

```{r}
#| echo: true

Pm2 = 80 
# create vector of prices
P = c(Pm2, Pm1, P0) 
# calculate vector of 1-month returns from vector of prices 
R = (P[2:3] - P[1:2])/P[1:2] 
R  
# calculate 2-month return from 2 1-month returns 
(1 + R[1])*(1 + R[2]) - 1 
# same calculation vectorized using R function cumprod() 
cumprod(1+R) - 1
```

In the above, `P[2:3]` represents the 2nd and 3rd elements of `P`, and
`P[1:2]` represents the 1st and 2nd elements of `P`.

\hfill{}$\blacksquare$

It is important to recognize that two-period simple returns are not
symmetric in the following sense. Suppose $R_{t-1}=-0.5$ and $R_{t}=0.5$
and initial wealth is \$1. After two-periods \$1 falls to
$$\$1 \times (1+R_{t-1})(1+R_{t}) = \$1 \times (1-0.5)(1+0.5) = \$0.75$$
That is, a 50% drop followed by a 50% gain does not return wealth to its
initial level. In the second period, the return has to be 100% in order
for wealth to return to its initial level.

In general, the $k$-month gross return is defined as the product of $k$
one-month gross returns: $$
\begin{align}1+R_{t}(k) & =(1+R_{t})(1+R_{t-1})\cdots(1+R_{t-k+1})\\
 & =\prod_{j=0}^{k-1}(1+R_{t-j}).\nonumber\end{align}
$$ {#eq-k-period-return}

If all returns are close to zero then
$$R_{t}(k) \approx \sum_{j=0}^{k-1} R_{t-j}$$

As with the one-month return, the multi-period return can also be
derived in the same way if multiple shares of an asset are initially
purchased. Using $V_{t-j} = s \times P_{t-j}$ we have 
\begin{align}
\frac{V_{t}}{V_{t-k}} &= \frac{V_{t}}{V_{t-1}} \times \frac{V_{t-1}}{V_{t-2}} \ldots \frac{V_{t-k+1}}{V_{t-k}} \\
&= \frac{s\times P_{t}}{s\times P_{t-1}} \times \frac{s\times P_{t-1}}{s\times P_{t-2}} \ldots \frac{s\times P_{t-k+1}}{s\times P_{t-k}} \\
&= (1+R_{t})(1+R_{t-1})\cdots(1+R_{t-k+1}).
\end{align} 
As a result, $V_{t}$ can be interpreted as the future value
of $V_{t-k}$ based on the returns $R_{t-k+1}, \cdots, R_{t}$:
$$V_{t} = V_{t-k} \times (1+R_{t-k+1})(1+R_{t-k+2}) \cdots (1+R_{t}).$$

#### Adjusting for dividends

If an asset pays a *dividend*\index{Dividend} (or any cash payment),
$D_{t}$, sometime between months $t-1$ and $t$, the *total net
return*\index{Total net return} calculation becomes: $$
R_{t}^{\textrm{total}}=\frac{P_{t}+D_{t}-P_{t-1}}{P_{t-1}}=\frac{P_{t}-P_{t-1}}{P_{t-1}}+\frac{D_{t}}{P_{t-1}},
$$ {#eq-dividend-adjusted-return} where $(P_{t}-P_{t-1})/P_{t-1}$ is referred as the
*capital gain* \index{Capital gain} and $D_{t}/P_{t-1}$ is referred to
as the *dividend yield*\index{Dividend yield}. The *total gross
return*\index{Total gross return} is: $$
1+R_{t}^{\textrm{total}}=\frac{P_{t}+D_{t}}{P_{t-1}}.
$$ {#eq-dividend-adjusted-gross-return} The formula @eq-k-period-return for computing
multi-period return remains the same except that one-period gross
returns are computed using @eq-dividend-adjusted-gross-return.

High dividend paying stocks are often called *income stocks* and have
characteristics similar to bonds where the regular dividends play a
similar role as coupon interest payments.

::: {#exm-example0109 .example name="Compute total return when dividends are paid"}
Consider a one-month investment in Microsoft stock. Suppose you buy the
stock in month $t-1$ at $P_{t-1}=\$85$ and sell the stock the next month
for $P_{t}=\$90$. Further assume that Microsoft pays a \$1 dividend
between months $t-1$ and $t$. The capital gain, dividend yield and total
return are then: \begin{align*}
R_{t} & =\frac{\$90+\$1-\$85}{\$85}=\frac{\$90-\$85}{\$85}+\frac{\$1}{\$85}\\
 & =0.0588+0.0118\\
 & =0.0707.
\end{align*} The one-month investment in Microsoft yields a 7.07% per
month total return. The capital gain component is $5.88\%$ and the
dividend yield component is 1.18%

\hfill{}$\blacksquare$
:::

#### Adjusting for inflation

The return calculations considered so far are based on the
*nominal*\index{Nominal price} or current prices of assets. Returns
computed from nominal prices are nominal returns. The *real
return*\index{Real return} on an asset over a particular horizon takes
into account the growth rate of the general price level over the
horizon. If the nominal price of the asset grows faster than the general
price level, then the nominal return will be greater than the inflation
rate and the real return will be positive. Conversely, if the nominal
price of the asset increases less than the general price level, then the
nominal return will be less than the inflation rate and the real return
will be negative. Real returns reflect the increase or decrease in
purchasing power from an investment.

The computation of real returns on an asset is a two-step process:

1. Deflate the nominal price of the asset by the general price level
2. Compute returns in the usual way using the deflated prices

To illustrate, consider computing the real simple one-period return on
an asset. Let $P_{t}$ denote the nominal price of the asset at time $t$
and let $CPI_{t}$ denote an index of the general price level (e.g.
consumer price index) at time $t$[^01-returncalculations-new-1]. The
deflated or real price at time $t$ is:
$$P_{t}^{\textrm{Real}}=\frac{P_{t}}{CPI_{t}},$$
and the real one-period return is: $$
\begin{align}R_{t}^{\textrm{Real}}
&=\frac{P_{t}^{\textrm{Real}}-P_{t-1}^{\textrm{Real}}}{P_{t-1}^{\textrm{Real}}}=\frac{\frac{P_{t}}{CPI_{t}}-\frac{P_{t-1}}{CPI_{t-1}}}{\frac{P_{t-1}}{CPI_{t-1}}} \nonumber \\
&=\frac{P_{t}}{P_{t-1}}\cdot\frac{CPI_{t-1}}{CPI_{t}}-1.\end{align}
$$ {#eq-simple-real-return} The one-period gross real return return is: $$
1+R_{t}^{\textrm{Real}}=\frac{P_{t}}{P_{t-1}}\cdot\frac{CPI_{t-1}}{CPI_{t}}.
$$ {#eq-simple-real-gross-return} If we define inflation between periods $t-1$ and $t$ as:
$$
\pi_{t}=\frac{CPI_{t}-CPI_{t-1}}{CPI_{t-1}}=\%\Delta CPI_{t},
$$ {#eq-inflation} then $1+\pi_{t}=CPI_{t}/CPI_{t-1}$ and
@eq-simple-real-return may be re-expressed as: $$
R_{t}^{\textrm{Real}}=\frac{1+R_{t}}{1+\pi_{t}}-1.
$$ {#eq-simple-real-return-2}

[^01-returncalculations-new-1]: The CPI is usually normalized to 1 or
    100 in some base year. We assume that the CPI is normalized to 1 in
    the base year for simplicity.

::: {#exm-example0110 .example name="Compute real return"}
Consider, again, a one-month investment in Microsoft stock. Suppose the
CPI in months $t-1$ and $t$ is 1 and 1.01, respectively, representing a
1% monthly growth rate in the overall price level. The real prices of
Microsoft stock are: 
$$
\begin{align*}
P_{t-1}^{\textrm{Real}}&=\frac{\$85}{1}=\$85, \\
P_{t}^{\textrm{Real}}&=\frac{\$90}{1.01}=\$89.1089, 
\end{align*}
$$
and the real monthly return is: 
$$
R_{t}^{\textrm{Real}}=\frac{\$89.10891-\$85}{\$85}=0.0483. 
$$
The nominal return and inflation over the month are: 
$$
\begin{align*}
R_{t}&=\frac{\$90-\$85}{\$85}=0.0588,\\
\pi_{t}&=\frac{1.01-1}{1}=0.01.
\end{align*}
$$ 
Then the real return computed using @eq-simple-real-return-2 is: 
$$
R_{t}^{\textrm{Real}}=\frac{1.0588}{1.01}-1=0.0483.
$$
Notice that simple real return is almost, but not quite, equal to the simple nominal return
minus the inflation rate: $R_{t}^{\textrm{Real}}\approx R_{t}-\pi_{t}=0.0588-0.01=0.0488.$

\hfill{}$\blacksquare$
:::

#### Adjusting for taxes

Any investment that generates a capital gain is potentially subject to
taxation by the government. This taxation reduces the investment return.
To see how this works, let $V_{t}$ denote an initial investment amount,
let $R_{t}>0$ denote the one-period rate of return on the investment,
and let $\tau$ denote the tax rate on investment income (capital gain).
In the U.S. the tax rate on long-term (holding period more than one
year) capital gains is either $0\%$, $15\%$, or $20\%$ (depending on
income) and the tax rate on short-term capital gains (holding period
less than one year) is the tax rate on ordinary income (which could be
as high as $35\%$). The future value without taxes is $V_{t}(1+R_{t})$,
and $V_{t}R_{t}$ is the capital gain. The amount owed in taxes is
$V_{t}R_{t}\tau$ and the amount remaining after taxes is
$V_{t}R_{t}(1 -\tau)$. The after tax return is $$
\begin{eqnarray}R_{t}^{Atax} &= \frac{V_{t}+V_{t}R_{t}(1 -\tau)-V_{t}}{V_{t}} \nonumber \\
&=\frac{V_{t}(1+R_{t}(1-\tau)-1)}{V_{t}} \nonumber \\
&= R_{t}(1-\tau).\end{eqnarray}
$$ {#eq-afterTaxReturn}

Hence, the after tax return is the before tax return multiplied by one
minus the tax rate.

::: {#exm-example0110b .example name="Compute after tax return"}
Let $V_{t}=\$100$, $R_{t}=0.05$, and $\tau=0.20$. Then the after tax
return is
$$R_{t}^{Atax} = (0.05)(1-0.20) = (0.05)(0.80) = 0.04.  $$
The amount paid in taxes is $\$1000\times(0.05)\times(0.2)=\$10$, and
the after tax profit is $\$50 - \$10 = \$40$.
:::

#### Return on foreign investment {#sec-return-on-foreign-investment}

\index{Return on foreign investment} It is common for an investor in one
country to buy an asset in another country as an investment. If the two
countries have different currencies then the exchange rate between the
two currencies influences the return on the asset. To illustrate,
consider an investment in a foreign stock (e.g. a stock trading on the
London stock exchange) by a U.S. national (domestic investor). The
domestic investor takes U.S. dollars, converts them to the foreign
currency (e.g. British Pound) via the exchange rate (price of foreign
currency in U.S. dollars) and then purchases the foreign stock using the
foreign currency. When the stock is sold, the proceeds in the foreign
currency must then be converted back to the domestic currency.

To be more precise, consider the information in the table below:

| Month | U.S. \$ Cost of 1 Pound $£$, $E_{t}^{US/UK}$ | Value of British Shares, $P_{t}^{UK}$ | Value in U.S. \$, $P_{t}^{US}$ |
|----------------|----------------------|------------------|----------------|
| t - 1 | \$ 1.50 | $£40$ | $\$1.50\times
      £40=\$60$ |
| t | \$ 1.30 | $£45$ | $\$1.30\times
    £45=\$58.5$ |

Let $P_{t}^{US}$, $P_{t}^{UK}$, and $E_{t}^{US/UK}$ denote the U.S.
dollar price of the asset, British Pound price of the asset, and
currency exchange rate between U.S. dollars and British Pounds, at the
end of month $t$, respectively. The U.S. dollar cost of one share of
British stock in month $t-1$ is
$$P_{t-1}^{US}=E_{t-1}^{US/UK}\times P_{t-1}^{UK}=\$1.50\times￡40=\$60,$$
and the U.S. dollar cost at time $t$ is
$$P_{t}^{US}=E_{t}^{US/UK}\times P_{t}^{UK}=\$1.30\times£45=\$58.5.$$
The U.S. dollar rate of return is then
$$R_{t}^{US}=\frac{P_{t}^{US}-P_{t-1}^{US}}{P_{t-1}^{US}}=\frac{\$58.5-\$60}{\$60}=-0.025.$$
Notice that $$
\begin{eqnarray}R_{t}^{US} & = & \frac{E_{t}^{US/UK}\times P_{t}^{UK}-E_{t-1}^{US/UK}\times P_{t-1}^{UK}}{E_{t-1}^{US/UK}\times P_{t-1}^{UK}} \nonumber \\
 & = & \frac{E_{t}^{US/UK}\times P_{t}^{UK}}{E_{t-1}^{US/UK}\times P_{t-1}^{UK}}-1 \nonumber \\
 & = & (1+R_{t}^{US/UK})(1+R_{t}^{UK})-1,\end{eqnarray}
$$ {#eq-domesticReturn2} where
$$\frac{E_{t}^{US/UK}}{E_{t-1}^{US/UK}}=1+R_{t}^{US/UK},\,\frac{P_{t}^{UK}}{P_{t-1}^{UK}}=1+R_{t}^{UK}.$$
Hence,
$$1+R_{t}^{US} = (1+R_{t}^{US/UK})(1+R_{t}^{UK}).$$
Here, $R_{t}^{US/UK}$ is the simple rate of return on foreign currency
and $R_{t}^{UK}$ is the simple rate of return on the British stock in
British Pounds: \begin{eqnarray*}
R_{t}^{US/UK} & = & \frac{E_{t}^{US/UK}-E_{t-1}^{US/UK}}{E_{t-1}^{US/UK}}=\frac{\$1.30-\$1.50}{\$1.50}=-0.133,\\
R_{t}^{UK} & = & \frac{P_{t}^{UK}-P_{t-1}^{UK}}{P_{t-1}^{UK}}=\frac{£45-£40}{£40}=0.125.
\end{eqnarray*} Hence, the gross return, $1+R_{t}^{US}$, for the U.S.
investor has two parts: (1) the gross return from an investment in
foreign currency, $1+R_{t}^{US/UK}$ ; and (2) the gross return of the
foreign asset in the foreign currency, $1+R_{t}^{UK}:$ \begin{eqnarray*}
1+R_{t}^{US} & = & (1+R_{t}^{US/UK})(1+R_{t}^{UK})\\
 & = & (0.867)(1.125)\\
 & = & 0.975.
\end{eqnarray*} In this example, the U.S. investor loses money even
though the British investment has a positive rate of return because of
the large drop in the exchange rate.

#### Annualizing returns

Very often returns over different horizons are annualized, i.e.,
converted to an annual return, to facilitate comparisons with other
investments. The annualization process depends on the holding period of
the investment and an implicit assumption about compounding. We
illustrate with several examples using monthly returns. The same
techniques work with different holding periods.

To start, if our investment horizon is one year then the annual gross
and net returns are computed directly from the 12 one-month returns:
\begin{align*}
1+R_{A} & =1+R_{t}(12)=\frac{P_{t}}{P_{t-12}}=(1+R_{t})(1+R_{t-1})\cdots(1+R_{t-11}),\\
R_{A} & =R_{t}(12).
\end{align*} 
In this case, no compounding is required to create an
annual return.

Next, consider a one-month investment in an asset with return $R_{t}$.
What is the annualized return on this investment? If we assume that we
receive the same return $R=R_{t}$ every month for the year, then the
gross annual return is:
$$1+R_{A}=1+R_{t}(12)=(1+R)^{12}.$$
That is, the *annual gross return* \index{Annual gross return} is
defined as the assumed common monthly return compounded for 12 months.
The net annual return is then:
$$R_{A}=(1+R)^{12}-1.$$
It is important to recognize that this annualization calculation assumes
the same monthly return $R$ for each of the 12 months. This can be
highly inaccurate and misleading if the actual future monthly returns
are much different than $R$. The following examples illustrate.

::: {#exm-example0111 .example name="Compute annualized return from one-month return"}
In the first example, the one-month return, $R_{t}$, on Microsoft stock
was 5.88%. If we assume that we can get this return for 12 months then
the annualized return is: $R_{A}=(1.0588)^{12}-1=1.9850-1=0.9850,$ 
or 98.50% per year. Pretty good! However, this is likely to be highly
inaccurate since getting 5\% every month for a year is highly unlikely
for a risky stock like Microsoft.

\hfill{}$\blacksquare$
:::

Now, consider a two-month investment with return $R_{t}(2).$ If we
assume that we receive the same two-month return $R(2)=R_{t}(2)$ for the
next six two-month periods, then the gross and net annual returns are:
\begin{align*}
1+R_{A} & =(1+R(2))^{6},\\
R_{A} & =(1+R(2))^{6}-1.
\end{align*} Here the annual gross return is defined as the two-month
return compounded for 6 months.

::: {#exm-example0112 .example name="Compute annualized return from two-month return"}
Suppose the two-month return, $R_{t}(2),$ on Microsoft stock is 12.5%.
If we assume that we can get this two-month return for the next 6
two-month periods then the annualized return is: $R_{A}=(1.1250)^{6}-1=2.0273-1=1.0273,$ 
or 102.73% per year. As with the previous example, this result is likely to be inaccurate 
as it is highly unlikely to get 12.5% every two months for the rest of the
year.

\hfill{}$\blacksquare$
:::

As the last example, suppose that our investment horizon is two years.
That is, we start our investment at time $t-24$ and cash out at time
$t$. The two-year gross return is then $1+R_{t}(24)=P_{t}/P_{t-24}$.
What is the annual return on this two-year investment? The process is
the same as computing the effective annual rate. To determine the annual
return we solve the following relationship for $R_{A}$: 
\begin{align*}
(1+R_{A})^{2} & =1+R_{t}(24)\rightarrow\\
R_{A} & =\left(1+R_{t}(24)\right)^{1/2}-1.
\end{align*} 
In this case, the annual return is compounded twice to get
the two-year return and the relationship is then solved for the annual
return.

::: {#exm-example0113 .example name="Compute annualized return from two-year return"}
Suppose that the price of Microsoft stock $24$ months ago is
$P_{t-24}=\$50$ and the price today is $P_{t}=\$90$. The two-year gross
return is $1+R_{t}(24)=\$90/\$50=1.8000$ which yields a two-year net
return of $R_{t}(24)=0.80=80\%$. The annual return for this investment
is defined as: $R_{A}=(1.800)^{1/2}-1=1.3416-1=0.3416$ or 34.16%
per year.

\hfill{}$\blacksquare$
:::

#### Average returns

For investments over a given horizon, it is often of interest to compute
a measure of the average rate of return over the horizon. To illustrate,
consider a sequence of monthly investments over $n$ months with monthly
returns $R_{1},R_{2},\ldots,R_{n}$. The $n-$month return is

$$R(n)=(1+R_{1})(1+R_{2})\cdots(1+R_{n})-1.$$
What is the average monthly return? There are two possibilities. The
first is the *arithmetic average
return*\index{Arithmetic average return}:
$$\bar{R}^{Arith}=\frac{1}{n}\left(R_{1}+R_{2}+\cdots+R_{n}\right).$$

The second is the *geometric average
return*\index{Geometric average return}:
$$\bar{R}^{Geo}=\left[(1+R_{1})(1+R_{2})\cdots(1+R_{n})\right]^{1/n}-1.$$
Notice that the geometric average return is the monthly return which
compounded monthly for $n$ months gives the gross $n-$month return
$$(1+\bar{R}^{Geo})^{n}=1+R(n).$$
As a measure of average investment performance, the geometric average
return is preferred to the arithmetic average return. To see why,
consider a two-month investment horizon with monthly returns $R_{1}=0.5$
and $R_{2}=-0.5$. One dollar invested over these two months grows to
$$
(1+R_{1})(1+R_{2})=(1.5)(0.5)=0.75,
$$
for a two-period return of $R(2)=0.75-1=-0.25$. Hence, the two-month investment 
loses 25%. The arithmetic average return is
$$\bar{R}^{Arith}=\frac{1}{2}\left(0.5+(-0.5)\right)=0.$$
As a measure of investment performance, this is misleading because the
investment loses money over the two-month horizon. The average return
indicates the investment breaks even. The geometric average return is
$$\bar{R}^{Geo}=\left[(1.5)(0.5)\right]^{1/2}-1=-0.134.$$
This is a more accurate measure of investment performance because it
indicates that the investment eventually loses money. In addition, the
compound two-month return using $\bar{R}^{Geo}=-0.134$ is equal to
$R(2)=-0.25.$

While the arithmetic average return may not be the most appropriate
measure of actual average investment performance, we shall see in the
following chapters that it is a useful estimate of expected performance.

### Continuously compounded returns

\index{Continuously compounded returns} In this section we define
continuously compounded returns from simple returns, and describe their
properties.

#### One-period returns

Let $R_{t}$ denote the simple monthly return on an investment. The
*continuously compounded monthly return*, $r_{t}$, is defined as:
$$
r_{t}=\ln(1+R_{t})=\ln\left(\frac{P_{t}}{P_{t-1}}\right),
$$ {#eq-cc-return} where $\ln(\cdot)$ is the natural log
function[^01-returncalculations-new-2]. To see why $r_{t}$ is called the
continuously compounded return, take the exponential of both sides of
@eq-cc-return to give:
$$e^{r_{t}}=1+R_{t}=\frac{P_{t}}{P_{t-1}}.$$
Rearranging we get $P_{t}=P_{t-1}e^{r_{t}}=P_{t-t}(1+R_{t})$, so that
$r_{t}$ is the continuously compounded growth rate in prices between
months $t-1$ and $t$. This is to be contrasted with $R_{t}$, which is
the simple growth rate in prices between months $t-1$ and $t$ without
any compounding. As a result, $r_{t}$ is always smaller than $R_{t}$.
Furthermore, since $\ln\left(x/y\right)=\ln(x)-\ln(y)$ it follows that:
\begin{align*}
r_{t} & =\ln\left(\frac{P_{t}}{P_{t-1}}\right)\\
& =\ln(P_{t})-\ln(P_{t-1})\\
& =p_{t}-p_{t-1},
\end{align*} where $p_{t}=\ln(P_{t})$. Hence, the continuously
compounded monthly return, $r_{t}$, can be computed simply by taking the
first difference of the natural logarithms of monthly prices.

[^01-returncalculations-new-2]: The continuously compounded return is
    always defined since asset prices, $P_{t}$, are always non-negative.
    Properties of logarithms and exponentials are discussed in the
    appendix to this chapter.

If the simple return $R_{t}$ is close to zero ($R_{t}\approx0$) then
$r_{t}=\mathrm{ln}(1+R_{t})\approx R_{t}$. This result can be shown
using a first order Taylor series approximation to the function
$f(x)=\mathrm{ln}(1+x).$ Recall, for a continuous and differentiable
function $f(x)$ the first order Taylor series approximation of $f(x)$
about the point $x=x_{0}$ is
$$f(x)=f(x_{0})+f^{\prime}(x_{0})(x-x_{0})+\mathrm{remainder},$$
where $f^{\prime}(x_{0})$ denotes the derivative of $f(x)$ evaluated at
$x_{0}.$ Let $f(x)=\mathrm{ln}(1+x)$ and set $x_{0}=0.$ Note that
$$f(0)=\mathrm{ln}(1)=0,\,f^{\prime}(x)=\frac{1}{1+x},\,f^{\prime}(0)=1.$$
Then, using the first order Taylor series approximation,
$\mathrm{ln}(1+x)\approx x$ when $x\approx0.$

::: {#exm-example0114 .example name="Compute continuously compounded returns"}
Using the price and return data from Example @exm-example0106, the
continuously compounded monthly return on Microsoft stock can be
computed in two ways: \begin{eqnarray*}
r_{t} & = & \mathrm{ln}(1.0588)=0.0571\\
r_{t} & = & \mathrm{\ln}(90)-\ln(85)=4.4998-4.4427=0.0571
\end{eqnarray*} Notice that $r_{t}$ is slightly smaller than $R_{t}$.
Using R, the calculations are
:::

```{r}
#| echo: true

c(log(1 + R0), log(P0) - log(Pm1))
```

Notice that the R function `log()` computes the natural logarithm.

\hfill{}$\blacksquare$

Given a monthly continuously compounded return $r_{t}$, it is
straightforward to solve back for the corresponding simple net return
$R_{t}$: $$
R_{t}=e^{r_{t}}-1.
$$ {#eq-simpleReturnFromCCreturn} Hence, nothing is lost by considering continuously
compounded returns instead of simple returns. Continuously compounded
returns are very similar to simple returns as long as the return is
relatively small, which it generally will be for monthly or daily
returns. Since $R_{t}$ is bounded from below by -1, the smallest value
for $r_{t}$ is $r_{t} = ln(1-1) = -\infty$. This, however, does not mean
that you could lose an infinite amount of money on an investment. The
actual amount of money lost is determined by the simple return
@eq-simpleReturnFromCCreturn.

::: {#exm-example0115 .example name="Determine simple returns from continuously compounded returns"}
In the previous example, the continuously compounded monthly return on
Microsoft stock is $r_{t}=5.71\%$. The simple net return is then: $R_{t}=e^{.0571}-1=0.0588.$

\hfill{}$\blacksquare$
:::

For asset return modeling and statistical analysis it is often more
convenient to use continuously compounded returns than simple returns.
One reason is that continuously compounded returns take values on the
entire real line, $-\infty < r_{t} < \infty$, whereas simple returns are
only defined for values larger than -1, $-1 \le R_{t} < \infty$. Another
reason is due to the additivity property of multi-period continuously
compounded returns discussed in the next sub-section.

#### Multi-period returns

The relationship between multi-period continuously compounded returns
and one-period continuously compounded returns is more simple than the
relationship between multi-period simple returns and one-period simple
returns. To illustrate, consider the two-month continuously compounded
return defined as:
$$r_{t}(2)=\ln(1+R_{t}(2))=\ln\left(\frac{P_{t}}{P_{t-2}}\right)=p_{t}-p_{t-2}.$$
Taking exponentials of both sides and rearranging gives:
$$P_{t}=P_{t-2}e^{r_{t}(2)}$$
so that $r_{t}(2)$ is the continuously compounded growth rate of prices
between months $t-2$ and $t$. Using
$P_{t}/P_{t-2}=(P_{t}/P_{t-1})\cdot(P_{t-1}/P_{t-2})$ and the fact that
$\ln(x\cdot y)=\ln(x)+\ln(y)$ it follows that: \begin{align*}
r_{t}(2) & =\ln\left(\frac{P_{t}}{P_{t-1}}\cdot\frac{P_{t-1}}{P_{t-2}}\right)\\
 & =\ln\left(\frac{P_{t}}{P_{t-1}}\right)+\ln\left(\frac{P_{t-1}}{P_{t-2}}\right)\\
 & =r_{t}+r_{t-1}.
\end{align*} Hence the continuously compounded two-month return is just
the sum of the two continuously compounded one-month returns. Recall,
with simple returns the two-month return is a multiplicative (geometric)
sum of two one-month returns.

::: {#exm-example0116 .example name="Compute two-month continuously compounded returns"}
Using the data from @exm-example0107, the continuously compounded
two-month return on Microsoft stock can be computed in two equivalent
ways. The first way uses the difference in the logs of $P_{t}$ and
$P_{t-2}$: 
$$
r_{t}(2)=\ln(90)-\ln(80)=4.4998-4.3820=0.1178.
$$  
The second way uses the sum of the two continuously compounded one-month
returns. Here $r_{t}=\ln(90)-\ln(85)=0.0571$ and
$r_{t-1}=\ln(85)-\ln(80)=0.0607$ so that: 
$$
r_{t}(2)=0.0571+0.0607=0.1178. 
$$
Notice that $r_{t}(2)=0.1178<R_{t}(2)=0.1250$. The vectorized calculations in R are:
:::

```{r}
#| echo: true

p = log(P) 
r = log(1 + R) 
c(r[1] + r[2], sum(r), p[3] - p[1], diff(p, 2))
```

\hfill{}$\blacksquare$

Previously, we saw that simple two-period returns were not symmetric.
Continuously compounded two-period returns, however, are symmetric. To
see this, suppose $r_{t-1}=-0.5$ and $r_{t}=0.5$. Notice that the
implied simple returns are not symmetric: $R_{t-1}=e^{-0.5}-1=-0.3935$
and $R_{t}=e^{0.5}-1=0.6487$. Then one dollar invested over two periods
becomes:
$$\$1 \times (1-0.3935)(1+0.6487) = \$1$$

The continuously compounded $k$-month return is defined by:
$$r_{t}(k)=\ln(1+R_{t}(k))=\ln\left(\frac{P_{t}}{P_{t-k}}\right)=p_{t}-p_{t-k}.$$
Using similar manipulations to the ones used for the continuously
compounded two-month return, we can express the continuously compounded
$k$-month return as the sum of $k$ continuously compounded monthly
returns: $$
r_{t}(k)=\sum_{j=0}^{k-1}r_{t-j}.
$$ {#eq-multiperiod-cc-return} The additivity of continuously compounded returns to form
multiperiod returns is an important property for statistical modeling
purposes.

#### Adjusting for dividends

The continuously compounded one-period return adjusted for dividends is
defined by @eq-cc-return, where $R_{t}$ is computed using
@eq-dividend-adjusted-return[^01-returncalculations-new-3].

[^01-returncalculations-new-3]: Show formula from Campbell, Lo and
    MacKinlay.

::: {#exm-example0118 .example name="Compute continuously compounded total return"}
From example @exm-example0109, the total simple return using
@eq-dividend-adjusted-return is $R_{t}=0.0707.$ The continuously
compounded total return is then: 
$$
r_{t}=\ln(1+R_{t})=\ln(1.0707)=0.0683.
$$

\hfill{}$\blacksquare$
:::

#### Adjusting for inflation

Adjusting continuously compounded nominal returns for inflation is
particularly simple. The continuously compounded one-period real return
is defined as: $$
r_{t}^{\textrm{Real}}=\ln(1+R_{t}^{\textrm{Real}}).
$$ {#eq-cc-return-real} Using @eq-simple-real-gross-return, it follows that:
$$
\begin{align}r_{t}^{\textrm{Real}} &
=\ln\left(\frac{P_{t}}{P_{t-1}}\cdot\frac{CPI_{t-1}}{CPI_{t}}\right)\\
&=\ln\left(\frac{P_{t}}{P_{t-1}}\right)+\ln\left(\frac{CPI_{t-1}}{CPI_{t}}\right)\nonumber\\
& =\ln(P_{t})-\ln(P_{t-1})+\ln(CPI_{t-1})-\ln(CPI_{t})\nonumber\\
&=\ln(P_{t})-\ln(P_{t-1})-\left(\ln(CPI_{t})-\ln(CPI_{t-1})\right)\nonumber\\
& =r_{t}-\pi_{t}^{c},\nonumber\end{align}
$$ {#eq-cc-return-real-2} where $r_{t}=\ln(P_{t})-\ln(P_{t-1})=\ln(1+R_{t})$ is the
nominal continuously compounded one-period return and
$\pi_{t}^{c}=\ln(CPI_{t})-\ln(CPI_{t-1})=\ln(1+\pi_{t})$ is the
one-period continuously compounded growth rate in the general price
level (continuously compounded one-period inflation rate). Hence, the
real continuously compounded return is simply the nominal continuously
compounded return minus the the continuously compounded inflation rate.

::: {#exm-example0119 .example name="Compute continuously compounded real returns"}
From Example @exm-example0110, the nominal simple return is
$R_{t}=0.0588$, the monthly inflation rate is $\pi_{t}=0.01$, and the
real simple return is $R_{t}^{\textrm{Real}}=0.0483.$ Using
@eq-cc-return-real, the real continuously compounded return is: 
$$
r\_{t}^{^\textrm{Real}}=\ln(1+R\_{t}{\textrm{Real}})=\ln(1.0483)=0.047.
$$ 
Equivalently, using (@eq-cc-return-real-2) the real return may also
be computed as: 
$$
r_{t}^{\textrm{Real}}=r_{t}-\pi_{t}{c}=\ln(1.0588)-\ln(1.01)=0.047.
$$

\hfill{}$\blacksquare$
:::

#### Adjusting for taxes

Using @eq-afterTaxReturn, the continuously compounded return adjusted
for taxes is \begin{equation}
r_{t}^{Atax} = ln \left(R_{t}(1-\tau) \right) = ln(R_{t}) + ln(1-\tau) = r_{t} + ln(1-\tau)
\end{equation}

#### Return on a foreign investment

Consider again, the US/UK investment example from sub-section
@sec-return-on-foreign-investment. Using @eq-domesticReturn2, the
continuously compounded US return is \begin{align}
r_{t}^{US} &= ln \left( (1+R_{t}^{US/UK})(1+R_{t}^{UK}) \right) \nonumber \\
&= ln(1+R_{t}^{US/UK}) + ln(1+R_{t}^{UK}) \nonumber \\
&= r_{t}^{US/UK} + r_{t}^{UK}
\end{align}

where $r_{t}^{US/UK} = ln(1+R_{t}^{US/UK})$ is the continuously
compounded currency return and $r_{t}^{UK} = ln(1+R_{t}^{UK})$ is the
continuously compounded UK return.

#### Annualizing continuously compounded returns

Just as we annualized simple monthly returns, we can also annualize
continuously compounded monthly returns. We illustrate using monthly
continuously compounded returns. First, if our investment horizon is one
year and we observe 12 monthly continuously compounded returns then the
annual continuously compounded return is just the sum of the twelve
monthly continuously compounded returns: \begin{align*}
r_{A} & =r_{t}(12)=r_{t}+r_{t-1}+\cdots+r_{t-11}\\
 & =\sum_{j=0}^{11}r_{t-j}.
\end{align*} The average continuously compounded monthly return is
defined as:
$$\overline{r}_{m}=\frac{1}{12}\sum_{j=0}^{11}r_{t-j}.$$
Notice that:
$$12\cdot\overline{r}_{m}=\sum_{j=0}^{11}r_{t-j}$$
so that we may alternatively express $r_{A}$ as:
$$r_{A}=12\cdot\overline{r}_{m}.$$
That is, the continuously compounded annual return is twelve times the
average of the continuously compounded monthly returns.

Next, consider a one-month investment in an asset with continuously
compounded return $r_{t} = r$. If we assume that we receive the same
return $r$ every month for the year, then the annual continuously
compounded return is just 12 times the monthly continuously compounded
return:
$$r_{A}=r_{t}(12)=r + r + \ldots + r = 12\cdot r.$$
If we have a two-month continuously compounded return of $r(2)$, and we
assume that we get this same return every two months for the next year,
then the annualized continuously compounded return is
$$r_{A} = 6\cdot r(2).$$

Finally, suppose we have a 2-year continuously compounded return
$r(24)$. Then the continuously compounded annual return can be defined
using
$$2\cdot r_{A} = r(24) \Longrightarrow  r_{A} = \frac{r(24)}{2}.$$

## Portfolios and Portfolio Returns {#sec-portfolios-and-portfolio-returns}

\index{Portfolio returns} Consider an investment in two assets named
asset $A$ and asset $B$ (e.g. Amazon and Boeing stock). Let $V_{A}$ and
$V_{B}$ denote the dollar amounts invested in assets A and B,
respectively. We call such an investment in multiple assets a
*portfolio*. The total dollar amount invested in the portfolio is
$V_p = V_A + V_B$. The shares of dollars invested in assets $A$ and $B$
are:
$$x_A = \frac{V_A}{V_p}, \space x_B = \frac{V_B}{V_p}.$$
Notice that $x_A + x_B = 1$ by construction ($100\%$ of all wealth is
invested in the two assets).

In this book, the collection of investment shares $(x_{A},x_{B})$ and
initial wealth invested $V_p$ defines a portfolio. If $V_p$ is not
specified then assume that $V_p=\$1$. For example, one portfolio may be
$(x_{A}=0.5,x_{B}=0.5)$ (equally weighted portfolio) and another may be
$(x_{A}=0,x_{B}=1)$ (everything invested in asset B). Negative values
for $x_{A}$ or $x_{B}$ are possible and represent short sales (to be
discussed in later chapters). Given $x_A$, $x_B$, and $V_p$, the dollar
amounts invested in assets $A$ and $B$ are:
$$V_A =  x_{A} \times V_p, V_B = x_{B} \times V_{p}.$$

The rate of return on a portfolio can be represented as a share weighted
average of the rates of returns on the assets in the portfolio. To see
this, let $R_{A}$ and $R_{B}$ denote the simple one-period returns on
assets $A$ and $B$. We wish to determine the simple one-period return on
the portfolio defined by $(x_{A},x_{B})$ and initial wealth $V_p$. To do
this, note that at the end of the holding period, the investments in
assets $A$ and $B$ are worth $V_p \times x_{A}(1+R_{A})$ and
$V_p\times x_{B}(1+R_{B})$, respectively. Hence, at the end of the
holding period the portfolio is worth:
$$V_p\times\left[x_{A}(1+R_{A})+x_{B}(1+R_{B})\right] = V_p \times (1 + R_{p}).$$
Hence, $x_{A}(1+R_{A})+x_{B}(1+R_{B})=1+R_p$ defines the *portfolio
gross return*. The portfolio gross return is equal to a weighted average
of the gross returns on assets $A$ and $B$, where the weights are the
portfolio shares $x_{A}$ and $x_{B}$.

To determine the portfolio rate of return, re-write the portfolio gross
return as:
$$1+R_{p}=x_{A}+x_{B}+x_{A}R_{A}+x_{B}R_{B}=1+x_{A}R_{A}+x_{B}R_{B},$$
since $x_{A}+x_{B}=1$ by construction. Then the *portfolio rate of
return* is:
$$R_{p}=x_{A}R_{A}+x_{B}R_{B},$$
which is equal to a weighted average of the simple returns on assets $A$
and $B$, where the weights are the portfolio shares $x_{A}$ and $x_{B}$.

In the portfolio rate of return, the components $x_{A}R_{A}$ and
$x_{B}R_{B}$ are the *contributions* of assets A and B to the portfolio
return, respectively. These contributions add up to the total portfolio
return $R_{p}$.

::: {#exm-example0108 .example name="Compute portfolio return"}
Consider a portfolio of Microsoft and Starbucks stock in which you
initially purchase ten shares of each stock at the end of month $t-1$ at
the prices $P_{\textrm{msft},t-1}=\$85$ and
$P_{\textrm{sbux},t-1}=\$30$, respectively. The initial value of the
portfolio is $V_{t-1}=10\times\$85+10\times30=\$1,150$. The portfolio
shares are $x_{\textrm{msft}}=850/1150=0.7391$ and
$x_{\textrm{sbux}}=30/1150=0.2609$. Suppose at the end of month $t$,
$P_{\textrm{msft},t}=\$90$ and $P_{\textrm{sbux},t}=\$28$. Assuming that
Microsoft and Starbucks do not pay a dividend between periods $t-1$ and
$t$, the one-period returns on the two stocks are: 
\begin{align*}
R_{\textrm{msft},t} & =\frac{\$90-\$85}{\$85}=0.0588,\\
R_{\textrm{sbux},t} & =\frac{\$28-\$30}{\$30}=-0.0667.
\end{align*} 
The one-month rate of return on the portfolio is then: 
$$
R_{p,t}=(0.7391)(0.0588)+(0.2609)(-0.0667)=0.02609,
$$
The contributions of assets $A$ and $B$ to the portfolio return are: 
$$
x_{A}R_{A} = (0.7391)(0.0588) = 0.0435, ~ x_{B}R_{B} = (0.2609)(-0.0667) = -0.0174.
$$ 
Asset $A$ contributes positively and asset $B$ contributes negatively
to the portfolio return. The portfolio value at the end of month $t$ is:
$$
V_{t}=V_{t-1}(1+R_{p,t})=\$1,150\times(1.02609)=\$1,180 
$$

In R, the portfolio calculations are:
:::

```{r}
#| echo: true

P.msft = c(85, 90) 
P.sbux = c(30, 28) 
V = P.msft[1]*10 + P.sbux[1]*10
x.msft = 10*P.msft[1]/V 
x.sbux = 10*P.sbux[1]/V 
R.msft = (P.msft[2] - P.msft[1])/P.msft[1] 
R.sbux = (P.sbux[2] - P.sbux[1])/P.sbux[1] 
R.p = x.msft*R.msft + x.sbux*R.sbux 
V1 = V*(1 + R.p) 
c(R.p, x.msft*R.msft, x.sbux*R.sbux, V1)
```

\hfill{}$\blacksquare$

In general, for a portfolio of $n$ assets with investment shares $x_{i}$
such that $x_{1}+\cdots+x_{n}=1,$ the one-period portfolio gross and
simple returns are defined as:

$$
1+R_{p,t} =\sum_{i=1}^{n}x_{i}(1+R_{i,t})
$$ {#eq-portfolio-gross-return}

$$
R_{p,t} =\sum_{i=1}^{n}x_{i}R_{i,t}.
$$ {#eq-portfolio-return}

where $R_{i,t}$ is the one-period simple return on asset
$i$. Asset $i$'s contribution to the portfolio return is
$$
\mathrm{ctr}_{i,p} = x_{i}R_{i,t}.
$$ {#eq-portfolio-contribution} and
$$\sum_{i=i}^n \mathrm{ctr}_{i,p} = \sum_{i=i}^n x_{i}R_{i,t} = R_{p,t}.$$

### Multiperiod portfolio returns and rebalancing

How to compute multiperiod portfolio returns depends on the assumptions
made about the portfolio weights each period. Several scenarios exist:

1.  Portfolio weights from the initial portfolio are allowed to change
    over time as prices of the underlying assets change over time. In
    this case no rebalancing of the portfolio is done. This is called a
    *buy-and-hold* portfolio.
2.  Portfolio weights remain constant over time, which implies that the
    portfolio is rebalanced at every time period (associated with
    holding period) to maintain constant weights.
3.  Portfolio weights from the initial portfolio are rebalanced at
    specific time intervals that are different than the holding period
    (e.g. rebalance quarterly for monthly holding periods). This is a
    hybrid of scenarios 1 and 2.
4.  Portfolio weights are actively changed at each time period
    associated with the holding period. This is called an *actively
    managed portfolio*.

How to deal with the above scenarios is best illustrated through
examples.

#### No rebalancing of portolio weights

*Prices and shares framework*

We illustrate the buy-and-hold scenario 1 using a simple portfolio
consisting of two assets $A$ and $B$ invested for three months starting
at time $t-3$ in a framework where we know the asset prices each period
and the initial amounts invested in each
asset[^01-returncalculations-new-4]. In particular, we assume that the
initial portfolio consists of $s_A = 100$ shares of asset $A$ and
$s_B = 50$ shares of asset $B$. The end-of-month prices and values
($V_{i,t} = s_i \times P_{i,t}, i=A,B$) of the assets as well as the
end-of-month values of the portfolio are given by

[^01-returncalculations-new-4]: This example is inspired by the
    `return.portfolio` vignette in **PerformanceAnalytics**).

```{r}
#| echo: true
s.A = 100
s.B = 50
P.A = c(5,7,6,7)
V.A = s.A*P.A
P.B = c(10, 11, 12, 8)
V.B = s.B*P.B
V.p = V.A + V.B
table.vals = cbind(P.A, V.A, P.B, V.B, V.p)
rownames(table.vals) = c("t-3", "t-2", "t-1", "t")
table.vals
```

In the above table the time index $t$ represents the end-of-month $t$.
Slightly abusing notation, we can also say that the time index $t-1$
represents the beginning of month $t$. This is important because for the
portfolio calculations the end-of-month values $V_{A,t-1}$, $V_{B,t-1}$
and $V_{p,t-1}$ are used to calculate the portfolio weights to be
applied in month $t$. That is,
$$x_{A,t-1} = \frac{V_{A,t-1}}{V_{p,t-1}}, ~ x_{B,t-1} = \frac{V_{B,t-1}}{V_{p,t-1}}$$
represent the portfolio weights for assets $A$ and $B$ at the beginning
of month $t$. Then, the portfolio return over month $t$ is computed as
\begin{align}
R_{p,t} &= \%\Delta V_{p,t} = \frac{V_{p,t} - V_{p,t-1}}{V_{p,t-1}}\\
&= x_{A,t-1}R_{A,t} + x_{B,t-1}R_{B,t},\\
\end{align} where \begin{align}
R_{A,t} &= \%\Delta V_{A,t} = \frac{V_{A,t} - V_{A,t-1}}{V_{A,t-1}}\\
&= \%\Delta P_{A,t} = \frac{P_{A,t} - P_{A,t-1}}{P_{A,t-1}}\\
R_{B,t} &= \%\Delta V_{B,t} = \frac{V_{B,t} - V_{B,t-1}}{V_{B,t-1}}\\
&= \%\Delta P_{B,t} = \frac{P_{B,t} - P_{B,t-1}}{P_{B,t-1}}.
\end{align}

The portfolio return, $R_{p,t}$, is still a share weighted average of
the individual asset returns but with time varying portfolio weights
$x_{A,t-1}$ and $x_{B,t-1}$.

The R calculations for returns are:

```{r}
#| echo: true
R.A = (V.A[2:4] - V.A[1:3])/V.A[1:3]
R.B = (V.B[2:4] - V.B[1:3])/V.B[1:3]
R.p = (V.p[2:4] - V.p[1:3])/V.p[1:3]
table.returns = cbind(R.A, R.B, R.p)
rownames(table.returns) =  c("t-2", "t-1", "t")
table.returns
```

The R calculations for the weights are:

```{r}
#| echo: true
x.A = V.A/V.p
x.B = V.B/V.p
table.weights = cbind(x.A, x.B)
rownames(table.weights) = c("t-3", "t-2", "t-1", "t")
table.weights
```

Here, the initial portfolio formed at the end of month $t-3$ is an
equally weighted portfolio of assets $A$ and $B$. However, the portfolio
weights at the end of month $t-2$ are not $0.5$ and $0.5$ as the prices
of assets $A$ and $B$ have changed over the month. Asset $A$'s price has
gone up more in percentage terms ($R_{A,t-2}=40\%$) than asset $B$'s
price ($R_{B,t-2}=10\%$) so the portfolio weight in asset $A$ at the
beginning of $t-1$ ($x_{A,t-2}=0.56$), has become bigger than the
portfolio weight in asset $B$ ($x_{B,t-2} = 0.44$).

For the calculation of the portfolio returns, the asset weights at time
$t-3$ apply to the asset returns at time $t-2$, the weights at time
$t-2$ apply to the returns at time $t-1$, and the weights at time $t-1$
apply to the returns at time $t$:

```{r}
#| echo: true
R.p = x.A[1:3]*R.A + x.B[1:3]*R.B
R.p
```

The asset contributions to the portfolio return at time $t$ are
\begin{align}
\mathrm{ctr}_{A,t} &= \frac{\Delta V_{A,t}}{V_{p,t-1}} = x_{A,t-1}\times R_{A,t}\\
\mathrm{ctr}_{B,t} &= \frac{\Delta V_{B,t}}{V_{p,t-1}} = x_{B,t-1}\times R_{B,t}
\end{align} The R calculations for the contributions derived from values
are:

```{r}
#| echo: true
ctr.A = (V.A[2:4] - V.A[1:3])/V.p[1:3]
ctr.B = (V.B[2:4] - V.B[1:3])/V.p[1:3]
table.ctr = cbind(ctr.A, ctr.B, ctr.A+ctr.B)
colnames(table.ctr)[3] = "R.p"
rownames(table.ctr) = c("t-2", "t-1", "t")
table.ctr
```

The R calculations for the contributions derived from returns are:

```{r}
#| echo: true
ctr.A = x.A[1:3]*R.A
ctr.B = x.B[1:3]*R.B
table.ctr = cbind(ctr.A, ctr.B, R.p)
rownames(table.ctr) = c("t-2", "t-1", "t")
table.ctr
```

*Returns and weights framework*

The above calculations are based on knowing asset prices and initial
shares purchased for each asset. In many situations, we only have return
information and initial portfolio weights on each asset and the initial
value of the portfolio. Using the above example, the initial portfolio
weights are $x_{A,t-3}=x_{B,t-3}=0.5$ and the initial value of the
portfolio is $V_{p,t-3}=1000$. We also know the returns, $R_{i,t}$, each
period for the two assets:

```{r}
#| echo: true
table.returns[, c("R.A", "R.B")]
```

We deduce the initial asset values using \begin{align}
V_{A,t-3} &= x_{A,t-3}\times V_{p,t-3} = 0.5\times 1000 = 500, \\
V_{B,t-3} &= x_{B,t-3}\times V_{p,t-3} = 0.5\times 1000 = 500.
\end{align}

The end-of-month asset values for $i=A,B$ are computed using
\begin{align*}
V_{i,t-2} &= V_{i,t-3}\times (1+R_{i,t-2}), \\
V_{i,t-1} &= V_{i,t-2}\times (1+R_{i,t-1}) = V_{i,t-3}\times (1+R_{i,t-2})(1+R_{i,t-1}), \\
V_{i,t} &= V_{i,t-1}\times (1+R_{i,t}) = V_{i,t-3}\times (1+R_{i,t-2})(1+R_{i,t-1})(1+R_{i,t}).
\end{align*}

The end-of-month portfolio values are the sum of the end-of-month asset
values. The R calculations are:

```{r}
#| echo: true
V.A = V.B = V.p = rep(0, 4)
V.p[1] = 1000
V.A[1] = V.B[1] = 0.5*V.p[1]
V.A[2:4] = cumprod(1+R.A)*V.A[1]
V.B[2:4] = cumprod(1+R.B)*V.B[1]
V.p[2:4] = V.A[2:4] + V.B[2:4]
table.vals = cbind(V.A, V.B, V.p)
rownames(table.vals) = c("t-3", "t-2", "t-1", "t")
table.vals
```

The portfolio weights at the end-of-month $t$ are computed as
$$x_{i,t} = \frac{V_{i,t}}{V_{p,t}},$$
and the R calculations are:

```{r}
#| echo: true
x.A = V.A/V.p
x.B = V.B/V.p
table.weights = cbind(x.A, x.B)
rownames(table.weights) = c("t-3", "t-2", "t-1", "t")
table.weights
```

The portfolio return over month $t$ can be computed using \begin{align}
R_{p,t} &= \%\Delta V_{p,t} = \frac{V_{p,t}-V_{p,t-1}}{V_{p,t-1}} \\
&= x_{A,t-1}\times R_{A,t} + x_{B,t-1}\times R_{B,t}
\end{align}

The R calculations for returns based on values are:

```{r}
#| echo: true
diff(table.vals[, "V.p"])/table.vals[1:3, "V.p"]
```

The calculation for returns based on weights is

```{r}
#| echo: true
x.A[1:3]*R.A + x.B[1:3]*R.B
```

#### Constant portfolio weights

If the initial portfolio weights are to be held constant over time then
the portfolio typically needs to be *rebalanced* at each time period to
adjust the portfolio weights back to the initial weights. Given the
constant portfolio weights $x_A$ and $x_B$ the return on the portfolio
at every time period is given by
$$R_{p,t} = x_A R_{A,t} + x_B R_{B,t}.$$

The rebalancing of portfolio weights can be illustrated with the example
from the previous sub-section. The initial portfolio is an equally
weighted portfolio, $x_{A,t-3} = x_{B,t-3} = 0.5$. To have constant
portfolio weights the portfolio will need to be rebalanced each period
so that $x_{A,t-2} = x_{A,t-1} = x_{A,t} = 0.5$ and
$x_{B,t-2} = x_{B,t-1} = x_{B,t} = 0.5$. Consider the portfolio at the
end-of-month $t-2$ where \begin{align}
V_{A,t-2} &= V_{A,t-3}\times (1 + R_{A,t-2}) = 500 \times (1 + 0.40) = 700 \\
V_{B,t-2} &= V_{B,t-3}\times (1 + R_{B,t-2}) = 500 \times (1 + 0.10) = 550 \\
V_{p,t-2} &= V_{A,t-2} + V_{B,t-2} = 700 + 550 = 1250 \\
x_{A,t-2} &= \frac{V_{A,t-2}}{V_{p,t-2}} = \frac{700}{1250} = 0.56 \\
x_{B,t-2} &= \frac{V_{B,t-2}}{V_{p,t-2}} = \frac{550}{1250} = 0.44 
\end{align} The end-of-month portfolio weights are $x_{A,t-2} = 0.56$
and $x_{B,t-2} = 0.44$, which are not equal to $0.5$. Since
$V_{p,t-2} = 1250$, to rebalance the portfolio so that
$x_{A,t-2} = x_{B,t-2} = 0.5$ we need
$V_{A,t-2} = V_{B,t-2} = V_{p,t-2}/2 = 625$. This requires selling
$700 - 625 = 75$ of asset A and purchasing $75$ of asset B at the
end-of-month $t-2$. Then, at the beginning of month $t-1$ we have
$V_{A,t-2} = V_{B,t-2} = 625$ and $x_{A,t-2} = x_{B,t-2} = 0.5$.

At the end-of-month $t-1$, for the rebalanced portfolio we have
\begin{align}
V_{A,t-1} &= V_{A,t-2}\times (1 + R_{A,t-1}) = 625 \times (1 - 0.1429) = 535.69 \\
V_{B,t-1} &= V_{B,t-2}\times (1 + R_{B,t-1}) = 625 \times (1 + 0.0909) = 681.81 \\
V_{p,t-1} &= V_{A,t-2} + V_{B,t-2} = 535.69 + 681.81 = 1217.50 \\
x_{A,t-1} &= \frac{V_{A,t-1}}{V_{p,t-1}} = \frac{535.69}{1217.50} = 0.44 \\
x_{B,t-1} &= \frac{V_{B,t-1}}{V_{p,t-1}} = \frac{681.81}{1217.50} = 0.56 
\end{align} The end-of-month portfolio weights are $x_{A,t-1} = 0.44$
and $x_{B,t-1} = 0.56$. Since $V_{p,t-1} = 1217.50$, to rebalance the
portfolio we need $V_{A,t-1} = V_{B,t-1} = 608.75$. This requires
selling $73.06$ of asset B and purchasing $73.06$ of asset A and the
end-of-month t-1. Then, at the beginning of month $t$ we have
$V_{A,t-1} = V_{B,t-1} = 608.75$ and $x_{A,t-1} = x_{B,t-1} = 0.5$.

Finally, at the end-of-month $t$, for the rebalanced portfolio we have
\begin{align}
V_{A,t} &= V_{A,t-1}\times (1 + R_{A,t}) = 608.75 \times (1 + 0.1667) = 710.23 \\
V_{B,t} &= V_{B,t-1}\times (1 + R_{B,t}) = 608.75 \times (1 - 0.3333) = 405.85 \\
V_{p,t} &= V_{A,t-1} + V_{B,t-1} = 710.23 + 405.85 = 1116.08 \\
x_{A,t} &= \frac{V_{A,t}}{V_{p,t}} = \frac{710.23}{1116.08} = 0.64 \\
x_{B,t} &= \frac{V_{B,t}}{V_{p,t}} = \frac{405.85}{1116.08} = 0.36 
\end{align} The end-of-month portfolio weights are $x_{A,t} = 0.64$ and
$x_{B,t} = 0.36$. Since $V_{p,t} = 1116.08$, to rebalance the portfolio
we need $V_{A,t} = V_{B,t} = 558.04$. This requires selling $152.19$ of
asset A and purchasing $152.19$ of asset B and the end-of-month t. Then,
at the beginning of month $t+1$ we have $V_{A,t} = V_{B,t} = 558.04$ and
$x_{A,t} = x_{B,t} = 0.5$.

The above calculations in R are:

```{r}
value.table = matrix(0, 4, 3)
rownames(value.table) = c("t-3", "t-2", "t-1", "t")
colnames(value.table) = c("V.A", "V.B", "V.p")
value.table["t-3", ] = c(V.A[1], V.B[1], V.p[1])
for (t in 2:4) {
  value.table[t, "V.A"] = value.table[t-1, "V.A"]*(1 + R.A[t-1])
  value.table[t, "V.B"] = value.table[t-1, "V.B"]*(1 + R.B[t-1])
  value.table[t, "V.p"] = value.table[t, "V.A"] + value.table[t, "V.B"]
  RB = max(value.table[t, "V.A"], value.table[t, "V.B"]) - value.table[t, "V.p"]/2
  if (value.table[t, "V.A"] > value.table[t, "V.B"]) {
    value.table[t, "V.A"] = value.table[t, "V.A"] - RB
    value.table[t, "V.B"] = value.table[t, "V.B"] + RB
    }
  else{
    value.table[t, "V.A"] = value.table[t, "V.A"] + RB
    value.table[t, "V.B"] = value.table[t, "V.B"] - RB
  }
  value.table[t, "V.p"] = value.table[t, "V.A"] + value.table[t, "V.B"]
}
value.table
```

Notice that the rebalanced portfolio slightly outperforms the
buy-and-hold portfolio.

The portfolio return each period can be computed using
$R_{p,t} = \%\Delta V_{p,t}$:

```{r}
diff(value.table[, "V.p"])/value.table[1:3, "V.p"]
```

It can also be computed using
$R_{p,t} = 0.5\times R_{A,t} + 0.5\times R_{B,t}$:

```{r}
0.5*R.A + 0.5*R.B
```

Here, we see that the rebalanced portfolio loses less money over month
$t-1$ than the buy-and-hold portfolio.

In the above example, at each rebalancing date we sell the asset that
has the best performance (sell high) and we buy the asset that has the
worst performance (buy low). Hence, rebalancing forces the sage
investing advice: "buy low and sell high".

#### Rebalance portfolio at specified dates

When the portfolio is rebalanced to a fixed initial portfolio, such as
an equally weighted portfolio, on specific dates the portfolio return
calculations become a hybrid of those used for scenarios 1 and 2. The
portfolio return is computed as
$$R_{p,t} = x_{A,t-1}\times R_{A,t} + x_{B,t-1}\times R_{B,t},$$
where $x_{i,t-1} (i=A,B)$ are equal to the initial portfolio weights on
the rebalancing dates and is computed like the buy-and-hold weights in
between the rebalancing dates.

#### Actively managed portfolio

In an actively managed portfolio, the portfolio weights are chosen by
the portfolio manager at each time period. Many mutual funds are managed
in this way. The return on the actively managed portfolio is computed as
$$R_{p,t} = x_{A,t-1}\times R_{A,t} + x_{B,t-1}\times R_{B,t}$$
where $x_{A,t-1}$ and $x_{B,t-1}$ are the portolio weights chosen by the
portfolio manager at the beginning of time $t$.

### Continuously Compounded Portfolio Returns

The continuously compounded portfolio return is defined by
@eq-cc-return, where $R_{t}$ is computed using the portfolio return
@eq-portfolio-return. However, notice that: $$
r_{p,t}=\ln(1+R_{p,t})=\ln\left(1+\sum_{i=1}^{n}x_{i}R_{i,t}\right)\neq\sum_{i=1}^{n}x_{i}r_{i,t},
$$ {#eq-cc-portfolio-return} where $r_{i,t}$ denotes the continuously compounded
one-period return on asset $i$. Hence, the continuously compounded
portfolio return is not a share weighted average of the individual asset
continuously compounded returns. However, if the portfolio return
$R_{p,t}=\sum_{i=1}^{n}x_{i}R_{i,t}$ is not too large then
$r_{p,t}\approx R_{p,t}$ otherwise, $R_{p,t}>r_{p,t}$.

::: {#exm-example0117 .example name="Compute continuously compounded portfolio returns"}
Consider a portfolio of Microsoft and Starbucks stock with
$x_{\textrm{msft}}=0.25$, $x_{\textrm{sbux}}=0.75$,
$R_{\textrm{msft},t}=0.0588$, $R_{\textrm{sbux},t}=-0.0503$ and
$R_{p,t}=-0.02302$. Using (@eq-cc-portfolio-return), the continuously
compounded portfolio return is: $r_{p,t}=\ln(1-0.02302)=\ln(0.977)=-0.02329.$ Using
$r_{\textrm{msft},t}=\ln(1+0.0588)=0.0572$ and
$r_{\textrm{sbux},t}=\ln(1-0.0503)=-0.05161,$ notice that: 
$x_{\textrm{msft}}r_{\textrm{msft}}+x_{\textrm{sbux}}r_{\textrm{sbux}}=-0.02442\neq r_{p,t}$


\hfill{}$\blacksquare$
:::

Because the continuously compounded portfolio return is not a share
weighted average of the individual asset continuously compounded
returns, the analysis of portfolios is typically performed using simple
returns and not continuously compounded returns.

## Return Calculations with Data in R {#sec-ReturnCalculationsR}

This section discusses representing time series data in R using `xts`
objects, the calculation of returns from historical prices in R, as well
as the graphical display of prices and returns.

### Representing time series data using `xts` objects

The examples in this section are based on the daily adjusted closing
price data for Microsoft and Starbucks stock over the period January 4,
1993 through December 31, 2014[^01-returncalculations-new-5]. These data
are available as the `xts` objects `msftDailyPrices` and
`sbuxDailyPrices` in the R package **introcompfinr**
[^01-returncalculations-new-6]

[^01-returncalculations-new-5]: These are prices that are adjusted for
    dividends and stock splits. That is, any dividend payments have been
    included in the prices and historical prices have been divided by
    the split ratio associated with any stock splits.

[^01-returncalculations-new-6]: See the Appendix Working with Time
    Series Data in R for an overview of `xts` objects and Zivot (2016)
    for a comprehensive coverage.

```{r}
#| echo: true
head(cbind(msftDailyPrices,sbuxDailyPrices), 3)
```

```{r}
class(msftDailyPrices)
```

There are many different ways of representing a time series of data in
R. For financial time series `xts` (extensible time series) objects from
the **xts** package are especially convenient and useful. An `xts`
object consists of two pieces of information: (1) a `matrix` of numeric
data with different time series in the columns, (2) an R object
representing the common time indexes associated with the rows of the
data. `xts` objects extend and enhance the `zoo` class of time series
objects from the **zoo** package written by Achim Zeileis (zoo stands
for Z's ordered observations).

The `matrix` of time series data can be extracted from the `xts` object
using the **xts** function `coredata()`:

```{r}
#| echo: true
matrix.data = coredata(msftDailyPrices)
head(matrix.data, 3)
```

The resulting `matrix` of data does not have any date information. The
date index can be extracted using the **xts** function `index()`:

```{r}
#| echo: true
date.index = index(msftDailyPrices)
head(date.index, 3)
```

The object `date.index` is of class `Date`:

```{r}
#| echo: true
class(date.index)
```

When a `Date` object is printed, the date is displayed in the format
`yyyy-mm-dd`. However, internally each date represents the number of
days since January 1, 1970:

```{r}
#| echo: true
as.numeric(date.index[1])
```

Here, `1993-01-04` (January 4, 1993) is $8,404$ days after January 1,
1970. This allows for simple *date arithmetic* (e.g. adding and
subtracting dates, etc). More sophisticated date arithmetic is available
using the functions in the **lubridate** package.

There are several advantages of using `xts` objects to represent
financial time series data. Most financial time series do not follow an
equally spaced regular calendar. The time index of a `xts` object can be
any strictly increasing date sequence which can match the times and
dates for which assets trade on an exchange. For example, assets traded
on the main US stock exchanges (e.g. NYSE and NASDAQ) trade only on
weekdays between 9:30AM and 4:00PM Eastern Standard Time and do not
trade on a number of holidays.

Another advantage of `xts` objects is that you can easily extract
observations at or between specific dates. For example, to extract the
daily price on January 3, 2014 use

```{r}
#| echo: true
msftDailyPrices["2014-01-03"]
```

To extract prices between January 3, 2014 and January 7, 2014, use

```{r}
#| echo: true

msftDailyPrices["2014-01-03::2014-01-07"]
```

Or to extract all of the prices for January, 2014 use:

```{r}
#| echo: true

msftDailyPrices["2014-01"]
```

Two or more `xts` objects can be merged together and aligned to a common
date index using the **xts** function `merge()`:

```{r}
#| echo: true

msftSbuxDailyPrices = merge(msftDailyPrices, sbuxDailyPrices) 
head(msftSbuxDailyPrices, n=3)
```

If two or more `xts` objects have a common date index then they can be
combined using `cbind()`.

#### Basic calculations on `xts` objects.

Because `xts` objects typically contain a matrix of numeric data, many R
functions that operate on matrices also operate on `xts` objects. For
example, to transform `msftDailyPrices` to log prices use

```{r}
#| echo: true
msftDailyLogPrices = log(msftDailyPrices)
head(msftDailyLogPrices, 3)
```

If two or more `xts` objects have the same time index then they can be
added, subtracted, multiplied, and divided. For example,

```{r}
#| echo: true
msftPlusSbuxDailyPrices = msftDailyPrices + sbuxDailyPrices
colnames(msftPlusSbuxDailyPrices) = "MSFT+SBUX"
head(msftPlusSbuxDailyPrices,3)
```

To compute the average of the Microsoft daily prices use

```{r}
mean(msftDailyPrices)
```

If an R function, for some unknown reason, is not operating correctly on
an `xts` object first extract the data using `coredata()` and then call
the R function. For example,

```{r}
mean(coredata(msftDailyPrices))
```

#### Changing the frequency of an `xts` object

Data at the daily frequency is the highest frequency of data considered
in this book. However, in many of the examples we want to use data at
the weekly or monthly frequency. The conversion of data at a daily
frequency to data at a monthly frequency is easy to do with `xts`
objects. For example, end-of-month prices can be extracted from the
daily prices using the **xts** function `to.monthly()`:

```{r}
#| echo: true

msftMonthlyPrices = to.monthly(msftDailyPrices, OHLC=FALSE) 
sbuxMonthlyPrices = to.monthly(sbuxDailyPrices, OHLC=FALSE)
msftSbuxMonthlyPrices = to.monthly(msftSbuxDailyPrices, OHLC=FALSE) 
head(msftMonthlyPrices, 3)
```

By default, `to.monthly()` extracts the data for the last day of the
month and creates a **zoo** `yearmon` date index. For monthly data, the
`yearmon` date index is convenient for printing and plotting as the
month and year are nicely printed. To preserve the `Date` class of the
time index and show the end-of-month date, use the optional argument
`indexAt = "lastof"` in the call to `to.monthly()`:

```{r}
#| echo: true
head(to.monthly(msftDailyPrices, OHLC=FALSE, indexAt="lastof"), 3) 
```

In the above calls to `to.monthly()`, the optional argument `OHLC=FALSE`
prevents the creation of open, high, low, and closing prices for the
month.

In a similar fashion, you can extract end-of-week prices using the `xts`
function `to.weekly()`:

```{r}
#| echo: true
msftWeeklyPrices = to.weekly(msftDailyPrices, OHLC=FALSE)
head(msftWeeklyPrices, 3)
```

Here, the weekly data are the closing prices on each Friday of the week:

```{r}
head(weekdays(index(msftWeeklyPrices)),3)
```

#### Plotting `xts` objects with `plot.xts()` and `plot.zoo`

Time plots of `xts` objects can be created with the generic `plot()`
function as there is a method function in the **xts** package for
objects of class `xts`. For example, Figure
@fig-MsftSbuxMonthlyPricesXts shows a basic time plot of the monthly
closing prices of Microsoft and Starbucks created with:

```{r}
#| label: fig-MsftSbuxMonthlyPricesXts
#| fig-cap: Single panel plot (default) with `plot.xts()`
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(msftSbuxMonthlyPrices, main="Monthly Closing Prices", 
     legend.loc="topleft")
```

The default plot style in `plot.xts()` is a single-panel plot with
multiple series. You can also create multi-panel plots, as in Figure
@fig-MsftSbuxPricesXtsMultiple, by setting the optional argument
`multi.panel = TRUE` in the call to `plot.xts()`

```{r}
#| label: fig-MsftSbuxPricesXtsMultiple
#| fig-cap: Multi-panel time series plot with `plot.xts()`
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(msftSbuxMonthlyPrices, main="Monthly Closing Prices", 
     multi.panel=TRUE)
```

See the help file for `plot.xts()` for more examples of plotting `xts`
objects.

Because the `xts` class inherits from the `zoo` class, the **zoo**
method function `plot.zoo()` can also be used to plot `xts` objects.
Figure @fig-MsftSbuxPricesZoo is created with

```{r}
#| label: fig-MsftSbuxPricesZoo
#| fig-cap: Single panel time series plot (default) with `plot.zoo()`
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot.zoo(msftSbuxMonthlyPrices, plot.type="single", 
         main="Monthly Closing Prices",         
         lwd = 2, col=c("black", "red"), ylab="Price") 
grid() 
legend(x="topleft", legend=colnames(msftSbuxMonthlyPrices),         
      lwd = 2, col=c("black", "red"))
```

A muti-panel plot, shown in Figure @fig-MsftSbuxPricesZooMultiple,
can be created using

```{r}
#| label: fig-MsftSbuxPricesZooMultiple
#| fig-cap: Multi-panel time series plot with `plot.zoo()`
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot.zoo(msftSbuxMonthlyPrices, plot.type="multiple", main="",
         lwd = 2, col=c("black", "red"), cex.axis=0.8)
```

Creating plots with `plot.zoo()` is a bit more involved than with
`plot.xts()`, as `plot.xts()` is newer. See the help file for
`plot.zoo()` for more examples.

#### Ploting `xts` objects using `autoplot()` from **ggplot2**

The plots created using `plot.xts()` and `plot.zoo()` are created using
the base graphics system in R. This allows for a lot of flexibility but
often the resulting plots do not look very exciting or modern. In
addition, these functions are not well suited for plotting many time
series together in a single plot.

Another very popular graphics system in R is provided by the **ggplot2**
package by Hadley Wickham of RStudio. For plotting `xts` objects,
especially with multiple columns (data series), the **ggplot2** function
`autoplot()` is especially convenient and easy:

```{r}
#| echo: true
autoplot(msftSbuxDailyPrices, facets = NULL) +
  ggtitle("Daily Closing Prices") +
  ylab("Closing Price Per Share") +
  xlab("Year")
```

The syntax for creating graphs with `autoplot()` uses the *grammar of
graphics* syntax from the **ggplot2** package. See the [R Graphics
Cookbook](https://r-graphics.org/) for more details and examples on
using the **ggplot2** package. The call to `autoplot()` first invokes
the method function `autoplot.zoo()` from the **zoo** package and
creates a basic time series plot. Additional layers showing a main title
and axes labels are added using `+`. The optional argument
`facets = NULL` specifies that all series are plotted together on a
single plot with different colors.

To produce a multi-panel plot call `autoplot()` with
`facets = Series ~ .`:

```{r}
autoplot(msftSbuxDailyPrices, facets = Series ~ .) +
  ggtitle("Daily Closing Prices") +
  ylab("Closing Price Per Share") +
  xlab("Year")
```

See the help file for `autoplot.zoo()` for more examples.

### Calculating returns

In this sub-section, we illustrate how to calculate a time series of
returns from a time series of prices.

#### Brute force return calculations

Consider computing simple monthly returns,
$R_{t}=\frac{P_{t}-P_{t-1}}{P_{t-1}}$, from historical prices using the
`xts` object `sbuxMonthlyPrices`. The R code for a brute force
calculation is:

```{r}
#| echo: true
sbuxMonthlyReturns = diff(sbuxMonthlyPrices)/stats::lag(sbuxMonthlyPrices)
head(sbuxMonthlyReturns, n=3)
```

Here, the `diff()` function computes the first difference in the prices,
$P_{t}-P_{t-1}$, and the `lag()` function computes the lagged price,
$P_{t-1}.$[^01-returncalculations-new-7] An equivalent calculation is
$R_{t}=\frac{P_{t}}{P_{t-1}}-1$:

[^01-returncalculations-new-7]: The method function `lag.xts()` works
    differently than generic `lag()` function and the method function
    `lag.zoo()`. In particular, `lag.xts()` with optional argument `k=1`
    gives the same result as `lag()` and `lag.zoo()` with `k=-1`.

```{r}
#| echo: true

head(sbuxMonthlyPrices/stats::lag(sbuxMonthlyPrices) - 1, n=3)
```

Notice that the return for January, 1993 is `NA` (missing value). To
automatically remove this missing value, use the R function `na.omit()`
or explicitly exclude the first observation:

```{r}
#| echo: true

head(na.omit(sbuxMonthlyReturns), n=3)
```

```{r}
#| echo: true

head(sbuxMonthlyReturns[-1], n=3)
```

To compute continuously compounded returns from simple returns, use:

```{r}
#| echo: true

sbuxMonthlyReturnsC = na.omit(log(1 + sbuxMonthlyReturns))
head(sbuxMonthlyReturnsC, n=3)
```

Or, equivalently, to compute continuously compounded returns directly
from prices, use:

```{r}
#| echo: true

sbuxMonthlyReturnsC = na.omit(diff(log(sbuxMonthlyPrices))) 
head(sbuxMonthlyReturnsC, n=3)
```

The above calculations used an `xts` object with a single column. If an
`xts` object has multiple columns the same calculations work for each
column. For example,

```{r}
msftSbuxMonthlyReturns = diff(msftSbuxMonthlyPrices)/stats::lag(msftSbuxMonthlyPrices)
head(msftSbuxMonthlyReturns, n=3)
```

Also,

```{r}
msftSbuxMonthlyReturnsC = diff(log(msftSbuxMonthlyPrices))
head(msftSbuxMonthlyReturnsC, n=3)
```

These are examples of *vectorized* calculations. That is, the
calculations are performed on all columns simultaneously. This is
computationally more efficient than looping the same calculation for
each column. In R, you should try to vectorize calculations whenever
possible.

#### Calculating returns using `Return.calculate()`

Simple and continuously compounded returns can be computed using the
**PerformanceAnalytics** function `Return.calculate()`:

```{r}
#| echo: true
# Simple returns
sbuxMonthlyReturns = Return.calculate(sbuxMonthlyPrices) 
head(sbuxMonthlyReturns, n=3)
```

```{r}
#| echo: true
# CC returns 
sbuxMonthlyReturnsC = Return.calculate(sbuxMonthlyPrices, method="log") 
head(sbuxMonthlyReturnsC, n=3)
```

`Return.calculate()` also works with `xts` objects with multiple
columns:

```{r}
msftSbuxMonthlyReturns = Return.calculate(msftSbuxMonthlyPrices)
head(msftSbuxMonthlyReturns, 3)
```

The advantages of using `Return.calculate()` instead of brute force
calculations are: (1) clarity of code (you know what is being computed),
(2) the same function is used for both simple and continuously
compounded returns, (3) built-in error checking.

#### Equity curves

\index{Equity curves} To directly compare the investment performance of
two or more assets, plot the simple multi-period cumulative returns of
each asset on the same graph. This type of graph, sometimes called an
*equity curve*, shows how a one dollar investment amount in each asset
grows over time. Better performing assets have higher equity curves. For
simple returns, the *k*-period returns are
$R_{t}(k)=\prod\limits _{j=0}^{k-1}(1+R_{t-j})$ and represent the growth
of one dollar invested for $k$ periods. For continuously compounded
returns, the *k*-period returns are
$r_{t}(k)=\sum\limits _{j=0}^{k-1}r_{t-j}$. However, this continuously
compounded $k$-period return must be converted to a simple $k$-period
return, using $R_{t}(k)=\exp\left(r_{t}(k)\right)-1$, to properly
represent the growth of one dollar invested for $k$ periods.

::: {#exm-example0506 .example name="Equity curves for Microsoft and Starbucks monthly returns"}
To create the equity curves for Microsoft and Starbucks based on simple
returns use (You can also use the **PerformanceAnalytics** function
`chart.CumReturns()`)
:::

```{r}
#| label: fig-DSmonthlyCumulativeReturns
#| fig-cap: Monthly equity curve for  Microsoft and Starbucks.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
msftMonthlyReturns = na.omit(Return.calculate(msftMonthlyPrices))
sbuxMonthlyReturns = na.omit(Return.calculate(sbuxMonthlyPrices))
equityCurveMsft = cumprod(1 + msftMonthlyReturns)  
equityCurveSbux = cumprod(1 + sbuxMonthlyReturns)  
dataToPlot = merge(equityCurveMsft, equityCurveSbux)
plot(dataToPlot, main="Monthly Equity Curves", legend.loc="topleft")
```

The R function `cumprod()` creates the cumulative products needed for
the equity curves. Figure @fig-DSmonthlyCumulativeReturns shows that
a one dollar investment in Starbucks dominated a one dollar investment
in Microsoft over the given period. In particular, \$1 invested in
Microsoft grew to about only \$22 (over about 20 years) whereas \$1
invested in Starbucks grew over \$70. Notice the huge increases and
decreases in value of Microsoft during the dot-com bubble and bust over
the period 1998 - 2001, and the enormous increase in value of Starbucks
from 2009 - 2014.

\hfill{}$\blacksquare$

### Calculating portfolio returns from time series data

As discussed in sub-section @sec-portfolios-and-portfolio-returns, the
calculation of portfolio returns depends on the assumptions made about
the portfolio weights over time.

#### Constant portfolio weights

The easiest case assumes that portfolio weights are constant over time.
To illustrate, consider an equally weighted portfolio of Microsoft and
Starbucks stock. A time series a portfolio monthly returns can be
created directly from the monthly returns on Microsoft and Starbucks
stock:

```{r}
x.msft = x.sbux = 0.5
msftMonthlyReturns = na.omit(Return.calculate(msftMonthlyPrices))
sbuxMonthlyReturns = na.omit(Return.calculate(sbuxMonthlyPrices))
equalWeightPortMonthlyReturns = x.msft*msftMonthlyReturns + 
  x.sbux*sbuxMonthlyReturns
head(equalWeightPortMonthlyReturns, 3)
```

The same calculation can be performed using the **PerformanceAnalytics**
function `Return.portfolio()`:

```{r}
#| echo: true
equalWeightPortMonthlyReturns = 
  Return.portfolio(na.omit(msftSbuxMonthlyReturns), 
                   weights = c(x.msft, x.sbux),
                   rebalance_on = "months")
head(equalWeightPortMonthlyReturns, 3)
```

Here, the optional argument `rebalance_on = "months"` specifies that the
portfolio is to be rebalanced to the specified weights at the beginning
of every month.

If you also want the asset contributions to the portfolio return set the
optional argument `contribution = TRUE` (recall, asset contributions sum
to the portfolio return):

```{r}
#| echo: true
equalWeightPortMonthlyReturns = 
  Return.portfolio(na.omit(msftSbuxMonthlyReturns),
                   weights = c(x.msft, x.sbux),
                   rebalance_on = "months",
                   contribution = TRUE)
head(equalWeightPortMonthlyReturns, 3)
```

You can see the rebalancing calculations by setting the optional
argument `verbose = TRUE`:

```{r}
#| echo: true
equalWeightPortMonthlyReturns = 
  Return.portfolio(na.omit(msftSbuxMonthlyReturns),
                   weights = c(x.msft, x.sbux),
                   rebalance_on = "months",
                   verbose = TRUE)
names(equalWeightPortMonthlyReturns)
```

In this case `Return.portfolio()` returns a list with the above
components. The component `BOP.Weight` shows the beginning-of-period
portfolio weights, and the component `EOP.Weight` shows the
end-of-period portfolio weights (before any rebalancing).

```{r}
#| echo: true
head(equalWeightPortMonthlyReturns$BOP.Weight, 3)
```

With monthly returns and with `rebalance_on = "months"`, the
beginning-of-period weights are always rebalanced to the initially
supplied weights.

```{r}
#| echo: true
head(equalWeightPortMonthlyReturns$EOP.Weight, 3)
```

Because the returns for Microsoft and Starbux are different every month
the end-of-period weights drift away from the beginning-of-period
weights.

#### Buy-and-hold portfolio

In the buy-and-hold portfolio, no rebalancing of the initial portfolio
is done and the portfolio weights are allowed to evolve over time based
on the returns of the assets. Hence, the computation of the portfolio
return each month requires knowing the weights on each asset each month.
You can use `Return.portfolio()` to compute the buy-and-hold portfolio
returns as follows:

```{r}
buyHoldPortMonthlyReturns = 
  Return.portfolio(na.omit(msftSbuxMonthlyReturns),
                   weights = c(x.msft, x.sbux))
head(buyHoldPortMonthlyReturns, 3)
```

Omitting the optional argument `rebalance_on` tells `Return.portfolio()`
to not rebalance the portfolio at all (default behavior of the
function).

#### Periodic rebalancing

In practice, a portfolio may be rebalanced at a specific frequency that
is different than the frequency of the returns. For example, with
monthly returns the portfolio might be rebalanced quarterly or annually.
For example, to rebalance the equally weighted portfolio of Microsoft
and Starbucks stock each quarter (every three months) use

```{r}
#| echo: true
equalWeightQtrPortMonthlyReturns = 
  Return.portfolio(na.omit(msftSbuxMonthlyReturns),
                   weights = c(x.msft, x.sbux),
                   rebalance_on = "quarters")
head(equalWeightQtrPortMonthlyReturns, 3)
```

In between quarters, the portfolio evolves as a buy-and-hold portfolio.
To see the details of the rebalancing set `verbose = TRUE` in the call
to `Return.Portfolio()`.

#### Active portfolio weights

If the weights are actively chosen at specific time periods, these
weights can be supplied to the function `Return.portfolio()` to compute
the portfolio return.

### Downloading financial data from the internet

The data for the examples in this book, available in the
**introcompFinr** package, were downloaded from
[finance.yahoo.com](https://finance.yahoo.com/) using the `getSymbols()`
function from the \index{quantmod} **quantmod** package.

For example, to download daily data on Amazon and Google stock (ticker
symbols AMZN and GOOG) from January 3, 2007 through January 3, 2020 use
`getSymbols()` as follows:

```{r}
#| echo: true
options("getSymbols.warning4.0"=FALSE)
suppressPackageStartupMessages(library(quantmod))
getSymbols(Symbols = c("AMZN", "GOOG"), from="2007-01-03", to="2020-01-03",
           auto.assign=TRUE, warnings=FALSE)
class(AMZN)
```

The `xts` objects `AMZN` and `GOOG` each have six columns containing the
daily open price for the day, high price for the day, low price for the
day, close price for the day, volume for the day, and (dividend and
split) adjusted closing price for the day:

```{r}
head(AMZN, 3)
```

```{r}
head(GOOG, 3)
```

## Further Reading: Return Calculations

This chapter describes basic asset return calculations with an emphasis
on equity calculations. Similar material is covered in Chapter 2 of
[@RuppertMatteson-2015]. Comprehensive coverage of general return
calculations is given in [@Bacon-2008] and [@CCF-2009]. The virtues of
portfolio rebalancing are discussed in [@Ang-2014] and [@Qian-2019].

This chapter introduced you to the R packages **PerformanceAnalytics**,
**quantmod**, and **xts** for working with financial time series. It is
recommended to read the vignettes included in each of these packages.
Jeff Ryan's original [quantmod page](http://www.quantmod.com/), while a
bit outdated, has many useful examples of working with **quantmod** and
**xts**. A good [online
tutorial](http://joshuaulrich.github.io/xts/index.html) for the **xts**
package is provided by Joshua Ulrich, one of the creators of **xts**.

There are many useful R packages for the analysis of financial data. An
up-to-date list of R packages relevant for finance is given on the
[Empirical Finance Task
View](https://cran.r-project.org/web/views/Finance.html) on the
comprehensive R archive network (CRAN).

\newpage{}

## Appendix: Properties of Exponentials and Logarithms

The computation of continuously compounded returns requires the use of
natural logarithms. The natural logarithm function, $\ln(\cdot),$ is the
inverse of the exponential function, $e^{(\cdot)}=\exp(\cdot),$ where
$e^{1}=2.718.$ That is, $\ln(x)$ is defined such that $x=\ln(e^{x}).$
Figure @fig-logexp plots $e^{x}$ and $\ln(x)$. Notice that $e^{x}$
is always positive and increasing in $x$. $\ln(x)$ is monotonically
increasing in $x$ and is only defined for $x>0.$ Also note that
$\ln(1)=0$ and $\ln(0)=-\infty.$

```{r}
#| echo: false
#| label: fig-logexp
#| fig-cap: Exponential and natural logarithm functions.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
x = seq(-1, 2, length.out = 100)
e.x = exp(x)
ln.x = log(x[x > 0])
lb = min(e.x, ln.x)
ub = max(e.x, ln.x)
plot(x, e.x, ylim = c(lb, ub), type = "l", lty = 2, lwd = 2, col = "blue",
     xlab = "x", ylab = "y")
lines(x[x > 0], ln.x, lwd = 2, col = "red")
abline(h=0, v=0)
legend(x = "topleft", legend = c("exp(x)", "ln(x)"), lty=c(2, 1), lwd = 2,
       col = c("blue", "red"))
```

The exponential and natural logarithm functions have the following
properties:

1.  $\ln(x\cdot y)=\ln(x)+\ln(y),$ $x,y>0$
2.  $\ln(x/y)=\ln(x)-\ln(y),$ $x,y>0$
3.  $\ln(x^{y})=y\ln(x),$ $x>0$
4.  $\frac{d\ln(x)}{dx}=\frac{1}{x},$ $x>0$
5.  $\frac{d}{dx}\ln(f(x))=\frac{1}{f(x)}\frac{d}{dx}f(x)$ (chain-rule)
6.  $e^{x}e^{y}=e^{x+y}$
7.  $e^{x}e^{-y}=e^{x-y}$
8.  $(e^{x})^{y}=e^{xy}$
9.  $e^{\ln(x)}=x$
10. $\frac{d}{dx}e^{x}=e^{x}$
11. $\frac{d}{dx}e^{f(x)}=e^{f(x)}\frac{d}{dx}f(x)$ (chain-rule)

\newpage{}

## Exercises

### Analytical Problems {.unnumbered}

```{r}
#| echo: false
suppressPackageStartupMessages(library(knitr))
```

::: {#exr-exercise0101}
In this question you will see how the future value of an investment,
$FV_n$ varies with the interest rate, $R$, and the investment horizon,
$n$. Let the initial amount invested be $V=\$1,000$. For a grid of
values $R=0.01, 0.02, \ldots , 0.10$ and $n=1,2,\ldots 10$, compute and
plot $FV_n$.
:::

::: {#exr-exercise0102}
Consider the following credit card offer from a hypothetical bank. The
card advertises an annual percentage rate (APR) of $19.34\%$. However,
the bank actually compounds interest charges on a daily basis.

1.  Assuming 365 days in the year, what is the daily periodic interest
    rate being charged?
2.  If you carried a \$1,000 balance throughout the year, how much would
    you owe at the end of the year?
:::

::: {#exr-exercise0103}
Consider "payday loans". Payday loans are short-term loans made to
consumers, often for less than two weeks, and are offered by companies
such as AmeriCash Advance and National Payday. The loan works like this:
You write a check today that is postdated and the company gives you an
amount of money less than the check. When the check date arrives, you go
to the store and pay cash for the check, or the company cashes the
check. For example, suppose AmeriCash Advance allows you to write a
postdated check for \$125 for 15 days later in return for a \$100 loan
today.

1.  What is the annual percentage rate (APR) on this loan?
2.  What is the effective annual rate (EAR)?
:::

::: {#exr-exercise0104}
Consider the following (actual) monthly adjusted closing price data for
Starbucks stock over the period December 2004 through December 2005

```{r}
#| echo: false
sbuxData.df = data.frame(Date = c("December, 2004 ", "January, 2005", "February, 2005", "March, 2005",
                                  "April, 2005", "May, 2005", "June, 2005", "July, 2005", "August, 2005",
                                  "September, 2005", "October, 2005", "November, 2005", "December, 2005"),
                         Price = c(31.18, 27.00, 25.91, 25.83, 24.76, 27.40, 25.83, 26.27, 24.51, 25.05,
                                   28.28, 30.45,30.51))
kable(sbuxData.df, format = "markdown", caption = "End of Month Price Data for Starbucks Stock")
```

For the following questions, first do the return calculations by hand.
That is, give the appropriate return calculation formula and evaluate
the result. Then, check your hand calculations using the corresponding R
calculations.

1.  Using the data in the table, what is the simple monthly return
    between the end of December 2004 and the end of January 2005? If you
    invested \$10,000 in Starbucks at the end of December 2004, how much
    would the investment be worth at the end of January 2005?
2.  Using the data in the table, what is the continuously compounded
    monthly return between December 2004 and January 2005? Convert this
    continuously compounded return to a simple return (you should get
    the same answer as in part 1).
3.  Assuming that the simple monthly return you computed in part 1 is
    the same for 12 months, what is the annual return with monthly
    compounding?
4.  Assuming that the continuously compounded monthly return you
    computed in part 2 is the same for 12 months, what is the
    continuously compounded annual return?
5.  Using the data in the table, compute the actual simple annual return
    between December 2004 and December 2005. If you invested \$10,000 in
    Starbucks at the end of December 2004, how much would the investment
    be worth at the end of December 2005? Compare with your result in
    part 3.
6.  Using the data in the table, compute the actual annual continuously
    compounded return between December 2004 and December 2005. Compare
    with your result in part 4. Convert this continuously compounded
    return to a simple return (you should get the same answer as in part
    5).
:::

::: {#exr-exercise0105}
Consider a one month investment in two Northwest stocks: Amazon and
Costco. Suppose you buy Amazon and Costco at the end of September at
$P_{A,t-1}=\$38.23$, $P_{C,t-1}=\$41.11$ and then sell at the end of the
October for $P_{A,t}=\$41.29$ and $P_{C,t}=\$41.74$. (Note: these are
actual closing prices for 2004 taken from Yahoo!)

For the following questions, first do the return calculations by hand.
That is, give the appropriate return calculation formula and evaluate
the result. Then, check your hand calculations using the corresponding R
calculations.

1.  What are the simple monthly returns for the two stocks?
2.  What are the continuously compounded returns for the two stocks?
3.  Suppose Costco paid a \$0.10 per share cash dividend at the end of
    October. What is the monthly simple total return on Costco? What is
    the monthly dividend yield?
4.  Suppose the monthly returns on Amazon and Costco from part 1 above
    are the same every month for 1 year. Compute the simple annual
    returns as well as the continuously compounded annual returns for
    the two stocks.
5.  At the end of September 2004, suppose you have \$10,000 to invest in
    Amazon and Costco over the next month. If you invest \$8000 in
    Amazon and \$2000 in Costco, what are your portfolio shares, $x_{A}$
    and $x_{C}$?
6.  Continuing with part 5, compute the monthly simple return and the
    monthly continuously compounded return on the portfolio. Assume that
    Costco does not pay a dividend.
:::

::: {#exr-exercise0106}
Consider a 48-month (4 year) investment in two assets: the Vanguard S&P
500 index (VFINX) and Amazon stock (AMZN). Suppose you buy one share of
the S&P 500 fund and one share of Amazon stock at the end of January,
2015 for $P_{VFINX,t-48}=165$, $P_{AMZN,t-48}=355$, and then sell these
shares at the end of January, 2019 for $P_{VFINX,t}=241$
,$P_{AMZN,t}=1719$ (Note: these are actual adjusted closing prices taken
from Yahoo!). In this question, you will see how much money you could
have made if you invested in these assets right before the COVID-19
pandemic.

1.  What are the simple 48-month (4-year) returns for the two
    investments?

2.  What are the continuously compounded (cc) 48-month (4-year) returns
    for the two investments? Why are the cc returns smaller?

3.  Suppose you invested \$10,000 in each asset at the end of
    January, 2015. How much would each investment be worth at the end of
    January, 2019?

4.  What are the compound annual returns on the two 4-year investments?

5.  At the end of January, 2015, suppose you plan to invest in a
    portfolio of VFINX and AMZN over the next 48 months (4 years).
    Suppose you purchase 5 shares of the VFINX mutual fund (at
    \$165/share) and 10 shares of AMZN stock (at \$355/share). What is
    the value of your initial investment? What are the portfolio weights
    in the two assets as of the end of January, 2015?

6.  Using the results from part 1, compute the 4-year simple and cc
    portfolio returns.

7.  What is the value of your portfolio at the end of January, 2019?
    What is the value of your investments in VFINX and AMZN? What is
    share of wealth invested in VFINX and AMZN (i.e., your portfolio
    weights)?
:::

::: {#exr-exercise0107}
Consider a 60-month (5 year) investment in two assets: the Vanguard S&P
500 index (VFINX) and Apple stock (AAPL). Suppose you buy one share of
the S&P 500 fund and one share of Apple stock at the end of January,
2010 for $P_{VFINX,t-60}=\$89.91$, $P_{AAPL,t-60}=\$25.88$, and then
sell these shares at the end of January, 2015 for $P_{VFINX,t}=\$184.2$,
$P_{AAPL,t}=\$116.7$. (Note: these are actual adjusted closing prices
taken from Yahoo!). In this question, you will see how much money you
could have made if you invested in these assets right after the 2008
financial crisis.

1.  What are the simple 60-month (5-year) returns for the two
    investments?
2.  What are the continuously compounded 60-month (5-year) returns for
    the two investments?
3.  Are the simple and continuously compounded returns close to each
    other?
4.  Suppose you invested \$1,000 in each asset at the end of January,
5.  How much would each investment be worth at the end of January, 2015?
6.  What is the compound annual return on the two 5 year investments?
7.  At the end of January, 2010, suppose you have $\$1,000$ to invest in
    VFINX and AAPL over the next 60 months (5 years). Suppose you
    purchase $\$400$ worth of VFINX and the remainder in AAPL. What are
    the portfolio weights in the two assets? Using the results from
    parts 1 and 2, compute the 5-year simple and continuously compounded
    portfolio returns.
:::

::: {#exr-exercise0108}
Consider a one month investment in two Northwest stocks: Amazon and
Costco. Suppose you buy Amazon and Costco at the end of September at
$P_{A,t-1} = \$38.23, P_{C,t-1} = \$41.11$ and then sell at the end of
the October for $P_{A,t} = \$41.29, P_{C,t} = \$41.74$ . (Note: these
are actual closing prices for 2004 taken from Yahoo!)

1.  What are the simple monthly returns for the two stocks?
2.  What are the continuously compounded returns for the two stocks?
3.  Suppose Costco paid a $\$0.10$ per share cash dividend at the end of
    October. What is the monthly simple total return on Costco? What is
    the monthly dividend yield?
4.  Suppose the monthly returns on Amazon and Costco from question (a)
    above are the same every month for 1 year. Compute the simple annual
    returns as well as the continuously compounded annual returns for
    the two stocks.
5.  At the end of September, 2004, suppose you have $\$10,000$ to invest
    in Amazon and Costco over the next month. If you invest $\$8,000$ in
    Amazon and $\$2,000$ in Costco, what are your portfolio shares,
    $x_A \text{ and } x_C$. (Assume partial share purchases are
    possible)
6.  Continuing with the previous question, compute the monthly simple
    return and the monthly continuously compounded return on the
    portfolio. Assume that Costco does not pay a dividend.
:::

::: {#exr-exercise0109}
Consider an investment in a foreign stock (e.g., a stock trading on the
London stock exchange) by a U.S. national (domestic investor). The
domestic investor takes U.S. dollars, converts them to the foreign
currency (e.g. British Pound) via the exchange rate (price of foreign
currency in U.S. dollars) and then purchases the foreign stock using the
foreign currency. When the stock is sold, the proceeds in the foreign
currency must then be converted back to the domestic currency. To be
more precise, consider the information in the table below:

| Time | Cost of 1 Pound | Value of UK Shares | Value in U.S. $\$$ |
|:-----|:----------------|:-------------------|:-------------------|
| 0    | $\$1.50$        | $£40$              | $1.5\times40=60$   |
| 1    | $\$1.30$        | $£45$              | $1.3\times45=58.5$ |

1.  Compute the simple rate of return, $R_{e}$, from the prices of
    foreign currency. This is the return to the domestic investor of
    investing in the foreign currency.
2.  Compute the simple rate of return, $R_{UK}$, from the UK stock
    prices.
3.  Compute the simple rate of return, $R_{US}$, from the prices in US
    dollars.
4.  What is the mathematical relationship between $R_{US}$, $R_{UK}$ and
    $R_{e}$?
:::

::: {#exr-exercise0110}
Consider an investing in a long-only portfolio of two assets A and B for
one period between months $t-1$ and $t$. The beginning of period
investment values in A and B are $V_{A,t-1}>0$ and $V_{B,t-1}>0$, and
the one-month returns are $R_{A,t-1}$ and $R_{B,t-1}$. The initial value
of the portfolio is $V_{p,t-1}=V_{A,t-1}+V_{B,t-1}$. The initial
investment weights are $x_{A,t-1}=V_{A,t-1}/V_{p,t-1}>0$ and
$x_{B,t-1}=V_{B,t-1}/V_{p,t-1}>0$ such that $x_{A,t-1}+x_{B,t-1}=1$.

1.  Show that
    $$x_{A,t} = x_{A,t-1} \times \frac{1+R_{A,t-1}}{1+R_{p,t-1}},~ x_{B,t} = x_{B,t-1} \times \frac{1+R_{B,t-1}}{1+R_{p,t-1}},$$
    and that $x_{A,t}+x_{B,t}=1$.
2.  Show that the change in portfolio weights are
    \begin{align*}
    \Delta x_{A,t} = x_{A,t} - x_{A,t-1} =   \frac{x_{A,t-1}(R_{A,t-1}-R_{p,t-1})}{1+R_{p,t-1}}, \\
    \Delta x_{B,t} = x_{B,t} - x_{B,t-1} = \frac{x_{B,t-1}(R_{B,t-1}-R_{p,t-1})}{1+R_{p,t-1}}.
    \end{align*}
3.  Under what conditions will $\Delta x_{A,t} >0$, $\Delta x_{A,t} < 0$
    and $\Delta x_{A,t} =0$?
4.  Using $R_{p,t-1}=x_{A,t-1}R_{A,t-1}+x_{B,t-1}R_{B,t-1}$ show that
    $$\Delta x_{A,t} = \frac{x_{A,t-1}x_{B,t-1}(R_{A,t-1}-R_{B,t-1})}{1+x_{A,t-1}R_{A,t-1}+x_{B,t-1}R_{B,t-1}}$$
5.  Under what conditions will $\Delta x_{A,t} >0$, $\Delta x_{A,t} < 0$
    and $\Delta x_{A,t} =0$?
6.  To rebalance the portfolio at the end-of-month, you need to adjust
    the portfolio weights such that $x_{A,t} = x_{A,t-1}$ and
    $x_{B,t} = x_{B,t-1}$. If $R_{A,t-1} > R_{B,t-1}$ do you buy or sell
    asset A to rebalance the portfolio?
:::

### Return Calculations using R Packages {.unnumbered}

- Give examples of using the **PerformanceAnalytics package**
- Download different asset data from Yahoo!. Extract adjusted closing
  price data. Compute returns and plot returns and equity curves.
