# Review of Random Variables {#sec-Review-of-Random-Variables}

Updated: August 30, 2026

Copyright © Eric Zivot 2026


In this chapter I give a review of probability concepts that are necessary for
the probability modeling of asset returns and for the statistical analysis of financial data presented in this book. Much of this material is typically covered in an introductory probability and statistics course using calculus. In the course of
the review, I will present many examples related to finance and risk management and
some important risk concepts, such as value-at-risk, will be introduced.
In the examples, I will also show how to use R for probability calculations
and for working with probability distributions.

The chapter is outlined as follows. Section @sec-Random-Variables
reviews univariate random variables and properties of univariate probability
distributions. Important univariate distributions, such as the normal
distribution, that are used throughout the book for modeling financial
data, are described in detail. Section @sec-Bivariate-Distributions
covers bivariate probability distributions for two random variables. The concepts 
of dependence and independence between two random variables are discussed. The measures of linear dependence between two random variables, covariance and
correlation, are defined and the bivariate normal distribution is
introduced. Properties of linear combinations of two random variables
are given and illustrated using an example to describe the return
and risk properties of a portfolio of two assets. More than two random variables and
Multivariate distributions are briefly discussed in section @sec-Multivariate-Distributions.

The R packages I use in this chapter are **mvtnorm** and **sn**. Make
sure these packages are downloaded and installed prior to replicating
the R examples in this chapter.

```{r}
#| include: false
options(digits=3)
library(mvtnorm)
library(sn)
```

## Random Variables {#sec-Random-Variables}

I start with the basic definition of a *random variable*\index{Random variable}:

::: {#def-random-variable name="Random variable"}
A *random variable*, $X$, is a variable that can
take on a given set of values, called the *sample space* and denoted
$S_{X}$, where the likelihood of the values in $S_{X}$ is determined
by $X$'s *probability distribution function* (pdf). 
:::

::: {#exm-example0201 name="Future price of Microsoft stock"}
Consider the price of Microsoft stock next month, $P_{t+1}$. Since $P_{t+1}$ is not known with certainty today ($t$), you can consider it as a random variable. $P_{t+1}$ must be positive, and realistically it can't get too large. Therefore, the sample space is the set of positive real numbers bounded above by some large number:
$S_{P_{t+1}}=\{P_{t+1}:P_{t+1}\in[0,M],~M>0\}$. It is an open question as
to what is the best characterization of the probability distribution
of stock prices. The log-normal distribution is one possibility.^[If $P_{t+1}$ is a positive random variable such that $\ln P_{t+1}$ is normally distributed then $P_{t+1}$ has a log-normal distribution.]
\hfill{}$\blacksquare$
:::

::: {#exm-example0202 name="Return on Microsoft stock"}
Consider a one-month investment in Microsoft stock. That is, you buy
one share of Microsoft stock at the end of month $t-1$ (e.g., end
of February) and plan to sell it at the end of month $t$ (e.g., end
of March). The simple return over month $t$ on this investment, $R_{t}=(P_{t}-P_{t-1})/P_{t-1}$, is a random variable because you do not know what the price will be at the end of the month. In contrast to prices, returns can be positive or negative and are bounded from below by -100\%. You can express the
sample space as $S_{R_{t}}=\{R_{t}:R_{t}\in[-1,M],~M>0\}$. The normal
distribution is often a good approximation to the distribution of
simple monthly returns, and is a better approximation to the distribution
of continuously compounded monthly returns.
\hfill{}$\blacksquare$
:::

::: {#exm-example0203 name="Up-down indicator variable"}
As a final example, consider a variable $X_{t+1}$ defined to be equal to
one if the monthly price change on Microsoft stock, $P_{t+1}-P_{t}$,
is positive, and is equal to zero if the price change is zero or negative.
Here, the sample space is the set $S_{X_{t+1}}=\{0,1\}$. If it is equally
likely that the monthly price change is positive or negative (including
zero) then the probability that $X_{t+1}=1$ or $X_{t+1}=0$ is 0.5. This is an
example of a Bernoulli random variable.
\hfill{}$\blacksquare$
:::

The next sub-sections define discrete and continuous random variables.

### Discrete random variables
\index{Discrete random variables}
Consider a random variable generically denoted $X$ and its set of
possible values or sample space denoted $S_{X}$.

::: {#def-discrete-random-variable name="Discrete random variable"}
A *discrete random variable* $X$ is one that can
take on a finite number of $n$ different values $S_{X}=\{x_{1},x_{2},\ldots,x_{n}\}$
or, at most, a countably infinite number of different values $S_{X}=\{x_{1},x_{2},\ldots.\}$.
:::

::: {#def-pdf name="Probability density function"}
The pdf of a discrete random variable, denoted
$p(x)$, is a function such that $p(x)=\Pr(X=x)$. The pdf must satisfy
(i) $p(x)\geq0$ for all $x\in S_{X}$; (ii) $p(x)=0$ for all $x\notin S_{X}$;
and (iii) $\sum_{x\in S_{X}}p(x)=1$. 
:::

::: {#exm-example0204 name="Annual return on Microsoft stock"}
------------------------------------------------------
State of Economy        Sample Space        $\Pr(X=x)$
----------------        ------------        ----------
  Depression              -0.30               0.05

  Recession               0.0                 0.20

  Normal                  0.10                0.50

  Mild Boom               0.20                0.20

  Major Boom              0.50                0.05
------------------------------------------------------

: Probability distribution for the annual return on Microsoft. {#tbl-msftdiscretedistribution} 


Let $X$ denote the annual return on Microsoft stock over the next
year. You might hypothesize that the annual return will be influenced
by the general state of the economy. Consider five possible states
of the economy: depression, recession, normal, mild boom and major
boom. A stock analyst might forecast different values of the return
for each possible state. Hence, $X$ is a discrete random variable
that can take on five different values. Table @tbl-msftdiscretedistribution
describes such a probability distribution of the return. A graphical
representation of the probability distribution is created with the following R code:

```{r}
#| label: ProbRevMsftDiscreteDistn
#| echo: true
#| fig-cap: Discrete distribution for Microsoft stock.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

r.msft = c(-0.3, 0, 0.1, 0.2, 0.5)
prob.vals = c(0.05, 0.20, 0.50, 0.20, 0.05)
plot(r.msft, prob.vals, type="h", lwd=4, xlab="annual return", 
     ylab="probability", xaxt="n")
points(r.msft, prob.vals, pch=16, cex=2, col="blue")
axis(1, at=r.msft)
```

This distribution is presented in Figure @fig-ProbRevMsftDiscreteDistn. The shape of the distribution is revealing. The center and most likely value is at
$0.1$ and the distribution is symmetric about this value (has the same shape to the 
left and to the right). Most of the probability is concentrated between $0$ and $0.2$
and there is very little probability associated with very large or small returns. 
\hfill{}$\blacksquare$
:::

#### The Bernoulli distribution
\index{Bernoulli distribution}
Let $X=1$ if the price next month of Microsoft stock goes up and
$X=0$ if the price goes down (assuming it cannot stay the same).
Then $X$ is clearly a discrete random variable with sample space
$S_{X}=\{0,1\}$. If the probability of the stock price going up or
down is the same then $p(0)=p(1)=1/2$ and $p(0)+p(1)=1$.

The probability distribution described above can be given an exact
mathematical representation known as the *Bernoulli *distribution.
Consider two mutually exclusive events generically called "success"
and "failure". For example, a success could be a stock
price going up or a coin landing heads and a failure could be a stock
price going down or a coin landing tails. The process creating the
success or failure is called a *Bernoulli trial*. In general,
let $X=1$ if success occurs and let $X=0$ if failure occurs. Let
$\Pr(X=1)=\pi$, where $0<\pi<1$, denote the probability of success.
Then $\Pr(X=0)=1-\pi$ is the probability of failure. A mathematical
model describing this distribution is: 
$$
p(x)=\Pr(X=x)=\pi^{x}(1-\pi)^{1-x},~x=0,1.
$$ {#eq-bernoulli}
When $x=0$, $p(0)=\pi^{0}(1-\pi)^{1-0}=1-\pi$ and when $x=1,p(1)=\pi^{1}(1-\pi)^{1-1}=\pi$.

#### The binomial distribution
\index{Binomial distribution}

Consider a sequence of independent Bernoulli trials with success probability
$\pi$ generating a sequence of 0-1 variables indicating failures
and successes. A *binomial* random variable $X$ counts the number
of successes in $n$ Bernoulli trials, and is denoted $X\sim B(n,\pi)$.
The sample space is $S_{X}=\{0,1,\ldots,n\}$ and: 
$$
\Pr(X=k)=\binom{n}{k}\pi^{k}(1-\pi)^{n-k}.
$$
The term $\binom{n}{k}$ is the binomial coefficient, and counts the
number of ways $k$ objects can be chosen from $n$ distinct objects.
It is defined by: 
$$
\binom{n}{k}=\frac{n!}{(n-k)!k!},
$$
where $n!$ is the factorial of $n$, or $n(n-1)\cdots2\cdot1$.

### Continuous random variables
\index{Continuous random variables}

::: {#def-continuous-random-variable name="Continuous random variable"}
A *continuous random variable* X is one that can
take on any real value. That is, $S_{X}=\{x:x\in\mathbb{R}\}$. 
:::

::: {#def-pdf-continuous-random-variable name="Probability density function for continuous random variable"}
The probability density function (pdf) of a continuous
random variable $X$ is a nonnegative function $f,$ defined on the
real line, such that for any interval $A$: 
$$
\Pr(X\in A)=\int_{A}f(x)dx.
$$
That is, $\Pr(X\in A)$ is the area under the probability curve
over the interval $A$. The pdf $f(x)$ must satisfy
(i) $f(x)\geq0$; and (ii) $\int_{-\infty}^{\infty}f(x)dx=1$. 
:::

A typical "bell-shaped" pdf is displayed in Figure
@fig-ProbReviewAreaUnderCurve and the shaded area under
the curve between -2 and 1 represents $\Pr(-2\leq X \leq 1)$. For a continuous
random variable, $f(x)\neq\Pr(X=x)$ but rather gives the height of
the probability curve at $x$. In fact, $\Pr(X=x)=0$ for all values
of $x$. That is, probabilities are not defined over single points.
They are only defined over intervals. As a result, for a continuous
random variable $X$: 
$$
\Pr(a\leq X\leq b)=\Pr(a<X\leq b)=\Pr(a<X<b)=\Pr(a\leq X<b).
$$

```{r}
#| label: fig-ProbReviewAreaUnderCurve
#| echo: false
#| fig-cap: $Pr(-2 \leq X \leq 1)$ is represented by the area under the probability curve.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

lb = -2
ub = 1
x.vals = seq(-4, 4, length=150)
d.vals = dnorm(x.vals)
plot(x.vals, d.vals, type="l", xlab="x", ylab="pdf")
i = x.vals >= lb & x.vals <= ub
# add shaded region between -2 and 1
polygon(c(lb, x.vals[i], ub), c(0, d.vals[i], 0), col="red")
```


#### The uniform distribution on an interval
\index{Uniform distribution}
Let $X$ denote the annual return on Microsoft stock and let $a$
and $b$ be two real numbers such that $a<b$. Suppose that the annual
return on Microsoft stock can take on any value between $a$ and $b$.
That is, the sample space is restricted to the interval $S_{X}=\{x\in\mathcal{R}:a\leq x\leq b\}$.
Further suppose that the probability that $X$ will belong to any
subinterval of $S_{X}$ is proportional to the length of the interval.
In this case, it is said that $X$ is *uniformly distributed* on
the interval $[a,b]$. The pdf of $X$ has a very simple mathematical
form: 
$$
f(x)=\left\{ \begin{array}{c}
\frac{1}{b-a}\\
0
\end{array}\right.\begin{array}{c}
\textrm{for }a\leq x\leq b,\\
\textrm{otherwise,}
\end{array}
$$
and is presented graphically as a rectangle. It is easy to see that
the area under the curve over the interval $[a,b]$ (area of rectangle)
integrates to one: 
$$
\int_{a}^{b}\frac{1}{b-a}~dx=\frac{1}{b-a}\int_{a}^{b}~dx=\frac{1}{b-a}\left[x\right]_{a}^{b}=\frac{1}{b-a}[b-a]=1.
$$

::: {#exm-example205 name="Uniform distribution on [-1 1]"}
Let $a=-1$ and $b=1,$ so that $b-a=2$. Consider computing the probability
that $X$ will be between -0.5 and 0.5. You must solve: 
$$
\Pr(-0.5\leq X\leq0.5)=\int_{-0.5}^{0.5}\frac{1}{2}~dx=\frac{1}{2}\left[x\right]_{-0.5}^{0.5}=\frac{1}{2}\left[0.5-(-0.5)\right]=\frac{1}{2}.
$$
Next, consider computing the probability that the return will fall
in the interval $[0,\delta]$ where $\delta$ is some small number
less than $b=1$: 
$$
\Pr(0\leq X\leq\delta)=\frac{1}{2}\int_{0}^{\delta}dx=\frac{1}{2}[x]_{0}^{\delta}=\frac{1}{2}\delta.
$$
As $\delta\rightarrow0$, $\Pr(0\leq X\leq\delta)\rightarrow\Pr(X=0)$.

Using the above result notice that 
$$
\lim_{\delta\rightarrow0}\Pr(0\leq X\leq\delta)=\Pr(X=0)=\lim_{\delta\rightarrow0}\frac{1}{2}\delta=0.
$$
Hence, for continuous random variables probabilities are defined on
intervals but not at distinct points. 
\hfill{}$\blacksquare$
:::

#### The standard normal distribution
\index{Standard normal distribution}
The normal or Gaussian distribution is perhaps the most famous and
most useful continuous distribution in all of statistics. The shape
of the normal distribution is the familiar "bell curve".
As you shall see, it can be used to describe the probabilistic behavior
of stock returns although other distributions may be more appropriate.

If a random variable $X$ follows a *standard normal distribution*
then you often write $X\sim N(0,1)$ as short-hand notation. This distribution
is centered at zero and has inflection points at $\pm1$.^[An inflection point is a point where a curve goes from concave up to concave down, or vice versa.] The pdf of a standard normal random variable is given by: 
$$
f(x)=\frac{1}{\sqrt{2\pi}}\cdot e^{-\frac{1}{2}x^{2}},~-\infty\leq x\leq\infty.
$$ {#eq-std-normal-density}
It can be shown via the change of variables formula in calculus that
the area under the standard normal curve is one: 
$$
\int_{-\infty}^{\infty}\frac{1}{\sqrt{2\pi}}\cdot e^{-\frac{1}{2}x^{2}}dx=1.
$$
The standard normal distribution is illustrated in Figure @fig-probReviewStdNormalDensity. Notice that the distribution is *symmetric* about zero; i.e., the distribution has exactly the same form to the left and right of
zero. Because the standard normal pdf formula is used so often it
is given its own special symbol $\phi(x)$.

```{r}
#| label: fig-probReviewStdNormalDensity
#| echo: false
#| fig-cap: Standard normal density.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.vals = seq(-4, 4, length=150)
plot(x.vals, dnorm(x.vals), type="l", lwd=2, col="blue", xlab="x", ylab="pdf")
```


The normal distribution has the annoying feature that the area under
the normal curve cannot be evaluated analytically. That is: 
$$
\Pr(a\leq X\leq b)=\int_{a}^{b}\frac{1}{\sqrt{2\pi}}\cdot e^{-\frac{1}{2}x^{2}}~dx,
$$
does not have a closed form solution. The above integral must be computed
by numerical approximation. Areas under the normal curve, in one form
or another, are given in tables in almost every introductory statistics
book and standard statistical software can be used to find these areas.
Some useful approximate results are: 
\begin{align*}
\Pr(-1 & \leq X\leq1)\approx0.67,\\
\Pr(-2 & \leq X\leq2)\approx0.95,\\
\Pr(-3 & \leq X\leq3)\approx0.99.
\end{align*}


### The cumulative distribution function
\index{Cumulative distribution function}

::: {#def-cdf name="Cumulative distribution function"}
The *cumulative distribution function* (cdf) of
a random variable $X$ (discrete or continuous), denoted $F_{X}$,
is the probability that $X\leq x$: 
$$
F_{X}(x)=\Pr(X\leq x),~-\infty\leq x\leq\infty.
$$
:::

The cdf has the following properties: 

(i) If $x_{1}<x_{2}$ then $F_{X}(x_{1})\leq F_{X}(x_{2})$ 
(ii) $F_{X}(-\infty)=0$ and $F_{X}(\infty)=1$ 
(iii) $\Pr(X>x)=1-F_{X}(x)$ 
(iv) $\Pr(x_{1}<X\leq x_{2})=F_{X}(x_{2})-F_{X}(x_{1})$ 
(v) $F_{X}^{\prime}(x)=\frac{d}{dx}F_{X}(x)=f(x)$ if $X$ is a continuous random variable and $F_{X}(x)$ is continuous and differentiable. 


::: {#exm-example206 name="CDF for a discrete random variable"}
The cdf for the discrete distribution of Microsoft from Table @tbl-msftdiscretedistribution
is given by: 
$$
F_{X}(x)=\left\{ \begin{array}{c}
0,\\
0.05,\\
0.25,\\
0.75\\
0.95\\
1
\end{array}\right.\begin{array}{c}
x<-0.3\\
-0.3\leq x<0\\
0\leq x<0.1\\
0.1\leq x<0.2\\
0.2\leq x<0.5\\
x>0.5
\end{array}
$$
and is illustrated in Figure @fig-figProbReviewDiscreteCDF.
\hfill{}$\blacksquare$
:::

```{r}
#| label: figProbReviewDiscreteCDF
#| echo: false
#| fig-cap: CDF of Discrete Distribution for Microsoft Stock Return.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

cdf = c(0, 0.05, 0.25, 0.75, 0.95, 1)
x.vals = c(-0.4, -0.3, 0, 0.1, 0.2, 0.5)
plot(x.vals, cdf, lwd=4, type="S", xaxt="n")
axis(1, at=x.vals)
```


::: {#exm-example207 name="CDF for a uniform random variable"}
The cdf for the uniform distribution over $[a,b]$ can be determined
analytically: 
\begin{align*}
F_{X}(x) & =\Pr(X<x)=\int_{-\infty}^{x}f(t)~dt\\
 & =\frac{1}{b-a}\int_{a}^{x}~dt=\frac{1}{b-a}[t]_{a}^{x}=\frac{x-a}{b-a}.
\end{align*}
You can determine the pdf of $X$ directly from the cdf via, 
$$
f(x)=F_{X}^{\prime}(x)=\frac{d}{dx}F_{X}(x)=\frac{1}{b-a}.
$$
\hfill{}$\blacksquare$
:::

::: {#exm-example208 name="CDF for a standard normal random variable"}
The cdf of standard normal random variable $X$ is used so often in
statistics that it is given its own special symbol: 
$$
\Phi(x)=F_{X}(x)=\int_{-\infty}^{x}\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}z^{2}}~dz.
$$ {#eq-standard-normal-cdf}
The cdf $\Phi(x)$, however, does not have an analytic representation
like the cdf of the uniform distribution and so the integral in @eq-standard-normal-cdf
must be approximated using numerical techniques. A graphical representation
of $\Phi(x)$ is given in Figure @fig-probReviewStdNormalCdf.
Because the standard normal pdf, $\phi(x)$, is symmetric about zero
and bell-shaped the standard normal CDF, $\Phi(x)$, has an "S"
shape where the middle of the "S" is at zero. 
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-probReviewStdNormalCdf
#| echo: false
#| fig-cap: Standard normal cdf $\Phi(x)$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.vals = seq(-4, 4, length=150)
plot(x.vals, pnorm(x.vals), type="l", lwd=2, col="blue", xlab="x", ylab="CDF")
```


### Quantiles of the distribution of a random variable
\index{Quantiles of distributions}

::: {#def-quantiles name="Quantiles of the distribution of a random variable"}
Consider a random variable $X$ with continuous cdf $F_{X}(x)$. For
$0\leq\alpha\leq1$, the $100\cdot\alpha\%$ *quantile* of the
distribution for $X$ is the value $q_{\alpha}$ that satisfies: 
$$
F_{X}(q_{\alpha})=\Pr(X\leq q_{\alpha})=\alpha.
$$
For example, the 5\% quantile of $X$, $q_{0.05}$, satisfies, 
$$
F_{X}(q_{0.05})=\Pr(X\leq q_{0.05})=0.05.
$$
:::

The *median*\index{Median} of the distribution is the 50\% quantile. That is, the
median, $q_{0.5}$, satisfies, 
$$
F_{X}(q_{0.5})=\Pr(X\leq q_{0.5})=0.5.
$$

If $F_{X}$ is invertible then $q_{\alpha}$ may be determined analytically as: 
$$
q_{\alpha}=F_{X}^{-1}(\alpha)
$$ {#eq-quantile-function}
where $F_{X}^{-1}$ denotes the inverse function of $F_{X}$.^[The inverse function of $F_{X}$ , 
denoted $F_{X}^{-1},$ has the property that $F_{X}^{-1}(F_{X}(x))=x.$ $F_{X}^{-1}$ will exist
if $F_{X}$ is strictly increasing and is continuous.] Hence,
the 5\% quantile and the median may be determined from 
$$
q_{0.05}=F_{X}^{-1}(.05),q_{0.5}=F_{X}^{-1}(.5).
$$
The inverse cdf $F_{X}^{-1}$ is sometimes called the *quantile function*\index{Quantile function}.

::: {#exm-example208 name="Quantiles from a uniform distribution"}
Let $X\sim U[a,b]$ where $b>a$. Recall, the cdf of $X$ is given
by: 
$$
F_{X}(x)=\frac{x-a}{b-a},~a\leq x\leq b,
$$
which is continuous and strictly increasing. Given $\alpha\in[0,1]$
such that $F_{X}(x)=\alpha$, solving for $x$ gives the inverse cdf:
$$
x=F_{X}^{-1}(\alpha)=\alpha(b-a)+a.
$$ {#eq-uniform-quantile}
Using @eq-uniform-quantile, the 5\% quantile and median, for
example, are given by: 
\begin{align*}
q_{0.05} & =F_{X}^{-1}(.05)=.05(b-a)+a=.05b+.95a,\\
q_{0.5} & =F_{X}^{-1}(.5)=.5(b-a)+a=.5(a+b).
\end{align*}
If $a=0$ and $b=1$, then $q_{0.05}=0.05$ and $q_{0.5}=0.5$.
\hfill{}$\blacksquare$
:::

::: {#exm-example209 name="Quantiles from a standard normal distribution"}
Let $X\sim N(0,1).$ The quantiles of the standard normal distribution
are determined by solving 
$$
q_{\alpha}=\Phi^{-1}(\alpha),
$$ {#eq-standard-normal-quantile}
where $\Phi^{-1}$ denotes the inverse of the cdf $\Phi$. This inverse
function must be approximated numerically and is available in most
spreadsheets and statistical software. Using the numerical approximation
to the inverse function, the 1\%, 2.5\%, 5\%, 10\% quantiles and median
are given by: 
\begin{align*}
q_{0.01} & =\Phi^{-1}(.01)=-2.33,~q_{0.025}=\Phi^{-1}(.025)=-1.96,\\
q_{0.05} & =\Phi^{-1}(.05)=-1.645,~q_{0.10}=\Phi^{-1}(.10)=-1.28,\\
q_{0.5} & =\Phi^{-1}(.5)=0.
\end{align*}
Often, the standard normal quantile is denoted $z_{\alpha}$.
\hfill{}$\blacksquare$
:::

### R functions for discrete and continuous distributions

R has built-in functions for a number of discrete and continuous distributions
that are commonly used in probability modeling and statistical analysis.
These are summarized in Table @tbl-probabilityDistributions.
For each distribution, there are four functions starting with `d`,
`p`, `q`, and `r` that compute `d`ensity
(pdf) values, cumulative `p`robabilities (cdf), `q`uantiles
(inverse cdf) and `r`andom draws, respectively. Consider, for
example, the functions associated with the normal distribution. The
functions `dnorm()`, `pnorm()` and `qnorm()`
evaluate the standard normal density @eq-std-normal-density,
the cdf @eq-standard-normal-cdf, and the inverse cdf or quantile
function @eq-standard-normal-quantile, respectively, with
the default values `mean = 0` and `sd = 1`. The function
`rnorm()` returns a specified number of simulated values from
the normal distribution. 

----------------------------------------------------------------------------------------
**Distribution**        **Function (root)**     Parameters              Defaults
------------------      ---------------------   --------------------    ----------------
beta                    `beta`                  shape1, shape2          \_, \_

binomial                `binom`                 size, prob              \_, \_

cauchy                  `cauchy`                location, scale         0, 1 

chi-squared             `chisq`                 df, ncp                 \_, 1 

F                       `f`                     df1, df2                \_, \_

gamma                   `gamma`                 shape, rate, scale      \_, 1, 1/rate

geometric               `geom`                  prob                    \_

hyper-geometric         `hyper`                 m, n, k                 \_, \_, \_

log-normal              `lnorm`                 meanlog, sdlog          0, 1 

logistic                `logis`                 location, scale         0, 1 

negative binomial       `nbinom`                size, prob, mu          \_,\_,\_

normal                  `norm`                  mean, sd                0, 1

Poisson                 `pois`                  Lambda                  1

Student's t             `t`                     df, ncp                 \_, 1

uniform                 `unif`                  min, max                0, 1

Weibull                 `weibull`               shape, scale            \_, 1 

Wilcoxon                `wilcoxon`              m, n                    \_, \_
----------------------------------------------------------------------------------------

: Probability distributions in base R. {#tbl-probabilityDistributions} 

Commonly used distributions used in finance application are the binomial, chi-squared,
log normal, normal, and Student's t. 

#### Finding areas under the standard normal curve

Let $X$ denote a standard normal random variable. Given that the
total area under the normal curve is one and the distribution is symmetric
about zero, the following results hold: 

-  $\Pr(X\leq z)=1-\Pr(X\geq z)$ and $\Pr(X\geq z)=1-\Pr(X\leq z)$ 
-  $\Pr(X\geq z)=\Pr(X\leq-z)$ 
-  $\Pr(X\geq0)=\Pr(X\leq0)=0.5$ 

The following examples show how to do probability calculations with
standard normal random variables.

::: {#exm-example210 name="Finding areas under the normal curve using R"}
First, consider finding $\Pr(X\geq2)$. By the symmetry of the normal
distribution, $\Pr(X\geq2)=\Pr(X\leq-2)=\Phi(-2)$. In R use:

```{r}
#| echo: true

pnorm(-2)
```

Next, consider finding $\Pr(-1\leq X\leq2)$. Using the cdf, you compute
$\Pr(-1\leq X\leq2)=\Pr(X\leq2)-\Pr(X\leq-1)=\Phi(2)-\Phi(-1)$. In
R use:

```{r}
#| echo: true

pnorm(2) - pnorm(-1)
```

Finally, using R the exact values for $\Pr(-1\leq X\leq1)$, $\Pr(-2\leq X\leq2)$
and $\Pr(-3\leq X\leq3)$ are:

```{r}
#| echo: true

pnorm(1) - pnorm(-1)
pnorm(2) - pnorm(-2)
pnorm(3) - pnorm(-3)
```

\hfill{}$\blacksquare$
:::

#### Plotting distributions

When working with a probability distribution, it is a good idea to
make plots of the pdf or cdf to reveal important characteristics.
The following examples illustrate plotting distributions using R.

::: {#exm-example211 name="Plotting the standard normal curve"}
The graphs of the standard normal pdf and cdf in Figures @fig-probReviewStdNormalDensity
and @fig-probReviewStdNormalCdf were created using the following
R code:

```{r}
#| echo: true
#| eval: false
# plot pdf
x.vals = seq(-4, 4, length=150)
plot(x.vals, dnorm(x.vals), type="l", lwd=2, col="blue", 
     xlab="x", ylab="pdf")
# plot cdf
plot(x.vals, pnorm(x.vals), type="l", lwd=2, col="blue", 
     xlab="x", ylab="CDF")
```

\hfill{}$\blacksquare$
:::

::: {#exm-example212 name="Shading a region under the standard normal curve"}
Figure @fig-ProbReviewAreaUnderCurve showing $\Pr(-2\leq X\leq1)$
as a red shaded area is created with the following code:

```{r}
#| echo: true
#| eval: false
lb = -2
ub = 1
x.vals = seq(-4, 4, length=150)
d.vals = dnorm(x.vals)
# plot normal density
plot(x.vals, d.vals, type="l", xlab="x", ylab="pdf")
i = x.vals >= lb & x.vals <= ub
# add shaded region between -2 and 1
polygon(c(lb, x.vals[i], ub), c(0, d.vals[i], 0), col="red")
```

\hfill{}$\blacksquare$
:::

### Shape characteristics of probability distributions

Very often you would like to know certain shape characteristics of
a probability distribution. You might want to know where the distribution
is centered, and how spread out the distribution is about the central
value. You might want to know if the distribution is symmetric about
the center or if the distribution has a long left or right tail. For
stock returns you might want to know about the likelihood of observing
extreme values for returns representing market crashes. This means
that you would like to know about the amount of probability in the
extreme tails of the distribution. In this section I will discuss four
important shape characteristics of a probability distribution: 

1.  *expected value* (mean): measures the center of mass of a distribution 
2.  *variance* and *standard deviation* (volatility): measures the spread about the mean 
3.  *skewness*: measures symmetry about the mean 
4.  *kurtosis*: measures "tail thickness"


#### Expected Value
\index{Expected value}

The *expected value* of a random variable $X,$ denoted $E[X]$ or $\mu_{X}$,
measures the center of mass of the pdf. 

::: {#def-expected-value name="Expected value of discrete random variable"}
For a discrete random variable $X$ with sample space $S_{X}$, the expected value is defined as:
$$
\mu_{X}=E[X]=\sum_{x\in S_{X}}x\cdot\Pr(X=x).
$$ {#eq-expected-value-discrete}
:::

Equation @eq-expected-value-discrete shows that $E[X]$ is
a probability weighted average of the possible values of $X$.

::: {#exm-example213 name="Expected value of discrete random variable"}
Using the discrete distribution for the return on Microsoft stock
in Table @tbl-Table-msftdiscretedistribution, the expected return
is computed as: 
\begin{align*}
E[X]  =&(-0.3)\cdot(0.05)+(0.0)\cdot(0.20)+(0.1)\cdot(0.5)\\
&+(0.2)\cdot(0.2)+(0.5)\cdot(0.05)\\
  =&0.10.
\end{align*}

\hfill{}$\blacksquare$
:::

::: {#exm-example214 name="Expected value of Bernoulli and binomial random variables"}
Let $X$ be a Bernoulli random variable with success probability $\pi$.
Then, 
$$
E[X]=0\cdot(1-\pi)+1\cdot\pi=\pi.
$$
That is, the expected value of a Bernoulli random variable is its
probability of success. Now, let $Y\sim B(n,\pi)$. It can be shown
that: 
$$
E[Y]=\sum_{k=0}^{n}k\binom{n}{k}\pi^{k}(1-\pi)^{n-k}=n\pi.
$$ {#eq-binomial-expected-value}

\hfill{}$\blacksquare$
:::

::: {#def-expected-value-continuous name="Expected value of a continuous random variable"}
For a continuous random variable $X$ with pdf $f(x)$, the *expected value* is defined as: 
$$
\mu_{X}=E[X]=\int_{-\infty}^{\infty}x\cdot f(x)~dx.
$$ {#eq-expected-value-cts}
:::

::: {#exm-example215 name="Expected value of a uniform random variable"} 
Suppose $X$ has a uniform distribution over the interval $[a,b]$.
Then, 
\begin{align*}
E[X] & =\frac{1}{b-a}\int_{a}^{b}x~dx=\frac{1}{b-a}\left[\frac{1}{2}x^{2}\right]_{a}^{b}\\
 & =\frac{1}{2(b-a)}\left[b^{2}-a^{2}\right]\\
 & =\frac{(b-a)(b+a)}{2(b-a)}=\frac{b+a}{2}.
\end{align*}
If $b=-1$ and $a=1$, then $E[X]=0$.
\hfill{}$\blacksquare$
:::

::: {#exm-example216 name="Expected value of a standard normal random variable"}
Let $X\sim N(0,1)$. Then it can be shown that: 
$$
E[X]=\int_{-\infty}^{\infty}x\cdot\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^{2}}~dx=0.
$$
Hence, the standard normal distribution is centered at zero.
\hfill{}$\blacksquare$
:::

#### Expectation of a function of a random variable
\index{Expectation}
The other shape characteristics of the distribution of a random variable
$X$ are based on expectations of certain functions of $X$. 

::: {#def-expected-value-function name="Expected value of a function of a random variable"}
Let $g(X)$ denote some function of the random variable $X$. If $X$ is a discrete
random variable with sample space $S_{X}$ then: 
$$
E[g(X)]=\sum_{x\in S_{X}}g(x)\cdot\Pr(X=x),
$$ {#eq-expectation-g-discrete}
and if $X$ is a continuous random variable with pdf $f$ then, 
$$
E[g(X)]=\int_{-\infty}^{\infty}g(x)\cdot f(x)~dx.
$$ {#eq-expectation-g-continuous}
:::

#### Variance and standard deviation
\index{Variance} \index{Standard deviation}
The variance of a random variable $X$, denoted $\mathrm{var}(X)$
or $\sigma_{X}^{2}$, measures the spread of the distribution about
the mean using the function $g(X)=(X-\mu_{X})^{2}$. If most values
of $X$ are close to $\mu_{X}$ then on average $(X-\mu_{X})^{2}$
will be small. In contrast, if many values of $X$ are far below and/or
far above $\mu_{X}$ then on average $(X-\mu_{X})^{2}$ will be large.
Squaring the deviations about $\mu_{X}$ guarantees a positive value.

::: {#def-variance name="Variance and standard deviation"}
For a discrete or continuous random variable $X$, the variance and standard deviation
of $X$ are defined as: 
$$
\sigma_{X}^{2}=\mathrm{var}(X)=E[(X-\mu_{X})^{2}],
$$ {#eq-variance}
$$
\sigma_{X} = \mathrm{sd}(X)= \sqrt{\sigma_{X}^{2}} = \sigma_{X}. 
$$ {#eq-standard-deviation}
:::

Because $\sigma_{X}^{2}$ represents an average squared deviation,
it is not in the same units as $X$. The standard deviation of $X$
is the square root of the variance and is in the same units as $X$. For "bell-shaped"
distributions, $\sigma_{X}$ measures the typical size of a deviation
from the mean value. 

The computation of @eq-variance can often be simplified by using the result: 
$$
\mathrm{var}(X)=E[(X-\mu_{X})^{2}]=E[X^{2}]-\mu_{X}^{2}
$$ {#eq-variance-2}
You will learn how to derive this result later in the chapter. 

::: {#exm-example217 name="Variance and standard deviation for a discrete random variable"}
Using the discrete distribution for the return on Microsoft stock
in Table @tbl-msftdiscretedistribution and the result that $\mu_{X}=0.1$, it follows that: 
\begin{align*}
\mathrm{var}(X) & =(-0.3-0.1)^{2}\cdot(0.05)+(0.0-0.1)^{2}\cdot(0.20)+(0.1-0.1)^{2}\cdot(0.5)\\
 & +(0.2-0.1)^{2}\cdot(0.2)+(0.5-0.1)^{2}\cdot(0.05)\\
 & =0.020.
\end{align*}
Alternatively, you can compute $\mathrm{var}(X)$ using @eq-variance-2:
\begin{align*}
E[X^{2}]-\mu_{X}^{2} & =(-0.3)^{2}\cdot(0.05)+(0.0)^{2}\cdot(0.20)+(0.1)^{2}\cdot(0.5)\\
 & +(0.2)^{2}\cdot(0.2)+(0.5)^{2}\cdot(0.05)-(0.1)^{2}\\
 & =0.020.
\end{align*}
The standard deviation is $\sqrt{0.020}=0.141$.
Given that the distribution is fairly bell-shaped you can say that
typical values deviate from the mean value of $10\%$ by about $\pm 14.1\%$.
\hfill{}$\blacksquare$
:::

::: {#exm-example218 name="Variance and standard deviation of a Bernoulli and binomial random variables"}
Let $X$ be a Bernoulli random variable with success probability $\pi$.
Given that $\mu_{X}=\pi$ it follows that: 
\begin{align*}
\mathrm{var}(X) & =(0-\pi)^{2}\cdot(1-\pi)+(1-\pi)^{2}\cdot\pi\\
 & =\pi^{2}(1-\pi)+(1-\pi)^{2}\pi\\
 & =\pi(1-\pi)\left[\pi+(1-\pi)\right]\\
 & =\pi(1-\pi),\\
\mathrm{sd}(X) & =\sqrt{\pi(1-\pi)}.
\end{align*}
Now, let $Y\sim B(n,\pi)$: It can be shown that: 
$$
\mathrm{var}(Y)=n\pi(1-\pi)
$$
and so $\mathrm{sd}(Y)=\sqrt{n\pi(1-\pi)}$.
\hfill{}$\blacksquare$
:::

::: {#exm-example219 name="Variance and standard deviation of a uniform random variable"}
Let $X\sim U[a,b]$. Using @eq-variance-2 and $\mu_{X}=\frac{a+b}{2}$,
after some algebra, it can be shown that: 
$$
\mathrm{var}(X)=E[X^{2}]-\mu_{X}^{2}=\frac{1}{b-a}\int_{a}^{b}x^{2}~dx-\left(\frac{a+b}{2}\right)^{2}=\frac{1}{12}(b-a)^{2},
$$
and $\mathrm{sd}(X)=(b-a)/\sqrt{12}$.
\hfill{}$\blacksquare$
:::

::: {#exm-example220 name="Variance and standard deviation of a standard normal random variable"}
Let $X\sim N(0,1)$. Here, $\mu_{X}=0$ and it can be shown that:
$$
\sigma_{X}^{2}=\int_{-\infty}^{\infty}x^{2}\cdot\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^{2}}~dx=1.
$$
It follows that sd$(X)=1$.
\hfill{}$\blacksquare$
:::

#### The general normal distribution
\index{General normal distribution}

Recall, if $X$ has a standard normal distribution then $E[X]=0$,
$\mathrm{var}(X)=1$. A *general normal* random variable $X$ has $E[X]=\mu_{X}$
and $\mathrm{var}(X)=\sigma_{X}^{2}$ and is denoted $X\sim N(\mu_{X},\sigma_{X}^{2})$.
Its pdf is given by: 
$$
f(x)=\frac{1}{\sqrt{2\pi\sigma_{X}^{2}}}\exp\left\{ -\frac{1}{2\sigma_{X}^{2}}\left(x-\mu_{X}\right)^{2}\right\} ,~-\infty\leq x\leq\infty.
$$ {#eq-general-normal-pdf}
Showing that $E[X]=\mu_{X}$ and $\mathrm{var}(X)=\sigma_{X}^{2}$
is a bit of work and is good calculus practice. As with the standard
normal distribution, areas under the general normal curve cannot be
computed analytically. Using numerical approximations, it can be shown
that: 
\begin{align*}
\Pr(\mu_{X}-\sigma_{X} & <X<\mu_{X}+\sigma_{X})\approx0.67,\\
\Pr(\mu_{X}-2\sigma_{X} & <X<\mu_{X}+2\sigma_{X})\approx0.95,\\
\Pr(\mu_{X}-3\sigma_{X} & <X<\mu_{X}+3\sigma_{X})\approx0.99.
\end{align*}
Hence, for a general normal random variable about 95\% of the time
you expect to see values within $\pm$ 2 standard deviations from its
mean. Observations more than three standard deviations from the mean
are very unlikely.

::: {#exm-example221 name="Normal distribution for monthly returns"}
Let $R$ denote the monthly return on an investment in Microsoft stock,
and assume that it is normally distributed with mean $\mu_{R}=0.01$
and standard deviation $\sigma_{R}=0.10$. That is, $R\sim N(0.01,(0.10)^{2})$.
Notice that $\sigma_{R}^{2}=0.01$ and is not in units of return per
month. Figure @fig-ProbReviewNormalReturns illustrates the
distribution and is created using:

```{r}
#| label: fig-ProbReviewNormalReturns
#| echo: true
#| fig-cap: 'Normal distribution for the monthly returns on Microsoft: $R\sim N(0.01,(0.10)^{2})$.'
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

mu.r = 0.01
sd.r = 0.1
x.vals = seq(-4, 4, length=150)*sd.r + mu.r
plot(x.vals, dnorm(x.vals, mean=mu.r, sd=sd.r), type="l", lwd=2, 
     col="blue", xlab="x", ylab="pdf")
```

Notice that essentially all of the probability lies between $-0.3$
and $0.3$. Using the R function `pnorm()`, you can easily compute
the probabilities $\Pr(R<-0.5)$, $\Pr(R<0)$, $\Pr(R>0.5)$ and $\Pr(R>1)$:

```{r}
#| echo: true

pnorm(-0.5, mean=0.01, sd=0.1)
pnorm(0, mean=0.01, sd=0.1)
1 - pnorm(0.5, mean=0.01, sd=0.1)
1 - pnorm(1, mean=0.01, sd=0.1)
```

Using the R function `qnorm()`, you can find the quantiles $q_{0.01}$,
$q_{0.05}$, $q_{0.95}$ and $q_{0.99}$:

```{r}
#| echo: true

a.vals = c(0.01, 0.05, 0.95, 0.99)
qnorm(a.vals, mean=0.01, sd=0.10)
```

Hence, over the next month, there are 1\% and 5\% chances of losing
more than 22.2\% and 15.5\%, respectively. In addition, there are
5\% and 1\% chances of gaining more than 17.5\% and 24.3\%, respectively.
\hfill{}$\blacksquare$
:::

::: {#exm-example222 name="Risk-return tradeoff"}
Consider the following investment problem. You can invest in two non-dividend
paying stocks, Amazon and Boeing, over the next month. Let $R_{A}$
denote the monthly return on Amazon and $R_{B}$ denote the monthly
return on Boeing. Assume that $R_{A}\sim N(0.02,(0.10)^{2})$ and
$R_{B}\sim N(0.01,(0.05)^{2})$. Figure @fig-probreviewRiskReturn
shows the pdfs for the two returns. Notice that $\mu_{A}=0.02>\mu_{B}=0.01$
but also that $\sigma_{A}=0.10>\sigma_{B}=0.05$. The return you expect
on Amazon is bigger than the return you expect on Boeing but the variability
of Amazon's return is also greater than the variability of Boeing's
return. The high return variability (volatility) of Amazon reflects
the higher risk associated with investing in Amazon compared to investing
in Boeing. If you invest in Boeing you get a lower expected return,
but you also get less return variability or risk. For example, the probability
of losing more than $10\%$ in one month for Amazon is
```{r}
pnorm(-0.1, 0.02, 0.1)
```
and the probability of losing more than $10\%$ in one month for Boeing is
```{r}
pnorm(-0.1, 0.01, 0.05)
```
Of course, there is also a higher probability for upside return for Amazon than for Boeing. 
This example illustrates the fundamental "no free lunch" principle of economics
and finance: you can't get something for nothing. In general, to get
a higher expected return you must be prepared to take on higher risk.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-probreviewRiskReturn
#| echo: false
#| fig-cap: Risk-return tradeoff between one-month investments in Amazon and Boeing stock.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
mu.r = 0.02
sd.r = 0.10
x.vals = seq(-3, 3, length=150)*sd.r + mu.r
plot(x.vals, dnorm(x.vals, mean=mu.r, sd=sd.r), type="l", lwd=2, 
     ylim=c(0, max(dnorm(x.vals, mean=0.01, sd=0.05))),
     col="black", xlab="x", ylab="pdf")
points(x.vals, dnorm(x.vals, mean=0.01, sd=0.05), type="l", lwd=2,
       col="blue", lty="dotted")
segments(0.02, 0, 0.02, dnorm(0.02, mean=0.02, sd=0.1), lwd=2)
segments(0.01, 0, 0.01, dnorm(0.01, mean=0.01, sd=0.05), lwd=2, 
         col="blue", lty="dotted")
legend(x="topleft", legend=c("Amazon", "Boeing"), lwd=2,
       col=c("black", "blue"), lty=c("solid","dotted"))
```


::: {#exm-example223 name="Why the normal distribution may not be appropriate for simple returns"}
Let $R_{t}$ denote the simple annual return on an asset, and suppose
that $R_{t}\sim N(0.05,(0.50)^{2})$. Because asset prices must be
non-negative, $R_{t}$ must always be larger than $-1$. However,
the normal distribution is defined for $-\infty\leq R_{t}\leq\infty$
and based on the assumed normal distribution $\Pr(R_{t}<-1)=0.018$.
That is, there is a 1.8\% chance that $R_{t}$ is smaller than $-1$.
This implies that there is a 1.8\% chance that the asset price at
the end of the year will be negative! This is one reason why the normal distribution
may not be appropriate for simple returns.
\hfill{}$\blacksquare$
:::

::: {#exm-example224 name="The normal distribution is more appropriate for continuously compounded returns"}
Let $r_{t}=\ln(1+R_{t})$ denote the continuously compounded annual
return on an asset, and suppose that $r_{t}\sim N(0.05,(0.50)^{2})$.
Unlike the simple return, the continuously compounded return can take
on values less than $-1$. In fact, $r_{t}$ is defined for $-\infty\leq r_{t}\leq\infty$.
For example, suppose $r_{t}=-2$. This implies a simple return of
$R_{t}=e^{-2}-1=-0.865$.\footnote{If $r_{t}=-\infty$ then $R_{t}=e^{r_{t}}-1=e^{-\infty}-1=0-1=-1$.}
Then $\Pr(r_{t}\leq-2)=\Pr(R_{t}\leq-0.865)=0.00002$. Although the
normal distribution allows for values of $r_{t}$ smaller than $-1$,
the implied simple return $R_{t}$ will always be greater than $-1$.
\hfill{}$\blacksquare$
:::

#### The Log-Normal Distribution
\index{Log normal distribution}
Let $X$ $\sim N(\mu_{X},\sigma_{X}^{2})$, which is defined for $-\infty<X<\infty$.
The *log-normally* distributed random variable $Y$ is determined from
the normally distributed random variable $X$ using the transformation
$Y=e^{X}$. Here, $Y$ is log-normally distributed and you write: 
$$
Y\sim\ln N(\mu_{X},\sigma_{X}^{2}),~0<Y<\infty.
$$ {#eq-lognormal-rv}
The pdf of the log-normal distribution for $Y$ can be derived from
the normal distribution for $X$ using the change-of-variables formula
from calculus and is given by: 
$$
f(y)=\frac{1}{y\sigma_{X}\sqrt{2\pi}}e^{-\frac{(\ln y-\mu_{X})^{2}}{2\sigma_{X}^{2}}}.
$$ {#eq-lognormal-density}
Due to the exponential transformation, $Y$ is only defined for non-negative
values. It can be shown that: 
$$
\begin{aligned}
\mu_{Y} & =E[Y]=e^{\mu_{X}+\sigma_{X}^{2}/2},\\
\sigma_{Y}^{2} & =\mathrm{var}(Y)=e^{2\mu_{X}+\sigma_{X}^{2}}(e^{\sigma_{X}^{2}}-1).\nonumber 
\end{aligned}
$$ {#eq-log-normal-moments}

::: {#exm-example225 name="Log-normal distribution for simple returns"}
Let $r_{t}=\ln(P_{t}/P_{t-1})$ denote the continuously compounded
monthly return on an asset and assume that $r_{t}\sim N(0.05,(0.50)^{2})$.
That is, $\mu_{r}=0.05$ and $\sigma_{r}=0.50$. Let $R_{t}=(P_{t}-P_{t-1})/P_{t}$
denote the simple monthly return. The relationship between $r_{t}$
and $R_{t}$ is given by $r_{t}=\ln(1+R_{t})$ and $1+R_{t}=e^{r_{t}}$.
Since $r_{t}$ is normally distributed, $1+R_{t}$ is log-normally
distributed. Notice that the distribution of $1+R_{t}$ is only defined
for positive values of $1+R_{t}$. This is appropriate since the smallest
value that $R_{t}$ can take on is $-1$. Using @eq-log-normal-moments,
the mean and variance for $1+R_{t}$ are given by: 
\begin{align*}
\mu_{1+R} & =e^{0.05+(0.5)^{2}/2}=1.191\\
\sigma_{1+R}^{2} & =e^{2(0.05)+(0.5)^{2}}(e^{(0.5)^{2}}-1)=0.563
\end{align*}
The pdfs for $r_{t}$ and $R_{t}$ are shown in Figure @fig-figNormalLognormal.
\hfill{}$\blacksquare$
:::

```{r}
#| label: figNormalLognormal
#| echo: false
#| fig-cap: Normal distribution for $r_{t}$ and log-normal distribution for $R_{t}=e^{r_{t}}-1$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

mu.x = 0.05
sd.x = 0.5
x.vals = seq(from=(mu.x - 3*sd.x), to=(mu.x + 3*sd.x), length = 100)
plot(x.vals, dnorm(x.vals, mean=mu.x, sd=sd.x), type="l", lwd=2, 
     xlim = c(min(x.vals), max(exp(x.vals))),
     ylim = c(0, 0.9),
     xlab="R", ylab="pdf")
points(exp(x.vals)-1, dlnorm(exp(x.vals), mean=mu.x, sd=sd.x), type="l", 
       lwd = 2, col="blue", lty="dashed",
     xlab="1 + R", ylab="density", main="Log-Normal(0.05, (0.5)^2)")
legend(x="topright", legend=c("Normal", "Log-Normal"), lwd=2,
       col=c("black", "blue"), lty = c("solid", "dashed"))
```


#### Skewness
\index{Skewness}

::: {#def-skewness name="Skewness"}
The *skewness* of a random variable $X$, denoted $\mathrm{skew}(X)$,
measures the symmetry of a distribution about its mean value using
the function $g(X)=(X-\mu_{X})^{3}/\sigma_{X}^{3}$, where $\sigma_{X}^{3}$
is just $\mathrm{sd}(X)$ raised to the third power: 
$$
\mathrm{skew}(X)=\frac{E[(X-\mu_{X})^{3}]}{\sigma_{X}^{3}}
$$ {#eq-skewness}
:::

When $X$ is far below its mean, $(X-\mu_{X})^{3}$ is a big negative
number. When $X$ is far above its mean, $(X-\mu_{X})^{3}$ is
a big positive number. Hence, if there are more big values of $X$
below $\mu_{X}$ then $\mathrm{skew}(X)<0$. Conversely, if there
are more big values of $X$ above $\mu_{X}$ then $\mathrm{skew}(X)>0$.
If $X$ has a symmetric distribution, then $\mathrm{skew}(X)=0$ since
positive and negative values in @eq-skewness cancel out. If
$\mathrm{skew}(X)>0$, then the distribution of $X$ has a *long
right tail* and if $\mathrm{skew}(X)<0$ the distribution
of $X$ has a *long left tail*. These cases are illustrated
in Figure @fig-ProbReviewSkewedPdfs.


```{r}
#| label: fig-ProbReviewSkewedPdfs
#| echo: false
#| fig-cap: Skewed distributions.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

suppressPackageStartupMessages(library(sn))
par(mfrow=c(1,2)) 
# plot positive skewed distribution 
x.vals = seq(-1, 4, length.out=100) 
plot(x.vals, dsn(x.vals, xi=0, omega=1, alpha=5), main="Positively Skewed Distribution",       
     type="l", lwd=2, col="blue", xlab="X", ylab="pdf")
# plot negatively skewed distribution 
x.vals = seq(-4, 1, length.out=100) 
plot(x.vals, dsn(x.vals, xi=0, omega=1, alpha=-5), main="Negatively Skewed Distribution",       
     type="l", lwd=2, col="blue", xlab="X", ylab="pdf") 
par(mfrow=c(1,1))
```

::: {#exm-example226 name="Skewness for a discrete random variable"}
From the discrete distribution for the return on Microsoft stock
in Table @tbl-msftdiscretedistribution, and the results
$\mu_{X}=0.1$ and $\sigma_{X}=0.141$, it follows that: 
\begin{align*}
\mathrm{skew}(X) & =[(-0.3-0.1)^{3}\cdot(0.05)+(0.0-0.1)^{3}\cdot(0.20)+(0.1-0.1)^{3}\cdot(0.5)\\
 & +(0.2-0.1)^{3}\cdot(0.2)+(0.5-0.1)^{3}\cdot(0.05)]/(0.141)^{3}\\
 & =0.0.
\end{align*}
\hfill{}$\blacksquare$
:::

::: {#exm-example227 name="Skewness for a normal random variable"}
Suppose $X$ has a general normal distribution with mean $\mu_{X}$
and variance $\sigma_{X}^{2}$. Then it can be shown that: 
$$
\mathrm{skew}(X)=\int_{-\infty}^{\infty}\frac{(x-\mu_{X})^{3}}{\sigma_{X}^{3}}\cdot\frac{1}{\sqrt{2\pi\sigma^{2}}}e^{-\frac{1}{2\sigma_{X}^{2}}\left(x-\mu_{X}\right)^{2}}~dx=0.
$$
This result is expected since the normal distribution is symmetric
about it's mean value $\mu_{X}$.
\hfill{}$\blacksquare$
:::

::: {#exm-example228 name="Skewness for a log-Normal random variable"}
Let $Y=e^{X},$ where $X\sim N(\mu_{X},\sigma_{X}^{2})$, be a log-normally
distributed random variable with parameters $\mu_{X}$ and $\sigma_{X}^{2}$.
Then it can be shown that: 
$$
\mathrm{skew}(Y)=\left(e^{\sigma_{X}^{2}}+2\right)\sqrt{e^{\sigma_{X}^{2}}-1}>0.
$$
Notice that $\mathrm{skew}(Y)$ is always positive, indicating that
the distribution of $Y$ has a long right tail, and that it is an
increasing function of $\sigma_{X}^{2}$. This positive skewness is
illustrated in Figure @fig-figNormalLognormal.
\hfill{}$\blacksquare$
:::

#### Kurtosis
\index{Kurtosis}

::: {#def-kurtosis name="Kurtosis"}
The *kurtosis* of a random variable $X$, denoted $\mathrm{kurt}(X)$,
measures the thickness in the tails of a distribution and is based
on $g(X)=(X-\mu_{X})^{4}/\sigma_{X}^{4}$: 
$$
\mathrm{kurt}(X)=\frac{E[(X-\mu_{X})^{4}]}{\sigma_{X}^{4}},
$$ {#eq-kurtosis}
where $\sigma_{X}^{4}$ is just $\mathrm{sd}(X)$ raised to the fourth
power.
:::

Since kurtosis is based on deviations from the mean raised
to the fourth power, large deviations get lots of weight. Hence, distributions
with large kurtosis values are ones where there is the possibility
of extreme values. In contrast, if the kurtosis is small then most
of the observations are tightly clustered around the mean and there
is very little probability of observing extreme values. 

::: {#exm-example229 name="Kurtosis for a discrete random variable"}
From the discrete distribution for the return on Microsoft stock
in Table @tbl-msftdiscretedistribution, and the results that
$\mu_{X}=0.1$ and $\sigma_{X}=0.141$, it follows that: 
\begin{align*}
\mathrm{kurt}(X) & =[(-0.3-0.1)^{4}\cdot(0.05)+(0.0-0.1)^{4}\cdot(0.20)+(0.1-0.1)^{4}\cdot(0.5)\\
 & +(0.2-0.1)^{4}\cdot(0.2)+(0.5-0.1)^{4}\cdot(0.05)]/(0.141)^{4}\\
 & =6.5.
\end{align*}
\hfill{}$\blacksquare$
:::

::: {#exm-example230 name="Kurtosis for a normal random variable"}
Suppose $X$ has a general normal distribution with mean $\mu_{X}$
and variance $\sigma_{X}^{2}$. Then it can be shown that: 
$$
\mathrm{kurt}(X)=\int_{-\infty}^{\infty}\frac{(x-\mu_{X})^{4}}{\sigma_{X}^{4}}\cdot\frac{1}{\sqrt{2\pi\sigma_{X}^{2}}}e^{-\frac{1}{2}(\frac{x-\mu_{X}}{\sigma_{X}})^{2}}~dx=3.
$$
Hence, a kurtosis of 3 is a benchmark value for tail thickness of
bell-shaped distributions. If a distribution has a kurtosis greater
than 3, then the distribution has thicker tails than the normal distribution.
If a distribution has kurtosis less than 3, then the distribution
has thinner tails than the normal.
\hfill{}$\blacksquare$
:::

Sometimes the kurtosis of a random variable is described relative
to the kurtosis of a normal random variable. This relative value of
kurtosis is referred to as *excess kurtosis* and is defined as:
$$
\mathrm{ekurt}(X)=\mathrm{kurt}(X)-3
$$ {#eq-excess-kurtosis}
If the excess kurtosis of a random variable is equal to zero then
the random variable has the same kurtosis as a normal random variable.
If excess kurtosis is greater than zero, then kurtosis is larger than
that for a normal; if excess kurtosis is less than zero, then kurtosis
is less than that for a normal.

#### The Student's t Distribution
\index{Studnet's t distribution}
The kurtosis of a random variable gives information on the tail thickness
of its distribution. The normal distribution, with kurtosis equal
to three, gives a benchmark for the tail thickness of symmetric distributions.
A distribution similar to the standard normal distribution but with
fatter tails, and hence larger kurtosis, is the Student's t distribution.
If $X$ has a Student's t distribution with degrees of freedom parameter
$v$, denoted $X\sim t_{v},$ then its pdf has the form: 
$$
f(x)=\frac{\Gamma\left(\frac{v+1}{2}\right)}{\sqrt{v\pi}\Gamma\left(\frac{v}{2}\right)}\left(1+\frac{x^{2}}{v}\right)^{-\left(\frac{v+1}{2}\right)},~-\infty<x<\infty,~v>0.
$$ {#eq-Students-t-pdf}
where $\Gamma(z)=\int_{0}^{\infty}t^{z-1}e^{-t}dt$ denotes the gamma
function. It can be shown that: 
\begin{align*}
E[X] & =0,~v>1\\
\mathrm{var}(X) & =\frac{v}{v-2},~v>2,\\
\mathrm{skew}(X) & =0,~v>3,\\
\mathrm{kurt}(X) & =\frac{6}{v-4}+3,~v>4.
\end{align*}
The parameter $v$ controls the scale and tail thickness of the distribution.
If $v$ is close to four, then the kurtosis is large and the tails
are thick. If $v<4$, then $\mathrm{kurt}(X)=\infty$. As $v\rightarrow\infty$
the Student's t pdf approaches that of a standard normal random variable
and $\mathrm{kurt}(X)=3$. Figure @fig-ProbReviewStudentT
shows plots of the Student's t density for various values of $v$.

```{r}
#| label: ProbReviewStudentT
#| echo: false
#| fig-cap: Student's t density with $v=1,5,10$ and $60$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.vals = seq(-5, 5, length=100)
ub = max( c(dt(0, df=c(1,5,10,60)), dnorm(0)) )
plot(x.vals, dt(x.vals, df=1), type="l", lwd=2, ylim=c(0, ub), ylab="pdf", xlab="x")
points(x.vals, dt(x.vals, df=5), type="l", lwd=2, lty=2, col=2)
points(x.vals, dt(x.vals, df=10), type="l", lwd=2, lty=3, col=3)
points(x.vals, dt(x.vals, df=60), type="l", lwd=2, lty=4, col=4)
legend(x="topleft", legend=paste("v=", c(1,5,10,60), sep=""), lty=1:4, col=1:4, lwd=2)
```

::: {#exm-example231 name="Computing tail probabilities and quantiles from the Student's t distribution"}
The R functions `pt()` and `qt()` can be used to compute
the cdf and quantiles of a Student's t random variable. For $v=1,2,5,10,60,100$
and $\infty$ the 1\% quantiles can be computed using:

```{r}
#| echo: true
v = c(1, 2, 5, 10, 60, 100, Inf)
qt(0.01, df=v)
```

For $v=1,2,5,10,60,100$ and $\infty$ the values of $\Pr(X<-3)$
are:

```{r}
#| echo: true
pt(-3, df=v)
```

\hfill{}$\blacksquare$
:::

### Linear functions of a random variable

Let $X$ be a random variable either discrete or continuous with $E[X]=\mu_{X}$,
$\mathrm{var}(X)=\sigma_{X}^{2}$, and let $a$ and $b$ be known
constants. Define a new random variable $Y$ via the linear function
of $X$: 
$$
Y=g(X)=aX+b.
$$
Then the following results hold: 
\begin{align*}
\mu_{Y} & =E[Y]=a\cdot E[X]+b=a\cdot\mu_{X}+b,\\
\sigma_{Y}^{2} & =\mathrm{var}(Y)=a^{2}\mathrm{var}(X)=a^{2}\sigma_{X}^{2},\\
\sigma_{Y} & =\mathrm{sd}(Y)=a\cdot\mathrm{sd}(X)=a\cdot\sigma_{X}.
\end{align*}
The first result shows that expectation is a linear operation. That
is, 
$$
E[aX+b]=aE[X]+b.
$$
The second result shows that adding a constant to $X$ does not affect
its variance, and that the effect of multiplying $X$ by the constant
$a$ increases the variance of $X$ by the square of $a.$ These results
will be used often enough that it is instructive to go through the derivations.

Consider the first result. Let $X$ be a discrete random variable.
By the definition of $E[g(X)]$, with $g(X)=b+aX$, you can write: 
\begin{align*}
E[Y] & =\sum_{x\in S_{X}}(ax+b)\cdot\Pr(X=x)\\
 & =a\sum_{x\in S_{X}}x\cdot\Pr(X=x)+b\sum_{x\in S_{X}}\Pr(X=x)\\
 & =a\cdot E[X]+b\cdot1\\
 & =a\cdot\mu_{X}+b.
\end{align*}
If $X$ is a continuous random variable then by the linearity of integration:
\begin{align*}
E[Y] & =\int_{-\infty}^{\infty}(ax+b)f(x)~dx=a\int_{-\infty}^{\infty}xf(x)~dx+b\int_{-\infty}^{\infty}f(x)~dx\\
 & =aE[X]+b.
\end{align*}
Next consider the second result. Since $\mu_{Y}=a\mu_{X}+b$:
\begin{align*}
\mathrm{var}(Y) & =E[(Y-\mu_{Y})^{2}]\\
 & =E[(aX+b-(a\mu_{X}+b))^{2}]\\
 & =E[(a(X-\mu_{X})+(b-b))^{2}]\\
 & =E[a^{2}(X-\mu_{X})^{2}]\\
 & =a^{2}E[(X-\mu_{X})^{2}]\textrm{ (by the linearity of }E[\cdot]\textrm{)}\\
 & =a^{2}\mathrm{var}(X)
\end{align*}
Notice that the derivation of the second result works for discrete
and continuous random variables. 

A normal random variable has the special property that a linear function
of it is also a normal random variable. The following proposition
states the result.

::: {#prp-prop-linear-normal name="Linear function of a normal random variable"}
Let $X\sim N(\mu_{X},\sigma_{X}^{2})$ and let
$a$ and $b$ be constants. Let $Y=aX+b$. Then $Y\sim N(a\mu_{X}+b,a^{2}\sigma_{X}^{2})$.
:::

The above property is special to the normal distribution and may or
may not hold for a random variable with a distribution that is not
normal.

::: {#exm-example232 name="Standardizing a random variable"}
Let $X$ be a random variable with $E[X]=\mu_{X}$ and $\mathrm{var}(X)=\sigma_{X}^{2}$.
Define a new random variable $Z$ as: 
$$
Z=\frac{X-\mu_{X}}{\sigma_{X}}=\frac{1}{\sigma_{X}}X-\frac{\mu_{X}}{\sigma_{X}},
$$
which is a linear function $aX+b$ where $a=\frac{1}{\sigma_{X}}$
and $b=-\frac{\mu_{X}}{\sigma_{X}}$. This transformation is called
*standardizing*\index{Standardizing} the random variable $X$ since, 
\begin{align*}
E[Z] & =\frac{1}{\sigma_{X}}E[X]-\frac{\mu_{X}}{\sigma_{X}}=\frac{1}{\sigma_{X}}\mu_{X}-\frac{\mu_{X}}{\sigma_{X}}=0,\\
\mathrm{var}(Z) & =\left(\frac{1}{\sigma_{X}}\right)^{2}\mathrm{var}(X)=\frac{\sigma_{X}^{2}}{\sigma_{X}^{2}}=1.
\end{align*}
Hence, standardization creates a new random variable with mean zero
and variance 1. In addition, if $X$ is normally distributed then
$Z\sim N(0,1)$.
\hfill{}$\blacksquare$
:::

::: {#exm-example233 name="Computing probabilities using standardized random variables"}
Let $X\sim N(2,4)$ and suppose you want to find $\Pr(X>5)$ but you
only know probabilities associated with a standard normal random variable
$Z\sim N(0,1)$. You can solve the problem by standardizing $X$ as follows:
$$
\Pr\left(X>5\right)=\Pr\left(\frac{X-2}{\sqrt{4}}>\frac{5-2}{\sqrt{4}}\right)=\Pr\left(Z>\frac{3}{2}\right)=0.06681.
$$
\hfill{}$\blacksquare$
:::

Standardizing a random variable is often done in the construction
of test statistics. For example, the so-called *t-statistic*
or *t-ratio* used for testing simple hypotheses on coefficients
is constructed by the standardization process. 

A non-standard random variable $X$ with mean $\mu_{X}$ and variance
$\sigma_{X}^{2}$ can be created from a standard random variable via
the linear transformation: 
$$
X=\mu_{X}+\sigma_{X}\cdot Z.
$$ {#eq-normal-rv}
This result is useful for modeling purposes as illustrated in the
next examples.

::: {#exm-example234 name="Quantile of general normal random variable"}
Let $X\sim N(\mu_{X},\sigma_{X}^{2})$. Quantiles of $X$ can be conveniently
computed using: 
$$
q_{\alpha}=\mu_{X}+\sigma_{X}z_{\alpha},
$$ {#eq-normal-quantile}
where $\alpha\in(0,1)$ and $z_{\alpha}$ is the $\alpha\times100\%$
quantile of a standard normal random variable. This formula is derived
as follows. By the definition $z_{\alpha}$ and @eq-normal-rv
$$
\alpha=\Pr(Z\leq z_{\alpha})=\Pr\left(\mu_{X}+\sigma_{X}\cdot Z\leq\mu_{X}+\sigma_{X}\cdot z_{\alpha}\right)=\Pr(X\leq\mu_{X}+\sigma_{X}\cdot z_{\alpha}),
$$
which implies @eq-normal-quantile.
\hfill{}$\blacksquare$
:::

::: {#exm-example235 name="Generalized Student's t distribution"}
The Student's t distribution defined in @eq-Students-t-pdf
has zero mean and variance equal to $v/(v-2)$ where $v$ is the degrees
of freedom parameter. You can define a Student's t random variable
with non-zero mean $\mu$ and variance $\sigma^{2}$ from a Student's
t random variable $X$ as follows:
$$
Y=\mu+\sigma\left(\frac{v-2}{v}\right)^{1/2}X=\mu+\sigma Z_{v},
$$
where $Z_{v}=\left(\frac{v-2}{v}\right)^{1/2}X$. The random variable
$Z_{v}$ is called a *standardized Student's t*\index{Standardized Student's t} 
random variable with $v$ degrees of freedom since $E[Z_{v}]=0$ and $\mathrm{var}(Z_{v})=\left(\frac{v-2}{v}\right)\mathrm{var}(X)=\left(\frac{v-2}{v}\right)\left(\frac{v}{v-2}\right)=1$.
Then
\begin{eqnarray*}
E[Y] & = & \mu+\sigma E[Z]=\mu,\\
\mathrm{var}(Y) & = & \sigma^{2}\mathrm{var}(Z)=\sigma^{2}.
\end{eqnarray*}

I use the notation $Y \sim t_{v}(\mu, \sigma^{2})$ to denote that $Y$ has a 
*generalized Student's t*\index{Generalized Student's t} distribution with mean $\mu$, variance $\sigma^{2}$, and degrees of freedom $v$.
\hfill{}$\blacksquare$
:::

::: {#exm-example236 name="Location scale return model for asset returns"}
Let $r$ denote the monthly continuously compounded return on an asset with
$E[r]=\mu$ and $\mathrm{var}(R)=\sigma^{2}$. If $r$ has a normal distribution or a 
general Student's t distribution, then $r$ can be expressed as: 
$$
r=\mu_{r}+\sigma_{r}\cdot\varepsilon,
$$
where $\varepsilon \sim N(0,1)$ or $\varepsilon \sim t_{v}(0,1)$. The random variable $\varepsilon$ can be interpreted as representing
the random news arriving in a given month that makes the observed
return differ from its expected value $\mu$. The fact that $\varepsilon$
has mean zero means that news, on average, is neutral. The value of
$\sigma_{r}$ represents the typical size of a news shock. The bigger
is $\sigma_{r}$, the larger is the impact of a news shock and vice-versa. The 
distribution of $\varepsilon$ impacts the likelihood of extreme new events (like
stock market crashes). If $\varepsilon \sim N(0,1)$ then extreme news events are not very likely. If $\varepsilon \sim t_{v}(0,1)$ with $v$ small (but bigger than 4) then extreme news events may happen from time-to-time. 
\hfill{}$\blacksquare$
:::

### Value-at-Risk: An introduction
\index{Value-at-Risk}
As an example of working with linear functions of a normal random
variable, and to illustrate the concept of *Value-at-Risk* (VaR),
consider an investment of \$10,000 in Microsoft stock over the next
month. Let $R$ denote the monthly simple return on Microsoft stock
and assume that $R\sim N(0.05,(0.10)^{2})$. That is, $E[R]=\mu_{R}=0.05$
and $\mathrm{var}(R)=\sigma_{R}^{2}=(0.10)^{2}$. Let $W_{0}$ denote
the investment value at the beginning of the month and $W_{1}$ denote
the investment value at the end of the month. In this example, $W_{0}=\$10,000$.
Consider the following questions:

  (i) What is the probability distribution of end-of-month wealth, $W_{1}?$ 
  (ii) What is the probability that end-of-month wealth is less than \$9,000, and what must the return on Microsoft be for this to happen? 
  (iii) What is the loss in dollars that would occur if the return on Microsoft stock is equal to its 5\% quantile, $q_{.05}$? That is, what is the monthly 5\% VaR on the \$10,000 investment in Microsoft? 

To answer (i), note that end-of-month wealth, $W_{1}$, is related
to initial wealth $W_{0}$ and the return on Microsoft stock $R$
via the linear function: 
$$
W_{1}=W_{0}(1+R)=W_{0}+W_{0}R=\$10,000+\$10,000\cdot R.
$$
Using the properties of linear functions of a random variable you get:
$$
E[W_{1}]=W_{0}+W_{0}E[R]=\$10,000+\$10,000(0.05)=\$10,500,
$$
and, 
\begin{align*}
\mathrm{var}(W_{1}) & =(W_{0})^{2}\mathrm{var}(R)=(\$10,000)^{2}(0.10)^{2},\\
\mathrm{sd}(W_{1}) & =(\$10,000)(0.10)=\$1,000.
\end{align*}
Further, since $R$ is assumed to be normally distributed it follows
that $W_{1}$ is normally distributed: 
$$
W_{1}\sim N(\$10,500,(\$1,000)^{2}).
$$
To answer (ii), you can use the above normal distribution for $W_{1}$
to get: 
$$
\Pr(W_{1}<\$9,000)=0.067.
$$
To find the return that produces end-of-month wealth of \$9,000, or
a loss of \$10,000-\$9,000=\$1,000, you solve 
$$
R=\frac{\$9,000-\$10,000}{\$10,000}=-0.10.
$$
If the monthly return on Microsoft is -10\% or less, then end-of-month
wealth will be \$9,000 or less. Notice that $R=-0.10$ is the 6.7\%
quantile of the distribution of $R$: 
$$
\Pr(R<-0.10)=0.067.
$$
Question (iii) can be answered in two equivalent ways. First, use
$R\sim N(0.05,(0.10)^{2})$ and solve for the 5\% quantile of Microsoft
Stock using @eq-normal-quantile^[Using R, $z_{.05}=\mathrm{qnorm}(0.05)=-1.645$.]:
\begin{align*}
\Pr(R & <q_{.05}^{R})=0.05\\
 & \Rightarrow q_{.05}^{R}=\mu_{R}+\sigma_{R}\cdot z_{.05}=0.05+0.10\cdot(-1.645)=-0.114.
\end{align*}
That is, with 5\% probability the return on Microsoft stock is -11.4\%
or less. Now, if the return on Microsoft stock is -11.4\% the loss
in investment value that occurs with 5\% probability is
$$
\mathrm{VaR}_{0.05} = W_{0}\cdot q_{.05}^{R} = \$10,000\cdot(0.114)= -\$1,144.
$$
Since losses are typically reported as positive numbers, $\mathrm{VaR}_{0.05} = \$1,144$
is the 5\% VaR over the next month on the \$10,000 investment in Microsoft
stock. 

For the second method, use $W_{1}$ ~ $N(\$10,500,(\$1,000)^{2})$
and solve for the $5\%$ quantile of end-of-month wealth directly:
$$
\Pr(W_{1}<q_{.05}^{W_{1}})=0.05\Rightarrow q_{.05}^{W_{1}}=\$8,856.
$$
This corresponds to a loss of investment value of \$10,000-\$8,856=\$1,144.
Hence, if $W_{0}$ represents the initial wealth and $q_{.05}^{W_{1}}$
is the 5\% quantile of the distribution of $W_{1}$ then the 5\% VaR
is: 
$$
\textrm{ VaR}_{.05}=W_{0}-q_{.05}^{W_{1}}=\$10,000-\$8,856=\$1,144.
$$

The following is the usual definition of VaR.

::: {#def-VaR name="Value-at-Risk"}
If $W_{0}$ represents the initial wealth in dollars and
$q_{\alpha}^{R}$ is the $\alpha\times100\%$ quantile of distribution
of the simple return $R$ for small values of $\alpha$, then the $\alpha\times100$\% VaR is defined
as: 
$$
\mathrm{VaR}_{\alpha}=-W_{0}\cdot q_{\alpha}^{R}.
$$ {#eq-VaR-simple-return}
:::

In words, VaR$_{\alpha}$ represents the dollar loss that could occur
with probability $\alpha$. By convention, it is reported as a positive
number (hence the multiplication by minus one as $q_{\alpha}^{R} < 0$ for small values of $\alpha$).

#### Value-at-Risk calculations for continuously compounded returns

The above calculations illustrate how to calculate VaR using
the normal distribution for simple returns. However, as argued in
Example 31, the normal distribution may not be appropriate for characterizing
the distribution of simple returns and may be more appropriate for
characterizing the distribution of continuously compounded returns.
Let $R$ denote the simple monthly return, let $r=\ln(1+R)$ denote
the continuously compounded return and assume that $r\sim N(\mu_{r},\sigma_{r}^{2})$.
The $\alpha\times100\%$ monthly VaR on an investment of \$$W_{0}$
is computed using the following steps: 

1.  Compute the $\alpha\cdot100\%$ quantile, $q_{\alpha}^{r}$, from the normal distribution for the continuously compounded return $r$:
$$
q_{\alpha}^{r}=\mu_{r}+\sigma_{r}z_{\alpha},
$$
where $z_{\alpha}$ is the $\alpha\cdot100\%$ quantile of the standard normal distribution. 
2.  Convert the continuously compounded return quantile, $q_{\alpha}^{r}$, to a simple return quantile using the transformation: 
$$
q_{\alpha}^{R}=e^{q_{\alpha}^{r}}-1
$$
3.  Compute VaR using the simple return quantile @eq-VaR-simple-return. 

Step 2 works because quantiles are preserved under monotonically increasing 
transformations of random variables. 

::: {#exm-example237 name="Computing VaR from simple and continuously compounded returns using R"}
Let $R$ denote the simple monthly return on Microsoft stock and assume
that $R\sim N(0.05,(0.10)^{2})$. Consider an initial investment of
$W_{0}=\$10,000$. To compute the 1\% and 5\% VaR values over the
next month use:

```{r}
#| echo: true

mu.R = 0.05
sd.R = 0.10
w0 = 10000
q.01.R = mu.R + sd.R*qnorm(0.01)
q.05.R = mu.R + sd.R*qnorm(0.05)
VaR.01 = -q.01.R*w0
VaR.05 = -q.05.R*w0
VaR.01
VaR.05
```

Hence with 1\% and 5\% probability the loss over the next month is
at least \$1,826 and \$1,145, respectively. 


Let $r$ denote the continuously compounded return on Microsoft stock
and assume that $r\sim N(0.05,(0.10)^{2})$. To compute the 1\% and
5\% VaR values over the next month use:

```{r}
#| echo: true

mu.r = 0.05
sd.r = 0.10
q.01.R = exp(mu.r + sd.r*qnorm(0.01)) - 1
q.05.R = exp(mu.r + sd.r*qnorm(0.05)) - 1
VaR.01 = -q.01.R*w0
VaR.05 = -q.05.R*w0
VaR.01
VaR.05
```

Notice that when $1+R=e^{r}$ has a log-normal distribution, the 1\%
and 5\% VaR values (losses) are slightly smaller than when $R$ is
normally distributed. This is due to the positive skewness of the
log-normal distribution.
\hfill{}$\blacksquare$
:::

## Bivariate Distributions {#sec-Bivariate-Distributions}
\index{Bivariate distribution}
So far I have only considered probability distributions for a single
random variable. In many situations, you want to be able to characterize
the probabilistic behavior of two or more random variables simultaneously.
For example, you might want to know about the probabilistic behavior of the 
returns on a two asset portfolio. In this section, I discuss bivariate distributions
and important concepts related to the analysis of two random variables.

### Discrete random variables
\index{Joint probability distribution}
Let $X$ and $Y$ be discrete random variables with sample spaces
$S_{X}$ and $S_{Y}$, respectively. The likelihood that $X$ and
$Y$ takes values in the joint sample space $S_{XY}=S_{X}\times S_{Y}$ (Cartesian 
product of the individual sample spaces) is determined by the *joint probability distribution* $p(x,y)=\Pr(X=x,Y=y).$ The function $p(x,y)$ satisfies: 

-  (i) $p(x,y)>0$ for $x,y\in S_{XY}$; 
-  (ii) $p(x,y)=0$ for $x,y\notin S_{XY}$; 
-  (iii) $\sum_{x,y\in S_{XY}}p(x,y)=\sum_{x\in S_{X}}\sum_{y\in S_{Y}}p(x,y)=1$. 

::: {#exm-example238 name="Bivariate discrete distribution for stock returns"}
Let $X$ denote the monthly return (in percent) on Microsoft stock
and let $Y$ denote the monthly return on Starbucks stock. For simplicity
suppose that the sample spaces for $X$ and $Y$ are $S_{X}=\{0,1,2,3\}$
and $S_{Y}=\{0,1\}$ so that the random variables $X$ and $Y$ are
discrete. The joint sample space is the two dimensional grid $S_{XY}=\{(0,0)$,
$(0,1)$, $(1,0)$, $(1,1)$, $(2,0)$, $(2,1)$, $(3,0)$, $(3,1)\}$.
Table @tbl-Table-bivariateDistn illustrates the joint distribution
for $X$ and $Y$. From the table, $p(0,0)=\Pr(X=0,Y=0)=1/8$. Notice
that the sum of all the entries in the table sum to unity. 
\hfill{}$\blacksquare$
:::

+------+--------------+---------+---------+--------------+
|      |              |**Y**    |         |              |
+======+==============+=========+=========+==============+
|      |**\%**        |**0**    |**1**    |**$\Pr(X)$**  |
+------+--------------+---------+---------+--------------+
|      |0             |1/8      |0        |1/8           |
+------+--------------+---------+---------+--------------+
|**X** |1             |2/8      |1/8      |3/8           |
+------+--------------+---------+---------+--------------+
|      |2             |1/8      |2/8      |3/8           |
+------+--------------+---------+---------+--------------+
|      |3             |0        |1/8      |1/8           |
+------+--------------+---------+---------+--------------+
|      |**$\Pr(Y)$**  |**4/8**  |**4/8**  |**1**         |
+------+--------------+---------+---------+--------------+

: Discrete bivariate distribution for Microsoft and Starbucks stock prices. {#tbl-bivariateDistn}


#### Marginal Distributions
\index{Marginal distributions}
The joint probability distribution tells us the probability of $X$ and
$Y$ occurring together. What if you only want to know about the probability
of $X$ occurring, or the probability of $Y$ occurring?

::: {#exm-example239 name="Find $\\Pr(X=0)$ and $\\Pr(Y=1)$ from joint distribution"}
Consider the joint distribution in Table @tbl-bivariateDistn.
What is $\Pr(X=0)$ regardless of the value of $Y$? Now $X$ can
occur if $Y=0$ or if $Y=1$ and since these two events are mutually
exclusive you get: 
$$
\Pr(X=0)=\Pr(X=0,Y=0)+\Pr(X=0,Y=1)=0+1/8=1/8.
$$
Notice that this probability is equal to the horizontal (row) sum
of the probabilities in the table at $X=0$. You can find $\Pr(Y=1)$
in a similar fashion: 
\begin{align*}
\Pr(Y=1)&=\Pr(X=0,Y=1)+\Pr(X=1,Y=1)+\Pr(X=2,Y=1)+\Pr(X=3,Y=1)\\
&=0+1/8+2/8+1/8=4/8.
\end{align*}
This probability is the vertical (column) sum of the probabilities
in the table at $Y=1$.
\hfill{}$\blacksquare$
:::

::: {#def-marginal-distribution name="Marginal distribution"}
The probability $\Pr(X=x)$ is called the *marginal probability*
of $X$ and is given by: 
$$
\Pr(X=x)=\sum_{y\in S_{Y}}\Pr(X=x,Y=y).
$$ {#eq-marginal-X-discrete}
Similarly, the marginal probability of $Y=y$ is given by: 
$$
\Pr(Y=y)=\sum_{x\in S_{X}}\Pr(X=x,Y=y).
$$ {#eq-marginal-Y-discrete}
:::

::: {#exm-example240 name="Marginal probabilities of discrete bivariate distribution"}
The marginal probabilities of $X=x$ are given in the last column
of Table @tbl-bivariateDistn, and the marginal probabilities
of $Y=y$ are given in the last row of Table @tbl-bivariateDistn.
Notice that these probabilities sum to 1. For future reference,
note that $E[X]=3/2,\mathrm{var}(X)=3/4,E[Y]=1/2$, and $\mathrm{var}(Y)=1/4$.
\hfill{}$\blacksquare$
:::

#### Conditional Distributions
\index{Conditional distributions}

For random variables in Table @tbl-bivariateDistn, suppose
you know that the random variable $Y$ takes on the value $Y=0$. How
does this knowledge affect the likelihood that $X$ takes on the values
0, 1, 2 or 3? For example, what is the probability that $X=0$ *given that* 
you know $Y=0$? To find this probability, you can use Bayes' law \index{Bayes' law}
and compute the *conditional probability*:\index{Conditional probability}
$$
\Pr(X=0|Y=0)=\frac{\Pr(X=0,Y=0)}{\Pr(Y=0)}=\frac{1/8}{4/8}=1/4.
$$
The notation $\Pr(X=0|Y=0)$ is read as "the probability that $X=0$
given that $Y=0$" . Notice that $\Pr(X=0|Y=0)=1/4>\Pr(X=0)=1/8$.
Hence, knowledge that $Y=0$ increases the likelihood that $X=0$.
Clearly, $X$ *depends* on $Y$. 


Now suppose that you know that $X=0$. How does this knowledge affect
the probability that $Y=0?$ To find out compute: 
$$
\Pr(Y=0|X=0)=\frac{\Pr(X=0,Y=0)}{\Pr(X=0)}=\frac{1/8}{1/8}=1.
$$
Notice that $\Pr(Y=0|X=0)=1>\Pr(Y=0)=1/2$. That is, knowledge that
$X=0$ makes it certain that $Y=0$. 

::: {#dfn-conditional-distribution name="Conditional distribution"}
The conditional probability that $X=x$ given that $Y=y$
(provided $\Pr(Y=y)\neq0)$ is, 
$$
p(x|y) = \Pr(X=x|Y=y)=\frac{\Pr(X=x,Y=y)}{\Pr(Y=y)} = \frac{p(x,y)}{p(y)},
$$ {#eq-conditional-X-discrete}
and the conditional probability that $Y=y$ given that $X=x$ (provided
$\Pr(X=x)\neq0$) is, 
$$
p(y|x) = \Pr(Y=y|X=x)=\frac{\Pr(X=x,Y=y)}{\Pr(X=x)} = \frac{p(x,y)}{p(x)}.
$$ {#eq-conditional-y-discrete}
:::

::: {#exm-example241 name="Conditional distributions"}
For the bivariate distribution in Table @tbl-bivariateDistn,
the conditional probabilities along with marginal probabilities are
summarized in Tables @tbl-ConditionalDistnX and @tbl-ConditionalDistnY.
Notice that the marginal distribution of $X$ is centered at $x=3/2$
whereas the conditional distribution of $X|Y=0$ is centered at $x=1$
and the conditional distribution of $X|Y=1$ is centered at $x=2$.
\hfill{}$\blacksquare$
:::

+---+------------+--------------+--------------+
|x  |$\Pr(X=x)$  |$\Pr(X|Y=0)$  |$\Pr(X|Y=1)$  |
+===+============+==============+==============+
|0  |1/8         |2/8           |0             |
+---+------------+--------------+--------------+
|1  |3/8         |4/8           |2/8           |
+---+------------+--------------+--------------+
|2  |3/8         |4/8           |4/8           |
+---+------------+--------------+--------------+
|3  |1/8         |0             |2/8           |
+---+------------+--------------+--------------+

: Conditional probability distribution of X from bivariate discrete distribution. {#tbl-ConditionalDistnX} 


+---+------------+--------------+--------------+--------------+--------------+
|y  |$\Pr(Y=y)$  |$\Pr(Y|X=0)$  |$\Pr(Y|X=1)$  |$\Pr(Y|X=2)$  |$\Pr(Y|X=3)$  |
+---+------------+--------------+--------------+--------------+--------------+
|0  |1/2         |1             |2/3           |1/3           |0             |
+---+------------+--------------+--------------+--------------+--------------+
|1  |1/2         |0             |1/3           |2/3           |1             |
+---+------------+--------------+--------------+--------------+--------------+

: Conditional distribution of Y from bivariate discrete distribution. {#tbl-ConditionalDistnY}


#### Conditional expectation and conditional variance

Just as I defined shape characteristics of the marginal distributions
of $X$ and $Y$, I can also define shape characteristics of the
conditional distributions of $X|Y=y$ and $Y|X=x$. The most important
shape characteristics are the *conditional expectation*\index{Conditional expectation} (*conditional mean*) and the *conditional variance*\index{Conditional variance}. 
The conditional mean of $X|Y=y$ is denoted by $\mu_{X|Y=y}=E[X|Y=y]$, and the conditional mean of $Y|X=x$ is denoted by $\mu_{Y|X=x}=E[Y|X=x]$. 

::: {#def-conditional-expectation name="Conditional expectation"}
For discrete random variables $X$ and $Y$, the conditional expectations are defined as:
\begin{align}
\mu_{X|Y=y} & =E[X|Y=y]=\sum_{x\in S_{X}}x\cdot\Pr(X=x|Y=y),\\
\mu_{Y|X=x} & =E[Y|X=x]=\sum_{y\in S_{Y}}y\cdot\Pr(Y=y|X=x).
\end{align}
:::

Similarly, the conditional variance of $X|Y=y$ is denoted by $\sigma_{X|Y=y}^{2}=\mathrm{var}(X|Y=y)$
and the conditional variance of $Y|X=x$ is denoted by $\sigma_{Y|X=x}^{2}=\mathrm{var}(Y|X=x)$.

::: {#def-conditional-variance name="Conditional variance"}
For discrete random variables $X$ and $Y$, the conditional variances are defined as:
\begin{align}
\sigma_{X|Y=y}^{2} & =\mathrm{var}(X|Y=y)=\sum_{x\in S_{X}}(x-\mu_{X|Y=y})^{2}\cdot\Pr(X=x|Y=y),\\
\sigma_{Y|X=x}^{2} & =\mathrm{var}(Y|X=x)=\sum_{y\in S_{Y}}(y-\mu_{Y|X=x})^{2}\cdot\Pr(Y=y|X=x).
\end{align}
The conditional volatilities are defined as:
\begin{align}
\sigma_{X|Y=y} & = \sqrt{\sigma_{X|Y=y}^{2}}, \\
\sigma_{Y|X=x} & = \sqrt{\sigma_{Y|X=x}^{2}}.                                                 
\end{align}
:::

Then conditional mean is the center of mass of the conditional distribution, 
and the conditional variance and volatility measure the spread of the conditional distribution about the conditional mean. 

::: {#exm-example242 name="Compute conditional expectation and conditional variance"}
For the random variables in Table @tbl-bivariateDistn, you can derive the following conditional moments for $X$: 
\begin{align*}
E[X|Y & =0]=0\cdot1/4+1\cdot1/2+2\cdot1/4+3\cdot0=1,\\
E[X|Y & =1]=0\cdot0+1\cdot1/4+2\cdot1/2+3\cdot1/4=2,\\
\mathrm{var}(X|Y & =0)=(0-1)^{2}\cdot1/4+(1-1)^{2}\cdot1/2+(2-1)^{2}\cdot1/2+(3-1)^{2}\cdot0=1/2,\\
\mathrm{var}(X|Y & =1)=(0-2)^{2}\cdot0+(1-2)^{2}\cdot1/4+(2-2)^{2}\cdot1/2+(3-2)^{2}\cdot1/4=1/2.
\end{align*}
Compare these values to $E[X]=3/2$ and $\mathrm{var}(X)=3/4$. Notice that $\mathrm{var}(X|Y) = 1/2 < \mathrm{var}(X)=3/4$ so that conditioning on information reduces
variability. Also, notice that as $y$ increases, $E[X|Y=y]$ increases. 

For $Y$, similar calculations gives: 
\begin{align*}
E[Y|X & =0]=0,E[Y|X=1]=1/3,E[Y|X=2]=2/3,E[Y|X=3]=1,\\
\mathrm{var}(Y|X & =0)=0,\mathrm{var}(Y|X=1)=0.2222,\mathrm{var}(Y|X=2)=0.2222,\mathrm{var}(Y|X=3)=0.
\end{align*}
Compare these values to $E[Y]=1/2$ and $\mathrm{var}(Y)=1/4$. Notice
that as $x$ increases $E[Y|X=x]$ increases.
\hfill{}$\blacksquare$
:::

#### Conditional expectation and the regression function

Consider the problem of predicting the value $Y$ given that you know
$X=x$. A natural predictor to use is the conditional expectation
$E[Y|X=x]$. In this prediction context, the conditional expectation
$E[Y|X=x]$ is called the *regression function*\index{Regression function}. The graph with
$E[Y|X=x]$ on the vertical axis and $x$ on the horizontal axis gives
the *regression line*\index{Regression line}. The relationship between $Y$ and the
regression function may be expressed using the trivial identity: 
$$
\begin{aligned}
Y & =E[Y|X=x]+Y-E[Y|X=x]\nonumber \\
 & =E[Y|X=x]+\varepsilon,
\end{aligned}
$$ {#eq-regression-function}
where $\varepsilon=Y-E[Y|X]$ is called the *regression error*.

::: {#exm-example243 name="Regression line for bivariate discrete distribution"}
For the random variables in Table @tbl-bivariateDistn,
the regression line is plotted in Figure @fig-figProbReviewDiscreteRegression.
Notice that there is a linear relationship between $E[Y|X=x]$ and
$x$. When such a linear relationship exists you call the regression
function a *linear regression*. Linearity of the regression function,
however, is not guaranteed. It may be the case that there is a non-linear
(e.g., quadratic) relationship between $E[Y|X=x]$ and $x$. In this
case, you call the regression function a *non-linear regression*\index{Non-linear regression}.
\hfill{}$\blacksquare$
:::

```{r}
#| label: figProbReviewDiscreteRegression
#| echo: false
#| fig-cap: Regression function $E[Y|X=x]$ from discrete bivariate distribution.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
# discrete distributions
x.probs = matrix(c(1/8,3/8,3/8,1/8,
                   2/8,4/8,2/8,0,
                   0,2/8,4/8,2/8), 4, 3)
y.probs = matrix(c(1/2,1/2,
                   1, 0,
                   2/3,1/3,
                   1/3,2/3,
                   0,1), 2, 5)       
x.vals = 0:3
y.vals = 0:1

mu.x = sum(x.probs[,1] *x.vals)
mu.x.y0 = sum(x.probs[,2] *x.vals)
mu.x.y1 = sum(x.probs[,3] *x.vals)

var.x = sum(x.probs[,1] *x.vals^2) - mu.x^2
var.x.y0 = sum(x.probs[,2] *x.vals^2) - mu.x.y0^2
var.x.y1 = sum(x.probs[,3] *x.vals^2) - mu.x.y1^2

mu.y = sum(y.probs[,1] *y.vals)
mu.y.x0 = sum(y.probs[,2] *y.vals)
mu.y.x1 = sum(y.probs[,3] *y.vals)
mu.y.x2 = sum(y.probs[,4] *y.vals)
mu.y.x3 = sum(y.probs[,5] *y.vals)

var.y = sum(y.probs[,1] *y.vals^2) - mu.y^2
var.y.x0 = sum(y.probs[,2] *y.vals^2) - mu.y.x0^2
var.y.x1 = sum(y.probs[,3] *y.vals^2) - mu.y.x1^2
var.y.x2 = sum(y.probs[,4] *y.vals^2) - mu.y.x2^2
var.y.x3 = sum(y.probs[,5] *y.vals^2) - mu.y.x3^2

# plot regression function of y on x

plot(x.vals, c(mu.y.x0, mu.y.x1, mu.y.x2, mu.y.x3), type="b", lwd=2,
     xlab = "x", ylab="E[Y|X=x]", col="blue")
```


#### Law of Total Expectations
\index{Law of total expectations}

Before $Y$ is observed, $E[X|Y]$ is a random variable because its value depends on
the value of $Y$ which is a random variable. Similarly, $E[Y|X]$ is also a random
variable. Since $E[X|Y]$ and $E[Y|X]$ are random variables you can consider their mean values. To illustrate, for the random variables in Table @tbl-bivariateDistn notice that: 
\begin{align*}
E[X] & = \sum_{y \in S_{Y}} E[X|Y=y]\cdot \Pr(Y=y)\\ 
& =E[X|Y=0]\cdot\Pr(Y=0)+E[X|Y=1]\cdot\Pr(Y=1)\\
 & =1\cdot1/2+2\cdot1/2=3/2,
\end{align*}
and, 
\begin{align*}
E[Y] & = \sum_{x \in S_{X}} E[Y|X=x]\cdot \Pr(X=x)\\
& =E[Y|X=0]\cdot\Pr(X=0)+E[Y|X=1]\cdot\Pr(X=1)\\
+E[Y|X & =2]\cdot\Pr(X=2)+E[Y|X=3]\cdot\Pr(X=3)=1/2.
\end{align*}
This result is known as the *law of iterated expectations* or the *law of total expectations*. 

::: {#prp-law-of-iterated-expectations names="Law of iterated expectations"}
For two random variables $X$ and $Y$ (discrete or continuous) it can be shown that: 
$$
\begin{aligned}
E[X] & =E[E[X|Y]],\\
E[Y] & =E[E[Y|X]],\nonumber 
\end{aligned}
$$ {#eq-law-of-iterated-expectations}
where the expectation on the first line is taken with respect to the random 
variable $Y$, and the the expectation on the second line is taken with respect 
to the random variable $X$.
:::

In words, the law of iterated expectations says that the mean of the conditional mean
is the unconditional mean.

### Bivariate distributions for continuous random variables

Let $X$ and $Y$ be continuous random variables defined over the
real line. You can characterize the joint probability distribution of $X$
and $Y$ using the joint probability function (pdf) $f(x,y)$ such
that $f(x,y)\geq0$ and, 
$$
\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(x,y)~dx~dy=1.
$$
The three-dimensional plot of the joint probability distribution gives
a probability surface whose total volume is unity. To compute joint
probabilities of $x_{1}\leq X\leq x_{2}$ and $y_{1}\leq Y\leq y_{2},$
you need to find the volume under the probability surface over the
grid where the intervals $[x_{1},x_{2}]$ and $[y_{1},y_{2}]$ overlap.
Finding this volume requries solving the double integral: 
$$
\Pr(x_{1}\leq X\leq x_{2},y_{1}\leq Y\leq y_{2})=\int_{x_{1}}^{x_{2}}\int_{y_{1}}^{y_{2}}f(x,y)~dx~dy.
$$

::: {#exm-example243 name="Standard bivariate normal distribution"}
A *standard bivariate normal*\index{Standard bivariate normal distribution} pdf for $X$ and $Y$ has the form: 
$$
f(x,y)=\frac{1}{2\pi}e^{-\frac{1}{2}(x^{2}+y^{2})},~-\infty\leq x,~y\leq\infty
$$ {#eq-standard-bivariate-normal}
and has the shape of a symmetric bell (think Liberty Bell) centered
at $x=0$ and $y=0$. To find $\Pr(-1<X<1,-1<Y<1)$ you must solve:
$$
\int_{-1}^{1}\int_{-1}^{1}\frac{1}{2\pi}e^{-\frac{1}{2}(x^{2}+y^{2})}~dx~dy,
$$
which, unfortunately, does not have an analytical solution. Numerical
approximation methods are required to evaluate the above integral.
The function `pmvnorm()` in the R package **mvtnorm** can be used
to evaluate areas under the bivariate standard normal surface. To
compute $\Pr(-1<X<1,-1<Y<1)$ use:

```{r}
#| echo: true
pmvnorm(lower=c(-1, -1), upper=c(1, 1))
```

Here, $\Pr(-1<X<1,-1<Y<1)=0.4661$. The attribute `error` gives
the estimated absolute error of the approximation, and the attribute
`message` tells the status of the algorithm used for the approximation.
See the online help for `pmvnorm` for more details.
\hfill{}$\blacksquare$
:::

#### Marginal and conditional distributions
\index{Marginal distributions}
The marginal pdf of $X$ is found by integrating $y$ out of the joint
pdf $f(x,y)$ and the marginal pdf of $Y$ is found by integrating
$x$ out of the joint pdf $f(x,y)$: 
\begin{align}
f(x) & =\int_{-\infty}^{\infty}f(x,y)~dy,\\
f(y) & =\int_{-\infty}^{\infty}f(x,y)~dx.
\end{align}
The conditional pdf of $X$ given that $Y=y$, denoted $f(x|y)$,
is computed as 
$$
f(x|y)=\frac{f(x,y)}{f(y)},
$$ {#eq-conditional-x-continuous}
and the conditional pdf of $Y$ given that $X=x$ is computed as,
$$
f(y|x)=\frac{f(x,y)}{f(x)}.
$$ {#eq-conditional-y-continuous}
The conditional means are computed as 
\begin{align}
\mu_{X|Y=y} & =E[X|Y=y]=\int x\cdot p(x|y)~dx,\\
\mu_{Y|X=x} & =E[Y|X=x]=\int y\cdot p(y|x)~dy,
\end{align}
and the conditional variances are computed as, 
\begin{align}
\sigma_{X|Y=y}^{2} & =\mathrm{var}(X|Y=y)=\int(x-\mu_{X|Y=y})^{2}p(x|y)~dx,\\
\sigma_{Y|X=x}^{2} & =\mathrm{var}(Y|X=x)=\int(y-\mu_{Y|X=x})^{2}p(y|x)~dy.
\end{align}

::: {#exm-example244 name="Conditional and marginal distributions from bivariate standard normal"}
Suppose $X$ and $Y$ are distributed bivariate standard normal. To
find the marginal distribution of $X$ you use @eq-marginal-x-continuous
and solve: 
$$
f(x)=\int_{-\infty}^{\infty}\frac{1}{2\pi}e^{-\frac{1}{2}(x^{2}+y^{2})}~dy=\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^{2}}\int_{-\infty}^{\infty}\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}y^{2}}~dy=\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^{2}}.
$$
Hence, the marginal distribution of $X$ is standard normal. Similiar
calculations show that the marginal distribution of $Y$ is also standard
normal. To find the conditional distribution of $X|Y=y$ use @eq-conditional-x-continuous
and solve: 
\begin{align*}
f(x|y) & =\frac{f(x,y)}{f(x)}=\frac{\frac{1}{2\pi}e^{-\frac{1}{2}(x^{2}+y^{2})}}{\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}y^{2}}}\\
 & =\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}(x^{2}+y^{2})+\frac{1}{2}y^{2}}\\
 & =\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^{2}}\\
 & =f(x)
\end{align*}
So, for the standard bivariate normal distribution $f(x|y)=f(x)$
which does not depend on $y$. Similar calculations show that $f(y|x)=f(y)$.
\hfill{}$\blacksquare$
:::

### Independence
\index{Independence}
Let $X$ and $Y$ be two discrete random variables. Intuitively, $X$
is independent of $Y$ if knowledge about $Y$ does not influence
the likelihood that $X=x$ for all possible values of $x\in S_{X}$
and $y\in S_{Y}$. Similarly, $Y$ is independent of $X$ if knowledge
about $X$ does not influence the likelihood that $Y=y$ for all values
of $y\in S_{Y}$. I represent this intuition formally for discrete
random variables as follows.

::: {#def-independence name="Independence of discrete random variables"}
Let $X$ and $Y$ be discrete random variables
with sample spaces $S_{X}$ and $S_{Y}$, respectively. $X$ and $Y$
are independent random variables if and only if (iff): 
\begin{align*}
\Pr(X & =x|Y=y)=\Pr(X=x),\textrm{ for all }x\in S_{X},y\in S_{Y}\\
\Pr(Y & =y|X=x)=\Pr(Y=y),\textrm{ for all }x\in S_{X},y\in S_{Y}
\end{align*}
:::

::: {#exm-example244 name="Check independence of bivariate discrete random variables"}
For the data in Table @tbl-bivariateDistn, you know that
$\Pr(X=0|Y=0)=1/4\neq\Pr(X=0)=1/8$ so $X$ and $Y$ are not independent.
\hfill{}$\blacksquare$
:::

::: {#prp-discrete-independence name="Independence of discrete random variables"}
Let $X$ and $Y$ be discrete random variables
with sample spaces $S_{X}$ and $S_{Y}$, respectively. $X$ and $Y$
are independent iff: 
$$
\Pr(X=x,Y=y)=\Pr(X=x)\cdot\Pr(Y=y),\textrm{ for all }x\in S_{X},y\in S_{Y}.
$$
:::

Intuition for the above result follows from: 
\begin{align*}
\Pr(X & =x|Y=y)&=\frac{\Pr(X=x,Y=y)}{\Pr(Y=y)}=\frac{\Pr(X=x)\cdot\Pr(Y=y)}{\Pr(Y=y)}\\
&=\Pr(X=x),\\
\Pr(Y & =y|X=x)&=\frac{\Pr(X=x,Y=y)}{\Pr(X=x)}=\frac{\Pr(X=x)\cdot\Pr(Y=y)}{\Pr(X=x)}\\
& =\Pr(Y=y),
\end{align*}
which shows that $X$ and $Y$ are independent. 


For continuous random variables, you have the following definition
of independence.

::: {#def-independence-continuous name="Independence of continuous random variables"}
Let $X$ and $Y$ be continuous random variables with marginal pdfs $f(x)$ and $f(y)$,
and conditional pdfs $f(x|y)$ and $f(y|x)$. Then $X$ and $Y$ are independent iff: 
\begin{align*}
f(x|y) & =f(x),\textrm{ for }-\infty<x,y<\infty,\\
f(y|x) & =f(y),\textrm{ for }-\infty<x,y<\infty.
\end{align*}
:::

As with discrete random variables, you have the following result for
continuous random variables.

::: {#prp-independence-continuous name="Independence of continuous random variables"}
Let $X$ and $Y$ be continuous random variables with marginal pdfs $f(x)$ and $f(y)$,
and joint pdf $f(x,y)$. Then $X$ and $Y$ are independent iff: 
$$
f(x,y)=f(x)f(y)
$$
:::

This result is extremely useful in practice because it gives us an easy way to compute the joint pdf for two independent random variables: you simply compute the product of the marginal distributions.

::: {#exm-example245 name="Constructing the bivariate standard normal distribution"}
Let $X\sim N(0,1)$, $Y\sim N(0,1)$ and let $X$ and $Y$ be independent.
Then, 
$$
f(x,y)=f(x)f(y)=\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^{2}}\frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}y^{2}}=\frac{1}{2\pi}e^{-\frac{1}{2}(x^{2}+y^{2})}.
$$
This result is a special case of the general bivariate normal distribution to
be discussed in the next sub-section. 
\hfill{}$\blacksquare$
:::

A useful property of the independence between two random variables
is the following Proposition. 

::: {#prp-independence-property name="Property of independent random variables"}
If $X$ and $Y$ are independent then $g(X)$ and
$h(Y)$ are independent for any functions $g(\cdot)$ and $h(\cdot)$.
:::

For example, if $X$ and $Y$ are independent then $X^{2}$ and $Y^{2}$
are also independent.

### Covariance and correlation

Let $X$ and $Y$ be two discrete random variables. Figure @fig-ProbReviewCovariancePlots
displays several bivariate probability scatterplots (where equal probabilities
are given on the dots). In panel (a) you see no linear relationship
between $X$ and $Y$. In panel (b) you see a perfect positive linear
relationship between $X$ and $Y$ and in panel (c) you see a perfect
negative linear relationship. In panel (d) you see a positive, but
not perfect, linear relationship; in panel (e) you see a negative,
but not perfect, linear relationship. Finally, in panel (f) you see
no systematic linear relationship but you see a strong nonlinear (parabolic)
relationship. The *covariance* \index{Covariance} between $X$ and $Y$ measures
the *direction* of the linear relationship between the two random
variables. The *correlation* \index{Correlation} between $X$ and $Y$ measures the
*direction* and the *strength* of the linear relationship
between the two random variables.


```{r}
#| label: fig-ProbReviewCovariancePlots
#| echo: false
#| fig-cap: Probability scatterplots illustrating dependence between $X$ and $Y$.
#| out-width: 80%
#| fig-asp: 1.25
#| fig-align: center
#| fig-width: 7
#| fig-height: 7

# create 6 scatterplots to illustrate covariance and correlation
par(mfrow=c(3,2)) 
set.seed(123) 
sigma = matrix(c(1, 0, 0, 1), 2, 2) 
x.mat = rmvnorm(50, sigma=sigma) 
plot(x.mat, pch = 16, main="(a)", cex=1.5, col="blue",xlab="x", ylab="y",      
     xlim=c(-1.5, 1.5), ylim=c(-2,2)) 
# panel (b): cor(x,y) = 1 
set.seed(123) 
sigma = matrix(c(1, 1, 1, 1), 2, 2) 
x.mat = rmvnorm(50, sigma=sigma) 
plot(x.mat, pch = 16, main="(b)",cex=1.5, col="blue",xlab="x", ylab="y",      
     xlim=c(-1.5, 1.5), ylim=c(-2,2)) 
# panel (c): cor(x,y) = -1 
sigma = matrix(c(1, -1, -1, 1), 2, 2) 
set.seed(123) 
x.mat = rmvnorm(50, sigma=sigma) 
plot(x.mat, pch = 16, main="(c)",cex=1.5, col="blue",xlab="x", ylab="y",      
     xlim=c(-1.5, 1.5), ylim=c(-2,2)) 
# panel (d): cor(x,y) > 0 
sigma = matrix(c(1, 0.75, 0.75, 1), 2, 2) 
set.seed(123) 
x.mat = rmvnorm(50, sigma=sigma) 
plot(x.mat, pch = 16, main="(d)",cex=1.5, col="blue",xlab="x", ylab="y",      
     xlim=c(-1.5, 1.5), ylim=c(-2,2)) 
# panel (e): cor(x,y) > 0 
sigma = matrix(c(1, -0.75, -0.75, 1), 2, 2) 
set.seed(123) 
x.mat = rmvnorm(50, sigma=sigma) 
plot(x.mat, pch = 16, main="(e)",cex=1.5, col="blue",xlab="x", ylab="y",      
     xlim=c(-1.5, 1.5), ylim=c(-2,2)) 
# panel (f): nonlinear 
sigma = matrix(c(0.5, 0, 0, 0.1), 2, 2) 
set.seed(123) 
x.mat = rmvnorm(50, sigma=sigma) 
x = x.mat[, 1] 
y = x^2 + x.mat[, 2] 
plot(x, y, pch = 16, main="(f)",cex=1.5, col="blue",xlab="x", ylab="y",      
     xlim=c(-1.5, 1.5), ylim=c(-2,2)) 
par(mfrow=c(1,1)) 
```

Let $X$ and $Y$ be two random variables with $E[X]=\mu_{X}$, $\mathrm{var}(X)=\sigma_{X}^{2},E[Y]=\mu_{Y}$
and $\mathrm{var}(Y)=\sigma_{Y}^{2}$.

::: {#def-covariance name="Covariance"}
The covariance between two random variables X
and Y is given by: 
\begin{align*}
\sigma_{XY} & =\mathrm{cov}(X,Y)=E[(X-\mu_{X})(Y-\mu_{Y})]\\
 & =\sum_{x\in S_{X}}\sum_{y\in S_{Y}}(x-\mu_{X})(y-\mu_{Y})\Pr(X=x,Y=y)\textrm{ for discrete } X, Y ,\\
 & =\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}(x-\mu_{X})(y-\mu_{Y})p(x,y)~dx~dy\textrm{ for continuous } X, Y.
\end{align*}
:::

::: {#def-correlation name="Correlation"}
The correlation between two random variables X and Y is given by: 
$$
\rho_{XY}=\mathrm{cor}(X,Y)=\frac{\mathrm{cov}(X,Y)}{\sqrt{\mathrm{var}(X)\mathrm{var}(Y)}}=\frac{\sigma_{XY}}{\sigma_{X}\sigma_{Y}}.
$$
The correlation coefficient, $\rho_{XY}$, is a scaled version of the covariance. 
:::

To see how covariance measures the direction of linear association,
consider the probability scatterplot in Figure @fig-ProbReviewCovariance.
In the figure, each pair of points occurs with equal probability.
The plot is separated into quadrants (right to left, top to bottom).
In the first quadrant (black circles), the realized values satisfy
$x<\mu_{X},y>\mu_{Y}$ so that the product $(x-\mu_{X})(y-\mu_{Y})<0$.
In the second quadrant (blue squares), the values satisfy $x>\mu_{X}$
and $y>\mu_{Y}$ so that the product $(x-\mu_{X})(y-\mu_{Y})>0$.
In the third quadrant (red triangles), the values satisfy $x<\mu_{X}$
and $y<\mu_{Y}$ so that the product $(x-\mu_{X})(y-\mu_{Y})>0$.
Finally, in the fourth quadrant (green diamonds), $x>\mu_{X}$ but
$y<\mu_{Y}$ so that the product $(x-\mu_{X})(y-\mu_{Y})<0$. Covariance
is then a probability weighted average all of the product terms in
the four quadrants. For the values in Figure @fig-ProbReviewCovariance,
this weighted average is positive because most of the values are in
the second and third quadrants.

```{r}
#| label: fig-ProbReviewCovariance
#| echo: false
#| fig-cap: Probability scatterplot of discrete distribution with positive covariance.
#|   Each pair $(X,Y)$ occurs with equal probability.
#| out-width: 80%
#| fig-asp: 1
#| fig-align: center

# Create plot to illustrate covariance calculation
sigma = matrix(c(1, 0.7, 0.7, 1), 2, 2)
set.seed(123)
x.mat = rmvnorm(20, sigma=sigma)
x = x.mat[, 1]
y = x.mat[, 2]
# define 4 quandrants 
q1 = (x < 0 & y > 0)
q2 = (x > 0 & y > 0)
q3 = (x < 0 & y < 0)
q4 = (x > 0 & y < 0)
# vector of plotting symbols
pch.vals = rep(16, 20)    # filled circles
pch.vals[q2] = 15         # filled squares
pch.vals[q3] = 17         # filled triangles
pch.vals[q4] = 18         # filled diamonds
col.vals = rep("black", 20)
col.vals[q2] = "blue"
col.vals[q3] = "red"
col.vals[q4] = "green"
plot(x, y, cex = 1.5, pch=pch.vals, col=col.vals,
     main = "Cov(x, y) > 0", xlim=c(-1.5, 1.5),
     xlab = expression(x - mu[x]), ylab = expression(y - mu[y]))
abline(h=0, v=0, lwd=2)
# 1st quadrant 
text(-1.5, 1.6, labels="QI") 
text(-1.35, 1.45, labels=expression(x - mu[x] < 0)) 
text(-1.35, 1.30, labels=expression(y - mu[y] > 0)) 
# 2nd quadrant 
text(0.15, 1.6, labels="QII") 
text(0.4, 1.45, labels=expression(x - mu[x] > 0)) 
text(0.4, 1.30, labels=expression(y - mu[y] > 0)) 
# 3rd quadrant 
text(-1.5, -0.2, labels="QIII") 
text(-1.35, -0.35, labels=expression(x - mu[x] < 0)) 
text(-1.35, -0.5, labels=expression(y - mu[y] < 0)) 
# 4th quadrant 
text(0.15, -0.2, labels="QIV") 
text(0.4, -0.35, labels=expression(x - mu[x] > 0)) 
text(0.4, -0.5, labels=expression(y - mu[y] < 0))
```

::: {#exm-example246 name="Calculate covariance and correlation for discrete random variables"}
For the data in Table @tbl-bivariateDistn, you get: 
\begin{gather*}
\sigma_{XY}=\mathrm{cov}(X,Y)=(0-3/2)(0-1/2)\cdot1/8+(0-3/2)(1-1/2)\cdot0\\
+\cdots+(3-3/2)(1-1/2)\cdot1/8=1/4\\
\rho_{XY}=\mathrm{cor}(X,Y)=\frac{1/4}{\sqrt{(3/4)\cdot(1/2)}}=0.577
\end{gather*}

\hfill{}$\blacksquare$
:::

#### Properties of covariance and correlation

Let $X$ and $Y$ be random variables and let $a$ and $b$ be constants.
Some important properties of $\mathrm{cov}(X,Y)$ are summarized in the following
Proposition: 

::: {#prp-covariance-properties name="Properties of covariance"}
1.  $\mathrm{cov}(X,X)=\mathrm{var}(X)$ 
2.  $\mathrm{cov}(X,Y)=\mathrm{cov}(Y,X)$ 
3.  $\mathrm{cov}(X,Y)=E[XY]-E[X]E[Y]=E[XY]-\mu_{X}\mu_{Y}$ 
4.  $\mathrm{cov}(aX,bY)=a\cdot b\cdot\mathrm{cov}(X,Y)$ 
5.  If $X$ and $Y$ are independent then $\mathrm{cov}(X,Y)=0$ (no association $\Longrightarrow$ no linear association). However, if $\mathrm{cov}(X,Y)=0$ then $X$ and $Y$ are not necessarily independent (no linear association $\nRightarrow$ no association). 
6.  If $X$ and $Y$ are jointly normally distributed and $\mathrm{cov}(X,Y)=0$, then $X$ and $Y$ are independent. 
:::

The first two properties are intuitive. The third property results
from expanding the definition of covariance: 
\begin{align*}
\mathrm{cov}(X,Y) & =E[(X-\mu_{X})(Y-\mu_{Y})]\\
 & =E\left[XY-X\mu_{Y}-\mu_{X}Y+\mu_{X}\mu_{Y}\right]\\
 & =E[XY]-E[X]\mu_{Y}-\mu_{X}E[Y]+\mu_{X}\mu_{Y}\\
 & =E[XY]-2\mu_{X}\mu_{Y}+\mu_{X}\mu_{Y}\\
 & =E[XY]-\mu_{X}\mu_{Y}
\end{align*}
The fourth property follows from the linearity of expectations: 
\begin{align*}
\mathrm{cov}(aX,bY) & =E[(aX-a\mu_{X})(bY-b\mu_{Y})]\\
 & =a\cdot b\cdot E[(X-\mu_{X})(Y-\mu_{Y})]\\
 & =a\cdot b\cdot\mathrm{cov}(X,Y)
\end{align*}
The fourth property shows that the value of $\mathrm{cov}(X,Y)$ depends
on the scaling of the random variables $X$ and $Y$. By simply changing
the scale of $X$ or $Y$ you can make $\mathrm{cov}(X,Y)$ equal to
any value that you want. Consequently, the numerical value of $\mathrm{cov}(X,Y)$
is not informative about the strength of the linear association between
$X$ and $Y$. However, the sign of $\mathrm{cov}(X,Y)$ is informative
about the direction of linear association between $X$ and $Y$. 

The fifth property should be intuitive. Independence between the random
variables $X$ and $Y$ means that there is no relationship, linear
or nonlinear, between $X$ and $Y$. However, the lack of a linear
relationship between $X$ and $Y$ does not preclude a nonlinear relationship.

The last result illustrates an important property of the normal distribution:
lack of covariance implies independence. 


Some important properties of $\mathrm{cor}(X,Y)$ are: 

::: {#prp-correlation-properties names="Properties of correlation"}
1.  $-1\leq\rho_{XY}\leq1$ 
2.  If $\rho_{XY}=1$ then $X$ and $Y$ are perfectly positively linearly related. That is, $Y=aX+b$ where $a>0$. 
3.  If $\rho_{XY}=-1$ then $X$ and $Y$ are perfectly negatively linearly related. That is, $Y=aX+b$ where $a<0$. 
4.  If $\rho_{XY}=0$ then $X$ and $Y$ are not linearly related but may be nonlinearly related. 
5.  $\mathrm{cor}(aX,bY)=\mathrm{cor}(X,Y)$ if $a>0$ and $b>0$; $\mathrm{cor}(X,Y)=-\mathrm{cor}(X,Y)$ if $a>0,b<0$ or $a<0,b>0$. 
:::

**TBD: discuss the results briefly**

### Bivariate normal distributions

Let $X$ and $Y$ be distributed bivariate normal. The joint pdf is
given by: 
$$
\begin{aligned}
&f(x,y)=\frac{1}{2\pi\sigma_{X}\sigma_{Y}\sqrt{1-\rho_{XY}^{2}}}\times\\
&\exp\left\{ -\frac{1}{2(1-\rho_{XY}^{2})}\left[\left(\frac{x-\mu_{X}}{\sigma_{X}}\right)^{2}+\left(\frac{y-\mu_{Y}}{\sigma_{Y}}\right)^{2}-\frac{2\rho_{XY}(x-\mu_{X})(y-\mu_{Y})}{\sigma_{X}\sigma_{Y}}\right]\right\} \nonumber 
\end{aligned}
$$ {#eq-bivariate-normal}
where $E[X]=\mu_{X}$, $E[Y]=\mu_{Y}$, $\mathrm{sd}(X)=\sigma_{X}$,
$\mathrm{sd}(Y)=\sigma_{Y}$, and $\rho_{XY}=\mathrm{cor}(X,Y)$.
The correlation coefficient $\rho_{XY}$ describes the dependence
between $X$ and $Y$. If $\rho_{XY}=0$ then the pdf collapses to
the pdf of the standard bivariate normal distribution. In this case, $X$ and $Y$
are independent random variables since $f(x,y)=f(x)f(y).$

It can be shown that the marginal distributions of $X$ and $Y$ are
normal: $X\sim N(\mu_{X},\sigma_{X}^{2})$, $Y\sim N(\mu_{Y},\sigma_{Y}^{2})$.
In addition, it can be shown that the conditional distributions $f(x|y)$
and $f(y|x)$ are also normal with means given by: 
\begin{align*}
\mu_{X|Y=y} & =\alpha_{X}+\beta_{X}\cdot y,\\
\mu_{Y|X=x} & =\alpha_{Y}+\beta_{Y}\cdot x,
\end{align*}
where, 
\begin{align*}
\alpha_{X} & =\mu_{X}-\beta_{X}\mu_{X},~\beta_{X}=\sigma_{XY}/\sigma_{Y}^{2},\\
\alpha_{Y} & =\mu_{Y}-\beta_{Y}\mu_{X},~\beta_{Y}=\sigma_{XY}/\sigma_{X}^{2},
\end{align*}
and variances given by, 
\begin{align*}
\sigma_{X|Y=y}^{2} & =\sigma_{X}^{2}-\sigma_{XY}^{2}/\sigma_{Y}^{2},\\
\sigma_{X|Y=y}^{2} & =\sigma_{Y}^{2}-\sigma_{XY}^{2}/\sigma_{X}^{2}.
\end{align*}
Notice that the conditional means (regression functions) are linear functions of $x$ and
$y$, respectively.

::: {#exm-example247 name="Expressing the bivariate normal distribution using matrix algebra"}
The formula for the bivariate normal distribution @eq-bivariate-normal
is a bit messy. you can greatly simplify the formula by using matrix
algebra. Define the $2\times1$ vectors $\mathbf{x}=(x,y)^{\prime}$
and $\boldsymbol{\mu}=(\mu_{X},\mu_{y})^{\prime},$ and the $2\times2$
matrix: 
$$
\boldsymbol{\Sigma}=\left(\begin{array}{cc}
\sigma_{X}^{2} & \sigma_{XY}\\
\sigma_{XY} & \sigma_{Y}^{2}
\end{array}\right)
$$
Then the bivariate normal distribution @eq-bivariate-normal
may be compactly expressed as: 
$$
f(\mathbf{x})=\frac{1}{2\pi\det(\boldsymbol{\Sigma})^{1/2}}e^{-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^{\prime}\boldsymbol{\Sigma}^{-1}(\mathbf{x}-\boldsymbol{\mu})}
$$ {#eq-bivariate-normal-matrix}
where, 
\begin{align*}
\det(\boldsymbol{\Sigma})&=\sigma_{X}^{2}\sigma_{Y}^{2}-\sigma_{XY}^{2}=\sigma_{X}^{2}\sigma_{Y}^{2}-\sigma_{X}^{2}\sigma_{Y}^{2}\rho_{XY}^{2}=\sigma_{X}^{2}\sigma_{Y}^{2}(1-\rho_{XY}^{2})\\
\boldsymbol{\Sigma}^{-1} &= \frac{1}{\det(\boldsymbol{\Sigma})} \left(\begin{array}{cc}
\sigma_{Y}^{2} & -\sigma_{XY}\\
-\sigma_{XY} & \sigma_{X}^{2}
\end{array}\right)
\end{align*}

\hfill{}$\blacksquare$
:::

::: {#exm-example248 name="Plotting the bivariate normal distribution"}
The R package **mvtnorm** contains the functions `dmvnorm()`,
`pmvnorm()`, and `qmvnorm()` which can be used to compute
the bivariate normal pdf, cdf and quantiles, respectively. Plotting
the bivariate normal distribution over a specified grid of $x$ and
$y$ values in R can be done with the `persp()` function. First,
I specify the parameter values for the joint distribution. Here,
I choose $\mu_{X}=\mu_{Y}=0$, $\sigma_{X}=\sigma_{Y}=1$ and $\rho=0.5$.
I will use the `dmvnorm()` function to evaluate the joint
pdf at these values. To do so, I must specify a covariance matrix
of the form: 
$$
\boldsymbol{\Sigma}=\left(\begin{array}{cc}
\sigma_{X}^{2} & \sigma_{XY}\\
\sigma_{XY} & \sigma_{Y}^{2}
\end{array}\right)=\left(\begin{array}{cc}
1 & 0.5\\
0.5 & 1
\end{array}\right).
$$
In R this matrix can be created using:

```{r}
#| echo: true

sigma = matrix(c(1, 0.5, 0.5, 1), 2, 2)
```

Next I specify a grid of $x$ and $y$ values between $-3$ and $3$:

```{r}
#| echo: true

x = seq(-3, 3, length=25)
y = seq(-3, 3, length=25)
```

To evaluate the joint pdf over the two-dimensional grid I can use
the `outer()` function:

```{r}
#| echo: true

# function to evaluate bivariate normal pdf on grid
bv.norm <- function(x, y, sigma) {
  z = cbind(x,y)
  return(dmvnorm(z, sigma=sigma))
}
# use outer function to evaluate pdf on 2D grid of x-y values
fxy = outer(x, y, bv.norm, sigma)
```

To create the 3D plot of the joint pdf, I use the `persp()` function:

```{r}
#| label: fig-ProbReviewBivariateNormal
#| echo: true
#| fig-cap: Bivariate normal pdf with $\mu_{X}=\mu_{Y}=0$, $\sigma_{X}\sigma_{Y}=1$ and
#|   $\rho=0.5$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

persp(x, y, fxy, theta=60, phi=30, expand=0.5, ticktype="detailed", 
      zlab="f(x,y)",col="grey")
```

The resulting plot is given in Figure @fig-ProbReviewBivariateNormal.
\hfill{}$\blacksquare$
:::

### Expectation and variance of the sum of two random variables

::: {#prp-sum-two-rv name="Expectation and variance of the sum of two random variables"}
Let $X$ and $Y$ be two random variables with well defined means,
variances and covariance and let $a$ and $b$ be constants. Then
the following results hold: 
\begin{align*}
E[aX+bY] & =aE[X]+bE[Y]=a\mu_{X}+b\mu_{Y}  \\
\mathrm{var}(aX+bY) & =a^{2}\mathrm{var}(X)+b^{2}\mathrm{var}(Y)+2\cdot a\cdot b\cdot\mathrm{cov}(X,Y)\\
 & =a^{2}\sigma_{X}^{2}+b^{2}\sigma_{Y}^{2}+2\cdot a\cdot b\cdot\sigma_{XY} 
\end{align*}
:::

The first result states that the expected value of a linear combination
of two random variables is equal to a linear combination of the expected
values of the random variables. This result indicates that the expectation
operator is a linear operator. In other words, expectation is additive.
The second result states that variance of a linear combination of
random variables is not a linear combination of the variances of the
random variables. In particular, notice that covariance comes up as
a term when computing the variance of the sum of two (not independent)
random variables. Hence, the variance operator is not, in general,
a linear operator. That is, variance, in general, is not additive.


It is instructive to go through the derivation of these results. Let
$X$ and $Y$ be discrete random variables. Then, 
\begin{align*}
E[aX+bY] & =\sum_{x\in S_{X}}\sum_{y\in S_{y}}(ax+by)\Pr (X=x,Y=y)\\
 & =\sum_{x\in S_{X}}\sum_{y\in S_{y}}ax\Pr(X=x,Y=y)+\sum_{x\in S_{X}}\sum_{y\in S_{y}}bx\Pr(X=x,Y=y)\\
 & =a\sum_{x\in S_{X}}x\sum_{y\in S_{y}}\Pr(X=x,Y=y)+b\sum_{y\in S_{y}}y\sum_{x\in S_{X}}\Pr(X=x,Y=y)\\
 & =a\sum_{x\in S_{X}}x\Pr(X=x)+b\sum_{y\in S_{y}}y\Pr(Y=y)\\
 & =aE[X]+bE[Y]=a\mu_{X}+b\mu_{Y}.
\end{align*}
The result for continuous random variables is similar. Effectively,
the summations are replaced by integrals and the joint probabilities
are replaced by the joint pdf. 

Next, let $X$ and $Y$ be discrete or continuous random variables. Then, using the
definition of variance you get:
\begin{align*}
\mathrm{var}(aX+bY) & =E[(aX+bY-E[aX+bY])^{2}]\\
 & =E[(aX+bY-a\mu_{X}-b\mu_{Y})^{2}]\\
 & =E[(a(X-\mu_{X})+b(Y-\mu_{Y}))^{2}]\\
 & =a^{2}E[(X-\mu_{X})^{2}]+b^{2}E[(Y-\mu_{Y})^{2}]+2\cdot a\cdot b\cdot E[(X-\mu_{X})(Y-\mu_{Y})]\\
 & =a^{2}\mathrm{var}(X)+b^{2}\mathrm{var}(Y)+2\cdot a\cdot b\cdot\mathrm{cov}(X,Y).
\end{align*}


#### Linear combination of two normal random variables

The following proposition gives an important result concerning a linear
combination of normal random variables.

::: {#prp-linear-combination-normal name="Linear combination of normal random variables"}
Let $X\sim N(\mu_{X},\sigma_{X}^{2})$, $Y\sim N(\mu_{Y},\sigma_{Y}^{2})$,
$\sigma_{XY}=\mathrm{cov}(X,Y)$ and $a$ and $b$ be constants. Define
the new random variable $Z$ as: 
$$
Z=aX+bY.
$$
Then, 
$$
Z\sim N(\mu_{Z},\sigma_{Z}^{2}),
$$
where $\mu_{Z}=a\mu_{X}+b\mu_{Y}$ and  $\sigma_{Z}^{2}=a^{2}\sigma_{X}^{2}+b^{2}\sigma_{Y}^{2}+2ab\sigma_{XY}$.
:::

This important result states that a linear combination of two normally
distributed random variables is itself a normally distributed random
variable. The proof of the result relies on the change of variables
theorem from calculus and is omitted. Not all random variables have
the nice property that their distributions are closed under addition.

::: {#exm-example248 name="Portfolio of two assets"}
Consider a portfolio of two stocks $A$ (Amazon) and $B$ (Boeing)
with investment shares $x_{A}$ and $x_{B}$ with $x_{A}+x_{B}=1$.
Let $R_{A}$ and $R_{B}$ denote the simple monthly returns on these
assets, and assume that $R_{A}\sim N(\mu_{A},\sigma_{A}^{2})$ and
$R_{B}\sim N(\mu_{B},\sigma_{B}^{2})$. Furthermore, let $\sigma_{AB}=\rho_{AB}\sigma_{A}\sigma_{B}=\mathrm{cov}(R_{A},R_{B})$.
The portfolio return is $R_{p}=x_{A}R_{A}+x_{B}R_{B}$, which is a
linear function of two random variables. Using the properties of linear
functions of random variables, you get: 
\begin{align*}
\mu_{p} & =E[R_{p}]=x_{A}E[R_{A}]+x_{B}E[R_{B}]=x_{A}\mu_{A}+x_{B}\mu_{B}\\
\sigma_{p}^{2} & =\mathrm{var}(R_{p})=x_{A}^{2}\mathrm{var}(R_{A})+x_{B}^{2}\mathrm{var}(R_{B})+2x_{A}x_{B}\mathrm{cov}(R_{A},R_{B})\\
 & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}.
\end{align*}

\hfill{}$\blacksquare$
:::

## Multivariate Distributions{#sec-Multivariate-Distributions}
\index{Multivariate distributions}
*Multivariate distributions* are used to characterize the joint distribution
of a collection of $N$ random variables $X_{1},X_{2},\ldots,X_{N}$
for $N>1$. The mathematical formulation of this joint distribution
can be quite complex and typically makes use of matrix algebra. Here,
I summarize some basic properties of multivariate distributions without
the use of matrix algebra. In chapter @sec-Matrix-Algebra-Review,
I will show how matrix algebra can greatly simplify the description of
multivariate distributions.

### Discrete random variables

Let $X_{1},X_{2},\ldots,X_{N}$ be $N$ discrete random variables
with sample spaces $S_{X_{1}},S_{X_{2}},\ldots,S_{X_{N}}$. The likelihood
that these random variables take values in the joint sample space
$S_{X_{1}}\times S_{X_{2}}\times\cdots\times S_{X_{N}}$ is given
by the joint probability function: 
$$
p(x_{1},x_{2},\ldots,x_{N})=\Pr(X_{1}=x_{1},X_{2}=x_{2},\ldots,X_{N}=x_{N}).
$$
For $N>2$ it is not easy to represent the joint probabilities in
a table like Table @tbl-bivariateDistn or to visualize
the distribution. 

Marginal distributions for each variable $X_{i}$ can be derived from
the joint distribution as in @eq-marginal-X-discrete by summing
the joint probabilities over the other variables $j\neq i$. For example,
$$
p(x_{1})=\sum_{x_{2}\in S_{X_{2}},\ldots,x_{N}\in S_{X_{N}}}p(x_{1},x_{2},\ldots,x_{N}).
$$
With $N$ random variables, there are numerous conditional distributions
that can be formed. For example, the distribution of $X_{1}$ given
$X_{2}=x_{2},\ldots,X_{N}=x_{N}$ is determined using: 
$$
\Pr(X_{1}=x_{1}|X_{2}=x_{2},\ldots,X_{N}=x_{N})=\frac{\Pr(X_{1}=x_{1},X_{2}=x_{2},\ldots,X_{N}=x_{N})}{\Pr(X_{2}=x_{2},\ldots,X_{N}=x_{N})}.
$$
Similarly, the joint distribution of $X_{1}$ and $X_{2}$ given $X_{3}=x_{3},\ldots,X_{N}=x_{N}$
is given by: 
\begin{align*}
\Pr(X_{1}=x_{1},X_{2}=x_{2}|X_{3}=x_{3},\ldots,X_{N}=x_{N}) \\
=\frac{\Pr(X_{1}=x_{1},X_{2}=x_{2},\ldots,X_{N}=x_{N})}{\Pr(X_{3}=x_{3},\ldots,X_{N}=x_{N})}.
\end{align*}


### Continuous random variables

Let $X_{1},X_{2},\ldots,X_{N}$ be $N$ continuous random variables
each taking values on the real line. The joint pdf is a function $f(x_{1},x_{2},\ldots,x_{N})\geq0$
such that: 
$$
\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\cdots\int_{-\infty}^{\infty}f(x_{1},x_{2},\ldots,x_{N})~dx_{1}~dx_{2}\cdots~dx_{N}=1.
$$
Joint probabilities of $x_{11}\leq X_{1}\leq x_{12},x_{21}\leq X_{2}\leq x_{22},\ldots,x_{N1}\leq X_{N}\leq x_{N2}$
are computed by solving the integral equation: 
$$
\int_{x_{11}}^{x_{12}}\int_{x_{21}}^{x_{22}}\cdots\int_{x_{N1}}^{x_{N2}}f(x_{1},x_{2},\ldots,x_{N})~dx_{1}~dx_{2}\cdots~dx_{N}.
$$ {#eq-joint-probability}
For most multivariate distributions, the integral in @eq-joint-probability
cannot be solved analytically and must be approximated numerically.


The marginal pdf for $x_{i}$ is found by integrating the joint pdf
with respect to the other variables. For example, the marginal pdf
for $x_{1}$ is found by solving: 
$$
f(x_{1})=\int_{-\infty}^{\infty}\cdots\int_{-\infty}^{\infty}f(x_{1},x_{2},\ldots,x_{N})~dx_{2}\cdots~dx_{N}.
$$
Conditional pdf for a single random variable or a collection of random
variables are defined in the obvious way.

### Independence

A collection of $N$ random variables are independent if their joint
distribution factors into the product of all of the marginal distributions:
\begin{align*}
p(x_{1},x_{2},\ldots,x_{N}) & =p(x_{1})p(x_{2})\cdots p(x_{N})\textrm{ for }X_{i}\textrm{ discrete},\\
f\left(x_{1},x_{2},\ldots,x_{N}\right) & =f(x_{1})f(x_{2})\cdots f(x_{N})\textrm{ for }X_{i}\textrm{ continuous}.
\end{align*}
In addition, if $N$ random variables are independent then any functions
of these random variables are also independent.

### Dependence concepts

In general, it is difficult to define dependence concepts for collections
of more than two random variables. Dependence is typically only defined
between pairwise random variables. Hence, covariance and correlation
are also useful concepts when dealing with more than two random variables.

For $N$ random variables $X_{1},X_{2},\ldots,X_{N}$, with mean values
$\mu_{i}=E[X_{i}]$ and variances $\sigma_{i}^{2}=\mathrm{var}(X_{i})$,
the pairwise covariances and correlations are defined as: 
\begin{align*}
\mathrm{cov}(X_{i},X_{j}) & =\sigma_{ij}=E[(X_{i}-\mu_{i})(X_{j}-\mu_{i})],\\
\mathrm{cov}(X_{i},X_{j}) & =\rho_{ij}=\frac{\sigma_{ij}}{\sigma_{i}\sigma_{j}},
\end{align*}
for $i\neq j$. There are $N(N-1)/2$ pairwise covariances and correlations.
Often, these values are summarized using matrix algebra in an $N\times N$
covariance matrix $\boldsymbol{\Sigma}$ and an $N\times N$ correlation
matrix $\mathbf{C}$: 
$$
\begin{aligned}
\boldsymbol{\Sigma} & =\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \cdots & \sigma_{1N}\\
\sigma_{12} & \sigma_{2}^{2} & \cdots & \sigma_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1N} & \sigma_{2N} & \cdots & \sigma_{N}^{2}
\end{array}\right),
\end{aligned}
$$ {#eq-covariance-matrix}
$$
\begin{aligned}
\mathbf{C} & =\left(\begin{array}{cccc}
1 & \rho_{12} & \cdots & \rho_{1N}\\
\rho_{12} & 1 & \cdots & \rho_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\rho_{1N} & \rho_{2N} & \cdots & 1
\end{array}\right).
\end{aligned}
$$ {#eq-correlation-matrix}
These matrices are formally defined in chapter @sec-Matrix-Algebra-Review.

### Linear combinations of $N$ random variables

Many of the results for manipulating a collection of random variables
generalize in a straightforward way to the case of more than two random
variables. The details of the generalizations are not important for
our purposes. However, the following results will be used repeatedly
throughout the book. 

::: {#prp-linear-combination name="Linear combination of random variables"}
Let $X_{1},X_{2},\ldots,X_{N}$ denote a collection of $N$ random
variables (discrete or continuous) with means $\mu_{i}$, variances
$\sigma_{i}^{2}$ and covariances $\sigma_{ij}$. Define the new random
variable $Z$ as a linear combination: 
$$
Z=a_{1}X_{1}+a_{2}X_{2}+\cdots+a_{N}X_{N}
$$
where $a_{1},a_{2},\ldots,a_{N}$ are constants. Then the following
results hold: 
$$
\begin{aligned}
\mu_{Z} & =E[Z]=a_{1}E[X_{1}]+a_{2}E[X_{2}]+\cdots+a_{N}E[X_{N}]\\
 & =\sum_{i=1}^{N}a_{i}E[X_{i}]=\sum_{i=1}^{N}a_{i}\mu_{i}.\nonumber 
\end{aligned}
$$ {#eq-mean-linear-combination}
$$
\begin{aligned}
\sigma_{Z}^{2} & =\mathrm{var}(Z)=a_{1}^{2}\sigma_{1}^{2}+a_{2}^{2}\sigma_{2}^{2}+\cdots+a_{N}^{2}\sigma_{N}^{2}\\
 & +2a_{1}a_{2}\sigma_{12}+2a_{1}a_{3}\sigma_{13}+\cdots+a_{1}a_{N}\sigma_{1N}\nonumber \\
 & +2a_{2}a_{3}\sigma_{23}+2a_{2}a_{4}\sigma_{24}+\cdots+a_{2}a_{N}\sigma_{2N}\nonumber \\
 & +\cdots+\nonumber \\
 & +2a_{N-1}a_{N}\sigma_{(N-1)N}\nonumber \\
 & =\sum_{i=1}^{N}a_{i}^{2}\sigma_{i}^{2}+2\sum_{i=1}^{N}\sum_{j\neq i}a_{i}a_{j}\sigma_{ij}.\nonumber 
\end{aligned}
$$ {#eq-variance-linear-combination}
:::

The derivation of these results is very similar to the bivariate case
and so is omitted. In addition, if all of the $X_{i}$ are normally
distributed then $Z$ is also normally distributed with mean $\mu_{Z}$
and variance $\sigma_{Z}^{2}$ as described above. 

The variance of a linear combination of $N$ random variables contains
$N$ variance terms and $N(N-1)$ covariance terms. For $N=2,5,10$
and 100 the number of covariance terms in $\mathrm{var}(Z)$ is $2,\,20,\,90$
and $9,900$, respectively. Notice that when $N$ is large there are
many more covariance terms than variance terms in $\mathrm{var}(Z)$.

The expression for $\mathrm{var}(Z)$ is messy. It can be simplified
using matrix algebra notation, as explained in detail in chapter @sec-Matrix-Algebra-Review.
To preview, define the $N\times1$ vectors $\mathbf{X}=(X_{1},\ldots,X_{N})^{\prime}$
and $\mathbf{a}=(a_{1},\ldots,a_{N})^{\prime}$. Then $Z=\mathbf{a}^{\prime}\mathbf{X}$
and $\mathrm{var}(Z)=\mathrm{var}(\mathbf{a}^{\prime}\mathbf{X})=\mathbf{a}^{\prime}\boldsymbol{\Sigma} \mathbf{a}$,
where $\boldsymbol{\Sigma}$ is the $N\times N$ covariance matrix, which is much
more compact than @eq-variance-linear-combination.

::: {#exm-example249 name="Square-root-of-time rule for multi-period continuously compounded returns"}
Let $r_{t}$ denote the continuously compounded monthly return on
an asset at times $t=1,\ldots,12.$ Assume that $r_{1},\ldots,r_{12}$
are independent and identically distributed (iid) $N(\mu,\sigma^{2}).$
Recall, the annual continuously compounded return is equal to the sum
of twelve monthly continuously compounded returns: $r_{A}=r(12)=\sum_{t=1}^{12}r_{t}.$ Since
each monthly return is normally distributed, the annual return is
also normally distributed. The mean of $r(12)$ is: 
\begin{align*}
E[r(12)] & =E\left[\sum_{t=1}^{12}r_{t}\right]\\
 & =\sum_{t=1}^{12}E[r_{t}]\textrm{ (by linearity of expectation)},\\
 & =\sum_{t=1}^{12}\mu\textrm{ (by identical distributions)},\\
 & =12\cdot\mu.
\end{align*}
Hence, the expected 12-month (annual) return is equal to 12 times
the expected monthly return. The variance of $r(12)$ is: 
\begin{align*}
\mathrm{var}(r(12)) & =\mathrm{var}\left(\sum_{t=1}^{12}r_{t}\right)\\
 & =\sum_{t=1}^{12}\mathrm{var}(r_{t})\textrm{ (by independence)},\\
 & =\sum_{j=0}^{11}\sigma^{2}\textrm{ (by identical distributions)},\\
 & =12\cdot\sigma^{2},
\end{align*}
so that the annual variance is also equal to 12 times the monthly
variance\footnote{This result often causes some confusion. It is easy to make the mistake
and say that the annual variance is $(12)^{2}=144$ time the monthly
variance. This result would occur if $r_{A}=12r_{t}$, so that $\mathrm{var}(r_{A})=(12)^{2}\mathrm{var}(r_{t})=144\mathrm{var}(r_{t})$.}. Hence, the annual standard deviation is $\sqrt{12}$ times the monthly
standard deviation: $\mathrm{sd}(r(12))=\sqrt{12}\sigma$ (this result
is known as the *square-root-of-time rule*). Therefore, $r(12)\sim N(12\mu,12\sigma^{2})$.
\hfill{}$\blacksquare$
:::

### Covariance between linear combinations of random variables

Consider the linear combinations of two random variables: 
\begin{align*}
Y & =aX_{1}+bX_{2},\\
Z & =cX_{3}+dX_{4},
\end{align*}
where $a,b,c$ and $d$ are constants. The covariance between $Y$
and $Z$ is, 
\begin{align*}
\mathrm{cov}(Y,Z) & =\mathrm{cov}(aX_{1}+bX_{2},cX_{3}+dX_{4})\\
 & =E[((aX_{1}+bX_{2})-(a\mu_{1}+b\mu_{2}))((cX_{3}+dX_{4})-(c\mu_{3}+d\mu_{4}))]\\
 & =E[(a(X_{1}-\mu_{1})+b(X_{2}-\mu_{2}))(c(X_{3}-\mu_{3})+d(X_{4}-\mu_{4}))]\\
 & =acE[(X_{1}-\mu_{1})(X_{3}-\mu_{3})]+adE[(X_{1}-\mu_{1})(X_{4}-\mu_{4})]\\
 & +bcE[(X_{2}-\mu_{2})(X_{3}-\mu_{3})]+bdE[(X_{2}-\mu_{2})(X_{4}-\mu_{4})]\\
 & =ac\mathrm{cov}(X_{1},X_{3})+ad\mathrm{cov}(X_{1},X_{4})+bc\mathrm{cov}(X_{2},X_{3})+bd\mathrm{cov}(X_{2},X_{4}).
\end{align*}
Hence, covariance is additive for linear combinations of random variables.
The result above extends in an obvious way to arbitrary linear combinations
of random variables.

## Further Reading: Review of Random Variables

This material in this chapter is a selected survey of topics, with
applications to finance, typically covered in an introductory probability
and statistics course using calculus. A good textbook for such a course
is [@degroot-2011]. Chapter 1 of [@carmona-2014] and
the appendix of [@RuppertMatteson-2015] give a slightly more advanced
review than is presented here. Two slightly more advanced books, oriented
towards economics students, are [@amemiya-1994] and [@goldberger-1991].
Good introductory treatments of probability and statistics using R
include [@dalgaard-2002] and [@verzani-2005]. [@pfaff-2013] discusses
some useful financial asset return distributions and their implementation
in R.

\newpage

## Exercises

::: {#exr-exercise1}
Suppose $X$ is a normally distributed random variable with mean $0.05$
and variance $(0.10)^{2}.$ That is, $X\sim N(0.05,\,(0,10)^{2})$.
Use the `pnorm()` and `qnorm()` functions in R to compute
the following:

1.  $\Pr(X\geq0.10)$, $\Pr(X\leq-0.10)$.
2.  $\Pr(-0.05\leq X\leq0.15)$.
3.  $1\%$ and $5\%$ quantiles, $q_{0.01}^{R}$ and $q_{0.05}^{R}$.
4.  $95\%$ and $99\%$ quantiles, $q_{0.95}^{R}$ and $q_{0.99}^{R}$.
:::

::: {#exr-exercise2}
Let $X$ denote the monthly return on Microsoft Stock and let $Y$
denote the monthly return on Starbucks stock. Assume that $X\sim N(0.05,\,(0.10)^{2})$
and $Y\sim N(0.025,\,(0.05)^{2})$. 

1.  Using a grid of values between $-0.25$ and $0.35$, plot the normal curves for $X$ and $Y$. Make sure that both normal curves are on the same plot.
2.  Plot the points $(\sigma_{X},\mu_{X})=(0.10,0.05)$ and $(\sigma_{Y},\mu_{Y})=(0.05,0.025)$ in an x-y plot.  
3.  Comment on the risk-return tradeoffs for the two stocks.
:::

::: {#exr-exercise3}
Let $X_{1},\,X_{2},\,Y_{1},$ and $Y_{2}$ be random variables. Show that 
\begin{align*}
\mathrm{Cov}(X_{1}+X_{2},Y_{1}+Y_{2}) = & \mathrm{Cov}(X_{1},Y_{1}) + \mathrm{Cov}(X_{1},Y_{2}) +\\
&\mathrm{Cov}(X_{2},Y_{1}) + \mathrm{Cov}(X_{2},Y_{2}).
\end{align*}
:::

::: {#exr-exercise4}
Let $X$ be a random variable with $\mathrm{var}(X)<\infty.$ Show that $\mathrm{var}(X)=E[(X-\mu_{X})^{2}]=E[X^{2}]-\mu_{X}^{2}.$
:::

::: {#exr-exercise5}
Let $X$ and $Y$ be a random variables such that $\mathrm{cov}(X,Y)<\infty$.

1.  If $Y=aX+b$ where $a>0$ show that $\rho_{XY}=\mathrm{cor}(X,Y)=1.$
2.  If $Y=aX+b$ where $a<0$ show that $\rho_{XY}=\mathrm{cor}(X,Y)=-1.$
:::

::: {#exr-exercise6}
Let $R$ denote the simple monthly return on Microsoft stock and let
$W_{0}$ denote initial wealth to be invested over the month. Assume
that $R\sim N(0.04,\,(0.09)^{2})$ and that $W_{0}=\$100,000$. Determine the 1\% and 5\% Value-at-Risk (VaR) over the month on the investment. That is, determine the loss in investment value that may occur over the next month with 1\% probability and with 5\% probability.
:::

::: {#exr-exercise7}
Let $R_{t}$ denote the simple monthly return and assume $R_{t}\sim iid\,N(\mu,\sigma^{2}).$
Consider the 2-period return $R_{t}(2)=(1+R_{t})(1+R_{t-1})-1$.

1.  Assuming that $\mathrm{cov}(R_{t},R_{t-1})=0,$ show that $E[R_{t}R_{t-1}]=\mu^{2}.$ Hint: use $\mathrm{cov}(R_{t},R_{t-1})=E[R_{t}R_{t-1}]-E[R_{t}]E[R_{t-1}]$.
2.  Show that $E[R_{t}(2)]=(1+\mu^{2})-1$.
3.  Is $R_{t}(2)$ normally distributed? Justify your answer.
:::


::: {#exr-exercise8}
Let $r$ denote the continuously compounded monthly return on Microsoft
stock and let $W_{0}$ denote initial wealth to be invested over the
month. Assume that $r\sim iid\,N(0.04,(0.09)^{2})$ and that $W_{0}=\$100,000.$

1.  Determine the 1\% and 5\% value-at-risk (VaR) over the month on the investment. That is, determine the loss in investment value that may occur over the next month with 1\% probability and with 5\% probability. (Hint: compute the 1\% and 5\% quantile from the Normal distribution
for $r$ and then convert the continuously compounded return quantile to a simple return quantile using the transformation $R=e^{r}-1$.)
2.  Let $r(12)=r_{1}+r_{2}+\cdots+r_{12}$ denote the 12-month (annual) continuously compounded return. Compute $E[r(12)]$, $\mathrm{var}(r(12))$ and $\mathrm{sd}(r(12))$. What is the probability distribution of $r(12)$? 
3.  Using the probability distribution for $r(12)$, determine the 1\% and 5\% value-at-risk (VaR) over the year on the investment. 
:::

::: {#exr-exercise9}
Let $R_{VFINX}$ and $R_{AAPL}$ denote the monthly simple returns
on VFINX (Vanguard S\&P 500 index) and AAPL (Apple stock). Suppose
that $R_{VFINX}\sim i.i.d.\,N(0.013,(0.037)^{2})$, $R_{AAPL}\sim i.i.d.\,N(0.028,(0.073)^{2})$.

1.  Sketch the normal distributions for the two assets on the same graph. Show the mean values and the ranges within 2 standard deviations of the mean. Which asset appears to be more risky?
2.  Plot the risk-return tradeoff for the two assets. That is, plot the mean values of each asset on the y-axis and the standard deviations on the x-axis. Comment on the relationship.
3.  Let $W_{0}=\$1,000$ be the initial wealth invested in each asset.  Compute the $1\%$ monthly Value-at-Risk values for each asset. (Hint: $q_{0.01}^{Z}=-2.326$).
4.  Continue with the above question, state in words what the $1\%$ Value-at-Risk numbers represent (i.e., explain what $1\%$ Value-at-Risk for a one month $\$1,000$ investment means).
5.  The normal distribution can be used to characterize the probability distribution of monthly simple returns or monthly continuously compounded returns. What are two problems with using the normal distribution for simple returns? Given these two problems, why might it be better to use the normal distribution for continuously compounded returns?
:::

::: {#exr-exercise10}
In this question, you will examine the chi-square and Student\textquoteright s
t distributions. A chi-square random variable with $n$ degrees of
freedom, denoted $X\sim\chi_{n}^{2},$ is defined as $X=Z_{1}^{2}+Z_{2}^{2}+\cdots+Z_{n}^{2}$
where $Z_{1},Z_{2},\ldots,Z_{n}$ are $iid\,N(0,1)$ random variables.
Notice that $X$ only takes positive values. A Student's t random
variable with $n$ degrees of freedom, denoted $t\sim t_{n}$, is
defined as $t=Z/\sqrt{X/n}$, where $Z\sim N(0,1)$, $X\sim\chi_{n}^{2}$
and $Z$ is independent of $X$. The Student's t distribution is similar
to the standard normal except that it has fatter tails.

1.  On the same graph, plot the probability curves of chi-squared distributed random variables with 1, 2, 5 and 10 degrees of freedom. Use different colors and line styles for each curve. Hint: In R the density of the chi-square distribution is computed using the function `dchisq()`.
2.  On the same graph, plot the probability curves of Student's t distributed random variables with 1, 2, 5 and 10 degrees of freedom. Also include the probability curve for the standard normal distribution. Use different colors and line styles for each curve. Hint: In R the density of the chi-square distribution is computed using the function `dt()`.
:::

::: {#exr-exercise11}
Consider the following joint distribution of $X$ and $Y$:

+---+---+---+---+
|X/Y|1  |2  |3  |
+===+===+===+===+
|1  |0.1|0.2|0  |
+---+---+---+---+
|2  |0.1|0  |0.2|
+---+---+---+---+
|3  |0  |0.1|0.3|
+---+---+---+---+

1.  Find the marginal distributions of $X$ and $Y$. Using these distributions, compute $E[X]$, $\mathrm{var}(X)$, $\mathrm{sd}(X)$, $E[Y]$, $\mathrm{var}(Y)$ and $\mathrm{sd}(Y)$.
2.  Compute $\mathrm{cov}(X,Y)$ and $\mathrm{cor}(X,Y)$. 
3.  Find the conditional distributions $\mathrm{Pr}(X|Y=y)$ for $y=1,2,3$ and the conditional distributions $\mathrm{Pr}(Y|X=x)$ for $x=1,2,3.$
4.  Using the conditional distributions $\mathrm{Pr}(X|Y=y)$ for $y=1,2,3$ and $\mathrm{Pr}(Y|X=x)$ for $x=1,2,3$ compute $E[X|Y=y]$ and $E[Y|X=x]$.
5.  Using the conditional distributions $\mathrm{Pr}(X|Y=y)$ for $y=1,2,3$ and $\mathrm{Pr}(Y|X=x)$ for $x=1,2,3$ compute $\mathrm{var}[X|Y=y]$ and $\mathrm{var}[Y|X=x]$.
6.  Are $X$ and $Y$ independent? Fully justify your answer.
:::

::: {#exr-exercise12}
Let $X$ and $Y$ be distributed bivariate normal with $\mu_{X}=0.05$,
$\mu_{Y}=0.025$, $\sigma_{X}=0.10$, and $\sigma_{Y}=0.05$.

1.  Using R package **mvtnorm** function `rmvnorm()`, simulate 100 observations from the bivariate normal distribution with $\rho_{XY}=0.9$. Using the `plot()` function create a scatterplot of the observations and comment on the direction and strength of the linear association. Using the function `pmvnorm()`, compute the joint probability
$Pr(X\leq,\,Y\leq0)$.
2.  Using R package **mvtnorm** function `rmvnorm()`, simulate 100 observations from the bivariate normal distribution with $\rho_{XY}=-0.9$. Using the `plot()` function create a scatterplot of the observations and comment on the direction and strength of the linear association. Using the function `pmvnorm()`, compute the joint probability $Pr(X\leq,\,Y\leq0)$.
3.  Using R package **mvtnorm** function `rmvnorm()`, simulate 100 observations from the bivariate normal distribution with $\rho_{XY}=0$. Using the `plot()` function create a scatterplot of the observations and comment on the direction and strength of the linear association. Using the function `pmvnorm()`, compute the joint probability $Pr(X\leq,\,Y\leq0)$.
:::


