# Matrix Algebra Review {#sec-Matrix-Algebra-Review}

Updated: September 3, 2026

Copyright © Eric Zivot 2026
 
This chapter reviews some basic matrix algebra concepts that we will
use throughout the book. Many of the formulas that involve multiple
variables that we derive in the book can be simplified using matrix
algebra. This aids in understanding general concepts. For example,
portfolio risk and return calculations for portfolios with more than
two assets is easy to express using matrix algebra. If returns are
jointly normally distributed, then this multivariate normal distribution
can be expressed concisely using matrix algebra. The first order conditions
involving the portfolio weights for mean-variance optimized portfolios
is a system of linear equations that can be set-up and solved using
matrix algebra. In addition, R is a matrix programming language. Pretty
much anything that you can represent in matrix algebra you can evaluate
in R using matrix calculations that have a syntax very much like the
matrix algebra representation. Lastly, many R calculations can be
efficiently evaluated if they are *vectorized* - that is, if they
operate on vectors of elements instead of looping over individual
elements. 


This chapter is organized as follows. Section @sec-Matrices-and-Vectors gives definitions
of matrices and vectors and shows how to represent these objects in
R. Section @sec-Basic-Matrix-Operations reviews basic matrix operations (addition, subtraction,
multiplication), and section @sec-Representing-Summation-Using-Matrix-Notation shows 
how common summation calculations can be represented using matrix notation. 
Section @sec-Systems-of-Linear-Equations covers systems
of linear equations and shows how the inverse of a matrix is related
to the solutions to a system of linear equations. Section @sec-Positive-Definite-Matrices 
discusses positive definite matrices and Cholesky decomposition. The use of
matrix algebra for representing multivariate probability distributions
is presented in section @sec-Multivariate-Probability-Distributions-Using-Matrix-Algebra, 
and portfolio calculations using matrix algebra is presented in section 
@sec-Portfolio-Math-Using-Matrix-Algebra. Section @sec-Derivatives-of-Simple-Matrix-Functions 
concludes with a discussion of derivatives of some simple matrix functions.

The R packages used in this chapter are **Matrix** and **mvtnorm**.
Make sure these packages are downloaded and installed prior to replicating
the R examples in this chapter.

```{r}
#| include: false
options(digits=3)
library(Matrix)
library(mvtnorm)
```


## Matrices and Vectors {#sec-Matrices-and-Vectors}

::: {#def-matrix .definition name="Matrix"}
A *matrix*\index{Matrices} is just an array of numbers. The *dimension*
of a matrix is determined by the number of its rows and columns. 
:::

For example, a matrix $\mathbf{A}$ with $n$ rows and $m$ columns is
illustrated below: 
$$
\underset{(n\times m)}{\mathbf{A}}=\left[\begin{array}{cccc}
a_{11} & a_{12} & \ldots & a_{1m}\\
a_{21} & a_{22} & \ldots & a_{2m}\\
\vdots & \vdots & \ddots & \vdots\\
a_{n1} & a_{n2} & \ldots & a_{nm}
\end{array}\right]
$$
where $a_{ij}$ denotes the $i^{th}$ row and $j^{th}$ column element
of $\mathbf{A}$. 

::: {#def-vector .definition name="Vector"}
A *vector*\index{Vectors} is simply a matrix with $1$ column. 
:::

For example,
$$
\underset{(n\times1)}{\mathbf{x}}=\left[\begin{array}{c}
x_{1}\\
x_{2}\\
\vdots\\
x_{n}
\end{array}\right]
$$
is an $n\times1$ vector with elements $x_{1},x_{2},\ldots,x_{n}$.
Vectors and matrices are often written in bold type (or underlined)
to distinguish them from scalars (single elements of vectors or matrices).

::: {#exm-matrix-creation-in-r .example name="Matrix creation in R"}
In R, matrix objects are created using the `matrix()` function.
For example, to create the $2\times3$ matrix: 
$$
\underset{(2\times3)}{\mathbf{A}}=\left[\begin{array}{ccc}
1 & 2 & 3\\
4 & 5 & 6
\end{array}\right]
$$
use,

```{r}
#| echo: true
matA = matrix(data=c(1,2,3,4,5,6),nrow=2,ncol=3,byrow=TRUE)
matA
class(matA)
```

The optional argument `byrow=TRUE` fills the matrix row-by-row.^[When specifying logical variables in R always spell out `TRUE` and
`FALSE` instead of using `T` and `F`. Upon startup
R defines the variables `T=TRUE` and `F=FALSE` so that
`T` and `F` can be used as substitutes for `TRUE`
and `FALSE`, respectively. However, this shortcut is not recommended
because the variables `T` and `F` could be reassigned
during subsequent programming.] The default is `byrow=FALSE` which fills the matrix 
column-by-column:

```{r}
#| echo: true
matrix(data=c(1,2,3,4,5,6),nrow=2,ncol=3)
```

Matrix objects (i.e., objects of class `"matrix"`) have
row and column dimension attributes which can be examined with the
`dim()` function:

```{r}
#| echo: true
dim(matA)
```

The rows and columns can be given names using:

```{r}
#| echo: true
dimnames(matA) = list(c("row1","row2"),c("col1","col2","col3"))
matA
```

or,

```{r}
#| echo: true
colnames(matA) = c("col1", "col2", "col3")
rownames(matA) = c("row1", "row2")
matA
```

The elements of a matrix can be extracted or subsetted as follows:

```{r}
#| echo: true
matA[1, 2]
matA["row1", "col1"]
matA[1, ]
matA[, 2]
```

To preserve the dimension attributes when subsetting, use the `drop=FALSE`
option:

```{r}
#| echo: true
matA[1, , drop=FALSE]
matA[, 2, drop=FALSE]
```

\hfill{}$\blacksquare$
:::

::: {#exm-creating-vectors-in-r .example name="Creating vectors in R"}
Vectors can be created in R using a variety of methods. The easiest
way is to use the combine function `c()`:

```{r}
#| echo: true
xvec = c(1,2,3)
xvec
xvec = 1:3
xvec
xvec = seq(from=1,to=3,by=1)
xvec
```

Vectors of numbers in R are of class `"numeric"` and do
not have a dimension attribute:

```{r}
#| echo: true
class(xvec)
dim(xvec)
```

The elements of a vector can be assigned names using the `names()`
function:

```{r}
#| echo: true
names(xvec) = c("x1", "x2", "x3")
xvec
```

To force a dimension attribute onto a vector, coerce it to a `"matrix"`
object using `as.matrix()`:

```{r}
#| echo: true
as.matrix(xvec)

```

\hfill{}$\blacksquare$
:::

::: {#def-transpose .definition name="Transpose"}
The *transpose*\index{Transpose} of an $n\times m$ matrix $\mathbf{A}$ is
a new matrix with the rows and columns of $\mathbf{A}$ interchanged,
and is denoted $\mathbf{A}^{\prime}$ or $\mathbf{A}^{\intercal}$.
:::

For example, 
$$
\begin{aligned}
\underset{(2\times3)}{\mathbf{A}} & =\left[\begin{array}{ccc}
1 & 2 & 3\\
4 & 5 & 6
\end{array}\right],~\underset{(3\times2)}{\mathbf{A}^{\prime}}=\left[\begin{array}{cc}
1 & 4\\
2 & 5\\
3 & 6
\end{array}\right]\\
\underset{(3\times1)}{\mathbf{x}} & =\left[\begin{array}{c}
1\\
2\\
3
\end{array}\right],~\underset{(1\times3)}{\mathbf{x}^{\prime}}=\left[\begin{array}{ccc}
1 & 2 & 3\end{array}\right].
\end{aligned}
$$

::: {#def-symmetric-matrix name="Symmetric matrix"}
A *symmetric* matrix $\mathbf{A}$ is such that $\mathbf{A}=\mathbf{A}^{\prime}$.
:::

Obviously, symmetry can only occur if $\mathbf{A}$ is a *square*
matrix; i.e., the number of rows of $\mathbf{A}$ is equal to the
number of columns. For example, consider the $2\times2$ matrix: 
$$
\mathbf{A}=\left[\begin{array}{cc}
1 & 2\\
2 & 1
\end{array}\right].
$$
Then, 
$$
\mathbf{A}^{\prime}=\mathbf{A}=\left[\begin{array}{cc}
1 & 2\\
2 & 1
\end{array}\right].
$$

::: {#exm-transpose-of-a-matrix-in-r name="Transpose of a matrix in R"}
To take the transpose of a matrix or vector, use the `t()`
function:

```{r}
#| echo: true
matA = matrix(data=c(1,2,3,4,5,6),nrow=2,ncol=3,byrow=TRUE)
matA
t(matA)
xvec = c(1,2,3)
t(xvec)
```

Notice that, when applied to a `"numeric"` vector with
$n$ elements, the `t()` function returns a `"matrix"`
object with dimension $1\times n$.

\hfill{}$\blacksquare$
:::

## Basic Matrix Operations {#sec-Basic-Matrix-Operations}

In this section we review the basic matrix operations of addition,
subtraction, scalar multiplication and multiplication.

### Addition and subtraction
\index{Matrix addition}

::: {#def-matrix-addition name="Matrix addition"}
Matrix addition and subtraction are element by element operations
and only apply to matrices of the same dimension. 
:::

For example, let,
$$
\mathbf{A}=\left[\begin{array}{cc}
4 & 9\\
2 & 1
\end{array}\right],~\mathbf{B}=\left[\begin{array}{cc}
2 & 0\\
0 & 7
\end{array}\right].
$$
Then, 
$$
\begin{aligned}
\mathbf{A+B} & =\left[\begin{array}{cc}
4 & 9\\
2 & 1
\end{array}\right]+\left[\begin{array}{cc}
2 & 0\\
0 & 7
\end{array}\right]=\left[\begin{array}{cc}
4+2 & 9+0\\
2+0 & 1+7
\end{array}\right]=\left[\begin{array}{cc}
6 & 9\\
2 & 8
\end{array}\right],\\
\mathbf{A-B} & =\left[\begin{array}{cc}
4 & 9\\
2 & 1
\end{array}\right]-\left[\begin{array}{cc}
2 & 0\\
0 & 7
\end{array}\right]=\left[\begin{array}{cc}
4-2 & 9-0\\
2-0 & 1-7
\end{array}\right]=\left[\begin{array}{cc}
2 & 9\\
2 & -6
\end{array}\right].
\end{aligned}
$$

::: {#exm-matrix-addition-and-subtraction-in-r name="Matrix addition and subtraction in R"}
Matrix addition and subtraction is straightforward in R:

```{r}
#| echo: true
matA = matrix(c(4,9,2,1),2,2,byrow=TRUE)
matB = matrix(c(2,0,0,7),2,2,byrow=TRUE)
matA
matB
# matrix addition
matC = matA + matB
matC
# matrix subtraction
matD = matA - matB
matD
```

\hfill{}$\blacksquare$
:::

### Scalar multiplication
\index{Scalar multiplication}

Here we refer to the multiplication of a matrix by a scalar number.
This is also an element-by-element operation. For example, let $c=2$
and, 
$$
\mathbf{A}=\left[\begin{array}{cc}
3 & -1\\
0 & 5
\end{array}\right].
$$
Then, 
$$
c\cdot\mathbf{A}=\left[\begin{array}{cc}
2\cdot3 & 2\cdot(-1)\\
2\cdot(0) & 2\cdot5
\end{array}\right]=\left[\begin{array}{cc}
6 & -2\\
0 & 10
\end{array}\right].
$$

::: {#exm-scalar-multiplication-in-r .example name="Scalar multiplication in R"}

```{r}
#| echo: true
matA = matrix(c(3,-1,0,5),2,2,byrow=TRUE)
matC = 2*matA
matC
```

\hfill{}$\blacksquare$
:::

### Matrix multiplication
\index{Matrix multiplication}

Matrix multiplication only applies to *conformable* matrices\index{Conformable matrices}.

::: {#def-conformable-matrices name="Conformable matrices"}
$\mathbf{A}$ and $\mathbf{B}$ are conformable matrices if the number
of columns in $\mathbf{A}$ is equal to the number of rows in $\mathbf{B}.$
:::

For example, if $\mathbf{A}$ is $n\times m$ and $\mathbf{B}$ is
$m\times p$ then $\mathbf{A}$ and $\mathbf{B}$ are conformable
and the matrix product of $\mathbf{A}$ and $\mathbf{B}$ has dimension
$n\times p$. The mechanics of matrix multiplication is best explained
by example. Let, 
$$
\underset{(2\times2)}{\mathbf{A}}=\left[\begin{array}{cc}
1 & 2\\
3 & 4
\end{array}\right]\textrm{ and}\underset{(2\times3)}{\mathbf{B}}=\left[\begin{array}{ccc}
1 & 2 & 1\\
3 & 4 & 2
\end{array}\right].
$$
Then, 
$$
\begin{aligned}
\underset{(2\times2)}{\mathbf{A}}\cdot\underset{(2\times3)}{\mathbf{B}} & =\left[\begin{array}{cc}
1 & 2\\
3 & 4
\end{array}\right]\cdot\left[\begin{array}{ccc}
1 & 2 & 1\\
3 & 4 & 2
\end{array}\right]\\
 & =\left[\begin{array}{ccc}
1\cdot1+2\cdot3 & 1\cdot2+2\cdot4 & 1\cdot1+2\cdot2\\
3\cdot1+4\cdot3 & 3\cdot2+4\cdot4 & 3\cdot1+4\cdot2
\end{array}\right]\\
 & =\left[\begin{array}{ccc}
7 & 10 & 5\\
15 & 22 & 11
\end{array}\right]=\underset{(2\times3)}{\mathbf{C}}
\end{aligned}
$$
The resulting matrix $\mathbf{C}$ has 2 rows and 3 columns. The $(1,1)$
element of $\mathbf{C}$ is the *dot product*\index{Dot product} of the first row
of $\mathbf{A}$ with the first column of $\mathbf{B}$: 
$$
\left(\begin{array}{cc}
1 & 2\end{array}\right)\cdot\left(\begin{array}{c}
1\\
3
\end{array}\right)=1\cdot1+2\cdot3=7.
$$
The $(1,2)$ element of $\mathbf{C}$ is the dot product of
the first row of $\mathbf{A}$ with the second column of $\mathbf{B}$:
$$
\left(\begin{array}{cc}
1 & 2\end{array}\right)\cdot\left(\begin{array}{c}
2\\
4
\end{array}\right)=1\cdot2+2\cdot4=10.
$$
In general, the $(i,j)$ element of $\mathbf{C}$ is the dot product
of the *i*th row of $\mathbf{A}$ with the *j*th column
of $\mathbf{B}$. If $\mathbf{A}$ is $n\times m$ and $\mathbf{B}$
is $m\times p$ then $\mathbf{C=A\cdot B}$ is $n\times p$. 

As another example, let, 
$$
\underset{(2\times2)}{\mathbf{A}}=\left[\begin{array}{cc}
1 & 2\\
3 & 4
\end{array}\right]\textrm{ and}\underset{(2\times1)}{\mathbf{B}}=\left[\begin{array}{c}
2\\
6
\end{array}\right].
$$
Then, 
$$
\begin{aligned}
\underset{(2\times2)}{\mathbf{A}}\cdot\underset{(2\times1)}{\mathbf{B}} & =\left[\begin{array}{cc}
1 & 2\\
3 & 4
\end{array}\right]\cdot\left[\begin{array}{c}
2\\
6
\end{array}\right]\\
 & =\left[\begin{array}{c}
1\cdot2+2\cdot6\\
3\cdot2+4\cdot6
\end{array}\right]\\
 & =\left[\begin{array}{c}
14\\
30
\end{array}\right].
\end{aligned}
$$
As a final example, let, 
$$
\mathbf{x}=\left[\begin{array}{c}
1\\
2\\
3
\end{array}\right],~\mathbf{y}=\left[\begin{array}{c}
4\\
5\\
6
\end{array}\right].
$$
Then 
$$
\mathbf{x}^{\prime}\mathbf{y}=\left[\begin{array}{ccc}
1 & 2 & 3\end{array}\right]\cdot\left[\begin{array}{c}
4\\
5\\
6
\end{array}\right]=1\cdot4+2\cdot5+3\cdot6=32
$$

::: {#exm-matrix-multiplication-in-r .example name="Matrix multiplication in R"}
In R, matrix multiplication is performed with the `% * %`
operator. For example:

```{r}
#| echo: true
matA = matrix(1:4,2,2,byrow=TRUE)
matB = matrix(c(1,2,1,3,4,2),2,3,byrow=TRUE)
matA
matB
dim(matA)
dim(matB)
matC = matA%*%matB
matC
```

The next example shows that matrix multiplication in R also works
on numeric vectors:

```{r}
#| echo: true
matA = matrix(c(1,2,3,4), 2, 2, byrow=TRUE)
vecB = c(2,6)
matA%*%vecB
vecX = c(1,2,3)
vecY = c(4,5,6)
t(vecX)%*%vecY
crossprod(vecX, vecY)
```

\hfill{}$\blacksquare$
:::

#### Miscellaneous properties

Matrix multiplication satisfies the distributive property: Let $\mathbf{A},\,\mathbf{B}\,$ and
$\mathbf{C}$ be conformable matrices. Then
$$
\mathbf{A}(\mathbf{B}+\mathbf{C})=\mathbf{AB}+\mathbf{AC}.
$$ {#eq-distributive-property-of-matrix-multiplication}
Also, the transpose of the product of two matrices is the product
of the transposes in opposite order:
$$
(\mathbf{AB})^{\prime}=\mathbf{B}^{\prime}\mathbf{A}^{\prime}.
$$ {#eq-transpose-of-a-product-of-matrices}


### The identity matrix
\index{Identity matrix}

The *identity matrix* plays a similar role as the number $1.$ Multiplying
any number by $1$ gives back that number. In matrix algebra, pre-multiplying
or post-multiplying a matrix $\mathbf{A}$ by a conformable identity
matrix $\mathbf{I}$ gives back the matrix $\mathbf{A}$. To illustrate, let 
$$
\mathbf{I}_{2}=\left[\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right]
$$
denote the 2-dimensional identity matrix and let 
$$
\mathbf{A}=\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right]
$$
denote an arbitrary $2\times2$ matrix. Then, 
$$
\begin{aligned}
\mathbf{I}_{2}\mathbf{\cdot A} & =\left[\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right]\cdot\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right]\\
 & =\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right]=\mathbf{A},
\end{aligned}
$$
and,
$$
\begin{aligned}
\mathbf{A\cdot I}_{2} & =\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right]\cdot\left[\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right]\\
 & =\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right]=\mathbf{A}.
\end{aligned}
$$

::: {#exm-the-identity-matrix-in-r .example name="The identity matrix in R"}
Use the `diag()` function to create an identity matrix:

```{r}
#| echo: true
matI = diag(2)
matI
matA = matrix(c(1,2,3,4), 2, 2, byrow=TRUE)
matI%*%matA
matA%*%matI
```

\hfill{}$\blacksquare$
:::

### Diagonal, lower triangular and upper triangular matrices

Consider an $n\times n$ matrix $\mathbf{A}$
$$
\mathbf{A}=\left[\begin{array}{ccccc}
d_{1} & u_{12} & u_{13} & \cdots & u_{1n}\\
l_{21} & d_{2} & u_{23} & \cdots & u_{2n}\\
l_{31} & l_{32} & d_{3} & \cdots & u_{3n}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
l_{n1} & l_{n2} & l_{n3} & \cdots & d_{n}
\end{array}\right].
$$
The main diagonal consists of the $n$ elements $\{d_{1},d_{2}\ldots d_{n}\}$.
The lower triangle consists of the $n(n-1)/2$ elements $\{l_{21},l_{31},l_{32},\ldots,l_{n(n-1)}\}$
below the main diagonal, and the upper triangle consists of the $n(n-1)/2$
elements $\{u_{12},u_{13},u_{23},\ldots,u_{(n-1)n}\}$ above the main diagonal.

An $n-$dimensional *diagonal matrix*\index{Diagonal matrix} $\mathbf{D}$ is an $n\times n$
square matrix with an $n\times1$ vector $\mathbf{d}=(d_{1},\ldots,d_{n})^{\prime}$
along the main diagonal and zeros elsewhere:
$$
\mathbf{D}= \mathrm{diag}(\mathbf{d}) = \left[\begin{array}{ccccc}
d_{1} & 0 & 0 & \cdots & 0\\
0 & d_{2} & 0 & \cdots & 0\\
0 & 0 & d_{3} & \cdots & 0\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
0 & 0 & 0 & \cdots & d_{n}
\end{array}\right].
$$
An $n\times n$ *lower triangular matrix*\index{Lower triangular matrix} $\mathbf{L}$ has all
values above the main diagonal equal to zero:
$$
\mathbf{L}=\left[\begin{array}{ccccc}
d_{1} & 0 & 0 & \cdots & 0\\
l_{21} & d_{2} & 0 & \cdots & 0\\
l_{31} & l_{32} & d_{3} & \cdots & 0\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
l_{n1} & l_{n2} & l_{n3} & \cdots & d_{n}
\end{array}\right].
$$
An $n\times n$ *upper triangular matrix*\index{Upper triangular matrix} $\mathbf{U}$ has all
values below the main diagonal equal to zero:
$$
\mathbf{U}=\left[\begin{array}{ccccc}
d_{1} & u_{12} & u_{13} & \cdots & u_{1n}\\
0 & d_{2} & u_{23} & \cdots & u_{2n}\\
0 & 0 & d_{3} & \cdots & u_{3n}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
0 & 0 & 0 & \cdots & d_{n}
\end{array}\right].
$$

::: {#exm-creating-a-diagonal-matrix-in-r .example name="Creating a diagonal matrix in R"}
Diagonal matrices can be created with the `diag()` function.
To create a $3\times3$ diagonal matrix with $\mbox{d}=(1,2,3)$ along
the main diagonal, use

```{r}
#| echo: true
matD = diag(1:3)
matD
```

\hfill{}$\blacksquare$
:::

::: {#exm-extracting-lower-and-upper-triangular-elements-of-a-square-matrix .example name="Extracting lower and upper triangular elements of a square matrix"}
The R functions `lower.tri()` and `upper.tri()` can
be used to extract the lower and upper triangular parts of a square
matrix:

```{r}
#| echo: true
matA = matrix(c(1,2,3,4,5,6,7,8,9), 3, 3)
matA
lower.tri(matA)
upper.tri(matA)
matA[lower.tri(matA)]
matA[upper.tri(matA)]
```

\hfill{}$\blacksquare$
:::

## Representing Summation Using Matrix Notation {#sec-Representing-Summation-Using-Matrix-Notation}

Consider the sum: 
$$
\sum_{k=1}^{n}x_{k}=x_{1}+\cdots+x_{n}.
$$
Let $\mathbf{x}=(x_{1},\ldots,x_{n})^{\prime}$ be an $n\times1$
vector and $\mathbf{1}=(1,\ldots,1)^{\prime}$ be an $n\times1$ vector
of ones. Then, 
$$
\mathbf{x}^{\prime}\mathbf{1}=\left[\begin{array}{ccc}
x_{1} & \ldots & x_{n}\end{array}\right]\cdot\left[\begin{array}{c}
1\\
\vdots\\
1
\end{array}\right]=x_{1}+\cdots+x_{n}=\sum_{k=1}^{n}x_{k},
$$
and, 
$$
\mathbf{1}^{\prime}\mathbf{x}=\left[\begin{array}{ccc}
1 & \ldots & 1\end{array}\right]\cdot\left[\begin{array}{c}
x_{1}\\
\vdots\\
x_{n}
\end{array}\right]=x_{1}+\cdots+x_{n}=\sum_{k=1}^{n}x_{k}.
$$
Next, consider the sum of squared $x$ values, 
$$
\sum_{k=1}^{n}x_{k}^{2}=x_{1}^{2}+\cdots+x_{n}^{2}.
$$
This sum can be conveniently represented as, 
$$
\mathbf{x}^{\prime}\mathbf{x}=\left[\begin{array}{ccc}
x_{1} & \ldots & x_{n}\end{array}\right]\cdot\left[\begin{array}{c}
x_{1}\\
\vdots\\
x_{n}
\end{array}\right]=x_{1}^{2}+\cdots+x_{n}^{2}=\sum_{k=1}^{n}x_{k}^{2}.
$$
Last, consider the sum of cross products, 
$$
\sum_{k=1}^{n}x_{k}y_{k}=x_{1}y_{1}+\cdots x_{n}y_{n}.
$$
This sum can be compactly represented by, 
$$
\mathbf{x}^{\prime}\mathbf{y}=\left[\begin{array}{ccc}
x_{1} & \ldots & x_{n}:\end{array}\right]\cdot\left[\begin{array}{c}
y_{1}\\
\vdots\\
y_{n}
\end{array}\right]=x_{1}y_{1}+\cdots x_{n}y_{n}=\sum_{k=1}^{n}x_{k}y_{k}.
$$
Note that $\mathbf{x}^{\prime}\mathbf{y=y}^{\prime}\mathbf{x}$.

::: {#exm-computing-sums-in-r .example name="Computing sums in R"}
In R, summing the elements in a vector can be done using matrix algebra.

```{r}
#| echo: true
# create vector of 1's and a vector of x
onevec = rep(1,3)
onevec
xvec = c(1,2,3)
# sum elements in x
t(xvec)%*%onevec
```

The functions `crossprod()` and `sum()` are generally
computationally more efficient:

```{r}
#| echo: true
crossprod(xvec,onevec)
sum(xvec)
```

Sums of squares are best computed using:

```{r}
#| echo: true
crossprod(xvec)
sum(xvec^2)
```

The dot-product two vectors can be conveniently computed using the
`crossprod()` function:

```{r}
#| echo: true
yvec = 4:6
xvec
yvec
crossprod(xvec,yvec)
crossprod(yvec,xvec)
```

\hfill{}$\blacksquare$
:::

## Systems of Linear Equations {#sec-Systems-of-Linear-Equations}

Consider the system of two linear equations: 
$$
x+y=1,
$$ {#eq-eq1}
$$
2x-y =1.
$$ {#eq-eq2}

As shown in Figure @fig-MatrixReview2eqns, equations @eq-eq1
and @eq-eq2 represent two straight lines which intersect at
the point $x=2/3$ and $y=1/3$. This point of intersection is determined
by solving for the values of $x$ and $y$ such that $x+y=2x-y$^[Soving for $x$ gives $x=2y$. Substituting this value into the equation
$x+y=1$ gives $2y+y=1$ and solving for $y$ gives $y=1/3$. Solving
for $x$ then gives $x=2/3$.]


```{r}
#| label: fig-MatrixReview2eqns
#| echo: false
#| fig-cap: System of two linear equations.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
curve(1-x, 0, 1, lwd=2, ylab="y")
abline(a=-1, b=2, lwd=2)
segments(x0=2/3, y0=0, x1 = 2/3, y1 = 1/3, col="red", lty = "dashed", lwd = 2)
segments(x0=0, y0=1/3, x1 = 2/3, y1 = 1/3, col="red", lty = "dashed", lwd = 2)
title("x+y=1, 2x-y=1")
```

The two linear equations can be written in matrix form as: 
$$
\left[\begin{array}{cc}
1 & 1\\
2 & -1
\end{array}\right]\left[\begin{array}{c}
x\\
y
\end{array}\right]=\left[\begin{array}{c}
1\\
1
\end{array}\right],
$$
or, 
$$
\mathbf{A\cdot z=b,}
$$
where, 
$$
\mathbf{A}=\left[\begin{array}{cc}
1 & 1\\
2 & -1
\end{array}\right],~\mathbf{z}=\left[\begin{array}{c}
x\\
y
\end{array}\right]\textrm{ and }\mathbf{b}=\left[\begin{array}{c}
1\\
1
\end{array}\right].
$$

If there was a $(2\times2)$ matrix $\mathbf{B},$ with elements $b_{ij}$,
such that $\mathbf{B\cdot A=I}_{2},$ where $\mathbf{I}_{2}$ is the
$(2\times2)$ identity matrix, then we could *solve* for the
elements in $\mathbf{z}$ as follows. In the equation $\mathbf{A\cdot z=b}$,
pre-multiply both sides by $\mathbf{B}$ to give: 
$$
\begin{aligned}
\mathbf{B\cdot A\cdot z} & =\mathbf{B\cdot b}\\
 & \Longrightarrow\mathbf{I\cdot z=B\cdot b}\\
 & \Longrightarrow\mathbf{z=B\cdot b},
\end{aligned}
$$
or 
$$
\left[\begin{array}{c}
x\\
y
\end{array}\right]=\left[\begin{array}{cc}
b_{11} & b_{12}\\
b_{21} & b_{22}
\end{array}\right]\left[\begin{array}{c}
1\\
1
\end{array}\right]=\left[\begin{array}{c}
b_{11}\cdot1+b_{12}\cdot1\\
b_{21}\cdot1+b_{22}\cdot1
\end{array}\right].
$$
If such a matrix $\mathbf{B}$ exists it is called the *inverse*
of $\mathbf{A}$ and is denoted $\mathbf{A}^{-1}$. Intuitively, the
inverse matrix $\mathbf{A}^{-1}$ plays a similar role as the inverse
of a number. Suppose $a$ is a number; e.g., $a=2$. Then we know
that $(1/a)\cdot a=a^{-1}a=1$. Similarly, in matrix algebra $\mathbf{A}^{-1}\mathbf{A=I}_{2}$
where $\mathbf{I}_{2}$ is the identity matrix. Now, consider solving
the equation $a\cdot x=b$. By simple division we have that $x=(1/a)\cdot b=a^{-1}\cdot b$.
Similarly, in matrix algebra, if we want to solve the system of linear
equations $\mathbf{Ax=b}$ we pre-multiply by $\mathbf{A}^{-1}$ and
get the solution $\mathbf{x=A}^{-1}\mathbf{b}$. 


Using $\mathbf{B=A}^{-1}$, we may express the solution for $\mathbf{z}$
as: 
$$
\mathbf{z=A}^{-1}\mathbf{b}.
$$
As long as we can determine the elements in $\mathbf{A}^{-1}$ then
we can solve for the values of $x$ and $y$ in the vector $\mathbf{z}$.
The system of linear equations has a solution as long as the two lines
intersect, so we can determine the elements in $\mathbf{A}^{-1}$
provided the two lines are not parallel. If the two lines are parallel,
then one of the equations is a multiple of the other. In this case,
we say that $\mathbf{A}$ is *not invertible*. \index{Invertible matrix}


There are general numerical algorithms for finding the elements of
$\mathbf{A}^{-1}$ (e.g., Gaussian elimination) and R has these algorithms available. However,
if $\mathbf{A}$ is a $(2\times2)$ matrix then there is a simple
formula for $\mathbf{A}^{-1}$. Let $\mathbf{A}$ be a $(2\times2)$
matrix such that: 
$$
\mathbf{A}=\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right].
$$
Then, 
$$
\mathbf{A}^{-1}=\frac{1}{\det(\mathbf{A})}\left[\begin{array}{cc}
a_{22} & -a_{12}\\
-a_{21} & a_{11}
\end{array}\right],
$$ {#eq-inverse-of-a-2x2-matrix}
where $\det(\mathbf{A})=a_{11}a_{22}-a_{21}a_{12}$ denotes the determinant\index{Matrix Determinant}
of $\mathbf{A}$ and is assumed to be not equal to zero. By brute
force matrix multiplication we can verify this formula: 
$$
\begin{aligned}
\mathbf{A}^{-1}\mathbf{A} & =\frac{1}{a_{11}a_{22}-a_{21}a_{12}}\left[\begin{array}{cc}
a_{22} & -a_{12}\\
-a_{21} & a_{11}
\end{array}\right]\left[\begin{array}{cc}
a_{11} & a_{12}\\
a_{21} & a_{22}
\end{array}\right]\\
 & =\frac{1}{a_{11}a_{22}-a_{21}a_{12}}\left[\begin{array}{cc}
a_{22}a_{11}-a_{12}a_{21} & a_{22}a_{12}-a_{12}a_{22}\\
-a_{21}a_{11}+a_{11}a_{21} & -a_{21}a_{12}+a_{11}a_{22}
\end{array}\right]\\
 & =\frac{1}{a_{11}a_{22}-a_{21}a_{12}}\left[\begin{array}{cc}
a_{22}a_{11}-a_{12}a_{21} & 0\\
0 & -a_{21}a_{12}+a_{11}a_{22}
\end{array}\right]\\
 & =\left[\begin{array}{cc}
\dfrac{a_{22}a_{11}-a_{12}a_{21}}{a_{11}a_{22}-a_{21}a_{12}} & 0\\
0 & \dfrac{-a_{21}a_{12}+a_{11}a_{22}}{a_{11}a_{22}-a_{21}a_{12}}
\end{array}\right]\\
 & =\left[\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right].
\end{aligned}
$$
Let's apply the above rule to find the inverse of $\mathbf{A}$ in
our example linear system @eq-eq1 and  @eq-eq2: 
$$
\mathbf{A}^{-1}=\frac{1}{-1-2}\left[\begin{array}{cc}
-1 & -1\\
-2 & 1
\end{array}\right]=\left[\begin{array}{cc}
\frac{1}{3} & \frac{1}{3}\\
\frac{2}{3} & \frac{-1}{3}
\end{array}\right].
$$
Notice that, 
$$
\mathbf{A}^{-1}\mathbf{A}=\left[\begin{array}{cc}
\frac{1}{3} & \frac{1}{3}\\
\frac{2}{3} & \frac{-1}{3}
\end{array}\right]\left[\begin{array}{cc}
1 & 1\\
2 & -1
\end{array}\right]=\left[\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right].
$$
Our solution for $\mathbf{z}$ is then, 
$$
\begin{aligned}
\mathbf{z} & =\mathbf{A}^{-1}\mathbf{b}\\
 & =\left[\begin{array}{cc}
\frac{1}{3} & \frac{1}{3}\\
\frac{2}{3} & \frac{-1}{3}
\end{array}\right]\left[\begin{array}{c}
1\\
1
\end{array}\right]\\
 & =\left[\begin{array}{c}
\frac{2}{3}\\
\frac{1}{3}
\end{array}\right]=\left[\begin{array}{c}
x\\
y
\end{array}\right],
\end{aligned}
$$
so that $x=2/3$ and $y=1/3$.

::: {#exm-solving-systems-of-linear-equations-in-r .example name="Solving systems of linear equations in R"}
In R, the `solve()` function is used to compute the inverse
of a matrix and solve a system of linear equations. The linear system
$x+y=1$ and $2x-y=1$ can be represented using:

```{r}
#| echo: true
matA = matrix(c(1,1,2,-1), 2, 2, byrow=TRUE)
vecB = c(1,1)
```

First we solve for $\mathbf{A}^{-1}$:^[Notice that the calculations in R do not show $\mathbf{A}^{-1}\mathbf{A}=\mathbf{I}$
exactly. The (1,2) element of $\mathbf{A}^{-1}\mathbf{A}$ is -5.552e-17,
which for all practical purposes is zero. However, due to the limitations
of machine calculations the result is not exactly zero.]

```{r}
#| echo: true
matA.inv = solve(matA)
matA.inv
matA.inv%*%matA
matA%*%matA.inv
```

Then we solve the system $\mathbf{z}=\mathbf{A}^{-1}\mathbf{b}$:

```{r}
#| echo: true
z = matA.inv%*%vecB
z
```

\hfill{}$\blacksquare$
:::

In general, if we have $n$ linear equations in $n$ unknown variables
we may write the system of equations as 
$$
\begin{aligned}
a_{11}x_{1}+a_{12}x_{2}+\cdots+a_{1n}x_{n} & =b_{1}\\
a_{21}x_{1}+a_{22}x_{2}+\cdots+a_{2n}x_{n} & =b_{2}\\
\vdots & =\vdots\\
a_{n1}x_{1}+a_{n2}x_{2}+\cdots+a_{nn}x_{n} & =b_{n}
\end{aligned}
$$
which we may then express in matrix form as 
$$
\left[\begin{array}{cccc}
a_{11} & a_{12} & \cdots & a_{1n}\\
a_{21} & a_{22} & \cdots & a_{2n}\\
\vdots & \vdots & \ddots & \vdots\\
a_{n1} & a_{n2} & \cdots & a_{nn}
\end{array}\right]\left[\begin{array}{c}
x_{1}\\
x_{2}\\
\vdots\\
x_{n}
\end{array}\right]=\left[\begin{array}{c}
b_{1}\\
b_{2}\\
\vdots\\
b_{n}
\end{array}\right]
$$
or 
$$
\underset{(n\times n)}{\mathbf{A}}\cdot\underset{(n\times1)}{\mathbf{x}}=\underset{(n\times1)}{\mathbf{b}}.
$$
The solution to the system of equations is given by: 
$$
\mathbf{x=A}^{-1}\mathbf{b},
$$
where $\mathbf{A}^{-1}\mathbf{A}=\mathbf{I}_n$ and $\mathbf{I}_{n}$ is the $(n\times n)$
identity matrix. If the number of equations is greater than two, then
we generally use numerical algorithms to find the elements in $\mathbf{A}^{-1}$.

### Rank of a matrix
\index{Matrix Rank}

::: {#def-linear-independence-of-vectors name="Linear independence of vectors"}
Consider an $n\times m$ matrix $\mathbf{A}$ with $m$ columns $\mathbf{a}_{1},\mathbf{a}_{2},\ldots,\mathbf{a}_{m}$
where each column is an $n\times1$ vector and $n$ rows where each
row is an $1\times m$ row vector:
$$
\underset{(n\times m)}{\mathbf{A}}=\left[\begin{array}{cccc}
a_{11} & a_{12} & \ldots & a_{1m}\\
a_{21} & a_{22} & \ldots & a_{2m}\\
\vdots & \vdots & \ddots & \vdots\\
a_{n1} & a_{n2} & \ldots & a_{nm}
\end{array}\right]=\left[\begin{array}{cccc}
\mathbf{a}_{1} & \mathbf{a}_{2} & \ldots & \mathbf{a}_{m}\end{array}\right].
$$
Two vectors $\mathbf{a}_{1}$ and $\mathbf{a}_{2}$ are linearly independent
if $c_{1}\mathbf{a}_{1}+c_{2}\mathbf{a}_{2}=0$ implies that $c_{1}=c_{2}=0$.
A set of vectors $\mathbf{a}_{1},\mathbf{a}_{2},\ldots,\mathbf{a}_{m}$
are linearly independent if $c_{1}\mathbf{a}_{1}+c_{2}\mathbf{a}_{2}+\cdots+c_{m}\mathbf{a}_{m}=0$
implies that $c_{1}=c_{2}=\cdots=c_{m}=0$. That is, no vector $\mathbf{a}_{i}$
can be expressed as a non-trivial linear combination of the other
vectors. 
::: 

::: {#def-rank-of-a-matrix name="Rank of a matrix"}
The *column rank*\index{Column rank} of the $n\times m$ matrix $\mathbf{A}$, denoted
$\mathrm{rank}(\mathbf{A})$, is equal to the maximum number of linearly
independent columns. The row rank of a matrix is equal to the maximum
number of linearly independent rows, and is given by $\mathrm{rank}(\mathbf{A}^{\prime})$.
:::

It turns out that the column rank and row rank of a matrix are the
same. Hence, $\mathrm{rank}(\mathbf{A})$$=\mathrm{rank}(\mathbf{A}^{\prime})\leq\mathrm{min}(m,n)$.
If $\mathrm{rank}(\mathbf{A})=m$ then $\mathbf{A}$ is called *full column rank*, and if $\mathrm{rank}(\mathbf{A})=n$ then $\mathbf{A}$
is called *full row rank*. If $\mathbf{A}$ is not full rank
then it is called *reduced rank*. 

::: {#exm-determining-the-rank-of-a-matrix-in-r .example name="Determining the rank of a matrix in R"}
Consider the $2\times3$ matrix 
$$
\mathbf{A}=\left(\begin{array}{ccc}
1 & 3 & 5\\
2 & 4 & 6
\end{array}\right)
$$
Here, $\mathrm{rank}(\mathbf{A})\leq min(3,2)=2$, the number of rows.
Now, $\mathrm{rank}(\mathbf{A})=2$ since the rows of $\mathbf{A}$
are linearly independent. In R, the rank of a matrix can be found
using the **Matrix** function `rankMatrix()`:

```{r}
#| echo: true

library(Matrix)
Amat = matrix(c(1,3,5,2,4,6), 2, 3, byrow=TRUE)
as.numeric(rankMatrix(Amat))
```

\hfill{}$\blacksquare$
:::

The rank of an $n\times n$ square matrix $\mathbf{A}$ is directly
related to its invertibility. If $\mathrm{rank}(\mathbf{A})=n$ then
$\mathbf{A}^{-1}$ exists. This result makes sense. Since $\mathbf{A}^{-1}$
is used to solve a system of $n$ linear equations in $n$ unknowns,
it will have a unique solution as long as no equation in the system
can be written as a linear combination of the other equations.

### Partitioned matrices and partitioned inverses
\index{Partitioned matrices}
Consider a general $n\times m$ matrix $\mathbf{A}$:
$$
\underset{(n\times m)}{\mathbf{A}}=\left[\begin{array}{cccc}
a_{11} & a_{12} & \ldots & a_{1m}\\
a_{21} & a_{22} & \ldots & a_{2m}\\
\vdots & \vdots & \ddots & \vdots\\
a_{n1} & a_{n2} & \ldots & a_{nm}
\end{array}\right].
$$
In some situations we might want to partition $\mathbf{A}$ into sub-matrices
containing sub-blocks of the elements of $\mathbf{A}$: 
$$
\mathbf{A}=\left[\begin{array}{cc}
\mathbf{A}_{11} & \mathbf{A}_{12}\\
\mathbf{A}_{21} & \mathbf{A}_{22}
\end{array}\right],
$$
where the sub-matrices $\mathbf{A}_{11}$, $\mathbf{A}_{12}$, $\mathbf{A}_{21}$,
and $\mathbf{A}_{22}$ contain the appropriate sub-elements of $\mathbf{A}$.
For example, consider
$$
\mathbf{A}=\left[\begin{array}{cc|cc}
1 & 2 & 3 & 4\\
5 & 6 & 7 & 8\\
\hline 9 & 10 & 11 & 12\\
13 & 14 & 15 & 16
\end{array}\right]=\left[\begin{array}{cc}
\mathbf{A}_{11} & \mathbf{A}_{12}\\
\mathbf{A}_{21} & \mathbf{A}_{22}
\end{array}\right],
$$
where 
\begin{eqnarray*}
\mathbf{A}_{11} & = & \left[\begin{array}{cc}
1 & 2\\
5 & 6
\end{array}\right],\,\mathbf{A}_{12}=\left[\begin{array}{cc}
3 & 4\\
7 & 8
\end{array}\right],\\
\mathbf{A}_{21} & = & \left[\begin{array}{cc}
9 & 10\\
13 & 14
\end{array}\right],\,\mathbf{A}_{22}=\left[\begin{array}{cc}
11 & 12\\
15 & 16
\end{array}\right].
\end{eqnarray*}

::: {#exm-combining-sub-matrices-in-r .example name="Combining sub-matrices in R"}
In R sub-matrices can be combined column-wise using `cbind()`,
and row-wise using `rbind()`. For example, to create the matrix
$\mathbf{A}$ from the sub-matrices use

```{r}
#| echo: true

A11mat = matrix(c(1,2,5,6), 2, 2, byrow=TRUE) 
A12mat = matrix(c(3,4,7,8), 2, 2, byrow=TRUE) 
A21mat = matrix(c(9,10,13,14), 2, 2, byrow=TRUE) 
A22mat = matrix(c(11,12,15,16), 2, 2, byrow=TRUE) 
Amat = rbind(cbind(A11mat, A12mat), cbind(A21mat, A22mat)) 
Amat
```

\hfill{}$\blacksquare$
:::

The basic matrix operations work on sub-matrices in the obvious way.
To illustrate, consider another matrix $\mathbf{B}$ that is conformable
with $\mathbf{A}$ and partitioned in the same way:
$$
\mathbf{B}=\left[\begin{array}{cc}
\mathbf{B}_{11} & \mathbf{B}_{12}\\
\mathbf{B}_{21} & \mathbf{B}_{22}
\end{array}\right].
$$
Then
$$
\mathbf{A}+\mathbf{B}=\left[\begin{array}{cc}
\mathbf{A}_{11} & \mathbf{A}_{12}\\
\mathbf{A}_{21} & \mathbf{A}_{22}
\end{array}\right]+\left[\begin{array}{cc}
\mathbf{B}_{11} & \mathbf{B}_{12}\\
\mathbf{B}_{21} & \mathbf{B}_{22}
\end{array}\right]=\left[\begin{array}{cc}
\mathbf{A}_{11}+\mathbf{B}_{11} & \mathbf{A}_{12}+\mathbf{B}_{12}\\
\mathbf{A}_{21}+\mathbf{B}_{21} & \mathbf{A}_{22}+\mathbf{B}_{22}
\end{array}\right].
$$
If all of the sub-matrices of $\mathbf{A}$ and $\mathbf{B}$ are
comformable then
$$
\mathbf{AB}=\left[\begin{array}{cc}
\mathbf{A}_{11} & \mathbf{A}_{12}\\
\mathbf{A}_{21} & \mathbf{A}_{22}
\end{array}\right]\left[\begin{array}{cc}
\mathbf{B}_{11} & \mathbf{B}_{12}\\
\mathbf{B}_{21} & \mathbf{B}_{22}
\end{array}\right]=\left[\begin{array}{cc}
\mathbf{A}_{11}\mathbf{B}_{11}+\mathbf{A}_{12}\mathbf{B}_{21} & \mathbf{A}_{11}\mathbf{B}_{12}+\mathbf{A}_{12}\mathbf{B}_{22}\\
\mathbf{A}_{21}\mathbf{B}_{11}+\mathbf{A}_{22}\mathbf{B}_{21} & \mathbf{A}_{21}\mathbf{B}_{12}+\mathbf{A}_{22}\mathbf{B}_{22}
\end{array}\right].
$$

The transpose of a partitioned matrix satisfies
$$
\mathbf{A^{\prime}}=\left[\begin{array}{cc}
\mathbf{A}_{11} & \mathbf{A}_{12}\\
\mathbf{A}_{21} & \mathbf{A}_{22}
\end{array}\right]^{\prime}=\left[\begin{array}{cc}
\mathbf{A}_{11}^{\prime} & \mathbf{A}_{21}^{\prime}\\
\mathbf{A}_{12}^{\prime} & \mathbf{A}_{22}^{\prime}
\end{array}\right].
$$
Notice the interchange of the two off-diagonal blocks. 

A partitioned matrix $\mathbf{A}$ with appropriate invertible sub-matrices
$\mathbf{A}_{11}$ and $\mathbf{A}_{22}$ has a partitioned inverse
of the form 
$$
\mathbf{A}^{-1}=\left[\begin{array}{cc}
\mathbf{A}_{11} & \mathbf{A}_{12}\\
\mathbf{A}_{21} & \mathbf{A}_{22}
\end{array}\right]^{-1}=\left[\begin{array}{cc}
\mathbf{A}_{11}^{-1}+\mathbf{A}_{11}^{-1}\mathbf{A}_{12}\mathbf{C}^{-1}\mathbf{A}_{21}\mathbf{A}_{11}^{-1} & -\mathbf{A}_{11}\mathbf{A}_{12}\mathbf{C}^{-1}\\
-\mathbf{C}^{-1}\mathbf{A}_{21}\mathbf{A}_{11}^{-1} & \mathbf{C}^{-1}
\end{array}\right],
$$
where $\mathbf{C}=\mathbf{A}_{22}-\mathbf{A}_{21}\mathbf{A}_{11}^{-1}\mathbf{A}_{12}$.
This formula can be verified by direct calculation.

## Positive Definite Matrices {#sec-Positive-Definite-Matrices}

::: {#def-positive-definite-matrix name="Positive definite matrix"}
Let $\mathbf{A}$ be an $n\times n$ symmetrix matrix. The matrix
$\mathbf{A}$ is *positive definite*\index{Positive definite} if for any $n\times1$ vector
$\mathbf{x\neq0}$
$$
\mathbf{x}^{\prime}\mathbf{A}\mathbf{x}>0.
$$
::: 

This condition implies that there is no non-zero vector $\mathbf{x}$
such that $\mathbf{Ax}=\mathbf{0}$, which implies that $\mathbf{A}$
has full rank $n$.

::: {#def-positive-semi-definite-matrix name="Positive semi-definite matrix"}
The matrix $\mathbf{A}$ is *positive semi-definite*\index{Positive semi-definite} if for any
$n\times1$ vector $\mathbf{x\neq0}$
$$
\mathbf{x}^{\prime}\mathbf{A}\mathbf{x}\geq0.
$$
:::

Hence, if $\mathbf{A}$ is positive semi-definite then there exists
some $n\times1$ vector $\mathbf{x}$ such that $\mathbf{Ax}=\mathbf{0}$,
which implies that $\mathbf{A}$ does not have full rank $n$.

### Matrix square root

To motivate the idea of the square root of a matrix, consider the
square root factorization of a positive number $a:$
$$
a=\sqrt{a}\times\sqrt{a}=a^{1/2}\times a^{1/2}.
$$
That is, we can factorize any positive number into the product of
its square root. It turns out we can do a similar factorization for
a positive definite and symmetric matrix.

::: {#prp-cholesky-factorization name="Cholesky factorization"}
Let $\mathbf{A}$ be an $n\times n$ positive definite and symmetric
matrix. Then it is possible to factor $\mathbf{A}$ as
$$
\mathbf{A}=\mathbf{C^{\prime}}\mathbf{C},
$$
where $\mathbf{C}$ is an $n\times n$ upper triangular matrix with
non-negative diagonal elements called the *Cholesky factor*\index{Cholesky factorization} of
$\mathbf{A}$. This factorization is also called a *matrix square root factorization*
by defining the square root matrix $\mathbf{A}^{1/2}=\mathbf{C}$.
Then we can write $\mathbf{A}=\mathbf{A}^{1/2\prime}\mathbf{A}^{1/2}$.
^[The Cholesky factorization is one way of defining the matrix square
root, where the square root matrix is upper triangular. Other matrix
square factorizations exist where the square root matrix is not necessarily
upper triangular.] If all of the diagonal elements of $\mathbf{C}$ are positive then
$\mathbf{A}$ is positive definite. Otherwise, $\mathbf{A}$ is positive
semi-definite.
:::

::: {#exm-cholesky-decomposition-in-r .example name="Cholesky decomposition in R"}
The R function `chol()` computes the Cholesky factorization
of a square symmetrix matrix. For example, consider the $2\times2$
correlation matrix:

```{r}
#| echo: true

Rho = matrix(c(1, 0.75, 0.75, 1), 2, 2)
Rho
```

The Cholesky matrix square root factorization is

```{r}
#| echo: true

C = chol(Rho)
C
t(C)%*%C
```

Here, all the diagonal elements of $\mathbf{C}$ are positive so that
covariance matrix is positive definite. Next, suppose the correlation
matrix is

```{r}
#| echo: true

Rho = matrix(c(1, 1, 1, 1), 2, 2)
Rho
```

so that the two variables are perfectly correlated. Then

```{r}
#| echo: true

chol(Rho, pivot=TRUE)
```

Here, we see that one of the diagonal elements of the Cholesky factor
is zero and that the correlation matrix is rank deficient.

\hfill{}$\blacksquare$
:::

## Multivariate Probability Distributions Using Matrix Algebra {#sec-Multivariate-Probability-Distributions-Using-Matrix-Algebra}

In this section, we show how matrix algebra can be used to simplify
many of the messy expressions concerning expectations and covariances
between multiple random variables, and we show how certain multivariate
probability distributions (e.g., the multivariate normal distribution)
can be expressed using matrix algebra.

### Random vectors

Let $X_{1},\ldots,X_{n}$ denote $n$ random variables. For $i=1,\ldots,n$
let $\mu_{i}=E[X_{i}]$ and $\sigma_{i}^{2}=\mathrm{var}(X_{i})$,
and let $\sigma_{ij}=\mathrm{cov}(X_{i},X_{j})$ for $i\neq j$. Define
the $n\times1$ random vector $\mathbf{X}=(X_{1},\ldots,X_{n})^{\prime}$.
Associated with $\mathbf{X}$ is the $n\times1$ vector of expected
values: 
$$
\underset{n \times 1}{\boldsymbol{\mu}}=E[\mathbf{X}]=\left(\begin{array}{c}
E[X_{1}]\\
\vdots\\
E[X_{n}]
\end{array}\right)=\left(\begin{array}{c}
\mu_{1}\\
\vdots\\
\mu_{n}
\end{array}\right). 
$$ {#eq-vector-expected-return}

### Covariance matrix

The covariance matrix\index{Covariance matrix}, $\boldsymbol{\Sigma}$, summarizes the variances
and covariances of the elements of the random vector $\mathbf{X}$.
In general, the covariance matrix of a random vector $\mathbf{X}$
(sometimes called the variance of the vector $\mathbf{X}$) with mean    
vector $\mu$ is defined as: 
$$
\mathrm{cov}(\mathbf{X})=E[(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}]=\boldsymbol{\Sigma}. 
$$ {#eq-vector-covariance}

If $\mathbf{X}$ has $n$ elements then $\boldsymbol{\Sigma}$ will be
the symmetric and positive semi-definite $n\times n$ matrix, 

$$
\underset{n\times n}{\boldsymbol{\Sigma}}=\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \cdots & \sigma_{1n}\\
\sigma_{12} & \sigma_{2}^{2} & \cdots & \sigma_{2n}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1n} & \sigma_{2n} & \cdots & \sigma_{n}^{2}
\end{array}\right) .
$$ {#eq-vector-covariance-matrix}

For the case $n=2$, we have by direct calculation 
$$
\begin{aligned}
E[(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}] & =E\left[\left(\begin{array}{l}
X_{1}-\mu_{1}\\
X_{2}-\mu_{2}
\end{array}\right)\cdot\left(X_{1}-\mu_{1},X_{2}-\mu_{2}\right)\right]\\
 & =E\left[\left(\begin{array}{ll}
(X_{1}-\mu_{1})^{2} & (X_{1}-\mu_{1})(X_{2}-\mu_{2})\\
(X_{2}-\mu_{2})(X_{1}-\mu_{1}) & (X_{2}-\mu_{2})^{2}
\end{array}\right)\right]\\
 & =\left(\begin{array}{ll}
E[(X_{1}-\mu_{1})^{2}] & E[(X_{1}-\mu_{1})(X_{2}-\mu_{2})]\\
E[(X_{2}-\mu_{2})(X_{1}-\mu_{1})] & E[(X_{2}-\mu_{2})^{2}]
\end{array}\right)\\
 & =\left(\begin{array}{ll}
\mathrm{var}(X_{1}) & \mathrm{cov}(X_{1},X_{2})\\
\mathrm{cov}(X_{2},X_{1}) & \mathrm{var}(X_{2})
\end{array}\right)=\left(\begin{array}{ll}
\sigma_{1}^{2} & \sigma_{12}\\
\sigma_{12} & \sigma_{2}^{2}
\end{array}\right)=\boldsymbol{\Sigma}.
\end{aligned}
$$
If $\rho_{12} \ne \pm 1$ then $\mathrm{det}(\boldsymbol{\Sigma}) \ne 0$, $\boldsymbol{\Sigma}$ has rank 2 and 
is positive definite. However, if $\rho_{12} = \pm 1$ then $\mathrm{det}(\boldsymbol{\Sigma})=0$, 
$\boldsymbol{\Sigma}$ has rank 1 and is positive semi-definite.

### Correlation matrix

The *correlation matrix*\index{Correlation matrix}, $\,\mathbf{C}$, summarizes all pairwise correlations
between the elements of the $n\times1$ random vector $\mathbf{X}$
and is given by
$$
\mathbf{C}=\left(\begin{array}{cccc}
1 & \rho_{12} & \cdots\  & \rho_{1n}\\
\rho_{12} & 1 & \cdots\  & \rho_{2n}\\
\vdots\  & \vdots\  & \ddots\  & \vdots\\
\rho_{1n} & \rho_{2n} & \cdots\  & 1
\end{array}\right) 
$$ {#eq-correlation-matrix}

Let $\boldsymbol{\Sigma}$ denote the covariance matrix of $\mathbf{X}$.
Then the correlation matrix $\mathbf{C}$ can be computed using
$$
\mathbf{C}=\mathbf{D}^{-1}\boldsymbol{\Sigma}\mathbf{D}^{-1},
$$ {#eq-MatrixReviewCov2Cor}
where $\mathbf{D}$ is an $n\times n$ diagonal matrix with the standard
deviations of the elements of $\mathbf{X}$ along the main diagonal:
$$
\mathbf{D}=\left(\begin{array}{cccc}
\sigma_{1} & 0 & \cdots & 0\\
0 & \sigma_{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & \sigma_{n}
\end{array}\right).
$$

::: {#exm-computing-correlation-matrix-from-covariance-matrix-in-r name="Computing correlation matrix from covariance matrix in R"}
In R the `cov2cor()` function can be used to compute the correlation
matrix given the covariance matrix using @eq-MatrixReviewCov2Cor:

```{r}
sigma.1 = 1
sigma.2 = 2
rho.12 = 0.5
sigma.12 = rho.12*sigma.1*sigma.2
Sigma = matrix(c(sigma.1^2, sigma.12, sigma.12, sigma.2^2), 
               2, 2, byrow=TRUE)
Sigma
cov2cor(Sigma)
```

\hfill{}$\blacksquare$
:::

### Variance of linear combination of random vectors

Consider the $n\times1$ random vector $\mathbf{X}$ with mean vector
$\mu$ and covariance matrix $\boldsymbol{\Sigma}$. Let $\mathbf{a}=(a_{1},\ldots,a_{n})^{\prime}$
be an $n\times1$ vector of constants and consider the random variable
$Y=\mathbf{a}^{\prime}\mathbf{X}=a_{1}X_{1}+\cdots+a_{n}X_{n}$. The expected value
of $Y$ is:

$$
\mu_{Y}=E[Y]=E[\mathbf{a}^{\prime}\mathbf{X}]=\mathbf{a}^{\prime}E[\mathbf{X}]=
\mathbf{a}^{\prime}\boldsymbol{\mu}.
$$ {#eq-mean-linear-combination}

The variance of $Y$ is 
$$
\mathrm{var}(Y)=\mathrm{var}(\mathbf{a}^{\prime}\mathbf{X}) = E[(\mathbf{a}^{\prime}\mathbf{X}-\mathbf{a}^{\prime}\boldsymbol{\mu})^{2}]
=E[(\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu}))^{2}]
$$ {#eq-variance-linear-combination}
since $Y=\mathbf{a}^{\prime}\mathbf{X}$ is a scalar. Now we use a
trick from matrix algebra. If $z$ is a scalar (think of $z=2$) then
$z^{\prime}z=z\cdot z^{\prime}=z^{2}$. Let $z=\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})$
and so $z\cdot z^{\prime}=\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}\mathbf{a}$.
Then, 
$$
\begin{aligned}
\mathrm{var}(Y) & =E[z^{2}]=E[z\cdot z^{\prime}]\\
 & =E[\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}\mathbf{a}]  \\
 & =\mathbf{a}^{\prime}E[(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}]\mathbf{a}  \\
 & =\mathbf{a}^{\prime}\mathrm{cov}(\mathbf{X})\mathbf{a}=\mathbf{a}^{\prime}\boldsymbol{\Sigma}\mathbf{a}.
\end{aligned}
$$ {#eq-variance-linear-combination}

### Covariance between linear combination of two random vectors

Consider the $n\times1$ random vector $\mathbf{X}$ with mean vector
$\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$. Let $\mathbf{a}=(a_{1},\ldots,a_{n})^{\prime}$
and $\mathbf{b}=(b_{1},\ldots,b_{n})^{\prime}$ be $n\times1$ vectors
of constants, and consider the random variable $Y=\mathbf{a}^{\prime}\mathbf{X}=a_{1}X_{1}+\cdots+a_{n}X_{n}$
and $Z=\mathbf{b}^{\prime}\mathbf{X}=b_{1}X_{1}+\cdots+b_{n}X_{n}$.
From the definition of covariance we have: 
$$
\mathrm{cov}(Y,Z)=E[(Y-E[Y])(Z-E[Z])]
$$
which may be rewritten in matrix notation as, 
$$
\begin{aligned}
\mathrm{cov}(\mathbf{a}^{\prime}\mathbf{X,b}^{\prime}\mathbf{X}) & =E[(\mathbf{a}^{\prime}\mathbf{X}-\mathbf{a}^{\prime}\boldsymbol{\mu})(\mathbf{b}^{\prime}\mathbf{X-b}^{\prime}\boldsymbol{\mu})] \\
 & =E[\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})\mathbf{b}^{\prime}(\mathbf{X}-\boldsymbol{\mu})] \\
 & =E[\mathbf{a}^{\prime}(\mathbf{X}-\mu)(\mathbf{X}-\boldsymbol{\mu})^{\prime}\mathbf{b}]  \\
 & =\mathbf{a}^{\prime}E[(\mathbf{X}\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}]\mathbf{b}  \\
 & =\mathbf{a}^{\prime}\boldsymbol{\Sigma} \mathbf{b}.
\end{aligned}
$$ {#eq-variance-two-linear-combinations}
Since $\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})$ and $\mathbf{b}^{\prime}(\mathbf{X}-\boldsymbol{\mu})$
are scalars, we can use the trick: 
$$
\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})\mathbf{b}^{\prime}(\mathbf{X}-\boldsymbol{\mu})=
\mathbf{a}^{\prime}(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}\boldsymbol{\mu})^{\prime}\mathbf{b}.
$$

### Variance of a vector of linear functions of a random vector

Consider the $n\times 1$ random vector $\mathbf{X}$ with mean vector
$\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$. Let $\mathbf{A}$
be an $m\times n$ matrix. Then the $m\times 1$ random vector $\mathbf{Y}=\mathbf{AX}$
represents $m$ linear functions of $\mathbf{X}$. By the linearity
of expectation, we have $E[\mathbf{Y}]=\mathbf{A}E[\mathbf{X}]=\mathbf{A}\boldsymbol{\mu}.$
The $m\times m$ covariance matrix of $\mathbf{Y}$ is given by
$$
\begin{aligned}
\mathrm{var}(\mathbf{Y}) & = E[(\mathbf{Y}-E[\mathbf{Y}])(\mathbf{Y}-E[\mathbf{Y}])^{\prime}]  \\
 & = E[(\mathbf{AX}-\mathbf{A}\boldsymbol{\mu})(\mathbf{AX}-\mathbf{A}\boldsymbol{\mu})^{\prime}]  \\
 & = E[\mathbf{A}(\mathbf{X}-\boldsymbol{\mu})(\mathbf{A}(\mathbf{X}-\boldsymbol{\mu}))^{\prime}]  \\
 & = E[\mathbf{A}(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}\mathbf{A}^{\prime}]  \\
 & = \mathbf{A}E[(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}]\mathbf{A}^{\prime}  \\
 & = \mathbf{A}\boldsymbol{\Sigma}\mathbf{A}^{\prime}.
\end{aligned}
$$ {#eq-variance-vector-linear-combinations}

### Multivariate normal distribution

Consider $n$ random variables $X_{1},\ldots,X_{n}$ and assume they
are jointly normally distributed. Define the $n\times1$ vectors $\mathbf{X}=(X_{1},\ldots,X_{n})^{\prime}$,
$\mathbf{x}=(x_{1},\ldots,x_{n})^{\prime}$ and $\mu=(\mu_{1},\ldots,\mu_{n})^{\prime}$,
and the $n\times n$ covariance matrix: 
$$
\boldsymbol{\Sigma}=\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \cdots & \sigma_{1n}\\
\sigma_{12} & \sigma_{2}^{2} & \cdots & \sigma_{2n}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1n} & \sigma_{2n} & \cdots & \sigma_{n}^{2}
\end{array}\right).
$$
Then $\mathbf{X}\sim N(\boldsymbol{\mu},\boldsymbol{\Sigma})$ means that
the random vector $\mathbf{X}$ has a *multivariate normal distribution*\index{Multivariate normal distribution}
with mean vector $\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$.
The pdf of the multivariate normal distribution can be compactly expressed
as: 
\begin{align*}
f(\mathbf{x}) & =\frac{1}{2\pi^{n/2}\det(\boldsymbol{\Sigma})^{1/2}}e^{-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^{\prime}\boldsymbol{\Sigma}^{-1}(\mathbf{x}-\boldsymbol{\mu})}\\
 & =(2\pi)^{-n/2}\det(\boldsymbol{\Sigma})^{-1/2}e^{-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^{\prime}\boldsymbol{\Sigma}^{-1}(\mathbf{x}-\boldsymbol{\mu})}.
\end{align*}

::: {#exm-creating-a-multivariate-normal-random-vector-from-a-vector-of-independent-standard-normal-random-variables .example name="Creating a multivariate normal random vector from a vector of independent standard normal random variables"}
Let $\mathbf{Z}=(Z_{1},\ldots,Z_{n})^{\prime}$ denote a vector of
$n$ iid standard normal random variables. Then $\mathbf{Z}\sim N(\mathbf{0},\,\mathbf{I}_{n})$
where $\mathbf{I}_{n}$ denotes the $n-$dimensional identity matrix.
Given $\mathbf{Z}$, we would like to create a random vector $\mathbf{X}$
such that $\mathbf{X}\sim N(\boldsymbol{\mu},\,\boldsymbol{\Sigma}).$ Let
$\boldsymbol{\Sigma}^{1/2}$ denote the upper triangular Cholesky factor of $\boldsymbol{\Sigma}$
such that $\boldsymbol{\Sigma}=\boldsymbol{\Sigma}^{1/2 \prime}\boldsymbol{\Sigma}^{1/2}.$
Define $\mathbf{X}$ using
$$
\mathbf{X}=\boldsymbol{\mu}+\boldsymbol{\Sigma}^{1/2 \prime}\mathbf{Z}.
$$
Then 
\begin{eqnarray*}
E[\mathbf{X}] & = & \boldsymbol{\mu}+\boldsymbol{\Sigma}^{1/2}E[\mathbf{Z}]=\boldsymbol{\mu},\\
\mathrm{var}(\mathbf{X}) & = & \mathrm{var}\left(\boldsymbol{\Sigma}^{1/2}\mathbf{Z}\right)=\boldsymbol{\Sigma}^{1/2 \prime}\mathrm{var}(\mathbf{Z})\boldsymbol{\Sigma}^{1/2}=\boldsymbol{\Sigma}^{1/2 \prime}\mathbf{I}_{n}\boldsymbol{\Sigma}^{1/2}=\boldsymbol{\Sigma}.
\end{eqnarray*}
Hence, $\mathbf{X}\sim N(\boldsymbol{\mu},\,\boldsymbol{\Sigma}).$ 

\hfill{}$\blacksquare$
:::

::: {#exm-simulating-multivariate-normal-random-vectors-in-r .example name="Simulating multivariate normal random vectors in R"}
Here, we illustrate how to simulate $\mathbf{X} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$ where $\boldsymbol{\mu} = (1,1)^{\prime}$ and
$$
\boldsymbol{\Sigma} = \left(\begin{array}{ll}
1 & 1\\
1 & 4
\end{array}\right)
$$
First, specify the mean vector and covariance matrix for the the bivariate normal distribution:
```{r}
mu.vec = c(1,1)
sigma.1 = 1
sigma.2 = 2
rho.12 = 0.5
sigma.12 = rho.12*sigma.1*sigma.2
Sigma = matrix(c(sigma.1^2, sigma.12, sigma.12, sigma.2^2), 
               2, 2, byrow=TRUE)
mu.vec
Sigma
```
Second, compute the Cholesky factorization of $\boldsymbol{\Sigma}$:
```{r}
Sigma.5 = chol(Sigma)
Sigma.5
```
Third, simulate $\mathbf{Z} \sim N(\mathbf{0}, \mathbf{I}_{2})$
```{r}
n = 2
set.seed(123)
Z = rnorm(n)
```

Fourth, compute $\mathbf{X} = \boldsymbol{\mu} + \boldsymbol{\Sigma}^{1/2 \prime}\mathbf{Z}$
```{r}
X = mu.vec + t(Sigma.5)%*%Z
X
```
\hfill{}$\blacksquare$
:::

## Portfolio Math Using Matrix Algebra {#sec-Portfolio-Math-Using-Matrix-Algebra}

Let $R_{i}$ denote the return on asset $i=A,B,C$ and assume that
$R_{A},R_{B}$ and $R_{C}$ are jointly normally distributed with
means, variances and covariances: 
$$
\mu_{i}=E[R_{i}],~\sigma_{i}^{2}=\mathrm{var}(R_{i}),~\mathrm{cov}(R_{i},R_{j})=\sigma_{ij}.
$$
Let $x_{i}$ denote the share of wealth invested in asset $i$ $(i=A,B,C)$,
and assume that all wealth is invested in the three assets so that
$x_{A}+x_{B}+x_{C}=1$. The portfolio return, $R_{p,x}$, is the random
variable: 
$$
R_{p,x}=x_{A}R_{A}+x_{B}R_{B}+x_{C}R_{C}.
$$ {#eq-Rpx}
The subscript "$x$" indicates that the portfolio
is constructed using the x-weights $x_{A},x_{B}$ and $x_{C}$. The
expected return on the portfolio is: 
$$
\mu_{p,x}=E[R_{p,x}]=x_{A}\mu_{A}+x_{B}\mu_{B}+x_{C}\mu_{C},
$$ {#eq-mu-px}
and the variance of the portfolio return is: 
$$
\begin{aligned}
\sigma_{p,x}^{2}&=\mathrm{var}(R_{p,x})=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+x_{C}^{2}\sigma_{C}^{2}+2x_{A}x_{B}\sigma_{AB} \nonumber\\
&+2x_{A}x_{C}\sigma_{AC}+2x_{B}x_{C}\sigma_{BC}.
\end{aligned}
$$ {#eq-var-px}
Notice that variance of the portfolio return depends on three variance
terms and six covariance terms. Hence, with three assets there are
twice as many covariance terms than variance terms contributing to
portfolio variance. Even with three assets, the algebra representing
the portfolio characteristics @eq-Rpx - @eq-var-px
is cumbersome. We can greatly simplify the portfolio algebra using
matrix notation. 


Define the following $3\times1$ column vectors containing the asset
returns and portfolio weights: 
$$
\mathbf{R}=\left(\begin{array}{c}
R_{A}\\
R_{B}\\
R_{C}
\end{array}\right),~\mathbf{x}=\left(\begin{array}{c}
x_{A}\\
x_{B}\\
x_{C}
\end{array}\right).
$$
The probability distribution of the random return vector $\mathbf{R}$
is simply the joint distribution of the elements of $\mathbf{R}$.
Here all returns are jointly normally distributed and this joint distribution
is completely characterized by the means, variances and covariances
of the returns. We can easily express these values using matrix notation
as follows. The $3\times1$ vector of portfolio expected values is:
$$
E[\mathbf{R}]=E\left[\left(\begin{array}{c}
R_{A}\\
R_{B}\\
R_{C}
\end{array}\right)\right]=\left(\begin{array}{c}
E[R_{A}]\\
E[R_{B}]\\
E[R_{C}]
\end{array}\right)=\left(\begin{array}{c}
\mu_{A}\\
\mu_{B}\\
\mu_{C}
\end{array}\right)=\boldsymbol{\mu},
$$
and the $3\times3$ covariance matrix of returns is, 
$$
\begin{aligned}
\mathrm{var}(\mathbf{R}) & =\left(\begin{array}{ccc}
\mathrm{var}(R_{A}) & \mathrm{cov}(R_{A},R_{B}) & \mathrm{cov}(R_{A},R_{C})\\
\mathrm{cov}(R_{B},R_{A}) & \mathrm{var}(R_{B}) & \mathrm{cov}(R_{B},R_{C})\\
\mathrm{cov}(R_{C},R_{A}) & \mathrm{cov}(R_{C},R_{B}) & \mathrm{var}(R_{C})
\end{array}\right)\\
 & =\left(\begin{array}{ccc}
\sigma_{A}^{2} & \sigma_{AB} & \sigma_{AC}\\
\sigma_{AB} & \sigma_{B}^{2} & \sigma_{BC}\\
\sigma_{AC} & \sigma_{BC} & \sigma_{C}^{2}
\end{array}\right)=\boldsymbol{\Sigma}.
\end{aligned}
$$
Notice that the covariance matrix is symmetric (elements off the diagonal
are equal so that $\boldsymbol{\Sigma} = \boldsymbol{\Sigma}^{\prime}$, where $\boldsymbol{\Sigma}^{\prime}$
denotes the transpose of $\boldsymbol{\Sigma}$) since $\mathrm{cov}(R_{A},R_{B})=\mathrm{cov}(R_{B},R_{A})$,
$\mathrm{cov}(R_{A},R_{C})=\mathrm{cov}(R_{C},R_{A})$ and $\mathrm{cov}(R_{B},R_{C})=\mathrm{cov}(R_{C},R_{B})$.


The return on the portfolio using vector notation is: 
$$
R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}=(x_{A},x_{B},x_{C})\cdot\left(\begin{array}{c}
R_{A}\\
R_{B}\\
R_{C}
\end{array}\right)=x_{A}R_{A}+x_{B}R_{B}+x_{C}R_{C}.
$$
Similarly, the expected return on the portfolio is: 
\begin{align*}
\mu_{p,x} & =E[\mathbf{x}^{\prime}\mathbf{R]=x}^{\prime}E[\mathbf{R}]=\mathbf{x}^{\prime}\mu\\
 & =(x_{A},x_{B},x_{C})\cdot\left(\begin{array}{c}
\mu_{A}\\
\mu_{B}\\
\mu_{C}
\end{array}\right)=x_{A}\mu_{A}+x_{B}\mu_{B}+x_{C}\mu_{C}.
\end{align*}
The variance of the portfolio is: 
\begin{align*}
\sigma_{p,x}^{2} & =\mathrm{var}(\mathbf{x}^{\prime}\mathbf{R})=\mathbf{x}^{\prime}\boldsymbol{\Sigma}\mathbf{x}=(x_{A},x_{B},x_{C})\cdot\left(\begin{array}{ccc}
\sigma_{A}^{2} & \sigma_{AB} & \sigma_{AC}\\
\sigma_{AB} & \sigma_{B}^{2} & \sigma_{BC}\\
\sigma_{AC} & \sigma_{BC} & \sigma_{C}^{2}
\end{array}\right)\left(\begin{array}{c}
x_{A}\\
x_{B}\\
x_{C}
\end{array}\right)\\
 & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+x_{C}^{2}\sigma_{C}^{2}+2x_{A}x_{B}\sigma_{AB}+2x_{A}x_{C}\sigma_{AC}+2x_{B}x_{C}\sigma_{BC}.
\end{align*}
Finally, the condition that the portfolio weights sum to one can be
expressed as: 
$$
\mathbf{x}^{\prime}\mathbf{1}=(x_{A},x_{B},x_{B})\cdot\left(\begin{array}{c}
1\\
1\\
1
\end{array}\right)=x_{A}+x_{B}+x_{C}=1,
$$
where $\mathbf{1}$ is a $3\times1$ vector with each element equal
to 1. 


Consider another portfolio with weights $\mathbf{y}=(y_{A},y_{B},y_{C})^{\prime}$. 
The return on this portfolio is: 
$$
R_{p,y}=\mathbf{y}^{\prime}\mathbf{R}=y_{A}R_{A}+y_{B}R_{B}+y_{C}R_{C}.
$$
We often need to compute the covariance between the return on portfolio
$\mathbf{x}$ and the return on portfolio $\mathbf{y}$, $\mathrm{cov}(R_{p,x},R_{p,y})$.
This can be easily expressed using matrix algebra: 
$$
\begin{aligned}
\sigma_{xy} & =\mathrm{cov}(R_{p,x},R_{p,y})=\mathrm{cov}(\mathbf{x}^{\prime}\mathbf{R},\mathbf{y}^{\prime}\mathbf{R})\\
 & =E[(\mathbf{x}^{\prime}\mathbf{R}-\mathbf{x}^{\prime}\boldsymbol{\mu}])(\mathbf{y}^{\prime}\mathbf{R}-\mathbf{y}^{\prime}\boldsymbol{\mu})]=E[\mathbf{x}^{\prime}(\mathbf{R}-\boldsymbol{\mu})(\mathbf{R}-\boldsymbol{\mu})^{\prime}\mathbf{y}]\\
 & =\mathbf{x}^{\prime}E[(\mathbf{R}-\boldsymbol{\mu})(\mathbf{R}-\boldsymbol{\mu})^{\prime}]\mathbf{y}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{y}.
\end{aligned}
$$
Notice that, 
$$
\begin{aligned}
\sigma_{xy} & =\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{y}=(x_{A},x_{B},x_{C})\cdot\left(\begin{array}{ccc}
\sigma_{A}^{2} & \sigma_{AB} & \sigma_{AC}\\
\sigma_{AB} & \sigma_{B}^{2} & \sigma_{BC}\\
\sigma_{AC} & \sigma_{BC} & \sigma_{C}^{2}
\end{array}\right)\left(\begin{array}{c}
y_{A}\\
y_{B}\\
y_{C}
\end{array}\right)\\
 & =x_{A}y_{A}\sigma_{A}^{2}+x_{B}y_{B}\sigma_{B}^{2}+x_{C}y_{C}\sigma_{C}^{2}\\
 & +(x_{A}y_{B}+x_{B}y_{A})\sigma_{AB}+(x_{A}y_{C}+x_{C}y_{A})\sigma_{AC}+(x_{B}y_{C}+x_{C}y_{B})\sigma_{BC},
\end{aligned}
$$
which is quite a messy expression! 


## Derivatives of Simple Matrix Functions {#sec-Derivatives-of-Simple-Matrix-Functions}

**Result**\index{Derivatives of matrices}: Let $\mathbf{A}$ be an $n\times n$ symmetric matrix,
and let $\mathbf{x}$ and $\mathbf{y}$ be $n\times1$ vectors. Then,
$$
\begin{aligned}
\underset{n\times1}{\frac{\partial}{\partial\mathbf{x}}}\mathbf{x}^{\prime}\mathbf{y} & =\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\mathbf{x}^{\prime}\mathbf{y}\\
\vdots\\
\frac{\partial}{\partial x_{n}}\mathbf{x}^{\prime}\mathbf{y}
\end{array}\right)=\mathbf{y},
\end{aligned}
$$ {#eq-matrix-deriv-1}
$$
\begin{aligned}
\underset{n\times1}{\frac{\partial}{\partial\mathbf{x}}}\mathbf{Ax} & =\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\left(\mathbf{Ax}\right)^{\prime}\\
\vdots\\
\frac{\partial}{\partial x_{n}}\left(\mathbf{Ax}\right)^{\prime}
\end{array}\right)=\mathbf{A},
\end{aligned}
$$ {#eq-matrix-deriv-2}
$$
\begin{aligned}
\underset{n\times1}{\frac{\partial}{\partial\mathbf{x}}}\mathbf{x}^{\prime}\mathbf{Ax} & =\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\mathbf{x}^{\prime}\mathbf{Ax}\\
\vdots\\
\frac{\partial}{\partial x_{n}}\mathbf{x}^{\prime}\mathbf{Ax}
\end{array}\right)=2\mathbf{Ax}.
\end{aligned}
$$ {#eq-matrix-deriv-3}

We will demonstrate these results with simple examples. Let, 
$$
\mathbf{A}=\left(\begin{array}{cc}
a & b\\
b & c
\end{array}\right),~\mathbf{x}=\left(\begin{array}{c}
x_{1}\\
x_{2}
\end{array}\right),\mathbf{y}=\left(\begin{array}{c}
y_{1}\\
y_{2}
\end{array}\right).
$$
First, consider @eq-matrix-deriv-1. Now, 
$$
\mathbf{x}^{\prime}\mathbf{y}=x_{1}y_{1}+x_{2}y_{2}.
$$
Then, 
$$
\frac{\partial}{\partial\mathbf{x}}\mathbf{x}^{\prime}\mathbf{y}=\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\mathbf{x}^{\prime}\mathbf{y}\\
\frac{\partial}{\partial x_{2}}\mathbf{x}^{\prime}\mathbf{y}
\end{array}\right)=\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\left(x_{1}y_{1}+x_{2}y_{2}\right)\\
\frac{\partial}{\partial x_{2}}\left(x_{1}y_{1}+x_{2}y_{2}\right)
\end{array}\right)=\left(\begin{array}{c}
y_{1}\\
y_{2}
\end{array}\right)=\mathbf{y}.
$$
Next, consider @eq-matrix-deriv-2. Note that, 
$$
\mathbf{Ax}=\left(\begin{array}{cc}
a & b\\
b & c
\end{array}\right)\left(\begin{array}{c}
x_{1}\\
x_{2}
\end{array}\right)=\left(\begin{array}{c}
ax_{1}+bx_{2}\\
bx_{1}+cx_{2}
\end{array}\right),
$$
and, 
$$
\left(\mathbf{Ax}\right)^{\prime}=\left(ax_{1}+bx_{2},bx_{1}+cx_{2}\right).
$$
Then, 
$$
\frac{\partial}{\partial\mathbf{x}}\mathbf{Ax}=\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\left(ax_{1}+bx_{2},bx_{1}+cx_{2}\right)\\
\frac{\partial}{\partial x_{2}}\left(ax_{1}+bx_{2},bx_{1}+cx_{2}\right)
\end{array}\right)=\left(\begin{array}{cc}
a & b\\
b & c
\end{array}\right)=\mathbf{A.}
$$
Finally, consider @eq-matrix-deriv-3. We have, 
$$
\mathbf{x}^{\prime}\mathbf{Ax}=\left(\begin{array}{cc}
x_{1} & x_{2}\end{array}\right)\left(\begin{array}{cc}
a & b\\
b & c
\end{array}\right)\left(\begin{array}{c}
x_{1}\\
x_{2}
\end{array}\right)=ax_{1}^{2}+2bx_{1}x_{2}+cx_{2}^{2}.
$$
Then, 
$$
\begin{aligned}
\frac{\partial}{\partial\mathbf{x}}\mathbf{x}^{\prime}\mathbf{Ax} & =\left(\begin{array}{c}
\frac{\partial}{\partial x_{1}}\left(ax_{1}^{2}+2bx_{1}x_{2}+cx_{2}^{2}\right)\\
\frac{\partial}{\partial x_{2}}\left(ax_{1}^{2}+2bx_{1}x_{2}+cx_{2}^{2}\right)
\end{array}\right)=\left(\begin{array}{c}
2ax_{1}+2bx_{2}\\
2bx_{1}+2cx_{2}
\end{array}\right)\\
 & =2\left(\begin{array}{cc}
a & b\\
b & c
\end{array}\right)\left(\begin{array}{c}
x_{1}\\
x_{2}
\end{array}\right)=2\mathbf{Ax}.
\end{aligned}
$$

::: {#exm-calculating-asset-marginal-contribution-to-portfolio-volatility name="Calculating an asset's marginal contribution to portfolio volatility"}
In portfolio risk budgeting (see chapter @sec-Risk-Budgeting),
asset *i*'s marginal contribution to portfolio volatility $\sigma_{p}=\left(\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\right)^{1/2}$
is given by
$$
\mathrm{MCR_{i}^{\sigma}}=\frac{\partial\sigma_{p}}{\partial x_{i}}=\frac{\partial \left(\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\right)^{1/2}}{\partial x_{i}},
$$
and approximates how much portfolio volatility changes when the allocation
to asset *i* increases by a small amount. Using the chain rule
and matrix derivatives we can compute the entire vector of asset marginal
contributions at once: 

$$
\begin{aligned}
\frac{\partial \left(\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\right)^{1/2}}{\partial\mathbf{x}} & = 
\frac{1}{2}\left(\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\right)^{-1/2} \frac{\partial \mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}}{\partial\mathbf{x}} = \frac{1}{2}\left(\mathbf{x}^{\prime} \boldsymbol{\Sigma} \mathbf{x} \right)^{-1/2} 2 \boldsymbol{\Sigma} \mathbf{x} \\
& = \left(\mathbf{x}^{\prime} \boldsymbol{\Sigma} \mathbf{x} \right)^{-1/2} \boldsymbol{\Sigma} \mathbf{x}= \frac{\boldsymbol{\Sigma} \mathbf{x}}{\sigma_{p}}.
\end{aligned}
$$
Then asset $i$'s marginal contribution is given by the $i$-th
row of $\frac{\boldsymbol{\Sigma} \mathbf{x}}{\sigma_{p}}$.

\hfill{}$\blacksquare$
:::

::: {#exm-finding-the-global-minimum-variance-portfolio .example name="Finding the global minimum variance portfolio"}
Let $\mathbf{R}$ denote an $n\times1$ random vector of asset returns
with $E[\mathbf{R}]=\mu$ and $\mathrm{var}(\mathbf{R})=\boldsymbol{\Sigma}$.
The global minimum variance portfolio (see Chapter @sec-Introduction-to-Portfolio-Theory, Section @sec-Efficient-portfolios-with-two-risky-assets)
$\mathbf{m}$ solves the constrained minimization problem: 
$$
\min_{\mathbf{m}}~\sigma_{p,m}^{2}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}\text{ s.t. }\mathbf{m}^{\prime}\mathbf{1}=1.
$$ {#eq-global-min-matrix-1}
The Lagrangian function is: 
$$
L(\mathbf{m},\lambda)=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}+\lambda\mathbf{(m}^{\prime}\mathbf{1}-1).
$$
The first order conditions can be expressed in matrix notation as,
$$
\begin{aligned}
\underset{(n\times1)}{\mathbf{0}} & =\frac{\partial L(\mathbf{m},\lambda)}{\partial\mathbf{m}}=\frac{\partial}{\partial\mathbf{m}}\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m+}\frac{\partial}{\partial\mathbf{m}}\lambda\mathbf{(m}^{\prime}\mathbf{1}-1)=2\cdot\boldsymbol{\Sigma} \mathbf{m}+\lambda\cdot\mathbf{1}
\end{aligned}
$$ {#eq-gmv-foc-1-1}
$$
\begin{aligned}
\underset{(1\times 1)}{0} & =\frac{\partial L(\mathbf{m},\lambda)}{\partial\lambda}=\frac{\partial}{\partial\lambda}\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m+}\frac{\partial}{\partial\lambda}\lambda\mathbf{(m}^{\prime}\mathbf{1}-1)=\mathbf{m}^{\prime}\mathbf{1}-1
\end{aligned}
$$ {#eq-gmv-foc-1-2}
These first order conditions represent a system of $n+1$ linear equations
in $n+1$ unknowns ($\mathbf{m}$ and $\lambda$). These equations
can be represented in matrix form as the system 
$$
\left[\begin{array}{cc}
2\boldsymbol{\Sigma} & \mathbf{1}\\
\mathbf{1}^{\prime} & 0
\end{array}\right]\left[\begin{array}{c}
\mathbf{m}\\
\lambda
\end{array}\right]=\left[\begin{array}{c}
\mathbf{0}\\
1
\end{array}\right],
$$
which is of the form $\mathbf{Az}=\mathbf{b}$ for 
$$
\mathbf{A}=\left[\begin{array}{cc}
2\boldsymbol{\Sigma} & \mathbf{1}\\
\mathbf{1}^{\prime} & 0
\end{array}\right],\,\mathbf{z}=\left[\begin{array}{c}
\mathbf{m}\\
\lambda
\end{array}\right],\,\mathbf{b}=\left[\begin{array}{c}
\mathbf{0}\\
1
\end{array}\right].
$$
 The portfolio weight vector $\mathbf{m}$ can be found as the first
$n$ elements of $\mathbf{z}=\mathbf{A}^{-1}\mathbf{b}$.

\hfill{}$\blacksquare$
:::

## Further Reading: Matrix Algebra Review

A classic textbook on linear algebra is [@strang-1980]. A review of
basic matrix algebra with applications in economics and finance is
given in [@chiang-1984]. Matrix algebra results that are useful for
statistical analysis are presented in [@searle-1982], Gentile (2010)
and [@banerjee-2014]. [@lutkepohl-1997] and [@abadir-2005] present advanced matrix algebra useful for econometrics and
statistics. [@fieller-2015] overviews the basics of matrix algebra
for statistics using R.

## Problems: Matrix Algebra Review

::: {#exr-01 .exercise}
Use R to answer the following questions.
1.  Create the following matrices and vectors
$$
\mathbf{A}=\left[\begin{array}{ccc}
1 & 4 & 7\\
2 & 4 & 8\\
6 & 1 & 3
\end{array}\right],\,\mathbf{B}=\left[\begin{array}{ccc}
4 & 4 & 0\\
5 & 9 & 1\\
2 & 2 & 5
\end{array}\right],\,\mathbf{x}=\left[\begin{array}{c}
1\\
2\\
3
\end{array}\right],\,\mathbf{y}=\left[\begin{array}{c}
5\\
2\\
7
\end{array}\right]
$$
2.  Compute the transposes of the above matrices and vectors.
3.  Compute $\mathbf{A}+\mathbf{B},\:\mathbf{A}-\mathbf{B},\,2\mathbf{A},\,\mathbf{Ax},\,\mathbf{AB,\,y^{\prime}Ax}$
:::

::: {#exr-02 .exercise}
Consider the system of linear equations
\begin{eqnarray*}
x+y & = & 1\\
2x+4y & = & 2
\end{eqnarray*}

1.  Plot the two lines and note the solution to the system of equations. Hint: you can use the R functions `abline()` or `curve()` to plot the lines. 
2.  Write the system in matrix notation as $\mathbf{Az}=\mathbf{b}$ and solve for $\mathbf{z}$ by computing $\mathbf{z}=\mathbf{A}^{-1}\mathbf{b}$. 
:::

::: {#exr-03 .exercise}
Consider creating an equally weighted portfolio of three assets denoted
A, B, and C. Assume the following information
$$
\mu=\left[\begin{array}{c}
0.01\\
0.04\\
0.02
\end{array}\right],\,\Sigma=\left[\begin{array}{ccc}
0.10 & 0.30 & 0.10\\
0.30 & 0.15 & -0.20\\
0.10 & -0.20 & 0.08
\end{array}\right].
$$
Use R to answer the following questions.
1.  Create the vector of portfolio weights. 
2.  Compute the expected return on the portfolio.
3.  Compute the variance and standard deviation of the portfolio.
:::

::: {#exr-04 .exercise}
Let $R_{i}$ be a random variable denoting the simple return on asset
$i\,(i=1,\ldots,N)$ with $E[R_{i}]=\mu_{i},$ $\mathrm{var}(R_{i})=\sigma_{i}^{2}$
and $\mathrm{cov}(R_{i},R_{j})=\sigma_{ij}.$ Define the $N\times1$
vectors $\mathbf{R}=(R_{1},\ldots,R_{N})^{\prime},$ $\boldsymbol{\mu}=(\mu_{1},\ldots,\mu_{N})^{\prime},$
$\mathbf{x}=(x_{1},\ldots,x_{N})^{\prime},$ $\mathbf{y}=(y_{1},\ldots,y_{N})^{\prime},$
and $\mathbf{1}=(1,\ldots,1)^{\prime}$, and the $N\times N$ covariance
matrix
$$
\boldsymbol{\Sigma}=\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \cdots & \sigma_{1N}\\
\sigma_{12} & \sigma_{2}^{2} & \cdots & \sigma_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1N} & \sigma_{2N} & \cdots & \sigma_{N}^{2}
\end{array}\right).
$$
The vectors $\mathbf{x}$ and $\mathbf{y}$ contain portfolio weights
(investment shares) that sum to one. Using simple matrix algebra,
answer the following questions.

1.  For the portfolios defined by the vectors $\mathbf{x}$ and $\mathbf{y}$ give the matrix algebra expressions for the portfolio returns $R_{p,x}$ and $R_{p,y}$ and the portfolio expected returns $\mu_{p,x}$ and $\mu_{p,y}$. 
2.  For the portfolios defined by the vectors $\mathbf{x}$ and $\mathbf{y}$ give the matrix algebra expressions for the constraint that the portfolio weights sum to one.
3.  For the portfolios defined by the vectors $\mathbf{x}$ and $\mathbf{y}$ give the matrix algebra expressions for the portfolio variances $\sigma_{p,x}^{2}$ and $\sigma_{p,y}^{2}$, and the covariance between $R_{p,x}$ and $R_{p,y}$.
4.  In the expression for the portfolio variance $\sigma_{p,x}^{2}$, how many variance terms are there? How many unique covariance terms are there?
:::


