# Time Series Concepts {#sec-Time-Series-Concepts}

Last updated on September 4, 2026

Copyright © Eric Zivot 2026


Financial variables such as asset prices and returns are naturally
ordered by time. That is, these variables are *time series* variables.
When we construct returns, the time index or data frequency becomes
the investment horizon associated with the return. Typical data frequencies
are daily, weekly, monthly and annual. In building probability models
for time series variables, the time ordering of the data matters because
we may think there are important temporal dependencies among the variables.
For example, we might have reason to believe that the return on an
asset this month is correlated with the return on the same asset in
the previous month. This autocorrelation can only be defined if the
time ordering of the data is preserved. A major complication in analyzing
time series data is that the usual assumption of random sampling from
a common population is not appropriate because it does not allow for
any kind of time dependence in the time series variables. We would
like to retain the notion that the observed data come from some population
model, perhaps with time-invariant parameters, but we would like to
allow the variables to have time dependencies. Fortunately, we can
do this if the time series data come from a *stationary* time series
process. 

This chapter reviews some basic times series concepts that are essential
for describing and modeling financial time series. Section @sec-Stochastic-Processes
defines univariate time series processes and introduces the important
concepts of stationarity and ergodicity. Covariance stationary time
series processes are defined, which gives meaning to measuring linear
time dependence using autocorrelation. The benchmark Gaussian White
Noise process and related processes are introduced and illustrated
using R. Some common non-stationary time series processes are also
discussed including the famous random walk model. Section @sec-Multivariate-Time-Series
introduces covariance stationary multivariate time series process.
Such processes allow for dynamic interactions among groups of time
series variables. Section @sec-Time-Series-Models discusses
time series model building and introduces the class of univariate
autoregressive-moving average time series models and multivariate
vector autoregression models. The properties of some simple models
are derived and it is shown how to simulate observations from these
models using R. The chapter concludes with a brief discussion of forecasting
from time series models. 


The R packages used in this chapter are **mvtnorm** and **vars**.
Make sure these packages are installed and loaded before running the
R examples in this chapter.

```{r}
#| include: false
options(digits=3)
library(mvtnorm)
library(vars)
```

## Stochastic Processes {#sec-Stochastic-Processes}

::: {#def-stochastic-process .definition name="Stochastic Process"}
A discrete-time *stochastic process*\index{Stochastic process} or *time series process*
$$
\{\ldots,Y_{1},Y_{2},\ldots,Y_{t},Y_{t+1},\ldots\}=\{Y_{t}\}_{t=-\infty}^{\infty}
$$
is a sequence of random variables indexed by time $t$^[To conserve on notation, we will represent the stochastic process$\{Y_{t}\}_{t=-\infty}^{\infty}$ simply as $\{Y_{t}\}.$].
:::

In most applications, the time index is a regularly spaced index
representing calendar time (e.g., days, months, years, etc.) but it
can also be irregularly spaced representing event time (e.g., intra-day
transaction times). In modeling time series data, the ordering imposed
by the time index is important because we often would like to capture
the temporal relationships, if any, between the random variables in
the stochastic process. In random sampling from a population, the
ordering of the random variables representing the sample does not
matter because they are independent. 

::: {#def-realization-of-stochastic-process .definition name="Realization of Stochastic Process"}
A realization of a stochastic process with $T$ observations is the
sequence of observed data 
$$
\{Y_{1}=y_{1},Y_{2}=y_{2},\ldots,Y_{T}=y_{T}\}=\{y_{t}\}_{t=1}^{T}.
$$
:::

The goal of time series modeling is to describe the probabilistic
behavior of the underlying stochastic process that is believed to
have generated the observed data in a concise way. In addition, we
want to be able to use the observed sample to estimate important characteristics
of a time series model such as measures of time dependence. In order
to do this, we need to make a number of assumptions regarding the
joint behavior of the random variables in the stochastic process such
that we may treat the stochastic process in much the same way as we
treat a random sample from a given population.

### Stationary stochastic processes

We often describe random sampling from a population as a sequence
of independent, and identically distributed (iid) random variables
$X_{1},X_{2}\ldots$ such that each $X_{i}$ is described by the same
probability distribution $F_{X}$, and write $X_{i}\sim F_{X}$. With
a time series process, we would like to preserve the identical distribution
assumption but we do not want to impose the restriction that each
random variable in the sequence is independent of all of the other
variables. In many contexts, we would expect some dependence between
random variables close together in time (e.g, $X_{1}$, and $X_{2})$
but little or no dependence between random variables far apart in
time (e.g., $X_{1}$ and $X_{100})$. We can allow for this type of
behavior using the concepts of *stationarity*\index{Stationarity} and *ergodicity*\index{Ergodicity}.


We start with the definition of strict stationarity.

::: {#def-strict-stationarity .definition name="Strict stationarity"}
A stochastic process $\{Y_{t}\}$ is *strictly stationary*\index{Strict stationary} if,
for any given finite integer $r$ and for any set of subscripts $t_{1},t_{2},\ldots,t_{r}$
the joint distribution of $(Y_{t_{1}},Y_{t_{2}},\ldots,Y_{t_{r}})$
depends only on $t_{1}-t,t_{2}-t,\ldots,t_{r}-t$ but not on $t$.
In other words, the joint distribution of $(Y_{t_{1}},Y_{t_{2}},\ldots,Y_{t_{r}})$
is the same as the joint distribution of $(Y_{t_{1}-t},Y_{t_{2}-t},\ldots,Y_{t_{r}-t})$
for any value of $t$. 
:::

In simple terms, the joint distribution of random variables in a strictly
stationary stochastic process is *time invariant*. For example, the
joint distribution of $(Y_{1},Y_{5},Y_{7})$ is the same as the distribution
of $(Y_{12},Y_{16},Y_{18})$. Just like in an iid sample, in a strictly
stationary process all of the individual random variables $Y_{t}$
($t=-\infty,\ldots,\infty$) have the same marginal distribution $F_{Y}$.
This means they all have the same mean, variance etc., assuming these
quantities exist. However, assuming strict stationarity does not make
any assumption about the correlations between $Y_{t},Y_{t_{1}},\ldots,Y_{t_{r}}$
other than that the correlation between $Y_{t}$ and $Y_{t_{r}}$
only depends on $t-t_{r}$ (the time between $Y_{t}$ and $Y_{t_{r}})$
and not on $t$. That is, strict stationarity allows for general temporal
dependence between the random variables in the stochastic process.


A useful property of strict stationarity is that it is preserved under
general transformations, as summarized in the following proposition.

::: {#prp-01 .proposition}
Let $\{Y_{t}\}$ be strictly stationary and let
$g(\cdot)$ be any function of the elements in $\{Y_{t}\}$. Then
$\{g(Y_{t})\}$, is also strictly stationary. 
:::

For example, if $\{Y_{t}\}$ is strictly stationary then $\{Y_{t}^{2}\}$
and $\{Y_{t}Y_{t-1}\}$ are also strictly stationary.

The following are some simple examples of strictly stationary processes.

::: {#exm-iid-sequence .example name="iid sequence"}
If $\{Y_{t}\}$ is an iid sequence\index{iid sequence}, then it is strictly stationary.
\hfill{}$\blacksquare$
:::

::: {#exm-non-iid-sequence .example name="Non iid sequence"}
Let $\{Y_{t}\}$ be an iid sequence and let $X\sim N(0,1)$ independent
of $\{Y_{t}\}$. Define $Z_{t}=Y_{t}+X$. The sequence $\{Z_{t}\}$
is not an independent sequence (because of the common $X)$ but is
an identically distributed sequence and is strictly stationary.
\hfill{}$\blacksquare$
:::

Strict stationarity places very strong restrictions on the behavior
of a time series. A related concept that imposes weaker restrictions
and is convenient for time series model building is 
*covariance stationarity*\index{Covariance stationarity} (sometimes called *weak stationarity*\index{Weak stationarity}). 

::: {#def-covariance-stationarity .definition name="Covariance stationarity"}
A stochastic process $\{Y_{t}\}$ is *covariance stationary*
if

1.  $E[Y_{t}]=\mu < \infty$ does not depend on $t$ 
2.  $\mathrm{var}(Y_{t})=\sigma^{2}<\infty$ does not depend on $t$ 
3.  $\mathrm{cov}(Y_{t},Y_{t-j})=\gamma_{j}<\infty$, and depends only on $j$ but not on $t$ for $j=0,1,2,\ldots$ 
:::

The term $\gamma_{j}$ is called the $j^{\textrm{th}}$ order *autocovariance.*\index{Autocovariance}
The $j^{\textrm{th}}$ order *autocorrelation*\index{Autocorrelation} is defined as:
$$
\rho_{j}=\frac{\mathrm{cov}(Y_{t},Y_{t-j})}{\sqrt{\mathrm{var}(Y_{t})\mathrm{var}(Y_{t-j})}}=\frac{\gamma_{j}}{\sigma^{2}}.
$$ {#eq-autocorrelation}

The autocovariances, $\gamma_{j},$ measure the direction of linear
dependence between $Y_{t}$ and $Y_{t-j}$. The autocorrelations,
$\rho_{j}$, measure both the direction and strength of linear dependence
between $Y_{t}$ and $Y_{t-j}$. With covariance stationarity, instead
of assuming the entire joint distribution of a collection of random
variables is time invariant we make a weaker assumption that only
the mean, variance and autocovariances of the random variables are
time invariant. A strictly stationary stochastic process $\{Y_{t}\}$
such that $\mu$, $\sigma^{2}$, and $\gamma_{ij}$
exist is a *covariance stationary* stochastic process. However,
a covariance stationary process need not be strictly stationary.

The autocovariances and autocorrelations are measures of the linear
temporal dependence in a covariance stationary stochastic process.
A graphical summary of this temporal dependence is given by the plot
of $\rho_{j}$ against $j$, and is called the *autocorrelation function* (ACF). Figure @fig-TimeSeriesConceptsACF illustrates
an ACF for a hypothetical covariance stationary time series with $\rho_{j}=(0.9)^{j}$
for $j=1,2,\ldots,10$ created with 

```{r}
#| label: fig-TimeSeriesConceptsACF
#| echo: true
#| fig-cap: ACF for time series with $\rho_{j}=(0.9)^{j}$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

rho = 0.9 
rhoVec = (rho)^(1:10) 
ts.plot(rhoVec, type="h", lwd=2, col="blue", xlab="Lag j", 
        ylab=expression(rho[j]))
```

For this process the strength of linear time dependence decays toward
zero geometrically fast as $j$ increases. 


The definition of covariance stationarity requires that $E[Y_{t}]<\infty$
and $\mathrm{var}(Y_{t})<\infty.$ That is, $E[Y_{t}]$ and $\mathrm{var}(Y_{t})$
must exist and be finite numbers. This is true if $Y_{t}$ is normally
distributed. However, it is not true if, for example, $Y_{t}$ has a Student's t
distribution with one degree of freedom.^[This is also called a Cauchy distribution. For this distribution $E[Y_{t}]=\mathrm{var}(Y_{t})=\mathrm{cov}(Y_{t},Y_{t-j})=\infty.$]
Hence, a strictly stationary stochastic process $\{Y_{t}\}$ where
the (marginal) pdf of $Y_{t}$ (for all $t$) has very fat tails may
not be covariance stationary. 

::: {#exm-gaussian-white-noise .example name="Gaussian White Noise"}
Let $Y_{t}\sim iid$ $N(0,\sigma^{2})$. Then $\{Y_{t}\}$ is called
a *Gaussian white noise*\index{Gaussian white noise} (GWN) process and is denoted $Y_{t}\sim\mathrm{GWN}(0,\sigma^{2})$.
Notice that: 
\begin{align*}
E[Y_{t}] & =0\textrm{ independent of }t,\\
\mathrm{var}(Y_{t}) & =\sigma^{2}\textrm{ independent of }t,\\
\mathrm{cov}(Y_{t},Y_{t-j}) & =0\text{ (for }j>0)\text{ independent of }t\textrm{ for all }j,
\end{align*}
so that $\{Y_{t}\}$ satisfies the properties of a covariance stationary
process. The defining characteristic of a GWN process is the lack
of any predictable pattern over time in the realized values of the
process. The term white noise comes from the electrical engineering
literature and represents the absence of any signal.^[As an example of white noise, think of tuning an AM radio. In between
stations there is no signal and all you hear is static. This is white
noise.]

Simulating observations from a GWN process in R is easy: just simulate
iid observations from a normal distribution. For example, to simulate
$T=250$ observations from the GWN(0,1) process use:

```{r}
#| echo: true

set.seed(123)
y = rnorm(250)
```

The simulated iid N(0,1) values are generated using the `rnorm()`
function. The command `set.seed(123)` initializes R's internal
random number generator using the seed value 123. Every time the random
number generator seed is set to a particular value, the random number
generator produces the same set of random numbers. This allows different
people to create the same set of random numbers so that results are
reproducible. The simulated data is illustrated in Figure @fig-TimeSeriesConcpetsGWN
created using:

```{r}
#| label: fig-TimeSeriesConcpetsGWN
#| echo: true
#| fig-cap: Realization of a GWN(0,1) process.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

ts.plot(y,main="Gaussian White Noise Process",xlab="time",
        ylab="y(t)", col="blue", lwd=2)
abline(h=0)
```

The function `ts.plot()` creates a time series line plot with
a dummy time index. An equivalent plot can be created using the generic
`plot()` function with optional argument `type="l"`.
The data in Figure @fig-TimeSeriesConcpetsGWN fluctuate
randomly about the mean value zero and exhibit a constant volatility
of one (typical magnitude of a fluctuation about zero). There is no
visual evidence of any predictable pattern in the data. 
\hfill{}$\blacksquare$
:::

::: {#exm-gwn-model-for-continuously-compounded-returns .example name="GWN model for continuously compounded returns"}
Let $r_{t}$ denote the continuously compounded monthly return on
Microsoft stock and assume that $r_{t}\sim\mathrm{iid}\,N(0.01,\,(0.05)^{2})$.
We can represent this distribution in terms of a GWN process as follows
$$
r_{t}=0.01+\varepsilon_{t},\,\varepsilon_{t}\sim\mathrm{GWN}(0,\,(0.05)^{2}).
$$
Here, $r_{t}$ is constructed as a $\mathrm{GWN}(0,\sigma^{2})$ process plus a constant. 
Hence, $\{r_{t}\}$ is a GWN process with a non-zero mean: $r_{t}\sim\mathrm{GWN}(0.01,\,(0.05)^{2}).$
$T=60$ simulated values of $\{r_{t}\}$ are computed using:

```{r}
#| label: fig-TimeSeriesConceptsGWNreturns
#| echo: true
#| fig-cap: Simulated returns from GWN(0.01,(0.05)$^{2})$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

set.seed(123)
y = rnorm(60, mean=0.01, sd=0.05)
ts.plot(y,main="GWN Process for Monthly Continuously Compounded Returns",
        xlab="time",ylab="r(t)", col="blue", lwd=2)
abline(h=c(0,0.01,-0.04,0.06), lwd=2, 
       lty=c("solid","solid","dotted","dotted"), 
       col=c("black", "red", "red", "red"))
```

and are illustrated in Figure @fig-TimeSeriesConceptsGWNreturns.
Notice that the returns fluctuate around the mean value of $0.01$ (solid red line)
and the size of a typical deviation from the mean is about $0.05$. The red dotted lines
show the values $0.1 \pm 0.05$ .

An implication of the GWN assumption for monthly returns is that non-overlapping
multi-period returns are also GWN. For example, consider the two-month
return $r_{t}(2)=r_{t}+r_{t-1}$. The non-overlapping process $\{r_{t}(2)\}=\{...,r_{t-2}(2),r_{t}(2),r_{t+2}(2),...\}$
is GWN with mean $E[r_{t}(2)]=2\cdot\mu=0.02$, variance $\mathrm{var}(r_{t}(2))=2\cdot\sigma^{2}=0.005$,
and standard deviation $\mathrm{sd}(r_{t}(2))=\sqrt{2}\sigma=0.071$.
\hfill{}$\blacksquare$
:::

::: {#exm-independent-white-noise .example name="Independent white noise"}
Let $Y_{t}\sim\mathrm{iid}$ $(0,\sigma^{2})$. Then $\{Y_{t}\}$
is called an *independent white noise* (IWN) process and is denoted
$Y_{t}\sim\mathrm{IWN}(0,\sigma^{2})$. The difference between GWN
and IWN is that with IWN we don't specify that all random variables
are normally distributed. The random variables can have any distribution
with mean zero and variance $\sigma^{2}.$ To illustrate, suppose
$Y_{t}=\frac{1}{\sqrt{3}}\times t_{3}$ where $t_{3}$ denotes a Student's
t distribution with $3$ degrees of freedom. This process has $E[Y_{t}]=0$
and $\mathrm{var}(Y_{t})=1.$\footnote{Recall, if $Y_{t}\sim t_{v}$ then $\mathrm{var}(Y_{t})=\frac{v}{v-2}$
for $v>2.$} Figure @fig-TimeSeriesConceptsIWN shows simulated observations
from this process created using the R commands

```{r}
#| label: fig-TimeSeriesConceptsIWN
#| echo: true
#| fig-cap: 'Simulation of IWN(0,1) process: $Y_{t}\sim\frac{1}{\sqrt{3}}\times t_{3}$'
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

set.seed(123) 
y = (1/sqrt(3))*rt(250, df=3) 
ts.plot(y, main="Independent White Noise Process", xlab="time", ylab="y(t)",         
        col="blue", lwd=2) 
abline(h=0)
```

The simulated IWN process resembles the GWN in Figure @fig-TimeSeriesConceptsIWN
but has more extreme observations. 
\hfill{}$\blacksquare$
:::

::: {#exm-weak-white-noise .example name="Weak white noise"}
Let $\{Y_{t}\}$ be a sequence of uncorrelated random variables each
with mean zero and variance $\sigma^{2}$. Then $\{Y_{t}\}$ is called
a *weak white noise*\index{Weak white noise} (WWN) process and is denoted $Y_{t}\sim\mathrm{WWN}(0,\sigma^{2})$.
With a WWN process, the random variables are not independent, only
uncorrelated. This allows for potential non-linear temporal dependence
between the random variables in the process.
\hfill{}$\blacksquare$
:::

### Non-Stationary processes
\index{Non-stationarity-process}
In a covariance stationary stochastic process it is assumed that the
means, variances and autocovariances are independent of time. In a
*non-stationary* process, one or more of these assumptions is not true.
The following examples illustrate some typical non-stationary time
series processes.

::: {#exm-deterministically-trending-process .example name="Deterministically trending process"}
Suppose $\{Y_{t}\}$ is generated according to the *deterministically trending process*: 
\begin{align*}
Y_{t} & =\beta_{0}+\beta_{1}t+\varepsilon_{t},~\varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}),\\
t & =0,1,2,\ldots
\end{align*}
Then $\{Y_{t}\}$ is nonstationary because the mean of $Y_{t}$ depends
on $t$: 
$$
E[Y_{t}]=\beta_{0}+\beta_{1}t.
$$
Figure @fig-TimeSeriesConceptsTrend shows a realization
of this process with $\beta_{0}=0,\beta_{1}=0.1$ and $\sigma_{\varepsilon}^{2}=1$
created using the R commands:

```{r}
#| label: fig-TimeSeriesConceptsTrend
#| echo: true
#| fig-cap: Deterministically trending nonstationary process $Y_{t}=0.1\times t+\varepsilon_{t},\varepsilon_{t}\sim\mathrm{GWN}(0,1)$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

set.seed(123)
e = rnorm(250)
y.dt = 0.1*seq(1,250) + e
ts.plot(y.dt, lwd=2, col="blue", main="Deterministic Trend + Noise")
abline(a=0, b=0.1)
```

Here the non-stationarity is created by the deterministic trend $\beta_{0}+\beta_{1}t$
in the data. The non-stationary process $\{Y_{t}\}$ can be transformed
into a stationary process by simply subtracting off the trend: 
$$
X_{t}=Y_{t}-\beta_{0}-\beta_{1}t=\varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}).
$$
The detrended process $X_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2})$
is obviously covariance stationary.
\hfill{}$\blacksquare$
:::

::: {#exm-random-walk .example name="Random walk"}
A random walk\index{Random walk} (RW) process $\{Y_{t}\}$ is defined as: 
\begin{align*}
Y_{t} & =Y_{t-1}+\varepsilon_{t},~\varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}),\\
 & Y_{0}\textrm{ is fixed (non-random)}.
\end{align*}
By recursive substitution starting at $t=1$, we have: 
\begin{align*}
Y_{1} & =Y_{0}+\varepsilon_{1},\\
Y_{2} & =Y_{1}+\varepsilon_{2}=Y_{0}+\varepsilon_{1}+\varepsilon_{2},\\
 & \vdots\\
Y_{t} & =Y_{t-1}+\varepsilon_{t}=Y_{0}+\varepsilon_{1}+\cdots+\varepsilon_{t}\\
 & =Y_{0}+\sum_{j=1}^{t}\varepsilon_{j}.
\end{align*}
Now, $E[Y_{t}]=Y_{0}$ which is independent of $t$. However, 
$$
\mathrm{var}(Y_{t})=\mathrm{var}\left(\sum_{j=1}^{t}\varepsilon_{j}\right)=\sum_{j=1}^{t}\sigma_{\varepsilon}^{2}=\sigma_{\varepsilon}^{2}\times t,
$$
which depends on $t$, and so $\{Y_{t}\}$ is not stationary. 

Figure @fig-TimeSeriesConceptsRW shows a realization of
the RW process with $Y_{0}=0$ and $\sigma_{\varepsilon}^{2}=1$ created
using the R commands:

```{r}
#| label: fig-TimeSeriesConceptsRW
#| echo: true
#| fig-cap: 'Random walk process: $Y_{t}=Y_{t-1}+\varepsilon_{t},\varepsilon_{t}\sim\mathrm{GWN}(0,1)$.'
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

set.seed(321)
e = rnorm(250)
y.rw = cumsum(e)
ts.plot(y.rw, lwd=2, col="blue", main="Random Walk")
abline(h=0)
```

The RW process looks much different from the GWN process in Figure
@fig-TimeSeriesConcpetsGWN. As the variance of the process
increases linearly with time, the uncertainty about where the process
will be at a given point in time increases with time. 


Although $\{Y_{t}\}$ is non-stationary, a simple first-differencing
transformation, however, yields a covariance stationary process: 
$$
X_{t}=Y_{t}-Y_{t-1}=\varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}).
$$

\hfill{}$\blacksquare$
:::

::: {#exm-random-walk-with-drift-model-for-log-stock-prices .example name="Random walk with drift model for log stock prices"}
Let $r_{t}$ denote the continuously compounded monthly return on
Microsoft stock and assume that $r_{t}\sim\mathrm{GWN}(\mu,\,\sigma^{2})$.
Since $r_{t}=\ln(P_{t}/P_{t-1})$ it follows that $\ln P_{t}=\ln P_{t-1}+r_{t}$.
Now, re-express $r_{t}$ as $r_{t}=\mu+\varepsilon_{t}$ where $\varepsilon_{t}\sim\mathrm{GWN}(0,\sigma^{2}).$
Then $lnP_{t}=lnP_{t-1}+\mu+\varepsilon_{t}.$ By recursive substitution
we have $lnP_{t}=lnP_{0}+\mu t+\sum_{t=1}^{t}\varepsilon_{t}$ and
so $\ln P_{t}$ follows a random walk process with drift value $\mu$.
Here, $E[lnP_{t}]=\mu t$ and $\mathrm{var}(lnP_{t})=\sigma^{2}t$
so $lnP_{t}$ is non-stationary because both the mean and variance
depend on $t$. In this model, prices, however, do not follow a random
walk since $P_{t}=e^{\ln P_{t}}=P_{t-1}e^{r_{t}}$.
\hfill{}$\blacksquare$
:::

### Ergodicity
\index{Ergodicity}
In a strictly stationary or covariance stationary stochastic process
no assumption is made about the strength of dependence between random
variables in the sequence. For example, in a covariance stationary
stochastic process it is possible that $\rho_{1}=\mathrm{cor}(Y_{t},Y_{t-1})=\rho_{100}=\mathrm{cor}(Y_{t},Y_{t-100})=0.5$,
say. However, in many contexts it is reasonable to assume that the
strength of dependence between random variables in a stochastic process
diminishes the farther apart they become. That is, $\rho_{1}>\rho_{2}\cdots$
and that eventually $\rho_{j}=0$ for $j$ large enough. This diminishing
dependence assumption is captured by the concept of *ergodicity*.

::: {#def-ergodicity-intuitive-definition .definition name="Ergodicity (intuitive definition)"}
Intuitively, a stochastic process $\{Y_{t}\}$ is *ergodic* if
any two collections of random variables partitioned far apart in the
sequence are essentially independent. 
:::

The formal definition of ergodicity is highly technical and requires
advanced concepts in probability theory. However, the intuitive definition
captures the essence of the concept. The stochastic process $\{Y_{t}\}$
is ergodic if $Y_{t}$ and $Y_{t-j}$ are essentially independent
if $j$ is large enough. 

If a stochastic process $\{Y_{t}\}$ is covariance stationary and
ergodic then strong restrictions are placed on the joint behavior
of the elements in the sequence and on the type of temporal dependence
allowed.

::: {#exm-white-noise-processes .example name="White noise processes"}
If $\{Y_{t}\}$ is GWN or IWN then it is both covariance stationary
and ergodic.
\hfill{}$\blacksquare$
:::

::: {#exm-covariance-stationary-but-not-ergodic-process-white-1984-page-41 .example name="Covariance stationary but not ergodic process (White 1984 page 41)"}
Let $Y_{t}\sim\mathrm{GWN}(0,1)$ and let $X\sim N(0,1)$ independent
of $\{Y_{t}\}$. Define $Z_{t}=Y_{t}+X$. Then $\{Z_{t}\}$ is covariance
stationary but not ergodic. To see why $\{Z_{t}\}$ is not ergodic,
note that for all $j>0$: 
\begin{align*}
\mathrm{var}(Z_{t}) & =\mathrm{var}(Y_{t}+X)=1+1=2,\\
\gamma_{j} & =\mathrm{cov}(Y_{t}+X,Y_{t-j}+X)=\mathrm{cov}(Y_{t},Y_{t-j})+\mathrm{cov}(Y_{t},X)+\mathrm{cov}(Y_{t-j},X)+\mathrm{cov}(X,X)\\
 & =\mathrm{cov}(X,X)=\mathrm{var}(X)=1,\\
\rho_{j} & =\frac{1}{2}\text{ for all }j.
\end{align*}
Hence, the correlation between random variables separated far apart
does not eventually go to zero and so $\{Z_{t}\}$ cannot be ergodic.
\hfill{}$\blacksquare$
:::

## Multivariate Time Series {#sec-Multivariate-Time-Series}
\index{Multivariate time series}
Consider $n$ time series variables $\{Y_{1t}\},\ldots,\{Y_{nt}\}$.
A *multivariate time series* is the $(n\times1)$ vector time
series $\{\mathbf{Y}_{t}\}$ where the $i^{th}$ row of $\{\mathbf{Y}_{t}\}$
is $\{Y_{it}\}$. That is, for any time $t$, $\mathbf{Y}_{t}=(Y_{1t},\ldots,Y_{nt})^{\prime}$.
Multivariate time series analysis is used when one wants to model
and explain the interactions and co-movements among a group of time
series variables. In finance, multivariate time series analysis is
used to model systems of asset returns, asset prices, exchange
rates, the term structure of interest rates, and economic variables, etc.. Many of the time series concepts described previously for univariate time series carry over to multivariate time series in a natural way. Additionally, there are some important time series concepts that are particular to multivariate time series. The following sections give the details of these extensions.

### Stationary and ergodic multivariate time series

A multivariate time series $\{\mathbf{Y}_{t}\}$ is covariance stationary\index{Covariance stationary}
and ergodic\index{Ergodicity} if all of its component time series are stationary and
ergodic. The mean of $\mathbf{Y}_{t}$ is defined as the $(n\times1)$
vector
$$
E[\mathbf{Y}_{t}]=(\mu_{1},\ldots,\mu_{n})^{\prime}=\boldsymbol{\mu},
$$
where $\mu_{i}=E[Y_{it}]$ for $i=1,\ldots,n$. The variance/covariance
matrix of $\mathbf{Y}_{t}$ is the $(n\times n)$ matrix
\begin{align*}
\mathrm{var}(\mathbf{Y}_{t}) & =\boldsymbol{\Sigma}=E[(\mathbf{Y}_{t}-\boldsymbol{\mu})(\mathbf{Y}_{t}-\boldsymbol{\mu})^{\prime}]\\
 & =\left(\begin{array}{cccc}
\mathrm{var}(Y_{1t}) & \mathrm{cov}(Y_{1t},Y_{2t}) & \cdots & \mathrm{cov}(Y_{1t},Y_{nt})\\
\mathrm{cov}(Y_{2t},Y_{1t}) & \mathrm{var}(Y_{2t}) & \cdots & \mathrm{cov}(Y_{2t},Y_{nt})\\
\vdots & \vdots & \ddots & \vdots\\
\mathrm{cov}(Y_{nt},Y_{1t}) & \mathrm{cov}(Y_{nt},Y_{2t}) & \cdots & \mathrm{var}(Y_{nt})
\end{array}\right).
\end{align*}
The matrix $\boldsymbol{\Sigma}$ has elements $\sigma_{ij}=$ cov$(Y_{it},Y_{jt})$ which measure
the contemporaneous linear dependence between $Y_{it}$ and $Y_{jt}$ that is time invariant.
The correlation matrix of $\mathbf{Y}_{t}$ is the $(n\times n)$
matrix
$$
\mathrm{cor}(\mathbf{Y}_{t})=\mathbf{C}=\mathbf{D}^{-1} \boldsymbol{\Sigma} \mathbf{D}^{-1},
$$
where $\mathbf{D}$ is an $(n\times n)$ diagonal matrix with $j^{th}$
diagonal element $\sigma_{j}=\mathrm{sd}(Y_{jt})$. 

#### Cross covariance and correlation matrices

For a univariate time series $\{Y_{t}\}$, the autocovariances, $\gamma_{k}$,
and autocorrelations, $\rho_{k}$, summarize the linear time dependence
in the data. With a multivariate time series $\{\mathbf{Y}_{t}\}$
each component has autocovariances and autocorrelations but there
are also cross lead-lag covariances and correlations between all possible
pairs of components. The lag $k$ autocovariances and autocorrelations of $Y_{jt}$,
for $j=1,\ldots,n$, are defined as
\begin{align*}
\gamma_{jj}^{k} & =\mathrm{cov}(Y_{jt},Y_{jt-k}),\\
\rho_{jj}^{k} & =\mathrm{corr}(Y_{jt},Y_{jt-k})=\frac{\gamma_{jj}^{k}}{\sigma_{j}^{2}},
\end{align*}
and these are symmetric in $k$: $\gamma_{jj}^{k}=\gamma_{jj}^{-k}$,
$\rho_{jj}^{k}=\rho_{jj}^{-k}$. The *cross lag-k covariances*
and *cross lag-k correlations* \index{cross lag correlation}between
$Y_{it}$ and $Y_{jt}$ are defined as
\begin{align*}
\gamma_{ij}^{k} & =\mathrm{cov}(Y_{it},Y_{jt-k}),\\
\rho_{ij}^{k} & =\mathrm{corr}(Y_{jt},Y_{jt-k})=\frac{\gamma_{ij}^{k}}{\sqrt{\sigma_{i}^{2}\sigma_{j}^{2}}},
\end{align*}
and they are not necessarily symmetric in $k$. In general, 
$$
\gamma_{ij}^{k}=\mathrm{cov}(Y_{it},Y_{jt-k})\neq\mathrm{cov}(Y_{jt},Y_{it-k})=\gamma_{ji}^{k}.
$$
If $\gamma_{ij}^{k}\neq0$ for some $k>0$ then $Y_{jt}$ is said
to *lead* $Y_{it}$. This implies that past values of $Y_{jt}$
are useful for predicting future values of $Y_{it}.$ Similarly, if
$\gamma_{ji}^{k}\neq0$ for some $k>0$ then $Y_{it}$ is said to
*lead* $Y_{jt}$. It is possible that $Y_{it}$ leads $Y_{jt}$
and vice-versa. In this case, there is said to be *dynamic feedback*
between the two series.


All of the lag $k$ cross covariances and correlations are summarized
in the $(n\times n)$ lag $k$ cross covariance and lag $k$ cross
correlation matrices
\begin{align*}
\boldsymbol{\Gamma}_{k} & =E[(\mathbf{Y}_{t}-\boldsymbol{\mu})(\mathbf{Y}_{t-k}-\boldsymbol{\mu})^{\prime}]\\
 & =\left(\begin{array}{cccc}
\mathrm{cov}(Y_{1t},Y_{1t-k}) & \mathrm{cov}(Y_{1t},Y_{2t-k}) & \cdots & \mathrm{cov}(Y_{1t},Y_{nt-k})\\
\mathrm{cov}(Y_{2t},Y_{1t-k}) & \mathrm{cov}(Y_{2t},Y_{2t-k}) & \cdots & \mathrm{cov}(Y_{2t},Y_{nt-k})\\
\vdots & \vdots & \ddots & \vdots\\
\mathrm{cov}(Y_{nt},Y_{1t-k}) & \mathrm{cov}(Y_{nt},Y_{2t-k}) & \cdots & \mathrm{cov}(Y_{nt},Y_{nt-k})
\end{array}\right),\\
\mathbf{C}_{k} & =\mathbf{D}^{-1}\boldsymbol{\Gamma}_{k}\mathbf{D}^{-1}.
\end{align*}
The matrices $\boldsymbol{\Gamma}_{k}$ and $\mathbf{C}_{k}$ are not
symmetric in $k$ but it is easy to show that $\boldsymbol{\Gamma}_{-k}=\boldsymbol{\Gamma}_{k}^{\prime}$
and $\mathbf{C}_{-k}=\mathbf{C}_{k}^{\prime}$. 

::: {#exm-multivariate-gaussian-white-noise-processes .example name="Multivariate Gaussian white noise processes"}
Let $\{\mathbf{Y}_{t}\}$ be an $n\times1$ vector time series process.
If $\mathbf{Y}_{t}\sim iid\,N(\mathbf{0},\,\boldsymbol{\Sigma})$ then
$\{\mathbf{Y}_{t}\}$ is called *multivariate Gaussian white noise*
and is denoted $\mathbf{Y}_{t}\sim\mathrm{GWN}(\mathbf{0},\,\boldsymbol{\Sigma})$.
Notice that
\begin{eqnarray*}
E[\mathbf{Y}_{t}] & = & \mathbf{0},\\
\mathrm{var}(\mathbf{Y}_{t}) & = & \boldsymbol{\Sigma},\\
\mathrm{cov}(Y_{jt},Y_{jt-k}) & = & \gamma_{jj}^{k}=0\,(\textrm{for}\,k>0)\\
\mathrm{cov}(Y_{it},Y_{jt-k}) & = & \gamma_{ij}^{k}=0\,(\textrm{for}\,k>0)
\end{eqnarray*}
Hence, the elements of $\{\mathbf{Y}_{t}\}$ are contemporaneously correlated but
exhibit no time dependence. That is,
each element of $\mathbf{Y}_{t}$ exhibits no time dependence and
there is no dynamic feedback between any two elements. Simulating
observations from $\mathrm{GWN}(\mathbf{0},\,\boldsymbol{\Sigma})$ requires
simulating from a multivariate normal distribution, which can be done
using the **mvtnorm** function `rmvnorm()`. For example,
to simulate and plot $T=250$ observation from a bivariate $\mathrm{GWN}(\mathbf{0},\,\boldsymbol{\Sigma})$
process with 
$$
\boldsymbol{\Sigma}=\left(\begin{array}{cc}
4 & 1\\
1 & 1
\end{array}\right)\Rightarrow \mathbf{C}=\left(\begin{array}{cc}
1 & 0.5\\
0.5 & 1
\end{array}\right)
$$
use:

```{r}
#| label: fig-TimeSeriesMultivariateGWN
#| echo: true
#| fig-cap: Simulated bivariate GWN process.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

library(mvtnorm) 
Sigma = matrix(c(4, 1, 1, 1), 2, 2) 
set.seed(123) 
Y = rmvnorm(250, sigma=Sigma) 
colnames(Y) = c("Y1", "Y2") 
ts.plot(Y, lwd=2, col=c("black", "blue")) 
abline(h=0)
legend("topleft", legend=c("Y1", "Y2"), 
       lwd=2, col=c("black", "blue"), lty=1)
```

The simulated values are shown on the same plot in Figure @fig-TimeSeriesMultivariateGWN.
Both series fluctuate randomly about zero, and the first series (black
line) has larger fluctuations (volatility) than the second series
(blue line). The two series are contemporaneously correlated $(\rho_{12}=0.5)$
but are both uncorrelated over time ($\rho_{11}^{k}=\rho_{22}^{k}=0,\,k>0)$
and are not cross-lag correlated ($\rho_{12}^{k}=\rho_{21}^{k}=0,\,k>0$). 
\hfill{}$\blacksquare$
:::

## Time Series Models {#sec-Time-Series-Models}

Time series models are probability models that are used to describe
the behavior of a stochastic process. In many cases of interest, it
is assumed that the stochastic process to be modeled is covariance
stationary and ergodic. Then, the main feature of the process to be
modeled is the time dependence between the random variables. In this
section, we illustrate some simple models for covariance stationary
and ergodic time series that exhibit particular types of linear time
dependence captured by autocorrelations. The univariate models, made
popular originally by [@box-1976], are called *autoregressive
moving average* (ARMA) models. The multivariate model, made popular
by [@sims-1980], is called the *vector autoregressive* (VAR) model.
These models are used extensively in economics and finance for modeling
univariate and multivariate time series.

### Moving average models

Moving average models are simple covariance stationary and ergodic
time series models built from linear functions of GWN that can capture
time dependence between random variables that lasts only for a finite
number of lags. 

#### MA(1) Model
\index{MA(1)}
Suppose you want to create a covariance stationary and ergodic stochastic
process $\{Y_{t}\}$ in which $Y_{t}$ and $Y_{t-1}$ are correlated
but $Y_{t}$ and $Y_{t-j}$ are not correlated for $j>1.$ That is,
the time dependence in the process only lasts for one period. Such
a process can be created using the *first order moving average*
(MA(1)) model: 
$$
Y_{t} =\mu+\varepsilon_{t}+\theta\varepsilon_{t-1},~-1<\theta<1, ~ \varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}). 
$$ {#eq-MA1}
The MA(1) model is a simple linear function of the GWN random variables
$\varepsilon_{t}$ and $\varepsilon_{t-1}.$ This linear structure
allows for easy analysis of the model. The moving average parameter
$\theta$ determines the sign and magnitude of the correlation between
$Y_{t}$ and $Y_{t-1}$. Clearly, if $\theta=0$ then $Y_{t}=\mu+\varepsilon_{t}$
so that $\{Y_{t}\}$ is GWN with non-zero mean $\mu$ and exhibits
no time dependence. As will be shown below, the MA(1) model produces
a covariance stationary and ergodic process for any (finite) value
of $\theta$. The restriction $-1<\theta<1$ is called the *invertibility* \index{Invertibility}
restriction and will be explained below. 


To verify that @eq-MA1 process is a covariance stationary
process we must show that the mean, variance and autocovariances are
time invariant. For the mean, we have: 
$$
E[Y_{t}]=\mu+E[\varepsilon_{t}]+\theta E[\varepsilon_{t-1}]=\mu,
$$
because $E[\varepsilon_{t}]=E[\varepsilon_{t-1}]=0$. 


For the variance, we have 
\begin{align*}
\mathrm{var}(Y_{t}) & =\sigma^{2}=E[(Y_{t}-\mu)^{2}]=E[(\varepsilon_{t}+\theta\varepsilon_{t-1})^{2}]\\
 & =E[\varepsilon_{t}^{2}]+2\theta E[\varepsilon_{t}\varepsilon_{t-1}]+\theta^{2}E[\varepsilon_{t-1}^{2}]\\
 & =\sigma_{\varepsilon}^{2}+0+\theta^{2}\sigma_{\varepsilon}^{2}=\sigma_{\varepsilon}^{2}(1+\theta^{2}).
\end{align*}
The term $E[\varepsilon_{t}\varepsilon_{t-1}]=\mathrm{cov}(\varepsilon_{t},\varepsilon_{t-1})=0$
because $\{\varepsilon_{t}\}$ is an independent process. 


For $\gamma_{1}=\mathrm{cov}(Y_{t},Y_{t-1}),$ we have: 
\begin{align*}
\mathrm{cov}(Y_{t},Y_{t-1}) & =E[(Y_{t}-\mu)(Y_{t-1}-\mu)]\\
 & =E[(\varepsilon_{t}+\theta\varepsilon_{t-1})(\varepsilon_{t-1}+\theta\varepsilon_{t-2})]\\
 & =E[\varepsilon_{t}\varepsilon_{t-1}]+\theta E[\varepsilon_{t}\varepsilon_{t-2}]\\
 & +\theta E[\varepsilon_{t-1}^{2}]+\theta^{2}E[\varepsilon_{t-1}\varepsilon_{t-2}]\\
 & =0+0+\theta\sigma_{\varepsilon}^{2}+0=\theta\sigma_{\varepsilon}^{2}.
\end{align*}
Here, the sign of $\gamma_{1}$ is the same as the sign of $\theta$.


For $\rho_{1}=\mathrm{cor}(Y_{t},Y_{t-1})$ we have: 
$$
\rho_{1}=\frac{\gamma_{1}}{\sigma^{2}}=\frac{\theta\sigma_{\varepsilon}^{2}}{\sigma_{\varepsilon}^{2}(1+\theta^{2})}=\frac{\theta}{(1+\theta^{2})}.
$$
Clearly, $\rho_{1}=0$ if $\theta=0$; $\rho_{1}>0$ if $\theta>0;\rho_{1}<0$
if $\theta<0$. Also, the largest value for $|\rho_{1}|$ is 0.5 which
occurs when $|\theta|=1$. Hence, a MA(1) model cannot describe a
stochastic process that has $|\rho_{1}|>0.5$. Also, note that there
is more than one value of $\theta$ that produces the same value of
$\rho_{1}.$ For example, $\theta$ and 1/$\theta$ give the same
value for $\rho_{1}$. The *invertibility restriction* $-1<\theta<1$
provides a unique mapping between $\theta$ and $\rho_{1}.$ 

For $\gamma_{2}=\mathrm{cov}(Y_{t},Y_{t-2}),$ we have:

\begin{align*}
\mathrm{cov}(Y_{t},Y_{t-2}) &= E[(Y_{t}-\mu)(Y_{t-2}-\mu)] \\
&= E[(\varepsilon_{t}+\theta\varepsilon_{t-1})(\varepsilon_{t-2}+\theta\varepsilon_{t-3})] \\
&= E[\varepsilon_{t}\varepsilon_{t-2}]+\theta E[\varepsilon_{t}\varepsilon_{t-3}] \\
&+ \theta E[\varepsilon_{t-1}\varepsilon_{t-2}]+\theta^{2}E[\varepsilon_{t-1}\varepsilon_{t-3}] \\
&= 0+0+0+0=0.
\end{align*}

Similar calculations can be used to show that $\mathrm{cov}(Y_{t},Y_{t-j})=\gamma_{j}=0\text{ for }j>1.$
Hence, for $j>1$ we have $\rho_{j}=0$ and there is only time dependence
between $Y_{t}$ and $Y_{t-1}$ but no time dependence between $Y_{t}$
and $Y_{t-j}$ for $j>1$. Because $\rho_{j}=0$ for $j>1$ the MA(1)
process is trivially ergodic.

::: {#exm-simulating-values-from-ma-1-process .example name="Simulating values from MA(1) process"}
Consider simulating $T=250$ observations from @eq-MA1 with
$\mu=1$, $\theta=0.9$ and $\sigma_{\varepsilon}=1$. When simulating
an MA(1) process, you need to decide how to start the simulation.
The value of $Y_{t}$ at $t=0$, $y_{0},$ is called the *initial
value* and is the starting value for the simulation. Now, the first
two observations from @eq-MA1 starting at $t=0$ are
\begin{eqnarray*}
y_{0} & = & \mu+\varepsilon_{0}+\theta\varepsilon_{-1},\\
y_{1} & = & \mu+\varepsilon_{1}+\theta\varepsilon_{0}.
\end{eqnarray*}
It is common practice is to set $\varepsilon_{-1}=\varepsilon_{0}=0$
so that $y_{0}=\mu$, $y_{1}=\mu+\varepsilon_{1}=y_{0}+\varepsilon_{1}$
and $\varepsilon_{1}$ is the first simulated error term. The remaining
observations for $t=2,\ldots,T$ are created from @eq-MA1.
We can implement the simulation in a number of ways in R. The most
straightforward way is to use a simple loop:

```{r}
#| echo: true

n.obs = 250 
mu = 1 
theta = 0.9 
sigma.e = 1 
set.seed(123) 
e = rnorm(n.obs, sd = sigma.e) 
y = rep(0, n.obs) 
y[1] = mu + e[1] 
for (i in 2:n.obs) {   
	y[i] = mu + e[i] + theta*e[i-1] 
}
head(y, n=3)
```

The "for loop" in R can be slow, however, especially for a very
large number of simulations. The simulation can be more efficiently
implemented using *vectorized* calculations as illustrated below:

```{r}
#| echo: true

set.seed(123) 
e = rnorm(n.obs, sd = sigma.e) 
em1 = c(0, e[1:(n.obs-1)]) 
y = mu + e + theta*em1 
head(y, n=3)
```

The vectorized calculation avoids looping all together and computes
all of the simulated values at the same time. This can be considerably
faster than the "for loop" calculation.


The MA(1) model is a special case of the more general *autoregressive
integrated moving average* (ARIMA) model. R has many built-in functions
and several packages for working with ARIMA models. In particular,
the R function `arima.sim()` can be used to simulate observations
from a MA(1) process. It essentially implements the simulation loop
described above. The arguments of `arima.sim()` are:

```{r}
#| echo: true

args(arima.sim)
```

where `model` is a list with named components describing the
ARIMA model parameters (excluding the mean value), `n` is the
number of simulated observations, `rand.gen` specifies the
pdf for $\varepsilon_{t},$ `innov` is a vector $\varepsilon_{t}$ values
of length `n`, `n.start` is the number of pre-simulation
(burn-in) values for $\varepsilon_{t}$, `start.innov` is a
vector of `n.start` pre-simulation values for $\varepsilon_{t}$,
and `...` specify any additional arguments for `rand.gen`.
For example, to perform the same simulations as above use:

```{r}
#| echo: true

ma1.model = list(ma=0.9)
set.seed(123)
y = mu + arima.sim(model=ma1.model, n=250, 
                   n.start=1, start.innov=0) 
head(y, n=3)
```

The `ma` component of the `"list"` object `ma1.model`
specifies the value of $\theta$ for the MA(1) model, and is used
as an input to the function `arima.sim()`. The options `n.start
= 1` and `start.innov = 0` sets the start-up initial value
$\varepsilon_{0}=0$. By default, `arima.sim()` sets $\mu=0$,
specifies $\varepsilon_{t}\sim\mathrm{GWN}(0,1)$, and returns $\varepsilon_{t}+\theta\varepsilon_{t-1}$ for $t=1,\ldots,T$. The simulated value for $Y_{t}$ is constructed by
adding the value of `mu` ($\mu=1)$ to the output of `arima.sim()`.


The function `ARMAacf()` can be used to compute the theoretical
autocorrelations, $\rho_{j},$ from the MA(1) model (recall, $\rho_{1}=\theta/(1+\theta^{2})$
and $\rho_{j}=0$ for $j>1)$. For example, to compute $\rho_{j}$
for $j=1,\ldots,10$ use:

```{r}
#| echo: true

ma1.acf = ARMAacf(ar=0, ma=0.9, lag.max=10)
ma1.acf
```

Figure @fig-TimeSeriesConceptsMA1 shows the simulated data
and the theoretical ACF created using:

```{r}
#| label: fig-TimeSeriesConceptsMA1
#| echo: true
#| fig-cap: Simulated values and theoretical ACF from MA(1) process with $\mu=1$, $\theta=0.9$
#|   and $\sigma_{\varepsilon}^{2}=1$.
#| out-width: 80%
#| fig-asp: 1
#| fig-align: center

par(mfrow=c(2,1))
ts.plot(y,main="MA(1) Process: mu=1, theta=0.9",
xlab="time",ylab="y(t)", col="blue", lwd=2)
abline(h=c(0,1))
plot(0:10, ma1.acf,type="h", col="blue", lwd=2,
main="ACF for MA(1): theta=0.9",xlab="lag",ylab="rho(j)")
abline(h=0)
par(mfrow=c(1,1))
```

Compared to the GWN process in @fig-TimeSeriesConcpetsGWN,
the MA(1) process is a bit smoother in its appearance. This is due
to the positive one-period time dependence captured by $\rho_{1}=0.4972$.
\hfill{}$\blacksquare$
:::

::: {#exm-ma-1-model-for-overlapping-continuously-compounded-returns .example name="MA(1) model for overlapping continuously compounded returns"}
Let $R_{t}$ denote the one-month continuously compounded return and
assume that: 
$$
R_{t}\sim ~\mathrm{GWN}(\mu,\sigma^{2}).
$$
Consider creating a monthly time series of two-month continuously
compounded returns using: 
$$
R_{t}(2)=R_{t}+R_{t-1}.
$$
The time series of these two-month returns, observed monthly, overlap
by one month: 
\begin{align*}
R_{t}(2) & =R_{t}+R_{t-1},\\
R_{t-1}(2) & =R_{t-1}+R_{t-2},\\
R_{t-2}(2) & =R_{t-2}+R_{t-3},\\
 & \vdots
\end{align*}
The one-month overlap in the two-month returns implies that $\{R_{t}(2)\}$
follows an MA(1) process. To show this, we need to show that the autocovariances
of $\{R_{t}(2)\}$ behave like the autocovariances of an MA(1) process.

To verify that $\{R_{t}(2)\}$ follows an MA(1) process, first we
have: 
\begin{align*}
E[R_{t}(2)] & =E[R_{t}]+E[R_{t-1}]=2\mu,\\
\mathrm{var}(R_{t}(2)) & =\mathrm{var}(R_{t}+R_{t-1})=2\sigma^{2}.
\end{align*}
Next, we have: 
$$
\mathrm{cov}(R_{t}(2),R_{t-1}(2))=\mathrm{cov}(R_{t}+R_{t-1},R_{t-1}+R_{t-2})=\mathrm{cov}(R_{t-1},R_{t-1})=\mathrm{var}(R_{t-1})=\sigma^{2},
$$
and, 
\begin{align*}
\mathrm{cov}(R_{t}(2),R_{t-2}(2)) & =\mathrm{cov}(R_{t}+R_{t-1},R_{t-2}+R_{t-3})=0,\\
\mathrm{cov}(R_{t}(2),R_{t-j}(2)) & =0\text{ for }j>1.
\end{align*}
Hence, the autocovariances of $\{R_{t}(2)\}$ are those of an MA(1)
process. 

Notice that 
$$
\rho_{1}=\frac{\sigma^{2}}{2\sigma^{2}}=\frac{1}{2}.
$$
What MA(1) process describes $\{R_{t}(2)\}$? Because $\rho_{1}=\frac{\theta}{1+\theta^{2}}=0.5$
it follows that $\theta=1$. Hence, the MA(1) process has mean $2\mu$
and $\theta=1$ and can be expressed as the MA(1) model: 
\begin{align*}
R_{t}(2) & =2\mu+\varepsilon_{t}+\varepsilon_{t-1},\\
 & \varepsilon_{t}\sim\mathrm{GWN}(0,\sigma^{2}).
\end{align*}

Notice that this is a *non-invertible* MA(1) model. 
\hfill{}$\blacksquare$
:::

#### MA(q) Model
\index{MA(q)}
The MA($q$) model has the form
$$
Y_{t}=\mu+\varepsilon_{t}+\theta_{1}\varepsilon_{t-1}+\cdots+\theta_{q}\varepsilon_{t-q}, 
$$ {#eq-ma-q}
where
$$
\varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}).
$$
The MA(q) model is stationary and ergodic provided $\theta_{1},\ldots,\theta_{q}$
are finite. The moments of the MA($q$) (see end-of-chapter exercises)
are
\begin{align*}
E[Y_{t}] & =\mu,\\
\gamma_{0} & =\sigma^{2}(1+\theta_{1}^{2}+\cdots+\theta_{q}^{2}),\\
\gamma_{j} & =\left\{ \begin{array}{c}
\left(\theta_{j}+\theta_{j+1}\theta_{1}+\theta_{j+2}\theta_{2}+\cdots+\theta_{q}\theta_{q-j}\right)\sigma^{2}\text{ for }j=1,2,\ldots,q\\
0\text{ for }j>q
\end{array}.\right.
\end{align*}
Hence, the ACF of an MA($q$) is non-zero up to lag $q$ and is zero
afterward.

::: {#exm-overlapping-returns-and-ma-q-models .example name="Overlapping returns and MA(q) models"}
MA($q$) models often arise in finance through data aggregation transformations.
For example, let $R_{t}=\ln(P_{t}/P_{t-1})$ denote the monthly continuously
compounded return on an asset with price $P_{t}$. Define the annual
return at time $t$ using monthly returns as $R_{t}(12)=\ln(P_{t}/P_{t-12})=\sum_{j=0}^{11}R_{t-j}$.
Suppose $R_{t}\sim\mathrm{GWN}(\mu,\sigma^{2})$ and consider a sample
of monthly returns of size $T$, $\{R_{1},R_{2},\ldots,R_{T}\}$.
A sample of annual returns may be created using *overlapping*
or *non-overlapping* returns. Let $\{R_{12}(12),R_{13}(12),$
$\ldots,R_{T}(12)\}$ denote a sample of $T^{\ast}=T-11$ monthly
overlapping annual returns and $\{R_{12}(12),R_{24}(12),\ldots,R_{T}(12)\}$
denote a sample of $T/12$ non-overlapping annual returns. Researchers
often use overlapping returns in analysis due to the apparent larger
sample size. One must be careful using overlapping returns because
the monthly annual return sequence $\{R_{t}(12)\}$ is not a Gaussian
white noise process even if the monthly return sequence $\{R_{t}\}$
is. To see this, straightforward calculations give:
\begin{eqnarray*}
E[R_{t}(12)] & = & 12\mu,\\
\gamma_{0} & = & \mathrm{var}(R_{t}(12))=12\sigma^{2},\\
\gamma_{j} & = & \mathrm{cov}(R_{t}(12),R_{t-j}(12))=(12-j)\sigma^{2}\text{ for }j<12,\\
\gamma_{j} & = & 0\text{ for }j\geq12.
\end{eqnarray*}
Since $\gamma_{j}=0$ for $j\geq12$ notice that $\{R_{t}(12)\}$
behaves like an MA(11) process:
\begin{eqnarray*}
R_{t}(12) & = & 12\mu+\varepsilon_{t}+\theta_{1}\varepsilon_{t-1}+\cdots+\theta_{11}\varepsilon_{t-11},\\
\varepsilon_{t} & \sim & \mathrm{GWN}(0,\sigma^{2}).
\end{eqnarray*}

\hfill{}$\blacksquare$
:::

### Autoregressive Models

Moving average models can capture almost any kind of autocorrelation
structure. However, this may require many moving average terms in
@eq-ma-q. Another type of simple time series model is the
*autoregressive model*. This model can capture complex autocorrelation
patterns with a small number of parameters and is used more often
in practice than the moving average model.

#### AR(1) Model
\index{AR(1)}
Suppose you want to create a covariance stationary and ergodic stochastic
process $\{Y_{t}\}$ in which $Y_{t}$ and $Y_{t-1}$ are correlated,
$Y_{t}$ and $Y_{t-2}$ are slightly less correlated, $Y_{t}$ and
$Y_{t-3}$ are even less correlated and eventually $Y_{t}$ and $Y_{t-j}$
are uncorrelated for $j$ large enough. That is, the time dependence
in the process decays to zero as the random variables in the process
get farther and farther apart. Such a process can be created using
the *first order autoregressive* (AR(1)) model: 
$$
Y_{t}-\mu =\phi(Y_{t-1}-\mu)+\varepsilon_{t},~-1<\phi<1,~ \varepsilon_{t}\sim\mathrm{iid} ~ N(0,\sigma_{\varepsilon}^{2}).
$$ {#eq-AR1}

It can be shown that the AR(1) model is covariance stationary and
ergodic provided $-1<\phi<1$. We will show that the AR(1) process
has the following properties: 
$$
E[Y_{t}]  =\mu,
$$ {#eq-AR1-mean}
$$
\mathrm{var}(Y_{t}) =\sigma^{2}=\sigma_{\varepsilon}^{2}/(1-\phi^{2}),
$$ {#eq-AR1-variance}
$$
\mathrm{cov}(Y_{t},Y_{t-1}) =\gamma_{1}=\sigma^{2}\phi,
$$ {#eq-AR1-gamma1}
$$
\mathrm{cor}(Y_{t},Y_{t-1}) =\rho_{1}=\gamma_{1}/\sigma^{2}=\phi,
$$ {#eq-AR1-rho1}
$$
\mathrm{cov}(Y_{t},Y_{t-j}) =\gamma_{j}=\sigma^{2}\phi^{j},
$$ {#eq-AR1-gamma-j}
$$
\mathrm{cor}(Y_{t},Y_{t-j}) =\rho_{j}=\gamma_{j}/\sigma^{2}=\phi^{j}.
$$ {#eq-AR1-rho-j}
Notice that the restriction $|\phi|<1$ implies that: 
$$
\lim_{j\rightarrow\infty}\rho_{j}=\phi^{j}=0,
$$
so that $Y_{t}$ is essentially independent of $Y_{t-j}$ for large
$j$ and so $\{Y_{t}\}$ is ergodic$.$ For example, if $\phi=0.5$
then $\rho_{10}=(0.5)^{10}=0.001$; if $\phi=0.9$ then $\rho_{10}=(0.9)^{10}=0.349$.
Hence, the closer $\phi$ is to unity the stronger is the time dependence
in the process. If $\phi=1$, then @eq-AR1 becomes the random
walk model $Y_{t}=Y_{t-1}+\varepsilon_{t}$ and is a non-stationary
process. 


Verifying covariance stationarity for the AR(1) model is more involved
than for the MA(1) model, and establishing the properties @eq-AR1-mean
- @eq-AR1-rho-j involves some tricks. In what follows, we
will assume that $\{Y_{t}\}$ is a covariance stationary process and
that $|\phi|<1.$ First, consider the derivation for @eq-AR1-mean.
We have: 
\begin{align*}
E[Y_{t}] & =\mu+\phi(E[Y_{t-1}]-\mu)+E[\varepsilon_{t}]\\
 & =\mu+\phi E[Y_{t-1}]-\phi\mu.
\end{align*}
Here, we use the first trick. Given that $\{Y_{t}\}$ is covariance stationary it follows that $E[Y_{t}]=E[Y_{t-1}].$
Substituting $E[Y_{t}]=E[Y_{t-1}]$ into the above and solving for
$E[Y_{t}]$ gives @eq-AR1-mean. 

A similar trick can be used to derive @eq-AR1-variance:
$$
\mathrm{var}(Y_{t})=\phi^{2}(\mathrm{var}(Y_{t-1}))+\mathrm{var}(\varepsilon_{t})=\phi^{2}(\mathrm{var}(Y_{t}))+\sigma_{\varepsilon}^{2},
$$
which uses the fact that $Y_{t-1}$ is independent of $\varepsilon_{t}$
(because $Y_{t-1}$ only depends on $t-1$ values) and $\mathrm{var}(Y_{t})=\mathrm{var}(Y_{t-1})$
given that $\{Y_{t}\}$ is covariance stationary. Solving for $\sigma^{2}=\mathrm{var}(Y_{t})$
gives @eq-AR1-variance. 

To determine @eq-AR1-gamma1, we use another trick. Multiply both sides of @eq-AR1 by $Y_{t-1}-\mu$ and take expectations to give: 
$$
\gamma_{1}=E\left[\left(Y_{t}-\mu\right)\left(Y_{t-1}-\mu\right)\right]=\phi E\left[(Y_{t-1}-\mu)^{2}\right]+E\left[\varepsilon_{t}\left(Y_{t-1}-\mu\right)\right]=\phi\sigma^{2},
$$
which uses the fact that $Y_{t-1}$ is independent of $\varepsilon_{t}$,
and $\mathrm{var}(Y_{t})=\mathrm{var}(Y_{t-1})=\sigma^{2}$. Finally,
to determine @eq-AR1-gamma-j, multiply both sides of @eq-AR1
by $Y_{t-j}-\mu$ and take expectations to give: 
\begin{align*}
\gamma_{j} & =E\left[\left(Y_{t}-\mu\right)\left(Y_{t-j}-\mu\right)\right]=\phi E\left[(Y_{t-1}-\mu)(Y_{t-j}-\mu)\right]+E\left[\varepsilon_{t}\left(Y_{t-j}-\mu\right)\right]\\
 & =\phi\gamma_{j-1},
\end{align*}
which uses the fact that $Y_{t-j}$ is independent of $\varepsilon_{t}$,
and $E\left[(Y_{t-1}-\mu)(Y_{t-j}-\mu)\right]=\gamma_{j-1}$ provided
$\{Y_{t}\}$ is covariance stationary. Using recursive substitution
and $\gamma_{0}=\sigma^{2}$ gives @eq-AR1-gamma-j. 

The AR(1) model @eq-AR1 is written in what is called the *mean-adjusted form*. The 
mean-adjusted form can be re-expressed in the form of a linear regression
model as follows: 
$$
\begin{aligned}
Y_{t}-\mu & =\phi(Y_{t-1}-\mu)+\varepsilon_{t}\Rightarrow\\
Y_{t} & =\mu-\phi\mu+\phi Y_{t-1}+\varepsilon_{t}\\
 & =c+\phi Y_{t-1}+\varepsilon_{t},
\end{aligned}
$$ {#eq-AR1-regression} 
where $c=(1-\phi)\mu\Rightarrow\mu=c/(1-\phi)$ is the intercept of the linear
regression. This regression model
form is convenient for estimation by ordinary least squares.

::: {#exm-simulating-values-from-ar-1-process .example name="Simulating values from AR(1) process"}
Consider simulating 250 observations from @eq-AR1 with $\mu=1$,
$\phi=0.9$ and $\sigma_{\varepsilon}=1$. To start the simulation,
an initial value or start-up value for $Y_{0}$ is required. A commonly
used initial value is the mean value $\mu$ so that $Y_{1}=\mu+\varepsilon_{1}.$
As with the MA(1) model, this can be performed using a simple "for loop" in R:

```{r}
#| echo: true

phi = 0.9 
mu = 1 
sigma.e = 1 
n.obs = 250 
y = rep(0, n.obs)
set.seed(123) 
e = rnorm(n.obs, sd=sigma.e) 
y[1] = mu + e[1]  
for (i in 2:n.obs) {
     y[i] = mu + phi*(y[i-1] - mu) + e[i] 
} 
head(y, 3)
```

Unfortunately, there is no easy way to vectorize the loop calculation.
However, the R function `filter()`, with optional argument
`method = "recursive"`, implements the AR(1) recursion
efficiently in C code and so is more efficient than the for loop code
in R above:

```{r}
#| echo: true
y = mu + stats::filter(e, phi, method="recursive") 
head(y, 3)
```

The R function `arima.sim()`, which internally uses the `filter()`
function, can also be used to simulate observations from an AR(1)
process. For the AR(1) model, the function `arima.sim()` simulates
the *components form* of the AR(1) model
\begin{eqnarray*}
Y_{t} & = & \mu+u_{t,}\\
u_{t} & = & \phi u_{t-1}+\epsilon_{t}.
\end{eqnarray*}
Hence, to replicate the "for loop" simulation with `arima.sim()`
use:

```{r}
#| echo: true

ar1.model = list(ar=0.9)
mu = 1
set.seed(123)
y = mu + arima.sim(model=ar1.model,n=250,n.start=1, start.innov=0)
head(y, 3)
```

The R function `ARMAacf()` can be used to compute the theoretical
ACF for an AR(1) model as follows

```{r}
#| echo: true

ar1.acf = ARMAacf(ar = 0.9, ma = 0, lag.max = 10)
```

The simulated AR(1) values and the ACF are shown in Figure @fig-TimeSeriesConceptsAR1.
Compared to the MA(1) process in Figure @fig-TimeSeriesConceptsMA1,
the realizations from the AR(1) process are much smoother. That is,
when $Y_{t}$ wanders high above its mean it tends to stay above the
mean for a while and when it wanders low below the mean it tends to
stay below for a while.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-TimeSeriesConceptsAR1
#| echo: false
#| fig-cap: Simulated values and ACF from AR(1) model with $\mu=1,\phi=0.9$ and $\sigma_{\varepsilon}^{2}=1$
#| out-width: 80%
#| fig-asp: 1
#| fig-align: center

par(mfrow=c(2,1))
ts.plot(y,main="AR(1) Process: mu=1, phi=0.9",
xlab="time",ylab="y(t)", col="blue", lwd=2)
abline(h=0)
abline(h=1)
# ACF for AR(1) model
plot(0:10, ar1.acf,type="h", col="blue", lwd=2,
main="ACF for AR(1): phi=0.9",xlab="lag",ylab="rho(j)")
abline(h=0)
par(mfrow=c(1,1))
```

::: {#exm-ar-1-models-and-the-speed-of-mean-reversion .example name="AR(1) models and the speed of mean reversion"}

* insert example showing AR(1) models with different values of $\rho$
* discuss the concept of mean reversion
* discuss RW model as special case of AR(1)

\hfill{}$\blacksquare$
:::

#### AR(p) Model
\index{AR(p)}
The covariance stationary AR($p$) model in mean-adjusted form is
$$
\begin{aligned}
Y_{t}-\mu & =\phi_{1}(Y_{t-1}-\mu)+\cdots+\phi_{p}(Y_{t-p}-\mu)+\varepsilon_{t},\\
 & \varepsilon_{t}\sim\mathrm{GWN}(0,\sigma_{\varepsilon}^{2}),
\end{aligned}
$$ {#eq-AR-p}
where $\mu=E[Y_{t}].$ Like the AR(1), restrictions on the autoregressive
parameters $\phi_{1},\ldots,\phi_{p}$ are required for $\{Y_{t}\}$
to be covariance stationary and ergodic. A detailed treatment of these
restrictions is beyond the scope of this book. However, one simple
necessary condition for $\{Y_{t}\}$ to be covariance stationary is
$|\phi|<1$ where $\phi=\phi_{1}+\cdots+\phi_{p}.$ Hence, in the
AR(p) model the sum of the autoregressive components $\phi$ has a
similar interpretation as the single autoregressive coefficient in
the AR(1) model.

The regression form of the AR(p) is
$$
Y_{t}=c+\phi_{1}Y_{t-1}+\cdots+\phi_{p}Y_{t-p}+\varepsilon_{t},
$$
where $c=\mu/(1-\phi_{1}-\cdots-\phi_{p})=\mu/(1-\phi).$ This form
is convenient for estimation purposes because it is in the form of
a linear regression.

The regression form of the AR(p) model is used very often in practice because of its simple
linear structure and because it can capture a wide variety of autocorrelation
patterns such as exponential decay, damped cyclical patterns, and
oscillating damped cyclical patterns. Unfortunately, the mathematical
derivation of the autocorrelations in the AR(p) model is complicated
and tedious (and beyond the scope of this book). The exercises at the end of the chapter illustrate some of the calculations for the AR(2) model.

### Autoregressive Moving Average Models
\index{ARMA}
Autoregressive and moving average models can be combined into a general
model called the *autoregressive moving average* (ARMA) model. The ARMA
model with $p$ autoregressive components and $q$ moving average
components, denoted ARMA(p,q) is given by 
$$
\begin{aligned}
Y_{t}-\mu & =\phi_{1}(Y_{t-1}-\mu)+\cdots+\phi_{p}(Y_{t-p}-\mu) \\
 & +\varepsilon_{t}+\theta_{1}\varepsilon_{t-1}+\cdots+\theta_{q}\varepsilon_{t-q}  \\
\varepsilon_{t} & \sim\mathrm{GWN}(0,\sigma^{2}) 
\end{aligned}
$$ {#eq-ARMA-pq}
The regression formulation is
$$
Y_{t}=c+\phi_{1}Y_{t-1}+\cdots+\phi_{p}Y_{t-p}+\varepsilon_{t}+\theta\varepsilon_{t-1}+\cdots+\theta\varepsilon_{t-q}
$$
where $c=\mu/(1-\phi_{1}-\cdots-\phi_{p})=\mu/(1-\phi)$ and $\phi=\phi_{1}+\cdots+\phi_{p}.$
This model combines aspects of the pure moving average models and
the pure autoregressive models and can capture many types of autocorrelation
patterns. For modeling typical non-seasonal economic and financial
data, it is seldom necessary to consider models in which $p>2$ and
$q>2$. For example, the simple ARMA(1,1) model

$$
Y_{t}-\mu = \phi_{1}(Y_{t-1}-\mu) + \varepsilon_{t}+\theta_{1}\varepsilon_{t-1} 
$$ {#eq-ARMA-11}

can capture many realistic autocorrelations observed in data.

### Vector Autoregressive Models
\index{VAR}
The most popular multivariate time series model is the \emph{vector
autoregressive} (VAR) model. The VAR model is a multivariate extension
of the univariate autoregressive model @eq-AR-p. For example,
a bivariate VAR(1) model for $\mathbf{Y}_{t}=(Y_{1t},Y_{2t})^{\prime}$
has the form
$$
\left(\begin{array}{c}
Y_{1t}\\
Y_{2t}
\end{array}\right)=\left(\begin{array}{c}
c_{1}\\
c_{2}
\end{array}\right)+\left(\begin{array}{cc}
a_{11}^{1} & a_{12}^{1}\\
a_{21}^{1} & a_{22}^{1}
\end{array}\right)\left(\begin{array}{c}
Y_{1t-1}\\
Y_{2t-1}
\end{array}\right)+\left(\begin{array}{c}
\varepsilon_{1t}\\
\varepsilon_{2t}
\end{array}\right),
$$
or
\begin{align*}
Y_{1t} & =c_{1}+a_{11}^{1}Y_{1t-1}+a_{12}^{1}Y_{2t-1}+\varepsilon_{1t},\\
Y_{2t} & =c_{2}+a_{21}^{1}Y_{1t-1}+a_{22}^{1}Y_{2t-1}+\varepsilon_{2t},
\end{align*}
where 
$$
\left(\begin{array}{c}
\varepsilon_{1t}\\
\varepsilon_{2t}
\end{array}\right)\sim\mathrm{iid}\text{ }N\left(\left(\begin{array}{c}
0\\
0
\end{array}\right),\left(\begin{array}{cc}
\sigma_{11} & \sigma_{12}\\
\sigma_{12} & \sigma_{22}
\end{array}\right)\right).
$$
In the equations for $Y_{1}$ and $Y_{2}$, the lagged values of both
$Y_{1}$ and $Y_{2}$ are present. Hence, the VAR(1) model allows
for dynamic feedback between $Y_{1}$ and $Y_{2}$ and can capture
cross-lag correlations between the variables. In matrix notation,
the model is
\begin{eqnarray*}
\mathbf{Y}_{t} & = & \mathbf{A}\mathbf{Y}_{t-1}+\boldsymbol{\varepsilon}_{t},\\
\boldsymbol{\varepsilon}_{t} & \sim & N(\mathbf{0},\,\boldsymbol{\Sigma}),
\end{eqnarray*}
where
$$
\mathbf{A}=\left(\begin{array}{cc}
a_{11}^{1} & a_{12}^{1}\\
a_{21}^{1} & a_{22}^{1}
\end{array}\right),\,\boldsymbol{\Sigma}=\left(\begin{array}{cc}
\sigma_{11} & \sigma_{12}\\
\sigma_{12} & \sigma_{22}
\end{array}\right).
$$
 

The general VAR($p$) model for $\mathbf{Y}_{t}=(Y_{1t},Y_{2t},\ldots,Y_{nt})^{\prime}$
has the form
$$
\mathbf{Y}_{t}=\mathbf{c+A}_{1}\mathbf{Y}_{t-1}+\mathbf{A}_{2}\mathbf{Y}_{t-2}+\cdots+\mathbf{A}_{p}\mathbf{Y}_{t-p}+\boldsymbol{\varepsilon}_{t},
$$
where $\mathbf{A}_{i}$ are $(n\times n)$ coefficient matrices and
$\mathbf{\varepsilon}_{t}$ is an $(n\times1)$ unobservable zero
mean white noise vector process with covariance matrix $\Sigma$.
VAR models are capable of capturing much of the complicated dynamics
observed in stationary multivariate time series.

::: {#exm-simulated-values-from-a-bivariate-var-1-process .example name="Simulated values from a bivariate VAR(1) process"}
To be completed
:::

## Forecasting

-  One of the main practical uses of time series models is for forecasting future observations
-  The existence of time dependence in a covariance stationary time series means that we can exploit this time dependence to obtain forecasts of future observations that are superior to the unconditional mean. 
-  Describe basic forecasting problem 
-  Show result that conditional mean is optimal mse forecast
-  Show chain-rule of forecasting for AR(1) processes


## Further Reading: Time Series Concepts

This chapter gives a very brief introduction to time series modeling.
More thorough treatments of time series analysis with an orientation
towards economics and finance with examples in R are given in 
[@RuppertMatteson-2015], [@tsay-2010], and [@zivot-2016]. The CRAN task
view for Time Series is an excellent summary of R packages used for
time series analysis.

## Problems: Time Series Concepts

::: {#exr-01 .exercise}
Suppose the time series $\left\{ X_{t}\right\} _{t=-\infty}^{\infty}$
is independent white noise. That is $X_{t}\sim i.i.d.\,(0,\sigma^{2})$.
Define two new time series $\left\{ Y_{t}\right\} _{t=-\infty}^{\infty}$
and $\left\{ Z_{t}\right\} _{t=-\infty}^{\infty}$, where $Y_{t}=X_{t}^{2}$
and $Z_{t}=|X_{t}|$. Are $\left\{ Y_{t}\right\} _{t=-\infty}^{\infty}$
and $\left\{ Z_{t}\right\} _{t=-\infty}^{\infty}$ also independent
white noise processes? Why or why not?
:::

::: {#exr-02 .exercise}
Realizations from four stochastic processes are
given in Figure . Which processes appear to be covariance stationary
and which processes appear to be non-stationary? Briefly justify your
answers.
:::

::: {#exr-03 .exercise}
Consider the MA(1) model
\begin{eqnarray*}
Y_{t} & = & 0.05+\varepsilon_{t}+\theta\varepsilon_{t-1},\,-1<\theta<1\\
\varepsilon_{t} & \sim & iid\,N(0,(0.10)^{2}).
\end{eqnarray*}
This process has mean $E[Y_{t}]=0.05.$

1.  Calculate $\mathrm{var}(Y_{t})$ and $\rho_{1}=\mathrm{cor}(Y_{t},Y_{t-1})$ for $\theta=0.5$ and $\theta=0.9$.
2.  Using the R function `arima.sim()`, simulate and plot 250 observations of the MA(1) process with $\theta=0.5$ and $\theta=0.9$ . Briefly comment on the behavior of the simulated data series. Does it look covariance stationary? Does it show evidence of time dependence?
:::

::: {#exr-04 .exercise}
Consider the AR(1) model
\begin{eqnarray*}
Y_{t}-0.05 & =\phi(Y_{t-1}-0.05)+\varepsilon_{t},\,-1<\phi<1\\
 & \varepsilon_{t}\sim iid\,N(0,(0.10)^{2}).
\end{eqnarray*}
This process has mean $E[Y_{t}]=0.05.$

1.  Calculate $\mathrm{var}(Y_{t})$ for $\phi=0.5$ and $\phi=0.9$.
2.  Calculate $\rho_{j}=\mathrm{cor}(Y_{t},Y_{t-j})$ for $\phi=0.5$ and $\phi=0.9$ and for $j=1,\ldots,5$.
3.  Using the R function `arima.sim()`, simulate and plot 250 observations of the AR(1) process with $\theta=0.5$ and $\theta=0.9$ . Briefly comment on the behavior of the simulated data series. Does it look covariance stationary? Does it show evidence of time dependence? How is it different from the MA(1) process
:::

::: {#exr-05 .exercise}
Figure shows a realization of a stochastic process
representing a monthly time series of overlapping 2-month continuously
compounded returns $r_{t}(2)=r_{t}+r_{t-1}$, where the 1-month continuously
compounded returns ${r_{t}}$ follow a Gaussian White noise process
with variance 1.

1.  Based on the sample autocorrelations, which time series process is most appropriate for describing the series: MA(1) or AR(1)? Justify
your answer.
2.  If you think the process is an AR(1) process, what do you think is the value of the autoregressive parameter? If you think the process is a MA(1) process, what do you think is the value of the moving average parameter?
:::

::: {#exr-06 .exercise}
Let ${Y_t}$ represent a stochastic process. Under what conditions is ${Y_t}$ covariance stationary?
:::

::: {#exr-07 .exercise}
Realizations from four stochastic processes are given in the Figures below:

```{r}
#| echo: false
set.seed(123)
y1 = rnorm(150)
y2 = 0.01*seq(1:150) + 0.1*y1
y3 = c(y1[1:75], y1[76:150]+2)
y4 = c(y1[1:75], y1[76:150]*3)

par(mfrow=c(2,2))
  ts.plot(y1, lwd=2, col="blue", main="Process 1")
  abline(h=0)
  ts.plot(y2, lwd=2, col="blue", main="Process 2")
  abline(h=0)
  ts.plot(y3, lwd=2, col="blue", main="Process 3")
  abline(h=0)
  ts.plot(y4, lwd=2, col="blue", main="Process 4")
  abline(h=0)
par(mfrow=c(1,1))
```

Which processes appear to be covariance stationary and which processes appear to be non-stationary? For those processes that you think are non-stationary, explain why the process is non-stationary. 
:::

::: {#exr-08 .exercise}
Consider the following model:
$$
Y_t = 10 - 0.67 Y_{t-1} + \epsilon_t, \,\,\,\, \epsilon \sim N(0, 1)  
$$

1. Is it stationary? Why or why not?

2. Find the mean and the variance of this process.
:::



