# Descriptive Statistics for Financial Data {#sec-Descriptive-Statistics-for-Financial-Data}

Updated: September 4, 2026

Copyright © Eric Zivot 2026

In this chapter I use graphical and numerical descriptive statistics
to study the distribution and dependence properties of daily and monthly
asset returns on two representative assets representing a single stock and a 
large diversified portfolio of stocks. The purpose of
this chapter is to introduce the techniques of exploratory data analysis
for financial time series and to document a set of stylized facts
for monthly and daily asset returns that will be used in later chapters
to motivate probability models for asset returns. 

This chapter is organized as follows. Section @sec-Univariate-Descriptive-Statistic
introduces the example data and reviews graphical and numerical descriptive
statistics for univariate data. Graphical descriptive statistics include
time plots, histograms, QQ-plots and boxplots. Numerical statistics
are the sample statistics associated with the analagous characteristics
of univariate distributions reviewed in @sec-Review-of-Random-Variables. 
Section @sec-Time-Series-Descriptive covers univariate time series descriptive 
statistics including sample autocovariances, autocorrelations and the sample 
autocorrelation function. Section @sec-Bivariate-Descriptive-Statistics presents 
bivariate descriptive statistics including scatterplots, sample covariance and 
correlation, and sample cross-lag covariances and correlations. Section @sec-Stylized-facts 
concludes with a summary of some stylized facts for monthly and daily returns. 

The R packages used in this chapter are **car**, **corrplot, IntroCompFinR,
PerformanceAnalytics, sn, tseries, zoo** and **xts**.
Make sure these packages are installed and loaded before running the
R examples in the chapter.

```{r}
#| label: libraries
#| include: false
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(methods))
suppressPackageStartupMessages(library(PerformanceAnalytics))
suppressPackageStartupMessages(library(sn))
suppressPackageStartupMessages(library(xts))
options(digits=3)
```


## Univariate Descriptive Statistics {#sec-Univariate-Descriptive-Statistic}

Let $\{R_{t}\}$ denote a univariate time series\index{Univariate time series} of asset returns
(simple or continuously compounded). Throughout this chapter I will
assume that $\{R_{t}\}$ is a covariance stationary and ergodic stochastic
process such that: 
\begin{align*}
E[R_{t}] & =\mu\textrm{ independent of }t,\\
\mathrm{var}(R_{t}) & =\sigma^{2}\textrm{ independent of }t,\\
\mathrm{cov}(R_{t},R_{t-j}) & =\gamma_{j}\textrm{ independent of }t,\\
\mathrm{corr}(R_{t},R_{t-j}) & =\gamma_{j}/\sigma^{2}=\rho_{j}\textrm{ independent of }t.
\end{align*}
In addition, I will assume that each $R_{t}$ is identically distributed
with unknown pdf $f_{R}(r)$. 


An observed sample of size $T$ of historical asset returns $\{r_{t}\}_{t=1}^{T}$
is assumed to be a realization from the stochastic process $\{R_{t}\}$
for $t=1,\ldots,T.$ That is, 
$$
\{r_{t}\}_{t=1}^{T}=\{R_{1}=r_{1},\ldots,R_{T}=r_{T}\}
$$
The goal of *exploratory data analysis*\index{Exploratory data analysis} (eda) is to use the observed
sample $\{r_{t}\}_{t=1}^{T}$ to learn about the unknown pdf $f_{R}(r)$
as well as the time dependence properties of $\{R_{t}\}$.

### Example data

I illustrate the descriptive statistical analysis of financial data
using daily and monthly adjusted closing prices on Microsoft stock
(ticker symbol `msft`) and the S\&P 500 index (ticker symbol
`^gspc`) over the period January 2, 1998 and May
31, 2012.^[An adjusted closing price is adjusted for dividend payments and stock
splits. Any dividend payment received between closing dates are added
to the close price. If a stock split occurs between the closing dates
then the all past prices are divided by the split ratio. The ticker
symbol `^gspc` refers to the actual S\&P 500 index, which
is not a tradeable security. There are several mutual funds (e.g.,
Vanguard's S\&P 500 fund with ticker VFINF) and exchange traded funds
(e.g., State Street's SPDR S\&P 500 ETF with ticker SPY) which track
the S\&P 500 index that are investable.]
These data are obtained from (finance.yahoo.com) and are available
in the R package **introcompfinr.** I first use the daily and
monthly data to illustrate descriptive statistical analysis and to
establish a number of stylized facts about the distribution and time
dependence in daily and monthly returns.

::: {#exm-example0501 name="Constructing example data"}
Daily adjusted closing price data on Microsoft and the S\&P 500 index
over the period January 4, 1993 through December 31, 2014 are available
in the **IntroCompFinR** package as the `xts` objects
`msftDailyPrices` and `sp500DailyPrices`, respectively. 

I restrict the sample period to January 2, 1998 and May 31, 2012
using: 

```{r}
#| echo: true

smpl = "1998-01::2012-05" 
msftDailyPrices = msftDailyPrices[smpl] 
sp500DailyPrices = sp500DailyPrices[smpl]
```

End-of-month prices are extracted from the daily prices using the
**xts** function `to.monthly()`:

```{r}
#| echo: true

msftMonthlyPrices = to.monthly(msftDailyPrices, OHLC=FALSE) 
sp500MonthlyPrices = to.monthly(sp500DailyPrices, OHLC=FALSE)
```

It will also be convenient to create a merged `xts` object
containing both the Microsoft and S\&P 500 index prices:

```{r}
#| echo: true

msftSp500DailyPrices = merge(msftDailyPrices, sp500DailyPrices) 
msftSp500MonthlyPrices = merge(msftMonthlyPrices, sp500MonthlyPrices)
```

I create `xts` objects containing simple returns using
the **PerformanceAnalytics** function `Return.calculate()` (and remove the first `NA` 
value with `na.omit()`):

```{r}
#| echo: true

msftMonthlyRetS = na.omit(Return.calculate(msftMonthlyPrices, 
                                           method="simple")) 
msftDailyRetS = na.omit(Return.calculate(msftDailyPrices, 
                                         method="simple")) 
sp500MonthlyRetS = na.omit(Return.calculate(sp500MonthlyPrices, 
                                            method="simple")) 
sp500DailyRetS = na.omit(Return.calculate(sp500DailyPrices, 
                                          method="simple")) 
msftSp500MonthlyRetS = na.omit(Return.calculate(msftSp500MonthlyPrices, 
                                                method="simple")) 
msftSp500DailyRetS = na.omit(Return.calculate(msftSp500DailyPrices, 
                                              method="simple"))
```

I also create `xts` objects containing monthly continuously
compounded (cc) returns:

```{r}
#| echo: true

msftMonthlyRetC = log(1 + msftMonthlyRetS) 
sp500MonthlyRetC = log(1 + sp500MonthlyRetS) 
msftSp500MonthlyRetC = merge(msftMonthlyRetC, sp500MonthlyRetC)
```

```{r}
#| echo: false
my.panel <- function(...) {
  lines(...)
  abline(h=0)
}
```

\hfill{}$\blacksquare$
:::

### Time plots

A natural graphical descriptive statistic for time series data is
a *time plot*. 

::: {#def-time-plot .definition name="Time Plot"}
A time plot is simply a line plot with the time series
data on the y-axis and the time index on the x-axis. 
:::

Time plots are useful for quickly visualizing many features of the time series data.

::: {#exm-example0502 name="Time plots of monthly prices and returns"}
A two-panel plot showing the monthly prices is given in @fig-DSmonthlyPrices,
and is created using the `plot` method for `xts` objects:

```{r}
#| label: fig-DSmonthlyPrices
#| echo: false
#| fig-cap: End-of-month closing prices on Microsoft stock and the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(msftSp500MonthlyPrices, multi.panel=TRUE, yaxis.same=FALSE,
     main="", lwd=2, col="blue")
```

The prices exhibit random walk like behavior (no tendency to revert
to a time independent mean) and appear to be non-stationary. Both
prices show two large boom-bust periods associated with the dot-com
period of the late 1990s and the run-up to the financial crisis of
2008 reflecting some common trend behavior of the two price series.

A time plot for the monthly simple returns is created using:

```{r}
#| label: fig-DSmonthlyReturns
#| echo: true
#| fig-cap: Monthly simple returns on Microsoft stock and the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(msftSp500MonthlyRetS, multi.panel=TRUE, yaxis.same=FALSE, 
     main="", lwd=2, col="blue")
```

and is given in @fig-DSmonthlyReturns. In contrast to prices, returns show 
clear mean-reverting behavior and the common monthly
mean values look to be very close to zero. Hence, the common mean
value assumption of covariance stationarity looks to be satisfied.
However, the volatility (i.e., fluctuation of returns about the mean)
of both series appears to change over time. Both series show higher
volatility over the periods 1998 - 2003 and 2008 - 2012 than over
the period 2003 - 2008. This is an indication of possible non-stationarity
in volatility.^[The returns can still be covariance stationary and exhibit time varying
*conditional* volatility. This is explored in chapter 10.] Also, the coincidence of high and low volatility periods across assets
suggests a common driver to the time varying behavior of volatility.
There does not appear to be any visual evidence of systematic time
dependence in the returns. Later on you will see that the estimated
autocorrelations are very close to zero. The returns for Microsoft
and the S\&P 500 tend to go up and down together suggesting a positive
correlation.
\hfill{}$\blacksquare$
:::

::: {#exm-example0503 name="Plotting returns on the same graph"}
In @fig-DSmonthlyReturns, the volatility of the returns
on Microsoft and the S\&P 500 looks to be similar but this is illusory.
The y-axis scale for Microsoft is much larger than the scale for the
S\&P 500 index and so the volatility of Microsoft returns is actually
much larger than the volatility of the S\&P 500 returns. @fig-DSmonthlyReturns2
shows both returns series on the same time plot created using:

```{r}
#| label: fig-DSmonthlyReturns2
#| fig-cap: Monthly simple returns for Microsoft and S\&P 500 index on the same graph.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(msftSp500MonthlyRetS, main="", multi.panel=FALSE, lwd=2, 
     col=c("red", "blue"), lty=c("dashed", "solid"), 
     legend.loc = "topright")
```

Now the higher volatility of Microsoft returns, especially before
2003, is clearly visible. However, after 2008 the volatilities of
the two series look quite similar. In general, the lower volatility
of the S\&P 500 index represents risk reduction due to holding a large
diversified portfolio.
\hfill{}$\blacksquare$
:::

::: {#exm-example0504 name="Comparing simple and continuously compounded returns"}
@fig-DScompareReturns compares the simple and continuously compounded (cc) monthly 
returns for Microsoft created using:

```{r}
#| label: fig-DScompareReturns
#| fig-cap: 'Monthly simple and cc returns on Microsoft. Top panel: simple returns; middle
#|   panel: cc returns; bottom panel: difference between simple and cc returns.'
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

retDiff = msftMonthlyRetS - msftMonthlyRetC
dataToPlot = merge(msftMonthlyRetS, msftMonthlyRetC, retDiff)
colnames(dataToPlot) = c("Simple", "cc", "Diff")
plot(dataToPlot, multi.panel=TRUE, yaxis.same=FALSE, main="", 
         lwd=2, col=c("black", "blue", "red"))
```

The top panel shows the simple returns, the middle panel shows the
cc returns, and the bottom panel shows the difference between the
simple and cc returns. Qualitatively, the simple and cc returns look
almost identical. The main differences occur when the simple returns
are large in absolute value (e.g., mid 2000, early 2001 and early
2008). In this case the difference between the simple and cc returns
can be fairly large (e.g., as large as 0.08 in mid 2000). When the
simple return is large and positive, the cc return is not quite as
large; when the simple return is large and negative, the cc return
is a larger negative number.
\hfill{}$\blacksquare$
:::

::: {#exm-example0505 name="Plotting daily returns"}
@fig-DSdailyReturns shows the daily simple returns
on Microsoft and the S\&P 500 index created with:

```{r}
#| label: fig-DSdailyReturns
#| fig-cap: Daily returns on Microsoft and the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(msftSp500DailyRetS, multi.panel=TRUE, yaxis.same=FALSE,
     main="", col="blue")
```

Compared with the monthly returns, the daily returns are closer to
zero and the magnitude of the fluctuations (volatility) is smaller.
However, the clustering of periods of high and low volatility is more
pronounced in the daily returns. As with the monthly returns, the
volatility of the daily returns on Microsoft is larger than the volatility
of the S\&P 500 returns. Also, the daily returns show some large positive and 
negative "spikes" that represent unusually large (inabsolute value) daily movements 
(e.g., Microsoft up 20\% in one day and down 15\% on another day). The monthly returns 
do not exhibit such extreme movements relative to typical volatility.
\hfill{}$\blacksquare$
:::

### Descriptive statistics for the distribution of returns

In this section, I consider graphical and numerical descriptive statistics
for the unknown marginal pdf, $f_{R}(r)$, of returns. Recall, I
assume that the observed sample $\{r_{t}\}_{t=1}^{T}$ is a realization
from a covariance stationary and ergodic time series $\{R_{t}\}$
where each $R_{t}$ is a continuous random variable with common pdf
$f_{R}(r)$. The goal is to use $\{r_{t}\}_{t=1}^{T}$ to describe
properties of $f_{R}(r)$. 

I study returns and not prices because prices are random walk non-stationary.
Sample descriptive statistics are only meaningful for covariance stationary
and ergodic time series.

#### Histograms
\index{Histogram}

:::{#def-histogram .definition name="Histogram"}
A *histogram* of returns is a graphical summary used to describe
the general shape of the unknown pdf $f_{R}(r).$ It is constructed
as follows. Order returns from smallest to largest. Divide the range
of observed values into $N$ equally sized bins. Show the number or
fraction of observations in each bin using a bar chart.
:::

::: {#exm-example0507 name="Histograms for the daily and monthly returns on Microsoft and the S&P 500 index"}
@fig-DSmsftsp500histograms shows the histograms of
the daily and monthly returns on Microsoft stock and the S\&P 500
index created using the R function `hist()`:

```{r}
#| label: fig-DSmsftsp500histograms
#| fig-cap: Histograms of monthly continuously compounded returns on Microsoft stock
#|   and S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))
hist(msftMonthlyRetS, main="", col="cornflowerblue")
hist(msftDailyRetS, main="", col="cornflowerblue")
hist(sp500MonthlyRetS, main="", col="cornflowerblue")
hist(sp500DailyRetS, main="", col="cornflowerblue")
```

All histograms have a bell-shape like the normal distribution. The
histograms of daily returns are centered around zero and those for
the monthly return are centered around values slightly larger than
zero. The bulk of the daily (monthly) returns for Microsoft and the
S\&P 500 are between -5\% and 5\% (-20\% and 20\%) and -3\% and 3\%
(-10\% and 10\%), respectively. The histogram for the S\&P 500 monthly
returns is slightly skewed left (long left tail) due to more large
negative returns than large positive returns whereas the histograms
for the other returns are roughly symmetric. 

When comparing two or more return distributions, it is useful to use
the same bins for each histogram. @fig-DSmsftsp500histogram2
shows the histograms for Microsoft and S\&P 500 returns using the
same 15 bins, created with the R code:

```{r}
#| label: fig-DSmsftsp500histogram2
#| fig-cap: Histograms for Microsoft and S\&P 500 returns using the same bins.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

msftHist = hist(msftMonthlyRetS, plot=FALSE, breaks=15)
par(mfrow=c(2,2))
hist(msftMonthlyRetS, main="", col="cornflowerblue")
hist(msftDailyRetS, main="", col="cornflowerblue",
     breaks=msftHist$breaks)
hist(sp500MonthlyRetS, main="", col="cornflowerblue", 
     breaks=msftHist$breaks)
hist(sp500DailyRetS, main="", col="cornflowerblue",
     breaks=msftHist$breaks)
```

The common bins for each histogram are specified using the optional argument 
`breaks=msftHist$breaks` in the call to `hist()`. Using the same bins for all 
histograms allows us to see more clearly that the distribution of monthly returns 
is more spread out than the distributions for daily returns, and that the distribution of S\&P
500 returns is more tightly concentrated around zero than the distribution
of Microsoft returns.
\hfill{}$\blacksquare$
:::

::: {#exm-example0508 name="Are Microsoft returns normally distributed? A first look"}
The shape of the histogram for Microsoft returns suggests that a normal
distribution might be a good candidate for the unknown distribution
of Microsoft returns. To investigate this conjecture, I simulate
random returns from a normal distribution with mean and standard deviation
calibrated to the Microsoft daily and monthly returns using:

```{r}
#| echo: true

set.seed(123)
gwnDaily = rnorm(length(msftDailyRetS), 
                 mean=mean(msftDailyRetS), sd=sd(msftDailyRetS))
gwnDaily = xts(gwnDaily, index(msftDailyRetS))
gwnMonthly = rnorm(length(msftMonthlyRetS), 
                   mean=mean(msftMonthlyRetS), sd=sd(msftMonthlyRetS))
gwnMonthly = xts(gwnMonthly, index(msftMonthlyRetS))
```

@fig-DScompareMSFTnormal shows the Microsoft monthly
returns together with the simulated normal returns.

```{r}
#| label: fig-DScompareMSFTnormal
#| echo: false
#| fig-cap: Comparison of Microsoft monthly returns with simulated normal returns with
#|   the same mean and standard deviation as the Microsoft returns.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))
plot.zoo(msftMonthlyRetS, main="Monthly Returns on MSFT",
         lwd=2, col="blue", ylim=c(-0.4, 0.4), ylab="Returns")
abline(h=0)
plot.zoo(gwnMonthly, main="Simulated Normal Returns",
         lwd=2, col="blue", ylim=c(-0.4, 0.4), ylab="Returns")
abline(h=0)
hist(msftMonthlyRetS, main="", col="cornflowerblue", 
     xlab="returns")
hist(gwnMonthly, main="", col="cornflowerblue", 
     xlab="returns", breaks=msftHist$breaks)
```

```{r}
#| label: fig-DScompareMsftNormalDaily
#| echo: false
#| fig-cap: Comparison of Microsoft daily returns with simulated normal returns with
#|   the same mean and standard deviation as the Microsoft returns.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

msftDailyHist = hist(msftDailyRetS, plot=FALSE, breaks=15)
par(mfrow=c(2,2))
plot.zoo(msftDailyRetS, main="Monthly Returns on MSFT", 
     lwd=2, col="blue", ylim=c(-0.15, 0.15), ylab="Returns")
abline(h=0)
plot.zoo(gwnDaily, main="Simulated Normal Returns",  
     lwd=2, col="blue", ylim=c(-0.15, 0.15), ylab="Returns")
abline(h=0)
hist(msftDailyRetS, main="", col="cornflowerblue", 
     xlab="returns")
hist(gwnDaily, main="", col="cornflowerblue", 
     xlab="returns", breaks=msftDailyHist$breaks)
```

The simulated normal returns shares many of the same features as the
Microsoft returns: both fluctuate randomly about zero. However, there
are some important differences. In particular, the volatility of Microsoft
returns appears to change over time (large before 2003, small between
2003 and 2008, and large again after 2008) whereas the simulated returns
has constant volatility. Additionally, the distribution of Microsoft
returns has fatter tails (more extreme large and small returns) than
the simulated normal returns. Apart from these features, the simulated
normal returns look remarkably like the Microsoft monthly returns.

@fig-DScompareMsftNormalDaily shows the Microsoft
daily returns together with the simulated normal returns. The daily
returns look much less like GWN than the monthly returns. Here the
constant volatility of simulated GWN does not match the volatility
patterns of the Microsoft daily returns, and the tails of the histogram
for the Microsoft returns are "fatter" than the tails
of the GWN histogram.
\hfill{}$\blacksquare$
:::

#### Smoothed histogram

Histograms give a good visual representation of the data distribution.
The shape of the histogram, however, depends on the number of bins
used. With a small number of bins, the histogram often appears blocky
and fine details of the distribution are not revealed. With a large
number of bins, the histogram might have many bins with very few observations.
The `hist()` function in R smartly chooses the number of bins
so that the resulting histogram typically looks good. 

The main drawback of the histogram as a descriptive statistic for the
underlying pdf of the data is that it is discontinuous. If it is believed
that the underlying pdf is continuous, it is desirable to have a continuous
graphical summary of the pdf. The *smoothed histogram*\index{Smoothed histogram} achieves
this goal. Given a sample of data $\{x_{t}\}_{t=1}^{T}$ the R function
`density()` computes a smoothed estimate of the underlying
pdf $f_X(x)$ at each point $x$ in the bins of the histogram using the formula:
$$
\hat{f}_{X}(x)=\frac{1}{Tb}\sum_{t=1}^{T}k\left(\frac{x-x_{t}}{b}\right),
$$
where $k(\cdot)$ is a continuous smoothing function (typically a
standard normal distribution) and $b$ is a bandwidth (or bin-width)
parameter that determines the width of the bin around $x$ in which
the smoothing takes place. The resulting pdf estimate $\hat{f}_{X}(x)$
is a two-sided weighted average of the histogram values around $x$.^[In this book, a "hat" on top of a symbol denotes that it is an estimate the underlying symbol.]

::: {#exm-example0509 name="Smoothed histogram for Microsoft monthly returns"}
@fig-DSsmoothedHistogramMsft shows the histogram of
Microsoft returns overlaid with the smoothed histogram created using:

```{r}
#| label: fig-DSsmoothedHistogramMsft
#| fig-cap: Histogram and smoothed density estimate for the monthly returns on Microsoft.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
par(mfrow=c(1,1))
MSFT.density = density(msftMonthlyRetS)
hist(msftMonthlyRetS, main="", xlab="Microsoft Monthly Returns",
     col="cornflowerblue", probability=TRUE, ylim=c(0,5.5))
points(MSFT.density,type="l", col="orange", lwd=2)
```

In @fig-DSsmoothedHistogramMsft, the histogram is
normalized (using the argument \texttt{probability=TRUE}), so that
its total area is equal to one. The smoothed density estimate transforms
the blocky shape of the histogram into a smooth continuous graph.
\hfill{}$\blacksquare$
:::

#### Empirical CDF

Recall, the CDF of a random variable $X$ is the function $F_{X}(x)=\Pr(X\leq x).$ The empirical CDF $\hat{F}_{X}(x)$ is the sample analogue of $F_{X}(x)$.

::: {#def-empirical-cdf .definition name="Empirical CDF"}
The *empirical CDF*\index{Empirical CDF} of a data sample $\{x_{t}\}_{t=1}^{T}$ is
the function that counts the fraction of observations less than or
equal to $x$: 
$$
\begin{aligned}
\hat{F}_{X}(x) & =\frac{1}{T}(\#x_{i}\leq x)\\
 & =\frac{\textrm{number of }x_{i}\textrm{ values }\leq x}{\textrm{sample size}}
\end{aligned} 
$$ {#eq-empiricalCDF}
:::

::: {#exm-example0510 name="Empirical CDF for monthly returns on Microsoft"}
Computing the empirical CDF of Microsoft monthly returns at a given
point, say $r=0$, is a straightforward calculation in R:

```{r}
#| echo: true
Fhat.0 = sum(msftMonthlyRetS <= 0)/nrow(msftMonthlyRetS) 
Fhat.0
```

Here, the expression `msftMonthlyRetS <= 0` creates
a logical vector the same length as `msftMonthlyRetS` that
is equal to `TRUE` when returns are less than or equal to 0
and `FALSE` when returns are greater than zero. Then `Fhat.0`
is equal to the fraction of `TRUE` values (returns less than
or equal to zero) in the data, which gives @eq-empiricalCDF
evaluated at zero. The R function `ecdf()` can be used to compute
and plot @eq-empiricalCDF for all of the observed returns.
For example, @fig-DSecdfMSFT shows the empirical CDF
for Microsoft monthly returns computed using:

```{r}
#| label: fig-DSecdfMSFT
#| fig-cap: Empirical CDF of monthly returns on Microsoft
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(ecdf(coredata(msftMonthlyRetS)), main="", col="blue")
```

The "S" shape of the empirical CDF of Microsoft monthly returns is due to the symmetric 
shape of the histogram of monthly returns and resembles the CDF of a normal distribution.
\hfill{}$\blacksquare$
:::

#### Empirical quantiles/percentiles

Recall, for $\alpha\in(0,1)$ the $\alpha\times100\%$ quantile of
the distribution of a continuous random variable $X$ with CDF $F_{X}$
is the point $q_{\alpha}^{X}$ such that $F_{X}(q_{\alpha}^{X})=\Pr(X\leq q_{\alpha}^{X})=\alpha$.
The *empirical quantile* is defined similarly using the empirical CDF of the data.

::: {#def-empirical-quantile .definition name="Empirical Quantile"}
The $\alpha\times100\%$ *empirical quantile* \index{Empirical quantile}(or $100\times\alpha^{\textrm{th}}$ *percentile*) of a data sample
$\{x_{t}\}_{t=1}^{T}$ is the data value $\hat{q}_{\alpha}$ such
that $\alpha\cdot100\%$ of the data are less than or equal to $\hat{q}_{\alpha}$.
:::

Empirical quantiles can be easily determined by ordering the data
from smallest to largest giving the ordered sample (also known as *order statistics*): 
$$
x_{(1)}<x_{(2)}<\cdots<x_{(T)}.
$$
The empirical quantile $\hat{q}_{\alpha}$ is the order statistic
closest to $\alpha\times T$.^[There is no unique way to determine the empirical quantile from a sample of size $N$ for all values of $\alpha$. The R function `quantile()` can compute empirical quantile using one of seven different definitions.]

The *empirical quartiles*\index{Empirical quartiles} are the empirical quantiles for $\alpha=0.25,0.5$
and $0.75,$ respectively. The second empirical quartile $\hat{q}_{.50}$
is called the *sample median* and is the data point such that
half of the data is less than or equal to its value. 
The *interquartile range* \index{Interquartile range} (IQR) is the difference between the 3rd and 1st quartile: 
$$
\mathrm{IQR}=\hat{q}_{.75}-\hat{q}_{.25},
$$
and shows the size of the middle of the data distribution.

::: {#exm-example0511 name="Empirical quantiles of the Microsoft and S&P 500 monthly returns"}
The R function `quantile()` computes empirical quantiles for
a single data series. By default, `quantile()` returns the
empirical quartiles as well as the minimum and maximum values:

```{r}
#| echo: true

apply(msftSp500MonthlyRetS, 2, quantile)
```
In the above code, the R function `apply()` is used to compute the empirical quantiles on each column of the `xts` object `msftSp500MonthlyRetS`. The syntax for `apply()` is:
```{r}
args(apply)
```
where `X` is a multi-column data object, `MARGIN` specifies looping over rows (`MARGIN=1`) or columns (`MARGIN=2`), `FUN` specifies the R function to be applied, and `...` represents any optional arguments to be passed to `FUN`.

Here you see that the lower (upper) quartiles of the Microsoft monthly returns are smaller (larger) than the respective quartiles for the S\&P 500 index. 

To compute quantiles for a specified $\alpha$ use the `probs`
argument. For example, to compute the 1\% and 5\% quantiles of the
monthly returns use:

```{r}
#| echo: true

apply(msftSp500MonthlyRetS, 2, quantile, probs=c(0.01,0.05))
```
Historically, there is a $1\%$ chance of losing more than $19\%$ ($12\%$) on Microsoft
(S\&P 500) in a given month, and a $5\%$ chance of losing more than $14\%$ ($8\%$) on Microsoft (S\&P 500) in a given month.

To compute the median and IQR values for monthly returns use the R
functions `median()` and `IQR()`, respectively:

```{r}
#| echo: true

med = apply(msftSp500MonthlyRetS, 2, median)
iqr = apply(msftSp500MonthlyRetS, 2, IQR)
rbind(med, iqr)
```
The median returns are similar (about 0.8\% per month) but the IQR
for Microsoft is about twice as large as the IQR for the S\&P 500
index.
\hfill{}$\blacksquare$
:::

#### Empirical/Historical VaR

Recall, the $\alpha\times100\%$ value-at-risk (VaR) of an investment
of $\$W$ is $\mathrm{VaR}_{\alpha}=-\$W\times q_{\alpha}^{R},$ where
$q_{\alpha}^{R}$ is the $\alpha\times100\%$ quantile of the probability
distribution of the investment simple rate of return $R$ (recall, you multiply by $-1$ to represent loss as a positive number). The *empirical VaR* or *historical VaR* is defined similarly using the *empirical quantile*. 

::: {#def-empirical-historical-var .definition name="Empirical/Historical VaR"}
The $\alpha\times100\%$
*empirical VaR*\index{Historical VaR} (sometimes called *historical VaR* or *historical simulation VaR*) of an investment of $\$W$ is defined as: 
$$
\mathrm{VaR}_{\alpha}^{HS}=-\$W\times\hat{q}_{\alpha}^{R},
$$ {#eq-empiricalVaR}
where $\hat{q}_{\alpha}^{R}$ is the empirical $\alpha$ quantile
of a sample of simple returns $\{R_{t}\}_{t=1}^{T}$. For a sample
of continuously compounded returns $\{r_{t}\}_{t=1}^{T}$ with empirical
$\alpha$ quantile $\hat{q}_{\alpha}^{r}$, 
$$
\mathrm{VaR}_{\alpha}^{HS}=-\$W\times\left(\exp(\hat{q}_{\alpha}^{r})-1\right).
$$ {#eq-empiricalVaRcc}
:::

Historical VaR is based on the distribution of the observed returns
and not on any assumed distribution for returns (e.g., the normal
distribution or the log-normal distribution).

::: {#exm-example0512 name="Using empirical quantiles to compute historical VaR"}
Consider investing $W=\$100,000$ in Microsoft and the S\&P 500 over
a month. The 1\% and 5\% historical VaR values for these investments
based on the historical samples of monthly returns are:

```{r}
#| echo: true

W = 100000
msftQuantiles = quantile(msftMonthlyRetS, probs=c(0.01, 0.05))
sp500Quantiles = quantile(sp500MonthlyRetS, probs=c(0.01, 0.05))
msftVaR = -W*msftQuantiles
sp500VaR = -W*sp500Quantiles
cbind(msftVaR, sp500VaR)
```
Based on the empirical distribution of the monthly returns, a \$100,000
monthly investment in Microsoft will lose \$13,694 or more with 5\% probability and will lose \$18,929 or more with 1\% probability. The corresponding values for the S\&P
500 are \$8,183 and \$12,039, respectively. The historical VaR values for the S\&P 500 are considerably
smaller than those for Microsoft. In this sense, investing in Microsoft
is riskier than investing in the S\&P 500 index.
\hfill{}$\blacksquare$
:::

### QQ-plots
\index{QQ-plots}
Often it is of interest to see if a given data sample could be viewed
as a random sample from a specified probability distribution. One
easy and effective way to do this is to compare the empirical quantiles
of a data sample to the quantiles from a reference probability distribution.
If the quantiles match up, then this provides strong evidence that
the reference distribution is appropriate for describing the distribution
of the observed data. If the quantiles do no,t match up, then the observed
differences between the empirical quantiles and the reference quantiles
can be used to determine a more appropriate reference distribution.
It is common to use the normal distribution as the reference distribution,
but any distribution can in principle be used. 

:::{#def-quantile-quantile-plot .definition name="Quantile-Quantile Plot"}
The *quantile-quantile plot* (QQ-plot) gives a graphical comparison
of the empirical quantiles of a data sample to those from a specified
reference distribution. The QQ-plot is an xy-plot with the reference
distribution quantiles on the x-axis and the empirical quantiles on
the y-axis. If the quantiles exactly match up then the QQ-plot is
a straight line. If the quantiles do not match up, then the shape
of the QQ-plot indicates which features of the data are not captured
by the reference distribution.
:::

```{r}
#| echo: false
# compute simulated data from skewed and fat tailed distributions
# Data for student t with 3 df: tails fatter than normal
set.seed(123)
n.sim = 500
tdata = rt(n.sim,df=3)		# Student-t with 3 df
gdata = rnorm(n.sim)			# N(0,1) data
xx = seq(from=-5,to=5,length=100)
# data for log-normal: asymmetric distribution
lndata = rlnorm(n.sim)
yy = seq(from=-3, to = 3, length=100)
# data for skew normal: asymmetric distribution
set.seed(321)
rightSkewData = rsn(n.sim, alpha = 2)
set.seed(123)
leftSkewData = rsn(n.sim, alpha = -2)
```

```{r}
#| label: fig-DSqqplotfat
#| echo: false
#| fig-cap: Normal QQ-Plot for Fat-tailed Data
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(1,2))
# densities
	plot(xx,dnorm(xx),type="l", lwd=2,
	     main="Normal and Student-t with 3 df", xlab = "z, t", ylab = "pdf", col="cornflowerblue")
	points(xx,dt(xx,df=3), type="l", col="orange", lwd=3)
	legend(x="topright", legend=c("Normal","Student-t"), lty=c(1,1), col=c("cornflowerblue","orange"),
	       lwd=c(2,3))
	# QQ-plot
	qqnorm(tdata, col="cornflowerblue")
	qqline(tdata)
par(mfrow=c(1,1))
```

::: {#exm-example0512b name="QQ-plots for non-normal data"}
@fig-DSqqplotfat shows a normal QQ-plot for data simulated from a Student's t
distribution with 3 degrees-of-freedom which has fatter tails than the normal distribution.
Notice how the normal QQ-plot deviates from the straight line for both large and small quantiles of the normal
distribution. This S-shape tells us that both extremely small and extremely large empirical quantiles (on the vertical axis) are larger (in absolute value) than the corresponding theoretical quantiles of the normal distribution. 
:::

```{r}
#| label: fig-DSqqplotrightskew
#| echo: false
#| fig-cap: Normal QQ-Plot for Positively Skewed Data
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(1,2))
 # 2nd plot
	plot(yy,dnorm(yy,sd=1),type="l", lwd=2, ylim=c(0,0.7),
	     main="Normal and Positive Skew", xlab = "z, skew-z", ylab = "pdf",
	     col="cornflowerblue")
	points(yy,dsn(yy, alpha=2), type="l", lwd=3, col="orange")
	legend(x="topleft", legend=c("Normal","Positive Skew"), lty=c(1,1), col=c("cornflowerblue","orange"),
	       lwd=c(2,3))
 	# QQ-plots
	qqnorm(rightSkewData, col="cornflowerblue")
	qqline(rightSkewData)
  
par(mfrow=c(1,1))
```

@fig-DSqqplotrightskew shows a normal QQ-plot for data simulated from a distribution with positive
skewness which has more density in the right tail than in the left tail.
Here the normal QQ-plot has a convex shape relative to the straight line indicating that the lower (upper) empirical quantiles are closer to (farther from) zero than the corresponding theoretical quantiles of the normal distribution. 

```{r}
#| label: fig-DSqqplotleftskew
#| echo: false
#| fig-cap: Normal QQ-Plot for Negatively Skewed Data
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(1,2))
# densities
  plot(yy,dnorm(yy,sd=1),type="l", lwd=2, ylim=c(0,0.7),
       main="Normal and Negative Skew", xlab = "z, skew-z", ylab = "pdf", col="cornflowerblue")
  points(yy,dsn(yy, alpha=-2), type="l", lwd=3, col="orange")
  legend(x="topright", legend=c("Normal","Negative Skew"), lty=c(1,1), col=c("cornflowerblue","orange"),
         lwd=c(2,3))
  # QQ-plots
  qqnorm(leftSkewData, col="cornflowerblue")
  qqline(leftSkewData)
par(mfrow=c(1,1))
```

@fig-DSqqplotleftskew shows a normal QQ-plot for data simulated from a distribution 
with negative skewness which has more density in the left tail than in the right tail.
Here the normal QQ-plot has a concave shape relative to the straight line indicating that 
the lower (upper) empirical quantiles are farther from (closer to) zero than the corresponding 
theoretical quantiles of the normal distribution. 
\hfill{}$\blacksquare$
:::

::: {#exm-example0513 name="Normal QQ-plots for GWN Microsoft and S\&P 500 returns"}
The R function `qqnorm()` creates a QQ-plot for a data sample
using the normal distribution as the reference distribution. @fig-DSqqplots shows 
normal QQ-plots for the simulated GWN data, Microsoft returns and S\&P 500 returns.

```{r}
#| label: fig-DSqqplots
#| echo: false
#| fig-cap: Normal QQ-plots for GWN, Microsoft returns and S&P 500 returns.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,3))
qqnorm(gwnMonthly, main="GWN Monthly", col="cornflowerblue")
qqline(gwnMonthly)
qqnorm(msftMonthlyRetS, main="MSFT Monthly Returns", 
       col="cornflowerblue")
qqline(msftMonthlyRetS)
qqnorm(sp500MonthlyRetS, main="SP500 Monthly Returns", 
       col="cornflowerblue")
qqline(sp500MonthlyRetS)
qqnorm(gwnDaily, main="GWN Daily", col="cornflowerblue")
qqline(gwnDaily)
qqnorm(msftDailyRetS, main="MSFT Daily Returns", 
       col="cornflowerblue")
qqline(msftDailyRetS)
qqnorm(sp500DailyRetS, main="SP500 Daily Returns", 
       col="cornflowerblue")
qqline(sp500DailyRetS)
```

The normal QQ-plot for the simulated GWN data is very close to a straight
line, as it should be since the data are simulated from a normal 
distribution.^[Even with data simulated from a normal distribution, the empirical quantiles
in the extreme left and right tails can deviate a bit from the straight line due to 
estimation error. For this reason, small deviations from linearity in the tails should be ignored.]
The `qqline()` function draws a straight line through the points
to help determine if the quantiles match up. The normal QQ-plots for
the Microsoft and S\&P 500 monthly returns are linear in the middle
of the distribution but deviate from linearity in the tails of the
distribution. In the normal QQ-plot for Microsoft monthly returns,
the theoretical normal quantiles on the x-axis are too small in both
the left and right tails because the points fall below the straight
line in the left tail and fall above the straight line in the right
tail. Hence, the normal distribution does not match the empirical
distribution of Microsoft monthly returns in the extreme tails of
the distribution. In other words, the Microsoft returns have *fatter tails* than 
the normal distribution. For the S\&P 500 monthly returns,
the theoretical normal quantiles are too small only for the left tail
of the empirical distribution of returns (points fall below the straight
line in the left tail only). This reflects the long left tail (negative
skewness) of the empirical distribution of S\&P 500 returns. The normal
QQ-plots for the daily returns on Microsoft and the S\&P 500 index
both exhibit a pronounced tilted S-shape with extreme departures from
linearity in the left and right tails of the distributions. The daily
returns clearly have much fatter tails than the normal distribution.
\hfill{}$\blacksquare$
:::

::: {#exm-example0514 name="Student's t QQ-plot for Microsoft returns"}
The function `qqPlot()` from the **car** package or the
function `chart.QQPlot()` from the **PerformanceAnalytics**
package can be used to create a QQ-plot against any reference distribution
that has a corresponding quantile function implemented in R. For example,
a QQ-plot for the Microsoft returns using a Student's t reference
distribution with 5 degrees of freedom can be created using:

```{r}
#| label: fig-DSqqplotStudent
#| cache: true
#| fig-cap: QQ-plot of Microsoft returns using Student's t distribution with 5 degrees
#|   of freedom as the reference distribution.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

qqPlot(coredata(msftMonthlyRetS), distribution="t", 
       df=5, ylab="MSFT quantiles", envelope=FALSE, id=FALSE)
```

In the function `qqPlot()`, the argument `distribution= "t"`
specifies that the quantiles are to be computed using the R function
`qt()`.^[The `coredata()` function is used to extract the data as a
`"numeric"` object because `qqPlot()`
does not work correctly with `"zoo"` or `xts` objects.] @fig-DSqqplotStudent 
shows the resulting graph. Here, with a reference distribution with fatter tails 
than the normal distribution the QQ-plot for Microsoft returns is closer to a straight line. This
indicates that the Student's t distribution with 5 degrees of freedom
is a better reference distribution for Microsoft returns than the
normal distribution.
\hfill{}$\blacksquare$
:::

### Shape characteristics of the empirical distribution

Recall, for a random variable $X$ the shape measures of center, spread,
asymmetry and tail thickness of the pdf are: 
\begin{align*}
\textrm{center} & :\mu_{X}=E[X],\\
\text{spread} & :\sigma_{X}^{2}=\mathrm{var}(X)=E[(X-\mu_{X})^{2}],\\
\text{spread} & :\sigma_{X}=\sqrt{\mathrm{var}(X)}\\
\text{asymmetry} & :\mathrm{skew}_{X}=E[(X-\mu_{X})^{3}]/\sigma^{3},\\
\text{tail thickness} & :\mathrm{kurt}_{X}=E[(X-\mu_{X})^{4}]/\sigma^{4}.
\end{align*}

:::{#def-sample-statistics .definition name="Sample Statistics"}
The shape measures for the empirical distribution (e.g.,
as measured by the histogram) of a data sample $\{x_{t}\}_{t=1}^{T}$
are the *sample statistics*: 
$$
\hat{\mu}_{x} =\bar{x}=\frac{1}{T}\sum_{t=1}^{T}x_{t},
$$ {#eq-sample-mean}
$$
\hat{\sigma}_{x}^{2} =s_{x}^{2}=\frac{1}{T-1}\sum_{t=1}^{T}(x_{t}-\bar{x})^{2},
$$ {#eq-sample-variance}
$$
\hat{\sigma}_{x}  =\sqrt{\hat{\sigma}_{x}^{2}},
$$ {#eq-sample-sd}
$$
\widehat{\mathrm{skew}}_{x} =\frac{\frac{1}{T-1}\sum_{t=1}^{T}(x_{t}-\bar{x})^{3}}{s_{x}^{3}},
$$ {#eq-sample-skewness}
$$
\widehat{\mathrm{kurt}}_{x} =\frac{\frac{1}{T-1}\sum_{t=1}^{T}(x_{t}-\bar{x})^{4}}{s_{x}^{4}}.
$$ {#eq-sample-kurtosis}
:::

The sample mean, $\hat{\mu}_{X}$, measures the center of the histogram;
the sample standard deviation, $\hat{\sigma}_{x}$, measures the spread
of the data about the mean in the same units as the data; the sample
skewness\index{Skewness}, $\widehat{\mathrm{skew}}_{x}$, measures the asymmetry of
the histogram; the sample kurtosis\index{Kurtosis}, $\widehat{\mathrm{kurt}}_{x}$,
measures the tail-thickness of the histogram. The sample excess kurtosis\index{Excess kurtosis},
defined as the sample kurtosis minus 3: 
$$
\widehat{\mathrm{ekurt}}_{x}=\widehat{\mathrm{kurt}}_{x}-3,
$$ {#eq-sample-ekurt}
measures the tail thickness of the data sample relative to that of
a normal distribution.^[Values with hats, `"^"`, denote
sample estimates of the corresponding population quantity. For example,
the sample mean $\hat{\mu}_{x}$ is the sample estimate of the population
expected value $\mu_{X}$.] 

Notice that the divisor in @eq-sample-variance - @eq-sample-kurtosis
is $T-1$ and not $T$. This is called a *degrees-of-freedom*\index{Degree of freedom}
correction. In computing the sample variance, skewness and kurtosis,
one degree-of-freedom in the sample is used up in the computation
of the sample mean so that there are effectively only $T-1$ observations
available to compute the statistics.^[If there is only one observation in the sample then it is impossible
to create a measure of spread in the sample. You need at least two
observations to measure deviations from the sample average. Hence
the effective sample size for computing the sample variance is $T-1$.]

::: {#exm-example0515 name="Sample shape statistics for the returns on Microsoft and S&P 500"}
The R functions for computing @eq-sample-mean - @eq-sample-kurtosis
are `mean()`, `var()` and `sd()`, respectively.
There are no functions for computing @eq-sample-skewness and
@eq-sample-kurtosis in base R. The functions `skewness()`
and `kurtosis()` in the **PerformanceAnalytics** package
compute the sample skewness @eq-sample-skewness and the sample excess kurtosis
@eq-sample-ekurt, respectively.^[Similar functions are available in the **moments** package.]
The sample statistics for the Microsoft and S\&P 500 monthly returns
are computed using:

```{r}
#| echo: true

statsMonthly = rbind(apply(msftSp500MonthlyRetS, 2, mean),
                     apply(msftSp500MonthlyRetS, 2, var),
                     apply(msftSp500MonthlyRetS, 2, sd),
                     apply(msftSp500MonthlyRetS, 2, skewness),
                     apply(msftSp500MonthlyRetS, 2, kurtosis))
rownames(statsMonthly) = c("Mean", "Variance", "Std Dev",
                           "Skewness",  "Excess Kurtosis")
round(statsMonthly, digits=4)
```

The mean and standard deviation for Microsoft monthly returns are
0.9\% and 10\%, respectively. Annualized, these values are approximately
10.8\% (.009 $\times12$) and 34.6\% (.10 $\times$ $\sqrt{12}$),
respectively. The corresponding monthly and annualized values for
S\&P 500 returns are .3\% and 4.7\%, and 4.8\% and 16.6\%, respectively.
Microsoft has a higher mean and volatility than S\&P 500. The lower
volatility for the S\&P 500 reflects risk reduction due to *diversification* (holding many assets in a portfolio).
The sample skewness for Microsoft, 0.485, is slightly positive and
reflects the approximate symmetry in the histogram in @fig-DSmsftsp500histograms.
The skewness for S\&P 500, however, is moderately negative at -0.563
which reflects the somewhat long left tail of the histogram in @fig-DSmsftsp500histograms. 
The sample excess kurtosis values for Microsoft and S\&P 500 are 1.915 and 0.654, respectively, and
indicate that the tails of the histograms are slightly fatter than
the tails of a normal distribution. 

The sample statistics for the daily returns are:

```{r}
#| echo: true

statsDaily = rbind(apply(msftSp500DailyRetS, 2, mean),
                   apply(msftSp500DailyRetS, 2, var),
                   apply(msftSp500DailyRetS, 2, sd),
                   apply(msftSp500DailyRetS, 2, skewness),
                   apply(msftSp500DailyRetS, 2, kurtosis))
rownames(statsDaily) = rownames(statsMonthly)
round(statsDaily, digits=4)
```

For a daily horizon, the sample mean for both series is zero (to three
decimals). As with the monthly returns, the sample standard deviation
for Microsoft, 2.2\%, is about twice as big as the sample standard
deviation for the S\&P 500 index, 1.4\%. Neither daily return series
exhibits much skewness, but the excess kurtosis values are much larger
than the corresponding monthly values indicating that the daily returns
have much fatter tails and are more non-normally distributed than
the monthly return series. 

The daily simple returns are very close to the daily cc returns so
the square-root-of-time rule (with 20 trading days within the month)
can be applied to the sample mean and standard deviation of the simple
returns:

```{r}
#| echo: true

20*statsDaily["Mean", ]
sqrt(20)*statsDaily["Std Dev", ]
```

Comparing these values to the corresponding monthly sample statistics
shows that the square-root-of-time-rule applied to the daily mean
and standard deviation gives results that are fairly close to the
actual monthly mean and standard deviation.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-DSoutlier
#| echo: false
#| fig-cap: Illustration of an "outlier" in a data sample.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# create nice systematic relationship with an outlier 
n.obs = 20 
set.seed(123) 
x = seq(0, 10, length.out=n.obs) 
e = rnorm(n.obs, 0, 0.1) 
y = 1 + 0.5*x + e 
y[5] = 5 
plot(x, y, type="p", lwd=2, col="blue", pch=16, cex=1.5) 
points(x[5], y[5], pch=16, cex=1.5, col="red")
```

### Outliers

@fig-DSoutlier nicely illustrates the concept of an
outlier in a data sample. All of the points are following a nice systematic
relationship except one - the red dot "outlier". Outliers
can be thought of in two ways. First, an outlier can be the result
of a data entry error. In this view, the outlier is not a valid observation
and should be removed from the data sample. Second, an outlier can
be a valid data point whose behavior is seemingly unlike the other
data points. In this view, the outlier provides important information
and should not be removed from the data sample. For financial market
data, outliers are typically extremely large or small values that
could be the result of a data entry error (e.g. price entered as 10
instead of 100) or a valid outcome associated with some unexpected
bad or good news. Outliers are problematic for data analysis because
they can greatly influence the value of certain sample statistics.


+----------------+---------+----------------------+------------+
|**Statistic**   |**GWN**  |**GWN with Outlier**  |\%$\Delta$  |
+================+=========+======================+============+
|Mean            |-0.0054  |-0.0072               |32\%        |
+----------------+---------+----------------------+------------+
|Variance        |0.0086   |0.0101                |16\%        |
+----------------+---------+----------------------+------------+
|Std. Deviation  |0.0929   |0.1003                |8\%         |
+----------------+---------+----------------------+------------+
|Skewness        |-0.2101  |-1.0034               |378\%       |
+----------------+---------+----------------------+------------+
|Kurtosis        |2.7635   |7.1530                |159\%       |
+----------------+---------+----------------------+------------+
|Median          |-0.0002  |-0.0002               |0\%         |
+----------------+---------+----------------------+------------+
|IQR             |0.1390   |0.1390                |0\%         |
+----------------+---------+----------------------+------------+

: Sample statistics for GWN with and without outlier. {#tbl-tableOutlier} 


::: {#exm-example0516 name="Effect of outliers on sample statistics"}
To illustrate the impact of outliers on sample statistics, the simulated
monthly GWN data is polluted by a single large negative outlier:

```{r}
#| echo: true

gwnMonthlyOutlier = gwnMonthly
gwnMonthlyOutlier[20] = -sd(gwnMonthly)*6
```

@fig-DSgwnOutlier shows the resulting data. Visually,
the outlier is much smaller than a typical negative observation and
creates a pronounced asymmetry in the histogram. Table @tbl-tableOutlier
compares the sample statistics @eq-sample-mean - @eq-sample-kurtosis
of the unpolluted and polluted data. All of the sample statistics
are influenced by the single outlier with the skewness and kurtosis
being influenced the most. The mean decreases by 32\% and the variance
and standard deviation increase by 16\% and 8\%, respectively. The
skewness changes by 378\% and the kurtosis by 159\%. Table @tbl-tableOutlier
also shows the median and the IQR, which are quantile-based statistics
for the center and spread, respectively. Notice that these statistics
are unaffected by the outlier.

```{r}
#| label: fig-DSgwnOutlier
#| echo: false
#| fig-cap: Monthly GWN data polluted by a single outlier.
#| out-width: 80%
#| fig-asp: 1
#| fig-align: center

par(mfrow=c(2,1))
	plot.zoo(gwnMonthlyOutlier,main="", lwd=2, col="blue", ylab="")
	abline(h=0)
	hist(gwnMonthlyOutlier, main="", col="cornflowerblue", 
       xlab="GWN with outlier")
par(mfrow=c(1,1))
```

\hfill{}$\blacksquare$
:::

The previous example shows that the common sample statistics @eq-sample-mean
- @eq-sample-kurtosis based on the sample average and deviations
from the sample average can be greatly influenced by a single outlier,
whereas quantile-based sample statistics are not. Sample statistics
that are not greatly influenced by a single outlier are called (outlier)
*robust statistics*\index{Robust statistics}.

#### Defining outliers

A commonly used rule-of-thumb defines an outlier as an observation
that is beyond the sample mean plus or minus three times the sample
standard deviation. The intuition for this rule comes from the fact
that if $X\sim N(\mu,\sigma^{2})$ then $\Pr(\mu-3\sigma\leq X\leq\mu+3\sigma)\approx0.99$.
While the intuition for this definition is clear, the practical implementation
of the rule using the sample mean, $\hat{\mu}$, and sample standard
deviation, $\hat{\sigma}$, is problematic because these statistics
can be heavily influenced by the presence of one or more outliers
(as illustrated in the previous sub-section). In particular, both
$\hat{\mu}$ and $\hat{\sigma}$ become larger (in absolute value)
when there are outliers and this may cause the rule-of-thumb rule
to miss identifying an outlier or to identify a non-outlier as an
outlier. That is, the rule-of-thumb outlier detection rule is not
robust to outliers! 


A natural fix to the above outlier detection rule-of-thumb is to use
outlier robust statistics, such as the median and the IQR, instead
of the mean and standard deviation. To this end, a commonly used definition
of a moderate outlier in the right tail of the distribution (value
above the median) is a data point $x$ such that: 
$$
\hat{q}_{.75}+1.5\cdot\mathrm{IQR}<x<\hat{q}_{.75}+3\cdot\mathrm{IQR}.
$$ {#eq-moderate-right-outlier}
To understand this rule, recall that the IQR is where the middle 50\%
of the data lies. If the data were normally distributed then $\hat{q}_{.75}\approx\hat{\mu}+0.67\times\hat{\sigma}$
and $\mathrm{IQR}\approx1.34\times\hat{\sigma}$. Then @eq-moderate-right-outlier
is approximately: 
$$
\hat{\mu}+2.67\times\hat{\sigma}<x<\hat{\mu}+4.67\times\hat{\sigma}
$$
Similarly, a moderate outlier in the left tail of the distribution
(value below the median) is a data point such that, 
$$
\hat{q}_{.25}-3\cdot IQR<x<\hat{q}_{.25}-1.5\cdot\mathrm{IQR}.
$$
Extreme outliers are those observations $x$ even further out in the
tails of the distribution and are defined using: 
\begin{align*}
x & >\hat{q}_{.75}+3\cdot\mathrm{IQR},\\
x & <\hat{q}_{.25}-3\cdot\mathrm{IQR}.
\end{align*}
For example, with normally distributed data an extreme outlier would
be an observation that is more than $4.67$ standard deviations above
the mean.

### Box plots
\index{Box plots}
A boxplot of a data series describes the distribution using outlier
robust statistics. The basic boxplot is illustrated in @fig-DSboxplot.
The middle of the distribution is shown by a rectangular box whose
top and bottom sides are approximately the first and third quartiles,
$\hat{q}_{.25}$ and $\hat{q}_{.75},$ respectively. The line roughly
in the middle of the box is the sample median, $\hat{q}_{.5}.$ The
top whisker is a horizontal line that represents either the largest
data value or $\hat{q}_{.75}+1.5\cdot IQR$, whichever is smaller.
Similarly, the bottom whisker is located at either the smallest data
value or $\hat{q}_{.25}-1.5\cdot IQR,$ whichever is closer to $\hat{q}_{.25}$.
Data values above the top whisker and below the bottom whisker are
displayed as circles and represent extreme observations. If a data
distribution is symmetric then the median is in the center of the
box and the top and bottom whiskers are approximately the same distance
from the box. Fat tailed distributions will have observations above
and below the whiskers. 


```{r}
#| label: fig-DSboxplot
#| echo: false
#| fig-cap: Boxplot of return distribution.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

q.25 = quantile(msftMonthlyRetS, 0.25) 
q.75 = quantile(msftMonthlyRetS, 0.75) 
q.5 = quantile(msftMonthlyRetS, 0.5) 
iqr = IQR(msftMonthlyRetS) 
boxplot(coredata(msftMonthlyRetS), col="cornflowerblue") 
text(x=1.25, y = q.25, labels="q.25") 
text(x=1.25, y = q.5, labels="q.50") 
text(x=1.25, y = q.75, labels="q.75") 
text(x=1.25, y = q.25-iqr, labels="q.25 - 1.5*IQR") 
text(x=1.25, y = q.75+iqr, labels="q.75 + 1.5*IQR") 
text(x=1.25, y = 0.3, labels="extreme values") 
text(x=1.25, y = -0.3, labels="extreme values")
```

::: {#exm-example0517 name="Boxplots of return distributions"}
Boxplots are a great tool for visualizing the return distribution
of one or more assets. @fig-DSboxplotReturns shows
the boxplots of the monthly continuously compounded and simple returns
on Microsoft and the S\&P 500 index computed using:^[The `coredata()` function is used to extract the data from
the zoo object because `boxplot()` doesn't like `"xts"`
or `"zoo"` objects.]

```{r}
#| label: fig-DSboxplotReturns
#| echo: true
#| fig-cap: Boxplots of the continuously compounded and simple monthly returns on Microsoft
#|   stock and the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

dataToPlot = merge(msftMonthlyRetS, msftMonthlyRetC, 
                   sp500MonthlyRetS, sp500MonthlyRetC) 
colnames(dataToPlot) = c("msftRetS", "msftRetC", 
                         "sp500RetS", "sp500RetC") 
boxplot(coredata(dataToPlot), main="Simple vs. CC Monthly Returns",         
        col="cornflowerblue")
```

The Microsoft returns have both positive and negative outliers whereas
the S\&P 500 index returns only have negative outliers. The boxplots
also clearly show that the distributions of the continuously compounded
and simple returns are very similar except for the extreme tails of
the distributions.
\hfill{}$\blacksquare$
:::

## Time Series Descriptive Statistics {#sec-Time-Series-Descriptive}

In this section, I consider graphical and numerical descriptive statistics
for summarizing the linear time dependence in a time series.

### Sample autocovariances and autocorrelations
\index{Sample Autocovariances}
For a covariance stationary time series process $\{X_{t}\}$, the
autocovariances $\gamma_{j}=\mathrm{cov}(X_{t},X_{t-j})$ and autocorrelations
$\rho_{j}=\mathrm{cov}(X_{t},X_{t-j})/\sqrt{\mathrm{var}(X_{t})\mathrm{var}(X_{t-j})}=\gamma_{j}/\sigma^{2}$
describe the linear time dependencies in the process. For a sample
of data $\{x_{t}\}_{t=1}^{T}$, the linear time dependencies are captured
by the *sample autocovariances* and *sample autocorrelations*: 
\begin{align*}
\hat{\gamma}_{j} & =\frac{1}{T-1}\sum_{t=j+1}^{T}(x_{t}-\bar{x})(x_{t-j}-\bar{x}),~j=1,2,\ldots,T-J+1,\\
\hat{\rho}_{j} & =\frac{\hat{\gamma}_{j}}{\hat{\sigma}^{2}},~j=1,2,\ldots,T-J+1,
\end{align*}
where $\hat{\sigma}^{2}$ is the sample variance @eq-sample-variance.
The *sample autocorrelation function*\index{Sample autocorrelation function} (SACF) is a plot of $\hat{\rho}_{j}$
vs. $j$, and gives a graphical view of the linear time dependencies
in the observed data.

::: {#exm-example0518 name="SACF for the Microsoft and S&P 500 returns"}
The R function `acf()` computes and plots the sample autocorrelations
$\hat{\rho}_{j}$ of a time series for a specified number of lags.
For example, to compute $\hat{\rho}_{j}$ for $j=1,\ldots,5$ for
the monthly and daily returns on Microsoft use:

```{r}
#| echo: true

acf(coredata(msftMonthlyRetS), lag.max=5, plot=FALSE)
acf(coredata(msftDailyRetS), lag.max=5, plot=FALSE)
```

None of these sample autocorrelations are very big and indicate lack
of linear time dependence in the monthly and daily returns. 

Sample autocorrelations can also be visualized using the function
`acf()`. @fig-DSsacf shows the SACF for the
monthly and daily returns on Microsoft and the S\&P 500 created using:

```{r}
#| label: fig-DSsacf
#| echo: true
#| fig-cap: SACFs for the monthly and daily returns on Microsoft and the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))
acf(coredata(msftMonthlyRetS), main="msftMonthlyRetS", lwd=2)
acf(coredata(sp500MonthlyRetS), main="sp500MonthlyRetS", lwd=2)
acf(coredata(msftDailyRetS), main="msftDailyRetS", lwd=2)
acf(coredata(sp500DailyRetS), main="sp500DailyRetS", lwd=2)
```

None of the series show strong evidence for linear time dependence.
The monthly returns have slightly larger sample autocorrelations at
low lags than the daily returns. The dotted lines on the SACF plots
are 5\% critical values for testing the null hypothesis that the true
autocorrelations $\rho_{j}$ are equal to zero (see chapter @sec-Hypothesis-Testing).
If $\hat{\rho}_{j}$ crosses the dotted line then the null hypothesis
that $\rho_{j}=0$ can be rejected at the 5\% significance level.
\hfill{}$\blacksquare$
:::

### Time dependence in volatility

The daily and monthly returns of Microsoft and the S\&P 500 index
do not exhibit much evidence for linear time dependence. Their returns
are essentially uncorrelated over time. However, the lack of autocorrelation
does not mean that returns are independent over time. There can be
nonlinear time dependence. Indeed, there is strong evidence for a
particular type of nonlinear time dependence in the daily returns
of Microsoft and the S\&P 500 index that is related to time varying
volatility. This nonlinear time dependence, however, is less pronounced
in the monthly returns.

::: {#exm-example0519 name="GWN and lack of time independence"}
Let $y_{t}\sim\mathrm{GWN}(0,\sigma^{2})$. Then $y_{t}$ is, by construction,
independent over time. This implies that any function of $y_{t}$
is also independent over time. To illustrate, @fig-DSgwnIndep
shows $y_{t}$, $y_{t}^{2}$ and $|y_{t}|$ together with their SACFs
where $y_{t}\sim \mathrm{GWN}(0,1)$ for $t=1,\ldots,500$. Notice that all
of the sample autocorrelations are essentially zero which confirms
the lack of linear time dependence in $y_{t}$, $y_{t}^{2}$ and $|y_{t}|$.

```{r}
#| label: fig-DSgwnIndep
#| echo: false
#| fig-cap: The left panel (top to bottom) shows $y_{t}$, $y_{t}^{2}$ and $|y_{t}|$ where
#|   $y_{t}$ is similar to GWN(0,1) for $t=1,...,500.$ The right panel shows the corresponding
#|   SACFs.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

n.obs = 500
set.seed(123)
gwn = rnorm(n.obs)
par(mfrow=c(3, 2))
  ts.plot(gwn)
  abline(h=0)
  acf(gwn, lwd=2)
  ts.plot(gwn^2)
  acf(gwn^2, lwd=2)
  ts.plot(abs(gwn))
  acf(abs(gwn), lwd=2)
par(mfrow=c(1,1))
```

\hfill{}$\blacksquare$
:::

#### Nonlinear time dependence in asset returns

@fig-DSdailyReturns and @fig-DSmonthlyReturns
show that the daily and monthly returns on Microsoft and the S\&P
500 index appear to exhibit periods of volatility clustering. That
is, high periods of volatility tend to be followed by periods of high
volatility and periods of low volatility appear to be followed by
periods of low volatility. In other words, the volatility of the returns
appears to exhibit some time dependence. This time dependence in volatility
causes linear time dependence in the absolute value and squared value
of returns.

::: {#exm-example0520 name="Nonlinear time dependence in the S&P 500 returns"}
To illustrate nonlinear time dependence in asset returns, @fig-DSsp500Vol
shows the daily returns, absolute returns and squared returns for
the S\&P 500 index. Notice how the absolute and squared returns are
large (small) when the daily returns are more (less) volatile. 
@fig-DSsp500vol2 shows the SACFs for the absolute and squared
daily and monthly returns. There is clear evidence of time dependence
in the daily absolute and squared returns while there is some evidence
of time dependence in the monthly absolute and squared returns. Since
the absolute and squared daily returns reflect the volatility of the
daily returns, the SACFs show a strong positive linear time dependence
in daily return volatility.

```{r}
#| label: fig-DSsp500Vol
#| echo: false
#| fig-cap: Daily returns, absolute returns, and squared returns on the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

dataToPlot = merge(sp500DailyRetS, abs(sp500DailyRetS), sp500DailyRetS^2)
colnames(dataToPlot) = c("Returns", "abs(Returns)", "Returns^2")
plot.zoo(dataToPlot, main="Daily Returns", col="blue")
```

```{r}
#| label: fig-DSsp500vol2
#| echo: false
#| fig-cap: SACFs of daily and monthly absolute and squared returns on the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))
acf(abs(coredata(sp500DailyRetS)), main="Daily Absolute Returns", lwd=2)
acf(coredata(sp500DailyRetS)^2, main="Daily Squared Returns", lwd=2)
acf(abs(coredata(sp500MonthlyRetS)), main="Monthly Absolute Returns", lwd=2)
acf(coredata(sp500MonthlyRetS)^2, main="Monthly Squared Returns", lwd=2)
par(mfrow=c(1,1))
```

\hfill{}$\blacksquare$
:::

## Bivariate Descriptive Statistics {#sec-Bivariate-Descriptive-Statistics}

In this section, I consider graphical and numerical descriptive statistics
for summarizing two or more data series.

### Scatterplots
\index{Scatterplots}
The contemporaneous dependence properties between two data series
$\{x_{t}\}_{t=1}^{T}$ and $\{y_{t}\}_{t=1}^{T}$ can be displayed
graphically in a *scatterplot*, which is simply an xy-plot of the bivariate
data.

::: {#exm-example0521 name="Scatterplot of Microsoft and S&P 500 returns"}
@fig-DSscatterplot shows the scatterplots between
the Microsoft and S\&P 500 monthly and daily returns created using:

```{r}
#| label: fig-DSscatterplot
#| echo: true
#| fig-cap: Scatterplot of Monthly returns on Microsoft and the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(1,2)) 
plot(coredata(sp500MonthlyRetS),coredata(msftMonthlyRetS),
     main="Monthly returns", xlab="S&P500", ylab="MSFT", lwd=2,
     pch=16, cex=1.25, col="blue") 
abline(v=mean(sp500MonthlyRetS))   
abline(h=mean(msftMonthlyRetS)) 
plot(coredata(sp500DailyRetS),coredata(msftDailyRetS),           
     main="Daily returns", xlab="S&P500", ylab="MSFT", lwd=2, 
     pch=16, cex=1.25, col="blue") 
abline(v=mean(sp500DailyRetS))   
abline(h=mean(msftDailyRetS))
```

The S\&P 500 returns are put on the x-axis and the Microsoft returns
on the y-axis because the "market", as proxied by the S\&P 500,
is often thought as an independent variable driving individual asset
returns. The upward sloping orientation of the scatterplots indicate
a positive linear dependence between Microsoft and S\&P 500 returns
at both the monthly and daily frequencies.
\hfill{}$\blacksquare$
:::

::: {#exm-example0522 name="Pair-wise scatterplots for multiple series"}
For more than two data series, the R function `pairs()` plots
all pair-wise scatterplots in a single plot. For example, to plot
all pair-wise scatterplots for the GWN, Microsoft returns and S\&P
500 returns use:

```{r}
#| label: fig-DSpairs
#| echo: true
#| fig-cap: Pair-wise scatterplots between simulated GWN, Microsoft returns and S\&P
#|   500 returns.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

dataToPlot = merge(gwnMonthly,msftMonthlyRetS,sp500MonthlyRetS)
pairs(coredata(dataToPlot), col="blue", pch=16, cex=1.25, cex.axis=1.25)
```

The top row of @fig-DSpairs shows the scatterplots
between the pairs (MSFT, GWN) and (SP500, GWN), the second row shows
the scatterplots between the pairs (GWN, MSFT) and (SP500, MSFT),
the third row shows the scatterplots between the pairs (GWN, SP500)
and (MSFT, SP500). The plots in the lower triangle are the same as the plots in the upper
triangle except the axes are reversed. 
\hfill{}$\blacksquare$
:::

### Sample covariance and correlation
\index{Sample covirance}
For two random variables $X$ and $Y,$ the direction of linear dependence
is captured by the covariance, $\sigma_{XY}=E[(X-\mu_{X})(Y-\mu_{Y})],$
and the direction and strength of linear dependence is captured by
the correlation, $\rho_{XY}=\sigma_{XY}/\sigma_{X}\sigma_{Y}.$ 

::: {#def-sample-covariance-and-correlation name="Sample Covariance and Correlation"}
For two data series $\{x_{t}\}_{t=1}^{T}$ and $\{y_{t}\}_{t=1}^{T},$
the *sample covariance*, 
$$
\hat{\sigma}_{xy}=\frac{1}{T-1}\sum_{t=1}^{T}(x_{t}-\bar{x})(y_{t}-\bar{y}),
$$ {#eq-sample-covariance}
measures the direction of linear dependence, and the *sample correlation*\index{Sample correlation},
$$
\hat{\rho}_{xy}=\frac{\hat{\sigma}_{xy}}{\hat{\sigma}_{x}\hat{\sigma}_{x}},
$$ {#eq-sample-correlation}
measures the direction and strength of linear dependence. In @eq-sample-correlation,
$\hat{\sigma}_{x}$ and $\hat{\sigma}_{y}$ are the sample standard
deviations of $\{x_{t}\}_{t=1}^{T}$ and $\{y_{t}\}_{t=1}^{T},$ respectively,
defined by @eq-sample-sd. 
:::

When more than two data series are being analyzed, it is often convenient
to compute all pair-wise covariances and correlations at once using
matrix algebra. Recall, for a vector of $N$ random variables $\mathbf{X}=(X_{1},\ldots,X_{N})^{\prime}$
with mean vector $\mu=(\mu_{1},\ldots,\mu_{N})^{\prime}$
the $N\times N$ variance-covariance matrix is defined as: 
$$
\boldsymbol{\Sigma}=\mathrm{var}(\mathbf{X})=E[(\mathbf{X}-\boldsymbol{\mu})(\mathbf{X}-\boldsymbol{\mu})^{\prime}]=\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \cdots & \sigma_{1N}\\
\sigma_{12} & \sigma_{2}^{2} & \cdots & \sigma_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1N} & \sigma_{2N} & \cdots & \sigma_{N}^{2}
\end{array}\right).
$$

::: {#def-sample-covariance-and-correlation-matrices name="Sample Covariance and Correlation Matrices"}
For $N$ data series $\{\mathbf{x}_{t}\}_{t=1}^{T},$ where $x_{t}=(x_{1t},\ldots,x_{Nt})^{\prime},$
the *sample covariance matrix* is computed using: 
$$
\hat{\boldsymbol{\Sigma}}=\frac{1}{T-1}\sum_{t=1}^{T}(\mathbf{x}_{t}-\hat{\boldsymbol{\mu}})(\mathbf{x}_{t}-\hat{\boldsymbol{\mu}})^{\prime}=\left(\begin{array}{cccc}
\hat{\sigma}_{1}^{2} & \hat{\sigma}_{12} & \cdots & \hat{\sigma}_{1N}\\
\hat{\sigma}_{12} & \hat{\sigma}_{2}^{2} & \cdots & \hat{\sigma}_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\hat{\sigma}_{1N} & \hat{\sigma}_{2N} & \cdots & \hat{\sigma}_{N}^{2}
\end{array}\right),
$$ {#eq-sample-covariance-matrix}
where $\hat{\boldsymbol{\mu}}$ is the $N\times1$ sample mean vector.
Define the $N\times N$ diagonal matrix: 
$$
\mathbf{\hat{D}}=\left(\begin{array}{cccc}
\hat{\sigma}_{1} & 0 & \cdots & 0\\
0 & \hat{\sigma}_{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & \hat{\sigma}_{N}
\end{array}\right).
$$ {#eq-DSdhat}
Then the $N\times N$ *sample correlation matrix* $\mathbf{\hat{C}}$
is computed as: 
$$
\mathbf{\hat{C}}=\mathbf{\hat{D}}^{-1}\hat{\boldsymbol{\Sigma}}\mathbf{\hat{D}}^{-1}=\left(\begin{array}{cccc}
1 & \hat{\rho}_{12} & \cdots & \hat{\rho}_{1N}\\
\hat{\rho}_{12} & 1 & \cdots & \hat{\rho}_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\hat{\rho}_{1N} & \hat{\rho}_{2N} & \cdots & 1
\end{array}\right).
$$ {#eq-sample-correlation-matrix}
:::

::: {#exm-example0523 name="Sample covariance and correlation between Microsoft and S&P 500 returns"}
The scatterplots of Microsoft and S\&P 500 returns in @fig-DSscatterplot
suggest positive linear relationships in the data. You can confirm
this by computing the sample covariance and correlation using the
R functions `cov()` and `cor()`. For the monthly returns,
you get:

```{r}
#| echo: true

cov(sp500MonthlyRetS, msftMonthlyRetS)
cor(sp500MonthlyRetS, msftMonthlyRetS)
```

Indeed, the sample covariance is positive and the sample correlation
shows a moderately strong linear relationship. For the daily returns
you get:

```{r}
#| echo: true

cov(sp500DailyRetS, msftDailyRetS) 
cor(sp500DailyRetS, msftDailyRetS)
```

Here, the daily sample covariance is about twenty times smaller than
the monthly covariance (recall the square-root-of time rule), but
the daily sample correlation is similar to the monthly sample correlation.


When passed a matrix of data, the `cov()` and `cor()`
functions can also be used to compute the sample covariance and correlation
matrices $\hat{\boldsymbol{\Sigma}}$ and $\mathbf{\hat{C}}$, respectively.
For example, 

```{r}
#| echo: true

cov(msftSp500MonthlyRetS)
cor(msftSp500MonthlyRetS)
```

The function `cov2cor()` transforms a sample covariance matrix
to a sample correlation matrix using @eq-sample-correlation-matrix:

```{r}
#| echo: true

cov2cor(cov(msftSp500MonthlyRetS))
```

\hfill{}$\blacksquare$
:::

::: {#exm-example0524 name="Visualizing correlation matrices"}
The R package **corrplot** contains functions for visualizing
correlation matrices. This is particularly useful for summarizing
the linear dependencies among many data series. For example, 
@fig-DScorrplot shows the correlation plot from the **corrplot**
function `corrplot.mixed()` created using:

```{r}
#| label: fig-DScorrplot
#| echo: true
#| fig-cap: Correlation plot created with `corrplot()`.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

dataToPlot = merge(gwnMonthly, msftMonthlyRetS, sp500MonthlyRetS)
cor.mat = cor(dataToPlot)
corrplot.mixed(cor.mat, lower="number", upper="ellipse")
```

The color scheme shows the magnitudes of the correlations (blue for
positive and red for negative) and the orientation of the ellipses
show the magnitude and direction of the linear associations.
\hfill{}$\blacksquare$
:::

### Sample cross-lag covariances and correlations

The dynamic interactions between two observed time series $\{x_{t}\}_{t=1}^{T}$and
$\{y_{t}\}_{t=1}^{T}$ can be measured using the *sample cross-lag covariances* and *correlations* defined by: 
\begin{align*}
\hat{\gamma}_{xy}^{k} & =\widehat{\mathrm{cov}}(X_{t},Y_{t-k}),\\
\hat{\rho}_{xy}^{k} & =\widehat{\mathrm{corr}}(X_{t},Y_{t-k})=\frac{\hat{\gamma}_{xy}^{k}}{\sqrt{\hat{\sigma}_{x}^{2}\hat{\sigma}_{y}^{2}}}.
\end{align*}
When more than two data series are being analyzed, all pairwise cross-lag
sample covariances\index{Sample cross-lag covariances} and sample correlations\index{Sample cross-lag correlations} can be computed at once
using matrix algebra. For a time series of $N$ data series $\{\mathbf{x}_{t}\}_{t=1}^{T},$
where $\mathbf{x}_{t}=(x_{1t},\ldots,x_{Nt})^{\prime},$ the sample
lag $k$ cross-lag covariance and correlation matrices are computed
using:
$$
\hat{\boldsymbol{\Gamma}}_{k} =  \frac{1}{T-1}\sum_{t=k+1}^{T}(\mathbf{x}_{t}-\hat{\boldsymbol{\mu}})(\mathbf{x}_{t-k}-\hat{\boldsymbol{\mu}})^{\prime},
$$ {#eq-DSGammaHatk}
$$
\hat{\mathbf{C}}_{k} = \hat{\mathbf{D}}^{-1}\hat{\boldsymbol{\Gamma}}_{k}\hat{\mathbf{D}}^{-1},
$$ {#eq-DSChatk}
where $\hat{\mathbf{D}}$ is defined in @eq-DSdhat. 

::: {#exm-example0525 name="Sample cross-lag covariances and correlations between Microsoft and S&P 500 returns"}
Consider computing the cross-lag covariance and correlation matrices
@eq-DSGammaHatk and @eq-DSChatk for $k=0,1,\ldots,5$
between Microsoft and S\&P 500 monthly returns. These matrices may
be computed using the R function `acf()` as follows:

```{r}
#| echo: true

Ghat = acf(coredata(msftSp500MonthlyRetS), type="covariance",
           lag.max=5, plot=FALSE) 
Chat = acf(coredata(msftSp500MonthlyRetS), type="correlation", 
           lag.max=5, plot=FALSE) 
names(Ghat)
```

Here, `Ghat` and `Chat` are objects of class `"acf"`
for which there are print and plot methods. The `acf` components
of `Ghat` and `Chat` are 3-dimensional arrays containing
the cross lag matrices @eq-DSGammaHatk and @eq-DSChatk,
respectively. For example, to extract $\hat{\mathbf{C}}_{0}$ and
$\hat{\mathbf{C}}_{1}$ use:

```{r}
#| echo: true

# Chat0 
Chat$acf[1,,] 
# Chat1 
Chat$acf[2,,]
```

The print method shows the sample autocorrelations of each variable
as well as the pairwise cross-lag correlations:

```{r}
#| echo: true

Chat
```

These values can also be visualized using the plot method:

```{r}
#| label: fig-DScrosslagcorrelations
#| echo: true
#| fig-cap: Sample cross-lag correlations between Microsoft and S\&P 500 returns.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(Chat, lwd=2)
```

@fig-DScrosslagcorrelations shows the resulting four-panel
plot. The top-left and bottom-right (diagonal) panels give the SACFs
for Microsoft and S\&P 500 returns. The top-right plot gives the cross-lag
correlations $\hat{\rho}_{msft,sp500}^{k}$ for $k=0,1,\ldots,5$
and the bottom-left panel gives the cross-lag correlations $\hat{\rho}_{sp500,msft}^{k}$
for $k=0,1,\ldots,5$. The plots show no evidence of any dynamic feedback between the return series. 
\hfill{}$\blacksquare$
:::

## Rolling descriptive statistics

\index{Rolling estimators}
The assumption that asset returns are covariance stationary implies that the asset return means ($\mu_i$), variances ($\sigma_i^2$), covariances ($\sigma_{ij}$), and correlations ($\rho_{ij}$) are constant over time for all assets $i$ and $j$. Essentially, this means that the economic environment remains constant over time. Clearly, this is a strong assumption and you have seen evidence in the graphical descriptive statistics that this assumption might be suspect. For example, you have seen that the volatility of asset returns appears to change over time. In this section, I introduce *rolling descriptive statistics* to further investigate the appropriateness of the covariance stationary assumption of asset returns.  

A commonly used technique to investigate whether certain asset return characteristics are constant over time is to compute sample descriptive statistics over rolling sub-sample windows of fixed length. To illustrate, suppose one is interested in determining if the expected return parameter of an asset return $\mu$ is constant over time. Assume returns are observed over a sample of size $T$. An informal way to do this is to compute estimates of $\mu$ over rolling windows of
length $n<T$:

$$
\hat{\mu}_{t}(n) =\frac{1}{n}\sum_{k=0}^{n-1}r_{t-k}=\frac{1}{n}(r_{t}+r_{t-1}+\cdots+r_{t-n+1}),
$$ {#eq-DSrollingmean}


for $t=n,\ldots,T$. Here $\hat{\mu}_{n}(n)$ is the sample mean of
the returns $\{r_{t}\}_{t=1}^{n}$ over the first sub-sample window of
size $n$. Then $\hat{\mu}_{n+1}(n)$ is the sample mean of the returns
$\{r_{t}\}_{t=2}^{n+1}$ over the second sub-sample window of size $n$.
Continuing in this fashion I roll the sample window forward one observation
at a time and compute $\hat{\mu}_{t}(n)$ until I arrive at the last
sub-sample window of size $n$. The end result is $T-n+1$ rolling estimates
$\left\{\hat{\mu}_{n}(n),\,\hat{\mu}_{n+1}(n),\ldots,\hat{\mu}_{T}(n)\right\}$.

If returns are covariance stationary then $\mu$ is constant over
time and you would expect that $\hat{\mu}_{n}(n)\approx\hat{\mu}_{n+1}(n)\approx\cdots\approx\hat{\mu}_{T}(n)$.
That is, the estimates of $\mu$ over each rolling window should be
similar. I would not expect them to be exactly equal due to sampling uncertainty but they should not be drastically different. In contrast, if returns are not covariance stationary due to time
variation in $\mu$ then you would expect to see meaningful time variation
in the rolling estimates $\left\{\hat{\mu}_{n}(n),\,\hat{\mu}_{n+1}(n),\ldots,\hat{\mu}_{T}(n)\right\}$.

A natural way to see the time variation in $\{\hat{\mu}_{t}(n)\}_{t=n}^T$ is to make a time series
plot of these rolling estimates as illustrated in the next example. 

::: {#exm-example0526 name="Rolling mean estimates for Microsoft S&P 500 Index and simulated GWN"}
Consider computing rolling estimates of $\mu$ for Microsoft, Starbucks
and the S\&P 500 index returns. You can easily do this using the **zoo**
function `rollapply()`, which applies a user-specified function
to a time series over rolling windows of fixed size. The arguments
to `rollapply()` are:

```{r}
#| echo: true

args(zoo:::rollapply.zoo)
```

where `data` is either a `"zoo"` or `"xts"` object, `width` is the window width, `FUN` is the R
function to be applied over the rolling windows, `...` are any optional arguments to be supplied to `FUN`, `by`
specifies how many observations to increment the rolling windows, `by.column` specifies if `FUN` is to be applied to each
column of data or to be applied to all columns of data, and `align` specifies the location of the time index in the window. The function returns an "xts/zoo" object containing the rolling estimates.

For example, to compute 24-month rolling estimates of $\mu$ for Microsoft, S\&P 500 index, and simulated GWN data (that matches the mean and volatility of Microsoft returns), incremented by one month with the time index aligned to the end of the window use:

```{r}
#| echo: true

roll.data = merge(msftMonthlyRetS,sp500MonthlyRetS,gwnMonthly)
colnames(roll.data) = c("MSFT", "SP500", "GWN")
roll.muhat = rollapply(roll.data, width=24, by=1, by.column=TRUE,                       
                       FUN=mean, align="right") 
class(roll.muhat) 

```
The object `roll.muhat` is an "xts" object with three columns giving the rolling mean estimates for the GWN data, Microsoft, and the S\&P 500 index. The first 23 months of data are NA as you need at least 24 months of data to compute the first rolling estimates:

```{r}
#| echo: true
head(roll.muhat, n=3)

```
The first rolling estimates are for the 24-months just prior to January,
2000 and the last rolling estimates are for the 24-months just prior
to May 2012. 

```{r}
head(na.omit(roll.muhat), n=3)
tail(na.omit(roll.muhat), n=3)
```

The rolling means are shown in @fig-HypothesisRollMuhat created using: 

```{r}
#| label: fig-HypothesisRollMuhat
#| echo: true
#| fig-cap: 24-month rolling estimates of $\mu$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(roll.muhat, main="", multi.panel=FALSE, lwd=2, 
     col=c("black", "red", "green"), lty=c("solid", "solid", "solid"), 
     major.ticks="years", grid.ticks.on="years",
     legend.loc = "topright")
```

All of the rolling estimates show time variation over the sample, including the simulated GWN returns which have a constant mean by construction. The ranges of the rolling estimates are:

```{r}
#| echo: true
ans = apply(na.omit(roll.muhat),2,range)
rownames(ans) = c("Min", "Max")
ans
```
Interestingly, the rolling estimates for Microsoft and the S\&P 500 index exhibit a similar pattern of time variation. The estimates start positive at the end of the
dot-com boom, decrease from 2000 to mid 2002 (Microsoft and the S\&P 500 index become negative), 
increase from 2003 to 2006 during the booming economy, fall sharply during the financial
crisis period 2007 to 2009, and increase from 2009 to 2011, and decrease afterward. The
Microsoft and S\&P 500 estimates follow each other very closely due to their moderately 
strong positive correlation of 0.614. Somewhat surprisingly, the rolling means for the 
simulated GWN data also show substantial time variation over the sample. Since the true 
mean of the simulated GWN data is constant, the variation observed in the rolling means 
represents sampling uncertainty due to the small sample sizes of the rolling means. The 
magnitude of the variation in the rolling mean estimates for the GWN data should make us 
pause in concluding there is substantial change in the mean estimates of Microsoft 
(and possibly the S\&P 500). However, the fact that the rolling estimates for Microsoft 
and the S\&P 500 index behave as expected during the booms and busts strongly suggests 
that the underlying mean values for these assets are not constant over the full sample.
\hfill{}$\blacksquare$
:::

To investigate time variation in return volatility, I compute
rolling estimates of $\sigma^{2}$ and $\sigma$ over fixed windows
of size $n<T$:
$$
\hat{\sigma}_{t}^{2}(n) =\frac{1}{n-1}\sum_{k=0}^{n-1}(r_{t-k}-\hat{\mu}_{t}(n))^{2},
$$ {#eq-DSrollingsigma2}
$$
\hat{\sigma}_{t}(n) =\sqrt{\hat{\sigma}_{t}^{2}(n)}.
$$ {#eq-DSrollingsigma}
If return volatility is constant over time, then you would expect that
$\hat{\sigma}_{n}(n)\approx\hat{\sigma}_{n+1}(n)\approx\cdots\approx\hat{\sigma}_{T}(n)$.
That is, the estimates of $\sigma$ over each rolling window should
be similar. In contrast, if volatility is not constant over time then
you would expect to see meaningful time variation in the rolling estimates
$\left\{\hat{\sigma}_{n}(n),\,\hat{\sigma}_{n+1}(n),\ldots,\hat{\sigma}_{T}(n)\right\}$.

::: {#exm-example0527 name="Rolling volatility estimates for Microsoft Starbucks and the S&P 500 Index"}
To compute and plot 24-month rolling estimates of $\sigma$ for Microsoft,
S\&P 500 index, and GWN returns incremented by one month with the
time index aligned to the beginning of the window use:

```{r}
roll.sigmahat = rollapply(roll.data,width=24, by=1, by.column=TRUE,
                          FUN=sd, align="right")
```

The rolling estimates are illustrated in @fig-HypothesisRollSigma.

```{r}
#| label: fig-HypothesisRollSigma
#| echo: false
#| fig-cap: 24-month rolling estimates of $\sigma$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(roll.sigmahat, main="", multi.panel=FALSE, lwd=2, 
     col=c("black", "red", "green"), lty=c("solid", "solid", "solid"), 
     major.ticks="years", grid.ticks.on="years",
     legend.loc = "topright")

```

The rolling volatilities for Microsoft and the S\&P 500 index show a similar pattern of 
time variation over the sample but the Microsoft volatilities are considerably more variable:

```{r}
#| echo: true
ans = apply(na.omit(roll.sigmahat),2,range)
rownames(ans) = c("Min", "Max")
ans
```

The rolling volatilities for Microsoft peak at about $18\%$ per month at the end of 
the dot-com boom, fall sharply to about $4\%$ per month at the beginning of 2005, then 
level off until the beginning of the financial
crisis, and increase during the financial crisis to about $10\%$ per month and fall 
slightly afterward. The S\&P 500 volatilities have a similar pattern as the Microsoft 
volatilities but their movements are not as extreme. In contrast, the rolling volatilities 
of the GWN series are fairly stable bouncing between $7\%$ and $12\%$. The
time variation exhibited by the rolling volatilities of Microsoft and the 
S\&P 500 index places additional doubt on the covariance stationarity assumption for returns.
\hfill{}$\blacksquare$
:::

To investigate time variation in the correlation between two asset
returns, $\rho_{ij}$, one can compute the rolling covariances and correlations over fixed windows
of size $n<T$:
$$
\hat{\sigma}_{ij,t}(n)  =\frac{1}{n-1}\sum_{k=0}^{n-1}(r_{i,t-k}-\hat{\mu}_{i}(n))(r_{j,t-k}-\hat{\mu}_{j}(n)),
$$ {#eq-DSrollingcovariance}
$$
\hat{\rho}_{ij,t}(n)  =\frac{\hat{\sigma}_{ij,t}(n)}{\hat{\sigma}_{i,t}(n)\hat{\sigma}_{j,t}(n)}.
$$ {#eq-DSrollingcorrelation}

::: {#exm-example0528 name="Rolling correlation estimates for Microsoft Starbucks and the S&P 500 index"}
A function to compute the pair-wise sample correlations between Microsoft,
Starbucks and the S\&P 500 index is

```{r}
#| echo: true

rhohat = function(x) {   
   corhat = cor(x)   
   corhat.vals = corhat[lower.tri(corhat)]   
   names(corhat.vals) = c("MSFT.SP500", "MSFT.GWN", "SP500.GWN")   
   corhat.vals 
}
```

Using the function `rhohat()`, 24-month rolling estimates of
$\rho_{ij}$ for Microsoft, Starbucks and the S\&P 500 index incremented
by one month with the time index aligned to the beginning of the window
can be computed using

```{r}
#| echo: true
roll.rhohat = rollapply(roll.data, width=24, FUN=rhohat,                          
   by=1, by.column=FALSE, align="right") 
head(na.omit(roll.rhohat),n=3)
```

Notice that `by.column=FALSE` in the call to `rollapply()`.
This is necessary as the function `rhohat()` operates on all
columns of the data object `roll.data`. The rolling estimates
are illustrated in @fig-HypothesisRollRhohat.

```{r}
#| label: fig-HypothesisRollRhohat
#| echo: false
#| fig-cap: 24-month rolling estimates of $\rho_{ij}$ for Microsoft, Starbucks and the
#|   S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(roll.rhohat, main="", multi.panel=FALSE, lwd=2, 
     col=c("black", "red", "green"), lty=c("solid", "solid", "solid"), 
     major.ticks="years", grid.ticks.on="years",
     legend.loc = "bottomleft")

```

The ranges of the rolling correlations are:

```{r}
#| echo: true
ans = apply(na.omit(roll.rhohat),2,range)
rownames(ans) = c("Min", "Max")
ans
```

The rolling correlations between Microsoft and the S\&P 500 index show an interesting pattern. They start moderately high, around 0.6, then plunge to about 0.05 in early 2005, and then gradually pick back up to about 0.8 at the end of the sample. The rolling correlations are highest during the financial crisis periods and lowest during the calm periods. Interestingly, the rolling correlations between the GWN series and Microsoft, and the GWN series and the S\&P 500 index fluctuate somewhat randomly between -0.5 and 0.5 and show no particular pattern. There is compelling evidence that the correlation between Microsoft and the S&P 500 index is not constant over time.
\hfill{}$\blacksquare$
:::

It must be kept in mind that the rolling descriptive statistics are *informal diagnostics* for evaluating the assumption of covariance stationarity. As you have seen, there is considerably sampling variability in the rolling statistics due to the small window size which can make it difficult to untangle statistical noise from true changes in the underlying parameters. In Chapter @sec-Hypothesis-Testing I will consider formal statistical tests for constant parameters that can give better evidence for or against the assumption of covariance stationarity.

### Practical issues associated with rolling estimates

To be completed

*  Discuss choice of window width. Small window widths tend to produce choppy estimates with a lot of estimation error. Large window widths produce smooth estimates with less estimation error
*  Discuss choice of increment. For a given window width, the roll-by increment determines the number of overlapping returns in each window. Number of overlapping returns is $n-incr$ where $n$ is the window width and $incr$ is the roll-by increment. Incrementing the window by 1 gives the largest number of over lapping returns. If $incr\geq n$ then the rolling windows are non-overlapping.  
*  A large positive or negative return in a window can greatly influence the estimate and cause a drop-off effect as the window rolls past the large observation. 


## Stylized facts for daily and monthly asset returns {#sec-Stylized-facts}

A *stylized fact*\index{Stylized fact} is something that is generally true but not always.
From the data analysis of the daily and monthly returns on Microsoft
and the S\&P 500 index I observe a number of stylized facts that
tend to be true for other individual assets and portfolios of assets.
For monthly data I observe the following stylized facts: 

1. M1. Prices appear to be random walk non-stationary and returns appear to be mostly covariance stationary. There is evidence that return volatility changes over time. 
2. M2. Returns appear to be approximately normally distributed. There is some negative skewness and excess kurtosis. 
3. M3. Assets that have high average returns tend to have high standard deviations (volatilities) and vice-versa. This is the *no free lunch* principle. 
4. M4. Returns on individual assets (stocks) have higher standard deviations than returns on diversified portfolios of assets (stocks). 
5. M5. Returns on different assets tend to be positively correlated. It is unusual for returns on different assets to be negatively correlated. 
6. M6. Returns are approximately uncorrelated over time. That is, there is little evidence of linear time dependence in asset returns. 
7. M7. Returns do not exhibit dynamic feedback. That is, there are no strong lead-lag effects between pairs of returns.
8. M8  There is some informal evidence of non-constant volatilities and correlations, particularly during financial crisis periods.


Daily returns have some features in common with monthly returns and
some not. The common features are M1 and M3-M7 above. The stylized
facts that are specific to daily returns are: 

1. D2. Returns are not normally distributed. Empirical distributions have much fatter tails than the normal distribution (excess kurtosis). 
2. D8. Returns are not independent over time. Absolute and squared returns are positively auto correlated and the correlation dies out very slowly. Volatility appears to be auto correlated and, hence, predictable. 


These stylized facts of daily and monthly asset returns are the main
features of returns that models of assets returns should capture.
A *good model*\index{Good model} is one that can explain many stylized facts of the data.
A bad model does not capture important stylized facts.

## Further Reading: Descriptive Statistics

[@fama-1976] is the first book that provides a comprehensive descriptive
statistical analysis of asset returns. Much of the material in this
chapter, and this book, is motivated by Fama's book. Several recent
textbooks discuss exploratory data analysis of finance time series.
This discussion in this chapter is similar to that presented in [@carmona-2014], [@JPR-2007], [@RuppertMatteson-2015], [@tsay-2010], [@tsay-2014], and [@zivot-2016]. 

* Mention books by Carol Alexander

* Mention R packages for descriptive statistics: dataExplorer, 

## Problems: Descriptive Statistics

::: {#exr-01}
Compute time plots with *ggplot2* function `autoplot()`
:::

::: {#exr-02}
In this question, you will compute descriptive statistics
for the continuously compounded monthly return data on the Vanguard
long term bond index fund (VBLTX), Fidelity Magellan stock mutual
fund (FMAGX), and Starbucks stock (SBUX). You will use the R packages
**corrplot**, **IntroCompFinR**, **PerformanceAnalytics**,
**tseries** and **zoo**.

1. Downloading data and return calculations. 
    a. Use the **tseries** function `get.hist.quote()` to download from Yahoo! the monthly adjusted closing prices on VBLTX, FMAGX and SBUX over the period 1998-01-01 through 2009-12-31. Download these prices as `"zoo"` objects. 
    b. Use the **PerformanceAnalytics** function `Return.calculate()` to compute continuously compounded monthly returns and remove the first NA observation.

2. Univariate graphical analysis. 
    a. Make time plots of the return data using the **zoo** function `plot.zoo()`. Comment on any relationships between the returns suggested by the plots. Pay particular attention to the behavior of returns toward the end of 2008 at the beginning of the financial crisis.
    b. Make a equity curve/cumulative return plot (future of \$1 invested in each asset) and comment. Which assets gave the best and worst future values over the investment horizon?
    c. For each return series, use the **IntroCompFinR** function `fourPanelPlot()` to make a four-panel plot containing a histogram, boxplot, normal QQ-plot, and SACF plot. Do the return series look normally distributed?  Briefly compare the return distributions. Do you see any evidence of time dependence in the returns?

3. Univariate numerical summary statistics and historical VaR.
    a. Compute numerical descriptive statistics for all assets using the R functions `summary()`, `mean()`, `var()`, `stdev()`, and the **PerformanceAnalytics** functions `skewness()`
and `kurtosis()`. Compare and contrast the descriptive statistics for the three assets. Which asset appears to be the riskiest asset?
    b. Using the mean monthly return for each asset, compute an estimate of the annual continuously compounded return (i.e., recall the relationship between the expected monthly cc return and the expected annual cc return). Convert this annual continuously compounded return into a simple annual return. Are there any surprises?
    c. Using the estimate of the monthly return standard deviation for each asset, compute an estimate of the annual return standard deviation. Briefly comment on the magnitude of the annual standard deviations. 
    d. For each asset compute the empirical 1\% and 5\% quantiles of the cc returns. Using these quantiles compute the 1\% and 5\% historical (monthly) VaR values based on an initial \$100,000 investment. Which asset has the highest and lowest VaR values? Are you surprised?

4. Bivariate graphical analysis and numerical summary statistics
    a. Use the R `pairs()` function to create all pair-wise scatterplots of returns Comment on the direction and strength of the linear relationships in these plots. 
    b. Use the **corrplot** functions `corrplot()` and `corrplot.mixed()` to plot the correlation matrix of the returns on the three assets.
    c. Use the R functions `var()`, `cov()`, and `cor()` to compute the sample covariance matrix and sample correlation matrix of the returns. Comment on the direction and strength of the linear
relationships suggested by the values of the covariances and correlations.
:::

::: {#exr-03}
In this question, you will compute descriptive statistics
for the continuously compounded weekly return data on the Vanguard
long term bond index fund (VBLTX), Fidelity Magellan stock mutual
fund (FMAGX), and Starbucks stock (SBUX). Repeat the analysis for
the monthly returns but use instead weekly returns computed from end-of-week
adjusted closing prices. 
:::

::: {#exr-04}
In this question, you will compute descriptive statistics
for the continuously compounded daily return data on the Vanguard
long term bond index fund (VBLTX), Fidelity Magellan stock mutual
fund (FMAGX), and Starbucks stock (SBUX). Repeat the analysis for
the monthly returns but use instead daily returns computed from end-of-day
adjusted closing prices. 
:::

::: {#exr-05}
In this problem you will analyze the monthly simple
returns on the Vanguard S\&P 500 index (VFINX) and Amazon stock (AMZN)
over the 5-year period January 2010, through January 2015. For this
period there are T=60 monthly returns. The monthly prices for VFINX
are in the **IntroCompFinR** object `VangardPrices` and
the daily prices for AMZN are in the **IntroCompFinR** object
`amznDailyPrices`. 

1. Compute simple monthly returns on AMZN and VFINX and plot these returns together in a single plot. Do the monthly returns from the two assets look like realizations from a covariance stationary stochastic process? Why or why not?
2. Compare and contrast the return characteristics of the two assets. In addition, comment on any common features, if any, of the two return series. 
3. Plot the cumulative simple returns (equity curve) for each fund, which represents the growth of \$1 invested in each fund over the sample period, and plot these returns in the same graph. Which fund performed better over the sample? 
4. Using the **IntroCompFinR** function `fourPanelPlot()`, create descriptive statistics plots for each asset. In addition, compute Based on statistics, do the returns on VFINX and AMZN look normally distributed? Is there any evidence of linear time dependence? Briefly justify your answer.
5. Which asset appears to be riskier? Briefly justify your answer. 
:::

::: {#exr-descriptive-statistics-for-weekly-data  name="Descriptive statistics for weekly data"}
* Create weekly time series for MSFT, SP500

* Repeat graphical and numerical descriptive statistics

* Deduce stylized facts for weekly data
:::

::: {#exr-rolling-descriptive-statistics-for-daily-data  name="Rolling descriptive statistics for daily data"}
TBD
:::

::: {#exr-06}

Loading data and computing returns

Load the daily price data from **IntroCompFinR**, and create monthly returns over the period Jan 1998 through Dec 2014:

```{r}
data(amznDailyPrices, baDailyPrices, costDailyPrices, jwnDailyPrices, sbuxDailyPrices)
fiveStocks = merge(amznDailyPrices, baDailyPrices, costDailyPrices, jwnDailyPrices, sbuxDailyPrices)
fiveStocks = to.monthly(fiveStocks, OHLC=FALSE)
```

Let's look at the data

```{r}
head(fiveStocks, n=3)
```

```{r}
tail(fiveStocks, n=3)
```

Next, let's compute monthly simple returns using the **PerformanceAnalytics** function `Return.Calculate()`

```{r}
fiveStocksRet = na.omit(Return.calculate(fiveStocks, method = "simple"))
head(fiveStocksRet, n=3)
```

I removed the missing January return using the function `na.omit()`.

Part I: Univariate Graphical Analysis

1.	Make time plots of the return data using the **zoo** function `plot.zoo()`. Comment on any relationships between the returns suggested by the plots. 

2. Make a cumulative return plot (equity curve) showing the future value of $1 invested in each asset. Which assets gave the best and worst future values over the investment horizon? Hint: you can use the `cumprod()` function or the **PerformanceAnalytics** function `chart.CumReturns()`. 

3. For each return series, make a four panel plot containing a histogram, boxplot, normal QQ-plot and SACF. Do the return series look normally distributed? Is there any evidence of time dependence? Briefly compare the return distributions. Hint: the `fourPanelPlot()` function only works on one asset at a time. So call the function for each asset. For example, `fourPanelPlot(fiveStocksRet[, "AMZN])`

Part II: Univariate Numerical Summaries

1. Compute numerical descriptive statistics for all assets using the R functions `mean()`, `var()`, `stdev()`, and the **PerformanceAnalytics** functions `skewness()` and `kurtosis()`. Compare and contrast the descriptive statistics for the five assets. Which asset appears to be the riskiest asset?

2. Using the monthly mean return for each asset, compute an estimate of the mean annual return. Hint: use the square root of time rule for cc returns. (The square root of time rule is exact for continuously compounded returns but can be applied as an approximation to simple returns.)

3. Using the monthly return standard deviation for each asset, compute an estimate of the annual return standard deviation. Hint: use the square root of time rule for cc returns. (The square root of time rule is exact for continuously compounded returns but can be applied as an approximation to simple returns.)

Part III: Historical VaR

1.	For each asset compute the empirical $1\%$ and $5\%$ quantiles of the returns. Using these quantiles compute the $1\%$ and $5\%$ historical (monthly) VaR values based on an initial $100,000 investment. Which asset has the highest and lowest VaR values? Are you surprised?

Part IV. Bivariate Graphical Analysis

1.	Use the `pairs()` function to create all pair-wise scatter plots of returns. Comment on the direction and strength of the linear relationships in these plots.

2.	Use the **corrplot** function `corrplot.mixed()` to plot the correlation matrix of the returns on the five assets. Comment on what you see.

Part V. Bivariate Numerical Summary Statistics

Use the R functions `cov()`, and `cor()` to compute the sample covariance matrix and sample correlation matrix of the returns. Comment on the direction and strength of the linear relationships suggested by the values of the covariances and correlations.

Part VI. Stylized Facts

1. Based on your analysis of the five Northwest stocks, what stylized facts do you see?

:::






