# Estimation of The GWN Model {#sec-Estimation-of-The-GWN-Model}

Updated: September 5, 2026

Copyright © Eric Zivot 2026

The GWN model of asset returns presented in the previous chapter gives
you a simple framework for interpreting the time series behavior of
asset returns and prices. At the beginning of time $t-1$, $\mathbf{R}_{t}$
is an $N\times1$ random vector representing the returns (simple or
continuously compounded) on assets $i=1,\ldots,N$ to be realized
at time $t$. The GWN model states that $\mathbf{R}_{t}\sim iid~N(\boldsymbol{\mu},\boldsymbol{\Sigma})$.
Your best guess for the return at $t$ on asset $i$ is $E[R_{it}]=\mu_{i}$,
your measure of uncertainty about our best guess is captured by $\mathrm{sd}(R_{it})=\sigma_{i}$,
and your measures of the direction and strength of linear association
between $R_{it}$ and $R_{jt}$ are $\sigma_{ij}=\mathrm{cov}(R_{it},R_{jt})$
and $\rho_{ij}=\mathrm{cor}(R_{it},R_{jt})$, respectively. The GWN
model assumes that the economic environment is constant over time (i.e., covariance stationary)
so that the multivariate normal distribution characterizing returns
is the same for all time periods $t$. 


Your life would be very easy if you knew the exact values of $\mu_{i},\sigma_{i}^{2}$,
$\sigma_{ij}$ and $\rho_{ij}$: the parameters of the GWN model.
Then you could use the GWN model for risk and portfolio analysis. In
actuality, however, you do not know these values with certainty. Therefore,
a key task in *financial econometrics* is estimating these values from
a history of observed return data. Given estimates of the GWN model
parameters you can then apply the model to risk and portfolio analysis.


Suppose you observe returns on $N$ different assets over the sample
$t=1,\ldots,T$. Denote this sample $\{\mathbf{r}_{1},\ldots,\mathbf{r}_{T}\}=\{\mathbf{r}_{t}\}_{t=1}^{T}$,
where $\mathbf{r}_{t}=({r}_{1t},\ldots,r_{Nt})^{\prime}$
is the $N\times1$ vector of returns on $N$ assets observed at time
$t$. It is assumed that the observed returns are realizations of
the random variables $\{\mathbf{R}_{t}\}_{t=1}^{T}$, where $\mathbf{R}_{t}=(R_{1t},\ldots,R_{Nt})^{\prime}$
is a vector of $N$ asset returns described by the GWN model: 

$$
\mathbf{R}_{t}=\boldsymbol{\mu}+\boldsymbol{\varepsilon}_{t},~\boldsymbol{\varepsilon}_{t}\sim \mathrm{GWN}(\mathbf{0},\boldsymbol{\Sigma}).
$$ {#eq-GWN-model-matrix}

Under these assumptions, you can use the observed returns $\{\mathbf{r}_{t}\}_{t=1}^{T}$
to estimate the unknown parameters in $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$
of the GWN model. However, before you describe the estimation of the
GWN model in detail, it is necessary to review some fundamental concepts
in the statistical theory of estimation. This is important because
estimates of parameters have *estimation error* and statistical theory
allows you to evaluate characteristics of estimation error and to develop
quantitative measures of the typical magnitude of estimation error.
As you shall see, some parameters of the GWN model are precisely estimated
(have small estimation error) and some parameters are not (have large estimation
error).


The chapter is outlined as follows. In @sec-Estimators-and-Estimates
I review some of the statistical theory of estimation and discusses
some important finite sample properties of estimators such as bias
and precision. For many estimators it is difficult to derive exact finite
sample properties which motivates the need to look at asymptotic properties
of estimators. These are properties that hold exactly as the sample
size goes to infinity but are used as approximations for a finite
sample size. I present estimators of the GWN model parameters
in @sec-Estimators-for-the-Parameters-of-the-GWN-Model
and I investigate the statistical properties of these estimators in
@sec-Statistical-Properties-of-GWN-Model-Estimates. In @sec-Using-Monte-Carlo-Simulation 
I illustrates how Monte Carlo simulation can be used to evaluate and understand the statistical
properties of the GWN model estimators. 

To be added:

* Which asymptotic distribution formulas depend on the normality assumption? Make 
clear which results depend on GWN model and which results do not.  

In this chapter I use the R packages **introcompfinr**, **mvtnorm**,
and **PerformanceAnalytics**. Make sure these packages are installed
and loaded prior to running the R examples in the chapter.

```{r}
#| include: false
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(mvtnorm))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```

## Estimators and Estimates {#sec-Estimators-and-Estimates}

Let $R_{t}$ be the return on a single asset (simple or continuously compounded) described by the GWN
model and let $\theta$ denote some characteristic (parameter) of
the GWN model you are interested in estimating. For simplicity, assume
that $\theta\in\mathbb{R}$ is a single parameter. For example, if
you are interested in the expected return on the asset, then $\theta=\mu$;
if you are interested in the variance of the asset returns, then $\theta=\sigma^{2}$;
if you are interested in the first lag autocorrelation of returns then $\theta=\rho_{1}$.
The goal is to estimate $\theta$ based on a sample of size $T$ of the observed data when
you believe the data is generated from the GWN return model.

::: {#def-estimator name="Estimator"}
Let $\{R_{1},\ldots,R_{T}\}$ denote a collection
of $T$ random returns from the GWN model, and let $\theta$ denote
some characteristic of the model. An *estimator*\index{Estimator} of $\theta$, denoted
$\hat{\theta}$, is a rule or algorithm for estimating $\theta$ as
a function of the random variables $\{R_{1},\ldots,R_{T}\}$. Here,
$\hat{\theta}$ is a random variable. 
:::


::: {#def-estimate name="Estimate"}
Let $\{r_{1},\ldots,r_{T}\}$ denote an observed
sample of size $T$ from the GWN model, and let $\theta$ denote some
characteristic of the model. An *estimate*\index{Estimate} of $\theta$, denoted $\hat{\theta}$,
is simply the value of the estimator for $\theta$ based on the observed
sample $\{r_{1},\ldots,r_{T}\}$. Here, $\hat{\theta}$ is a number.
:::


::: {#exm-example0701 name="The sample average as an estimator and an estimate"}
Let $R_{t}$ be the return on a single asset described by the GWN
model, and suppose you are interested in estimating $\theta=\mu=E[R_{t}]$
from the sample of observed returns $\{r_{t}\}_{t=1}^{T}$. The sample
average is an algorithm for computing an estimate of the expected return $\mu$. Before the
sample is observed, you can think of $\hat{\mu}=\frac{1}{T}\sum_{t=1}^{T}R_{t}$
as a simple linear function of the random variables $\{R_{t}\}_{t=1}^{T}$
and so is itself a random variable. You can study the properties of $\hat{\mu}$ using 
the tools of probability theory presented in chapter 2. 

After the sample is observed,
the sample average can be evaluated using the observed data $\{r_{t}\}_{t=1}^{T}$
giving $\hat{\mu}=\frac{1}{T}\sum_{t=1}^{T}r_{t}$, which produces the numerical estimate of $\mu$. For example, suppose $T=5$
and the realized values of the returns are $r_{1}=0.1,r_{2}=0.05,r_{3}=0.025,r_{4}=-0.1,r_{5}=-0.05$.
Then the estimate of $\mu$ using the sample average is: 
$$
\hat{\mu}=\frac{1}{5}(0.1+0.05+0.025+-0.1+-0.05)=0.005.
$$
\hfill{}$\blacksquare$
:::

The example above illustrates the distinction between an estimator
and an estimate of a parameter $\theta$. However, typically in the
statistics literature you use the same symbol, $\hat{\theta}$, to
denote both an estimator and an estimate. When $\hat{\theta}$ is
treated as a function of the random returns it denotes an estimator
and is a random variable. When $\hat{\theta}$ is evaluated using
the observed data it denotes an estimate and is simply a number. The
context in which you discuss $\hat{\theta}$ will determine how to
interpret it.

## Finite Sample Properties of Estimators {#sec-Finite-Sample-Properties-of-Estimators}

Consider $\hat{\theta}$ as a random variable. In general, the pdf
of $\hat{\theta}$, $f(\hat{\theta})$, depends on the pdf's of the
random variables $\{R_{t}\}_{t=1}^{T}$. The exact form of $f(\hat{\theta})$
may be very complicated. Sometimes you can use analytic calculations
to determine the exact form of $f(\hat{\theta})$. This can be done, for example, 
when $\hat{\theta} = \hat{\mu}$. In general, the
exact form of $f(\hat{\theta})$ is often too difficult to derive
exactly. When $f(\hat{\theta})$ is too difficult to compute you can
often approximate $f(\hat{\theta})$ using either *Monte Carlo simulation*
techniques or the *Central Limit Theorem* (CLT). In Monte Carlo
simulation, you use the computer to simulate many different realizations
of the random returns $\{R_{t}\}_{t=1}^{T}$ and on each simulated
sample you evaluate the estimator $\hat{\theta}$. The Monte Carlo
approximation of $f(\hat{\theta})$ is the empirical distribution of
$\hat{\theta}$ over the different simulated samples. For a given
sample size $T$, Monte Carlo simulation gives a very accurate approximation
to $f(\hat{\theta})$ if the number of simulated samples is very large.
The CLT approximation of $f(\hat{\theta})$ is a normal distribution
approximation that becomes more accurate as the sample size $T$ gets
very large. An advantage of the CLT approximation is that it is often
easy to compute. The disadvantage is that the accuracy of the approximation
depends on the estimator $\hat{\theta}$ and sample size $T$. 

For analysis purposes, you often focus on certain characteristics of
$f(\hat{\theta})$, like its expected value (center), variance and
standard deviation (spread about expected value)$.$ The expected
value of an estimator is related to the concept of *estimator bias*,
and the variance/standard deviation of an estimator is related to
the concept of *estimator precision*. Different realizations
of the random variables $\{R_{t}\}_{t=1}^{T}$ will produce different
values of $\hat{\theta}$. Some values of $\hat{\theta}$ will be
bigger than $\theta$ and some will be smaller. Intuitively, a *good estimator*
of $\theta$ is one that is on average correct (unbiased)
and never gets too far away from $\theta$ (small variance). That
is, a good estimator will have small bias and high precision.

### Bias
\index{Bias}

*Bias* concerns the location or center of $f(\hat{\theta})$ in relation
to $\theta$. If $f(\hat{\theta})$ is centered away from $\theta$,
then you say $\hat{\theta}$ is a *biased* estimator of $\theta$.
If $f(\hat{\theta})$ is centered at $\theta$, then you say that $\hat{\theta}$
is an *unbiased* estimator of $\theta$. Formally, you have
the following definitions:

::: {#def-estimation-error name="Estimation Error"}
The *estimation error* is the difference between the estimator and the parameter being estimated: 
$$
\mathrm{error}(\hat{\theta},\theta)=\hat{\theta}-\theta.
$$ {#eq-error}
:::

::: {#def-bias name="bias"}
The *bias* of an estimator $\hat{\theta}$ of $\theta$
is the expected estimation error: 
$$
\mathrm{bias}(\hat{\theta},\theta)=E[\mathrm{error}(\hat{\theta},\theta)]=E[\hat{\theta}]-\theta.
$$ {#eq-bias}
:::

::: {#def-unbiased name="Unbiased Estimator"}
An estimator $\hat{\theta}$ of $\theta$ is *unbiased*
if $\mathrm{bias}(\hat{\theta},\theta)=0$; i.e., if $E[\hat{\theta}]=\theta$
or $E[\mathrm{error}(\hat{\theta},\theta)]=0$. 
:::

Unbiasedness is a desirable property of an estimator. It means that
the estimator produces the correct answer "on average",
where "on average" means over many hypothetical realizations
of the random variables $\{R_{t}\}_{t=1}^{T}$. It is important to
keep in mind that an unbiased estimator for $\theta$ may not be very
close to $\theta$ for a particular sample, and that a biased estimator
may actually be quite close to $\theta$. For example, consider two
estimators of $\theta$, $\hat{\theta}_{1}$ and $\hat{\theta}_{2}$.
The pdfs of $\hat{\theta}_{1}$ and $\hat{\theta}_{2}$ are illustrated
in @fig-CerCompareEstimators. $\hat{\theta}_{1}$
is an unbiased estimator of $\theta$ with a large variance, and $\hat{\theta}_{2}$
is a biased estimator of $\theta$ with a small variance. Consider
first, the pdf of $\hat{\theta}_{1}$. The center of the distribution
is at the true value $\theta=0$, $E[\hat{\theta}_{1}]=0$, but the
distribution is very widely spread out about $\theta=0$. That is,
$\mathrm{var}(\hat{\theta}_{1})$ is large. On average (over many
hypothetical samples) the value of $\hat{\theta}_{1}$ will be close
to $\theta$, but in any given sample the value of $\hat{\theta}_{1}$
can be quite a bit above or below $\theta$. Hence, unbiasedness by
itself does not guarantee a good estimator of $\theta$. Now consider
the pdf for $\hat{\theta}_{2}$. The center of the pdf is slightly
higher than $\theta=0$, i.e., $\mathrm{bias}(\hat{\theta}_{2},\theta)>0$,
but the spread of the distribution is small. Although the value of
$\hat{\theta}_{2}$ is not equal to 0 *on average* you might
prefer the estimator $\hat{\theta}_{2}$ over $\hat{\theta}_{1}$
because it is generally closer to $\theta=0$ *on average*
than $\hat{\theta}_{1}$. 

While unbiasedness is a desirable property of an estimator $\hat{\theta}$
of $\theta$, it, by itself, is not enough to determine if $\hat{\theta}$
is a good estimator. Being correct on average means that $\hat{\theta}$
is seldom exactly correct for any given sample. In some samples $\hat{\theta}$
is less than $\theta$, and some samples $\hat{\theta}$ is greater
than $\theta$. More importantly, you need to know how far $\hat{\theta}$
typically is from $\theta$. That is, you need to know about the magnitude
of the spread of the distribution of $\hat{\theta}$ about its average
value. This will tell you the precision of $\hat{\theta}$.

```{r}
#| label: fig-CerCompareEstimators
#| echo: false
#| fig-cap: Distributions of competing estimators for $\theta=0$. $\hat{\theta}_{1}$
#|   is unbiased but has high variance, and $\hat{\theta}_{2}$ is biased but has low
#|   variance.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.min = -3
x.max = 3
npts = 100
x.vals = seq(from=x.min,to=x.max,length=npts)
pdf1 = dnorm(x.vals,mean=0,sd=1) 
pdf2 = dnorm(x.vals,mean=0.1,sd=0.25)
ylimits = range(pdf1,pdf2)

plot(x.vals,pdf1,ylim=ylimits,type="l",lty=1,ylab="pdf",
     xlab="estimate value", lwd=2)
segments(x0=0, y0=0, x1=0, y1=dnorm(0), lwd=2)
lines(x.vals, pdf2, lty=2, lwd=2, col="red")
# legend(x="topleft",legend=c("theta.hat 1","theta.hat 2"), lty=1:2,
#        lwd=c(2,2), col=c(1,2))
legend(x="topleft",legend=c(expression(f(hat(theta[1]))),expression(f(hat(theta[2])))), lty=1:2,
        lwd=c(2,2), col=c(1,2))
segments(x0=0.1, y0=0, x1=0.1, y1=dnorm(0.1, mean=0.1, sd=0.25), lwd=2, col="red")
```


### Precision
\index{Precision}

An estimate is, hopefully, our best guess of the true (but unknown)
value of $\theta$. Our guess most certainly will be wrong, but you
hope it will not be too far off. A *precise estimate* is one in which
the variability of the estimation error is small. The variability
of the estimation error is captured by the *mean squared error*.

::: {#def-mean-squared-error .definition name="Mean Squared Error"}
The *mean squared error* of an estimator $\hat{\theta}$
of $\theta$ is given by: 
$$
\mathrm{mse}(\hat{\theta},\theta)=E[(\hat{\theta}-\theta)^{2}]=E[\mathrm{error}(\hat{\theta},\theta)^{2}]
$$ {#eq-mse}
:::

The mean squared error measures the expected squared deviation of
$\hat{\theta}$ from $\theta$. If this expected deviation is small,
then you know that $\hat{\theta}$ will almost always be close to $\theta$.
Alternatively, if the mean squared is large then it is possible to
see samples for which $\hat{\theta}$ is quite far from $\theta$.
A useful decomposition of $\mathrm{mse}(\hat{\theta},\theta)$ is:
$$
\mathrm{mse}(\hat{\theta},\theta)=E[(\hat{\theta}-E[\hat{\theta}])^{2}]+\left(E[\hat{\theta}]-\theta\right)^{2}=\mathrm{var}(\hat{\theta})+\mathrm{bias}(\hat{\theta},\theta)^{2}
$$
The derivation of this result is straightforward. Use the *add and subtract trick* and write 
$$
\hat{\theta}-\theta=\hat{\theta}-E[\hat{\theta}]+E[\hat{\theta}]-\theta.
$$
Then square both sides giving 
$$
(\hat{\theta}-\theta)^{2}=\left(\hat{\theta}-E[\hat{\theta}]\right)^{2}+2\left(\hat{\theta}-E[\hat{\theta}]\right)\left(E[\hat{\theta}]-\theta\right)+\left(E[\hat{\theta}]-\theta\right)^{2}.
$$
Taking expectations of both sides yields, 

\begin{align*}
\mathrm{mse}(\hat{\theta},\theta) & =E\left[\left(\hat{\theta}-E[\hat{\theta}]\right)\right]^{2}+2\left(E[\hat{\theta}]-E[\hat{\theta}]\right)\left(E[\hat{\theta}]-\theta\right)+E\left[\left(E[\hat{\theta}]-\theta\right)^{2}\right]\\
 & =E\left[\left(\hat{\theta}-E[\hat{\theta}]\right)\right]^{2}+E\left[\left(E[\hat{\theta}]-\theta\right)^{2}\right]\\
 & =\mathrm{var}(\hat{\theta})+\mathrm{bias}(\hat{\theta},\theta)^{2}.
\end{align*}

The result states that for any estimator $\hat{\theta}$ of $\theta$,
$\mathrm{mse}(\hat{\theta},\theta)$ can be split into a variance
component, $\mathrm{var}(\hat{\theta})$, and a squared bias component, $\mathrm{bias}(\hat{\theta},\theta)^{2}$.
Clearly, $\mathrm{mse}(\hat{\theta},\theta)$ will be small only if
both components are small. If an estimator is unbiased then $\mathrm{mse}(\hat{\theta},\theta)=\mathrm{var}(\hat{\theta})=E[(\hat{\theta}-\theta)^{2}]$
is just the squared deviation of $\hat{\theta}$ about $\theta$.
Hence, an unbiased estimator $\hat{\theta}$ of $\theta$ is *good,*
if it has a small variance. 

The $\mathrm{mse}(\hat{\theta},\theta)$ and $\mathrm{var}(\hat{\theta})$
are based on squared deviations and so are not in the same units of
measurement as $\theta$. Measures of precision that are in the same
units as $\theta$ are the *root mean squared error* and the *standard error*.

::: {#def-root-mean-squared .definition name="Root Mean Squared Error and Standard Error"}
The *root mean squared error* and the *standard error* of an estimator $\hat{\theta}$
are:
$$
\mathrm{rmse}(\hat{\theta},\theta)=\sqrt{\mathrm{mse}(\hat{\theta},\theta)},
$$ {#eq-rmse}
 
$$
\mathrm{se}(\hat{\theta})=\sqrt{\mathrm{var}(\hat{\theta})}.
$$ {#eq-se}
:::

If $\mathrm{bias}(\hat{\theta},\theta)\approx0$ then the precision
of $\hat{\theta}$ is typically measured by $\mathrm{se}(\hat{\theta})$.

With the concepts of bias and precision in hand, I can state what
defines a good estimator.

::: {#def-good-estimator .definition name="Good Estimator"}
A good estimator $\hat{\theta}$ of $\theta$
has a small bias @eq-bias and a small standard error @eq-se.
:::

### Estimated standard errors

Standard error formulas $\mathrm{se}(\hat{\theta})$ that I will derive in this chapter often depend on the unknown values of the GWN model parameters. For example, later on I will show that $\mathrm{se}(\hat{\mu}) = \sigma^2 / T$ which depends on the GWN parameter $\sigma^2$ which is unknown. Hence, the formula for $\mathrm{se}(\hat{\mu})$ is a *practically useless* quantity as I
cannot compute its value because I don't the know the value of $\sigma^2$. Fortunately, I can create a *practically useful* formula by replacing the unknown quantity $\sigma^2$ with a good estimate $\hat{\sigma}^2$ that I compute from the data. This gives rise the *estimated standard error* $\widehat{\mathrm{se}}(\hat{\mu}) = \hat{\sigma}^2 / T$.

## Asymptotic Properties of Estimators {#sec-Asymptotic-Properties-of-Estimators}

Estimator bias and precision are finite sample properties. That is,
they are properties that hold for a fixed sample size $T$. Very often
you are also interested in properties of estimators when the sample
size $T$ gets very large. For example, analytic calculations may
show that the bias and mse of an estimator $\hat{\theta}$ depend
on $T$ in a decreasing way. That is, as $T$ gets very large the
bias and mse approach zero. So for a very large sample, $\hat{\theta}$
is effectively unbiased with high precision. In this case I say that
$\hat{\theta}$ is a *consistent* estimator of $\theta$. In
addition, for large samples the *Central Limit Theorem* (CLT) says that
$f(\hat{\theta})$ can often be well approximated by a normal distribution.
In this case, I say that $\hat{\theta}$ is *asymptotically normally distributed*. 
The word "asymptotic" means" in an infinitely large sample" or "as the sample
size $T$ goes to infinity" . Of course, in the real
world I don't have an infinitely large sample and so the asymptotic
results are only approximations. How good these approximations are
for a given sample size $T$ depends on the context. Monte Carlo simulations
(see @sec-Using-Monte-Carlo-Simulation) can often be
used to evaluate asymptotic approximations in a given context.

### Consistency
\index{Consistency}

Let $\hat{\theta}$ be an estimator of $\theta$ based on the random
returns $\{R_{t}\}_{t=1}^{T}$. 

::: {#def-consistent-estimator .definition name="Consistent Estimator"}
$\hat{\theta}$ is consistent for $\theta$ (converges
in probability to $\theta$) if for any $\varepsilon>0$: 
$$
\lim_{T\rightarrow\infty}\Pr(|\hat{\theta}-\theta|>\varepsilon)=0.
$$
:::

Intuitively, consistency says that as you get enough data then $\hat{\theta}$
will eventually equal $\theta$. In other words, if you have enough
data then you know the truth. 

Theorems in probability theory known as *Laws of Large Numbers*
are used to determine if an estimator is consistent or not. In general,
you have the following result.

::: {#prp-estimator-consistency .proposition name="Estimator Consistency"}
An estimator $\hat{\theta}$ is consistent for $\theta$ if:

-  $\mathrm{bias}(\hat{\theta},\theta)=0$ as $T\rightarrow\infty$. 
-  $\mathrm{se}(\hat{\theta})=0$ as $T\rightarrow\infty$. 

Equivalently, $\hat{\theta}$ is consistent for $\theta$ if $\mathrm{mse}(\hat{\theta},\theta)\rightarrow0$
as $T\rightarrow\infty$. 
:::

Intuitively, if $f(\hat{\theta})$ collapses
to $\theta$ as $T\rightarrow\infty$ then $\hat{\theta}$ is consistent
for $\theta$.

### Asymptotic Normality
\index{Asymptotic normality}

Let $\hat{\theta}$ be an estimator of $\theta$ based on the random
returns $\{R_{t}\}_{t=1}^{T}$.

::: {#def-asymptotically-normal-estimator .definition name="Asymptotically Normal Estimator"}
An estimator $\hat{\theta}$ is *asymptotically normally distributed* if: 
$$
\hat{\theta}\sim N(\theta,\mathrm{se}(\hat{\theta})^{2})
$$ {#eq-asymptotic-normal-distribution}
for large enough $T$. 
:::

Asymptotic normality means that $f(\hat{\theta})$ is well approximated
by a normal distribution with mean $\theta$ and variance $\mathrm{se}(\hat{\theta})^{2}$.
The justification for asymptotic normality comes from the famous *Central Limit Theorem*.

In the definition of an asymptotically normal estimator, the variance of the normal distribution, $\mathrm{se}(\hat{\theta})^{2}$, often depends on unknown GWN model parameters and so is practically useless. Fortunately, you can create a practically useful result if you replace the unknown parameters in $\mathrm{se}(\hat{\theta})^{2}$ with consistent estimates. 

::: {#prp-practically-useful-asymptotic-normality .proposition name="Practically Useful Asymptotic Normality"}
Supoose an estimator $\hat{\theta}$ is asymptotically normally distributed with variance $\mathrm{se}(\hat{\theta})^{2}$ that depends on unknown GWN model parameters. If you replace these unknown parameters with consistent estimates, creating the estimated variance $\widehat{\mathrm{se}}(\hat{\theta})^{2}$, then 
$$
\hat{\theta}\sim N(\theta,\widehat{\mathrm{se}}(\hat{\theta})^{2})
$$ {#eq-asymptotic-normal-distribution2}
for large enough $T$.
:::

### Central Limit Theorem
\index{Central Limit Theorem}

There are actually many versions of the CLT with different assumptions.
^[White (1984) gives a comprehensive discussion of CLTs useful in econometrics.]
In its simplest form, the CLT says that the sample average of a collection
of iid random variables $X_{1},\ldots,X_{T}$ with $E[X_{i}]=\mu$
and $var(X_{i})=\sigma^{2}$ is asymptotically normal with mean $\mu$
and variance $\sigma^{2}/T$. In particular, the CDF of the standardized
sample mean: 
$$
\frac{\bar{X}-\mu}{\mathrm{se}(\bar{X})}=\frac{\bar{X}-\mu}{\sigma/\sqrt{T}}=\sqrt{T}\left(\frac{\bar{X}-\mu}{\sigma}\right),
$$
converges to the CDF of a standard normal random variable $Z$ as
$T\rightarrow\infty$. This result can be expressed as: 
$$
\sqrt{T}\left(\frac{\bar{X}-\mu}{\sigma}\right)\sim Z\sim N(0,1),
$$
for large enough $T$. Equivalently, 
$$
\bar{X}\sim\mu+\frac{\sigma}{\sqrt{T}}\times Z\sim N\left(\mu,\frac{\sigma^{2}}{T}\right)=N\left(\mu,\mathrm{se}(\bar{X})^{2}\right),
$$
for large enough $T$. This form shows that $\bar{X}$ is asymptotically
normal with mean $\mu$ and variance $\mathrm{se}(\bar{X})^{2} = \sigma^{2}/T$.

::: {.callout-note}
## Remarks
1. The CLT result is truly remarkable because the iid random variables $X_i$ can come from *any distribution* that has mean $\mu$ and variance $\sigma^2$. For example, $X_i$ could be binomial, Student's t with df > 2, chi-square, etc. 

2. While the basic CLT shows that the sample average $\bar{X}$ is asymptotically normally distributed, it can be extended to general estimators of parameters of the GWN model because, as you shall see, these estimators are essentially averages of iid random variables. 

3. The iid assumption of the basic CLT can be relaxed to allow $\{X_i\}$ to be covariance stationary and ergodic. This allows the asymptotic normality result to be extended to estimators computed from time series data that could be serially correlated. 

4. The CLT result depends on knowing $\sigma^2$, which is practically useless because you almost never know $\sigma^2$. However, the CLT result holds if you replace $\sigma^2$ with a consistent estimate $\hat{\sigma}^2$. 

:::

### Asymptotic Confidence Intervals
\index{Asymptotic confidence intervals}

For an asymptotically normal estimator $\hat{\theta}$ of $\theta$,
the precision of $\hat{\theta}$ is measured by $\widehat{\mathrm{se}}(\hat{\theta})$
but is best communicated by computing a (asymptotic) *confidence
interval* for the unknown value of $\theta$. A confidence interval
is an *interval estimate* of $\theta$ such that you can put an
explicit probability statement about the likelihood that the interval
covers $\theta$. 

The construction of an asymptotic confidence interval for $\theta$
uses the asymptotic normality result: 
$$
\frac{\hat{\theta}-\theta}{\widehat{\mathrm{se}}(\hat{\theta})}=Z\sim N(0,1).
$$ {#eq-z-ratio}
Then, for $\alpha\in(0,1)$, you compute a $(1-\alpha)\cdot100\%$
confidence interval for $\theta$ using @eq-z-ratio and the
$1-\alpha/2$ standard normal quantile (critical value) $q_{(1-\alpha/2)}^{Z}$
to give: 
$$
\Pr\left(-q_{(1-\alpha/2)}^{Z}\leq\frac{\hat{\theta}-\theta}{\widehat{\mathrm{se}}(\hat{\theta})}\leq q_{(1-\alpha/2)}^{Z}\right)=1-\alpha,
$$
which can be rearranged as, 
$$
\Pr\left(\hat{\theta}-q_{(1-\alpha/2)}^{Z}\cdot\widehat{\mathrm{se}}(\hat{\theta})\leq\mu_{i}\leq\hat{\theta}+q_{(1-\alpha/2)}^{Z}\cdot\widehat{\mathrm{se}}(\hat{\theta})\right)=1-\alpha.
$$
Hence, the random interval, 
$$
[\hat{\theta}-q_{(1-\alpha/2)}^{Z}\cdot\widehat{\mathrm{se}}(\hat{\theta}),~\hat{\theta}+q_{(1-\alpha/2)}^{Z}\cdot\widehat{\mathrm{se}}(\hat{\theta})]=\hat{\theta}\pm q_{(1-\alpha/2)}^{Z}\cdot\widehat{\mathrm{se}}(\hat{\theta})
$$ {#eq-asymp-ci}
covers the true unknown value of $\theta$ with probability $1-\alpha$.

In practice, typical values for $\alpha$ are 0.05 and 0.01 for which
$q_{(0.975)}^{Z}=1.96$ and $q_{(0.995)}^{Z}=2.58$. Then, approximate
95\% and 99\% asymptotic confidence intervals for $\theta$ have the
form $\hat{\theta}\pm2\cdot\widehat{\mathrm{se}}(\hat{\theta})$ and $\hat{\theta}\pm2.5\cdot\widehat{\mathrm{se}}(\hat{\theta})$,
respectively.

## Estimators for the Parameters of the GWN Model {#sec-Estimators-for-the-Parameters-of-the-GWN-Model}

Let $\{\mathbf{r}_{t}\}_{t=1}^{T}$ denote a sample of size $T$ of
observed returns on $N$ assets from the GWN model @eq-GWN-model-matrix.
To estimate the unknown GWN model parameters $\mu_{i},\sigma_{i}^{2}$,
$\sigma_{ij}$ and $\rho_{ij}$ from $\{\mathbf{r}_{t}\}_{t=1}^{T}$
you can use the *plug-in principle* from statistics: 

::: {#def-plug-in-principle .definition name="Plug-in-Principle"}
The *plug-in-principle* says to estimate model parameters using corresponding
sample statistics. 
:::

::: {#prp-plug-in-principle-estimates-for-gwn-model-parameters .proposition name="Plug-in Principle Estimates for GWN Model Parameters"}
For the GWN model parameters, the plug-in principle estimates are
the following sample descriptive statistics discussed in Chapter @sec-Descriptive-Statistics-for-Financial-Data:

$$
\hat{\mu}_{i} =\frac{1}{T}\sum_{t=1}^{T}r_{it},
$$ {#eq-muhat}
$$
\hat{\sigma}_{i}^{2} =\frac{1}{T-1}\sum_{t=1}^{T}(r_{it}-\hat{\mu}_{i})^{2},
$$ {#eq-sigma2hat}
$$
\hat{\sigma}_{i} =\sqrt{\hat{\sigma}_{i}^{2}},
$$ {#eq-sigmahat}
$$
\hat{\sigma}_{ij} =\frac{1}{T-1}\sum_{t=1}^{T}(r_{it}-\hat{\mu}_{i})(r_{jt}-\hat{\mu}_{j}),
$$ {#eq-covhat}
$$
\hat{\rho}_{ij} =\frac{\hat{\sigma}_{ij}}{\hat{\sigma}_{i}\hat{\sigma}_{j}}.
$$ {#eq-rhohat}
:::

The plug-in principle is appropriate because the GWN model parameters
$\mu_{i},\sigma_{i}^{2}$, $\sigma_{ij}$ and $\rho_{ij}$ are characteristics
of the underlying distribution of returns that are naturally estimated
using sample statistics. 

The plug-in principle sample statistics @eq-muhat - @eq-rhohat
are given for a single asset and the statistics @eq-covhat
- @eq-rhohat are given for one pair of assets. However, these
statistics can be computed for a collection of $N$ assets using the
matrix sample statistics: 

$$
\underset{(N\times1)}{\hat{\mu}}  =\frac{1}{T}\sum_{t=1}^{T}\mathbf{r}_{t}=\left(\begin{array}{c}
\hat{\mu}_{1}\\
\vdots\\
\hat{\mu}_{N}
\end{array}\right),
$$ {#eq-muhat-vector}
$$
\underset{(N\times N)}{\hat{\Sigma}}  =\frac{1}{T-1}\sum_{t=1}^{T}(\mathbf{r}_{t}-\hat{\mu})(\mathbf{r}_{t}-\hat{\mu})^{\prime}=\left(\begin{array}{cccc}
\hat{\sigma}_{1}^{2} & \hat{\sigma}_{12} & \cdots & \hat{\sigma}_{1N}\\
\hat{\sigma}_{12} & \hat{\sigma}_{2}^{2} & \cdots & \hat{\sigma}_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\hat{\sigma}_{1N} & \hat{\sigma}_{2N} & \cdots & \hat{\sigma}_{N}^{2}
\end{array}\right).
$$ {#eq-Sigmahat-matrix}

Here $\hat{\mu}$ is called the *sample mean vector* and $\hat{\Sigma}$
is called the *sample covariance matrix*. The sample variances are the
diagonal elements of $\hat{\Sigma}$, and the sample covariances
are the off diagonal elements of $\hat{\Sigma}.$ To get
the sample correlations, define the $N\times N$ diagonal matrix:
$$
\mathbf{\hat{D}}=\left(\begin{array}{cccc}
\hat{\sigma}_{1} & 0 & \cdots & 0\\
0 & \hat{\sigma}_{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & \hat{\sigma}_{N}
\end{array}\right).
$$

Then the sample correlation matrix $\mathbf{\hat{C}}$ is computed
as:

$$
\mathbf{\hat{C}}=\mathbf{\hat{D}}^{-1}\mathbf{\hat{\Sigma}}\mathbf{\hat{D}}^{-1} = \left(\begin{array}{cccc}
1 & \hat{\rho}_{12} & \cdots & \hat{\rho}_{1N}\\
\hat{\rho}_{12} & 1 & \cdots & \hat{\rho}_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\hat{\rho}_{1N} & \hat{\rho}_{2N} & \cdots & 1
\end{array}\right).
$$ {#eq-Rhat}  

Here, the sample correlations are the off diagonal elements of $\mathbf{\hat{C}}$.

::: {#exm-estimating-the-gwn-model-parameters-for-microsoft-starbucks-and-the-s-p500-index .example name="Estimating the GWN model parameters for Microsoft Starbucks and the S&P500 index"}
To illustrate typical estimates of the GWN model parameters, I use
data on monthly simple and continuously compounded (cc) returns for Microsoft, Starbucks
and the S \& P 500 index over the period January 1998 through May
2012 from the **introcompfinr** package. This is same data used in Chapter @sec-Descriptive-Statistics-for-Financial-Data and is constructed as follows:

```{r}
#| echo: true

msftPrices = to.monthly(msftDailyPrices, OHLC=FALSE) 
sp500Prices = to.monthly(sp500DailyPrices, OHLC=FALSE) 
sbuxPrices = to.monthly(sbuxDailyPrices, OHLC=FALSE) 
# set sample to match examples in other chapters
smpl = "1998-01::2012-05" 
msftPrices = msftPrices[smpl] 
sp500Prices = sp500Prices[smpl] 
sbuxPrices = sbuxPrices[smpl] 
# calculate returns 
msftRetS = na.omit(Return.calculate(msftPrices, method="simple"))
sp500RetS = na.omit(Return.calculate(sp500Prices, method="simple"))
sbuxRetS = na.omit(Return.calculate(sbuxPrices, method="simple"))
msftRetC = log(1 + msftRetS) 
sp500RetC = log(1 + sp500RetS) 
sbuxRetC = log(1 + sbuxRetS) 
# merged data set 
gwnRetS = merge(msftRetS, sbuxRetS, sp500RetS) 
gwnRetC = merge(msftRetC, sbuxRetC, sp500RetC) 
colnames(gwnRetS) = colnames(gwnRetC) = c("MSFT", "SBUX", "SP500") 
```

These data are illustrated in @fig-CerModelEstimation3returns
and @fig-CerModelEstimation3returnsScatter. 

```{r}
#| label: fig-CerModelEstimation3returns
#| echo: false
#| results: hide
#| cache: true
#| fig-cap: Monthly cc returns on Microsoft stock, Starbucks stock, and the S\&P 500
#|   index, over the period January 1998 through May 2012.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# panel function for drawing horizontal line at zero in each panel
# my.panel <- function(...) {
#   lines(...)
#   abline(h=0)
# }
# plot.zoo(gwnRetC, col="blue", lwd=2, main="", panel=my.panel)
plot(gwnRetC, multi.panel=TRUE, main="",
     col="blue", lwd=2, yaxis.same=TRUE,
     major.ticks="years", grid.ticks.on="years")
```

```{r}
#| label: fig-CerModelEstimation3returnsScatter
#| echo: false
#| fig-cap: Monthly cc returns on Microsoft stock, Starbucks stock, and the S\&P 500
#|   index, over the period January 1998 through May 2012.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

pairs(coredata(gwnRetC), col="blue", cex=1.25, pch=16)
```

The estimates of $\mu_{i}$ ($i=\textrm{msft,sbux,sp500})$ using
@eq-muhat or @eq-muhat-vector can be computed using
the R functions `apply()` and `mean()`:

```{r}
#| echo: true

muhat = apply(gwnRetC,2,mean)
muhat
```

Starbucks has the highest average monthly return at 1.5\% and the
S\&P 500 index has the lowest at 0.2\%. 


The estimates of the parameters $\sigma_{i}^{2},\sigma_{i}$, using
@eq-sigma2hat and @eq-sigmahat can be computed using
`apply()`, `var()` and `sd()`:

```{r}
#| echo: true

sigma2hat = apply(gwnRetC,2,var)
sigmahat = apply(gwnRetC,2,sd)
rbind(sigma2hat, sigmahat)
```

Starbucks has the most variable monthly returns at 11\%, and the S\&P
500 index has the smallest at 5\%. 

The scatterplots of the returns are illustrated in @fig-CerModelEstimation3returnsScatter.
All returns appear to be positively related. The covariance and correlation
matrix estimates using @eq-Sigmahat-matrix and @eq-Rhat can
be computed using the functions `var()` (or `cov()`)
and `cor()`:

```{r}
#| echo: true

covmat = var(gwnRetC)
covmat
cormat = cor(gwnRetC)
cormat
```

To extract the unique pairwise values of $\sigma_{ij}$ and $\rho_{ij}$
from the matrix objects `covmat` and `cormat` use:

```{r}
#| echo: true

covhat = covmat[lower.tri(covmat)]
rhohat = cormat[lower.tri(cormat)]
names(covhat) <- names(rhohat) <- 
c("msft,sbux","msft,sp500","sbux,sp500")
covhat
rhohat
```

The pairs (MSFT, SP500) and (SBUX, SP500) are the most correlated.
These estimates confirm the visual results from the scatterplot matrix
in @fig-CerModelEstimation3returnsScatter.

\hfill{}$\blacksquare$
:::

## Statistical Properties of the GWN Model Estimates {#sec-Statistical-Properties-of-GWN-Model-Estimates}

To determine the statistical properties of plug-in principle estimators
$\hat{\mu}_{i},\hat{\sigma}_{i}^{2},\hat{\sigma}_{i},\hat{\sigma}_{ij}$
and $\hat{\rho}_{ij}$ in the GWN model, you treat them as functions
of the random variables $\{\mathbf{R}_{t}\}_{t=1}^{T}$ where $\mathbf{R}_{t}$
is assumed to be generated by the GWN model @eq-GWN-model-matrix.

### Bias
\index{Bias}

::: {#prp-bias-of-gwn-model-estimators .proposition name="Bias of GWN Model Estimators"}
Assume that returns are generated by the GWN model @eq-GWN-model-matrix. Then the estimators
$\hat{\mu}_{i},$ $\hat{\sigma}_{i}^{2}$ and $\hat{\sigma}_{ij}$
are unbiased estimators:^[The matrix sample statistics @eq-muhat-vector and @eq-Sigmahat-matrix
are unbiased estimators of $\mu$ and $\Sigma$,
respectively.]
\begin{align*}
E[\hat{\mu}_{i}] & =\mu_{i}\\
E[\hat{\sigma}_{i}^{2}] & =\sigma_{i}^{2},\\
E[\hat{\sigma}_{ij}] & =\sigma_{ij}.
\end{align*}
The estimators $\hat{\sigma}_{i}$ and $\hat{\rho}_{ij}$ are biased estimators:
\begin{align*}
E[\hat{\sigma}_{i}] & \neq\sigma_{i},\\
E[\hat{\rho}_{ij}] & \neq\rho_{ij}.
\end{align*}
:::

It can be shown that the biases in $\hat{\sigma}_{i}$ and $\hat{\rho}_{ij}$,
are very small and decreasing in $T$ such that $\mathrm{bias}(\hat{\sigma}_{i},\sigma_{i})=\mathrm{bias}(\hat{\rho}_{ij},\rho_{ij})=0$
as $T\rightarrow\infty$. The proofs of these results are beyond the
scope of this book and may be found, for example, in Goldberger (1991).
As you shall see, these results about bias can be easily verified using
Monte Carlo methods. 

It is instructive to illustrate how to derive the result $E[\hat{\mu}_{i}]=\mu_{i}$.
Using $\hat{\mu}_{i} = \frac{1}{T}\sum_{t=1}^{T}R_{it}$ and results about the expectation of a linear combination of random variables, it follows that: 
\begin{align*}
E[\hat{\mu}_{i}] & =E\left[\frac{1}{T}\sum_{t=1}^{T}R_{it}\right]\\
  & =\frac{1}{T}\sum_{t=1}^{T} E[R_{it}]~ (\textrm{by the linearity of }E[\cdot])\\
 & =\frac{1}{T}\sum_{t=1}^{T}\mu_{i}~ (\textrm{since }E[R_{it}]=\mu_{i},~t=1,\ldots,T)\\
 & =\frac{1}{T}T\cdot\mu_{i}=\mu_{i}.
\end{align*}
The derivation of the results $E[\hat{\sigma}_{i}^{2}]=\sigma_{i}^{2}$
and $E[\hat{\sigma}_{ij}]=\sigma_{ij}$ are similar but are considerably
more involved and so are omitted.

### Precision
\index{Precision}
Because the GWN model estimators are either unbiased or the bias is
very small, the precision of these estimators is measured by their
standard errors. Here, I give the mathematical formulas for their standard errors. 

<!-- ::: {.proposition name="Standard error for $\\hat{\\mu_i}$"} -->
::: {#prp-standard-error-for-muhat .proposition name="Standard error for $\\hat{\\mu_i}$"}
The standard error for $\hat{\mu}_{i},$ $\mathrm{se}(\hat{\mu}_{i}),$
can be calculated exactly and is given by: 
$$
\mathrm{se}(\hat{\mu}_{i})=\frac{\sigma_{i}}{\sqrt{T}}.
$$ {#eq-semuhat}
:::

The derivation of this result is straightforward. Using the results
for the variance of a linear combination of uncorrelated random variables,
you have: 
\begin{align*}
\mathrm{var}(\hat{\mu}_{i}) & =\mathrm{var}\left(\frac{1}{T}\sum_{t=1}^{T}R_{it}\right)\\
 & =\frac{1}{T^{2}}\sum_{t=1}^{T}\mathrm{var}(R_{it})~ (\textrm{since }R_{it}\textrm{ is independent over time})\\
 & =\frac{1}{T^{2}}\sum_{t=1}^{T}\sigma_{i}^{2}~ (\textrm{since }\mathrm{var}(R_{it})=\sigma_{i}^{2},~t=1,\ldots,T)\\
 & =\frac{1}{T^{2}}T\sigma_{i}^{2}\\
 & =\frac{\sigma_{i}^{2}}{T}.
\end{align*}

Then $\mathrm{se}(\hat{\mu}_{i})=\mathrm{sd}(\hat{\mu}_{i})=\frac{\sigma_{i}}{\sqrt{T}}$.
 
::: {.callout-note}
## Remarks
1. The value of $\mathrm{se}(\hat{\mu}_{i})$ is in the same units as $\hat{\mu}_{i}$ and measures the precision of $\hat{\mu}_{i}$ as an estimate. If $\mathrm{se}(\hat{\mu}_{i})$ is small relative to $\hat{\mu}_{i}$ then $\hat{\mu}_{i}$ is a relatively precise estimate of $\mu_{i}$ because $f(\hat{\mu}_{i})$ will be tightly concentrated around $\mu_{i}$; if $\mathrm{se}(\hat{\mu}_{i})$ is large relative to $\mu_{i}$ then $\hat{\mu}_{i}$ is a relatively imprecise estimate of $\mu_{i}$ because $f(\hat{\mu}_{i})$ will be spread out about $\mu_{i}$. 
2. The magnitude of $\mathrm{se}(\hat{\mu}_{i})$ depends positively on the volatility of returns, $\sigma_{i}=\mathrm{sd}(R_{it})$. For a given sample size $T$, assets with higher return volatility have larger values of $\mathrm{se}(\hat{\mu}_{i})$ than assets with lower return volatility. In other words, estimates of expected return for high volatility assets are less precise than estimates of expected returns for low volatility assets. 
3. For a given return volatility $\sigma_{i}$, $\mathrm{se}(\hat{\mu}_{i})$ is smaller for larger sample sizes $T$. In other words, $\hat{\mu}_{i}$ is more precisely estimated for larger samples. 
4. $\mathrm{se}(\hat{\mu}_{i})\rightarrow0$ as $T\rightarrow\infty$. This together with $E[\hat{\mu_{i}}]=\mu_{i}$ shows that $\hat{\mu_{i}}$ is consistent for $\mu_{i}$. 
:::

The derivations of the standard errors for $\hat{\sigma}_{i}^{2},\hat{\sigma}_{i}$,
$\hat{\sigma}_{ij}$ and $\hat{\rho}_{ij}$ are complicated, and the
exact results are extremely messy and hard to work with. However,
there are simple approximate formulas for the standard errors of $\hat{\sigma}_{i}^{2},\hat{\sigma}_{i}$, $\hat{\sigma_{ij}}$, and $\hat{\rho}_{ij}$ based on the CLT under the GWN model that are valid if the sample
size, $T$, is reasonably large. 

::: {#prp-approximate-standard-error-formulas .proposition name="Approximate standard error formulas for $\\hat{\\sigma}_{i}^{2}$, $\\hat{\\sigma}_{i}$, $\\hat{\\sigma}_{ij}$, and $\\hat{\\rho}_{ij}$"}
The approximate standard error formulas for $\hat{\sigma}_{i}^{2},\hat{\sigma}_{i}$, $\hat{\sigma_{ij}}$, and $\hat{\rho}_{ij}$ are given by: 

$$
\mathrm{se}(\hat{\sigma}_{i}^{2}) \approx\frac{\sqrt{2}\sigma_{i}^{2}}{\sqrt{T}}=\frac{\sigma_{i}^{2}}{\sqrt{T/2}},
$$ {#eq-sesigma2hat}
$$
\mathrm{se}(\hat{\sigma}_{i}) \approx\frac{\sigma_{i}}{\sqrt{2T}},
$$ {#eq-sesigmahat}
$$
\mathrm{se}(\hat{\sigma}_{ij}) \approx \frac{\sqrt{\sigma_i^2 \sigma_j^2 + \sigma_{ij}^2}}{\sqrt{T}} = \frac{\sqrt{\sigma_i^2 \sigma_j^2 (1 + \rho_{ij}^2)}}{\sqrt{T}} 
$$ {#eq-sesigmaijhat}
$$
\mathrm{se}(\hat{\rho}_{ij}) \approx\frac{1-\rho_{ij}^{2}}{\sqrt{T}},
$$ {#eq-serhohat}
where "$\approx$" denotes approximately equal. The
approximations are such that the approximation error goes to zero
as the sample size $T$ gets very large. 
:::

I make the following remarks: 

1.  As with the formula for the standard error of the sample mean, the formulas for $\mathrm{se}(\hat{\sigma}_{i}^{2})$ and $\mathrm{se}(\hat{\sigma}_{i})$ depend on $\sigma_{i}^{2}$. Larger values of $\sigma_{i}^{2}$ imply less precise estimates of $\hat{\sigma}_{i}^{2}$ and $\hat{\sigma}_{i}$. 
2.  The formula for $\mathrm{se}(\hat{\sigma}_{ij})$ depends on $\sigma_i^2$, $\sigma_j^2$, and $\rho_{ij}^2$. Given $\sigma_i^2$ and $\sigma_j^2$, the standard error is smallest when $\rho_{ij}=0$. 
3.  The formula for $\mathrm{se}(\hat{\rho}_{ij})$, does not depend on $\sigma_{i}^{2}$ but rather depends on $\rho_{ij}^{2}$ and is smaller the closer $\rho_{ij}^{2}$ is to unity. Intuitively, this makes sense because as $\rho_{ij}^{2}$ approaches one the linear dependence between $R_{it}$ and $R_{jt}$ becomes almost perfect and this will be easily recognizable in the data (scatterplot will almost follow a straight line). 
4.  The formulas for the standard errors above are inversely related to the square root of the sample size, $\sqrt{T}$, which means that larger sample sizes imply smaller values of the standard errors. 
5.  Interestingly, $\mathrm{se}(\hat{\sigma}_{i})$ goes to zero the fastest and $\mathrm{se}(\hat{\sigma}_{i}^{2})$ goes to zero the slowest. Hence, for a fixed sample size, these formulas suggest that $\sigma_{i}$ is generally estimated more precisely than $\sigma_{i}^{2}$ and $\rho_{ij}$, and $\rho_{ij}$ is estimated generally more precisely than $\sigma_{i}^{2}$.  


The above formulas @eq-sesigma2hat - @eq-serhohat
are not practically useful, however, because they depend on the unknown
quantities $\sigma_{i}^{2},\sigma_{i}$ and $\rho_{ij}$. Practically
useful formulas make use of the plug-in principle and replace $\sigma_{i}^{2},\sigma_{i}$ and $\rho_{ij}$
by the estimates $\hat{\sigma}_{i}^{2},\hat{\sigma}_{i}$ and $\hat{\rho}_{ij}$
and give rise to the *estimated standard errors*: 
$$
\widehat{\mathrm{se}}(\hat{\mu}_{i}) =\frac{\widehat{\sigma}_{i}}{\sqrt{T}}
$$ {#eq-estimated-semuhat}
$$
\widehat{\mathrm{se}}(\hat{\sigma}_{i}^{2}) \approx\frac{\hat{\sigma}_{i}^{2}}{\sqrt{T/2}},
$$ {#eq-estimated-ssesigma2hat}
$$
\widehat{\mathrm{se}}(\hat{\sigma}_{i}) \approx\frac{\hat{\sigma}_{i}}{\sqrt{2T}},
$$ {#eq-estimated-sesigmahat}
$$
\widehat{\mathrm{se}}(\hat{\sigma}_{ij}) \approx \frac{\sqrt{\hat{\sigma}_i^2 \hat{\sigma}_j^2 (1 + \hat{\rho}_{ij}^2)}}{\sqrt{T}} 
$$ {#eq-estimated-sesigmaijhat}
$$
\widehat{\mathrm{se}}(\hat{\rho}_{ij}) \approx\frac{1-\hat{\rho}_{ij}^{2}}{\sqrt{T}}.
$$ {#eq-estimated-serhohat}

It is good practice to report estimates together with their estimated
standard errors. In this way the precision of the estimates is transparent
to the user. Typically, estimates are reported in a table with the
estimates in one column and the estimated standard errors in an adjacent
column.

::: {#exm-widehat-mathrm-se-hat-mu-i-values-for-microsoft-starbucks-and-the-s-p-500-index .example name="$\\widehat{\\mathrm{se}}(\\hat{\\mu}_{i})$ values for Microsoft Starbucks and the S&P 500 index"}
For Microsoft, Starbucks and S\&P 500, the values of $\widehat{\mathrm{se}}(\hat{\mu}_{i})$
are easily computed in R using:

```{r}
#| echo: true

n.obs = nrow(gwnRetC)
seMuhat = sigmahat/sqrt(n.obs)
```

The values of $\hat{\mu}_{i}$ and $\widehat{\mathrm{se}}(\hat{\mu}_{i})$
shown together are:

```{r}
#| echo: true

cbind(muhat, seMuhat)
```

For Microsoft and Starbucks, the values of $\widehat{\mathrm{se}}(\hat{\mu}_{i})$
are similar because the values of $\hat{\sigma}_{i}$ are similar,
and $\widehat{\mathrm{se}}(\hat{\mu}_{i})$ is smallest for the S\&P
500 index. This occurs because $\hat{\sigma}_{\textrm{sp500}}$ is
much smaller than $\hat{\sigma}_{\textrm{msft}}$ and $\hat{\sigma}_{\textrm{sbux}}$.
Hence, $\hat{\mu}_{i}$ is estimated more precisely for the S\&P 500
index (a highly diversified portfolio) than it is for Microsoft and
Starbucks stock (individual assets). 

It is tempting to compare the magnitude of $\hat{\mu}_{i}$ to $\widehat{\mathrm{se}}(\hat{\mu}_{i})$ to evaluate if $\hat{\mu}_{i}$ is
a precise estimate. A common way to do this is to compute the so-called *t-ratio*:

```{r}
#| echo: true

muhat/seMuhat
```
The t-ratio shows the number of $\widehat{\mathrm{se}}(\hat{\mu}_{i})$ values $\hat{\mu}_{i}$ is from zero. For example, $\hat{\mu}_{msft} = 0.004$ is 0.54 values of
$\widehat{\mathrm{se}}(\hat{\mu}_{msft}) = 0.008$ above zero. This is not very far from zero. In contrast, $\hat{\mu}_{sbux} = 0.015$ is 1.72 values of
$\widehat{\mathrm{se}}(\hat{\mu}_{sbux}) = 0.008$ above zero. This is moderately far
from zero. Because, $\hat{\mu_{i}}$ represents an average rate of return on an asset it informative to know how far above zero it is likely to be. The farther $\hat{\mu_{i}}$ 
is from zero, in units of $\widehat{\mathrm{se}}(\hat{\mu}_{i})$, the more sure you are 
that the asset has a positive average rate of return. 
\hfill{}$\blacksquare$
:::

::: {#exm-computing-estimated-standard-errors .example name="Computing $\\widehat{\\mathrm{se}}(\\hat{\\sigma}_{i}^{2})$, $\\widehat{\\mathrm{se}}(\\hat{\\sigma}_{i})$, $\\widehat{\\mathrm{se}}(\\hat{\\sigma}_{ij})$, and $\\widehat{\\mathrm{se}}(\\hat{\\rho}_{ij})$ for Microsoft Starbucks and the S&P 500"}

For Microsoft, Starbucks and S\&P 500, the values of $\widehat{\mathrm{se}}(\hat{\sigma}_{i}^{2})$ and
$\widehat{\mathrm{se}}(\hat{\sigma}_{i})$ (together with the estimates $\hat{\sigma}_{i}^{2}$ and $\hat{\sigma}_{i}$ are:

```{r}
#| echo: true

seSigma2hat = sigma2hat/sqrt(n.obs/2)
seSigmahat = sigmahat/sqrt(2*n.obs)
cbind(sigma2hat, seSigma2hat, sigmahat, seSigmahat)
```

Notice that $\sigma^{2}$ and $\sigma$ for the S\&P 500 index are
estimated much more precisely than the values for Microsoft and Starbucks.
Also notice that $\sigma_{i}$ is estimated more precisely than $\mu_{i}$
for all assets: the values of $\widehat{\mathrm{se}}$($\hat{\sigma}_{i})$
relative to $\hat{\sigma}_{i}$ are much smaller than the values of
$\widehat{\mathrm{se}}$($\hat{\mu}_{i})$ to $\hat{\mu}_{i}$. 

The values of $\widehat{\mathrm{se}}(\hat{\sigma}_{ij})$ and $\widehat{\mathrm{se}}(\hat{\rho}_{ij})$ (together
with the values of $\sigma_{ij}$ and $\hat{\rho}_{ij})$ are:

```{r}
#| echo: true

seCovhat = sqrt(sigma2hat[c("MSFT","MSFT","SBUX")]*sigma2hat[c("SBUX","SP500","SP500")]*(1 + rhohat^2))/sqrt(n.obs)
seRhohat = (1-rhohat^2)/sqrt(n.obs)
cbind(covhat, seCovhat, rhohat, seRhohat)
```

The values of $\widehat{\mathrm{se}}(\hat{\sigma}_{ij})$ and $\widehat{\mathrm{se}}(\hat{\rho}_{ij})$ are moderate
in size (relative to $\hat{\sigma}_{ij}$ and $\hat{\rho}_{ij})$. Notice that $\hat{\rho}_{\textrm{msft,sp500}}$
has the smallest estimated standard error because $\hat{\rho}_{\textrm{msft,sp500}}^{2}$
is closest to one.
\hfill{}$\blacksquare$
:::

### Sampling Distributions and Confidence Intervals

#### Sampling Distribution for $\hat{\mu}_{i}$
\index{Sample distribution}
In the GWN model, $R_{it}\sim iid~N(\mu_{i},\sigma_{i}^{2})$ and
since $\hat{\mu}_{i}=\frac{1}{T}\sum_{t=1}^{T}R_{it}$ is an average
of these normal random variables, it is also normally distributed.
Previously I showed that $E[\hat{\mu}_{i}]=\mu_{i}$ and $\mathrm{var}(\hat{\mu}_{i}) =\frac{\sigma_{i}^{2}}{T}$.
Therefore, the exact probability distribution of $\hat{\mu}_{i}$,
$f(\hat{\mu}_{i})$, for a fixed sample size $T$ is the normal distribution
$\hat{\mu}_{i}\sim~N\left(\mu_{i},\frac{\sigma_{i}^{2}}{T}\right).$ Therefore, 
you have an exact formula for $f(\hat{\mu}_{i})$:

$$
f(\hat{\mu}_{i})=\left(\frac{2\pi\sigma_{i}^{2}}{T}\right)^{-1/2}\exp\left\{ -\frac{1}{2\sigma_{i}^{2}/T}(\hat{\mu}_{i}-\mu_{i})^{2}\right\} .
$$ {#eq-muhatdist}

The probability curve $f(\hat{\mu}_{i})$ is centered at the true
value $\mu_{i}$, and the spread about $\mu_{i}$ depends on the magnitude
of $\sigma_{i}^{2}$, the variability of $R_{it}$, and the sample
size, $T$. For a fixed sample size $T$, the uncertainty in $\hat{\mu}_{i}$
is larger (smaller) for larger (smaller) values of $\sigma_{i}^{2}$. Notice that the
variance of $\hat{\mu}_{i}$ is inversely related to the sample size
$T$. Given $\sigma_{i}^{2}$, $\mathrm{var}(\hat{\mu}_{i})$ is smaller
for larger sample sizes than for smaller sample sizes. This makes
sense since you expect to have a more precise estimator when you have
more data. If the sample size is very large (as $T\rightarrow\infty$)
then $\mathrm{var}(\hat{\mu}_{i})$ will be approximately zero and
the normal distribution of $\hat{\mu}_{i}$ given by @eq-muhatdist
will be essentially a spike at $\mu_{i}$. In other words, if the
sample size is very large then you essentially know the true value
of $\mu_{i}$. Hence, you have established that $\hat{\mu}_{i}$ is
a *consistent* estimator of $\mu_{i}$ as the sample size goes
to infinity.

The exact sampling distribution for $\hat{\mu}$ assumes that $\sigma^2$ is known, which is not practically useful. Below, I provide a practically useful result which holds when you use $\hat{\sigma}^2$.

::: {#exm-sampling-distribution-with-different-sample-sizes .example name="Sampling distribution of $\\hat{\\mu}$ with different sample sizes"}
The distribution of $\hat{\mu}_{i}$, with $\mu_{i}=0$ and $\sigma_{i}^{2}=1$
for various sample sizes is illustrated in @fig-CerModelEstimationMuhat.
Notice how fast the distribution collapses at $\mu_{i}=0$ as $T$
increases.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-CerModelEstimationMuhat
#| echo: false
#| fig-cap: $N(0,1/T)$ sampling distributions for $\hat{\mu}$ for $T=1,10$ and $50$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.vals = seq(3, -3, length=100)
pdf1 = dnorm(x.vals)
pdf2 = dnorm(x.vals, sd=1/sqrt(10))
pdf3 = dnorm(x.vals, sd=1/sqrt(50))
ub = max(pdf1, pdf2, pdf3)

# Figure
plot(x.vals, pdf1, ylab="pdf", xlab=expression(hat(mu)), ylim=c(0, ub), type="l", lwd=2)
lines(x.vals, pdf2, lwd=2, col=2, lty=2)
lines(x.vals, pdf3, lwd=2, col=3, lty=3)
legend(x="topleft", legend=c("pdf: T=1", "pdf: T=10", "pdf: T=50"),
       lty=1:3, col=1:3, lwd=2)
```


#### Confidence Intervals for $\mu_{i}$
\index{Confidence intervals}

In practice, the precision of $\hat{\mu}_{i}$ is measured by $\widehat{\mathrm{se}}(\hat{\mu}_{i})$
but is best communicated by computing a *confidence interval*
for the unknown value of $\mu_{i}$. A confidence interval is an *interval estimate* of $\mu_{i}$ such that you can put an explicit probability
statement about the likelihood that the interval covers $\mu_{i}$.

Because you know the exact finite sample distribution of $\hat{\mu}_{i}$ in the GWN
return model, you can construct an exact confidence interval for $\mu_{i}$.


::: {#prp-t-ratio-for-sample-mean .proposition name="t-ratio for sample mean"}
Let $\{R_{it}\}_{t=1}^{T}$ be generated from the
GWN model @eq-GWN-model-matrix. Define the $t$-ratio as:
$$
t_{i}=\frac{\hat{\mu}_{i}-\mu_{i}}{\widehat{\mathrm{se}}(\hat{\mu}_{i})}=\frac{\hat{\mu}_{i}-\mu_{i}}{\hat{\sigma}/\sqrt{T}},
$$ {#eq-t-ratio}
Then $t_{i}\sim t_{T-1}$ where $t_{T-1}$ denotes a Student's $t$
random variable with $T-1$ degrees of freedom. 
:::

The Student's $t$ distribution with $v>0$ degrees of freedom is
a symmetric distribution centered at zero, like the standard normal.
The tail-thickness (kurtosis) of the distribution is determined by
the degrees of freedom parameter $v$. For values of $v$ close to
zero, the tails of the Student's $t$ distribution are much fatter
than the tails of the standard normal distribution. As $v$ gets large,
the Student's $t$ distribution approaches the standard normal distribution.


For $\alpha\in(0,1)$, you compute a $(1-\alpha)\cdot100\%$ confidence
interval for $\mu_{i}$ using @eq-t-ratio and the $1-\alpha/2$
quantile (critical value) $t_{T-1}(1-\alpha/2)$ to give: 
$$
\Pr\left(-t_{T-1}(1-\alpha/2)\leq\frac{\hat{\mu}_{i}-\mu_{i}}{\widehat{\mathrm{se}}(\hat{\mu}_{i})}\leq t_{T-1}(1-\alpha/2)\right)=1-\alpha,
$$
which can be rearranged as, 
$$
\Pr\left(\hat{\mu}_{i}-t_{T-1}(1-\alpha/2)\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i})\leq\mu_{i}\leq\hat{\mu}_{i}+t_{T-1}(1-\alpha/2)\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i})\right)=1-\alpha.
$$
Hence, the interval, 
$$
 [\hat{\mu}_{i}-t_{T-1}(1-\alpha/2)\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i}),~\hat{\mu}_{i}+t_{T-1}(1-\alpha/2)\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i})]
 = \hat{\mu}_{i}\pm t_{T-1}(1-\alpha/2)\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i})
$$ {#eq-exact-ci}
covers the true unknown value of $\mu_{i}$ with probability $1-\alpha$.

::: {#exm-computing-95-confidence-intervals-for-mu .example name="Computing 95% confidence intervals for $\\mu_{i}$"}

Suppose you want to compute a 95\% confidence interval for $\mu_{i}$.
In this case $\alpha=0.05$ and $1-\alpha=0.95$. Suppose further
that $T-1=60$ (e.g., five years of monthly return data) so that $t_{T-1}(1-\alpha/2)=t_{60}(0.975)=2$.
This can be verified in R using the function `qt()`.

Then the 95\% confidence for $\mu_{i}$ is given by: 

$$
\hat{\mu}_{i}\pm2\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i}).
$$ {#eq-approx-95-ci}


The above formula for a 95\% confidence interval is often used as
a rule of thumb for computing an approximate 95\% confidence interval
for moderate sample sizes. It is easy to remember and does not require
the computation of the quantile $t_{T-1}(1-\alpha/2)$ from the Student's
$t$ distribution. It is also an approximate 95\% confidence interval
that is based the asymptotic normality of $\hat{\mu}_{i}$. Recall,
for a normal distribution with mean $\mu$ and variance $\sigma^{2}$
approximately 95\% of the probability lies between $\mu\pm2\sigma$.
\hfill{}$\blacksquare$
:::

The coverage probability associated with the confidence interval for
$\mu_{i}$ is based on the fact that the estimator $\hat{\mu}_{i}$
is a random variable. Since the confidence interval is constructed as
$\hat{\mu}_{i}\pm t_{T-1}(1-\alpha/2)\cdot\widehat{\mathrm{se}}(\hat{\mu}_{i})$
it is also a random variable. An intuitive way to think about the
coverage probability associated with the confidence interval is to
think about the game of "horseshoes"^[Horse shoes is a game commonly played at county fairs. See (http://en.wikipedia.org/wiki/Horseshoes)
for a complete description of the game.]. The horseshoe is the confidence interval and the parameter $\mu_{i}$
is the post at which the horse shoe is tossed. Think of playing the game
100 times. If the thrower
is 95\% accurate (if the coverage probability is 0.95) then 95 of
the 100 tosses should ring the post (95 of the constructed confidence
intervals should contain the true value $\mu_{i})$.

::: {#exm-95-confidence-intervals-for-mu .example name="95% confidence intervals for $\\mu_{i}$ for Microsoft Starbucks and the S&P 500 index"}

Consider computing 95\% confidence intervals for $\mu_{i}$ using
@eq-exact-ci based on the estimated results for the Microsoft,
Starbucks and S\&P 500 data. The degrees of freedom for the Student's
$t$ distribution is $T-1=171$. The 97.5\% quantile, $t_{99}(0.975)$,
can be computed using the R function `qt()`:

```{r}
#| echo: true

t.975 = qt(0.975, df=(n.obs-1))
t.975
```

Notice that this quantile is very close to $2.$ Then the exact 95\%
confidence intervals are given by:

```{r}
#| echo: true

lower = muhat - t.975*seMuhat
upper = muhat + t.975*seMuhat
width = upper - lower
ans= cbind(lower, upper, width)
colnames(ans) = c("2.5%", "97.5%", "Width")
ans
```

With probability 0.95, the above intervals will contain the true mean
values assuming the GWN model is valid. The 95\% confidence intervals
for Microsoft and Starbucks are fairly wide (about 3\%) and contain
both negative and positive values. The confidence interval for the
S\&P 500 index is tighter but also contains negative and positive
values. For Microsoft, the confidence interval is $[-1.1\%~1.9\%]$.
This means that with probability 0.95, the true monthly expected return
is somewhere between -1.1\% and 1.9\%. The economic implications of
a -1.1\% expected monthly return and a 1.9\% expected return are vastly
different. In contrast, the 95\% confidence interval for the SP500
is about half the width of the intervals for Microsoft or Starbucks.
The lower limit is near -0.5\% and the upper limit is near 1\%. This
result clearly shows that the monthly mean return for the S\&P 500
index is estimated much more precisely than the monthly mean returns
for Microsoft or Starbucks.
\hfill{}$\blacksquare$
:::

#### Sampling Distributions for $\hat{\sigma}_{i}^{2}$, $\hat{\sigma}_{i}$, $\hat{\sigma}_{ij}$, and $\hat{\rho}_{ij}$
\index{Sample distribution}
The exact distributions of $\hat{\sigma}_{i}^{2}$, $\hat{\sigma}_{i}$, $\hat{\sigma}_{ij}$,
and $\hat{\rho}_{ij}$ based on a fixed sample size $T$ are not normal
distributions and are difficult to derive^[For example, the exact sampling distribution of $(T-1)\hat{\sigma}_{i}^{2}/\sigma_{i}^{2}$
is chi-square with $T-1$ degrees of freedom.]. However, approximate normal distributions of the form @eq-asymptotic-normal-distribution
based on the CLT are readily available: 

$$
 \hat{\sigma}_{i}^{2}\sim N\left(\sigma_{i}^{2},\widehat{\mathrm{se}}(\hat{\sigma}_{i}^{2})^{2}\right)=N\left(\sigma_{i}^{2},\frac{4\hat{\sigma}_{i}^{4}}{T}\right),
$$ {#eq-pdf-sigma2}
$$
 \hat{\sigma}_{i}\sim N\left(\sigma_{i},\widehat{\mathrm{se}}(\hat{\sigma}_{i})^{2}\right)=N\left(\sigma_{i},\frac{\hat{\sigma}_{i}^{2}}{2T}\right),
$$ {#eq-pdf-sigma}
 $$
 \hat{\sigma}_{ij} \sim  N\left(\sigma_{ij},\widehat{\mathrm{se}}(\hat{\sigma}_{ij})^{2}\right)= N\left(\sigma_{ij}, \frac{\hat{\sigma}_i^2 \hat{\sigma}_j^2(1 + \hat{\rho}_{ij}^2)}{T}  \right) 
 $$ {#eq-pdf-sigmaij}
 $$
 \hat{\rho}_{ij}\sim N\left(\rho_{ij},\widehat{\mathrm{se}}(\hat{\rho}_{ij})^{2}\right)=N\left(\rho_{ij},\frac{(1-\hat{\rho}_{ij}^{2})^{2}}{T}\right).
 $$ {#eq-pdf-rho}

These approximate normal distributions can be used to compute approximate
confidence intervals for $\sigma_{i}^{2}$, $\sigma_{i}$, $\sigma_{ij}$ and $\rho_{ij}$.

#### Approximate Confidence Intervals for $\sigma_{i}^{2},\sigma_{i}$, $\sigma_{ij}$, and $\rho_{ij}$

Approximate 95\% confidence intervals for $\sigma_{i}^{2},\sigma_{i}$
and $\rho_{ij}$ are given by: 
$$
\hat{\sigma}_{i}^{2}\pm2\cdot\widehat{\mathrm{se}}(\hat{\sigma}_{i}^{2})  =\hat{\sigma}_{i}^{2}\pm2\cdot\frac{\hat{\sigma}_{i}^{2}}{\sqrt{T/2}},
$$ {#eq-ci-sigma2}
$$
\hat{\sigma}_{i}\pm2\cdot\widehat{\mathrm{se}}(\hat{\sigma}_{i}) =\hat{\sigma}_{i}\pm2\cdot\frac{\hat{\sigma}_{i}}{\sqrt{2T}},
$$ {#eq-ci-sigma}
$$
\hat{\sigma}_{ij}\pm2\cdot\widehat{\mathrm{se}}(\hat{\sigma}_{ij}) = \hat{\sigma}_{ij}\pm2\cdot\frac{\sqrt{\hat{\sigma}_i^2 \hat{\sigma}_j^2(1+\hat{\rho}_{ij}^2)}}{\sqrt{T}}.
$$ {#eq-ci-sigmaij}
$$ 
\hat{\rho}_{ij}\pm2\cdot\widehat{\mathrm{se}}(\hat{\rho}_{ij}) =\hat{\rho}_{ij}\pm2\cdot\frac{(1-\hat{\rho}_{ij}^{2})}{\sqrt{T}}.
$$ {#eq-ci-rho}

::: {#exm-approximate-95-confidence-intervals-for-vol-parameters .example name="Approximate 95% confidence intervals for $\\sigma_{i}^{2}$, $\\sigma_{i}$, $\\sigma_{ij}$ and $\\rho_{ij}$ for Microsoft Starbucks and the S&P 500"}

Using @eq-ci-sigma2 - @eq-ci-sigma, the approximate
95\% confidence intervals for $\sigma_{i}^{2}$ and $\sigma_{i}$
($i=$ Microsoft, Starbucks, S\&P 500) are:

```{r}
#| echo: true

# 95% confidence interval for variance
lowerSigma2 = sigma2hat - 2*seSigma2hat
upperSigma2 = sigma2hat + 2*seSigma2hat
widthSigma2 = upperSigma2 - lowerSigma2
ans = cbind(lowerSigma2, upperSigma2, widthSigma2)
colnames(ans) = c("2.5%", "97.5%", "Width")
ans

```

```{r}
#| echo: true
# 95% confidence interval for volatility
lowerSigma = sigmahat - 2*seSigmahat
upperSigma = sigmahat + 2*seSigmahat
widthSigma = upperSigma - lowerSigma
ans = cbind(lowerSigma, upperSigma, widthSigma)
colnames(ans) = c("2.5%", "97.5%", "Width")
ans
```
The 95\% confidence intervals for $\sigma$ and $\sigma^{2}$ are
larger for Microsoft and Starbucks than for the S\&P 500 index. For
all assets, the intervals for $\sigma$ are fairly narrow (2\% for
Microsoft and Starbucks and 1\% for S\&P 500 index) indicating that
$\sigma$ is precisely estimated. 

The approximate 95\% confidence intervals for $\sigma_{ij}$ are:

```{r}
#| echo: true
lowerCov = covhat - 2*seCovhat
upperCov = covhat + 2*seCovhat
widthCov = upperCov - lowerCov
ans = cbind(lowerCov, upperCov, widthCov)
colnames(ans) = c("2.5%", "97.5%", "Width")
ans
```


The approximate 95\% confidence intervals for $\rho_{ij}$ are:

```{r}
#| echo: true

lowerRho = rhohat - 2*seRhohat
upperRho = rhohat + 2*seRhohat
widthRho = upperRho - lowerRho
ans = cbind(lowerRho, upperRho, widthRho)
colnames(ans) = c("2.5%", "97.5%", "Width")
ans
```

The 95\% confidence intervals for $\sigma_{ij}$ $\rho_{ij}$ are not too wide and all contain just positive values away from zero. The smallest interval is for $\rho_{\textrm{msft,sp500}}$ because $\hat{\rho}_{\textrm{msft,sp500}}$ is closest to 1.
\hfill{}$\blacksquare$
:::

## Using Monte Carlo Simulation to Understand the Statistical Properties of Estimators {#sec-Using-Monte-Carlo-Simulation}
\index{Monte Carlo Simulation}
Let $R_{t}$ be the return on a single asset described by the GWN
model, let $\theta$ denote some characteristic (parameter) of the
GWN model you are interested in estimating, and let $\hat{\theta}$
denote an estimator for $\theta$ based on a sample of size $T$.
The exact meaning of estimator bias, $\mathrm{bias}(\hat{\theta},\theta)$,
the interpretation of $\mathrm{se}(\hat{\theta})$ as a measure of
precision, the sampling distribution $f(\hat{\theta})$, and the interpretation
of the coverage probability of a confidence interval for $\theta$,
can all be a bit hard to grasp at first. If $\mathrm{bias}(\hat{\theta},\theta)=0$
so that $E[\hat{\theta}]=\theta$ then over an infinite number of
repeated samples of $\{R_{it}\}_{t=1}^{T}$ the average of the $\hat{\theta}$
values computed over the infinite samples is equal to the true value
$\theta$. The value of $\mathrm{se}(\hat{\theta})$ represents the
standard deviation of these $\hat{\theta}$ values. The sampling distribution
$f(\hat{\theta})$ is the smoothed histogram of these $\hat{\theta}$
values. And the 95\% confidence intervals for $\theta$ will actually
contain $\theta$ in 95\% of the samples. You can think of these hypothetical
samples as different Monte Carlo simulations of the GWN model. In
this way you can approximate the computations involved in evaluating
$E[\hat{\theta}]$, $\mathrm{se}(\hat{\theta})$, $f(\hat{\theta})$,
and the coverage probability of a confidence interval using a large,
but finite, number of Monte Carlo simulations.

### Evaluating the Statistical Properties of $\hat{\mu}$ Using Monte Carlo Simulation

Consider the GWN model: 
$$
\begin{aligned}
R_{t} & =0.05+\varepsilon_{t},t=1,\ldots,100\\
 & \varepsilon_{t}\sim\mathrm{GWN}(0,(0.10)^{2}). 
\end{aligned}
$$ {#eq-GWN-simulation-model}
Here, the true parameter values are $\mu=0.05$ and $\sigma=0.10$.
Using Monte Carlo simulation, you can simulate $N=1000$ different
samples of size $T=100$ from @eq-GWN-simulation-model giving
the sample realizations $\{r_{t}^{j}\}_{t=1}^{100}$ for $j=1,\ldots,1000$. The R code to perform the Monte Carlo simulation is:

```{r}
#| echo: true

mu = 0.05
sigma = 0.10
n.obs = 100
n.sim = 1000
set.seed(111)
sim.means = rep(0,n.sim)
mu.lower = rep(0,n.sim)
mu.upper = rep(0,n.sim)
qt.975 = qt(0.975, n.obs-1)
for (sim in 1:n.sim) {
	sim.ret = rnorm(n.obs,mean=mu,sd=sigma)
	sim.means[sim] = mean(sim.ret)
	se.muhat = sd(sim.ret)/sqrt(n.obs)
	mu.lower[sim] = sim.means[sim]-qt.975*se.muhat
	mu.upper[sim] = sim.means[sim]+qt.975*se.muhat
}
```

The first 10 of these simulated samples are illustrated in @fig-CerModelEstimationCerMcSim. Notice that there is considerable variation in the appearance of the simulated samples, but that all of the simulated samples fluctuate about the true mean value of $\mu=0.05$ and have a typical deviation from the mean of about $0.10$. 

```{r}
#| label: fig-CerModelEstimationCerMcSim
#| echo: false
#| fig-cap: Ten simulated samples of size $T=100$ from the GWN model $R_{t}=0.05+\varepsilon_{t}$,
#|   $\varepsilon_{t}\sim iid~N(0,(0.10)^{2})$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

my.panel <- function(...) {
  lines(...)
  abline(h=0)
}
sim.ret.10 = matrix(0, n.obs, 10)
colnames(sim.ret.10) = paste("Sim", 1:10, sep=".")
# 10 simulation for figure figure
set.seed(111)
for (sim in 1:10) {
	sim.ret.10[, sim] = rnorm(n.obs,mean=mu,sd=sigma)
}
plot(as.zoo(sim.ret.10), col="blue", lwd=2, main="", panel=my.panel)
```

For each of the $1000$ simulated samples you can estimate $\hat{\mu}$ giving $1000$ mean estimates $\{\hat{\mu}^{1},\ldots,\hat{\mu}^{1000}\}$. A histogram of these $1000$ mean values is illustrated in  @fig-CerModelEstimationMcMuhat. 

```{r}
#| label: fig-CerModelEstimationMcMuhat
#| echo: false
#| results: hide
#| fig-cap: Distribution of $\hat{\mu}$ computed from $1000$ Monte Carlo simulations
#|   from the GWN model. White dashed line is the average of the $\mu$ values, and orange
#|   curve is the true $f(\hat{\mu})$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# plot histogram and overlay true sampling distribution
hist(sim.means, col="cornflowerblue", ylim=c(0,40), 
     main="", xlab="muhat", probability=T)
abline(v=mean(sim.means), col="white", lwd=4, lty=2)
# overlay normal curve
x.vals = seq(0.02, 0.08, length=100)
lines(x.vals, dnorm(x.vals, mean=mu, sd=sigma/sqrt(100)), col="orange", lwd=2)
```

The histogram of the estimated means, $\hat{\mu}^{j}$, can be thought of as an estimate of the underlying pdf, $f(\hat{\mu})$,
of the estimator $\hat{\mu}$ which you know from @eq-muhatdist
is a normal pdf centered at $E[\hat{\mu}]=\mu=0.05$ with $\mathrm{se}(\hat{\mu}_{i})=0.10/\sqrt{100}=0.01$. This normal curve (solid orange line) is superimposed on the histogram in @fig-CerModelEstimationMcMuhat. Notice that the
center of the histogram (white dashed vertical line) is very close
to the true mean value $\mu=0.05$. That is, on average over the $1000$
Monte Carlo samples the value of $\hat{\mu}$ is about 0.05. In some
samples, the estimate is too big and in some samples the estimate
is too small but on average the estimate is correct. In fact, the
average value of $\{\hat{\mu}^{1},\ldots,\hat{\mu}^{1000}\}$ from
the $1000$ simulated samples is: 
$$
\overline{\hat{\mu}}=\frac{1}{1000}\sum_{j=1}^{1000}\hat{\mu}^{j}=0.0497,
$$
which is very close to the true value $0.05$. If the number of simulated
samples is allowed to go to infinity then the sample average $\overline{\hat{\mu}}$
will be exactly equal to $\mu=0.05$: 
$$
\lim_{N\rightarrow\infty}\frac{1}{N}\sum_{j=1}^{N}\hat{\mu}^{j}=E[\hat{\mu}]=\mu=0.05.
$$

The typical size of the spread about the center of the histogram represents
$\mathrm{se}(\hat{\mu}_{i})$ and gives an indication of the precision
of $\hat{\mu}_{i}$. The value of $\mathrm{se}(\hat{\mu}_{i})$ may
be approximated by computing the sample standard deviation of the
$1000$ $\hat{\mu}^{j}$ values: 
$$
\hat{\sigma}_{\hat{\mu}}=\sqrt{\frac{1}{999}\sum_{j=1}^{1000}(\hat{\mu}^{j}-0.04969)^{2}}=0.0104.
$$
Notice that this value is very close to the exact standard error:  $\mathrm{se}(\hat{\mu}_{i})=0.10/\sqrt{100}=0.01$. If the number of simulated sample goes to infinity then: 
$$
\lim_{N\rightarrow\infty}\sqrt{\frac{1}{N-1}\sum_{j=1}^{N}(\hat{\mu}^{j}-\overline{\hat{\mu}})^{2}}=\mathrm{se}(\hat{\mu}_{i})=0.01.
$$

The coverage probability of the 95\% confidence interval for $\mu$
can also be illustrated using Monte Carlo simulation. For each simulation
$j$, the interval $\hat{\mu}^j \pm t_{100}(0.975)\times\widehat{\mathrm{se}}(\hat{\mu}^{j})$
is computed. The coverage probability is approximated by the fraction
of intervals that contain (cover) the true $\mu=0.05$. For the 1000
simulated samples, this fraction turns out to be 0.931. As the number
of simulations goes to infinity, the Monte Carlo coverage probability
will be equal to 0.95.

Here are the R results to support the above calculations. The mean and standard deviation of $\{\hat{\mu}^{1},\ldots,\hat{\mu}^{1000}\}$ are:

```{r}
#| echo: true

ans = c(mean(sim.means), sd(sim.means))
names(ans) = c("Mu", "Sigma")
ans
```
To evaluate the coverage probability of the 95\% confidence intervals,
you count the number of times each interval actually contains the true
value of $\mu$:

```{r}
#| echo: true

in.interval = mu >= mu.lower & mu <= mu.upper
sum(in.interval)/n.sim
```
### Evaluating the Statistical Properties of $\hat{\sigma}_{i}^{2}$ and $\hat{\sigma}_{i}$ Using Monte Carlo simulation

You can evaluate the statistical properties of $\hat{\sigma}_{i}^{2}$
and $\hat{\sigma}_{i}$ by Monte Carlo simulation in the same way
that you evaluated the statistical properties of $\hat{\mu}_{i}$.
You use the simulation model @eq-GWN-simulation-model and $N=1000$
simulated samples of size $T=100$ to compute the estimates $\{\left(\hat{\sigma}^{2}\right)^{1},\ldots,\left(\hat{\sigma}^{2}\right)^{1000}\}$
and $\{\hat{\sigma}^{1},\ldots,\hat{\sigma}^{1000}\}$. The R code
for the simulation loop is:

```{r}
#| echo: true

mu = 0.05
sigma = 0.10
n.obs = 100
n.sim = 1000
set.seed(111)
sim.sigma2 = rep(0,n.sim)
sigma2.lower = rep(0,n.sim)
sigma2.upper = rep(0,n.sim)
sim.sigma = rep(0,n.sim)
sigma.lower = rep(0,n.sim)
sigma.upper = rep(0,n.sim)
for (sim in 1:n.sim) {
	sim.ret = rnorm(n.obs,mean=mu,sd=sigma)
	sim.sigma2[sim] = var(sim.ret)
	sim.sigma[sim] = sqrt(sim.sigma2[sim])
    sigma2.lower[sim] = sim.sigma2[sim] - 2*sim.sigma2[sim]/sqrt(n.obs/2)
	sigma2.upper[sim] = sim.sigma2[sim] + 2*sim.sigma2[sim]/sqrt(n.obs/2)
	sigma.lower[sim] = sim.sigma[sim] - 2*sim.sigma[sim]/sqrt(n.obs*2)
	sigma.upper[sim] = sim.sigma[sim] + 2*sim.sigma[sim]/sqrt(n.obs*2)
}

```

The histograms of these values, with the asymptotic normal distributions
overlayed, are displayed in @fig-CerModelEstimationMcSigma.
The histogram for the $\hat{\sigma}^{2}$ values is bell-shaped and
slightly right skewed but is centered very close to $\sigma^{2}=0.010$.
The histogram for the $\hat{\sigma}$ values is more symmetric and
is centered near $\sigma=0.10$. 

```{r}
#| label: fig-CerModelEstimationMcSigma
#| echo: false
#| results: hide
#| fig-cap: Histograms of $\hat{\sigma}^{2}$ and $\hat{\sigma}$ computed from $N=1000$
#|   Monte Carlo samples from GWN model.
#| out-width: 100%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(1,2))
	hist(sim.sigma2, col="cornflowerblue", xlab="sigma2 hat values", 
       main="sigma2 hat", probability=T)
	abline(v=mean(sim.sigma2), col="white", lwd=4, lty=2)
  # overlay normal distribution
  x.vals = seq(0.005, 0.016, length=100)
  lines(x.vals, dnorm(x.vals, mean=sigma^2, sd=sigma^2/sqrt(100/2)), col="orange", lwd=2)
	hist(sim.sigma, col="cornflowerblue", xlab="sigma hat values", 
       main="sigma hat", probability=T)
	abline(v=mean(sim.sigma), col="white", lwd=4, lty=2)
  # overlay normal distribution
  x.vals = seq(0.07, 0.13, length=100)
  lines(x.vals, dnorm(x.vals, mean=sigma, sd=sigma/sqrt(2*100)), col="orange", lwd=2)
par(mfrow=c(1,1))
```

The average values of $\hat{\sigma}^{2}$
and $\hat{\sigma}$ from the $1000$ simulations are: 
\begin{align*}
\overline{\hat{\sigma}^{2}} & =\frac{1}{1000}\sum_{j=1}^{1000}\left(\hat{\sigma}^{2}\right)^{j}=0.00999,\\
\overline{\hat{\sigma}} & =\frac{1}{1000}\sum_{j=1}^{1000}\hat{\sigma}^{j}=0.0997.
\end{align*}
The Monte Carlo estimate of the bias for $\hat{\sigma}^{2}$ is $0.00999-0.01=-0.0000$, and the estimate of bias for $\hat{\sigma}$ is $0.0997-0.010=-0.0003$. This confirms that $\hat{\sigma}^{2}$ is unbiased and that the bias in $\hat{\sigma}$ is extremely small. If the number of simulated
samples, $N$, goes to infinity then $\overline{\hat{\sigma}^{2}}\rightarrow E[\hat{\sigma}^{2}]=\sigma^{2}=0.01$, and $\overline{\hat{\sigma}}\rightarrow E[\hat{\sigma}]=\sigma+\mathrm{bias}(\hat{\sigma},\sigma)$.

The sample standard deviation values of the Monte Carlo estimates
of $\sigma^{2}$ and $\sigma$ give approximations to $\mathrm{se}(\hat{\sigma}^{2})$
and $\mathrm{se}(\hat{\sigma})$: 
\begin{align*}
\hat{\sigma}_{\hat{\sigma}^{2}} & =\sqrt{\frac{1}{999}\sum_{j=1}^{1000}((\hat{\sigma}^{2})^{j}-0.00999)^{2}}=0.00135,\\
\hat{\sigma}_{\hat{\sigma}} & =\sqrt{\frac{1}{999}\sum_{j=1}^{1000}(\hat{\sigma}^{j}-0.0997)^{2}}=0.00676.
\end{align*}
The approximate values for $\mathrm{se}(\hat{\sigma}^{2})$ and $\mathrm{se}(\hat{\sigma})$
based on the CLT are: 
\begin{align*}
\mathrm{se}(\hat{\sigma}^{2}) & =\frac{(0.10)^{2}}{\sqrt{100/2}}=0.00141,\\
\mathrm{se}(\hat{\sigma}^{2}) & =\frac{0.10}{\sqrt{2\times100}}=0.00707.
\end{align*}
Notice that the Monte Carlo estimates of $\mathrm{se}(\hat{\sigma}^{2})$
and $\mathrm{se}(\hat{\sigma})$ are a bit different from the CLT
based estimates. The reason is that the CLT based estimates are approximations
that hold when the sample size $T$ is large. Because $T=100$ is
not too large, the Monte Carlo estimates of s$\mathrm{e}(\hat{\sigma}^{2})$
and $\mathrm{se}(\hat{\sigma})$ are likely more accurate (and will
be more accurate if the number of simulations is larger). 

For each simulation $j$, the approximate 95\% confidence intervals
$(\hat{\sigma}^{2})^{j}\pm2\times\widehat{\mathrm{se}}((\hat{\sigma}^{2})^{j})$
and $\hat{\sigma}^{j}\pm2\times\widehat{\mathrm{se}}(\hat{\sigma}^{j})$
are computed. The coverage probabilities of these intervals is approximated by the
fractions of intervals that contain (cover) the true values $\sigma^{2}=0.01$
and $\sigma=0.10$ respectively. For the 1000 simulated samples, these
fractions turn out to be 0.951 and 0.963, respectively. As the number
of simulations and the sample size go to infinity, the Monte Carlo
coverage probability will be equal to 0.95.

Here is the R code to support the above calculation. The mean values and biases of the estimates are:

```{r}
#| echo: true
ans = rbind(c(sigma^2,mean(sim.sigma2),mean(sim.sigma2) - sigma^2),
            c(sigma, mean(sim.sigma), mean(sim.sigma) - sigma))
colnames(ans) = c("True Value", "Mean", "Bias")
rownames(ans) = c("Variance", "Volatility")
round(ans, digits=5)
```
The Monte Carlo and analytic standard error estimates are:

```{r}
#| echo: true
ans = rbind(c(sd(sim.sigma2),sigma^2/sqrt(n.obs/2)),
            c(sd(sim.sigma), sigma/sqrt(2*n.obs)))
colnames(ans) = c("MC Std Error", "Analytic Std Error")
rownames(ans) = c("Variance", "Volatility")
ans
```

The coverage probabilities of the Monte Carlo 95% Confidence intervals are:

```{r}
#| echo: true

in.interval.sigma2 = sigma^2 >= sigma2.lower & sigma^2 <= sigma2.upper
in.interval.sigma = sigma >= sigma.lower & sigma <= sigma.upper
ans = rbind(c(sum(in.interval.sigma2)/n.sim,0.95),
            c(sum(in.interval.sigma)/n.sim,0.95))
colnames(ans) = c("MC Coverage", "True Coverage")
rownames(ans) = c("Variance", "Volatility")
ans
```


### Evaluating the Statistical Properties of $\hat{\sigma}_{ij}$ and  $\hat{\rho}_{ij}$ by Monte Carlo simulation

To evaluate the statistical properties of $\hat{\sigma}_{ij}$ and  $\hat{\rho}_{ij}=\mathrm{cor}(R_{it},R_{jt})$, you must simulate from the GWN model in matrix form @eq-GWN-model-matrix. For example, consider the bivariate GWN model: 
$$
\left(\begin{array}{c}
R_{1t}\\
R_{2t}
\end{array}\right)=\left(\begin{array}{c}
0.05\\
0.03
\end{array}\right)+\left(\begin{array}{c}
\varepsilon_{1t}\\
\varepsilon_{2t}
\end{array}\right),~t=1,\ldots,100,
$$ {#eq-bivariate-GWN-model-1}
$$
\left(\begin{array}{c}
\varepsilon_{1t}\\
\varepsilon_{2t}
\end{array}\right)\sim iid~N\left(\left(\begin{array}{c}
0\\
0
\end{array}\right),\left(\begin{array}{cc}
(0.10)^{2} & (0.75)(0.10)(0.05)\\
(0.75)(0.10)(0.05) & (0.05)^{2}
\end{array}\right)\right),
$$ {#eq-bivariate-GWN-model-2}

where $\mu_{1}=0.05$, $\mu_{2}=0.03$, $\sigma_{1}=0.10$, $\sigma_{2}=0.05$,
and $\rho_{12}=0.75$. You use the simulation model @eq-bivariate-GWN-model-1
- @eq-bivariate-GWN-model-2 with $N=1000$ simulated samples
of size $T=100$ to compute the estimates $\{\hat{\rho}_{12}^{1},\ldots,\hat{\rho}_{12}^{1000}\}$. The R code to peform the Monte Carlo simulation is:

```{r}
#| echo: true

library(mvtnorm)
n.obs = 100
n.sim = 1000
set.seed(111)
sim.corrs = rep(0, n.sim)	# initialize vectors
sim.covs = rep(0, n.sim)
muvec = c(0.05, 0.03)
sigmavec = c(0.1, 0.05)
names(muvec) = names(sigmavec) = c("Asset.1", "Asset.2")
rho12 = 0.75
cov12 = rho12*sigmavec[1]*sigmavec[2]
Sigma = diag(sigmavec^2)
Sigma[1,2] = Sigma[2,1] = cov12
rho.lower = rep(0,n.sim)  
rho.upper = rep(0,n.sim)
cov.lower = rep(0,n.sim)
cov.upper = rep(0,n.sim)
for (sim in 1:n.sim) {
	sim.ret = rmvnorm(n.obs,mean=muvec, sigma=Sigma)
	sim.corrs[sim] = cor(sim.ret)[1,2]
	sim.covs[sim] = cov(sim.ret)[1,2]
	sig1hat2 = cov(sim.ret)[1,1]
	sig2hat2 = cov(sim.ret)[2,2]
	se.rhohat = (1 - sim.corrs[sim]^2)/sqrt(n.obs)
	se.covhat = sqrt((sig1hat2*sig2hat2)*(1+sim.corrs[sim]^2))/sqrt(n.obs)
	rho.lower[sim] = sim.corrs[sim] - 2*se.rhohat
	rho.upper[sim] = sim.corrs[sim] + 2*se.rhohat
	cov.lower[sim] = sim.covs[sim] - 2*se.covhat
	cov.upper[sim] = sim.covs[sim] + 2*se.covhat
}
```

The histograms of these values, with the asymptotic normal distributions
overlaid, is displayed in @fig-CerModelEstimationMcRho.


```{r}
#| label: fig-CerModelEstimationMcRho
#| echo: false
#| results: hide
#| fig-cap: Histograms of $\hat{\rho}_{12}$ computed from $N=1000$ Monte Carlo samples
#|   from the bivariate GWN model
#| out-width: 100%
#| fig-asp: 0.75
#| fig-align: center
secovhat.a = sqrt((sigmavec[1]^2*sigmavec[2]^2)*(1+rho12^2))/sqrt(n.obs)
serhohat.a = (1 - rho12^2)/sqrt(n.obs)
par(mfrow=c(1,2))
hist(sim.covs, xlab="Covhat values", col="cornflowerblue", 
     main="covhat", probability=TRUE)
abline(v=cov12, lwd=4, col="white")
x.vals = seq(0.0019, 0.006, length=100)
lines(x.vals, dnorm(x.vals, mean=cov12, sd=secovhat.a), col="orange", lwd=2)
hist(sim.corrs, xlab="rhohat values", col="cornflowerblue", 
     main="rhohat", probability=TRUE)
abline(v=rho12, lwd=4, col="white")
x.vals = seq(0.5, 0.9, length=100)
lines(x.vals, dnorm(x.vals, mean=rho12, sd=serhohat.a), col="orange", lwd=2)
```

The histogram for the $\hat{\sigma}_{12}$ values is bell-shaped with
and centered at the true value $\sigma_{12}=0.00375$. The
sample mean and standard deviation values of $\hat{\sigma}_{12}$ across
the 1000 simulations are, respectively, 
\begin{align*}
\overline{\hat{\sigma}}_{12} & =\frac{1}{1000}\sum_{j=1}^{1000}\hat{\sigma}_{12}^{j}=0.00376,\\
\hat{\sigma}_{\hat{\sigma}_{12}} & =\sqrt{\frac{1}{999}\sum_{j=1}^{1000}(\hat{\sigma}_{12}^{j}-0.00376)^{2}}=0.000628.
\end{align*}

Here, you see that $\hat{\sigma_{12}}$ is essentially unbiased and the Monte Carlo standard error is almost identical to the analytic standard based the asymptotic distribution of $\hat{\sigma}_{12}$.

The histogram for the $\hat{\rho}_{12}$ values is bell-shaped with
a mild left skewness and centered close to $\rho_{12}=0.75$. The
sample mean and standard deviation values of $\hat{\rho}_{12}$ across
the 1000 simulations are, respectively, 
\begin{align*}
\overline{\hat{\rho}}_{12} & =\frac{1}{1000}\sum_{j=1}^{1000}\hat{\rho}_{12}^{j}=0.747,\\
\hat{\sigma}_{\hat{\rho}_{12}} & =\sqrt{\frac{1}{999}\sum_{j=1}^{1000}(\hat{\rho}_{12}^{j}-0.747)^{2}}=0.045.
\end{align*}

There is very slight downward bias in $\hat{\rho}_{12}$. The Monte
Carlo standard deviation, $\hat{\sigma}_{\hat{\rho}_{12}}$, is very
close to the approximate standard error $\mathrm{se}(\hat{\rho}_{12})=(1-0.75^{2})/\sqrt{100}=0.044$.

For each simulation $j$, the approximate 95\% confidence intervals
\begin{align*}
\hat{\sigma}_{12}^{j} & \pm2\times\widehat{\mathrm{se}}(\hat{\rho}_{12}^{j}),, \\
\hat{\rho}_{12}^{j} & \pm2\times\widehat{\mathrm{se}}(\hat{\rho}_{12}^{j}),
\end{align*}
are computed. The coverage probabilities of these interval is approximated by the fractions of intervals that contain (cover) the true values $\sigma_{12} = 0.00375$ and $\rho_{12}=0.75$, respectively. For the 1000 simulated samples, this fractions turns out to be 0.942 adn 0.952, respectively.

Here is the R code to support the above calculations. The Monte Carlo means, standard deviations, and analytic standard errors are:

```{r}
#| echo: true
secovhat.a = sqrt((sigmavec[1]^2*sigmavec[2]^2)*(1+rho12^2))/sqrt(n.obs)
serhohat.a = (1 - rho12^2)/sqrt(n.obs)
ans = rbind(c(cov12, mean(sim.covs),sd(sim.covs),secovhat.a),
            c(rho12, mean(sim.corrs), sd(sim.corrs), serhohat.a))
colnames(ans) = c("True Value", "MC Mean", "MC SE", "Analytic SE")
rownames(ans) = c("Covariance", "Correlation")
ans
```

The fractions of confidence intervals that cover the true parameters are:

```{r}
#| echo: true

in.interval.rho = (rho12 >= rho.lower) & (rho12 <= rho.upper)
in.interval.cov = (cov12 >= cov.lower) & (cov12 <= cov.upper)
ans = rbind(c(sum(in.interval.cov)/n.sim,0.95),
            c(sum(in.interval.rho)/n.sim,0.95))
colnames(ans) = c("MC Coverage", "True Coverage")
rownames(ans) = c("Covariance", "Correlation")
ans
```

## Sampling distributions of $\hat{\boldsymbol{\mu}}$ and $\hat{\boldsymbol{\Sigma}}$

Consider the GWN model in matrix form @eq-GWN-model-matrix. The estimator of the $N \times 1$ matrix $\boldsymbol{\mu}$ is the sample mean vector @eq-muhat-vector and the estimator of the $N \times N$ matrix $\boldsymbol{\Sigma}$ is the sample covariance matrix @eq-SigmaHat. In this section I give the sampling distributions for the mean vector $\hat{\boldsymbol{\mu}}$ and for the elements of the covariance matrix $\hat{\boldsymbol{\Sigma}}$.

### Sampling distribution of $\hat{\boldsymbol{\mu}}$

The following proposition gives the sampling distribution of the $N \times 1$ vector $\hat{\boldsymbol{\mu}}$.

::: {#prp-sampling-distribution-of-hat-boldsymbol-mu .proposition name="Sampling distribution of $\\hat{\\boldsymbol{\\mu}}$"}
Under the GWN model and for large enough T, the sampling distribution for $\hat{\boldsymbol{\mu}}$ is *multivariate normal*:

$$
\hat{\boldsymbol{\mu}} \sim N\left(\boldsymbol{\mu}, ~ \frac{1}{T}\hat{\boldsymbol{\Sigma}}\right).
$$
:::

This result shows that the elements of the vector $\hat{\boldsymbol{\mu}}$ are correlated, with the correlations being the same as the asset return correlations. 

::: {#exm-joint-distribution-of-muhat1-and-muhat2 .example name="Joint distribution of $\\hat{\\mu}_1$ and $\\hat{\\mu}_2$"}
Consider the bivariate case ($N=2$). Proposition 7.8 tells us the joint distribution between $\hat{\mu}_1$ and $\hat{\mu}_2$ is bivariate normal:

$$
\left( \begin{array}{c} \hat{\mu}_1 \\
\hat{\mu}_2 \end{array} \right) \sim N\left(  \left(\begin{array}{c}  \mu_1 \\ \mu_2 \end{array} \right), ~ \frac{1}{T} \left( \begin{array}{cc} \hat{\sigma}_1^2 & \hat{\sigma}_{12} \\
\hat{\sigma}_{12} & \hat{\sigma}_{2}^{2} \end{array} \right)
\right).
$$

Hence, $\widehat{\mathrm{cov}}(\hat{\mu}_1, \hat{\mu}_2) = \frac{1}{T}\hat{\sigma}_{12}$ and $\widehat{\mathrm{cor}}(\hat{\mu}_1, \hat{\mu}_2)=\hat{\rho}_{12}$.

The fact that the elements of $\hat{\mu}$ are correlated and the correlations are equal to the correlations between the corresponding assets is a useful result. For example, if assets 1 and 2 are strongly positively (negatively) correlated, then $\hat{\mu}_1$ and $\hat{\mu}_2$ are strongly positively (negatively) correlated. As a result, a positive (negative) estimate of $\hat{\mu}_1$ will tend to be associated with a positive (negative) estimate of $\hat{\mu}_2$.
\hfill{}$\blacksquare$
:::

### Sampling distribution of the elements of $\hat{\boldsymbol{\Sigma}}$

You can also determine the sampling distribution for the vector of unique elements of $\hat{\boldsymbol{\Sigma}}$. This requires a bit of new matrix notation because the joint distribution of a $N \times N$ symmetric matrix of random variables is defined as the joint distribution of the $N(N+1)/2$ elements of the matrix stacked into a $N(N+1)/2 \times 1$ vector. 

Define $\hat{\boldsymbol{\theta}}$ as the $N(N+1)/2 \times 1$ vector of unique elements of $\boldsymbol{\Sigma}$. These elements can be extracted using the $\mathrm{vech}(\cdot)$ operator, which extracts the elements on and below the main diagonal of $\hat{\boldsymbol{\Sigma}}$ column-wise and stacks them into a $N(N+1)/2$ vector. For example, if $N=3$ then $\hat{\boldsymbol{\theta}}$ is the $6 \times 1$ vector 

$$
\hat{\boldsymbol{\theta}} = \mathrm{vech}(\hat{\boldsymbol{\Sigma}}) = \mathrm{vech} \left(\begin{array}{ccc}
\hat{\sigma}_1^2 & \hat{\sigma}_{12} & \hat{\sigma}_{13} \\
\hat{\sigma}_{12} & \hat{\sigma}_2^2 & \hat{\sigma}_{23} \\
\hat{\sigma}_{13} & \hat{\sigma}_{23} & \hat{\sigma}_{3}^2
\end{array} \right) = 
\left(\begin{array}{c} \hat{\sigma}_1^2 \\ \hat{\sigma}_{12}\\ \hat{\sigma}_{13}\\ \hat{\sigma}_2^2\\ \hat{\sigma}_{23}\\ \hat{\sigma}_{3}^2 \end{array} \right).
$$

::: {#prp-sampling-distribution-of-the-unique-elements-of-sigmahat .proposition name="Sampling distribution of the unique elements of $\\hat{\\boldsymbol{\\Sigma}}$"}
Under the GWN model and for large enough T, the sampling distribution for the $N(N+1)/2 \times 1$ vector $\hat{\boldsymbol{\theta}} = \mathrm{vech}(\hat{\boldsymbol{\Sigma}})$ is multivariate normal:

$$
\hat{\boldsymbol{\theta}} \sim N\left(\boldsymbol{\theta}, ~ \frac{1}{T}\hat{\boldsymbol{\Omega}}\right).
$$
where $\frac{1}{T}\hat{\boldsymbol{\Omega}}$ is an $N(N+1)/2 \times N(N+1)/2$ matrix giving the estimated variances and covariances of the unique elements of $\hat{\boldsymbol{\Sigma}}$. The element of $\frac{1}{T}\hat{\boldsymbol{\Omega}}$ corresponding to $\widehat{\mathrm{cov}}(\hat{\sigma}_{ij},\hat{\sigma}_{lm})$ is given by $T^{-1}(\hat{\sigma}_{il}\hat{\sigma}_{jm} + \hat{\sigma}_{im}\hat{\sigma}_{jl})$ for all $i,j,l,m = 1,2,\ldots,N$, including $i=j=l=m.$  If $i=j$ and $l=m$ then 

\begin{align*}
\widehat{\mathrm{cov}}(\hat{\sigma}_{ij},\hat{\sigma}_{lm}) & = \widehat{\mathrm{cov}}(\hat{\sigma}_{ii},\hat{\sigma}_{ll}) = \widehat{\mathrm{cov}}(\hat{\sigma}_i^2,\hat{\sigma}_l^2) \\  & =\frac{\hat{\sigma}_{il}\hat{\sigma}_{il} + \hat{\sigma}_{il}\hat{\sigma}_{il}}{T} 
= \frac{\hat{\sigma}_{il}^2 + \hat{\sigma}_{il}^2}{T} = \frac{2\hat{\sigma}_{il}^2}{T}
\end{align*}.

If $i=j=l=m$ then 

\begin{align*}
\widehat{\mathrm{cov}}(\hat{\sigma}_{ij},\hat{\sigma}_{lm}) & = \widehat{\mathrm{cov}}(\hat{\sigma}_i^2,\hat{\sigma}_i^2) = \widehat{\mathrm{var}}(\hat{\sigma}_i^2) \\  & =\frac{\hat{\sigma}_i^2\hat{\sigma}_i^2 + \hat{\sigma}_i^2\hat{\sigma}_i^2}{T} = \frac{2\hat{\sigma}_i^4}{T}
\end{align*}.
:::

In Proposition 7.9, it is not easy to see what are the elements of $\hat{\boldsymbol{\Omega}}$. A simple example can make things more clear. Consider the bivariate case $N=2$. Then the result in Proposition 7.9 gives:

\begin{align*}
\hat{\boldsymbol{\theta}} = \left(\begin{array}{c} \hat{\sigma}_1^2 \\
\hat{\sigma}_{12} \\ \hat{\sigma}_2^2 \end{array} \right) & \sim
N\left( \left(\begin{array}{c} \sigma_1^2\\ 
\sigma_{12} \\ \sigma_2^2 \end{array} \right),~ \frac{1}{T}\left( \begin{array}{ccc}
\hat{\omega}_{11} & \hat{\omega}_{12} & \hat{\omega}_{13} \\ \hat{\omega}_{21} & \hat{\omega}_{22} & \hat{\omega}_{23} \\ \hat{\omega}_{31} & \hat{\omega}_{32} & \hat{\omega}_{33} 
\end{array} \right) \right) \\
 & \sim
N\left( \left(\begin{array}{c} \sigma_1^2\\ 
\sigma_{12} \\ \sigma_2^2 \end{array} \right),~ \left( \begin{array}{ccc}
\widehat{\mathrm{var}}(\hat{\sigma}_1^2) & \widehat{\mathrm{cov}}(\hat{\sigma}_1^2, \hat{\sigma}_{12}) & \widehat{\mathrm{cov}}(\hat{\sigma}_1^2, \hat{\sigma}_2^2) \\ \widehat{\mathrm{cov}}(\hat{\sigma}_{12}, \hat{\sigma}_1^2) & \widehat{\mathrm{var}}(\hat{\sigma}_{12}) & \widehat{\mathrm{cov}}(\hat{\sigma}_{12}, \hat{\sigma}_2^2) \\ \widehat{\mathrm{cov}}(\hat{\sigma}_1^2, \hat{\sigma}_2^2) & \widehat{\mathrm{cov}}(\hat{\sigma}_2^2, \hat{\sigma}_{12}) & \widehat{\mathrm{var}}(\hat{\sigma}_2^2) 
\end{array} \right) \right) \\
 & \sim
N\left( \left(\begin{array}{c} \sigma_1^2\\ 
\sigma_{12} \\ \sigma_2^2 \end{array} \right),~ \frac{1}{T}\left( \begin{array}{ccc}
2\hat{\sigma}_1^4 & 2\hat{\sigma}_1^2\hat{\sigma}_{12} & 2\hat{\sigma}_{12}^2 \\ 2\hat{\sigma}_1^2 \hat{\sigma}_{12} & \hat{\sigma}_1^2
\hat{\sigma}_2^2 + \hat{\sigma}_{12}^2 & 2\hat{\sigma}_{12}\hat{\sigma}_2^2 \\ 2\hat{\sigma}_{12} & 2\hat{\sigma}_{12} \hat{\sigma}_2^2 & 2\hat{\sigma}_2^4 
\end{array} \right) \right).
\end{align*}

Notice that the results for the diagonal elements of $\frac{1}{T}\hat{\boldsymbol{\Omega}}$ match those given earlier (see equations @eq-sesigma2hat and @eq-sesigmaijhat). What is new here are formulas for the off-diagonal elements.

### Joint sampling distribution between $\hat{\boldsymbol{\mu}}$ and $\mathrm{vech}(\hat{\boldsymbol{\Sigma}})$

When you study hypothesis testing in Chapter @sec-Hypothesis-Testing and the statistical analysis of portfolios in Chapter @sec-Statistical-Analysis-of-Portfolios you will need to use results on the joint sampling distribution between $\hat{\boldsymbol{\mu}}$ and $\mathrm{vech}(\hat{\boldsymbol{\Sigma}})$. The follow proposition states the result:

::: {#prp-joint-sampling-distribution-between-mu-and-sigma .proposition name="Joint sampling distribution between $\\hat{\\boldsymbol{\\mu}}$ and $\\mathrm{vech}(\\hat{\\boldsymbol{\\Sigma}})$"}
Under the GWN model and for large enough T, the joint sampling distribution for the $N+N(N+1)/2 \times 1$ vector $(\hat{\boldsymbol{\mu}}', \hat{\boldsymbol{\theta}}')'$, where $\hat{\boldsymbol{\theta}} = \mathrm{vech}(\hat{\boldsymbol{\Sigma}})$ is multivariate normal
$$
\left(\begin{array}{c} \hat{\boldsymbol{\mu}} \\ \hat{\boldsymbol{\theta}} \end{array} \right) \sim N \left( \left( \begin{array}{c} \boldsymbol{\mu} \\ \boldsymbol{\theta} \end{array} \right), ~ \frac{1}{T} \left( \begin{array}{cc} \hat{\boldsymbol{\Sigma}} & 0 \\ 0 & \hat{\boldsymbol{\Omega}} \end{array} \right) \right).
$$
:::

Proposition 7.10 tells us that the vectors $\hat{\boldsymbol{\mu}}$ and $\hat{\boldsymbol{\theta}}$ are jointly multivariate normally distributed and that all elements of $\hat{\boldsymbol{\mu}}$ are independent of all the elements of $\hat{\boldsymbol{\theta}}$. 

::: {#exm-joint-distribution-of-muhat-and-sigmahat .example name="Joint distribution of $\\hat{\\mu}$, and $\\hat{\\sigma}$"}
Consider the GWN model for a single asset and suppose you are interested in the joint distribution of the vector $(\hat{\mu}, \hat{\sigma}^2)'$. Then Proposition 7.10 gives us the result:

$$
\left( \begin{array}{c} 
\hat{\mu} \\ 
\hat{\sigma}^2 
\end{array} \right) \sim N \left( \left( \begin{array}{c} 
\mu \\ 
\sigma^2 
\end{array} \right), ~ \left( \begin{array}{cc} 
\frac{\hat{\sigma}^2}{T} & 0 \\ 
0 & \frac{2\hat{\sigma}^4}{T} 
\end{array} \right) \right). 
$$ {#eq-EST-joint-mu-sigma2}

Here, you see that $\mathrm{cov}(\hat{\mu}, \hat{\sigma}^2)=0$, which implies that $\hat{\mu}$ and $\hat{\sigma}^2$ are independent since they are jointly normally distributed. Furthermore, since $\hat{\mu}$ and $\hat{\sigma}^2$ are independent any function of $\hat{\mu}$ is independent of any function of $\hat{\sigma}^2$. For example, $\hat{\sigma} = \sqrt{\hat{\sigma}^2}$ is independent of $\hat{\mu}$. Using @eq-pdf-sigma you can then deduce the joint distribution of $(\hat{\mu}, \hat{\sigma})'$:

$$
\left( \begin{array}{c} 
\hat{\mu} \\ 
\hat{\sigma} 
\end{array} \right) \sim N \left( \left( \begin{array}{c} 
\mu \\ 
\sigma 
\end{array} \right), ~ \left( \begin{array}{cc} 
\frac{\hat{\sigma}^2}{T} & 0 \\ 
0 & \frac{\hat{\sigma}^2}{2T} 
\end{array} \right) \right). 
$$ {#eq-EST-joint-mu-sigma}

\hfill{}$\blacksquare$
:::

## Further Reading: Estimation of The GWN Model

-  Good introductory statistical theory is given in [@degroot-2011]. For an econometrics focus see [@goldberger-1991]. More advanced books - [@cb-2002], Bickell and Dokson. 
-  R books on statistics, Monte Carlo and bootstrapping: [@vinod-2010] (bootstrapping) 
-  Surprisingly, the statistical properties of VaR are often not discussed much in standard textbook treatments. See, for example [@jorion-1997].


## Problems: Estimation of The GWN Model
::: {#exr-01 .exercise}
Show that the matrix sample statistics @eq-muhat-vector and
@eq-Sigmahat-matrix are unbiased estimators for $\boldsymbol{\mu}$ and
$\boldsymbol{\Sigma}$, respectively. 
:::


