# Estimating Functions of GWN Model Parameters {#sec-The-Bootstrap}

Updated: September 5, 2026

Copyright © Eric Zivot 2026

In the previous chapter, I considered estimation of the parameters of the GWN model
for asset returns and the construction of estimated standard errors and confidence intervals for the estimated parameters. In this chapter, I consider estimation of functions of the parameters of the GWN model, and the construction of estimated standard errors and confidence intervals for these estimates. This is an important topic as you are often more interested in a function of the GWN model parameters (e.g., $\mathrm{VaR}_{\alpha}$) than the GWN model parameters themselves. In order to do this, I introduce some new statistical tools including the analytic delta method based on asymptotic distribution theory, and two computer intensive resampling methods called the jackknife and the bootstrap. 

The R packages I use in this chapter are **boot**, **bootstrap**, **car**, 
**introcompfinr**, and **PerformanceAnalytics**. Have these packages installed and 
loaded in R before replicating the chapter examples.

```{r}
#| include: false
suppressPackageStartupMessages(library(boot))
suppressPackageStartupMessages(library(bootstrap))
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```

## Functions of GWN Model Parameters {#sec-Functions-of-GWN-Model-Parameters}

Consider the GWN model for returns (simple or continuously compounded):

\begin{align*}
R_{it} & =\mu_{i}+\varepsilon_{it},\\
\{\varepsilon_{it}\}_{t=1}^{T} & \sim\mathrm{GWN}(0,\sigma^{2}), \\
\mathrm{cov}(\varepsilon_{it},\varepsilon_{js}) & =\left\{ \begin{array}{c}
\sigma_{ij}\\
0
\end{array}\right.\begin{array}{c}
~ t=s\\
t\neq s
\end{array}.
\end{align*}

Let $\boldsymbol{\theta}$ denote a $k \times 1$ vector of GWN model parameters from which you want to compute a function of interest (I drop the asset subscript $i$ to simplify notation). Let $f:R^k \rightarrow R$ be a continuous and differentiable function of $\boldsymbol{\theta}$, and define $\eta = f(\boldsymbol{\theta})$ as the scalar parameter of interest. To make things concrete, let $\boldsymbol{\theta} = (\mu,\sigma)^{\prime}$. Consider the following functions of $\boldsymbol{\theta}$:

$$
f_{1}(\boldsymbol{\theta}) = \mu + \sigma \times q_{\alpha}^{Z} = q_{\alpha}^R 
$$ {#eq-f1} 
$$
f_{2}(\boldsymbol{\theta}) = -W_{0}\times \left( \mu + \sigma \times q_{\alpha}^{Z} \right) = W_{0} \times f_{1}(\boldsymbol{\theta}) = \mathrm{VaR}_{\alpha}^N 
$$ {#eq-f2} 
$$ 
f_{3}(\boldsymbol{\theta}) = -W_{0}\times \left( \exp \left( \mu + \sigma \times q_{\alpha}^{Z} \right) - 1 \right) = W_{0} \times (\exp(f_{1}(\boldsymbol{\theta})) - 1) = \mathrm{VaR}_{\alpha}^{LN} 
$$ {#eq-f3}

The first function @eq-f1 is the $\alpha$-quantile, $q_{\alpha}^R$, of the $N(\mu,\sigma^{2})$
distribution for simple returns. The second function @eq-f2 is $\mathrm{VaR}_{\alpha}$ of an initial investment of $W_{0}$ based on the normal distribution for simple returns. The third function @eq-f3 is $\mathrm{VaR}_{\alpha}$ based on the log-normal distribution for simple returns (normal distribution for continuously compounded returns). The first two functions are linear functions of the elements of $\boldsymbol{\theta}$, and the third function is a nonlinear function of the elements of $\boldsymbol{\theta}$.

Given the plug-in estimate of $\boldsymbol{\theta}$, $\hat{\boldsymbol{\theta}} = (\hat{\mu},\hat{\sigma})'$, I want to estimate each of the above functions of $\boldsymbol{\theta}$, and construct estimated standard errors and confidence intervals. 

::: {#exm-data-for-examples .example name="Data for examples"}
For the examples in this chapter, I use data on monthly continuously
compounded and simple returns for Microsoft over the period January 1998, through
May 2012, from the **introcompfinr** package. The data is the
same as that used in Chapter @sec-Descriptive-Statistics-for-Financial-Data
and is constructed as follows:

```{r}
#| echo: true
msftPrices = to.monthly(msftDailyPrices, OHLC=FALSE) 
smpl = "1998-01::2012-05" 
msftPrices = msftPrices[smpl] 
msftRetS = na.omit(Return.calculate(msftPrices, method="discrete"))
msftRetC = log(1 + msftRetS)
```

The GWN model estimates for $\mu_{msft}$ and $\sigma_{msft}$, with estimated standard
errors, are:

```{r}
#| echo: true
n.obs = nrow(msftRetC)
muhatC = mean(msftRetC) 
sigmahatC = sd(msftRetC) 
muhatS = mean(msftRetS)
sigmahatS = sd(msftRetS)
estimates = c(muhatC, sigmahatC, muhatS, sigmahatS)
se.muhatC = sigmahatC/sqrt(n.obs) 
se.muhatS = sigmahatS/sqrt(n.obs) 
se.sigmahatC = sigmahatC/sqrt(2*n.obs) 
se.sigmahatS = sigmahatS/sqrt(2*n.obs) 
stdErrors = c(se.muhatC, se.sigmahatC, se.muhatS, se.sigmahatS)
ans = rbind(estimates, stdErrors) 
colnames(ans) = c("mu.cc", "sigma.cc", "mu.simple", "sigma.simple") 
ans
```

\hfill{}$\blacksquare$
:::

## Estimation of Functions of GWN Model Parameters

In Chapter @sec-Estimation-of-The-GWN-Model, I used the *plug-in principle* to motivate estimators of the GWN model
parameters $\mu_i$ $\sigma_{i}^2$, $\sigma_{ij}$, and $\rho_{ij}$. I can also use the plug-in principle to estimate *functions* of the GWN model parameters.

::: {#def-plug-in-principle-for-functions-of-gwn-model-parameters .definition name="Plug-in principle for functions of GWN model parameters"}
Let $\boldsymbol{\theta}$ denote a $k \times 1$ vector of GWN model parameters, and let $\eta = f(\boldsymbol{\theta})$
where $f:\mathbb{R}^k \rightarrow \mathbb{R}$ is a scalar function of $\boldsymbol{\theta}$. Let $\hat{\boldsymbol{\theta}}$ denote the plug-in estimator of $\boldsymbol{\theta}$. Then the plug-in estimator of $\eta$ is $\hat{\eta} = f(\hat{\boldsymbol{\theta}})$.
:::

::: {#exm-plug-in-estimates-of-example-functions-for-example-data .example name="Plug-in estimates of example functions for example data"}
Let $W_0 = \$100,000$, and $r_f = 0.03/12 = 0.0025$. Using the plug-in estimates of the GWN model parameters, the plug-in estimates of the functions @eq-f1 - @eq-f3 are:

```{r}
W0 = 100000
r.f = 0.03/12
f1.hat = muhatS + sigmahatS*qnorm(0.05)
f2.hat = -W0*f1.hat
f3.hat = -W0*(exp(muhatC + sigmahatC*qnorm(0.05)) - 1)
fhat.vals = cbind(f1.hat, f2.hat, f3.hat)
colnames(fhat.vals) = c("f1", "f2", "f3")
rownames(fhat.vals) = "Estimate"
fhat.vals
```

\hfill{}$\blacksquare$
:::

Plug-in estimators of functions are random variables and are subject to estimation error. Just as you studied the statistical properties of the plug-in estimators of the GWN model parameters you can study the statistical properties of the plug-in estimators of functions of GWN model parameters.  

### Bias

As discussed in Chapter @sec-Estimation-of-The-GWN-Model, a desirable finite sample property of an estimator $\hat{\boldsymbol{\theta}}$ of $\boldsymbol{\theta}$ is *unbiasedness*: on average over many hypothetical samples $\hat{\boldsymbol{\theta}}$ is equal to $\boldsymbol{\theta}$. When estimating $f(\boldsymbol{\theta})$, unbiasedness of $\hat{\boldsymbol{\theta}}$ may or may not carry over to $f(\hat{\boldsymbol{\theta}})$. 

Suppose $f(\boldsymbol{\theta})$ is a linear function of the elements of $\boldsymbol{\theta}$ so that

$$
f(\boldsymbol{\theta}) = a + b_1 \theta_1 + b_2 \theta_2 + \cdots + b_k \theta_k.
$$

where $a, b_1, b_2, \ldots b_k$ are constants.  For example, the functions @eq-f1 and @eq-f2 are linear functions of the elements of $\boldsymbol{\theta} = (\mu, \sigma)^{\prime}.$ Let $\hat{\boldsymbol{\theta}}$ denote an estimator of $\boldsymbol{\theta}$ and let $f(\hat{\boldsymbol{\theta}})$ denote the plug-in estimator of $f(\boldsymbol{\theta})$. If $E[\hat{\boldsymbol{\theta}}]=\boldsymbol{\theta}$ for all elements of $\boldsymbol{\theta}$ ($\hat{\theta_i}$ is unbiased for $\theta_i$, $i=1,\ldots, k$), then 

\begin{align*}
E[f(\hat{\boldsymbol{\theta}})] & = a + b_1 E[\hat{\theta_1}] + b_2 E[\hat{\theta_2}] + \cdots + b_k E[\hat{\theta_k}] \\
 & = a + b_1 \theta_1 + b_2 \theta_2 + \cdots + b_k \theta_k = f(\boldsymbol{\theta})
\end{align*}

and so $f(\hat{\boldsymbol{\theta}})$ is unbiased for $f(\boldsymbol{\theta})$.

Now suppose $f(\boldsymbol{\theta})$ is a nonlinear function of the elements of $\boldsymbol{\theta}$. For example, the function @eq-f3 is a nonlinear function of the elements of $\boldsymbol{\theta} = (\mu, \sigma)^{\prime}.$ Then, in general, $f(\hat{\boldsymbol{\theta}})$ is not unbiased for $f(\boldsymbol{\theta})$ even if $\hat{\boldsymbol{\theta}}$ is unbiased for $\boldsymbol{\theta}$. The direction of the bias depends on the properties of $f(\boldsymbol{\theta})$. If $f(\cdot)$ is convex at the scalar parameter $\theta$ or concave at $\theta$ then you can say something about the direction of the bias in $f(\hat{\theta})$. 

::: {#def-convex-function .definition name="Convex function"}
Let $\theta$ be a scalar parameter. A scalar function $f(\theta)$ is *convex* (in two dimensions) if all points on a straight line connecting any two points on the graph of $f(\theta)$ is above or on that graph. More formally, $f(\theta)$ is convex if for all $\theta_1$ and $\theta_2$ and for all $0 \le \alpha \le1$
  
$$
f(\alpha \theta_1 + (1-\alpha) \theta_2) \le \alpha f(\theta_1) + (1 - \alpha)f(\theta_2)
$$ 
  
:::

For example, the function $f(\theta)=\theta^2$ is a convex function. Another common convex function is $\exp(\theta)$.

::: {#def-concave-function .definition name="Concave function"}
Let $\theta$ be a scalar parameter. A scalar function $f(\theta)$ is *concave* if $-f(\theta)$ is convex. 
:::

For example, the function $f(\theta) = - \theta^2$ is concave. Other common concave functions are $log(\theta)$ and $\sqrt{\theta}$.

::: {#prp-jensen-s-inequality .proposition name="Jensen's inequality"}
Let $\theta$ be a scalar parameter and let $\hat{\theta}$ be an unbiased estimate of $\theta$ so that $E[\hat{\theta}]=\theta$. Let $f(\theta)$ be a convex function of $\theta$. Then
$$
E[f(\hat{\theta)}] \ge f(E[\hat{\theta}]) = f(\theta).
$$
:::

Jensen's inequality tells you that if $f$ is convex (concave) and $\hat{\theta}$ is unbiased then $f(\hat{\theta})$ is positively (negatively) biased. For example, you know from Chapter @sec-Estimation-of-The-GWN-Model that $\hat{\sigma}^2$ is unbiased for $\sigma^2$. Now, $\hat{\sigma} = f(\hat{\sigma}^2) = \sqrt{\hat{\sigma}^2}$ is a concave function of $\hat{\sigma}$. From Jensen's inequality, you know that $\hat{\sigma}$ is negatively biased (i.e., too small). 

\hfill{}$\blacksquare$


### Consistency

An important property of the plug-in estimators of the GWN model parameters is that they are *consistent*: as the sample size gets larger and larger the estimators get closer and closer to their true values and eventually equal their true values for an infinitely large sample. 

Consistency also holds for plug-in estimators of functions of GWN model parameters. The justification comes from *Slutsky's Theorem*. 

::: {#prp-slutsky-s-theorem .proposition name="Slutsky's Theorem"}
Let $\hat{\boldsymbol{\theta}}$ be consistent for $\boldsymbol{\theta}$, and let $f:\mathbb{R}^k \rightarrow \mathbb{R}$ be continuous at $\boldsymbol{\theta}$. Then $f(\hat{\boldsymbol{\theta}})$ is consistent for $f(\boldsymbol{\theta})$.

:::

The key condition on $f(\boldsymbol{\theta})$ is continuity at $\boldsymbol{\theta}$.^[Recall, the intuitive definition of continuity is that you can draw the function on paper without lifting up your pen.]

::: {#exm-consistency-of-functions-of-gwn-model-parameters .example name="Consistency of functions of GWN model parameters"}

All of the functions @eq-f1 - @eq-f3 are continuous at $\boldsymbol{\theta} = (\mu, \sigma)'$. By Slutsky's Theorem, all of the plug-in estimators of the functions are also consistent.

\hfill{}$\blacksquare$
:::

## Asymptotic normality and the delta method

Let $\theta$ denote a scalar parameter of the GWN model for returns and let $\hat{\theta}$ denote the plug-in estimator of $\theta$. In Chapter @sec-Estimation-of-The-GWN-Model you learned that $\hat{\theta}$ is asymptotically normally distributed

$$
\hat{\theta}\sim N(\theta,\widehat{\mathrm{se}}(\hat{\theta})^{2}),
$$

for large enough $T$, where $\widehat{\mathrm{se}}(\hat{\theta})$ denotes the estimated standard error of $\hat{\theta}.$ This result is justified by the CLT. The following proposition establishes the asymptotic distribution of the plug-in estimator of a function of $\theta$ using a technique called the *delta method*.

::: {#prp-delta-method .proposition name="The delta method"}
Let $f(\theta)$ denote a real-valued continuous and differentiable scalar function of $\theta$. Using a technique called the *delta method*, you can show that the plug-in estimator $f(\hat{\theta})$ is also asymptotically normally distributed: 

$$
f(\hat{\theta}) \sim N(f(\theta),\widehat{\mathrm{se}}(f(\hat{\theta}))^{2}),
$$

for large enough $T$, where $\widehat{\mathrm{se}}(f(\hat{\theta}))$ is the estimated standard error for $f(\hat{\theta})$. 

The *delta method* shows you how to compute $\widehat{\mathrm{se}}(f(\hat{\theta}))^2$:

$$
\widehat{\mathrm{se}}(f(\hat{\theta}))^{2} = f'(\hat{\theta})^2 \times \widehat{\mathrm{se}}(\hat{\theta})^{2},
$$

where $f'(\hat{\theta})$ denotes the derivative of $f(\theta)$ evaluated at $\hat{\theta}$. Then,

$$
\widehat{\mathrm{se}}(f(\hat{\theta})) = \sqrt{f'(\hat{\theta})^2 \times \widehat{\mathrm{se}}(\hat{\theta})^{2}}.
$$ {#eq-delta-method-se}

:::

::: {#exm-delta-method-standard-error-for-hat-sigma .example name="Delta method standard error for $\\hat{\\sigma}$"}

The results from Chapter @sec-Estimation-of-The-GWN-Model showed, for large enough $T$, 

\begin{equation*}
\hat{\sigma}^2 \sim N(\sigma^2, \mathrm{se}(\hat{\sigma}^2)^2), ~ \hat{\sigma} \sim N(\sigma,\mathrm{se}(\hat{\sigma})^2),
\end{equation*}

where $\mathrm{se}(\sigma^2)^2 = 2\sigma^4/T$ and $\mathrm{se}(\sigma)^2 = \sigma^2/2T$. The formula for $\mathrm{se}(\hat{\sigma})^2$ is the result of the delta method for the function $f(\sigma^2)= \sqrt{\sigma^2}=\sigma$ applied to the asymptotic distribution of $\hat{\sigma}^2$. To see this, first note that by the chain rule

\begin{equation*}
f'(\sigma^2) = \frac{1}{2} (\sigma^2)^{-1/2},
\end{equation*}

so that $f'(\sigma^2)^2 = \frac{1}{4}(\sigma^2)^{-1}$. Then, 

\begin{align*}
\mathrm{se}(f(\hat{\theta}))^{2} = f'(\sigma^2)^2 \times \mathrm{se}(\hat{\sigma}^2)^2 & = \frac{1}{4}(\sigma^2)^{-1} \times 2\sigma^4/T \\
& = \frac{1}{2}\sigma^2/T = \sigma^2/2T \\
& = \mathrm{se}(\hat{\sigma})^2.
\end{align*}

The estimated squared standard error replaces $\sigma^2$ with its estimate $\hat{\sigma}^2$ and is given by
\begin{equation*}
\widehat{\mathrm{se}}(\hat{\sigma}^2)^2 =  \hat{\sigma}^2/2T.
\end{equation*}

\hfill{}$\blacksquare$
:::

Now suppose $\boldsymbol{\theta}$ is a $k \times 1$ vector of GWN model parameters, and $f:\mathbb{R}^k \rightarrow \mathbb{R}$ is a continuous and differentiable function of $\boldsymbol{\theta}$. Define the $k \times 1$ *gradient function* $\mathbf{g}(\boldsymbol{\theta})$ as:

$$
\mathbf{g}(\boldsymbol{\theta}) = \frac{\partial f(\boldsymbol{\theta})}{\partial \boldsymbol{\theta}} = \left( \begin{array}{c} 
\frac{\partial f(\boldsymbol{\theta})}{\partial \theta_1} \\
\vdots \\
\frac{\partial f(\boldsymbol{\theta})}{\partial \theta_k}
\end{array} \right).
$$

Assume that $\hat{\boldsymbol{\theta}}$ is asymptotically normally distributed:

$$
\hat{\boldsymbol{\theta}} \sim N(\boldsymbol{\theta}, \widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})),
$$

where $\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})$ is the $k \times k$ estimated variance-covariance matrix
of $\hat{\boldsymbol{\theta}}$. Then the delta method shows you that $f(\hat{\boldsymbol{\theta}})$ is asymptotically normally distributed:

$$
f(\hat{\boldsymbol{\theta}}) \sim N(f(\boldsymbol{\theta}), \widehat{\mathrm{se}}(f(\hat{\boldsymbol{\theta}}))^{2}),
$$

where

\begin{equation*}
\widehat{\mathrm{se}}(f(\hat{\boldsymbol{\theta}}))^{2} = \mathbf{g}(\hat{\boldsymbol{\theta}})'\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})\mathbf{g}(\hat{\boldsymbol{\theta}}).
\end{equation*}


::: {#exm-delta-method-standard-error-for-hat-q-alpha-z .example name="Delta method standard error for $\\hat{q}_{\\alpha}^Z$"}

Let $\boldsymbol{\theta} = (\mu, \sigma)'$ and consider $f_1(\boldsymbol{\theta}) = q_{\alpha}^R = \mu + \sigma q_{\alpha}^Z$ (normal simple return quantile). The gradient function $\mathbf{g}_1(\boldsymbol{\theta})$ for $f_1(\boldsymbol{\theta})$ is:

\begin{align*}
\mathbf{g}_1(\boldsymbol{\theta}) = \frac{\partial f_1(\boldsymbol{\theta})}{\partial \boldsymbol{\theta}} = \left( \begin{array}{c} 
\frac{\partial f_1(\boldsymbol{\theta})}{\partial \mu} \\
\frac{\partial f_1(\boldsymbol{\theta})}{\partial \sigma}
\end{array} \right) = \left( \begin{array}{c}
1 \\
q_{\alpha}^Z
\end{array} \right).
\end{align*}

Regarding $\mathrm{var}(\hat{\boldsymbol{\theta}})$, recall from Chapter @sec-Estimation-of-The-GWN-Model, in the GWN model: 
\begin{equation*}
\left(\begin{array}{c}
\hat{\mu}\\
\hat{\sigma}^2
\end{array}\right)\sim N\left(\left(\begin{array}{c}
\mu\\
\sigma^2
\end{array}\right),\left(\begin{array}{cc}
\mathrm{se}(\hat{\mu})^{2} & 0\\
0 & \mathrm{se}(\hat{\sigma}^2)^{2}
\end{array}\right)\right)
\end{equation*}

for large enough $T$. I show below, using the delta method, that 

\begin{equation*}
\left(\begin{array}{c}
\hat{\mu}\\
\hat{\sigma}
\end{array}\right)\sim N\left(\left(\begin{array}{c}
\mu\\
\sigma
\end{array}\right),\left(\begin{array}{cc}
\mathrm{se}(\hat{\mu})^{2} & 0\\
0 & \mathrm{se}(\hat{\sigma})^{2}
\end{array}\right)\right)
\end{equation*}

Hence, 

$$
\mathrm{var}(\hat{\boldsymbol{\theta}}) = \left(\begin{array}{cc}
\mathrm{se}(\hat{\mu})^{2} & 0\\
0 & \mathrm{se}(\hat{\sigma})^{2}
\end{array}\right) = \left(\begin{array}{cc}
\frac{\sigma^2}{T} & 0\\
0 & \frac{\sigma^2}{2T}
\end{array}\right)
$$

Then,

\begin{align*}
\mathbf{g}_1(\boldsymbol{\theta})'\mathrm{var}(\hat{\boldsymbol{\theta}})\mathbf{g}_1(\boldsymbol{\theta}) &= \left( \begin{array}{cc}
1 & q_{\alpha}^Z \end{array} \right) \left(\begin{array}{cc}
\mathrm{se}(\hat{\mu})^{2} & 0\\
0 & \mathrm{se}(\hat{\sigma})^{2}
\end{array}\right) \left( \begin{array}{c} 1 \\ q_{\alpha}^Z \end{array} \right) \\
&= \mathrm{se}(\hat{\mu})^{2} + (q_{\alpha}^Z)^2 \mathrm{se}(\hat{\sigma})^{2} \\
& =\frac{\sigma^{2}}{T}+\frac{\left(q_{\alpha}^{Z}\right)^{2}\sigma^{2}}{2T}\\
& =\frac{\sigma^{2}}{T}\left[1+\frac{1}{2}\left(q_{\alpha}^{Z}\right)^{2}\right] \\
&= \mathrm{se}(\hat{q}_{\alpha}^Z)^2.
\end{align*}

The standard error for $\hat{q}_{\alpha}^Z$ is: 
$$
\mathrm{se}(\hat{q}_{\alpha}^{R})=\sqrt{\mathrm{var}(\hat{q}_{\alpha}^{R})}=\frac{\sigma}{\sqrt{T}}\left[1+\frac{1}{2}\left(q_{\alpha}^{Z}\right)^{2}\right]^{1/2}.
$$ {#eq-se-qhat}

The formula @eq-se-qhat shows that $\mathrm{se}(\hat{q}_{\alpha}^{R})$ increases with $\sigma$ and $q_{\alpha}^{Z}$, and decreases with $T$. In addition, for fixed $T$, $\mathrm{se}(\hat{q}_{\alpha}^{R})\rightarrow\infty$ as $\alpha\rightarrow0$ or as $\alpha\rightarrow1$ because $q_{\alpha}^{Z}\rightarrow-\infty$ as $\alpha\rightarrow0$ and $q_{\alpha}^{Z}\rightarrow\infty$ as $\alpha\rightarrow1$. Hence, there is very large estimation error in $\hat{q}_{\alpha}^{R}$ when $\alpha$ is very close to zero or one (for fixed values of $\sigma$ and $T$). This makes intuitive sense as you have very few observations in the extreme tails of the empirical distribution of the data for a fixed sample size and so you expect a larger estimation error.

Replacing $\sigma$ with its estimate $\hat{\sigma}$ gives the *practically useful* estimated standard error for $\hat{q}_{\alpha}^{R}$:

$$
\widehat{\mathrm{se}}(\hat{q}_{\alpha}^{R})=\frac{\hat{\sigma}}{\sqrt{T}}\left[1+\frac{1}{2}\left(q_{\alpha}^{Z}\right)^{2}\right]^{1/2}.
$$ {#eq-se-qhat}

Using the above results, the sampling distribution of $\hat{q}_{\alpha}^{R}$
can be approximated by the normal distribution: 
$$
\hat{q}_{\alpha}^{R}\sim N(q_{\alpha}^{R},~\widehat{\mathrm{se}}(\hat{q}_{\alpha}^{R})^{2}),
$$ {#eq-qhat-distn-simple}

for large enough $T$. 
\hfill{}$\blacksquare$
:::

::: {#exm-applying-the-delta-method-to-hat-q-alpha-r-for-microsoft .example name="Applying the delta method to $\\hat{q}_{\\alpha}^{R}$ for Microsoft"}

The estimates of $q_{\alpha}^{R}$, $\mathrm{se}(\hat{q}_{\alpha}^{R})$, and 95% confidence intervals, for $\alpha=0.05,\,0.01$ and $0.001$, from the simple monthly returns for Microsoft are:

```{r}
#| echo: true
qhat = muhatS + sigmahatS*qnorm(c(0.05,0.01,0.001))
seQhat = (sigmahatS/sqrt(n.obs))*sqrt(1 + 0.5*qnorm(c(0.05,0.01,0.001))^2)
lowerQhat = qhat - 2*seQhat
upperQhat = qhat + 2*seQhat
width = upperQhat - lowerQhat
ans = cbind(qhat,seQhat,lowerQhat,upperQhat,width)
colnames(ans) = c("Estimates", "Std Errors", "2.5%", "97.5%", "Width")
rownames(ans) = c("q.05", "q.01", "q.001")
ans
```

For Microsoft the values of $\widehat{\mathrm{se}}(\hat{q}_{\alpha}^{R})$
are about 1.2\%, 1.5\%, and 1.9\% for $\alpha=0.05$, $\alpha=0.01$
and $\alpha=0.001$, respectively. The 95\% confidence interval widths increase with decreases in $\alpha$ indicating more uncertainty about the true value of $q_{\alpha}^R$ for smaller values of $\alpha$. For example, the 95% confidence for interval $q_{0.01}^R$ indicates that the true loss with $1\%$ probability (one month out of 100 months) is between $19.7\%$ and $25.7\%$.  
\hfill{}$\blacksquare$
:::

::: {#exm-delta-method-standard-errors-for-widehat-mathrm-var-alpha-n-and-widehat-mathrm-var-alpha-ln .example name="Delta method standard errors for $\\widehat{\\mathrm{VaR}}_{\\alpha}^{N}$ and $\\widehat{\\mathrm{VaR}}_{\\alpha}^{LN}$"}

Let $\boldsymbol{\theta} = (\mu, \sigma)'$ and consider the functions $f_2(\boldsymbol{\theta}) = \mathrm{VaR}_{\alpha}^N = -W_0(\mu + \sigma q_{\alpha}^Z)$ (normal VaR), and $f_3(\boldsymbol{\theta}) = \mathrm{VaR}_{\alpha}^{LN} = -W_0(\exp((\mu + \sigma q_{\alpha}^Z)) - 1)$ (log-normal VaR). It is straightforward to show (see end-of-chapter exercises) that the gradient functions are:

\begin{align*}
\mathbf{g}_2(\boldsymbol{\theta}) & = \frac{\partial f_2(\boldsymbol{\theta})}{\partial \boldsymbol{\theta}} = \left( \begin{array}{c} 
-W_0 \\ 
-W_0 q_{\alpha}^Z
\end{array} \right), \\
\mathbf{g}_3(\boldsymbol{\theta}) & = \frac{\partial f_3(\boldsymbol{\theta})}{\partial \boldsymbol{\theta}} = \left( \begin{array}{c} 
-W_0 \exp(\mu + \sigma q_{\alpha}^Z)  \\ 
-W_0 \exp(\mu + \sigma q_{\alpha}^Z)q_{\alpha}^Z 
\end{array} \right).
\end{align*}

The delta method variances are:

\begin{align*}
\mathbf{g}_2(\boldsymbol{\theta})'\mathrm{var}(\hat{\boldsymbol{\theta}})\mathbf{g}_2(\boldsymbol{\theta}) &= W_0^2 \mathrm{se}(\hat{q}_{\alpha}^R)^2 = \mathrm{se} \left( \widehat{\mathrm{VaR}}_{\alpha}^{N} \right )^2 \\
\mathbf{g}_3(\boldsymbol{\theta})'\mathrm{var}(\hat{\boldsymbol{\theta}})\mathbf{g}_3(\boldsymbol{\theta}) &= W_0^2 \exp(q_{\alpha}^R)^2 \mathrm{se}(\hat{q}_{\alpha}^R)^2 = \mathrm{se} \left( \widehat{\mathrm{VaR}}_{\alpha}^{LN} \right )^2.
\end{align*}

The corresponding standard errors are:

\begin{align*}
\mathrm{se} \left( \widehat{\mathrm{VaR}}_{\alpha}^{N} \right ) &= W_0 \mathrm{se}(\hat{q}_{\alpha}^R) \\
\mathrm{se} \left( \widehat{\mathrm{VaR}}_{\alpha}^{LN} \right ) &= W_0 \exp(q_{\alpha}^R) \mathrm{se}(\hat{q}_{\alpha}^R).
\end{align*}

where 

\begin{equation*}
\mathrm{se}(\hat{q}_{\alpha}^Z) = \sqrt{\frac{\sigma^{2}}{T}\left[1+\frac{1}{2}\left(q_{\alpha}^{Z}\right)^{2}\right]}.
\end{equation*}

Remarks: 

1.  $\mathrm{se}(\widehat{\mathrm{VaR}}_{\alpha}^{N})$ and $\mathrm{se}(\widehat{\mathrm{VaR}}_{\alpha}^{LN})$ increase with $W_{0}$, $\sigma$, and $q_{\alpha}^{Z}$, and decrease with $T$. 

2.  For fixed values of $W_{0}$, $\sigma$, and $T$, $\mathrm{se}(\widehat{\mathrm{VaR}}_{\alpha}^N)$ and $\mathrm{se}(\widehat{\mathrm{VaR}}_{\alpha}^{LN})$ become very large as $\alpha$ approaches zero. That is, for very small loss probabilities you will have very imprecise estimates of $\mathrm{VaR}_{\alpha}^N$ and $\mathrm{VaR}_{\alpha}^{LN}$.

3.  The practically useful estimated standard errors replace the unknown value of $q_{\alpha}^R$ with its estimated value $\hat{q}_{\alpha}^R$ and are given by:

$$
\mathrm{\widehat{se}} \left( \widehat{\mathrm{VaR}}_{\alpha}^{N} \right ) = W_0 \mathrm{\widehat{se}}(\hat{q}_{\alpha}^R) 
$$ {#eq-se-VaR-normal}
$$
\mathrm{\widehat{se}} \left( \widehat{\mathrm{VaR}}_{\alpha}^{LN} \right ) = W_0 \exp(\hat{q}_{\alpha}^R) \mathrm{\widehat{se}}(\hat{q}_{\alpha}^R)
$$ {#eq-se-VaR-logNormal}


\hfill{}$\blacksquare$
:::

::: {#exm-applying-the-delta-method-microsoft .example name="Applying the delta method to $\\widehat{VaR}_{\\alpha}^{N}$ and $\\widehat{VaR}_{\\alpha}^{LN}$ for Microsoft"}

Consider a $\$100,000$ investment for one month in Microsoft. The estimates of $\mathrm{VaR}_{\alpha}^{N}$, its standard error, and $95\%$ confidence intervals, for $\alpha=0.05$, $\alpha=0.01$, and $\alpha=0.001$ are:

```{r}
VaR.N = -qhat*W0
seVaR.N = W0*seQhat
lowerVaR.N = VaR.N - 2*seVaR.N
upperVaR.N = VaR.N + 2*seVaR.N
width = upperVaR.N - lowerVaR.N
ans = cbind(VaR.N, seVaR.N, lowerVaR.N, upperVaR.N, width)
colnames(ans) = c("Estimate", "Std Error", "2.5%", "97.5%", "Width")
rownames(ans) = c("VaR.N.05", "VaR.N.01", "VaR.N.001")
ans
```

The estimated values of $\mathrm{VaR}_{.05}^N$, $\mathrm{VaR}_{.01}^N$, and $\mathrm{VaR}_{.001}^N$ are $\$15,780$, $\$22,696$, and $\$30,450$ respectively, with estimated standard errors of $\$1,187$, $\$1,490$, and $\$1,860$. The estimated standard errors are fairly small compared with the estimated VaR values, so you can say that VaR is estimated fairly precisely here. Notice that the estimated standard error for $\mathrm{VaR}_{.001}$ is about $56\%$ larger than the estimated standard error for $\mathrm{VaR}_{.05}^N$ indicating you have much less precision for estimating VaR for very small values of $\alpha$.

The estimates of $\mathrm{VaR}_{\alpha}^{LN}$ and its standard error
for $\alpha=0.05$ and $\alpha=0.01$ are:

```{r}
qhat.c = muhatC + sigmahatC*qnorm(c(0.05,0.01,0.001))
seQhat.c = (sigmahatC/sqrt(n.obs))*sqrt(1 + 0.5*qnorm(c(0.05,0.01,0.001))^2)
VaR.LN = -W0*(exp(qhat.c)-1)
seVaR.LN = W0*exp(qhat.c)*seQhat.c
lowerVaR.LN = VaR.LN - 2*seVaR.LN
upperVaR.LN = VaR.LN + 2*seVaR.LN
width = upperVaR.LN - lowerVaR.LN
ans = cbind(VaR.LN, seVaR.LN, lowerVaR.LN, upperVaR.LN, width)
colnames(ans) = c("Estimate", "Std Error", "2.5%", "97.5%", "Width")
rownames(ans) = c("VaR.LN.05", "VaR.LN.01", "VaR.LN.001")
ans
```
The estimates of $\mathrm{VaR}_{\alpha}^{LN}$ (and its standard error) are slightly smaller than the estimates of $\mathrm{VaR}_{\alpha}^{N}$ (and its standard error) due to the positive skewness in the log-normal distribution.
\hfill{}$\blacksquare$
:::


### The numerical delta method 

The delta method is an advanced statistical technique and requires a bit of calculus to implement, especially for nonlinear vector-valued functions. It can be easy to make mistakes when working out the math! Fortunately, in R there is an easy way to implement the delta method that does not require you to work out the math. The **car** function `deltaMethod` implements the delta method for scalar valued functions *numerically*, utilizing the **stats** function `D()` which implements symbolic derivatives. The arguments expected by `deltaMethod` are:

```{r}
args(car:::deltaMethod.default)
```
Here, `object` is the vector of named estimated parameters, $\hat{\boldsymbol{\theta}}$, `g.` is a quoted string that is the function of the parameter estimates to be evaluated, $f(\boldsymbol{\theta})$, and `vcov` is the named estimated covariance matrix of the coefficient estimates, $\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})$. The function returns an object of class "deltaMethod" for which there is a `print` method. The "deltaMethod" object is essentially a `data.frame` with columns giving the parameter estimate(s), the delta method estimated standard error(s), and lower and upper confidence limits. The following example illustrates how to use `deltaMethod` for the example functions. 


::: {#exm-numerical-delta-method-standard-errors-for-example-data .example name="Numerical delta method standard errors for example data"}

I first create the named inputs $\hat{\boldsymbol{\theta}}$ and $\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})$ from the GWN model for simple returns:

```{r}
thetahatS = c(muhatS, sigmahatS)
names(thetahatS) = c("mu", "sigma")
var.thetahatS = matrix(c(se.muhatS^2,0,0,se.sigmahatS^2), 2, 2, byrow=TRUE)
rownames(var.thetahatS) = colnames(var.thetahatS) = names(thetahatS)
thetahatS
var.thetahatS
```

I do the same for the inputs from the GWN model for continuously compounded returns:

```{r}
thetahatC = c(muhatC, sigmahatC)
names(thetahatC) = c("mu", "sigma")
var.thetahatC = matrix(c(se.muhatC^2,0,0,se.sigmahatC^2), 2, 2, byrow=TRUE)
rownames(var.thetahatC) = colnames(var.thetahatC) = names(thetahatC)
```

It is important to name the elements in `thetahat` and `var.thetahat` as these names will be used in defining $f(\boldsymbol{\theta})$. To implement the numerical delta method for the normal quantile, you need to supply a quoted string specifying the function $f_1(\boldsymbol{\theta}) = \mu + \sigma q_{\alpha}^Z$. Here I use the string `"mu+sigma*q.05"`, where `q.05 = qnorm(0.05)`. For, $\alpha = 0.05$ the R code is:
```{r}
q.05 = qnorm(0.05)
dm1 = deltaMethod(thetahatS, "mu+sigma*q.05", var.thetahatS)
class(dm1)
```

The returned object `dm1` is of class "deltaMethod", inhereting from "data.frame", and has named columns:

```{r}
names(dm1)
```
The `print` method shows the results of applying the delta method:

```{r}
dm1
```
The results match the analytic calculations performed above.^[The 95\% confidence interval computed in `deltaMethod` uses 1.96 instead of the rule-of-thumb value of 2 used in the analytic calculations.] 

The delta method for the normal VaR function $f_2(\boldsymbol{\theta})$ is:

```{r}
dm2 = deltaMethod(thetahatS, "-W0*(mu+sigma*q.05)", var.thetahatS)
dm2
```

The delta method for the log-normal VaR function $f_3(\boldsymbol{\theta})$ is:

```{r}
dm3 = deltaMethod(thetahatC, "-W0*(exp(mu+sigma*q.05)-1)", var.thetahatC)
dm3
```

:::

### The delta method for vector valued functions (advanced)

Let $\boldsymbol{\theta}$ denote a $k \times 1$ vector of GWN model parameters. For example, $\boldsymbol{\theta} = (\mu, \sigma^2)'$.  Assume that $\hat{\boldsymbol{\theta}}$ is asymptotically normally distributed:
$$
\hat{\boldsymbol{\theta}} \sim N( \boldsymbol{\theta}, \widehat{\mathrm{var}}(\boldsymbol{\theta})),
$$
for large enough $T$ where $\widehat{\mathrm{var}}(\boldsymbol{\theta})$ is the $k \times k$ variance-covariance matrix of $\hat{\boldsymbol{\theta}}$. Sometimes you need to consider the asymptotic distribution of a vector-valued function of $\boldsymbol{\theta}$. Let $\mathbf{f}:\mathbb{R}^k \rightarrow \mathbb{R}^j$ where $j \le k$ be a continuous and differentiable function, and define the $j \times 1$ vector $\boldsymbol{\eta}=\mathbf{f}(\boldsymbol{\theta})$. It is useful to express $\boldsymbol{\eta}$ as:

$$
\boldsymbol{\eta} = \begin{pmatrix} \eta_1 \\ \eta_2 \\ \vdots \\ \eta_j \end{pmatrix}
= \begin{pmatrix} f_1(\boldsymbol{\theta}) \\ f_2(\boldsymbol{\theta}) \\ \vdots \\ f_j(\boldsymbol{\theta}) \end{pmatrix}.
$$

::: {#exm-example-vector-functions-of-gwn-model-parameters .example name="Example vector functions of GWN model parameters"}

Let $\boldsymbol{\theta}=(\sigma_1^2, \sigma_2^2)'$, and define the $2 \times 1$ vector of volatilities: $\boldsymbol{\eta} = \mathbf{f}(\boldsymbol{\theta})=(\sigma_1, \sigma_2)'$. Then, $\eta_1 = f_1(\boldsymbol{\theta}) = \sqrt{\sigma_2^2}$ and $\eta_2 = f_2(\boldsymbol{\theta}) = \sqrt{\sigma_2^2}$. 

\hfill{}$\blacksquare$
:::

Define the $j \times k$ Jacobian matrix:
$$
\mathbf{G}(\boldsymbol{\theta})' = \frac{\partial \mathbf{f}(\boldsymbol{\theta})}{\partial \boldsymbol{\theta}'} =
\begin{pmatrix}
\frac{\partial f_1(\boldsymbol{\theta})}{\partial \theta_1} &  \frac{\partial f_1(\boldsymbol{\theta})}{\partial \theta_2} & \cdots & \frac{\partial f_1(\boldsymbol{\theta})}{\partial \theta_k} \\
\frac{\partial f_2(\boldsymbol{\theta})}{\partial \theta_1} &  \frac{\partial f_2(\boldsymbol{\theta})}{\partial \theta_2} & \cdots & \frac{\partial f_2(\boldsymbol{\theta})}{\partial \theta_k} \\
\vdots & \vdots & \ldots & \vdots \\
\frac{\partial f_j(\boldsymbol{\theta})}{\partial \theta_1} &  \frac{\partial f_j(\boldsymbol{\theta})}{\partial \theta_2} & \cdots & \frac{\partial f_j(\boldsymbol{\theta})}{\partial \theta_k} \\
\end{pmatrix}
$$ {#eq-jacob}
Then, by the delta method, the asymptotic normal distribution of $\hat{\boldsymbol{\eta}} = \mathbf{f}(\hat{\boldsymbol{\theta}})$ is:
$$
\hat{\eta} \sim N(\eta, \widehat{\mathrm{var}}(\hat{\eta})),
$$
where
$$
\widehat{\mathrm{var}}(\hat{\eta}) = \mathbf{G}(\hat{\boldsymbol{\theta}})'\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})\mathbf{G}(\hat{\boldsymbol{\theta}}).
$$ {#eq-var-etahat}



::: {#exm-asymptotic-distributions-of-example-functions .example name="Asymptotic distributions of example functions"}

Consider the first example function where $\boldsymbol{\theta} = (\sigma_1^2,\sigma_2^2)'$ and $\boldsymbol{\eta} = \mathbf{f}(\boldsymbol{\theta})=(\sigma_1, \sigma_2)'$. Now, using the chain-rule:

\begin{align*}
\frac{\partial f_1(\boldsymbol{\theta})}{\partial \sigma_1^2} & = \frac{1}{2}\sigma_1^{-1}, ~ \frac{\partial f_1(\boldsymbol{\theta})}{\partial \sigma_2^2} = 0 \\
\frac{\partial f_2(\boldsymbol{\theta})}{\partial \sigma_1^2} & = 0, ~ \frac{\partial f_2(\boldsymbol{\theta})}{\partial \sigma_2^2} = \frac{1}{2}\sigma_1^{-1}
\end{align*}

The Jacobian matrix is

$$
\mathbf{G}(\boldsymbol{\theta})' = \begin{pmatrix} \frac{1}{2}\sigma_1^{-1} & 0 \\
0 & \frac{1}{2}\sigma_2^{-1} \end{pmatrix}
$$

From Chapter @sec-Estimation-of-The-GWN-Model, Proposition 7.9, 

$$
\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}}) = \frac{1}{T}\begin{pmatrix} 2\hat{\sigma}_1^4 & 2\hat{\sigma}_{12}^2 \\ 2\hat{\sigma}_{12}^2 & 2\hat{\sigma}_2^4 \end{pmatrix}
$$
Then

\begin{align*}
\widehat{\mathrm{var}}(\hat{\eta}) & = \mathbf{G}(\hat{\boldsymbol{\theta}})'\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})\mathbf{G}(\hat{\boldsymbol{\theta}}) \\
& = \begin{pmatrix} \frac{1}{2}\hat{\sigma}_1^{-1} & 0 \\
0 & \frac{1}{2}\hat{\sigma}_2^{-1} \end{pmatrix} \frac{1}{T}\begin{pmatrix} 2\hat{\sigma}_1^4 & 2\hat{\sigma}_{12}^2 \\ 2\hat{\sigma}_{12}^2 & 2\hat{\sigma}_2^4 \end{pmatrix} \begin{pmatrix} \frac{1}{2}\hat{\sigma}_1^{-1} & 0 \\
0 & \frac{1}{2}\hat{\sigma}_2^{-1} \end{pmatrix} \\
&= \frac{1}{T} \begin{pmatrix} \frac{1}{2}\hat{\sigma}_1^2 & \frac{1}{2}\hat{\sigma}_{12}^2 \hat{\sigma}_{1}^{-1} \hat{\sigma}_{2}^{-1}\\
\frac{1}{2}\hat{\sigma}_{12}^2 \hat{\sigma}_{1}^{-1} \hat{\sigma}_{2}^{-1} & \frac{1}{2}\hat{\sigma}_2^2 \end{pmatrix}
= \frac{1}{T} \begin{pmatrix} \frac{1}{2}\hat{\sigma}_1^2 & \frac{1}{2}\hat{\sigma}_{12} \hat{\rho}_{12} \\
\frac{1}{2}\hat{\sigma}_{12} \hat{\rho}_{12} & \frac{1}{2}\hat{\sigma}_2^2 \end{pmatrix}.
\end{align*}

\hfill{}$\blacksquare$
:::

### The numerical delta method for vector valued functions (advanced)

* TBD
* Use the delta method implemented in **metafor** and **msm**

### The delta method explained

The delta method deduces the asymptotic distribution of $f(\hat{\boldsymbol{\theta}})$ using a first order Taylor series approximation of $f(\hat{\boldsymbol{\theta}})$ evaluated at the true value $\boldsymbol{\theta}$. For simplicity, assume that $\theta$ is a scalar parameter. Then, the first order Taylor series expansion gives

\begin{align*}
f(\hat{\theta}) & = f(\theta) + f'(\theta)(\hat{\theta} - \theta) + \text{remainder} \\
& = f(\theta) + f'(\theta)\hat{\theta} + f'(\theta)\theta + \text{remainder} \\
& \approx  f(\theta) + f'(\theta)\hat{\theta} + f'(\theta)\theta
\end{align*}

The first order Taylor series expansion shows that $f(\hat{\theta})$ is an approximate linear function of the random variable $\hat{\theta}$. The values $\theta$, $f(\theta)$, and $f'(\theta)$ are constants. Assuming that, for large enough T, 

$$
\hat{\theta}\sim N(\theta,\widehat{\mathrm{se}}(\hat{\theta})^{2}),
$$

it follows that

$$
f(\hat{\theta}) \sim N(f(\theta), f'(\theta)^2 \widehat{\mathrm{se}}(\hat{\theta})^{2})
$$

Since $\hat{\theta}$ is consistent for $\theta$ you can replace $f'(\theta)^2$ with $f'(\hat{\theta})^2$ giving the practically useful result

$$
f(\hat{\theta}) \sim N(f(\theta), f'(\hat{\theta})^2 \widehat{\mathrm{se}}(\hat{\theta})^{2})
$$

## Resampling {#sec-Resampling}

As described in the previous sections, the delta method can be used to produce analytic expressions for standard errors of functions of estimated parameters that are asymptotically normally distributed. The formulas are specific to each function
and are approximations that rely on large samples and the validity of the Central Limit Theorem to be accurate. 

Alternatives to the analytic asymptotic distributions of estimators and the delta method are computer intensive *resampling methods* such as the *jackknife* and the *bootstrap*. These resampling methods can be used to produce numerical estimated standard errors for estimators and functions of estimators without the use of mathematical formulas.
Most importantly, these jackknife and bootstrap standard errors are: (1) easy to compute, and (2) often numerically very close to the analytic standard errors based on CLT approximations. In some cases the bootstrap standard errors can even be more reliable than asymptotic approximations. 

There are many advantages of the jackknife and bootstrap over analytic formulas. The
most important are the following:

-  *Fewer assumptions*. The jackknife and boostrap procedures typically requires fewer assumptions than are needed to derive analytic formulas. In particular, they do not require the data to be normally distributed or the sample size to be large enough so that the Central Limit Theorem holds.
-  *Greater accuracy*. In small samples the bootstrap is often more reliable than analytic approximations based on the Central Limit Theorem. In large samples when the Central Limit Theorem holds the bootstrap can be even be more accurate.
-  *Generality*. In many situations, the jackknife and bootstrap procedures are applied in the same way regardless of the estimator under consideration. That is, you don't need a different jackknife or bootstrap procedure for each estimator.

## The Jackknife {#sec-The-Jackknife}

Let $\theta$ denote a scalar parameter of interest. It could be a scalar parameter of the GWN model (e.g. $\mu$ or $\sigma$) or it could be a scalar function of the parameters of the GWN model (e.g., VaR). Let $\hat{\theta}$ denote the plug-in estimator of $\theta$. The goal is to compute a numerical estimated standard error for $\hat{\theta}$, $\widehat{\mathrm{se}}(\hat{\theta})$, using the jackknife resampling method. 

The jackknife resampling method utilizes $T$ resamples of the original data $\left\{ R_t \right\}_{t=1}^{T}$. The resamples are created by leaving out a single observation. For example, the first jackknife resample leaves out the first observation and is given by $\left\{ R_t \right\}_{t=2}^{T}$. This sample has $T-1$ observations. In general, the $t^{th}$ jackknife resample leaves out the $t^{th}$ return $R_{t}$ for $t \le T$. Now, on each of the $t=1,\ldots, T$ jackknife resamples you compute an estimate of $\theta$. Denote this jackknife estimate $\hat{\theta}_{-t}$ to indicate that $\theta$ is estimated using the resample that omits observation $t$. The estimator $\hat{\theta}_{-t}$ is called the *leave-one-out* \index{Leave-one-out estimator} estimator of $\theta$. Computing $\hat{\theta}_{-t}$ for $t=1,\ldots, T$ gives $T$ leave-one-out estimators $\left\{ \hat{\theta}_{-t} \right\}_{t=1}^{T}$. 

The *jackknife estimator* of $\widehat{\mathrm{se}}(\hat{\theta})$ is the scaled standard deviation of $\left\{ \hat{\theta}_{-t} \right\}_{t=1}^{T}$:

$$
\widehat{\mathrm{se}}_{jack}(\hat{\theta}) =  \sqrt{  \frac{T-1}{T}\sum_{t=1}^T (\hat{\theta}_{-t} - \bar{\theta})^2 }, 
$$ {#eq-se-jackknife}

where $\bar{\theta} = T^{-1}\sum_{t=1}^T \hat{\theta}_{-t}$ is the sample mean of $\left\{ \hat{\theta}_{-t} \right\}_{t=1}^{T}$. The estimator @eq-se-jackknife is easy to compute and does not require any analytic calculations from the asymptotic distribution of $\hat{\theta}$.

The intuition for the jackknife standard error formula can be illustrated by considering the jackknife standard error estimate for the sample mean $\hat{\theta} = \hat{\mu}$. You know that the analytic estimated squared standard error for $\hat{\mu}$ is 

$$
\widehat{\mathrm{se}}(\hat{\mu})^2 = \frac{\hat{\sigma}^2}{T} = \frac{1}{T} \frac{1}{T-1} \sum_{t=1}^T (R_t - \hat{\mu})^2. 
$$

Now consider the jackknife squared standard error for $\hat{\mu}$:

$$
\widehat{\mathrm{se}}_{jack}(\hat{\mu})^2= \frac{T-1}{T}\sum_{t=1}^T (\hat{\mu}_{-t} - \bar{\mu})^2, 
$$

where $\bar{\mu} = T^{-1}\sum_{t=1}^T \hat{\mu}_{-t}$. Here, I will show that $\widehat{\mathrm{se}}(\hat{\mu})^2 = \widehat{\mathrm{se}}_{jack}(\hat{\mu})^2$. 

First, note that 

\begin{align*}
\hat{\mu}_{-t} & = \frac{1}{T-1}\sum_{j \ne t} R_j = \frac{1}{T-1} \left( \sum_{j=1}^T R_j - R_t \right) \\
&= \frac{1}{T-1} \left( T \hat{\mu} - R_t \right) \\
&= \frac{T}{T-1}\hat{\mu} - \frac{1}{T-1}R_t.
\end{align*}

Next, plugging in the above to the definition of $\bar{\mu}$ I get 

\begin{align*}
\bar{\mu} &= \frac{1}{T} \sum_{j=1}^T \hat{\mu}_{-t} = \frac{1}{T}\sum_{t=1}^T \left( 
\frac{T}{T-1}\hat{\mu} - \frac{1}{T-1}R_t \right) \\
&= \frac{1}{T-1} \sum_{t=1}^T \hat{\mu} - \frac{1}{T}\frac{1}{T-1} \sum_{t=1}^T R_t \\
&= \frac{T}{T-1} \hat{\mu} - \frac{1}{T-1} \hat{\mu} \\
&= \hat{\mu} \left( \frac{T}{T-1} - \frac{1}{T-1} \right) \\
&= \hat{\mu}.
\end{align*}

Hence, the sample mean of $\hat{\mu}_{-t}$ is $\hat{\mu}$. Then I can write

\begin{align*}
\hat{\mu}_{-t} - \bar{\mu} & = \hat{\mu}_{-t} - \hat{\mu} \\
& = \frac{T}{T-1}\hat{\mu} - \frac{1}{T-1}R_t - \hat{\mu} \\
&= \frac{T}{T-1}\hat{\mu} - \frac{1}{T-1}R_t - \frac{T-1}{T-1}\hat{\mu} \\
&= \frac{1}{T-1}\hat{\mu} - \frac{1}{T-1}R_t \\
&= \frac{1}{T-1}(\hat{\mu} - R_t).
\end{align*}

Using the above I can write

\begin{align*}
\widehat{\mathrm{se}}_{jack}(\hat{\mu})^2 &= \frac{T-1}{T}\sum_{t=1}^T (\hat{\mu}_{-t} - \bar{\mu})^2 \\
&= \frac{T-1}{T}\left(\frac{1}{T-1}\right)^2 \sum_{t=1}^T (\hat{\mu} - R_t)^2 \\
&= \frac{1}{T}\frac{1}{T-1} \sum_{t=1}^T (\hat{\mu} - R_t)^2 \\
&= \frac{1}{T} \hat{\sigma}^2 \\
&= \widehat{\mathrm{se}}(\hat{\mu})^2.
\end{align*}

::: {#exm-jackknife-standard-error-for-hat-mu-for-microsoft .example name="Jackknife standard error for $\\hat{\mu}$ for Microsoft"}

You can compute the leave-one-out estimates $\{\hat{\mu}_{-t}\}_{t=1}^T$ using a simple `for` loop as follows:

```{r}
n.obs = length(msftRetS)
muhat.loo = rep(0, n.obs)
for(i in 1:n.obs) {
  muhat.loo[i] = mean(msftRetS[-i])
}
```

In the above loop, `msftRetS[-i]` extracts all observations except observation $i$. The mean of the leave-one-out estimates is

```{r}
#| echo: true
mean(muhat.loo)
```
which is the sample mean of `msftRetS`. It is informative to plot a histogram of the leave-one-out estimates:

```{r}
#| label: fig-BootLooHist
#| echo: false
#| fig-cap: Histogram of leave-one-out estimator of $\mu$ for Microsoft.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
hist(muhat.loo, col="cornflowerblue", main="")
abline(v=muhatS, col="white", lty="dotted", lwd=2)
```
The white dotted line is at the point $\hat{\mu} = 0.00915$. The jackknife estimated standard error is the scaled standard deviation of the $T$ $\hat{\mu}_{-t}$ values:^[the scaling of the standard deviation of $\hat{\mu}_{-t}$ is approximately $\sqrt{T}$.]

```{r}
se.jack.muhat = sqrt(((n.obs - 1)^2)/n.obs)*sd(muhat.loo)
se.jack.muhat
```

This value is exactly the same as the analytic standard error:

```{r}
se.muhatS
```

You can also compute jackknife estimated standard errors using the **bootstrap** function `jackknife`:

```{r}
library(bootstrap)
muhat.jack = jackknife(coredata(msftRetS), mean)
names(muhat.jack)
```
By default, `jackknife()` takes as input a vector of data and the name of an R function and returns a list with components related to the jackknife procedure. The component `jack.values` contains the leave-one-out estimators, and the component `jack.se` gives you the jackknife estimated standard error:

```{r}
muhat.jack$jack.se
```

\hfill{}$\blacksquare$
:::

The fact that the jackknife standard error estimate for $\hat{\mu}$ equals the analytic standard error is motivates using the jackknife to compute standard errors for general estimators. For estimated parameters other than $\hat{\mu}$, and general estimated functions of GWN model parameters, the jackknife standard errors are often numerically very close to the delta method standard errors. The following example illustrates this point for the example functions @eq-f1 - @eq-f3.

::: {#exm-jackknife-standard-errors-for-plug-in-estimators-of-example-functions .example name="Jackknife standard errors for plug-in estimators of example functions"}

Here, I use the **bootstrap** function `jackknife()` to compute jackknife estimated standard errors for the plug-in estimates of the example functions @eq-f1 - @eq-f3 and I compare the jackknife standard errors with the delta method standard errors. To use `jackknife()` with user-defined functions involving multiple parameters you must create the function such that the first input is an integer vector of index observations and the second input is the data. For example, a function for computing @eq-f1 to be passed to `jackknife()` is:

```{r}
theta.f1 = function(x, xdata) {
  mu.hat = mean(xdata[x, ])
  sigma.hat = sd(xdata[x, ])
  fun.hat = mu.hat + sigma.hat*qnorm(0.05)
  fun.hat
}
```

To compute the jackknife estimated standard errors for the estimate of @eq-f1 use:

```{r}
f1.jack = jackknife(1:n.obs, theta=theta.f1, coredata(msftRetS))
ans = cbind(f1.jack$jack.se, dm1$SE)
colnames(ans) = c("Jackknife", "Delta Method")
rownames(ans) = "f1"
ans
```
Here, you see that the estimated jackknife standard error for $f_1(\hat{\boldsymbol{\theta}})$ is very close to the delta method standard error computed earlier.

To compute the jackknife estimated standard error for $f_2(\hat{\boldsymbol{\theta}})$ use:

```{r}
theta.f2 = function(x, xdata, W0) {
  mu.hat = mean(xdata[x, ])
  sigma.hat = sd(xdata[x, ])
  fun.hat = -W0*(mu.hat + sigma.hat*qnorm(0.05))
  fun.hat
}
f2.jack = jackknife(1:n.obs, theta=theta.f2, coredata(msftRetS), W0)
ans = cbind(f2.jack$jack.se, dm2$SE)
colnames(ans) = c("Jackknife", "Delta Method")
rownames(ans) = "f2"
ans
```
Again, the jackknife standard error is close to the delta method standard error.

To compute the jackknife estimated standard error for $f_3(\hat{\boldsymbol{\theta}})$ use:

```{r}
theta.f3 = function(x, xdata, W0) {
  mu.hat = mean(xdata[x, ])
  sigma.hat = sd(xdata[x, ])
  fun.hat = -W0*(exp(mu.hat + sigma.hat*qnorm(0.05)) - 1)
  fun.hat
}
f3.jack = jackknife(1:n.obs, theta=theta.f3, coredata(msftRetC), W0)
ans = cbind(f3.jack$jack.se, dm3$SE)
colnames(ans) = c("Jackknife", "Delta Method")
rownames(ans) = "f3"
ans
```
For log-normal VaR, the jackknife standard error is a bit larger than the delta method standard error.

Here, the jackknife and delta method standard errors are very close.
\hfill{}$\blacksquare$
:::

### The jackknife for vector-valued estimates

The jackknife can also be used to compute an estimated variance matrix of a vector-valued estimate of parameters. Let $\boldsymbol{\theta}$ denote a $k \times 1$ vector of 
GWN model parameters (e.g., $\boldsymbol{\theta}=(\mu,\sigma)'$), and let $\hat{\boldsymbol{\theta}}$ denote the plug-in estimate of $\boldsymbol{\theta}$. Let $\hat{\boldsymbol{\theta}}_{-t}$ denote the leave-one-out estimate of $\boldsymbol{\theta}$, and $\bar{\boldsymbol{\theta}}=T^{-1}\sum_{t=1}^T \hat{\boldsymbol{\theta}}_{-t}$. Then the jackknife estimate of the $k \times k$ variance matrix  $\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})$ is the scaled sample covariance of $\{\hat{\boldsymbol{\theta}}_{-t}\}_{t=1}^T$:

$$
\widehat{\mathrm{var}}_{jack}(\hat{\boldsymbol{\theta}}) = \frac{T-1}{T} \sum_{t=1}^T (\hat{\boldsymbol{\theta}}_{-t} - \bar{\boldsymbol{\theta}})(\hat{\boldsymbol{\theta}}_{-t} - \bar{\boldsymbol{\theta}})'. 
$$ {#eq-jack-varmatrix}

The jackknife estimated standard errors for the elements of $\hat{\boldsymbol{\theta}}$ are given by the square root of the diagonal elements of $\widehat{\mathrm{var}}_{jack}(\hat{\boldsymbol{\theta}})$.

The same principle works for computing the estimated variance matrix of a vector of functions of GWN model parameters. Let $\mathbf{f}:R^k \rightarrow R^l$ with $l \le k$ and define $\boldsymbol{\eta} = \mathbf{f}(\boldsymbol{\theta})$ denote a $l \times 1$ vector of functions of GWN model parameters. Let $\hat{\eta}=f(\hat{\boldsymbol{\theta}})$ denote the plug-in estimator of $\boldsymbol{\eta}$. Let $\hat{\boldsymbol{\eta}}_{-t}$ denote the leave-one-out estimate of $\boldsymbol{\eta}$, and $\bar{\boldsymbol{\eta}}=T^{-1}\sum_{t=1}^T \hat{\boldsymbol{\eta}}_{-t}$. Then the jackknife estimate of the $l \times l$ variance matrix  $\widehat{\mathrm{var}}(\hat{\boldsymbol{\eta}})$ is the scaled sample covariance of $\{\hat{\boldsymbol{\eta}}_{-t}\}_{t=1}^T$:

$$
\widehat{\mathrm{var}}_{jack}(\hat{\boldsymbol{\eta}}) = \frac{T-1}{T} \sum_{t=1}^T (\hat{\boldsymbol{\eta}}_{-t} - \bar{\boldsymbol{\eta}})(\hat{\boldsymbol{\eta}}_{-t} - \bar{\boldsymbol{\eta}})'. 
$$ {#eq-jack-varmatrix-eta}

::: {#exm-jackknife-variance-estimate-for-hat-boldsymbol-theta-hat-mu-hat-sigma-for-microsoft-simple-returns .example name="Jackknife variance estimate for $\\hat{\\boldsymbol{\\theta}}=(\\hat{\\mu}, \\hat{\\sigma})'$ for Microsoft simple returns"}

The **bootstrap** function `jackknife()` function only works for scalar-valued parameter estimates, so you have to compute the jackknife variance estimate by hand using a `for` loop. The R code is:

```{r}
n.obs = length(msftRetS)
muhat.loo = rep(0, n.obs)
sigmahat.loo = rep(0, n.obs)
for(i in 1:n.obs) {
  muhat.loo[i] = mean(msftRetS[-i])
  sigmahat.loo[i] = sd(msftRetS[-i])
}
thetahat.loo = cbind(muhat.loo, sigmahat.loo)
colnames(thetahat.loo) = c("mu", "sigma")
varhat.jack = (((n.obs-1)^2)/n.obs)*var(thetahat.loo)
varhat.jack
```

The jackknife variance estimate is somewhat close to the analytic estimate:

```{r}
var.thetahatS
```
Since the first element of $\hat{\boldsymbol{\theta}}$ is $\hat{\mu}$, the (1,1) elements of the two variance matrix estimates match exactly. However, the `varhat.jack` has non-zero off diagonal elements and the (2,2) element corresponding to $\hat{\sigma}$ is considerably bigger than (2,2) element of the analytic estimate. Converting `var.thetahatS` to an estimated correlation matrix gives:

```{r}
cov2cor(varhat.jack)
```
Here, you see that the jackknife shows a small positive correlation between $\hat{\mu}$ and $\hat{\sigma}$ whereas the analytic matrix shows a zero correlation.
\hfill{}$\blacksquare$
:::

### Pros and Cons of the Jackknife

The jackknife gets its name because it is a useful statistical device.^[A jackknife is also the name of a very useful folding knife.] The main advantages (pros) of the jacknife are:

* It is a good "quick and dirty" method for computing numerical estimated standard errors of estimates that does not rely on asymptotic approximations.

* The jackknife estimated standard errors are often close to the delta method standard errors.

This jackknife, however, has some limitations (cons). The main limitations are:

* It can be computationally burdensome for very large samples (e.g., if used with ultra high frequency data).

* It does not work well for non-smooth functions such as empirical quantiles. 

* It is not well suited for computing confidence intervals.

* It is not as accurate as the bootstrap.

## The Nonparametric Bootstrap {#sec-The-Nonparametric-Bootstrap}
\index{Nonparametric Bootstrap}
In this section, I describe the easiest and most common form of the bootstrap: the *nonparametric bootstrap*. As you shall see, the
nonparametric bootstrap procedure is very similar to a Monte Carlo simulation experiment. The main difference is how the random data
is generated. In a Monte Carlo experiment, the random data are computer simulated observations from an assumed model (e.g., the GWN model). In the nonparametric bootstrap, the random data is created by resampling with replacement from the original data.

The procedure for the nonparametric bootstrap of an estimate $\hat{\theta}$ of a scalar parameter $\theta$ is as follows:

1.  *Resample*. Create $B$ bootstrap samples by sampling with replacement from the original data $\{r_{1},\ldots,r_{T}\}$.^[You can also bootstrap data from multiple assets by defining $r_t$ as a vector of asset returns.] Each bootstrap sample has $T$ observations (same as the original sample) 
\begin{eqnarray*}
\left\{r_{11}^{*},r_{12}^{*},\ldots,r_{1T}^{*}\right\}  & = & \mathrm{1st\,bootstrap\,sample}\\
 & \vdots\\
\left\{r_{B1}^{*},r_{B2}^{*},\ldots,r_{BT}^{*}\right\}  & = & \mathrm{Bth\,boostrap\,sample}
\end{eqnarray*}
2.  *Estimate* $\theta$. From each bootstrap sample estimate $\theta$ and denote the resulting estimate $\hat{\theta}^{*}$. There will be $B$ values of $\hat{\theta}^{*}$: $\left\{\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}\right\}$.
3.  *Compute statistics*. Using $\left\{\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}\right\}$ compute estimate of bias, standard error, and approximate 95\% confidence interval. 

This procedure is very similar to the procedure used to perform a
Monte Carlo experiment described in Chapter @sec-Estimation-of-The-GWN-Model.
The main difference is how the hypothetical samples are created. With
the nonparametric bootstrap, the hypothetical samples are created
by resampling with replacement from the original data. In this regard
the bootstrap treats the sample as if it were the population. This
ensures that the bootstrap samples inherit the same distribution as
the original data -- *whatever that distribution may be*. If the
original data is normally distributed, then the nonparametric bootstrap
samples will also be normally distributed. If the data is Student's
t distributed, then the nonparametric bootstrap samples will be Student's
t distributed. With Monte Carlo simulation, the hypothetical samples
are simulated under an assumed model and distribution. This requires
one to specify values for the model parameters and the distribution
from which to simulate.

### Bootstrap bias estimate
\index{Bootstrap bias}
The nonparametric bootstrap can be used to estimate the bias of an
estimator $\hat{\theta}$ using the bootstrap estimates {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$}.
The key idea is to treat the empirical distribution (i.e., histogram)
of the bootstrap estimates {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$}
as an approximate to the unknown distribution of $\hat{\theta}.$


Recall, the bias of an estimator is defined as

$$
E[\hat{\theta}]-\theta.
$$ 

The bootstrap estimate of bias is given by

$$ 
\begin{aligned}
\widehat{\mathrm{bias}}_{boot}(\hat{\theta},\theta)= \bar{\theta}^{\ast} - \hat{\theta} = \frac{1}{B}\sum_{j=1}^{\ B}\hat{\theta}_{j}^{\ast}-\hat{\theta.}\\
\text{(bootstrap mean - estimate)}
\end{aligned}
$$ {#eq-BootBootstrapBias}

The bootstrap bias estimate @eq-BootBootstrapBias is the difference between the mean of the bootstrap estimates of $\theta$ and the sample
estimate of $\theta$. This is similar to the Monte Carlo estimate of bias discussed in Chapter @sec-Estimation-of-The-GWN-Model.
However, the Monte Carlo estimate of bias is the difference between the mean of the Monte Carlo estimates of $\theta$ and the true value
of $\theta.$ The bootstrap estimate of bias does not require knowing the true value of $\theta$. Effectively, the bootstrap treats the
sample estimate $\hat{\theta}$ as the population value $\theta$ and the bootstrap mean $\bar{\theta}^{\ast} = \frac{1}{B}\sum_{j=1}^{\ B}\hat{\theta}_{j}^{\ast}$ as an approximation to $E[\hat{\theta}]$. Here, $\widehat{\mathrm{bias}}_{boot}(\hat{\theta},\theta)=0$
if the center of the histogram of {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$} is at $\hat{\theta}.$

Given the bootstrap estimate of bias @eq-BootBootstrapBias, you can compute a *bias-adjusted* estimate:

$$
\hat{\theta}_{adj} = \hat{\theta} - (\bar{\theta}^{\ast} - \hat{\theta}) = 2\hat{\theta} - \bar{\theta}^{\ast} 
$$ {#eq-BiasAdjustedEstimate}

if the bootstrap estimate of bias is large, it may be tempting to use the bias-adjusted estimate @eq-BiasAdjustedEstimate in place of the original estimate. This is generally not done in practice because the bias adjustment introduces extra variability into the estimator. So what use is the bootstrap bias estimate? It provides information to you that your estimate contains bias (or not) and this information can influence your decision making based on the estimate.

### Bootstrap standard error estimate
\index{Bootstrap SE}
for a scalar estimate $\hat{\theta}$, the bootstrap estimate of $\widehat{\mathrm{se}}(\hat{\theta}$) is
given by the sample standard deviation of the bootstrap estimates
{$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$}:

$$
\widehat{\mathrm{se}}_{boot}(\hat{\theta})=\sqrt{\frac{1}{B-1}\sum_{j=1}^{B}\left(\hat{\theta}_{j}^{\ast} - \bar{\theta}^{\ast} \right)^{2}}.
$$ {#eq-BootBootstrapSE}

where $\bar{\theta}^{\ast} = \frac{1}{B}\sum_{j=1}^{\ B}\hat{\theta}_{j}^{\ast}$. 
Here, $\widehat{\mathrm{se}}_{boot}(\hat{\theta})$ is the size of
a typical deviation of a bootstrap estimate, $\hat{\theta}^{*}$, from
the mean of the bootstrap estimates (typical deviation from the middle
of the histogram of the bootstrap estimates). This is very closely
related to the Monte Carlo estimate of $\widehat{\mathrm{se}}(\hat{\theta}$), which is the sample standard deviation of the estimates of $\theta$ from the Monte Carlo samples. 

For a $k \times 1$ vector estimate $\hat{\theta}$, the bootstrap estimate of $\widehat{\mathrm{var}}(\hat{\theta})$ is

$$
\widehat{\mathrm{var}}(\hat{\theta}) = \frac{1}{B-1}\sum_{j=1}^B (\hat{\theta}_{j}^{\ast} - \bar{\theta}^{\ast})(\hat{\theta}_{j}^{\ast} - \bar{\theta}^{\ast})'.
$$ {#eq-BootBootstrapVar}


### Bootstrap confidence intervals
\index{Bootstrap CI}

In addition to providing standard error estimates, the bootstrap is commonly used to compute confidence intervals for scalar parameters. There are several ways of computing confidence intervals with the bootstrap and which confidence interval to use in practice depends on the characteristics of the distribution of the bootstrap estimates {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$}. 

If the bootstrap distribution appears to be normal then you can compute a 95\% confidence interval for $\theta$ using two simple methods. The first method mimics the rule-of-thumb that is justified by the CLT to compute a 95\% confidence interval but uses the bootstrap standard error estimate @eq-BootBootstrapSE:

$$
\hat{\theta}\pm2\times\widehat{\mathrm{se}}_{boot}(\hat{\theta}) = [\hat{\theta} - 2\times \widehat{\mathrm{se}}_{boot}(\hat{\theta}), ~ \hat{\theta} + 2\times \widehat{\mathrm{se}}_{boot}(\hat{\theta})].
$$ {#eq-BootNormalBootCI}

The 95\% confidence interval @eq-BootNormalBootCI is called the *normal approximation* 95\% confidence interval. 

The second method directly uses the distribution of the bootstrap estimates {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$} to form a 95\% confidence interval:

$$
[\hat{q}_{.025}^{*},\,\hat{q}_{.975}^{*}],
$$ {#eq-BootPercentileBootCI}

where $\hat{q}_{.025}^{*}$ and $\hat{q}_{.975}^{*}$ are the 2.5\%
and 97.5\% empirical quantiles of {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$},
respectively. By construction, 95\% of the bootstrap estimates lie
in the quantile-based interval @eq-BootPercentileBootCI. This 95\% confidence interval is called the *bootstrap percentile* 95\% confidence interval. 

The main advantage of @eq-BootPercentileBootCI over @eq-BootNormalBootCI is *transformation invariance*. This means that if @eq-BootPercentileBootCI is a 95\% confidence interval for $\theta$ and $f(\theta)$ is a monotonic function of $\theta$ then $[f(\hat{q}_{.025}^{*}),\,f(\hat{q}_{.975}^{*})]$ is a 95\% confidence interval for $f(\theta)$. The interval @eq-BootPercentileBootCI can also be asymmetric whereas @eq-BootNormalBootCI is symmetric.

The bootstrap normal approximation and percentile 95\% confidence intervals @eq-BootNormalBootCI and @eq-BootPercentileBootCI will perform well (i.e., will have approximately correct coverage probability) if the bootstrap distribution looks normally distributed. However, if the bootstrap distribution is asymmetric then these 95\% confidence intervals may have coverage probability different from 95\%, especially if the asymmetry is substantial. The basic shape of the bootstrap distribution can be visualized by the histogram of {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$}, and the normality of the distribution can be visually evaluated using the normal QQ-plot of {$\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}$}. If the histogram is asymmetric and/or not bell-shaped or if the normal QQ-plot is not linear then normality is suspect and the confidence intervals @eq-BootPercentileBootCI and @eq-BootNormalBootCI may be inaccurate. 

If the bootstrap distribution is asymmetric then it is recommended to use Efron's *bias and skewness adjusted bootstrap percentile confidence* (BCA) interval. The details of how to compute the BCA interval are a bit complicated, and so are omitted, and the interested reader is referred to Chapter 10 of Hansen (2011) for a good explanation.  Suffice it to say that the 95\% BCA method adjusts the 2.5\% and 97.5\% quantiles of the bootstrap distribution for bias and skewness. The BCA confidence interval can be computed using the **boot** function `boot.ci()` with optional argument `type="bca"`.^[The bootstrap percentile-t confidence interval is another bias and skewness adjusted confidence interval.] 

### The bootstrap for vector-valued estimates

* TBD

### Performing the nonparametric bootstrap in R

The nonparametric bootstrap procedure is easy to perform in R. You can implement the procedure by "brute force" in very much the same way as you perform a Monte Carlo experiment. In this approach
you program all parts of the bootstrapping procedure. Alternatively, you can use the R package **boot** which contains functions for automating certain parts of the bootstrapping procedure. 

#### Brute force implementation

In the brute force implementation you program all parts of the bootstrap
procedure. This typically involves three steps:

::: {callout-note}
## Brute force nonparametric bootstrap procedure
1.  Sample with replacement from the original data using the R function `sample()` to create $\{r_{1}^{*},r_{2}^{*},\ldots,r_{T}^{*}\}$. Do this $B$ times.
2.  Compute $B$ values of the statistic of interest $\hat{\theta}$ from each bootstrap sample giving $\{\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}\}$. $B=1,000$ for routine applications of the bootstrap. To minimize simulation noise it may be required to have $B=10,000$ or higher. 
3.  Compute the bootstrap bias and SE values from $\{\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}\}$ using the R functions `mean()` and `sd()`, respectively.
4.  Compute the histogram and normal QQ-plot of $\{\hat{\theta}_{1}^{*},\ldots,\hat{\theta}_{B}^{*}\}$ using the R functions `hist()` and `qqnorm()`, respectively, to see if the bootstrap distribution looks normal.

:::

Typically steps 1 and 2 are performed inside a `for` loop or within
the R function `apply()`. 

In step 1 sampling with replacement for the original data is performed
with the R function `sample()`. To illustrate, consider sampling
with replacement from the $5\times1$ return vector: 

```{r}
#| echo: true

r = c(0.1,0.05,-0.02,0.03,-0.04)
r
```

Recall, in R you can extract elements from a vector by subsetting based on their location, or index, in the vector. Since the vector `r`
has five elements, its index is the integer vector `1:5`. Then to extract the first, second and fifth elements of `r` use the index vector `c(1, 2, 5)`: 

```{r}
#| echo: true

r[c(1, 2, 5)]
```

Using this idea, you can extract a random sample (of any given size)
with replacement from `r` by creating a random sample with
replacement of the integers $\{1,2,\ldots,5\}$ and using this set
of integers to extract the sample from `r`. The R fucntion
`sample()` can be used to do this process. When you pass a
positive integer value `n` to `sample()`, with the optional
argument `replace=TRUE`, it returns a random sample from the
set of integers from $1$ to `n`. For example, to create a
random sample with replacement of size 5 from the integers$\{1,2,\ldots,5\}$
use:^[I use `set.seed(123)` to initialize the random number generator in R and allow for reproducible results.]

```{r}
#| echo: true

set.seed(123)
idx = sample(5, replace=TRUE)
idx
```

You can then get a random sample with replacement from the vector `r`
by subsetting using the index vector `idx`:

```{r}
#| echo: true

r[idx]
```

This two step process is automated when you pass a vector of observations
to `sample()`:

```{r}
#| echo: true

set.seed(123)
sample(r, replace=TRUE)
```

::: {#exm-brute-force-nonparametric-bootstrap-of-gwn-model-in-r .example name="Brute force nonparametric bootstrap of GWN model in R"}

Consider using the nonparametric bootstrap to compute estimates of
the bias and standard error for $\hat{\mu}$ in the GWN model for
Microsoft. The R code for the brute force "for loop" to implement
the nonparametric bootstrap is:

```{r}
#| echo: true

B = 1000 
muhat.boot = rep(0, B) 
n.obs = nrow(msftRetS) 
set.seed(123) 
for (i in 1:B) {   
   boot.data = sample(msftRetS, n.obs, replace=TRUE)   
   muhat.boot[i] = mean(boot.data) 
}
```

The bootstrap bias estimate is:

```{r}
#| echo: true

mean(muhat.boot) - muhatS
```

which is very close to zero and confirms that $\hat{\mu}$ is unbiased.
The bootstrap standard error estimate is

```{r}
#| echo: true

sd(muhat.boot)
```

and is equal (to three decimals) to the analytic standard error estimate
computed earlier. This confirms that the nonparametric bootstrap accurately
computes $\widehat{\mathrm{se}}(\hat{\mu}).$ Figure @fig-BootBruteForce
shows the histogram and normal QQ-plot of the bootstrap estimates
created with: 

```{r}
#| label: fig-BootBruteForce
#| echo: true
#| fig-cap: Histogram and normal QQ-plot of bootstrap estimates of $\mu$ for Microsoft.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(1,2))   
hist(muhat.boot, col="cornflowerblue")   
abline(v=muhatS, col="white", lwd=2)   
qqnorm(muhat.boot, col="cornflowerblue")   
qqline(muhat.boot) 
```

The distribution of the bootstrap estimates looks normal. As a result,
the bootstrap 95\% confidence interval for $\mu$ has the form:

```{r}
#| echo: true

se.boot = sd(muhat.boot)
lower = muhatS + se.boot
upper = muhatS - se.boot
ans = cbind(muhatS, se.boot, lower, upper)
colnames(ans)=c("Estimate", "Std Error", "2.5%", "97.5%")
rownames(ans) = "mu"
ans
```

\hfill{}$\blacksquare$
:::

It is important to keep in mind that the bootstrap resampling algorithm draws *random samples* from the observed data. In the example above, the random number generator in R was initially set with `set.seed(123)`. If the seed is set to a different integer, say 345, then the $B=1,000$ bootstrap samples will be different and you will get different numerical results for the bootstrap bias and estimated standard error. How different these results will be depends on number of bootstrap samples $B$ and the statistic $\hat{\theta}$ being bootstrapped. Generally, the smaller is $B$ the more variable the bootstraps results will be. It is good practice to repeat the bootstrap analysis with several different random number seeds to see how variable the results are. If the results vary considerably for different random number seeds then it is advised to increase the number of bootstrap simulations $B$. It is common to use $B=1,000$ for routine calculations, and $B=10,000$ for final results.

#### R package boot

The R package **boot** implements a variety of bootstrapping
techniques including the basic non-parametric bootstrap described
above. The **boot** package was written to accompany the textbook *Bootstrap Methods and Their Application* by [@DH-1997].
The two main functions in **boot** are `boot()` and
`boot.ci()`, respectively. The `boot()` function implements
the bootstrap for a statistic computed from a user-supplied function.
The `boot.ci()` function computes bootstrap confidence intervals
given the output from the `boot()` function. 

The arguments to `boot()` are:

```{r}
#| echo: true

library(boot)
args(boot)
```

where `data` is a data object (typically a vector or matrix
but can be a time series object too), `statistic` is a use-specified
function to compute the statistic of interest, and `R` is the
number of bootstrap replications. The remaining arguments are not
important for the basic non-parametric bootstrap. The function assigned
to the argument `statistic` has to be written in a specific
form, which is illustrated in the next example. The `boot()`
function returns an object of class `boot` for which
there are print and plot methods. 

The arguments to `boot.ci()` are

```{r}
#| echo: true

args(boot.ci)
```

where `boot.out` is an object of class `boot`,
`conf` specifies the confidence level, and `type` is
a subset from `c("norm", "basic", "stud", "perc", "bca")` indicating the type of confidence interval to compute.
The choices `norm`, `perc`, and "bca" compute the
normal confidence interval @eq-BootNormalBootCI, the percentile
confidence interval @eq-BootPercentileBootCI, and the BCA confidence interval, respectively. The remaining arguments are not important for the computation of these
bootstrap confidence intervals.

::: {#exm-nonparametric-bootstrap-of-gwn-model-using-the-r-package-boot .example name="Nonparametric bootstrap of GWN model using the R package **boot**"}

To use the `boot()` function to implement the bootstrap for
$\hat{\mu},$ a function must be specified to compute $\hat{\mu}$
for each bootstrap sample. The function must have two arguments: `x`
and `idx`. Here, `x` represents the original data and
`idx` represents the random integer index (created internally
by `boot()`) to subset `x` for each bootstrap sample.
For example, a function to be passed to `boot()` for $\hat{\mu}$ is


```{r}
#| echo: true

mean.boot = function(x, idx) { 
# arguments: 
# x      data to be resampled 
# idx    vector of scrambled indices created by boot() function 
# value: 
# ans	mean value computed using resampled data      
	ans = mean(x[idx])      
	ans 
}
```

To implement the nonparametric bootstrap for $\hat{\mu}$ with 999
samples use 

```{r}
#| echo: true

set.seed(123)
muhat.boot = boot(msftRetS, statistic = mean.boot, R=999) 
class(muhat.boot) 
names(muhat.boot)
```

The returned object `muhat.boot` is of class `boot`.
The component `t0` is the sample estimate $\hat{\mu}$, and
the component t is a $999\times1$ matrix containing the bootstrap
estimates $\{\hat{\mu}_{1}^{*},\ldots,\hat{\mu}_{999}^{*}\}$. The
print method shows the sample estimate, the bootstrap bias and the
bootstrap standard error:

```{r}
#| echo: true

muhat.boot
```

These statistics can be computed directly from the components of `muhat.boot`

```{r}
#| echo: true

c(muhat.boot$t0,mean(muhat.boot$t) - muhat.boot$t0, sd(muhat.boot$t))
```

A histogram and nornal QQ-plot of the bootstrap values, shown in Figure
@fig-BootPlotMethod, can be created using the `plot` method

```{r}
#| label: fig-BootPlotMethod
#| echo: true
#| fig-cap: plot method for objects of class `boot`
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(muhat.boot)
```

Normal, percentile, and bias-adjusted (bca) percentile 95\% confidence intervals can be computed using

```{r}
#| echo: true

boot.ci(muhat.boot, conf=0.95, type=c("norm", "perc", "bca"))
```

Because the bootstrap distribution looks normal, the normal, percentile, and bca confidence intervals are very similar. 


The GWN model estimate $\hat{\sigma}$ can be "bootstrapped" in a similar fashion. First, I write a function to compute $\hat{\sigma}$
for each bootstrap sample:


```{r}
#| echo: true

sd.boot = function(x, idx) {    
	ans = sd(x[idx])      
	ans 
}
```

Then I call `boot()` with `statistic=sd.boot`

```{r}
#| echo: true

set.seed(123)
sigmahat.boot = boot(msftRetS, statistic = sd.boot, R=999) 
sigmahat.boot
```

```{r}
#| label: fig-BootPlotMethodSD
#| echo: false
#| fig-cap: Bootstrap distribution for $\hat{\sigma}$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sigmahat.boot)
```

The bootstrap distribution, shown in Figure @fig-BootPlotMethodSD,
looks a bit non-normal.  The 95\% confidences intervals are computed using

```{r}
#| echo: true

boot.ci(sigmahat.boot, conf=0.95, type=c("norm", "perc", "bca"))
```

You see that all of these intervals are quite similar.
\hfill{}$\blacksquare$
:::

::: {#exm-nonparametric-bootstrap-for-example-functions .example name="Nonparametric bootstrap for example functions"}

The real power of the bootstrap procedure comes when you apply it to plug-in estimates of functions like @eq-f1-@eq-f3 for which there are no easy analytic formulas for estimating bias and standard errors. In this situation, the bootstrap procedure easily computes numerical estimates of the bias, standard error, and 95\% confidence interval. All you need is an R function for computing the plug-in function estimate in a form suitable for `boot()`. The R functions for the example functions are:

```{r}
#| echo: true

f1.boot = function(x, idx, alpha=0.05) {
  q = mean(x[idx]) + sd(x[idx])*qnorm(alpha)
  q
}

f2.boot = function(x, idx, alpha=0.05, w0=100000) {
  q = mean(x[idx]) + sd(x[idx])*qnorm(alpha)
  VaR = -w0*q
  VaR
}

f3.boot = function(x, idx, alpha=0.05, w0=100000) { 
	q = mean(x[idx]) + sd(x[idx])*qnorm(alpha) 	
	VaR = -(exp(q) - 1)*w0 	
	VaR 
}
```

The functions $f_1$, $f_2$, and $f_3$ have the additional argument `alpha`
which specifies the tail probability for the quantile, and the functions $f_2$, and $f_3$ have the additional argument `w0` for the initial wealth invested.  

To compute the nonparametric bootstrap for $f_1$ with $\alpha=0.05$ use:

```{r}
#| echo: true

set.seed(123)
f1.bs = boot(msftRetS, statistic = f1.boot, R=999) 
f1.bs
```
Here, the bootstrap estimate of bias is small and the bootstrap estimated standard error is similar to the delta method and jackknife standard errors. 

```{r}
#| label: fig-BootPlotMethodf1
#| echo: false
#| fig-cap: Bootstrap distribution for $\hat{f}_1$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(f1.bs)
```

The bootstrap distribution looks normal, and the three methods for computing 95\% confidence intervals are essentially the same:

```{r}
boot.ci(f1.bs, type=c("norm", "perc", "bca"))
```

To compute the nonparametric bootstrap for $f_2$ with $\alpha=0.05$ and $W_0=\$100,000$ use:

```{r}
#| echo: true
set.seed(123)
f2.bs = boot(msftRetS, statistic = f2.boot, R=999) 
f2.bs
```

The bootstrap estimate of bias is small and the bootstrap estimated standard error is similar to the delta method and jackknife standard errors.

```{r}
#| label: fig-BootPlotMethodf2
#| echo: false
#| fig-cap: Bootstrap distribution for $\hat{f}_2$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(f2.bs)
```

The bootstrap distribution looks normal, and the three bootstrap 95\% confidence intervals are all very similar:

```{r}
boot.ci(f2.bs, type=c("norm", "perc", "bca"))
```
To compute the nonparametric bootstrap for $f_3$ with $\alpha=0.05$ and $W_0=\$100,000$ use:

```{r}
#| echo: true
set.seed(123)
f3.bs = boot(msftRetC, statistic = f3.boot, R=999) 
f3.bs
```
The bias is small and the bootstrap estimated standard error is larger than the delta method standard error and is closer to the jackknife standard error. 

```{r}
#| label: fig-BootPlotMethodf3
#| echo: false
#| fig-cap: Bootstrap distribution for $\hat{f}_3$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(f3.bs)
```

The bootstrap distribution looks a little asymmetric and the BCA confidence interval is a bit different from the normal and percentile intervals:

```{r}
boot.ci(f3.bs, type=c("norm", "perc", "bca"))
```

\hfill{}$\blacksquare$
:::

### Pros and cons of the nonparametric bootstrap

The nonparametric bootstrap is extremely useful and powerful statistical technique. The main advantages (pros) are:

* General procedure to estimate bias and standard errors, and to compute confidence intervals, that does not rely on asymptotic distributions. 

* Can be used for scalar and vector estimates.

* Under general conditions, bootstrap distributions of estimators converge to the asymptotic distribution of the estimators. 

* In certain situations, bootstrap results can be more accurate than analytic results computed from asymptotic distributions.

There are some things to watch out for when using the bootstrap (cons):

* It be computational intensive but this is really not much of an issue with today's fast computers.

* With small $B$, bootstrap results can vary substantially across simulations with different random number seeds. 

* There are situations where the bootstrap does not work. A leading case is when the bootstrap is applied to a function that can be become unbounded (e.g. a ratio of means when the denominator mean is close to zero). In these cases, a trimmed version of the bootstrap is recommended. See Hansen (2011) for a discussion of this issue.

* You have to be careful applying the bootstrap to time series data. If there is serial correlation in the data then the nonparametric bootstrap may give poor results. There are versions of the bootstrap (e.g. the block bootstrap) that work with serial correlated data. For example, the *tseries* function `tsbootstrap()` is designed to to work with serially correlated time series data. 

## Using Monte Carlo Simulation to Evaluate Delta Method, Jackknife and Bootstrap Standard Errors {#sec-Using-Monte-Carlo}

* You have studied three ways of computing standard errors for functions of GWN model parameters

* Give similar results for the example functions

* What are the preferred methods in practice?

* Which methods are most accurate?

* Use MC Simulation to evaluate the three methods in stylized model in which the GWN model is true. MC can give the best numerical standard error with which you can compare to the delta method, jackknife and bootstrap standard errors. 

You can evaluate the statistical properties of $f_j(\hat{\boldsymbol{\theta}})$, for $j=1,\ldots,3$,
by Monte Carlo simulation in the same way that you evaluated the statistical properties of $\hat{\mu}$ in Chapter @sec-Estimation-of-The-GWN-Model. You use the simulation model @eq-GWN-simulation-model and $N=1000$ simulated samples of size $T=100$ to compute the estimates $\{\left(f_{j}(\hat{\boldsymbol{\theta}})\right)^{1},\ldots,\left(f_j(\hat{\boldsymbol{\theta}})\right)^{1000}\}$
The true values for $f_j(\boldsymbol{\theta})$ are:
\begin{align*}
f_1(\boldsymbol{\theta}) & = 0.05+(0.10)\times(-1.645)=-0.114,\\
f_2(\boldsymbol{\theta}) & = - \$100,000 \times f_1(\boldsymbol{\theta}) = \$11,449, \\
f_3(\boldsymbol{\theta}) & = \$100,000 \times \left(\exp(f_1(\boldsymbol{\theta})) -1  \right) = \$10,818.
\end{align*}

The R code for the simulation loop is:

```{r}
#| echo: true
#| cache: true

mu = 0.05 
sigma = 0.10 
W0 = 100000 
f1 = mu + sigma*qnorm(0.05) 
f2 = -W0*f1 
f3 = -W0*(exp(f1) - 1)
n.obs = 172 
n.sim = 1000 
set.seed(111) 
sim.f = matrix(0,n.sim,3)  
dm.f = matrix(0, n.sim, 3)
jack.f = matrix(0, n.sim, 3)
boot.f = matrix(0, n.sim, 3)
boot.bias = matrix(0, n.sim, 3)
colnames(sim.f) = colnames(dm.f) = colnames(jack.f) = c("f1", "f2","f3") 
colnames(boot.f) = colnames(boot.bias) = c("f1", "f2","f3") 
 
for (sim in 1:n.sim) { 	
sim.ret = rnorm(n.obs,mean=mu,sd=sigma) 	
muhat = mean(sim.ret) 	
sigmahat = sd(sim.ret) 	
sim.f[sim, "f1"] = muhat + sigmahat*qnorm(0.05) 	
sim.f[sim, "f2"] = -W0*sim.f[sim, "f1"]  	
sim.f[sim, "f3"] = -W0*(exp(sim.f[sim, "f1"]) - 1) 	
# delta method se
dm.f[sim, "f1"] = (sigmahat/sqrt(n.obs))*sqrt(1 + 0.5*qnorm(0.05)^2)
dm.f[sim, "f2"] = W0*dm.f[sim, "f1"]
dm.f[sim, "f3"] = W0*exp(sim.f[sim, "f1"])*dm.f[sim, "f1"]
# jackknife se
f1.jack = jackknife(1:n.obs, theta=theta.f1, as.matrix(sim.ret))
jack.f[sim, "f1"] = f1.jack$jack.se
f2.jack = jackknife(1:n.obs, theta=theta.f2, as.matrix(sim.ret), W0)
jack.f[sim, "f2"] = f2.jack$jack.se
f3.jack = jackknife(1:n.obs, theta=theta.f3, as.matrix(sim.ret), W0)
jack.f[sim, "f3"] = f3.jack$jack.se
# bootstrap se
f1.bs = boot(sim.ret, statistic = f1.boot, R=999) 
boot.f[sim, "f1"] = sd(f1.bs$t)
f2.bs = boot(sim.ret, statistic = f2.boot, R=999) 
boot.f[sim, "f2"] = sd(f2.bs$t)
f3.bs = boot(sim.ret, statistic = f3.boot, R=999) 
boot.f[sim, "f3"] = sd(f3.bs$t)
# bootstrap bias
boot.bias[sim, "f1"] = summary(f1.bs)$bootBias
boot.bias[sim, "f2"] = summary(f2.bs)$bootBias
boot.bias[sim, "f3"] = summary(f3.bs)$bootBias
}
```

The histograms of $f_j(\hat{\boldsymbol{\theta}})$ values are displayed in Figure @fig-CerModelEstimationMCVaR. The dotted white lines indicate the true values, $f_j(\boldsymbol{\theta})$.  


```{r}
#| label: fig-CerModelEstimationMCVaR
#| echo: false
#| cache: true
#| fig-cap: Histograms of $\hat{f}_{j}$ (for $j=1,\ldots, 4$) computed from $N=1000$
#|   Monte Carlo samples from the GWN model.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))
# q.05   
hist(sim.f[, "f1"], col="cornflowerblue", xlab="f1 values",         
     main="f1", probability=T) 	
abline(v=f1, col="white", lwd=4, lty=2)   
# overlay normal distribution   
#x.vals = seq(-0.16, 0.06, length=100)   
#lines(x.vals, dnorm(x.vals, mean=q.05, sd=((sigma/sqrt(n.obs))*sqrt(1 + #0.5*qnorm(0.05)^2))),
#      col="orange", lwd=2)   
# q.01 	
hist(sim.f[, "f2"], col="cornflowerblue", xlab="f2 values",         
     main="f2", probability=T) 	
abline(v=f2, col="white", lwd=4, lty=2)   
# overlay normal distribution   
# x.vals = seq(-0.24, -0.12, length=100)   
# lines(x.vals, dnorm(x.vals, mean=q.01, sd=((sigma/sqrt(n.obs))*sqrt(1 + 0.5*qnorm(0.01)^2))),
#       col="orange", lwd=2)   
# VaR.05   
hist(sim.f[, "f3"], col="cornflowerblue", xlab="f3 values",         
     main="f3", probability=T) 	
abline(v=f3, col="white", lwd=4, lty=2)   
# overlay normal distribution   
# x.vals = seq(-16000, 6000, length=100)   
# lines(x.vals, dnorm(x.vals, mean=VaR.05, sd=((W0*sigma/sqrt(n.obs))*sqrt(1 + 0.5*qnorm(0.05)^2))),
#       col="orange", lwd=2)   
# VaR.01   
par(mfrow=c(1,1))
```

The histograms for the $f_j(\hat{\boldsymbol{\theta}})$ values are bell-shaped for $j=1,2,3$ and the histogram for $j=4$ is slightly right skewed. All of the histograms are centered very close to the respective true values. Recall, the standard deviation of the Monte Carlo estimates (spread of the histograms about true value) approximates the true standard error of the plug-in estimates.

The Monte Carlo estimates of bias are:

```{r}
#| echo: true

mc.bias = colMeans(sim.f) - c(f1, f2, f3)
round(mc.bias, digits=3)
```

which confirm that $f_j$ are essentially unbiased estimators. It is often useful to also look at the bias relative to the true value to eliminate units:

```{r}
mc.bias/c(f1, f2, f3)
```
In relative terms, the largest bias is xx\%, which occurs for the estimate of xxx.

The average bootstrap estimate of bias is:

```{r}
boot.bias.vals = colMeans(boot.bias) 
round(boot.bias.vals, digits=3)
```
The bootstrap bias values are very close to the Monte Carlo estimates of bias which indicates that the bootstrap estimate of bias is very accruate in this context. 

The sample standard deviations of the Monte Carlo estimates approximate the true standard errors of the estimated functions:

```{r}
#| echo: true

mc.se = apply(sim.f, 2, sd) 
round(mc.se, digits=3)
```
The sample means of the delta method, jackknife, and bootstrap standard errors give the average performance of these estimated standard errors. It is informative to compare these average standard errors to the Monte Carlo standard errors: 

```{r}
dm.se = colMeans(dm.f)
jack.se = colMeans(jack.f)
boot.se = colMeans(boot.f)
ans = rbind(dm.se/mc.se, jack.se/mc.se, boot.se/mc.se)
rownames(ans) = c("Delta Method", "Jackknife", "Bootstrap")
ans
```
In the table above, a value of $1$ means the estimated standard error is the same as the Monte Carlo standard error (on average), and a value greater than $1$ means that the estimated standard error is larger than the Monte Carlo standard error (on average). Here, you see that all of the estimated standard errors are very close to the Monte Carlo standard errors, and overall the bootstrap estimated standard errors are the closest. The worst performance is for $f_xx(\hat{\boldsymbol{\theta}})$ where the estimated standard errors are about xx\% to xx\% larger than the Monte Carlo standard errors. 

It is also useful to look at the standard deviations of the estimated standard errors over the Monte Carlo simulations. This tells you how variable the estimated standard errors are across different simulations:

```{r}
dm.sd = apply(dm.f,2,sd)
jack.sd = apply(jack.f,2,sd)
boot.sd = apply(boot.f,2,sd)
ans = rbind(dm.sd, jack.sd, boot.sd)
rownames(ans) = c("Delta Method", "Jackknife", "Bootstrap")
ans

```
Here, you see that, across the simulations, the delta method is the least variable and the jackknife and the bootstrap have similar variability that is about twice as large as the delta method variability. The variability of the bootstrap can be reduced by increasing the number of bootstrap simulations (here it is 1000). 

Overall, the Monte Carlo experiment confirms that the delta method, jackknife and bootstrap all produce accurate standard error estimates for the example plug-in functions.

## Further Reading: Delta Method and Resampling

Chernick and LaBudde (2011) give a comprehensive overview of the bootstrap with applications in R. 

## Exercises

::: {#exr-delta-method .exercise name="Delta method"}
* TBD
:::

::: {#exr-gradients-for-example-functions .exercise name="Gradients for example functions"}
* TBD
:::

::: {#exr-standard-error-of-sharpe-ratio .exercise name="Standard error of Sharpe ratio"}
* Look at determinants of se of Sharpe ratio. See discussion from Lo's paper
:::


::: {#exr-correlation-between-estimated-sharpe-ratios .exercise name="Correlation between estimated Sharpe ratios"}
* Look at the formula for the covariance between two Sharpe ratios
:::


::: {#exr-bootstrap-percentile-confidence-interval .exercise name="Bootstrap percentile confidence interval"}

Illustrate the transformation preserving property of percentile interval using the example of computing the confidence interval for $\sigma$ based on the percentile interval for $\sigma^2$. Show that asymptotic confidence interval is not transformation preserving.
:::

::: {#exr-bootstrap-variability .exercise name="Bootstrap variability"}

Repeat the bootstrap simulations from Example 8.16 using 5 different random number seeds: 124, 125, 126, 127 and 128. Make a table to compare the bootstrap estimated standard errors for the example functions across the 5 bootstrap simulations. Briefly comment on your results.
:::

::: {#exr-failure-of-the-bootstrap .exercise name="Failure of the bootstrap"}

* Run the bootstrap for calculating a standard error for $\mu_i/\mu_j$.

:::

::: {#exr-bootstrap-accuracy .exercise name="Bootstrap accuracy"}

Repeat the Monte Carlo simulation in @sec-Using-Monte-Carlo using 1,000 simulated samples and 10,000 bootstrap simulations for the bootstrap calculations. That is, set R=10000 in the call to `boot()` inside the Monte Carlo simulation loop. 

1. How does the increase in simulations impact the bootstrap bias estimates?

2. How does the increase in simulations impact the bootstrap standard error estimates?

:::


