# Hypothesis Testing in the GWN Model {#sec-Hypothesis-Testing}

Updated: September 7, 2026

Copyright © Eric Zivot 2026


The core of statistical analysis involves estimation and inference. The previous chapters considered estimation of the GWN model parameters, and the computation of standard error estimates and confidence intervals. In this chapter I discuss issues of statistical inference within the GWN model. Statistical inference is about asking questions about a model and using statistical procedures to obtain answers to these questions with a given level of confidence. For example, you might find that our estimate of the monthly mean of an asset return is positive but due to large estimation error we might ask the question: is the true monthly mean return positive?  As another example, you might find that a normal qq-plot of returns shows some evidence of non-normality and you ask the question: are returns normally distributed? As a third example, you might find that rolling estimates of the mean and volatility show some evidence of non-constant behavior and you might ask the question: are returns covariance stationary? Statistical hypothesis testing gives us a rigorous framework for answering such questions with a given level of confidence.

The R packages used in this chapter are **introcompfinr**, **PerformanceAnalytics**, **tseries**, and **zoo**. Make sure this packages are installed and loaded before replicating the examples.

```{r}
#| include: false
suppressPackageStartupMessages(library(introcompfinr))
# suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(PerformanceAnalytics))
suppressPackageStartupMessages(library(tseries))
options(digits=3)
```


1.  Hypothesis Testing in the GWN Model
    a.  Specification tests
        i.  Normal distribution
            A.  Tests for skewness and kurtosis
            B.  JB test
            C.  What to do if returns are not normally distributed?
        ii.  No autocorrelation
            A.  individual and joint tests
        iii.  Constant parameters (covariance stationarity)
            A.  Formal tests: testing same mean, volatility in sub-samples
            B.  Informal diagnostics - standard error bands on rolling estimators
    b.  Using Monte Carlo Simulation to Understand Hypothesis tests
        i.  Size
        ii.  Power
    c. Using the bootstrap for hypothesis tests
        a. exploit duality between hypothesis tests and confidence intervals

2.  Appendix: Distributions for test statistics
    a.  Student's t distribution
    b.  Chi-square distribution

## Review of General Principles {#sec-Review-of-General-Principles}

Classical hypothesis testing is discussed in almost all introductory textbooks
on statistics, and the reader is assumed be familiar with the basic
concepts.^[In this book, I do not consider evaluating hypotheses from a 
Bayesian perspective.]  In this section I provide a brief and concise review of
the general principles of classical hypothesis testing. In the following
sections I apply these principles to answer questions about the GWN
model parameters and assumptions.

### Steps for hypothesis testing 

The main steps for conducting a classical hypothesis test are as follows:
\index{Hypothesis testing}

1.  Specify the hypotheses to be tested:
$$
H_{0}:\textrm{null hypothesis vs. }H_{1}:\textrm{alternative hypothesis.}
$$
The *null hypothesis* is the maintained hypothesis (what is assumed to be true) and is the hypothesis to be tested against the data. The *alternative hypothesis* specifies what is assumed to be true if the null hypothesis is false. This can be specific or vague depending on the context.

2.  Specify the significance level of the test:
$$
\alpha=\textrm{significance level}=\Pr(\textrm{Reject }H_{0}|H_{0}\,\textrm{is true}).
$$
The *significance level* of a test specifies the probability of making a certain type of decision error: rejecting the null hypothesis when the null hypothesis is, in fact, true. In practice, the significance level $\alpha$ is chosen to be a small number like $0.01$ or $0.05$. 

3.  Construct a *test statistic*, $S$, from the observed data whose probability distribution is known when $H_{0}$ is true. 

4.  Use the test statistic $S$ to evaluate the data evidence regarding the validity of $H_{0}$. Typically, if $S$ is a big number then there is strong data evidence against $H_{0}$ and $H_{0}$ should be rejected; otherwise, there is not strong data evidence against $H_{0}$ and $H_{0}$ should not be rejected. 

     + Decide to reject $H_{0}$ at the specified significance level $\alpha$ if the value of the test statistic $S$ falls in the *rejection region* of the test. Usually, the rejection region for $S$ is determined by a critical value $cv_{\alpha}$ such that: 
\begin{eqnarray*}
S & > & cv_{\alpha}\Rightarrow\mathrm{reject}\,H_{0},\\
S & \leq & cv_{\alpha}\Rightarrow\mathrm{do\,not\,reject}\,H_{0}.
\end{eqnarray*}
Smaller values of $\alpha$ typically make $cv_{\alpha}$ larger and require more data evidence (i.e., larger $S$) to reject $H_{0}.$

    + Alternatively, decide to reject $H_{0}$ at the specified significance level $\alpha$ if the p-value of the test statistic $S$ is less than $\alpha$. The *p-value* of the statistic $S$ is defined as the significance level at which the test is just rejected.

### Hypothesis tests and decisions

Hypothesis testing involves making a decision: reject $H_{0}$ or
do not reject $H_{0}.$ Notice that if the data evidence strongly
favors $H_{0}$ then you say "I do not reject $H_{0}$" instead
of "accept $H_{0}$". You can rarely know the truth with absolute
certainty with a finite amount of data so it is more appropriate to
say "do not reject $H_{0}$" than to say "accept $H_{0}$".^[As an extreme example, consider testing $H_{0}:\textrm{"all swans are white"}$.
To logically accept this hypothesis you would have to show that every
swan that ever existed is white. In contrast, if you have a data set
that only contains white swans then saying you cannot reject the null
hypothesis that all swans are white is consistent with your data set. ] 


+-----------+----------------------------+-------------------------+--------------------------+
|           |                            |    Reality                                         |
+===========+============================+=========================+==========================+
|           |                            |    **$H_{0}$ is true**  |    **$H_{0}$ is false**  |
+-----------+----------------------------+-------------------------+--------------------------+
|Decision   |**Reject $H_{0}$**          |    Type I error         |    No error              |
+-----------+----------------------------+-------------------------+--------------------------+
|           |**Do not reject $H_{0}$**   |    No error             |    Type II error         |
+-----------+----------------------------+-------------------------+--------------------------+

: Decision table for hypothesis testing {#tbl-HTdecisionTable}


Table @tbl-HTdecisionTable shows the $2\times2$ decision
table associated with hypothesis testing. In reality there are two
states of the world: either $H_{0}$ is true or it is not.^[It is important to emphasize that there is a "true state of the world" in this context. That is, the null hypothesis $H_{0}$ is
either true or it is not. As a result, $\Pr\left\{H_{0}\,\textrm{is true}\right\} =1$
or $0$ depending on whether $H_{0}$ is true or not. ] There are two decisions to make: reject $H_{0}$ or don't reject
$H_{0}.$ If the decision corresponds with reality then the correct
decision is made and there is no decision error. This occurs in the
off-diagonal elements of the table. If the decision and reality disagree
then there is a decision error. These errors are in the diagonal elements
of the table. Type I error\index{Type I error} results when the decision to reject $H_{0}$
occurs when, in fact, $H_{0}$ is true. Type II error\index{Type II error} happens when
the decision not to reject $H_{0}$ occurs when $H_{0}$ is false.
To put these types of errors in context, consider a jury in a capital
murder trial in the US.^[A capital murder trial is one in which the defendant is eligible for
the death penalty.] In the US, a defendant is considered innocent until proven guilty.
Here, the null hypothesis to be tested is:
$$
H_{0}:\textrm{defendant is innocent}.
$$
The alternative is:
$$
H_{1}:\textrm{defendant is guilty}.
$$

There is a true state of the world here: the defendant is either innocent
or guilty. The jury must decide what the state of the world is based
on evidence presented to them. The decision table for the jury is
shown in Table @tbl-HTdecisionTableJury. Type I error occurs
when the jury convicts an innocent defendant and puts that person
on "death row". Clearly, this is a terrible mistake. To avoid
this type of mistake the jury should have a very small (close to zero)
significance level for evaluating evidence so that Type I error very
rarely occurs. This is why typical jury instructions are to convict
only if evidence of guilt is presented beyond reasonable doubt. Type
II error happens when the jury does not convict (acquits) a guilty
defendant and sets that person free. This is also a terrible mistake,
but perhaps it is not as terrible as convicting the innocent person.
Notice that there is a conflict between Type I error and Type II error.
In order to completely eliminate Type I error, you can never reject
$H_{0}$. That is, to avoid ever sending innocent people to "death
row" you must never convict anyone. But if you never convict anyone
then you never convict guilty people either and this increases the
occurrences of Type II errors. 

### Significance level and power

The *significance level*\index{Significance level} of a test is the probability of Type
I error:
$$
\alpha=\Pr(\textrm{Type I error})=\Pr(\textrm{Reject}\,H_{0}|H_{0}\,\textrm{is true}).
$$
The *power*\index{Power of a test} of a test is one minus the probability of Type II
error:
$$
\pi=1-\Pr(\textrm{Type II error})=\Pr(\textrm{Reject }H_{0}|H_{0}\,\textrm{is false}).
$$
In classical hypothesis tests, the goal is to construct a test that
has a small significance level ($\alpha$ close to zero) that you
can control and has high power ($\pi$ close to one). That is, a good
test is one that has a small and controllable probability of rejecting
the null when it is true and, at the same time, has a very high probability
of rejecting the null when it is false. In general, an *optimal test*\index{Optimal hypothesis test}
is one that has the highest possible power for a given significance
level. In the jury example, an optimal test would be the one in which
the jury has the highest probability of convicting a guilty defendant
while at the same time has a very low probability of convicting an
innocent defendant. A difficult problem indeed! 


+-----------+----------------+-----------------------------+---------------------------+
|           |                |    Reality                                              |
+===========+================+=============================+===========================+
|           |                |  **Defendant is innocent**  |  **Defendant is guilty**  |
+-----------+----------------+-----------------------------+---------------------------+
|Decision   |**Convict**     |  Type I error               |  No error                 |
+-----------+----------------+-----------------------------+---------------------------+
|           |**Acquit**      |  No error                   |  Type II error            |
+-----------+----------------+-----------------------------+---------------------------+

: Decision table for jury on capital murder trial {#tbl-HTdecisionTableJury}

## Hypothesis Testing in the GWN Model

In this section, I discuss hypothesis testing in the context of the
GWN model for asset returns:

$$
\begin{aligned}
R_{it} & =\mu_{i}+\varepsilon_{it},\,i=1\ldots,N;\,t=1,\ldots,T\\
\{\varepsilon_{it}\}_{t=1}^{T} & \sim\mathrm{GWN}(0,\sigma_{i}^{2}), \\
\mathrm{cov}(\varepsilon_{it},\varepsilon_{js}) & =\left\{\begin{array}{c}
\sigma_{ij}\\
0
\end{array}\right.\begin{array}{c}
~ t=s\\
t\neq s
\end{array}.
\end{aligned}
$$ {#eq-HTcermodel}

In @eq-HTcermodel, $R_{it}$ is the monthly return (continuously
compounded or simple) on asset $i$ in month $t$. Define $\mathbf{R}_{t}=(R_{1t},\ldots,R_{Nt})^{\prime}$.
The GWN model in matrix form is:

$$
\begin{aligned}
\mathbf{R}_{t}=\mu+\varepsilon_{t},\\
\varepsilon_{t}\sim\mathrm{iid}~N(\mathbf{0},\Sigma). 
\end{aligned}
$$ {#eq-HTcermatrix}

I consider two types of hypothesis tests: (1) tests for model coefficients or functions of model coefficients;
and (2) tests for model assumptions.^[Tests for model assumptions are sometimes called model specification
tests.] These tests can be specified for the GWN model for a single asset
using @eq-HTcermodel, or they can be specified for the GWN
model for all assets using @eq-HTcermatrix. 

### Coefficient tests

Here, I assume that the GWM is the true model for our observed return data and I are interested in asking
questions about the parameters of the model. The most common tests for model coefficients and functions of model coefficients for a single asset are
tests for a specific value. Some typical hypotheses to be tested are:

\begin{align*}
H_{0} & :\mu_{i}=\mu_{i}^{0}\text{ vs. }H_{1}:\mu_{i}\neq\mu_{i}^{0},\\
H_{0} & :\sigma_{i}=\sigma_{i}^{0}\text{ vs. }H_{1}:\sigma_{i}\neq\sigma_{i}^{0},\\
H_{0} & :\rho_{ij}=\rho_{ij}^{0}\text{ vs. }H_{1}:\rho_{ij}\neq\rho_{ij}^{0}, \\
H_{0} & :\mathrm{SR}_{i}=\mathrm{SR}_{i}^{0}\text{ vs. }H_{1}:\mathrm{SR}_{i}\neq\mathrm{SR}_{i}^{0},
\end{align*}

where $\mathrm{SR}_{i} = \frac{\mu_i - r_f}{\sigma_i}$ is the Sharpe ratio of asset $i$ and $r_f$ is the risk-free rate of interest. A hypothesis about the Sharpe ratio is a hypothesis about the ratio of two GWN model parameters. In the above hypotheses, the alternative hypotheses are *two-sided*. That is, under
$H_{1}$ the model parameters are either larger or smaller than what
is specified under $H_{0}$.

Also common are tests for specific sign. For example:

\begin{align*}
H_{0} & :\mu_{i}=0\text{ vs. }H_{1}:\mu_{i}>0\text{ or }H_{1}:\mu_{i}<0,\\
H_{0} & :\rho_{ij}=0\text{ vs. }H_{1}:\rho_{ij}>0\text{ or }H_{1}:\rho_{ij}<0, \\
H_{0} & :\mathrm{SR}_{i}=0\text{ vs. }H_{1}:\mathrm{SR}_{i}>0\text{ or }H_{1}:\mathrm{SR}_{i}<0,
\end{align*}

Here, the alternative hypotheses are *one-sided*. That is, under
$H_{1}$ the model parameter is either greater than zero or less than
zero. 

Sometimes you are interested in testing hypotheses about the
relationship between model coefficients for two or more assets. For
example: 

\begin{align*}
H_{0} & :\mu_{1}=\mu_{2}\,\textrm{vs. } H_{1}:\mu_{1}\neq\mu_{2}, \\
H_{0} & :\mathrm{SR}_{1}=\mathrm{SR}_{2}\,\textrm{vs. } H_{1}:\mathrm{SR}_{1}\neq\ \mathrm{SR}_{2}.
\end{align*}

It is often convenient to re-express these hypotheses in terms of differences of coefficients under the null and alternative hypotheses:

\begin{align*}
H_{0} & :\mu_{1}-\mu_{2}=0 \,\textrm{vs. } H_{1}:\mu_{1}-\mu_{2} \neq 0, \\
H_{0} & :\mathrm{SR}_{1}-\mathrm{SR}_{2}=0 \,\textrm{vs. } H_{1}:\mathrm{SR}_{1}- \mathrm{SR}_{2} \neq 0.
\end{align*}

### Model specification tests

In the GWN model @eq-HTcermodel, asset returns are assumed
to be normally distributed. The hypotheses to be tested for this assumption
are: 

\begin{align*}
H_{0}: & R_{it}\sim N(\mu_{i},\sigma_{i}^{2})\,\textrm{vs. }H_{1}:R_{it}\sim\textrm{ not normal.}
\end{align*}

Notice that the distribution of $R_{t}$ under $H_{1}$ is not specified
(e.g., Student's t with 5 degrees of freedom). 

In addition, it is assumed that asset returns are uncorrelated over time. One can test the hypothesis that each autocorrelation (up to some maximum lag $J$) is zero:

\begin{align*}
H_{0} & :\rho_{j}=\mathrm{cor}(r_{it},r_{i,t-j})=0\text{, }j=1,\ldots,J\,\mathrm{vs}.\\
H_{1} & :\rho_{j}=\mathrm{cor}(r_{it},r_{i,t-j})\neq0,
\end{align*}

or one can test the joint hypothesis that all autocorrelation coefficients
(up to some maximum lag $J$) are zero:

\begin{eqnarray*}
H_{0}:\rho_{1} & = & \rho_{2}=\cdots=\rho_{J}=0\,vs.\\
H_{1}:\rho_{j} & \neq & 0\,\textrm{for some }j\leq J
\end{eqnarray*}

Lastly, it is assumed that asset returns are covariance stationary.
This assumption implies that all model parameters are constant over
time. The hypotheses of interest are then:

\begin{align*}
H_{0} & :\mu_{i},\sigma_{i}\text{ and }\rho_{ij}\text{ are constant over entire sample, vs.}\\
H_{1} & :\mu_{i}\text{ }\sigma_{i}\text{ or }\rho_{ij}\text{ changes in some sub-sample.}
\end{align*}

The following sections describe test statistics for the above listed
hypotheses. 

### Data for hypothesis testing examples

The examples in this chapter use the monthly continuously compounded
returns on Microsoft, Starbucks and the S\&P 500 index over the period
January 1998 through May 2012 created using:

```{r}
#| echo: true
gwnDailyPrices = merge(msftDailyPrices, sbuxDailyPrices, sp500DailyPrices)
gwnMonthlyPrices = to.monthly(gwnDailyPrices, OHLC=FALSE) 
smpl = "1998-01::2012-05" 
gwnMonthlyPrices = gwnMonthlyPrices[smpl] 
gwnMonthlyRetC = na.omit(Return.calculate(gwnMonthlyPrices, method="log"))
```

The monthly continuously compounded returns, with a common y-axis,
are illustrated in Figure @fig-HypothesisExampleReturns created
using:

```{r}
#| label: fig-HypothesisExampleReturns
#| echo: true
#| fig-cap: Monthly cc returns on Microsoft, Starbucks and the S\&P 500 Index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(gwnMonthlyRetC, multi.panel=TRUE, main="",
     col="blue", lwd=2, yaxis.same=TRUE,
     major.ticks="years", grid.ticks.on="years")
```

## Tests for Individual Parameters: t-tests and z-scores

In this section, I present test statistics for testing hypotheses
that certain model coefficients equal specific values. In some cases,
I can derive *exact tests*. For an exact test, the pdf of the
test statistic assuming the null hypothesis is true is known exactly
for a finite sample size $T$. More generally, however, I rely on
*asymptotic tests*\index{Asymptotic tests}. For an asymptotic test, the pdf of the test
statistic assuming the null hypothesis is true is not known exactly
for a finite sample size $T$, but can be approximated by a known
pdf. The approximation is justified by the Central Limit Theorem (CLT)
and the approximation becomes exact as the sample size becomes infinitely
large. 

### Exact tests under normality of data

In the GWN model @eq-HTcermodel, consider testing the hypothesis
that the mean return, $\mu_{i}$, is equal to a specific value $\mu_{i}^{0}$:

$$
H_{0}:\mu_{i}=\mu_{i}^{0}.
$$ {#eq-HT-H0mean}

For example, an investment analyst may have provided an expected return forecast of $\mu_{i}^{0}$
and you would like to see if past data was consistent with such a forecast. 

The alternative hypothesis can be either two-sided or one-sided. The
two-sided alternative is:
$$
H_{1}:\mu_{i}\neq\mu_{i}^{0}.
$$ {#eq-HT-H1mean2sided}
Two-sided alternatives are used when you don't care about the sign
of $\mu_{i}-\mu_{i}^{0}$ under the alternative. With one-sided alternatives,
you care about the sign of $\mu_{i}-\mu_{i}^{0}$ under the alternative:
$$
H_{1}:\mu_{i}>\mu_{i}^{0}\,\textrm{ or }\,H_{1}:\mu_{i}<\mu_{i}^{0}.
$$ {#eq-HT-H1mean1sided}

How do you come up with a test statistic $S$ for testing @eq-HT-H0mean
against @eq-HT-H1mean2sided or @eq-HT-H1mean1sided?
You generally use two criteria: (1) you know the pdf of $S$ assuming
@eq-HT-H0mean is true; and (2) the value of $S$ should be
big if the alternative hypotheses @eq-HT-H1mean2sided or @eq-HT-H1mean1sided
are true. 

#### Two-sided test 

To simplify matters, let's assume that the value of $\sigma_{i}$
in @eq-HTcermodel is known and does not need to be estimated. This assumption 
is unrealistic and will be relaxed later. Consider testing @eq-HT-H0mean against 
the two-sided alternative @eq-HT-H1mean2sided using a 5\% significance level. 
Two-sided alternatives are typically more commonly used than one-sided alternatives.
Let $\{R_{t}\}_{t=1}^{T}$ denote a random sample from the GWN model. You estimate $\mu_{i}$
using the sample mean:

$$
\hat{\mu}_{i}=\frac{1}{T}\sum_{t=1}^{T}R_{it}.
$$ {#eq-HT-muhat}

Under the null hypothesis @eq-HT-H0mean, it is assumed that
$\mu_{i}=\mu_{i}^{0}$. Hence, if the null hypothesis is true then
$\hat{\mu}_{i}$ computed from data should be close to $\mu_{i}^{0}$.
If $\hat{\mu}_{i}$ is far from $\mu_{i}^{0}$
(either above or below $\mu_{i}^{0}$) then this
evidence casts doubt on the validity of the null hypothesis. Now,
there is estimation error in @eq-HT-muhat which is measured
by: 

$$
\mathrm{se}(\hat{\mu}_{i})=\frac{\sigma_{i}}{\sqrt{T}}.
$$ {#eq-HT-semuhat}

Because you assume $\sigma_{i}$ is known, you also know $\mathrm{se}(\hat{\mu}_{i})$.
Because of estimation error, you don't expect $\hat{\mu}_{i}$ to equal
$\mu_{i}^{0}$ even if the null hypothesis @eq-HT-H0mean
is true. How far from $\mu_{i}^{0}$ can $\hat{\mu}_{i}$
be if the null hypothesis is true? To answer this question, recall
from Chapter @sec-Estimation-of-The-GWN-Model that the exact
(finite sample) pdf of $\hat{\mu}_{i}$ is the normal distribution:

$$
\hat{\mu}_{i}\sim N(\mu_{i},\mathrm{se}(\hat{\mu}_{i})^{2})=N\left(\mu_{i},\frac{\sigma_{i}^{2}}{T}\right).
$$ {#eq-HT-pdfmuhat}

Under the null hypothesis @eq-HT-H0mean, the normal distribution
for $\hat{\mu}_{i}$ is centered at $\mu_{i}^{0}$:

$$
\hat{\mu}_{i}\sim N(\mu_{i}^{0},\mathrm{se}(\hat{\mu}_{i})^{2})=N\left(\mu_{i}^{0},\frac{\sigma_{i}^{2}}{T}\right).
$$ {#eq-HT-pdfmuhatH0}

From properties of the normal distribution you have:

$$
\Pr\left(\mu_{i}^{0}-1.96\times\mathrm{se}(\hat{\mu}_{i})\leq\hat{\mu}_{i}\leq\mu_{i}^{0}+1.96\times\mathrm{se}(\hat{\mu}_{i})\right)=0.95.
$$ {#eq-HT-probH0}

Hence, if the null hypothesis @eq-HT-H0mean is true you only
expect to see $\hat{\mu}_{i}$ more than $1.96$ values of $\mathrm{se}(\hat{\mu}_{i})$
away from $\mu_{i}^{0}$ with probability 0.05.^[This means that in an infinite number of hypothetical samples from
the GWN model in which $H_{0}$ is true only 5 percent of the samples
produce an estimate $\hat{\mu}_{i}$ that is more than $1.96$ values
of $\mathrm{se}(\hat{\mu}_{i})$ away from $\mu_{i}^{0}$.] Therefore, it makes intuitive sense to base a test statistic for
testing @eq-HT-H0mean on a measure of distance between $\hat{\mu}_{i}$
and $\mu_{i}^{0}$ relative to $\mathrm{se}(\hat{\mu}_{i})$.
Such a statistic is the *z-score*\index{Z-score}:

$$
z_{\mu=\mu^{0}}=\frac{\hat{\mu}_{i}-\mu_{i}^{0}}{\mathrm{se}(\hat{\mu}_{i})}=\frac{\hat{\mu}_{i}-\mu_{i}^{0}}{\sigma/\sqrt{T}}.
$$ {#eq-HT-zscore}

Assuming the null hypothesis @eq-HT-H0mean is true, from @eq-HT-pdfmuhatH0
it follows that:
$$
z_{\mu=\mu^{0}}\sim N(0,1).
$$

The intuition for using the z-score @eq-HT-zscore to test @eq-HT-H0mean is straightforward. If $z_{\mu=\mu^{0}}\approx0$ then $\hat{\mu}_{i}\approx\mu_{i}^{0}$ and @eq-HT-H0mean
should not be rejected. In contrast, if $z_{\mu=\mu^{0}}>1.96$ or $z_{\mu=\mu^{0}}<-1.96$
then $\hat{\mu}_{i}$ is more than $1.96$ values of $\mathrm{se}(\hat{\mu}_{i})$
away from $\mu_{i}^{0}$. From @eq-HT-probH0, this is very
unlikely (less than 5\% probability) if @eq-HT-H0mean is true.
In this case, there is strong data evidence against @eq-HT-H0mean
and you should reject it. Notice that the condition $z_{\mu=\mu^{0}}>1.96\textrm{ or }z_{\mu=\mu^{0}}<-1.96$
can be simplified as the condition $\left|z_{\mu=\mu^{0}}\right|>1.96$.
When $\left|z_{\mu=\mu^{0}}\right|$ is big, larger than 1.96, you
have data evidence against @eq-HT-H0mean. Using the value
$1.96$ to determine the rejection region for the test ensures that
the significance level (probability of Type I error) of the test is
exactly 5\%:

\begin{align*}
\Pr(\textrm{Reject } H_{0}|H_{0}\,\textrm{is true}) &= \Pr(S>1.96|\mu_{i}=\mu_{i}^{0}) \\
&=\Pr\left(\left|z_{\mu=\mu^{0}}\right|>1.96|\mu_{i}=\mu_{i}^{0}\right)=.05
\end{align*}

Hence, our formal test statistic for testing @eq-HT-H0mean
against @eq-HT-H1mean2sided is 

$$
S=\left|z_{\mu=\mu^{0}}\right|. 
$$ {#eq-testStatZscore}

The 5\% critical value is $cv_{.05}=1.96$, and you reject @eq-HT-H0mean
at the 5\% significance level if $S>1.96$. 

In summary, the steps for using $S=\left|z_{\mu=\mu^{0}}\right|$
to test the @eq-HT-H0mean against @eq-HT-H1mean2sided
are:

1.  Set the significance level $\alpha=\Pr(\textrm{Reject } H_{0}|H_{0}\,\textrm{is true})$ and determine the critical value $cv_{\alpha}$ such that $\Pr(S>cv_{\alpha})=\alpha$. Now, 

\begin{align*}
\Pr(S>cv_{\alpha}) &= \Pr\left(\left|z_{\mu=\mu^{0}}\right|>cv_{\alpha}\right) \\
&= \Pr\left(z_{\mu=\mu^{0}}>cv_{\alpha}\right)+\Pr\left(z_{\mu=\mu^{0}}<-cv_{\alpha}\right)=\alpha
\end{align*}

which implies that $cv_{\alpha}=-q_{\alpha/2}^{Z}=q_{1-\alpha/2}^{Z}$, where $q_{\alpha/2}^{Z}$ denotes the $\frac{\alpha}{2}-$quantile of $Z\sim N(0,1)$. For example, if $\alpha=.05$ then $cv_{.05}=-q_{.025}^{Z}=1.96.$ 

2. Reject @eq-HT-H0mean at the $\alpha\times100\%$ significance level if $S>cv_{\alpha}$. For example, if $\alpha=.05$ then reject @eq-HT-H0mean at the 5\% level if $S>1.96$.

3.  Equivalently, reject @eq-HT-H0mean at the $\alpha\times100\%$ significance level if the p-value for $S$ is less than $\alpha$. Here, the *p-value* is defined as the significance level at which the test is just rejected. Let $Z\sim N(0,1)$. The p-value is computed as:
$$
\textrm{p-value} = \Pr(|Z|>S)=\Pr(Z>S)+\Pr(Z<-S) = 2\times \Pr(Z>S)=2\times(1-\Pr(Z<S)). 
$$ {#eq-HT-pvalue2sided}

::: {#exm-testing-hypothesis-with-the-z-score-using-simulated-data-two-sided-tests .example name="Testing hypothesis with the z-score using simulated data: two-sided tests"}

Assume returns follow the GWN model

\begin{eqnarray*}
R_{t} & = & 0.05+\epsilon_{t},\,t=1,\ldots60,\\
\epsilon_{t} & \sim & GWN(0,0.10).
\end{eqnarray*}

and you are interested in testing the following hypotheses
\begin{align*}
H_{0}:\mu=0.05\,\,vs.\,\,H_{1}:\mu\neq.05 \\
H_{0}:\mu=0.06\,\,vs.\,\,H_{1}:\mu\neq.06 \\
H_{0}:\mu=0.10\,\,vs.\,\,H_{1}:\mu\neq.10 \\
\end{align*}

using the test statistic @eq-testStatZscore with a 5\% significance level. The 5\%
critical value is $cv_{.05}=1.96$. One hypothetical sample from the hypothesized model
is simulated using:

```{r}
#| echo: true
set.seed(123)
n.obs = 60
mu = 0.05
sigma = 0.10
ret.sim = mu + rnorm(n.obs, sd = sigma)
```

The estimate of $\mu$ and the value of $\mathrm{se}(\hat{\mu})$ are:

```{r}
muhat = mean(ret.sim)
se.muhat = sigma/sqrt(n.obs)
c(muhat, se.muhat)

```
The test statistic @eq-testStatZscore for testing $H_0:\mu=0.05$ computed from this sample is:

```{r}
z.score.05 = (muhat - 0.05)/se.muhat
S = abs(z.score.05)
S
```
The z-score tells us that the estimated mean, $\hat{\mu}=0.0566$, is 0.58 values of $\mathrm{se}(\hat{\mu})$ away from the hypothesized value $\mu^0=0.05$. This evidence does not contradict $H_0:\mu=0.05$. Since $S=0.508 < 1.96$, you do not reject $H_0:\mu=0.05$ at the 5\% significance level. The p-value of the test using @eq-HT-pvalue2sided is:

```{r}
#| echo: true
p.value = 2*(1 - pnorm(S))
p.value
```
Here, the p-value of 0.611 is less than the significance level $\alpha = 0.05$ so you do not reject the null at the 5\% significance level. The p-value tells us that you would reject $H0:\mu=0.05$ at the 61.1\% significance level.

The z-score information for testing $H_0:\mu=0.06$ is:

```{r}
z.score.06 = (muhat - 0.06)/se.muhat
S = abs(z.score.06)
S
```
Here, the z-score indicates that $\hat{\mu}=0.0566$ is just 0.266 values of $\mathrm{se}(\hat{\mu})$ away from the hypothesized value $\mu^0=0.06$. This evidence also does not contradict $H_0:\mu=0.06$. Since $S=0.226 < 1.96$, you do not reject $H_0:\mu=0.06$ at the 5\% significance level. The p-value of the test  is:

```{r}
#| echo: true
p.value = 2*(1 - pnorm(S))
p.value
```
The large p-value shows that there is insufficient data to reject $H_0:\mu=0.06$.

Last, the z-score information for testing $H_0:\mu=0.10$ is:

```{r}
z.score.10 = (muhat - 0.10)/se.muhat
S = abs(z.score.10)
S
```
Now, the z-score indicates that $\hat{\mu}=0.0566$ is 3.36 values of $\mathrm{se}(\hat{\mu})$ away from the hypothesized value $\mu^0=0.10$. This evidence is not in support of $H_0:\mu=0.10$. Since $S=3.36 > 1.96$, you reject $H_0:\mu=0.06$ at the 5\% significance level. The p-value of the test is:

```{r}
#| echo: true
p.value = 2*(1 - pnorm(S))
p.value
```
Here, the p-value is much less 5\% supporting the rejection of $H_0:\mu=0.06$.  
\hfill{}$\blacksquare$
:::

#### One-sided test

Here, I consider testing the null @eq-HT-H0mean against the one-sided alternative @eq-HT-H1mean1sided. For expositional purposes, consider $H_{1}:\mu_{i} > \mu_{i}^{0}$. This alternative can be equivalently represented as $H_{1}:\mu_{i} - \mu_{i}^{0} > 0$. That is, under the alternative hypothesis the sign of the difference $\mu_{i} - \mu_{i}^{0}$ is positive. This is why a test against a one-sided alternative is sometimes called a *test for sign*.  As in the previous sub-section, assume that $\sigma_i$ is known and let the significance level be $5\%$.  

The natural test statistic is the z-score @eq-HT-zscore. The intuition is straightforward. If $z_{\mu=\mu^0} \approx 0$ then $\hat{\mu}_i \approx \mu_i^0$ and the null should not be rejected. However, if the one-sided alternative is true then you would expect to see $\hat{\mu}_i > \mu_i^0$ and $z_{\mu=\mu^0} > 0$. How big $z_{\mu=\mu^0}$ needs to be for us to reject the null depends on the significance level. Since under the null $z_{\mu=\mu^0} \sim N(0,1) = Z$, with a $5\%$ significance level 

$$
\Pr(\text{Reject } H_0 | H_0 \text{ is true}) = \Pr(z_{\mu=\mu^0} > 1.645) = \Pr(Z > 1.645) = 0.05.
$$
Hence, our $5\%$ one-sided critical value is $cv_{.05}=q_{.95}^Z = 1.645$, and you reject the null at the $5\%$ significance level if $z_{\mu=\mu^0} >1.645$. 

In general, the steps for using $z_{\mu=\mu^0}$ to test @eq-HT-H0mean against the one-sided alternative $H_{1}:\mu_{i} > \mu_{i}^{0}$ are:

1. Set the significance level $\alpha=\Pr(\textrm{Reject } H_{0}|H_{0}\,\textrm{is true})$ and determine the critical value $cv_{\alpha}$ such that $\Pr(z_{\mu=\mu^0}>cv_{\alpha})=\alpha$. Since under the null $z_{\mu=\mu^0} \sim N(0,1) = Z$, $cv_{\alpha} = q_{1-\alpha}^Z$.

2. Reject the null hypothesis @eq-HT-H0mean at the $\alpha \times 100\%$ significance level if $z_{\mu=\mu^0} > cv_{\alpha}$. 

3. Equivalently, reject @eq-HT-H0mean at the $\alpha \times 100\%$ significance level if the one-sided p-value is less than $\alpha$, where

$$
\text{p-value} = \Pr(Z > z_{\mu=\mu^0}) = 1 - \Pr(Z \le z_{\mu=\mu^0})
$$

::: {#exm-testing-hypothesis-with-the-z-score-using-simulated-data-one-sided-tests .example name="Testing hypothesis with the z-score using simulated data: one-sided tests"}

Here, I use the simulated GWN return data from the previous example to test @eq-HT-H0mean against the one-sided alternative $H_{1}:\mu_{i} > \mu_{i}^{0}$ using a 5\% significance level. 

The z-scores and one-sided p-values for testing $H_0:\mu=0.05$, $H_0:\mu=0.06$, and $H_0:\mu=0.10$, computed from the simulated sample are:

```{r}
z.score.05 = (muhat - 0.05)/se.muhat
p.value.05 = 1 - pnorm(z.score.05)
z.score.06 = (muhat - 0.06)/se.muhat
p.value.06 = 1 - pnorm(z.score.06)
z.score.10 = (muhat - 0.10)/se.muhat
p.value.10 = 1 - pnorm(z.score.10)
ans = rbind(c(z.score.05, p.value.05),
            c(z.score.06, p.value.06),
            c(z.score.10, p.value.10))
colnames(ans) = c("Z-score", "P-value")
rownames(ans) = c("H0:mu=0.05", "H0:mu=0.06","H0:mu=0.10")
ans
```
Here, you do not reject any of the null hypotheses in favor of the one-sided alternative at the 5\% significance level. 
\hfill{}$\blacksquare$
:::

### Exact tests with $\sigma$ unknown 

In practice, $\sigma^2$ is unknown and is estimated with $\hat{\sigma}^2$, and so the exact tests described above are not feasible. Fortunately, an exact test is still available. Instead of using the z-score @eq-HT-zscore, we use the t-ratio (or t-score) 

$$
t_{\mu=\mu^0}=\frac{\hat{\mu}_{i}-\mu_{i}^0}{\widehat{\mathrm{se}}(\hat{\mu}_{i})}=\frac{\hat{\mu}_{i}-\mu_{i}^0}{\hat{\sigma}/\sqrt{T}}.
$$ {#eq-HT-t-ratio}

Assuming the null hypothesis @eq-HT-H0mean is true, Proposition 7.7 tells us that @eq-HT-t-ratio is distributed Student's t with $T-1$ degrees of freedom and is denoted by the random variable $t_{T-1}$. The steps for using the z-score and the t-ratio are the same for evaluating @eq-HT-H0mean, but now we use critical values and p-values from $t_{T-1}$. Our test statistic for the two-sided alternative $H_0:\mu_i \ne \mu_i^0$ is:
$$
S = \left|t_{\mu=\mu^{0}}\right|, 
$$ {#eq-HT-testStat-abs-tratio}

and our test statistic for the one-sided alternative $H_0:\mu_i > \mu_i^0$ is 

$$
S = t_{\mu=\mu^{0}}, 
$$ {#eq-HT-testStat-tratio}

Our critical values are determined from the quantiles of the Student's t with $T-1$ degrees of freedom, $t_{T-1}(1-\alpha/2)$. For example, if $\alpha=0.05$ and $T-1=60$ then the two-sided critical value is $cv_{.05} = t_{60}(0.975)=2$ (which can be verified using the R function `qt()`). The two-sided p-value is computed as:

\begin{eqnarray*}
\textrm{p-value} & = & \Pr(|t_{T-1}|>S)=\Pr(t_{T-1}>S)+\Pr(t_{T-1}<-S)\\
 & = & 2\times \Pr(t_{T-1}>S)=2\times(1-\Pr(t_{T-1}<S)).
\end{eqnarray*}

The one-sided critical value is $cv_{.05} = t_{60}(0.95)=1.67$, and the one-sided p-value is computed as:

$$
\text{p-value} = \Pr(t_{T-1} > S) = 1 - \Pr(t_{T-1} \le S).
$$

As the sample size gets larger $\hat{\sigma}_{i}$ gets closer to $\sigma_{i}$ and the Student's t distribution gets closer to the normal distribution. Decisions using the t-ratio and the z-score are almost the same for $T \ge 60$. 

::: {#exm-testing-hypothesis-with-the-t-ratio-using-simulated-data .example name="Testing hypothesis with the t-ratio using simulated data"}

I repeat the hypothesis testing from the previous example this time use the t-ratio @eq-HT-t-ratio instead of the z-score. Using $T-1=59$ the 5\% critical value for the two-sided test is

```{r}
#| echo: true
qt(0.975, df=59)
```

To compute the t-ratios we first estimate $\mathrm{se}(\hat{\mu})$:

```{r}
sigmahat = sd(ret.sim)
sehat.muhat = sigmahat/sqrt(n.obs)
c(sehat.muhat, se.muhat)
```
Here, $\widehat{\mathrm{se}}(\hat{\mu}) = 0.0118 < 0.0129 = \mathrm{se}(\hat{\mu})$ and so the t-ratios will be slightly larger than the z-scores. The t-ratios and test statistics for the three hypotheses are:

```{r}
t.ratio.05 = (muhat - 0.05)/sehat.muhat
t.ratio.06 = (muhat - 0.06)/sehat.muhat
t.ratio.10 = (muhat - 0.10)/sehat.muhat
S.05 = abs(t.ratio.05)
S.06 = abs(t.ratio.06)
S.10 = abs(t.ratio.10)
ans = c(S.05, S.06, S.10)
names(ans) = c("S.05", "S.06", "S.07")
ans
```
As expected the test statistics computed from the t-ratios are slightly larger than the tests computed from the z-scores. The first two statistics are less than 2, and the third statistic is bigger than 2 and so we reach the same decisions as before. The p-values of the three tests are:

```{r}
S.vals = c(S.05, S.06, S.10)
p.values = 2*(1 - pt(S.vals, df=59))
p.values
```
Here, the p-values computed from $t_{59}$ are very similar to those computed from $Z\sim N(0,1)$.
\hfill{}$\blacksquare$
:::

You can derive an exact test for testing hypothesis about the value of $\mu_{i}$ based on the z-score or the t-ratio, but you cannot derive exact tests for the values of $\sigma_{i}$ or for the values of $\rho_{ij}$ based on z-scores. Exact tests for these parameters are much more complicated. While t-ratios for the values of $\sigma_{i}$ or for the values of $\rho_{ij}$ do not have exact t-distributions in finite samples, as the sample size gets large the distributions of the t-ratios get closer and closer to the normal distribution due to the CLT. This motivates the use of so-called asymptotic z-scores discussed in the next sub-section.

### Z-scores under asymptotic normality of estimators 

Let $\hat{\theta}$ denote an estimator for $\theta$. Here, I allow $\theta$ to be a GWN model parameter or a function of GWN model parameters. For example, in the GWN model $\theta$ could be $\mu_{i}$, $\sigma_{i},$, $\rho_{ij}$, $q_{\alpha}^R$, $\mathrm{VaR}_{\alpha}$ or $\mathrm{SR}_i$. As you have seen, the CLT (and the delta method if needed) justifies the asymptotic normal distribution:

$$
\hat{\theta}\sim N(\theta,\widehat{\mathrm{se}}(\hat{\theta})^{2}),
$$ {#eq-HypothesisAsymptoticNormality}

for large enough sample size $T$, where $\widehat{\mathrm{se}}(\hat{\theta})$
is the estimated standard error for $\hat{\theta}$. Consider testing:

$$
H_{0}:\theta=\theta_{0}\text{ vs. }H_{1}:\theta\neq\theta_{0}.
$$ {#eq-HypothesisTwoSided}

Under $H_{0},$ the asymptotic normality result @eq-HypothesisAsymptoticNormality
implies that the z-score for testing @eq-HypothesisTwoSided
has a standard normal distribution for large enough sample size $T$:

$$
z_{\theta=\theta_{0}}=\frac{\hat{\theta}-\theta_{0}}{\widehat{\mathrm{se}}(\hat{\theta})}\sim N(0,1)=Z.
$$ {#eq-HypothesisTstatisticZ}

The intuition for using the z-score @eq-HypothesisTstatisticZ
is straightforward. If $z_{\theta=\theta_{0}}\approx0$ then $\hat{\theta}\approx\theta_{0},$
and $H_{0}:\theta=\theta_{0}$ should not be rejected. On the other
hand, if $|z_{\theta=\theta_{0}}|>2$, say, then $\hat{\theta}$ is
more than $2$ values of $\widehat{\mathrm{se}}(\hat{\theta})$ away
from $\theta_{0}.$ This is very unlikely if $\theta=\theta_{0}$
because $\hat{\theta}\sim N(\theta_{0},\mathrm{\widehat{se}}(\hat{\theta})^{2}),$
so $H_{0}:\theta\neq\theta_{0}$ should be rejected. Therefore, the test statistic
for testing @eq-HypothesisTstatisticZ is $S = |z_{\theta=\theta_{0}}|$.

The steps for using the z-score @eq-HypothesisTstatisticZ
with its critical value to test the hypotheses @eq-HypothesisTwoSided
are: 

1.  Set the significance level $\alpha$ of the test and determine the two-sided critical value $cv_{\alpha/2}$. Using @eq-HypothesisAsymptoticNormality, the critical value, $cv_{\alpha/2},$ is determined using: 
\begin{align*}
\Pr(|Z| & \geq cv_{\alpha/2})=\alpha\\
 & \Rightarrow cv_{\alpha/2}=-q_{\alpha/2}^{Z}=q_{1-\alpha/2}^{Z},
\end{align*}
where $q_{\alpha/2}^{Z}$ denotes the $\frac{\alpha}{2}-$quantile of $N(0,1)$. A commonly used significance level is $\alpha=0.05$ and the corresponding critical value is $cv_{.025}=-q_{.025}^{Z}=q_{.975}^{Z}=1.96\approx2$.

2.  Reject @eq-HypothesisTwoSided at the $100\times\alpha$\% significance level if: 
$$
S = |z_{\theta=\theta_{0}}|=\left\vert \frac{\hat{\theta}-\theta^{0}}{\widehat{\mathrm{se}}(\hat{\theta})}\right\vert >cv_{\alpha/2}.
$$
If the significance level is $\alpha=0.05$, then reject @eq-HypothesisTwoSided at the 5\% level using the rule-of-thumb: 
$$
S=|z_{\theta=\theta_{0}}|>2.
$$

The steps for using the z-score @eq-HypothesisTstatisticZ
with its p-value to test the hypotheses @eq-HypothesisTwoSided
are:

1.  Determine the two-sided p-value. The p-value of the two-sided test is the significance level at which the test is just rejected. From @eq-HypothesisAsymptoticNormality, the two-sided p-value is defined by
$$
\textrm{p-value}=\Pr\left(|Z|>|z_{\theta=\theta_{0}}|\right)=2\times(1-\Pr\left(Z\leq|z_{\theta=\theta_{0}}|\right).
$$ {#eq-HypothesisPvalue2sided}

2.  Reject @eq-HypothesisTwoSided at the $100\times\alpha$\% significance level if the p-value @eq-HypothesisPvalue2sided is less than $\alpha$. 


::: {#exm-using-the-z-score-to-test-hypothesis-about-mu-in-the-gwn-model .example name="Using the z-score to test hypothesis about \\(\mu\\) in the GWN model"}

Consider using the z-score @eq-HypothesisTstatisticZ to
test $H_{0}:\,\mu_{i}=0\,vs.\,H_{1}:\mu_{i}\neq0$ ($i=$Microsoft,
Starbucks, S\&P 500) using a 5\% significance level. First, calculate
the GWN model estimates for $\mu_{i}$:

```{r}
#| echo: true

n.obs = nrow(gwnMonthlyRetC) 
muhat.vals = apply(gwnMonthlyRetC, 2, mean) 
muhat.vals
```

Next, calculate the estimated standard errors:

```{r}
#| echo: true

sigmahat.vals = apply(gwnMonthlyRetC, 2, sd) 
se.muhat = sigmahat.vals/sqrt(n.obs) 
se.muhat
```

Then calculate the z-scores and test statistics:

```{r}
#| echo: true

z.scores = muhat.vals/se.muhat 
S.vals = abs(z.scores)
S.vals
```

Since the absolute value of all of the z-scores are less than
two, you do not reject $H_{0}:\,\mu_{i}=0$ at the 5\% level for all
assets. 

The p-values for all of the test statistics are computed using:

```{r}
#| echo: true

2*(1-pnorm(S.vals))
```

Since all p-values are greater than $\alpha=0.05$, you reject $H_{0}:\,\mu_{i}=0$
at the 5\% level for all assets. The p-value for Starbucks is the
smallest at 0.0851. Here, you can reject $H_{0}:\,\mu_{SBUX}=0$ at
the 8.51\% level.
\hfill{}$\blacksquare$
:::

::: {#exm-using-the-z-score-to-test-hypothesis-about-rho-in-the-gwn-model .example name="Using the z-score to test hypothesis about $\\rho$ in the GWN model"}

Consider using the z-score @eq-HypothesisTstatisticZ to
test the hypotheses:

$$
H_{0}:\,\rho_{ij}=0.5\,vs.\,H_{1}:\rho_{ij}\neq0.5.
$$ {#eq-HT-H0Rho}

using a 5\% significance level. Here, you use the result from the GWN
model that for large enough $T$:
$$
\hat{\rho}_{ij}\sim N\left(\rho_{ij},\,\widehat{\mathrm{se}}(\hat{\rho}_{ij})^2\right),\,\widehat{\mathrm{se}}(\hat{\rho}_{ij})=\frac{1-\hat{\rho}_{ij}^{2}}{\sqrt{T}}.
$$
Then the z-score for testing @eq-HT-H0Rho has the form:
$$
z_{\rho_{ij}=0.5}=\frac{\hat{\rho}_{ij}-0.5}{\widehat{\mathrm{se}}(\hat{\rho}_{ij})}=\frac{\hat{\rho}_{ij}-0.5}{\left(1-\hat{\rho}_{ij}^{2}\right)/\sqrt{T}}.
$$ {#eq-HT-RhoStat}
To compute the z-scores, first, calculate the GWN model estimates
for $\rho_{ij}$:

```{r}
#| echo: true

corhat.mat = cor(gwnMonthlyRetC) 
rhohat.vals = corhat.mat[lower.tri(corhat.mat)] 
names(rhohat.vals) = c("MSFT.SBUX", "MSFT.SP500", "SBUX.SP500") 
rhohat.vals
```

Next, calculate estimated standard errors:

```{r}
#| echo: true

se.rhohat = (1 - rhohat.vals^2)/sqrt(n.obs) 
se.rhohat
```

Then calculate the z-scores @eq-HT-RhoStat and test statistics:

```{r}
#| echo: true

z.scores = (rhohat.vals - 0.5)/se.rhohat 
S.vals = abs(z.scores)
S.vals
```

Here, the absolute value of the z-scores for $\rho_{MSFT,SBUX}$
and $\rho_{MSFT,SP500}$ are greater than 2 whereas the absolute value
of the z-score for $\rho_{SBUX,SP500}$ is less than 2. Hence,
for the pairs (MSFT, SBUX) and (MSFT, SP500) you cannot reject the
null @eq-HT-H0Rho at the 5\% level but for the pair (SBUX,
SP500) you can reject the null at the 5\% level. The p-values for the
test statistics are:

```{r}
#| echo: true

2*(1-pnorm(S.vals))
```

Here, the p-values for the pairs (MSFT,SBUX) and (MSFT,SP500) are
less than 0.05 and the p-value for the pair (SBUX, SP500) is much
greater than 0.05.
\hfill{}$\blacksquare$
:::

::: {#exm-using-the-z-score-to-test-hypothesis-about-asset-sharpe-ratios-in-the-gwn-model .example name="Using the z-score to test hypothesis about asset Sharpe ratios in the GWN model"}

Consider using the z-score to test the hypotheses:

$$
H_0: \mathrm{SR}_i = \frac{\mu_i - r_f}{\sigma_i} = 0 \text{ vs. } H_1: \mathrm{SR}_i > 0
$$
using a 5\% significance level. In Chapter @sec-The-Bootstrap I used the delta method to show that, for large enough $T$:

$$
\widehat{\mathrm{SR}}_i \sim N\left(\mathrm{SR}_i, \widehat{\mathrm{se}}\left(\widehat{\mathrm{SR}}_i\right)^2\right), ~  \widehat{\mathrm{se}}\left(\widehat{\mathrm{SR}}_i\right) = \frac{1}{\sqrt{T}}\sqrt{1+\frac{1}{2}\widehat{\mathrm{SR}}_i^2}
$$
Then, the z-score has the form

$$
z_{\mathrm{SR}=0} = \frac{\widehat{\mathrm{SR}_i} - 0}{\widehat{\mathrm{se}}\left(\widehat{\mathrm{SR}}_i\right)} = \frac{\widehat{\mathrm{SR}}_i}{\frac{1}{\sqrt{T}}\sqrt{1+\frac{1}{2}\widehat{\mathrm{SR}}_i^2}} = \frac{\sqrt{T} \times \widehat{\mathrm{SR}}_i}{\sqrt{1+\frac{1}{2}\widehat{\mathrm{SR}}_i^2}}
$$

\hfill{}$\blacksquare$
:::

### Relationship between hypothesis tests and confidence intervals

Consider testing the hypotheses @eq-HypothesisTwoSided at
the 5\% significance level using the z-score @eq-HypothesisTstatisticZ. 
The rule-of-thumb decision rule is to reject $H_{0}:\,\theta=\theta_{0}$ if $|z_{\theta=\theta_{0}}|>2$.
This implies that:

\begin{eqnarray*}
\frac{\hat{\theta}-\theta_{0}}{\widehat{\mathrm{se}}(\hat{\theta})} & > & 2\,\mathrm{~ or   }\,\frac{\hat{\theta}-\theta_{0}}{\widehat{\mathrm{se}}(\hat{\theta})}<-2,
\end{eqnarray*}

which further implies that:
$$
\theta_{0}<\hat{\theta}-2\times\widehat{\mathrm{se}}(\hat{\theta})\,\mathrm{\,or}\,\,\theta_{0}>\hat{\theta}+2\times\widehat{\mathrm{se}}(\hat{\theta}).
$$
Recall the definition of an approximate 95\% confidence interval for
$\theta$:
$$
\hat{\theta}\pm2\times\widehat{\mathrm{se}}(\hat{\theta})=\left[\hat{\theta}-2\times\widehat{\mathrm{se}}(\hat{\theta}),\,\,\hat{\theta}+2\times\widehat{\mathrm{se}}(\hat{\theta})\right].
$$
Notice that if $|z_{\theta=\theta_{0}}|>2$ then $\theta_{0}$ does
not lie in the approximate 95\% confidence interval for $\theta$.
Hence, you can reject $H_{0}:\,\theta=\theta_{0}$ at the 5\% significance
level if $\theta_{0}$ does not lie in the 95\% confidence for $\theta$.
This result allows you to have a deeper understanding of the 95\% confidence
interval for $\theta$: it contains all values of $\theta_{0}$ for
which you cannot reject $H_{0}:\,\theta=\theta_{0}$ at the 5\% significance
level.


This duality between two-sided hypothesis tests and confidence intervals
makes hypothesis testing for individual parameters particularly simple.
As part of estimation, you calculate estimated standard errors and
form approximate 95\% confidence intervals. Then when you look at the approximate 95\% confidence
interval, it gives us all values of $\theta_{0}$ for which you cannot
reject $H_{0}:\,\theta=\theta_{0}$ at the 5\% significance level (approximately).

::: {#exm-hypothesis-testing-for-mu-i-rho-ij-and-sharpe-ratios-using-95-confidence-intervals .example name="Hypothesis testing for $\\mu_{i}$, $\\rho_{ij}$, and $\\mathrm{SR}_i$ using 95% confidence intervals"}

For the example data, the approximate 95\% confidence intervals for
$\mu_{i}$ for Microsoft, Starbucks and the S\&P 500 index are:

```{r}
#| echo: true

lower = muhat.vals - 2*se.muhat 
upper = muhat.vals + 2*se.muhat 
cbind(lower, upper)
```

Consider testing $H_{0}:\mu_{i}=0$ at the 5\% level. Here you see
that $\mu_{i}^{0}=0$ lies in all of the 95\% confidence intervals
and so you do not reject $H_{0}:\mu_{i}=0$ at the 5\% level for any
asset. Each interval gives the values of $\mu_{i}^{0}$ for which
you cannot reject $H_{0}:\mu_{i}=\mu_{i}^{0}$ at the 5\% level. For
example, for Microsoft you cannot reject the hypothesis (at the 5\%
level) that $\mu_{MSFT}$ is as small as -0.011 or as large as 0.019.


Next, the approximate 95\% confidence intervals for $\rho_{ij}$ for
the pairs $\rho_{MSFT,SBUX}$, $\rho_{MSFT,SP500}$, and $\rho_{SBUX,SP500}$
are:

```{r}
#| echo: true

lower = rhohat.vals - 2*se.rhohat 
upper = rhohat.vals + 2*se.rhohat 
cbind(lower, upper)
```

Consider testing $H_{0}:\rho_{ij}=0.5$ for all pairs of assets. Here,
you see that $\rho_{ij}^{0}=0.5$ is not in the approximate 95\% confidence
intervals for $\rho_{MSFT,SBUX}$ and $\rho_{MSFT,SP500}$ but is in
the approximate 95\% confidence interval for $\rho_{SBUX,SP500}$.
Hence, you can reject $H_{0}:\rho_{ij}=0.5$ at the 5\% significance
level for $\rho_{MSFT,SBUX}$ and $\rho_{MSFT,SP500}$ but not for
$\rho_{SBUX,SP500}$. 

* Add SR info here

\hfill{}$\blacksquare$
:::

## Coefficient tests between model coefficients for two assets

In this section I consider hypothesis tests for coefficients in the bivariate GWN model for asset returns.

### Test for equal means between two assets

Often it is of interest to test if the expected returns on two assets are the same:

$$
H_{0}:\mu_{1}=\mu_{2}\,\textrm{vs. } H_{1}:\mu_{1}\neq\mu_{2}.
$$

Notice that this hypothesis is not of the form that the two coefficients equal a specific value (e.g. $\mu_1 = \mu_2 = 0.05$). However, you can transform the hypothesis into one in which a coefficient equals a specific value by considering the differences in the expected returns. In particular, the relation $\mu_1 = \mu_2$ is equivalent to the relation $\delta = \mu_1 - \mu_2 = 0$. Hence, you can reformulate the hypothesis of equal expected returns to the hypothesis that the difference in expected returns is equal to zero:

$$
H_{0}:\delta = \mu_{1}-\mu_{2}=0\,\textrm{vs. } H_{1}:\delta = \mu_{1}-\mu_{2} \neq 0. 
$$ {#eq-HT-equalmeans}

Now, you have test in the form of the coefficient tests you considered earlier. 

The parameter $\delta = \mu_1 - \mu_2$ can be estimated with the plug-in estimator:

$$
\hat{\delta} = \hat{\mu}_1 - \hat{\mu}_2.
$$

To form a z-score, you require an estimate of $\mathrm{se}(\hat{\delta})$. Now,

$$
\mathrm{var}(\hat{\delta}) = \mathrm{var}(\hat{\mu}_1-\hat{\mu}_2)
= \mathrm{var}(\hat{\mu}_1) + \mathrm{var}(\hat{\mu}_2) - 2\mathrm{cov}(\hat{\mu}_1,\hat{\mu}_2). 
$$ {#eq-HT-vardeltahat}

To evaluate @eq-HT-vardeltahat, you use the result from Chapter 7 that the joint distribution of the vector $(\hat{\mu}_1,\hat{\mu}_2)'$ is multivariate normal:
$$
\left( \begin{array}{c} \hat{\mu}_1 \\ \hat{\mu}_2 \end{array} \right) = N \left( \left( \begin{array}{c} \mu_1 \\ \mu_2 \end{array} \right), ~ \frac{1}{T} \left( \begin{array}{cc} \hat{\sigma}_1^2 & \hat{\sigma}_{12} \\ 
\hat{\sigma}_{12} & \hat{\sigma}_2^2 \end{array} \right) \right),
$$

for large enough $T$. Then

$$
\widehat{\mathrm{var}}(\hat{\delta}) = \frac{1}{T}(\hat{\sigma}_1^2 + \hat{\sigma}_2^2 - 2\hat{\sigma}_{12}),
$$

and 

$$
\widehat{\mathrm{se}}(\hat{\delta}) = \sqrt{\frac{1}{T}(\hat{\sigma}_1^2 + \hat{\sigma}_2^2 - 2\hat{\sigma}_{12})}.
$$

you can test @eq-HT-equalmeans using the z-score:

$$
z_{\delta=0} = \frac{\hat{\delta}}{\widehat{\mathrm{se}}(\hat{\delta})} = \frac{\hat{\mu}_1 - \hat{\mu}_2}{\sqrt{\frac{1}{T}(\hat{\sigma}_1^2 + \hat{\sigma}_2^2 - 2\hat{\sigma}_{12})}}, 
$$ {#eq-HT-zscoredeltahat}

with the test statistic $S=\left| z_{\delta=0} \right|$. 

The estimate $\widehat{\mathrm{var}}(\hat{\delta})$ can also be computed using the delta method. Define $\theta = (\mu_1,\mu_2)'$ and $f(\theta)=\delta = \mu_1 - \mu_2$. The gradient of $f(\theta)$ is: 

$$
g(\theta) = \left(\begin{array}{c} 1 \\ 1 \end{array}  \right),
$$

and $\widehat{\mathrm{var}}(\hat{\delta})$ can be computed as:

\begin{align*}
\widehat{\mathrm{var}}(\hat{\delta}) &= g(\hat{\theta})'\widehat{\mathrm{var}}(\hat{\theta})g(\hat{\theta})
 = (1, -1)' \left( \begin{array}{cc} \frac{\hat{\sigma}_1^2}{T} & \frac{\hat{\sigma}_{12}}{T} \\ \frac{\hat{\sigma}_{12}}{T} & \frac{\hat{\sigma}_2^2}{T} \end{array} \right) \left( \begin{array}{c} 1 \\ -1 \end{array}   \right) \\
&= \frac{1}{T}(\hat{\sigma}_1^2 + \hat{\sigma}_2^2 - 2\hat{\sigma}_{12}).
\end{align*}

::: {#exm-testing-that-microsoft-and-starbucks-have-the-same-expected-return .example name="Testing that Microsoft and Starbucks have the same expected return."}

Let's use the z-score @eq-HT-zscoredeltahat to test the hypothesis that the expected returns on Microsoft and Starbucks are the same using a 5\% significance level. The R code to compute the z-score and test statistic is:

```{r}
delta.hat = muhat.vals["MSFT"] - muhat.vals["SBUX"]
sig2hat.vals = sigmahat.vals^2
covhat = cov(gwnMonthlyRetC[, c("MSFT", "SBUX")])[1,2]
sehat.deltahat = sqrt((sig2hat.vals["MSFT"]+sig2hat.vals["SBUX"]-2*covhat)/n.obs)
z.score.delta = delta.hat/sehat.deltahat
S= abs(z.score.delta)
ans = c(delta.hat, sehat.deltahat, S)
names(ans) = c("Estimate", "Std Error", "Test Statistic")
ans
                      
```
Here, $\hat{\delta}=-0.0105$, $\widehat{\mathrm{se}}(\hat{\delta})=0.0093$ and $S=1.1323$. Since $S < 2$ you do not reject $H_0:\delta = \mu_{msft} - \mu_{sbux}=0$ at the 5\% level. 
\hfill{}$\blacksquare$
:::

### Test for equal Sharpe ratios between two assets

In the previous section, I considered testing if two assets have the same expected return. It is important to recognize that even if two assets have the same mean return they may have different standard deviations (volatilities) and, hence, different risks. Often, it is more interesting to test the hypothesis that two assets have the same reward-to-risk (i.e., Sharpe) ratios:

$$
H_{0}:\mathrm{SR}_1 = \frac{\mu_{1}-r_f}{\sigma_1}=\mathrm{SR}_2 = \frac{\mu_{2}-r_f}{\sigma_2}\,\textrm{vs. } H_{1}:\mathrm{SR}_1\neq\mathrm{SR}_2.
$$

where $r_f$ denotes the risk-free rate over the investment horizon. As with testing the equality between two means, it is convenient to re-formulate the hypothesis of equal Sharpe ratio to the hypothesis that the difference in the Sharpe ratios is equal to zero:

$$
H_{0}: \delta = \mathrm{SR}_1 - \mathrm{SR}_2 = 0\,\textrm{vs. } H_{1}: \delta = \mathrm{SR}_1 - \mathrm{SR}_2 \ne 0. 
$$ {#eq-HT-equalSR}

The plug-in estimator for $\delta$ is:

$$
\hat{\delta} = \widehat{\mathrm{SR}}_1 - \widehat{\mathrm{SR}}_2 = \frac{\hat{\mu}_1-r_f}{\hat{\sigma}_1} - \frac{\hat{\mu}_2-r_f}{\hat{\sigma}_2}
$$

To determine $\widehat{\mathrm{se}}(\hat{\delta})$ from @eq-HT-vardeltahat, I utilize the result from Chapter @sec-The-Bootstrap that the joint distribution of the vector $(\widehat{\mathrm{SR}}_1, \widehat{\mathrm{SR}}_2)$ is multivariate normal:

$$
\begin{pmatrix} \widehat{\mathrm{SR}}_1 \\ \widehat{\mathrm{SR}}_2 \end{pmatrix} \sim N \left( \begin{pmatrix} \mathrm{SR}_1 \\ \mathrm{SR}_1 \end{pmatrix}, ~ \begin{pmatrix} 
\frac{1}{T}(1+\frac{1}{2}\widehat{\mathrm{SR}}_1^2) & \frac{\hat{\rho}_{12}}{T}(1 + \frac{1}{2}\hat{\rho}_{12}\widehat{\mathrm{SR}}_1 \widehat{\mathrm{SR}}_2) \\ \frac{\hat{\rho}_{12}}{T}(1 + \frac{1}{2}\hat{\rho}_{12} \widehat{\mathrm{SR}}_1\widehat{\mathrm{SR}}_2) & \frac{1}{T}(1+\frac{1}{2}\widehat{\mathrm{SR}}_2^2)
\end{pmatrix} 
\right ),
$$

for large enough $T$. Then,

\begin{align*}
\widehat{\mathrm{var}}(\hat{\delta}) &= \widehat{\mathrm{var}}(\widehat{\mathrm{SR}}_1) + \widehat{\mathrm{var}}(\widehat{\mathrm{SR}}_2) - 2\widehat{\mathrm{cov}}(\widehat{\mathrm{SR}}_1,\widehat{\mathrm{SR}}_2) \\
&= \frac{1}{T}(1+\frac{1}{2}\widehat{\mathrm{SR}}_1^2) + \frac{1}{T}(1+\frac{1}{2}\widehat{\mathrm{SR}}_2^2) \\
&- 2\frac{\hat{\rho}_{12}}{T}(1 + \frac{1}{2}\hat{\rho}_{12}\widehat{\mathrm{SR}}_1\widehat{\mathrm{SR}}_2 )\\
&= \frac{1}{T}\left(2 + \frac{1}{2}(\widehat{\mathrm{SR}}_1^2 + \widehat{\mathrm{SR}}_2^2) - 2\hat{\rho}_{12}(1 +  \frac{1}{2}\hat{\rho}_{12}\widehat{\mathrm{SR}}_1\widehat{\mathrm{SR}}_2)    \right),
\end{align*}

and,

$$
\widehat{\mathrm{se}}(\hat{\delta}) = \sqrt{\frac{1}{T}\left(2 + \frac{1}{2}(\widehat{\mathrm{SR}}_1^2 + \widehat{\mathrm{SR}}_2^2) - 2\hat{\rho}_{12}(1 +  \frac{1}{2}\hat{\rho}_{12}\widehat{\mathrm{SR}}_1\widehat{\mathrm{SR}}_2)    \right)}.
$$

You can test @eq-HT-equalSR using the z-score:

$$
z_{\delta=0} = \frac{\hat{\delta}}{\widehat{\mathrm{se}}(\hat{\delta})} = \frac{\widehat{\mathrm{SR}}_1 - \widehat{\mathrm{SR}}_2}{\sqrt{\frac{1}{T}\left(2 + \frac{1}{2}(\widehat{\mathrm{SR}}_1^2 + \widehat{\mathrm{SR}}_2^2) - 2\hat{\rho}_{12}(1 +  \frac{1}{2}\hat{\rho}_{12}\widehat{\mathrm{SR}}_1\widehat{\mathrm{SR}}_2)    \right)}}, 
$$ {#eq-HT-zscoredeltahatSR}

with the test statistic $S=\left| z_{\delta=0} \right|$.

::: {#exm-testing-equal-sharpe-ratios-between-microsoft-and-starbucks .example name="Testing equal Sharpe ratios between Microsoft and Starbucks"}

Let's use the z-score @eq-HT-zscoredeltahatSR to test the hypothesis of equal Sharpe ratios between Microsoft and Starbucks using a 5\% significance level. The R code to compute the z-score and test statistic is:

```{r}
r.f = 0.03/12
SR.msft = (muhat.vals["MSFT"]-r.f)/sigmahat.vals["MSFT"]
SR.sbux = (muhat.vals["SBUX"]-r.f)/sigmahat.vals["SBUX"]
corhat = cor(gwnMonthlyRetC[, c("MSFT", "SBUX")])[1,2]
delta.hat = SR.msft - SR.sbux
var.deltahat = (2 + 0.5*(SR.msft^2 + SR.sbux^2)
                -2*corhat*(1+0.5*SR.msft*SR.sbux))
se.deltahat = sqrt(var.deltahat)
z.score.delta = delta.hat/se.deltahat
S = abs(z.score.delta)
ans = c(SR.msft, SR.sbux, delta.hat, se.deltahat, S)
names(ans) = c("SR.msft", "S.sbux", "deltahat", "se.deltahat", "S")
ans
```
Although the Sharpe ratio for Starbucks is much larger than the Sharpe ratio for Microsoft, there is substantial estimation error in $\hat{\delta}=\mathrm{SR}_{msft} - \mathrm{SR}_{sbux}$ which produces a small test statistic $S=0.0805$. Since $S < 2$ you do not reject the null hypothesis of equal Sharpe ratios at the 5\% level of significance. 
\hfill{}$\blacksquare$
:::

## Wald tests for general linear hypotheses (advanced)

* testing equal means across many assets

* testing equal Sharpe ratios across many assets (mention result from Jobson and Korke etc here)

## Model Specification Tests

-  Review GWN model assumptions
-  Discuss why it is interesting and useful to test these assumptions
-  What to do if certain assumptions are violated?


### Tests for normality

-  Why do you care about normality? 
    -  Return distribution only depends on mean and volatility (and correlation in multivariate case). 
    -  Justifies volatility as an appropriate measure of risk. 
    -  Gives formula for Value-at-risk calculations. 
    -  Justifies mean-variance portfolio theory and Black-Scholes option pricing formula. 
    -  GWN model parameter estimates and standard error formula do not require the data to be normally distributed. These estimates are consistent and asymptotically normally distributed for many non-normal distributions. There may be estimators that are more accurate (i.e., have smaller standard errors)
-  What to do if returns are found to not be normally distributed? 

    -  Return distribution characteristics in addition to mean and volatility may be important to investors. Volatility may not be the best measure of asset return risk. For example, skewness and kurtosis. If many returns are non-normal then this casts doubt on the appropriateness of mean-variance portfolio theory.
    -  Standard practice is to compute VaR based on a normal distribution. If returns are not normally distributed this is an incorrect way to compute VaR. Instead of using the normal distribution quantile to compute VaR you can use the empirical quantile to compute VaR. If returns are substantially non-normal in the tails of the distribution then the empirical quantile will be more appropriate than the normal quantile for computing VaR
    -  How are returns non-normal? This is important. 


In chapter @sec-Descriptive-Statistics-for-Financial-Data,
I used histograms, normal QQ-plots, and boxplots to assess whether
the normal distribution is a good characterization of the empirical
distribution of daily and monthly returns. For example, Figure @fig-HypothesisFourPanelMSFT
shows the four-panel plot for the monthly returns on Microsoft that
I used to summarize the empirical distribution of returns. The histogram
and smoothed histogram show that the empirical distribution of returns
is bell-shaped like the normal distribution, whereas the boxplot and
normal QQ-plot shows that returns have fatter tails than the normal
distribution. Overall, the graphical descriptive statistics suggest
that distribution of monthly Microsoft returns is not normally distributed.
However, graphical diagnostics are not formal statistical tests. There
are estimation errors in the statistics underlying the graphical diagnostics
and it is possible that the observed non-normality of the data is
due to sampling uncertainty. To be more definitive, I should use
an appropriate test statistic to test the null hypothesis that the
random return, $R_{t}$, is normally distributed. Formally, the null
hypothesis to be tested is: 

$$
H_{0}:R_{t}\, \textrm{is normally distributed}
$$ {#eq-HT-H0normal}

The alternative hypothesis is that $R_{t}$ is not normally distributed:

$$
H_{1}:R_{t}\,\textrm{is not normally distributed}
$$ {#eq-HT-H1nonnormal}

Here, I do not specify a particular distribution under the alternative
(e.g., a Student's t distribution with a 5 degrees of freedom). The
goal is to have a test statistic that will reject the null hypothesis
@eq-HT-H0normal for a wide range of non-normal distributions.

```{r}
#| label: fig-HypothesisFourPanelMSFT
#| echo: false
#| fig-cap: Four panel distribution plot for Microsoft monthly returns.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

four_panel_plot(gwnMonthlyRetC[, "MSFT"])
```


#### JB test based on skewness and kurtosis
\index{JB test}
```{r}
#| label: fig-HypothesisChiSq2
#| echo: false
#| fig-cap: pdf of $\chi^{2}(2)$
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.vals = seq(0,10,length.out=100) 
plot(x.vals, dchisq(x.vals, 2), type="l", lwd=2, col="blue", ylab="pdf", xlab="X")
```


In Chapter @sec-Review-of-Random-Variables, I showed that
the skewness and kurtosis of a normally distributed random variable
are equal to zero and three, respectively. I can use this information
to develop a simple test of @eq-HT-H0normal against @eq-HT-H1nonnormal
based on testing the hypotheses:
$$
H_{0}:\mathrm{skew}(R_{t}) = 0 ~ \mathrm{and} ~ \mathrm{kurt}(R_{t})=3
$$ {#eq-HypothesisH0normal1}
versus 
$$
H_{1}:\mathrm{skew}(R_{t}) \neq 0 ~ \mathrm{or} ~ \mathrm{kurt}(R_{t})\neq3 ~ \mathrm{or both}.
$$ {#eq-HypothesisH1normal1}

To develop a test statistic for testing @eq-HypothesisH0normal1
against @eq-HypothesisH1normal1, Jarque and Bera (1981) used
the result (see Kendall and Stuart, 1977) that if $\{R_{t}\}\sim\mathrm{iid}\,N(\mu,\sigma^{2})$
then by the CLT it can be shown that: 
$$
\left(\begin{array}{c}
\widehat{\mathrm{skew}}_{r}\\
\widehat{\mathrm{kurt}}_{r}-3
\end{array}\right)\sim N\left(\left(\begin{array}{c}
0\\
0
\end{array}\right),\left(\begin{array}{cc}
\frac{6}{T} & 0\\
0 & \frac{24}{T}
\end{array}\right)\right),
$$
for large enough $T$, where $\widehat{\mathrm{skew}}_{r}$ and $\widehat{\mathrm{kurt}}_{r}$
denote the sample skewness and kurtosis of the returns $\{r_{t}\}_{t=1}^{T}$,
respectively. This motivates Jarque and Bera's *JB statistic*:
$$
\mathrm{JB}=T\times\left[\frac{\left(\widehat{\mathrm{ske}w}_{r}\right)^{2}}{6}+\frac{\left(\widehat{\mathrm{kurt}}_{r}-3\right)^{2}}{24}\right],
$$ {#eq-HypothesisJB}
to test @eq-HT-H0normal against @eq-HT-H1nonnormal.
If returns are normally distributed then $\widehat{\mathrm{skew}}_{r}\approx0$
and $\widehat{\mathrm{kurt}}_{r}-3\approx0$ so that $\mathrm{JB}\approx0$.
In contrast, if returns are not normally distributed then $\widehat{\mathrm{skew}}_{r}\neq0$
or $\widehat{\mathrm{kurt}}_{r}-3\neq0$ or both so that $\mathrm{JB}>0$.
Hence, if JB is big then there is data evidence against the null hypothesis
that returns are normally distributed. 


To determine the critical value for testing @eq-HT-H0normal
at a given significance level I need the distribution of the JB statistic
@eq-HypothesisJB under the null hypothesis @eq-HT-H0normal.
To this end, Jarque and Bera show, assuming @eq-HT-H0normal
and large enough $T$, that:
$$
\mathrm{JB}\sim\chi^{2}(2),
$$
where $\chi^{2}(2)$ denotes a chi-square distribution with two degrees
of freedom (see Appendix). Hence, the critical values for the JB statistic
are the upper quantiles of $\chi^{2}(2)$. Figure @fig-HypothesisChiSq2
shows the pdf of $\chi^{2}(2)$. The 90\%, 95\% and 99\% quantiles
can be computed using the R function `qchisq()`:

```{r}
#| echo: true

 qchisq(c(0.90, 0.95, 0.99), 2)
```

Since the 95\% quantile of $\chi^{2}(2)$ is $5.99\approx6$, a simple
rule of thumb is to reject the null hypothesis @eq-HT-H0normal
at the 5\% significance level if the JB statistic @eq-HypothesisJB
is greater than 6.


Let $X\sim\chi^{2}(2)$. The p-value of the JB test is computed using:
$$
p-value=\Pr(X>JB).
$$
If the p-value() is less than $\alpha$ then the null () is rejected
at the 

::: {#exm-using-the-jb-statistic-to-test-the-normality-assumption-for-returns-on-microsoft-starbucks-and-the-s-p-500-index .example name="Using the JB statistic to test the normality assumption for returns  on Microsoft Starbucks and the S&P 500 index"}

The sample skewness and excess kurtosis values for the monthly returns
on Microsoft, Starbucks, and the S\&P 500 index are

```{r}
#| echo: true

skewhat = skewness(gwnMonthlyRetC) 
ekurthat = kurtosis(gwnMonthlyRetC) 
rbind(skewhat, ekurthat)
```

The sample skewness values for the assets are not too large but the
sample excess kurtosis values are fairly large. The JB statistics
for the three assets are easy to compute

```{r}
#| echo: true

n.obs = nrow(gwnMonthlyRetC) 
JB = n.obs*((as.numeric(skewhat)^2)/6 + (as.numeric(ekurthat)^2)/24) 
names(JB) = colnames(gwnMonthlyRetC) 
JB
```

Since all JB values are greater than 6, you reject the null hypothesis
that the monthly returns are normally distributed.

You can also use the **tseries** function `jarque.bera.test()`
to compute the JB statistic:

```{r}
#| echo: true

jarque.bera.test(gwnMonthlyRetC[, 1])
jarque.bera.test(gwnMonthlyRetC[, 2])
jarque.bera.test(gwnMonthlyRetC[, 3])
```

The test statistics are slightly different from the direct calculation
above due to differences in how the sample skewness and sample kurtosis
is computed within the function `jarque.bera.test()`.
\hfill{}$\blacksquare$
:::

#### Shipiro Wilks test
\index{Shipiro Wilks test}
To be complete...

### Tests for no autocorrelation
\index{Test for autocorrelation}
```{r}
#| label: fig-HypothesisSACFsp500
#| echo: false
#| fig-cap: SACF for monthly returns on the S\&P 500 index.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

acf(gwnMonthlyRetC[, "SP500"], main="", lwd=2)
```


A key assumption of the GWN model is that returns are uncorrelated
over time (not autocorrelated). In efficient financial markets where
prices fully reflect true asset value, returns of actively traded
assets should not be autocorrelated. In addition, if returns are autocorrelated
then they are predictable. That is, future returns can be predicted
from the past history of returns. If this is the case then trading
strategies can be easily developed to exploit this predictability.
But if many market participants act on these trading strategies then
prices will adjust to eliminate any predictability in returns. Hence,
for actively traded assets in well functioning markets I expect to
see that the returns on these assets are uncorrelated over time. 


For less frequently traded assets returns may appear to be autocorrelated
due to stale pricing. 


As you saw in Chapter @sec-The-Gaussian-White-Noise-Model, this allows
us to aggregate the GWN model over time and, in particular, derive
the square-root-of time rule for aggregating return volatility. If
returns are substantially correlated over time (i.e., autocorrelated)
then the square-root-of-time rule will not be an accurate way to aggregate
volatility. 


To measure autocorrelation in asset returns $\{R_{t}$\} I use the
j$^{th}$ lag autocorrelation coefficient 
\begin{align*}
\rho_{j} & =\mathrm{cor}(R_{t},R_{t-j})=\frac{\mathrm{cov}(R_{t},R_{t-j})}{\mathrm{var}(R_{t})},
\end{align*}
which is estimated using the sample autocorrelation 
$$
\hat{\rho}_{j}=\frac{\frac{1}{T-1}\sum_{t=j+1}^{T}(r_{t}-\hat{\mu})(r_{t-j}-\hat{\mu})}{\frac{1}{T-1}\sum_{t=1}^{T}(r_{t}-\hat{\mu})^{2}}.
$$
In chapter @sec-Descriptive-Statistics-for-Financial-Data,
I used the sample autocorrelation function (SACF) to graphically
evaluate the data evidence on autocorrelation for daily and monthly
returns. For example, Figure @fig-HypothesisSACFsp500 shows
the SACF for the monthly returns on the S\&P 500 index. Here you see
that none of the $\hat{\rho}_{j}$values are very large. You also see
a pair of blue dotted lines on the graph. As you shall see, these dotted
lines are related to critical values for testing the null hypothesis
that $\rho_{j}=0$. Hence, if 

#### Individual tests

Individual tests for autocorrelation are similar to coefficient tests
discussed earlier and are based on testing the hypotheses
$$
H_{0}:\rho_{j}=0 ~ vs. ~ H_{1}:\rho_{j}\neq0,
$$
for a given lag $j$. The tests rely on the result, justified by
the CLT if $T$ is large enough, that if returns are covariance stationary
and uncorrelated then 
\begin{align*}
\hat{\rho}_{j}\sim N\left(0,\frac{1}{T}\right)\text{ for all }j\geq1\,\mathrm{and}\,\mathrm{se}(\hat{\rho}_{j})=\frac{1}{\sqrt{T}}.
\end{align*}
This motivates the test statistics
$$
t_{\rho_{j=0}}=\frac{\hat{\rho}_{j}}{\mathrm{se}(\hat{\rho}_{j})}=\frac{\hat{\rho}_{j}}{1/\sqrt{T}}=\sqrt{T}\hat{\rho}_{j},\,j\geq1
$$
and the 95\% confidence intervals
$$
\hat{\rho}_{j}\pm2\cdot\frac{1}{\sqrt{T}}.
$$
If you use a 5\% significance level, then you reject the null hypothesis
$H_{0}:\rho_{j}=0$ if 
$$
|t_{\rho_{j=0}}|=\left\vert \sqrt{T}\hat{\rho}_{j}\right\vert >2.
$$
That is, reject if 
$$
\hat{\rho}_{j}>\frac{2}{\sqrt{T}}\text{ or }\hat{\rho}_{j}<\frac{-2}{\sqrt{T}}.
$$
Here, note that the dotted lines on the SACF are at the points $\pm2\cdot\frac{1}{\sqrt{T}}$.
If the sample autocorrelation $\hat{\rho}_{j}$ crosses either above
or below the dotted lines then you can reject the null hypothesis $H_{0}:\rho_{j}=0$
at the 5\% level. 

#### Joint tests

-  Have to be careful how to interpret the individual tests. Cannot simply use 20 individual tests at the 5\% level to test a joint hypothesis at the 5\% level. With independent individual tests each test has a 5\% probability of rejecting the null when it is true. So if you do 20 individual tests it would not be surprising if one tests rejects and this could be purely by chance (a type I error) if the null of no autocorrelation was true. 

To be complete

### Tests for constant parameters
\index{Test for constant parameters}
A key assumption of the GWN model is that returns are generated from
a covariance stationary time series process. As a result, all of the
parameters of the GWN model are constant over time. However, as shown
in Chapter @sec-Descriptive-Statistics-for-Financial-Data  there is evidence to suggest that the assumption of
covariance stationarity may not always be appropriate. For example,
the volatility of returns for some assets does not appear to be constant
over time. 

A natural null hypothesis associated with the covariance stationarity
assumption for the GWN model for an individual asset return has the
form:
$$
H_{0}:\mu_{i},\sigma_{i},\rho_{ij}\, \text{are constant over time}.
$$
To reject this null implied by covariance stationarity, one only needs
to find evidence that at least one parameter is not constant over
time.

#### Formal tests (two parameter structural change)
\index{Formal tests}

To be completed

* Do two-sample structural change tests based on known segments. Analysis is very similar to testing equality of parameters across two assets. 

### Tests for constant parameters based on rolling estimators
\index{Rolling-tests}

In Chapter @sec-Descriptive-Statistics-for-Financial-Data I introduced rolling descriptive statistics to informally investigate the assumption of covariance stationarity. Given any parameter $\theta$ of the GWN model (or any function of the parameters), you can compute a set of rolling estimates of $\theta$ over windows of length $n<T$: $\{\hat{\theta}_t(n) \}_{t=n}^{T}$. A time plot of these rolling estimates may show evidence of non-stationary behavior in returns.  

However, you must always keep in mind that estimates have estimation error and part of the observed time variation in the rolling estimates $\{\hat{\theta}_t(n) \}_{t=n}^{T}$ is due to random estimation
error. To account for this estimation error, rolling estimates are often displayed with estimated standard error bands of the form:

$$
\hat{\theta}_t(n) \pm c\times \widehat{\mathrm{se}}\left( \hat{\theta}_t(n)\right), 
$$ {#eq-HT-rolling}

where $c$ is a constant representing some multiple of the estimated standard error. Typically, $c$ is chosen so that @eq-HT-rolling represents an approximate confidence interval for $\theta$. For example, setting $c=2$ gives an approximate 95\% confidence interval.

::: {#exm-01 .example}

As in Chapter @sec-Descriptive-Statistics-for-Financial-Data, you can compute 24-month rolling estimates of $\mu$ and $\sigma$ for Microsoft, Starbucks and the S\&P 500 index using the **zoo** function `rollapply()`:

```{r}
roll.muhat = rollapply(gwnMonthlyRetC, width = 24, by = 1, 
                       by.column = TRUE, FUN = mean, align = "right")
roll.sigmahat = rollapply(gwnMonthlyRetC, width = 24, by = 1, 
                          by.column = TRUE, FUN = sd, align = "right")
```

Estimated standard errors for $\hat{\mu}_t(n)$ and $\hat{\sigma}_t(n)$ are computed using:
$$
\widehat{\mathrm{se}}(\hat{\mu}_t(n)) = \frac{\hat{\sigma}_t(n)}{\sqrt{T}}, ~ \widehat{\mathrm{se}}(\hat{\sigma}_t(n)) = \frac{\hat{\sigma}_t(n)}{\sqrt{2T}}
$$
Figure @fig-HTrolling shows the rolling estimates of $\mu$ (top panel) and $\sigma$ (bottom panel) for Starbucks together with 95% confidence intervals ($\pm2\times$ standard error bands) computed using:

```{r}
#| label: fig-HTrolling
#| echo: true
#| fig-cap: 24-month rolling estimates of $\mu$ and $\sigma$ for Starbucks with estimated
#|   standard error bands
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
se.muhat.SBUX = roll.sigmahat[, "SBUX"]/sqrt(24)
se.sigmahat.SBUX = roll.sigmahat[, "SBUX"]/sqrt(2 * 24)
lower.muhat.SBUX = roll.muhat[, "SBUX"] - 2 * se.muhat.SBUX
upper.muhat.SBUX = roll.muhat[, "SBUX"] + 2 * se.muhat.SBUX
lower.sigmahat.SBUX = roll.sigmahat[, "SBUX"] - 2 * se.sigmahat.SBUX
upper.sigmahat.SBUX = roll.sigmahat[, "SBUX"] + 2 * se.sigmahat.SBUX
# plot estimates with standard error bands
dataToPlot = merge(roll.muhat[, "SBUX"], lower.muhat.SBUX, 
                   upper.muhat.SBUX, roll.sigmahat[,"SBUX"],
                   lower.sigmahat.SBUX, upper.sigmahat.SBUX)
my.panel <- function(...) {
  lines(...)
  abline(h = 0)
}
plot.zoo(dataToPlot, main = "", screens = c(1, 1, 1, 2, 2, 2),
         lwd = 2, col = c("black","red", "red", "black", "red", "red"),
         lty = c("solid", "dashed", "dashed", "solid", "dashed", "dashed"),
         panel = my.panel, ylab = list("muhat", "sigmahat"))
```

The standard error bands for $\hat{\mu}_t(n)$ are fairly wide, especially during the high volatility periods of the dot-com boom-bust and the financial crisis. During the dot-com period the 95% confidence intervals for $\mu$ are wide and contain both positive and negative values reflecting a high degree of uncertainty about the value of $\mu$. However, during the economic boom around 2005, where $\hat{\mu}_t(n)$ is large and $\hat{\sigma}_t(n)$ is small, the 95% confidence interval for $\mu$ does not contain zero and one can infer that the mean is significantly positive. Also, at the height of the financial crisis, where $\hat{\mu}_t(n)$ is the largest negative value, the 95% confidence interval
for $\mu$ contains all negative values and one can infer that the mean is significantly negative. At the end of the sample, the 95% confidence intervals for $\mu$ contain all positive values.^[You have to be a little careful about interpreting the implicit hypotheses tests underlying the rolling 95\% confidence intervals. Each confidence interval represents a single hypothesis test with a 5\% significane level. However, we have many confidence intervals and so we are really doing many hypothesis tests. The joint significance level for all of these tests will be larger than the significance level for a single test.]

The standard error bands for $\hat{\sigma}_t(n)$ are smaller than the bands for $\hat{\mu}_t(n)$ and one can see more clearly that $\hat{\sigma}_t(n)$ significantly larger during the dot-com period and the financial crisis period than during the economic boom period and the period after the financial crisis.
\hfill{}$\blacksquare$
:::

Plotting rolling estimates with estimated standard error bands is good practice as it reminds you that there is estimation error in the rolling estimates and part of any observed time variation is this estimation error. 

## Using Monte Carlo Simulation to Understand Hypothesis Testing
\index{Monte Carlo Simulation}
In chapter @sec-Estimation-of-The-GWN-Model I used Monte Carlo
simulation to understand the statistical properties of estimators.
Now, I will use Monte Carlo Simulation to understand hypothesis testing.
To fix ideas, let $R_{t}$ be the return on a single asset described
by the GWN model, let $\theta$ denote some parameter of the GWN model
you are interested in testing an hypothesis about, let $\hat{\theta}$
denote an estimator for $\theta$ based on a sample of size $T$,
and let $\widehat{\mathrm{se}}(\hat{\theta)}$ denote the estimated
standard error for $\hat{\theta}$. Suppose you are interested in testing
$H_{0}:\theta=\theta_{0}$ against $H_{1}:\theta\neq\theta_{0}$ using
the z-score $z_{\theta=\theta_{0}}=\frac{\hat{\theta}-\theta_{0}}{\widehat{\mathrm{se}}(\hat{\theta)}}$.

To be completed

## Using the Bootstrap for Hypothesis Testing

To be completed

* duality between HT and CIs can be exploited by the bootstrap. Use bootstrap to compute CI and see if hypothesized values lies in the interval. Illustrate with testing equality of 2 Sharpe ratios

## Further Reading: Hypothesis Testing in the GWN Model

-  Mention books by Carol Alexander for discussion of rolling and EWMA
estimates
-  Formal tests for structural change require advanced probability and
statistics. See [@zivot-2016] for discussion. Many tests are implemented
in the **strucchange** package.

## Exercises

To be completed...
