# Introduction to Portfolio Theory {#sec-Introduction-to-Portfolio-Theory}

Updated: September 10, 2026
Copyright © Eric Zivot 2026


Chapters @sec-Estimation-of-The-GWN-Model through @sec-Hypothesis-Testing
focused on the development and estimation of, and the statistical
inference for, the GWN model for asset returns. In this chapter, I
use the GWN model for asset returns as the basis for the quantitative
analysis of portfolios and asset allocation. Specifically, I introduce
the mean-variance portfolio analysis framework originally developed
by Harry Markowitz.^[The mean-variance frame is described in detail in [@markowitz-1959], 
[@markowitz-1987]. Harry Markowitz was awarded the Nobel Prize in Economics in
1990, along with Merton Miller and William Sharpe, for his work on
portfolio theory.] This framework is the widely used in academia and industry and is
covered in most standard textbooks on investment analysis such as
[@bkm-2013], [@egbg-2014], and [@sharpe-1999]. In this
framework, asset returns are assumed to be normally distributed and
so investor preferences over portfolios of assets depend only on the
expected return and variance of the portfolio returns. Investors are
assumed to like portfolio with high mean returns but dislike portfolio
with high return variances. Since portfolios that have high (low)
expected returns tend to have high (low) variances investors face
a mean-variance tradeoff between different portfolios. Markowitz's
mean-variance portfolio theory quantifies this mean-variance tradeoff
and provides a methodology that investors can use to determine optimal
portfolio choice.

This chapter introduces Markowitz's mean-variance portfolio theory
in a simplified setting where there are only two risky assets and
a single risk-free asset. This allows us to present the main results
of mean-variance portfolio theory graphically and with simple algebra.
However, as you shall see in the next chapter, many of the analytic
results I derive for portfolios of two risky assets, and for portfolios
of two risky assets and a single risk-free asset, can be applied to
situations in which there are many risky assets and a single risk-free
asset. Therefore, it is highly beneficial to study and understand
well the portfolio theory results in the simplified framework of this
chapter. 

This chapter is organized as follows. In @sec-Portfolios-of-Two-Risky-Assets
I review expected return, variance and value-at-risk calculations for
portfolios of two risky assets. In @sec-Efficient-portfolios-with-two-risky-assets
I introduce mean-variance portfolio analysis for portfolios of two
risky assets. The portfolio risk-return frontier is introduced, efficient
portfolios are defined, and the global minimum variance portfolio
is defined. In @sec-Efficient-portfolios-with-a-risk-free-asset I 
present the analysis of portfolios with a single risky asset and
a risk-free asset. I also define the Sharpe ratio/slope of a risky asset and illustrate it graphically. 
I discuss portfolios of two risky assets and a risk-free asset, define the The tangency portfolio, 
present the Mutual Fund Separation Theorem in @sec-Efficient-portfolios-with-two-risky-assets-and-a-risk-free-asset
I give a brief illustration of the theory to a real world two asset portfolio of
stocks and bonds in @sec-Application-to-Real. In the Appendix I review some basic results on constrained
optimization of functions that I use in the chapter. 

The R packages I use in this chapter are **introcompfinr**, **kableExtra**, **knitr**,  
and **PerformanceAnalytics**. Make sure these packages are installed
and loaded in R before replicating the chapter examples.

```{r}
#| include: false
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```

## Assumptions About Returns and Investor Risk Preferences

To develop a quantitative methodology for investment analysis, I make the following assumptions regarding the probability distribution of asset returns and the behavior of
investors: 

1.  Returns are covariance stationary and ergodic, and jointly normally distributed over the investment horizon. This implies that means, variances and covariances of returns are constant over the investment horizon and completely characterize the joint distribution of returns. 
2.  Investors know the values of asset return means, variances and covariances. 
3.  Investors only care about portfolio expected return and portfolio variance. Investors like portfolios with high expected return but dislike portfolios with high return variance. 

The first assumption is supported by the stylized facts of asset returns for investment
horizons of at least a month as discussed in Chapter @sec-Descriptive-Statistics-for-Financial-Data. The second assumption can be thought of
as assuming that investors have access to high quality information that they can use to determine expected returns, variances, and covariances. For example, investors can
download asset price data from `finance.yahoo.com` and compute estimates of expected
returns, variances, and covariances using sample statistics. Two points are relevant for the third assumption. First, if the distribution of assets returns is jointly normal then
the probability distribution of returns is completely characterized by means, variances, and covariances. Second, the third assumption describes the mean-variance trade-off investors face. All else equal, investors prefer assets that have the highest expected return. However, as you have seen, assets with high expected
returns tend to also have high variances. High variance represents more uncertainty about future return and it is assumed that investors do not like uncertainty about future return. That is, I assume that investors are *risk averse*.

## Portfolios of Two Risky Assets {#sec-Portfolios-of-Two-Risky-Assets}
\index{Portfolio of two risky assets}
Consider the following investment problem. You can invest in two assets
A (Amazon stock) and B (Boeing stock) over the next year, and you want to know how much
to invest in each asset to make the investor most happy given the assumptions about return distributions and risk preferences from the previous sub-section. Let $R_{A}$ and $R_{B}$ denote the annual simple returns on assets A and  B, respectively.^[If the stocks pay a dividend then returns are computed as total returns.]
These returns are to be treated as random variables because the returns
will not be realized until the end-of-the year. I assume that the
returns $R_{A}$ and $R_{B}$ are jointly normally distributed, and
that I have the following information about the means, variances,
and covariances of the probability distribution of the two returns:

$$
\mu_{A} =E[R_{A}],~\sigma_{A}^{2}=\mathrm{var}(R_{A}),~\mu_{B}=E[R_{B}],~\sigma_{B}^{2}=\mathrm{var}(R_{B}),
$$ {#eq-portfolio-inputs-1}
$$
\sigma_{AB} =\mathrm{cov}(R_{A},R_{B}),~\rho_{AB}=\mathrm{cor}(R_{A},R_{B})=\frac{\sigma_{AB}}{\sigma_{A}\sigma_{B}}.
$$ {#eq-portfolio-inputs-2}

I assume that these values are taken as given. In practice, they can be
estimated from historical return data for the two stocks, or they can be subjective guesses by the investor. 

The expected returns, $\mu_{A}$ and $\mu_{B}$, are our best guesses
for the annual returns on each of the assets. However, because the
investment returns are random variables you must recognize that the
realized returns may be different from our expectations. The variances,
$\sigma_{A}^{2}$ and $\sigma_{B}^{2}$, provide measures of the uncertainty
associated with these annual returns. You can also think of the variances
as measuring the risk associated with the investments. Assets with
high return variability (or volatility) are often thought to be risky,
and assets with low return volatility are often thought to be safe.
The covariance $\sigma_{AB}$ gives us information about the direction
of any linear dependence between returns. If $\sigma_{AB}>0$ then
the two returns tend to move in the same direction; if $\sigma_{AB}<0$
the returns tend to move in opposite directions; if $\sigma_{AB}=0$
then the returns tend to move independently. The strength of the dependence
between the returns is measured by the correlation coefficient $\rho_{AB}$.
If $\rho_{AB}$ is close to one in absolute value then returns mimic
each other extremely closely, whereas if $\rho_{AB}$ is close to
zero then the returns may show very little relationship.

::: {#exm-two-risky-asset-portfolio-information .example name="Two risky asset portfolio information"}

@tbl-TableExampleData gives annual return distribution
parameters for the two hypothetical assets A and B that will be used in subsequent examples. Asset A is
the high risk asset with an annual return of $\mu_{A}=17.5\%$ and
annual standard deviation of $\sigma_{A}=25.8\%$. Asset B is a lower
risk asset with annual return $\mu_{B}=5.5\%$ and annual standard
deviation of $\sigma_{B}=11.5\%$. The assets are assumed to be slightly
negatively correlated with correlation coefficient $\rho_{AB}=-0.164$.^[The small negative correlation between the two stock returns is a
bit unusual, as most stock returns are positively correlated. The
negative correlation is assumed so that certain figures produced later
look nice and are easy to read. ] Given the standard deviations and the correlation, the covariance
can be determined from $\sigma_{AB}=\rho_{AB}\sigma_{A}\sigma_{B}=(-0.164)(0.0258)(0.115)=-0.004875.$
In R, the example data is:^[I use the **kableExtra** function `kbl()` to print out nice formatted tables from a `data.frame` of information.]

```{r}
#| echo: true
mu.A = 0.175
sig.A = 0.258
sig2.A = sig.A^2
mu.B = 0.055
sig.B = 0.115
sig2.B = sig.B^2
rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B
```

```{r}
#| echo: false
#| label: tbl-TableExampleData
#| tbl-cap: "Example data for two asset portfolio"

data.tbl = data.frame(t(c(mu.A, mu.B, sig2.A, sig2.B, sig.A, sig.B, sig.AB, rho.AB)))
col.names = c("\\(\\mu_{A}\\)","\\(\\mu_{B}\\)","\\(\\sigma^{2}_{A}\\)","\\(\\sigma^{2}_{B}\\)",
              "\\(\\sigma_{A}\\)","\\(\\sigma_{B}\\)","\\(\\sigma_{AB}\\)","\\(\\rho_{AB}\\)")
kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
```

\hfill{}$\blacksquare$
:::

### The Portfolio Problem

The portfolio problem is set-up as follows. The investor has a given amount
of initial wealth $W_{0}$ to be invested for one period (e.g., one month or one year). 
The investor can only invest in the two risky assets A and B and all wealth must be invested in the two assets. The investor has to decide how much wealth to put in asset A and how
much to put in asset B. Let $x_{A}$ and $x_{B}$ denote the shares of wealth
invested in assets A and B, respectively. The values of $x_{A}$ and $x_{B}$ can be positive or negative. Positive values denote *long* positions (purchases) in the assets. Negative values denote *short* positions (sales).^[To short an asset one borrows the asset, usually from a broker, and then sells it. The proceeds from 
the short sale are usually kept on account with a broker and there may be restrictions on the
use of these funds for the purchase of other assets. The short position
is closed out when the asset is repurchased and then returned to original
owner. If the asset drops in value then a gain is made on the short
sale and if the asset increases in value a loss is made. See Chapter  @sec-Portfolio-Theory-No-Shorts for more details on short selling assets] Since all wealth is put into the two investments it follows that
$x_{A}+x_{B}=1$. If asset A is shorted, then it is assumed that
the proceeds of the short sale are used to purchase more of asset
B and vice-versa.  

The investment in the two assets forms a *portfolio,* and the
shares $x_{A}$ and $x_{B}$ are referred to as *portfolio shares*
or *portfolio weights*. The return on the portfolio over the next year is a random
variable, and is given by: 

$$
R_{p}=x_{A}R_{A}+x_{B}R_{B},
$$ {#eq-Rp}

which is a linear combination or weighted average of the random variables
$R_{A}$ and $R_{B}$. Since $R_{A}$ and $R_{B}$ are assumed to
be normally distributed, $R_{p}$ is also normally distributed. 

Using the properties of linear combinations of random variables from 
Chapter @sec-Review-of-Random-Variables it follows that

$$
\mu_{p} =E[R_{p}]=x_{A}\mu_{A}+x_{B}\mu_{B},
$$ {#eq-port-ER}
$$
\sigma_{p}^{2} =\mathrm{var}(R_{p})=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB},
$$ {#eq-port-var}
$$
\sigma_{p} =\mathrm{SD}(R_{p})=\sqrt{x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}}.
$$ {#eq-port-sd}

Hence, the portfolio return has the following normal distribution: 
$$
R_{p}\sim N(\mu_{p},\sigma_{p}^{2}).
$$

Given the portfolio return distribution, end-of-period wealth is

$$
W_{1} = W_{0}\times (1 + R_{p}) = W_{0}+ W_{0}\times R_{p},
$$

which is a linear function of the random portfolio return $R_{p}$. It follows that 

\begin{align*}
E[W_{1}] &= W_{0} + W_{0}\times E[R_{p}] = W_{0}(1 + \mu_{p}), \\
\mathrm{var}(W_{1}) &= W_{0}^{2}\times \mathrm{var}(R_{p}) = W_{0}^{2}\times \sigma_{p}^2,
\end{align*}

so that the distribution of $W_{1}$ is normal:

$$
W_{1} \sim N(W_{0}(1 + \mu_{p}), W_{0}^{2}\times \sigma_{p}^2).
$$

The investor cares most about the distribution of end-of-period wealth $W_{1}$, which
depends on the values of $W_{0}$, $\mu_{p}$, and $\sigma_{p}^2$. Since $\mu_{p}$ and $\sigma_{p}^2$ are functions of $x_{A}$ and $x_{B}$, the portfolio problem is to find the values of $x_{A}$ and $x_{B}$ that makes the investor most happy with respect to the probability distribution of end-of-period wealth.^[In formal terms, the portfolio problem is to find the values of $x_{A}$ and $x_{B}$ that maximizes the investor's expected utility over end-of-period wealth]

The results @eq-port-ER and @eq-port-var are so important
to portfolio theory that it is worthwhile to review the derivations.^[See chapter @sec-Review-of-Random-Variables section @sec-Random-Variables.]
For the first result @eq-port-ER, you have: 
$$
E[R_{p}]=E[x_{A}R_{A}+x_{B}R_{B}]=x_{A}E[R_{A}]+x_{B}E[R_{B}]=x_{A}\mu_{A}+x_{B}\mu_{B},
$$
by the linearity of the expectation operator. For the second result
@eq-port-var, you have: 
\begin{align*}
\mathrm{var}(R_{p}) & =E[(R_{p}-\mu_{p})^{2}]=E[(x_{A}(R_{A}-\mu_{A})+x_{B}(R_{B}-\mu_{B}))^{2}]\\
 & =E[x_{A}^{2}(R_{A}-\mu_{A})^{2}+x_{B}^{2}(R_{B}-\mu_{B})^{2}+2x_{A}x_{B}(R_{A}-\mu_{A})(R_{B}-\mu_{B})]\\
 & =x_{A}^{2}E[(R_{A}-\mu_{A})^{2}]+x_{B}^{2}E[(R_{B}-\mu_{B})^{2}]+2x_{A}x_{B}E[(R_{A}-\mu_{A})(R_{B}-\mu_{B})]\\
 & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}.
\end{align*}

The third line uses linearity of the expectations operator, and the fourth line uses the definition of variance and covariance.

Notice that the variance of the portfolio is a weighted average of
the variances of the individual assets plus two times the product
of the portfolio weights times the covariance between the assets.
If the portfolio weights are both positive then a positive covariance
will tend to increase the portfolio variance, because both returns
tend to move in the same direction, and a negative covariance will
tend to reduce the portfolio variance as returns tend to move in opposite directions. Thus finding assets with negatively
correlated returns can be very beneficial when forming portfolios
because risk, as measured by portfolio standard deviation, can be
reduced. What is perhaps surprising is that forming portfolios with
positively correlated assets can also reduce risk as long as the correlation
is not too large. I illustrate these results with the following examples.

::: {#exm-two-asset-portfolios .example name="Two asset portfolios"}

Consider creating some portfolios using the asset information in @tbl-TableExampleData. The first portfolio is an equally weighted
portfolio with $x_{A}=x_{B}=0.5$. Using @eq-port-ER-@eq-port-sd,
you have: 
\begin{align*}
\mu_{p_{1}} & =(0.5)\cdot(0.175)+(0.5)\cdot(0.055)=0.115\\
\sigma_{p_{1}}^{2} & =(0.5)^{2}\cdot(0.067)+(0.5)^{2}\cdot(0.013)\\
 & +2\cdot(0.5)(0.5)(-0.004866)\\
 & =0.01751\\
\sigma_{p_{1}} & =\sqrt{0.01751}=0.1323
\end{align*}
This portfolio has expected return half-way between the expected returns
on assets A and B, but the portfolio standard deviation is less
than half-way between the asset standard deviations. This reflects
risk reduction via *diversification*. Here, diversification means investing wealth in two assets instead of investing wealth in a single asset. In R, the portfolio parameters
are computed using:

```{r}
#| echo: true
x.A = 0.5
x.B = 0.5
mu.p1 = x.A*mu.A + x.B*mu.B
sig2.p1 = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p1 = sqrt(sig2.p1)
```

```{r}
#| echo: false
#| results: asis

data.tbl = t(c(mu.p1, sig.p1, 0.5*sig.A+0.5*sig.B))
col.names = c("$\\mu_{p_{1}}$", "$\\sigma_{p_{1}}$", "$0.5\\sigma_{A}+0.5\\sigma_{B}$")
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```

Next, consider a long-short portfolio with $x_{A}=1.5$ and $x_{B}=-0.5$.
In this portfolio, asset B is sold short and the proceeds of the
short sale are used to *leverage* the investment in asset A.^[Leverage refers to an investment that is financed through borrowing. Here, the short sale of asset B produces borrowed funds used to purchase more of asset A. This increases the risk (portfolio standard deviation) of the investment.] The
portfolio characteristics are: 
\begin{align*}
\mu_{p_{2}} & =(1.5)\cdot(0.175)+(-0.5)\cdot(0.055)=0.235\\
\sigma_{p_{2}}^{2} & =(1.5)^{2}\cdot(0.067)+(-0.5)^{2}\cdot(0.013)\\
 & +2\cdot(1.5)(-0.5)(-0.004866)\\
 & =0.1604\\
\sigma_{p_{2}} & =\sqrt{0.1604}=0.4005
\end{align*}
This portfolio has both a higher expected return and standard deviation
than asset A. The high standard deviation is due to the short sale, which is a type of 
leverage, and the negative correlation between assets A and B. In R, the portfolio parameters are computed using:

```{r}
#| echo: true
x.A = 1.5
x.B = -0.5
mu.p2 = x.A*mu.A + x.B*mu.B
sig2.p2 = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p2 = sqrt(sig2.p2)
```

```{r}
#| echo: false
#| results: asis

data.tbl = t(c(mu.p2, sig.p2, 1.5*sig.A-0.5*sig.B))
col.names = c("$\\mu_{p_{2}}$", "$\\sigma_{p_{2}}$", "$1.5\\sigma_{A}-0.5\\sigma_{B}$")
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```


\hfill{}$\blacksquare$
:::

In the above example, the equally weighted portfolio has expected
return half way between the expected returns on assets A and B, but
has standard deviation (volatility) that is less than half way between
the standard deviations of the two assets. For long-only portfolios,
you can show that this is a general result as long as the correlation
between the two assets is not perfectly positive. 

::: {#prp-01 .proposition}
Consider a portfolio of the two assets A and B with portfolios weights $x_{A}\geq0$ and
$x_{B}\geq0$ such that $x_{A}+x_{B}=1.$ Then if $\rho_{AB}\neq1$
  
$$
\sigma_{p}=(x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB})^{1/2}<x_{A}\sigma_{A}+x_{B}\sigma_{B}.
$$

If $\rho_{AB}=1$ then 

$$
\sigma_{p}=x_{A}\sigma_{A}+x_{B}\sigma_{B}.
$$
:::

The proof is straightforward. Use $\sigma_{AB}=\rho_{AB}\sigma_{A}\sigma_{B}$ and
write

$$
\sigma_{p}^{2}=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\rho_{AB}\sigma_{A}\sigma_{B}.
$$

Now add and subtract $2x_{A}x_{B}\sigma_{A}\sigma_{B}$ from the right-hand
side to give
\begin{eqnarray*}
\sigma_{p}^{2} & = & (x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{A}\sigma_{B})-2x_{A}x_{B}\sigma_{A}\sigma_{B}+2x_{A}x_{B}\rho_{AB}\sigma_{A}\sigma_{B}\\
 & = & \left(x_{A}\sigma_{A}+x_{B}\sigma_{B}\right)^{2}-2x_{A}x_{B}\sigma_{A}\sigma_{B}(1-\rho_{AB}).
\end{eqnarray*}
If $\rho_{AB}\neq1$ then $2x_{A}x_{B}\sigma_{A}\sigma_{B}(1-\rho_{AB})>0$
and so 
$$
\sigma_{p}^{2}<\left(x_{A}\sigma_{A}+x_{B}\sigma_{B}\right)^{2}\Rightarrow\sigma_{p}<x_{A}\sigma_{A}+x_{B}\sigma_{B}.
$$

If $\rho_{AB}=1$ then $2x_{A}x_{B}\sigma_{A}\sigma_{B}(1-\rho_{AB})=0$
and 
$$
\sigma_{p}^{2}=\left(x_{A}\sigma_{A}+x_{B}\sigma_{B}\right)\Rightarrow\sigma_{p}=x_{A}\sigma_{A}+x_{B}\sigma_{B}
$$
This result shows that the volatility of a long-only two asset portfolio 
is a weighted average of individual asset volatility only when the
two assets have perfectly positively correlated returns. Hence, in
this case there is no risk reduction benefit from forming a portfolio.
However, if assets are not perfectly correlated then there can be
a risk reduction benefit from forming a portfolio. How big the risk
reduction benefit is depends on the magnitude of the asset return
correlation $\rho_{AB}$. In general, the closer $\rho_{AB}$ is to
-1 the larger is the risk reduction benefit. 


@fig-IntroPortRiskReturn shows the risk-return characteristics
of assets A and B as well as the equally weighted and long-short portfolios
(labeled P1 and P2, respectively). The dotted line connecting the
points A and B represents linear risk-return trade-offs of portfolios
of assets A and B, respectively, that occurs if $\rho_{AB}=1$. The point P1, which represents the
equally weighted long-only portfolio, has expected return half-way between the
expected returns on assets A and B but has standard deviation less
than half-way between the standard deviations of assets A and B. As
a result, P1 lies to the left of the dotted line indicating a risk
reduction benefit to the portfolio. In contrast, the point P2, which
represents the long-short portfolio, lies to the right of the dotted
line indicating a risk inflation cost of the portfolio due to leverage. 

```{r}
#| label: fig-IntroPortRiskReturn
#| echo: false
#| fig-cap: Risk-return characteristics of assets A and B, the equally weighted portfolio
#|   P1 and the long-short portfolio P2.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

sigma.vals = c(sig.A, sig.B, sig.p1, sig.p2) 
mu.vals = c(mu.A, mu.B, mu.p1, mu.p2) 
names(sigma.vals) = names(mu.vals) = c("A", "B", "P1", "P2") 
plot(sigma.vals, mu.vals,       
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab=1.5,      
     xlim = c(0, 0.40), ylim = c(0, 0.30),      
     pch=16, col="blue", cex=2) 
text(sigma.vals, mu.vals, labels=names(mu.vals), pos=2) 
slope = (mu.A - mu.B)/(sig.A - sig.B) 
intercept = mu.A - slope*sig.A 
abline(a=intercept, b=slope, lwd=2, lty="dotted")
```


### Portfolio Value-at-Risk
\index{Portfolio VaR}
Consider an initial investment of $\$W_{0}$ in the portfolio of assets
A and B with return given by @eq-Rp, expected return
given by @eq-port-ER) and variance given by @eq-port-var.
Then $R_{p}\sim N(\mu_{p},\sigma_{p}^{2})$. For $\alpha\in(0,1)$,
the $\alpha\times100\%$ portfolio value-at-risk is given by: 

$$
\mathrm{VaR}_{p,\alpha}= -q_{\alpha}^{R_{p}}W_{0},
$$ {#eq-portfolio-VaR}

where $q_{\alpha}^{R_{p}}$ is the $\alpha$-quantile of the distribution
of $R_{p}$ and is given by 

$$
q_{\alpha}^{R_{p}}=\mu_{p}+\sigma_{p}q_{\alpha}^{z},
$$ {#eq-portfolio-normal-quantile}

where $q_{\alpha}^{z}$ is the $\alpha$-quantile of the standard
normal distribution.^[If $R_{p}$ is a continuously compounded return then the implied simple
return quantile is $q_{\alpha}^{R_{p}}=\exp(\mu_{p}+\sigma_{p}q_{\alpha}^{z})-1$.] 

What is the relationship between portfolio VaR and the individual
asset VaRs? Is portfolio VaR a weighted average of the individual
asset VaRs? In general, portfolio VaR is not a weighted average of
the asset VaRs. To see this consider the portfolio weighted average
of the individual asset return quantiles: 

$$
\begin{aligned}
x_{A}q_{\alpha}^{R_{A}}+x_{B}q_{\alpha}^{R_{B}} & =x_{A}(\mu_{A}+\sigma_{A}q_{\alpha}^{z})+x_{B}(\mu_{B}+\sigma_{B}q_{\alpha}^{z}) \\
 & =x_{A}\mu_{A}+x_{B}\mu_{B}+(x_{A}\sigma_{A}+x_{B}\sigma_{B})q_{\alpha}^{z} \\
 & =\mu_{p}+(x_{A}\sigma_{A}+x_{B}\sigma_{B})q_{\alpha}^{z}.
\end{aligned}
$$ {#eq-weighted-asset-quantile}

The weighted asset quantile @eq-weighted-asset-quantile is
not equal to the portfolio quantile @eq-portfolio-normal-quantile
unless $\rho_{AB}=1$. Hence, weighted asset VaR is in general not
equal to portfolio VaR because the quantile @eq-weighted-asset-quantile
assumes a perfectly positive correlation between $R_{A}$ and $R_{B}$.

::: {#exm-portfolio-var .example name="Portfolio VaR"}

Consider an initial investment of $W_{0}=\$100,000$. Assuming that
returns are simple, the 5\% VaRs on assets A and B are: 

\begin{align*}
\mathrm{VaR}_{A,0.05} & = -q_{0.05}^{R_{A}}W_{0}= -\left(\mu_{A}+\sigma_{A}q_{.05}^{z}\right)W_{0}\\
 & = -(0.175+0.258(-1.645))\cdot100,000= 24,937,\\
\mathrm{VaR}_{B,0.05} & = -q_{0.05}^{R_{B}}W_{0}= -\left(\mu_{B}+\sigma_{B}q_{.05}^{z}\right)W_{0}\\
 & = -(0.055+0.115(-1.645))\cdot100,000= 13,416.
\end{align*}

The 5\% VaR on the equal weighted portfolio with $x_{A}=x_{B}=0.5$
is: 

\begin{align*}
\mathrm{VaR}_{p,0.05} & = -q_{0.05}^{R_{p}}W_{0}= -\left(\mu_{p}+\sigma_{p}q_{.05}^{z}\right)W_{0}\\
 & = -(0.115+0.1323(-1.645))\cdot100,000= 10,268,
\end{align*}

and the weighted average of the individual asset VaRs is, 
$$
x_{A}\mathrm{VaR}_{A,0.05}+x_{B}\mathrm{VaR}_{B,0.05}=0.5(24,937)+0.5(13,416)=19,177.
$$
The 5\% VaR on the long-short portfolio with $x_{A}=1.5$ and $x_{B}=-0.5$
is: 
$$
\mathrm{VaR}_{p,0.05}= -q_{0.05}^{R_{p}}W_{0}= -(0.235+0.4005(-1.645))\cdot100,000=42,371,
$$
and the weighted average of the individual asset VaRs is, 
$$
x_{A}\mathrm{VaR}_{A,0.05}+x_{B}\mathrm{VaR}_{B,0.05}=1.5(24,937)-0.5(13,416)=30,698.
$$
Notice that VaR$_{p,0.05}\neq x_{A}\mathrm{VaR}_{A,0.05}+x_{B}\mathrm{VaR}_{B,0.05}$
because $\rho_{AB}\neq1$. 

Using R, these computations are:

```{r}
#| echo: true
#| results: asis
w0 = 100000
VaR.A = -(mu.A + sig.A*qnorm(0.05))*w0
VaR.B = -(mu.B + sig.B*qnorm(0.05))*w0
VaR.p1 = -(mu.p1 + sig.p1*qnorm(0.05))*w0
VaR.p2 = -(mu.p2 + sig.p2*qnorm(0.05))*w0
```

```{r}
#| echo: false
#| results: asis
data.tbl = t(c(VaR.A,VaR.B,VaR.p1,VaR.p2))
col.names = c("$VaR_{A,0.05}$", "$VaR_{B,0.05}$", "$VaR_{p_{1},0.05}$",
              "$VaR_{p_{2},0.05}$")
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```


```{r}
#| echo: false
#| results: asis
data.tbl = t(c(0.5*VaR.A+0.5*VaR.B,1.5*VaR.A-0.5*VaR.B))
col.names = c("$0.5VaR_{A,0.05}+0.5VaR_{B,0.05}$",
              "$1.5VaR_{A,0.05}-0.5VaR_{B,0.05}$")
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```


\hfill{}$\blacksquare$
:::

::: {#exm-create-r-function-to-compute-normal-var .example name="Create R function to compute normal VaR"}

The previous example used repetitive R calculations to compute the
5\% VaR for a variety of investments. An alternative approach is to create
an R function to compute the VaR given $\mu$, $\sigma$, $\alpha$
(VaR probability) and $W_{0}$, and then apply the function using
the inputs of the different assets. A simple function to compute VaR
based on normally distributed asset returns is:

```{r}
#| echo: true
normalVaR = function(mu, sigma, w0, alpha = 0.05, invert=TRUE) {
## compute normal VaR for collection of assets given mean and sd vector
## inputs:
## mu         n x 1 vector of expected returns
## sigma      n x 1 vector of standard deviations
## w0         scalar initial investment in $
## alpha      scalar tail probability
## invert     logical. If TRUE report VaR as positive number
## output:
## VaR        n x 1 vector of left tail return quantiles
  if ( length(mu) != length(sigma) )
    stop("mu and sigma must have same number of elements")
  if ( alpha < 0 || alpha > 1)
    stop("alpha must be between 0 and 1")
  VaR = w0*(mu + sigma*qnorm(alpha))
  if (invert) {
    VaR = -VaR
  }
  return(VaR)
}
```

You create an R function using `function()`. Every R function has a set of arguments or inputs. Here, the arguments are the $n \times 1$ vectors `mu` and `sigma`, the scalars `w0` and `alpha`, and the logical value `invert`. Default values for `alpha` and `invert` are specified in the definition of the function. The body of the function is defined within the braces `{}`, and the return value of the function is specified using `return()`. Here, you first check that the inputs `mu` and `sigma` are of the same length, and that
the tail probability $\alpha$ is between 0 and 1. Then, you compute normal VaR using @eq-portfolio-VaR. If `invert=TRUE` you multiply the result by $-1$. Finally, you return the vector of VaR values. After running the code that defines the function `normalVaR()`, it will be available in the R session for use like any other R function. 

Using the `normalVaR()` function, the 5\% VaR values of asset
A, B and equally weighted portfolio are:

```{r}
#| echo: true
normalVaR(mu=c(mu.A, mu.B, mu.p1),
          sigma=c(sig.A, sig.B, sig.p1),
          w0=100000, alpha=0.05) 
```

\hfill{}$\blacksquare$
:::

### The set of feasible portfolios

The collection of all *feasible* portfolios, or the *investment possibilities set*, in the case of two assets is simply all possible portfolios that can be formed
by varying the portfolio weights $x_{A}$ and $x_{B}$ such that the
weights sum to one ($x_{A}+x_{B}=1)$. I summarize the expected return-risk
(mean-volatility) properties of the feasible portfolios in a plot with
portfolio expected return, $\mu_{p}$, on the vertical axis and portfolio
standard deviation, $\sigma_{p}$, on the horizontal axis. The portfolio
standard deviation is used instead of variance because standard deviation
is measured in the same units as the expected value (recall, variance
is the average squared deviation from the mean).

::: {#exm-investment-possibilities-set-for-example-data .example name="Investment possibilities set for example data"}

The investment possibilities set for the data
in @tbl-TableExampleData is illustrated in @fig-IntroPortEfficientPort.
Here the portfolio weight on asset A, $x_{A}$, is varied from -0.4
to 1.4 in increments of 0.1 and, since $x_{B}=1-x_{A}$, the weight
on asset B then varies from 1.4 to -0.4. This gives us 18 portfolios
with weights $(x_{A},x_{B})=(-0.4,1.4),(-0.3,1.3),...,(1.3,-0.3),(1.4,-0.4)$.
For each of these portfolios I use the formulas @eq-port-ER
and @eq-port-sd to compute $\mu_{p}$ and $\sigma_{p}$. In R, the calculations are:

```{r}
#| echo: true
#| results: asis
x.A = seq(from=-0.4, to=1.4, by=0.1)
x.B = 1 - x.A
mu.p = x.A*mu.A + x.B*mu.B
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p = sqrt(sig2.p)
port.names = paste("portfolio", 1:length(x.A), sep=" ")
data.tbl = as.data.frame(cbind(x.A, x.B, mu.p, sig.p))
rownames(data.tbl) = port.names
col.names = c("$x_{A}$","$x_{B}$", "$\\mu_{p}$", "$\\sigma_{p}$" )
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```

Portfolios 1-4 and 16-19 are the long-short portfolios and portfolios 5-15 are the long-only portfolios. In R, the risk-return properties of this set of feasible portfolios can be visualized using:

```{r}
#| label: fig-IntroPortEfficientPort
#| echo: true
#| fig-cap: Portfolio frontier of example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
cex.val = 1.5
plot(sig.p, mu.p, type="b", pch=16, cex=cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)), 
     xlab=expression(sigma[p]), ylab=expression(mu[p]),
     cex.lab=1.5, col=c(rep("blue", 4), rep("black", 2),
                        "green", rep("black", 8), rep("blue", 4)))
text(x=sig.A, y=mu.A, labels="Asset A", pos=4)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4)
```

\hfill{}$\blacksquare$
:::

Notice that the plot in ($\mu_{p},\sigma_{p})$-space looks like a
parabola turned on its side (in fact, it is one side of a hyperbola). The plot is
often referred to as the *Markowitz bullet* as it resembles the shape of a bullet speeding toward the y-axis. The black dots plus the green dot represent the long-only portfolios, and the blue dots represent long-short portfolios. The black dot labeled "Asset A" is the portfolio $(x_{A}=1, x_{B}=0)$, and the black dot labeled "Asset B" is the portfolio $(x_{A}=0, x_{B}=1)$. 

The bullet shape of the feasible set shows that portfolio risk can be reduced through diversification. Risk reduction through diversification can be understood as follows. Suppose an investor is initially invested 100\% in asset B (i.e., portfolio 5). Diversification involves rebalancing the portfolio to include asset A. For example, portfolio 6 is $(x_{A}=0.1, x_{B}=0.9)$. This portfolio has both a higher expected return and a lower standard deviation than portfolio 5. This is the benefit from diversification. Portfolio expected return increases and risk decreases until portfolio 7 is reached, which is at the tip of the Markowitz bullet (green dot). This portfolio has the property that it has the smallest variance/volatility among all feasible portfolios. Accordingly,
this portfolio is called the *global minimum variance portfolio*.\index{Global minimum variance portfolio} Continuing from portfolio 8 through portfolio 19, you see that both expected return and risk increase.

### Computing the global minimum variance portfolio

The global minimum variance portfolio plays a key role in mean-variance portfolio theory and it is important to know how to determine its weights. Here, I show that it is a simple exercise in calculus to find the global minimum variance portfolio weights. I solve the *constrained optimization* problem^[A review of optimization and constrained optimization is given in
the appendix to this chapter.]: 

\begin{align*}
\underset{x_{A},x_{B}}{\min}\sigma_{p}^{2} & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}\\
~s.t.~x_{A}+x_{B} & =1.
\end{align*}

This constrained optimization problem can be solved using two methods.
The first method, called the *method of substitution*\index{Method of substitution}, uses
the constraint to substitute out one of the variables to transform
the constrained optimization problem in two variables into an unconstrained
optimization problem in one variable. The second method, called the
*method of Lagrange multipliers*\index{Method of Lagrange multipliers}, introduces an auxiliary variable
called the *Lagrange multiplier* and transforms the constrained optimization
problem in two variables into an unconstrained optimization problem
in three variables. 

The substitution method is straightforward. Substituting $x_{B}=1-x_{A}$
into the formula for $\sigma_{p}^{2}$ reduces the problem to: 
$$
\min_{x_{A}}\sigma_{p}^{2}=x_{A}^{2}\sigma_{A}^{2}+(1-x_{A})^{2}\sigma_{B}^{2}+2x_{A}(1-x_{A})\sigma_{AB}.
$$
The first order conditions for a minimum, via the chain rule, are:
$$
0=\frac{d\sigma_{p}^{2}}{dx_{A}}=2x_{A}^{\min}\sigma_{A}^{2}-2(1-x_{A}^{\min})\sigma_{B}^{2}+2\sigma_{AB}(1-2x_{A}^{\min}),
$$
and straightforward calculations yield, 

$$
x_{A}^{\min}=\frac{\sigma_{B}^{2}-\sigma_{AB}}{\sigma_{A}^{2}+\sigma_{B}^{2}-2\sigma_{AB}},~x_{B}^{\min}=1-x_{A}^{\min}.
$$ {#eq-minvar-port}

The method of Lagrange multipliers involves two steps. In the first
step, the constraint $x_{A}+x_{B}=1$ is put into homogenous form
$x_{A}+x_{B}-1=0$. In the second step, the Lagrangian function is
formed by adding to $\sigma_{p}^{2}$ the homogenous constraint multiplied
by an auxiliary variable $\lambda$ (the Lagrange multiplier) giving:
$$
L(x_{A},x_{B},\lambda)=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}+\lambda(x_{A}+x_{B}-1).
$$
This function is then minimized with respect to $x_{A},$ $x_{B}$,
and $\lambda$. The first order conditions are: 

\begin{align*}
0 & =\frac{\partial L(x_{A},x_{B},\lambda)}{\partial x_{A}}=2x_{A}\sigma_{A}^{2}+2x_{B}\sigma_{AB}+\lambda,\\
0 & =\frac{\partial L(x_{A},x_{B},\lambda)}{\partial x_{B}}=2x_{B}\sigma_{B}^{2}+2x_{A}\sigma_{AB}+\lambda,\\
0 & =\frac{\partial L(x_{A},x_{B},\lambda)}{\partial\lambda}=x_{A}+x_{B}-1.
\end{align*}

The first two equations can be rearranged to give: 
$$
x_{B}=x_{A}\left(\frac{\sigma_{A}^{2}-\sigma_{AB}}{\sigma_{B}^{2}-\sigma_{AB}}\right).
$$
Substituting this value for $x_{B}$ into the third equation and rearranging
gives the solution @eq-minvar-port.

::: {#exm-global-minimum-variance-portfolio-for-example-data .example name="Global minimum variance portfolio for example data"}

Using the data in @tbl-TableExampleData and @eq-minvar-port
you have: 
$$
x_{A}^{\min}=\frac{0.01323-(-0.004866)}{0.06656+0.01323-2(-0.004866)}=0.2021,~x_{B}^{\min}=0.7979.
$$
Hence, the global minimum variance portfolio is essentially portfolio 7 in @fig-IntroPortEfficientPort and indeed lies at the tip of the Markowitz bullet.

The expected return, variance and standard deviation of this portfolio
are: 

\begin{align*}
\mu_{p} & =(0.2021)\cdot(0.175)+(0.7979)\cdot(0.055)=0.07925\\
\sigma_{p}^{2} & =(0.2021)^{2}\cdot(0.067)+(0.7979)^{2}\cdot(0.013)\\
 & +2\cdot(0.2021)(0.7979)(-0.004875)\\
 & =0.00975\\
\sigma_{p} & =\sqrt{0.00975}=0.09782.
\end{align*}

In R, the calculations to compute
the global minimum variance portfolio weights and its expected return and volatility are:

```{r}
#| echo: true
#| results: asis
xA.min = (sig2.B - sig.AB)/(sig2.A + sig2.B - 2*sig.AB)
xB.min = 1 - xA.min
mu.p.min = xA.min*mu.A + xB.min*mu.B
sig2.p.min = xA.min^2 * sig2.A + xB.min^2 * sig2.B + 2*xA.min*xB.min*sig.AB
sig.p.min = sqrt(sig2.p.min)
data.tbl = as.data.frame(t(c(xA.min, xB.min, mu.p.min, sig.p.min)))
col.names = c("$x_{A}^{\\min}$","$x_{B}^{\\min}$", "$\\mu_{p}^{\\min}$", 
              "$\\sigma_{p}^{\\min}$" )
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```

\hfill{}$\blacksquare$
:::

### Correlation and the shape of the portfolio frontier

The shape of the investment possibilities set is very sensitive to
the correlation between assets A and B given the other parameters. I illustrate this sensitivity by computing portfolio frontiers for the example data for $\rho_{AB} = -0.9, -0.5, 0, 0.5, 0.9$. @fig-IntroPortCorShape shows these portfolio frontiers.

```{r}
#| label: fig-IntroPortCorShape
#| echo: false
#| fig-cap: Portfolio frontier as a function of correlation.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
rho.AB.vals = c(-0.9,-0.5, 0, 0.5, 0.9)
x.A = seq(from=-0.4, to=1.4, by=0.1)
x.B = 1 - x.A
mu.p = x.A*mu.A + x.B*mu.B
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*rho.AB.vals[1]*sig.A*sig.B
sig.p = sqrt(sig2.p)

# create portfolio plot
cex.val = 1
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)),
     xlab=expression(sigma[p]), ylab=expression(mu[p]),
     cex.lab = cex.val, col=1)
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)

for (i in 2:length(rho.AB.vals)) {
  sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*rho.AB.vals[i]*sig.A*sig.B
  sig.p = sqrt(sig2.p)
  points(sig.p, mu.p, type="b", pch=16, col=i, cex=cex.val)
}
legend(x="bottomright", legend=paste("rho", as.character(rho.AB.vals), sep="="), 
       col=1:length(rho.AB.vals), lty=1, pch=16, cex = cex.val)
```
The curvature of the portfolio frontier is determined by the value of $\rho_{AB}$. The closer $\rho_{AB}$ is to -1 the more curved is the frontier toward the y-axis and the higher is the possible diversification benefit. What is perhaps surprising from @fig-IntroPortCorShape is that
there is noticeable curvature even for positive values of $\rho_{AB}$. 

```{r}
#| label: fig-IntroPortCorrelation
#| echo: false
#| results: hide
#| fig-cap: Portfolios with $\rho_{AB}=1$ and $\rho_{AB}=-1$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
rho.AB = 1
sig.AB = rho.AB*sig.A*sig.B
mu.p.1 = x.A*mu.A + x.B*mu.B
sig2.p.1 = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p.1 = sqrt(sig2.p.1)
rho.AB = -1
sig.AB = rho.AB*sig.A*sig.B
mu.p.m1 = x.A*mu.A + x.B*mu.B
sig2.p.m1 = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p.m1 = sqrt(sig2.p.m1)
cex.val = 1
plot(sig.p.1, mu.p.1, type="b", pch=16, cex = cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)),
     xlab=expression(sigma[p]), ylab=expression(mu[p]),
     cex.lab=1.5)
points(sig.p.m1, mu.p.m1, type="b", col="blue", pch=22, cex = cex.val)
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
legend(x="topleft", legend=c(expression(rho[AB]==1), expression(rho[AB]==-1)),
       col=c("black", "blue"), pch=c(16,22), cex = cex.val)
```

If $\rho_{AB}$ is close to 1 then the portfolio frontier approaches a straight line connecting the portfolio with all wealth invested in asset B, $(x_{A},x_{B})=(0,1)$, to the portfolio with all wealth invested in asset A, $(x_{A},x_{B})=(1,0)$. If $\rho_{AB}=-1$ then the
set actually touches the $\mu_{p}$ axis. What this means is that
if assets A and B are perfectly negatively correlated then there
exists a portfolio of A and B that has positive expected return
and zero variance! These cases are illustrated
in @fig-IntroPortCorrelation. 

To find the portfolio with $\sigma_{p}^{2}=0$
when $\rho_{AB}=-1$ I use @eq-minvar-port and the fact that
$\sigma_{AB}=\rho_{AB}\sigma_{A}\sigma_{B}$ to give: 
$$
x_{A}^{\min}=\frac{\sigma_{B}}{\sigma_{A}+\sigma_{B}},~x_{B}^{\min}=1-x_{A}.
$$

## Efficient portfolios with two risky assets {#sec-Efficient-portfolios-with-two-risky-assets}
\index{Efficient portfolios}

Given the assumptions on returns and investor risk preferences, I now characterize the set of *efficient portfolios* and *inefficient portfolios*\index{Inefficient portfolios}.

::: {#def-efficient-portfolios .definition name="Efficient Portfolios"}
*Efficient portfolios* are the feasible portfolios that have the highest expected return for a given level of risk as measured by portfolio standard deviation. These are the portfolios that investors are most interested in holding. Graphically, these are the portfolios starting with the global minimum variance portfolio that are above and to the right of the global minimum variance portfolio.
:::

::: {#def-inefficient-portfolios .definition name="Inefficient Portfolios"}
*Inefficient portfolios* are the portfolios such that there is another feasible portfolio that has the same risk but a higher expected return. Graphically, these are the portfolios that are below and to the right of the global minimum variance portfolio.
:::

```{r}
#| label: fig-IntroPortEfficientPorts
#| echo: false
#| results: hide
#| fig-cap: Efficient and inefficient portfolios for example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B
x.A = seq(from=-0.4, to=1.4, by=0.1)
x.B = 1 - x.A
mu.p = x.A*mu.A + x.B*mu.B
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p = sqrt(sig2.p)
plot(sig.p, mu.p, type="b", pch=16, ylim=c(0, max(mu.p)), cex = cex.val,
     xlim=c(0, max(sig.p)),
     xlab=expression(sigma[p]), ylab=expression(mu[p]),
     col=c(rep("red", 6), rep("green", 13)))
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
segments(x0=sig.p[3], y0=0, x1=sig.p[3], y1=mu.p[11], 
         lwd=3, lty="dotted")
text(x=sig.p[3], y=mu.p[3], labels="Inefficient portfolio (portfolio 3)", 
     pos=4, cex = cex.val)
text(x=sig.p[3], y=mu.p[11], labels="Efficient portfolio (portfolio 11)", 
     cex = cex.val, pos=4)
```

::: {#exm-efficient-and-inefficient-portfolios-for-example-data .example name="Efficient and inefficient portfolios for example data"}

@fig-IntroPortEfficientPorts shows the efficient and inefficient portfolios from the example data in @tbl-TableExampleData. The efficient portfolios are those that have the highest expected return for a given standard deviation value. These portfolios are the green dots starting with the global minimum variance portfolio at the tip of the Markowitz bullet. The inefficient portfolios are the red dots below the global minimum variance portfolio.  For example,  @fig-IntroPortEfficientPorts shows two feasible portfolios, portfolio 3 and portfolio 11, with the same standard deviation value but different expected return values. Portfolio 11 has the highest expected return and so is the efficient portfolio. Portfolio 3, with the lower expected return, is the inefficient portfolio.  
\hfill{}$\blacksquare$
:::

### Optimal portfolios
\index{Optimal portfolios}
Given the efficient set of portfolios as described in @fig-IntroPortEfficientPort,
which portfolio will an investor choose? Of the efficient portfolios,
investors will choose the one that accords with their *risk preferences*.
Very risk averse investors will want a portfolio that has low volatility
(risk) and will choose a portfolio very close to the global minimum
variance portfolio. In contrast, very risk tolerant investors will
ignore volatility and seek portfolios with high expected returns.
Hence, these investors will choose portfolios with large amounts of
asset A which may involve short-selling asset B.


## Efficient portfolios with a risk-free asset {#sec-Efficient-portfolios-with-a-risk-free-asset}

In the preceding section I constructed the efficient set of portfolios
involving two risky assets. Now I consider what happens
when you introduce a *risk-free* asset. In the present context, a risk-free
asset is equivalent to a default-free pure discount bond that matures
at the end of the assumed investment horizon.^[In the finance industry the risk-free asset is often referred to as
"cash". In general, "cash" doesn't refer to holding dollars
in your pocket but rather investing dollars in a very low risk asset
that pays a stable rate of return.] The *risk-free rate*, $r_{f}$, is then the nominal return on the bond.
For example, if the investment horizon is one month then the risk-free
asset is a 30-day U.S. Treasury bill (T-bill) and the risk-free rate
is the nominal rate of return on the T-bill.^[The default-free assumption of U.S. debt is often taken for granted. However, this assumption has been questioned due to the inability of the U.S. congress to address the long-term debt problems of the U.S. government.] If our holdings of the risk-free asset is positive then you are "lending money" to the government at the risk-free rate, and if our holdings are negative then you are "borrowing" from the government at the risk-free rate.

The assumption that investors can borrow and lend at the same risk-free rate is unrealistic in the real world due to differing credit worthiness of investors. Investors can certainly buy T-Bills and receive the risk-free rate. However, most investors can only borrow funds at rates much higher than the T-Bill rate. 

### Portfolios with one risky asset and one risk-free asset

In this sub-section, I consider portfolios of a single risky asset with random return $R \sim N(\mu, \sigma^2)$ and a risk-free asset with non-random return $r_{f}$. Since the risk-free rate is fixed over the investment horizon it is not a random variable. As a result, it has some special properties summarized in the following Proposition.

::: {#prp-properties-of-risk-free-rate .proposition name="Properties of Risk-Free Rate"}
Let $r_{f}$ denote the fixed non-random risk-free rate, and let $R$ denote the random return on a risky asset with $E[R]=\mu$ and $\mathrm{var}(R)=\sigma^{2}$. Then
\begin{align*}
E[r_{f}] &= r_{f},\\
\mathrm{var}(r_{f}) & = E[(r_{f}-E[r_{f}])^2] = 0,\\
\mathrm{cov}(R,r_{f}) &= E[(r_{f}-E[r_{f}])(R - E[R])] =0.
\end{align*}
:::

Consider an investment in the risky asset and the risk-free asset (henceforth
referred to as a T-bill). Let $x$ denote the share of wealth in the risky asset, and $x_{f}$
denote the share of wealth in T-bills such that $x + x_{f}=1$. Using $x_f = 1 - x$, the portfolio return can be written as: 
$$
R_{p}=x_{f}r_{f}+xR = (1-x)r_{f}+xR=r_{f}+x(R-r_{f}). 
$$ {#eq-RBplusTbillsReturn}
The quantity $R-r_{f}$ is called the *excess return*\index{Excess return} (over
the return on T-bills) on the risky asset. The portfolio expected return
is then: 

$$
\mu_{p}=r_{f}+x(E[R]-r_{f})=r_{f}+x(\mu-r_{f}),
$$ {#eq-muTbillsPlusRisky}

where the quantity $(\mu-r_{f})$ is called the *expected excess return* or *risk premium*\index{Risk premium} on the risky asset. The risk premium is typically positive indicating that investors expect
a higher return on the risky asset than the safe asset (otherwise,
why would investors hold the risky asset?). You may express the risk
premium on the portfolio in terms of the risk premium on the risky asset:
$$
\mu_{p}-r_{f}=x(\mu-r_{f}).
$$
The more you invest in the risky asset the higher the risk premium on the portfolio of T-Bills and the risky asset. 

Because the risk-free rate is constant, the portfolio variance only depends on the variability of the risky asset and is given by: 
$$
\sigma_{p}^{2}= \mathrm{var(R_{p})} = \mathrm{var(xR)}= x^{2}\sigma^{2}.
$$
The portfolio standard deviation is therefore proportional to the
standard deviation on the risky asset:^[Here, you can allow $x$ to be positive or negative but $\sigma$ is always positive. ]

$$
\sigma_{p}=|x|\sigma.
$$ {#eq-sigmaTbillsPlusRisky}

::: {#exm-portfolios-of-t-bills-and-asset-b-for-example-data .example name="Portfolios of T-Bills and asset B for example data"}

To illustrate the risk-return properties of portfolios of a T-Bill and a single risky asset, consider forming portfolios of T-Bills and asset B. Let $x_{B}$ and $1-x_{B}$ denote the shares of wealth in asset B and T-Bills, respectively. Assume the T-Bill rate is $r_f = 0.03$. Then the risk 
premium on asset B is $\mu_{B}-r_{r} = 0.055 - 0.03 = 0.025$, the expected return on portfolios of T-Bills and asset B is $\mu_{p} = 0.03 + x_{B}\times 0.025$, and the portfolio standard deviation is $\sigma_{p} = |x_{B}|\times 0.115$. @fig-IntroPortTbillsPlusRisky plots $\mu_{p}$ vs. $\sigma_{p}$ for portfolios with $x_{B} = -1.4, -1.3, \ldots 1.4$.


```{r}
#| label: fig-IntroPortTbillsPlusRisky
#| echo: false
#| results: hide
#| fig-cap: Portfolios of T-Bills and asset B for example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
cex.val = 1
r.f = 0.03
x.B = seq(-1.4, 1.4, by=0.1)
x.f = 1 - x.B
mu.p = r.f + x.B*(mu.B-r.f)
sig.p = abs(x.B)*sig.B
plot(sig.p, mu.p, type="b", pch=16, ylim=c(0, max(mu.p)), cex = cex.val,
     xlim=c(0, max(sig.p)), col=c(rep("blue", 14), rep("black", 11),
                                  rep("blue", 4)),
     xlab=expression(sigma[p]), ylab=expression(mu[p]))
abline(h=0)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = 1.25)
```

Notice that the set of feasible portfolios lie on two straight lines, one with a positive slope and one with a negative slope. The line with a positive slope represents portfolios with long positions in asset B, and the line with a negative slope represents portfolios with short positions in asset B. The blue dots represent portfolios with short positions and the black dots represent long-only portfolios. The blue dots above the point labeled "Asset B" represent long-short portfolios where the T-Bill is shorted to finance additional investment in asset B. 
\hfill{}$\blacksquare$
:::

If you assume that $x > 0$ (do not short the risky asset), you can use @eq-sigmaTbillsPlusRisky to solve for $x$: 
$$
x=\frac{\sigma_{p}}{\sigma}. 
$$ {#eq-solveForx}
Substituting @eq-solveForx into @eq-muTbillsPlusRisky gives the following equation: 

$$
\mu_{p}=r_{f}+\frac{\mu-r_{f}}{\sigma}\cdot\sigma_{p},
$$ {#eq-CAL}

Equation @eq-CAL describes a straight line in ($\mu_{p},\sigma_{p}$)-space with
intercept $r_{f}$ and slope $(\mu-r_{f})/\sigma$. This line characterizes the linear risk-return tradeoff of portfolios of T-Bills and a risky asset when the risky asset is not shorted. This line has a positive slope provided the risk premium is positive. 

This line described by @eq-CAL is often called the *capital allocation line*\index{Capital allocation line} (CAL) as it describes how investment capital is allocated between T-Bills and the risky asset. Points on the line close to the y-axis represent portfolios mostly allocated to T-Bills, and points far away from the y-axis represent portfolios mostly allocated to the risky asset. 
The slope of the CAL is called the *Sharpe ratio*\index{Sharpe ratio} (SR) (named after the economist William Sharpe), and it measures the risk premium on the asset per unit of risk (as measured by the
standard deviation of the asset). From @eq-CAL you get
$$
\frac{d\mu_{p}}{d\sigma_{p}}=\frac{\mu-r_{f}}{\sigma}=SR.
$$

::: {#exm-portfolios-of-t-bills-and-risky-assets-for-example-data .example name="Portfolios of T-Bills and risky assets for example data"}

Consider creating portfolios of asset A and T-bills and asset B and T-Bills with $r_{f}=0.03$:

```{r}
#| echo: true
r.f = 0.03
x.A = seq(from=0, to=1.4, by=0.1)
mu.p.A = r.f + x.A*(mu.A - r.f)
sig.p.A = x.A*sig.A
x.B = seq(from=0, to=1.4, by=0.1)
mu.p.B = r.f + x.B*(mu.B - r.f)
sig.p.B = x.B*sig.B
```

The risk-return properties of these portfolios are illustrated in @fig-IntroPortTbills.

```{r}
#| label: fig-IntroPortTbills
#| echo: false
#| results: hide
#| fig-cap: Portfolios of T-Bills and risky assets.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# plot portfolios of T-Bills and assets A and B
plot(sig.p.A, mu.p.A, type="b", col="green", ylim=c(0, max(mu.p.A)),
     xlim=c(0, max(sig.p.A, sig.p.B)), pch=16, cex = cex.val,
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab = cex.val)
points(sig.p.B, mu.p.B, type="b", col="red", pch=16, cex = cex.val)
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=1, cex = cex.val)
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = 1.25)
```

In @fig-IntroPortTbills the point labeled
"$r_{f}$" represents a portfolio that is 100\% invested in T-bills.
The points labeled "Asset A" and "Asset B" represent portfolios
that are 100\% invested in assets A and B, respectively. A point half-way
between "$r_{f}$" and "Asset A" is a portfolio that is 50\% invested
in T-bills and 50\% invested in asset A. A point above and to the
right of "Asset A" is a levered portfolio where more than 100\%
of asset A is financed by borrowing at the T-bill rate. Notice that as leverage increases (i.e, borrowing more at the T-Bill rate), portfolio volatility increases.

The portfolios which are combinations of asset A and T-bills have expected returns
uniformly higher than the portfolios consisting of asset B and T-bills.
This occurs because the Sharpe ratio for asset A is higher than
the ratio for asset B: 

\begin{align*}
\mathrm{SR}_{A} & =\frac{\mu_{A}-r_{f}}{\sigma_{A}}=\frac{0.175-0.03}{0.258}=0.562,\\
\mathrm{SR}_{B} & =\frac{\mu_{B}-r_{f}}{\sigma_{B}}=\frac{0.055-0.03}{0.115}=0.217.
\end{align*}
Hence, portfolios of asset A and T-bills are *efficient* relative
to portfolios of asset B and T-bills.
\hfill{}$\blacksquare$
:::

The previous example shows that the Sharpe ratio can be used to rank
the risk return properties of individual assets. Assets with a high
Sharpe ratio have a better risk-return tradeoff than assets with a
low Sharpe ratio. Accordingly, investment analysts routinely rank
assets based on their Sharpe ratios.

### Portfolios with one risky asset and one risk-free asset with different borrowing and lending rates

TBD

## Efficient portfolios with two risky assets and a risk-free asset {#sec-Efficient-portfolios-with-two-risky-assets-and-a-risk-free-asset}

Now I expand on the previous results by allowing our investor to
form portfolios of assets A, B and T-bills. The efficient set of portfolios
in this case will still be a straight line in ($\mu_{p},\sigma_{p})$-space
with intercept $r_{f}$. The slope of the efficient set, the maximum
Sharpe ratio, is such that it is tangent to the portfolio frontier constructed
just using the two risky assets A and B. @fig-IntroPortTangency
illustrates why this is so.

```{r}
#| label: fig-IntroPortTangency
#| echo: false
#| results: hide
#| fig-cap: Efficient portfolios of two risky assets and T-bills. Black dots represent
#|   portfolios of assets A and B; blue dots represent portfolios of T-bills and asset
#|   B; red dots represent portfolios of T-bills and asset A; green dots represent portfolios
#|   of the tangency portfolio and T-bills.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B
top = (mu.A - r.f)*sig2.B - (mu.B - r.f)*sig.AB
bot = (mu.A - r.f)*sig2.B + (mu.B - r.f)*sig2.A - (mu.A - r.f + mu.B - r.f)*sig.AB
x.A.tan = top/bot
x.B.tan = 1 - x.A.tan
# x.A.tan
# x.B.tan
mu.p.tan = x.A.tan*mu.A + x.B.tan*mu.B
sig2.p.tan = x.A.tan^2 * sig2.A + x.B.tan^2 * sig2.B + 2*x.A.tan*x.B.tan*sig.AB
sig.p.tan = sqrt(sig2.p.tan)
# mu.p.tan
# sig.p.tan

# T-bills plus tangency
x.tan = seq(from=0, to=2.4, by=0.1)
mu.p.tan.tbill = r.f + x.tan*(mu.p.tan - r.f)
sig.p.tan.tbill = x.tan*sig.p.tan
# plot portfolios with tangency portfolio
x.A = seq(from=-0.4, to=1.4, by=0.1)
x.B = 1 - x.A
mu.p = x.A*mu.A + x.B*mu.B
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p = sqrt(sig2.p)
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)),
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab = cex.val)
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
text(x=sig.p.min, y=mu.p.min, labels="Global min", pos=4, cex = cex.val)
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = 1.25)
text(x=sig.p.tan, y=mu.p.tan, labels="Tangency", pos=2, cex = cex.val)
points(sig.p.A, mu.p.A, type="b", col="red", pch=16, cex = cex.val)
points(sig.p.B, mu.p.B, type="b", col="blue", pch=16, cex = cex.val)
points(sig.p.tan.tbill, mu.p.tan.tbill, type="b", col="green", pch=16, cex = cex.val)
```


If you invest in only in asset B and T-bills then the Sharpe ratio
is $\mathrm{SR}_{B}=\frac{\mu_{B}-r_{f}}{\sigma_{B}}=0.217$, and the
CAL intersects the set of risky asset portfolios at point B. This
is clearly not the efficient set of portfolios. For example, you could
do uniformly better if you instead invest only in asset A and T-bills.
This gives us a Sharpe ratio of $\mathrm{SR}_{A}=\frac{\mu_{A}-r_{f}}{\sigma_{A}}=0.562$,
and the new CAL intersects the set of risky asset portfolios at point
A. However, you could do better still if you invest in T-bills and
some combination of assets A and B. Geometrically, it is easy
to see that the best you can do is obtained for the combination of
assets A and B such that the CAL is just tangent to the set
of risky asset portfolios. This point is labeled "Tangency"
on the graph and represents the *tangency portfolio*\index{Tangency portfolio} of assets
A and B. Portfolios of T-Bills and the tangency portfolio are
the set of efficient portfolios consisting of T-Bills, asset A
and asset B.

### Solving for the Tangency Portfolio

You can determine the proportions of each asset in the tangency portfolio
by finding the values of $x_{A}$ and $x_{B}$ that maximize the Sharpe
ratio of a portfolio that is on the envelope of the Markowitz bullet. Formally,
you solve the constrained maximization problem: 

\begin{align*}
 & \underset{x_{A},x_{B}}{\max}~\mathrm{SR}_{p}=\frac{\mu_{p}-r_{f}}{\sigma_{p}}~s.t.\\
\mu_{p} & =x_{A}\mu_{A}+x_{B}\mu_{B},\\
\sigma_{p}^{2} & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB},\\
1 & =x_{A}+x_{B}.
\end{align*}

Using the method of substitution with $x_{B} = 1 - x_{A}$ and rearranging, the above problem can be reduced to solving the unconstrained maximization:
$$
\underset{x_{A}}{\max}~\frac{x_{A}(\mu_{A}-r_{f})+(1-x_{A})(\mu_{B}-r_{f})}{\left(x_{A}^{2}\sigma_{A}^{2}+(1-x_{A})^{2}\sigma_{B}^{2}+2x_{A}(1-x_{A})\sigma_{AB}\right)^{1/2}}.
$$
This is a straightforward, albeit very tedious, calculus problem and
the solution can be shown to be (see Exercises): 
$$
\begin{aligned}
x_{A}^{tan} & =\frac{(\mu_{A}-r_{f})\sigma_{B}^{2}-(\mu_{B}-r_{f})\sigma_{AB}}{(\mu_{A}-r_{f})\sigma_{B}^{2}+(\mu_{B}-r_{f})\sigma_{A}^{2}-(\mu_{A}-r_{f}+\mu_{B}-r_{f})\sigma_{AB}},\\
x_{B}^{tan} & =1-x_{T}^{A}.
\end{aligned}
$$ {#eq-tangency-port}

::: {#exm-tangency-portfolio-for-example-data .example name="Tangency portfolio for example data"}

For the example data in @tbl-TableExampleData using @eq-tangency-port
with $r_{f}=0.03,$ you get $x_{A}^{tan}=0.4625$ and $x_{B}^{tan}=0.5375.$
The expected return, variance, standard deviation, and Sharpe ratio
of the tangency portfolio are: 

\begin{align*}
\mu_{tan}&=x_{A}^{tan}\mu_{A}+x_{B}^{tan}\mu_{B}\\
&=(0.4625)(0.175)+(0.5375)(0.055)=0.1105,\\
\sigma_{tan}^{2}&=\left(x_{A}^{tan}\right)^{2}\sigma_{A}^{2}+\left(x_{B}^{tan}\right)^{2}\sigma_{B}^{2}+2x_{A}^{tan} x_{B}^{tan}\sigma_{AB}\\
&=(0.4625)^{2}(0.06656)+(0.5375)^{2}(0.01323)+ 2(0.4625)(0.5375)(-0.004866)=0.01564,\\
\sigma_{tan}&=\sqrt{0.01564}=0.1251,\\
SR_{tan}&=\frac{\mu_{tan}-r_{f}}{\sigma_{tan}}=\frac{0.1105-0.03}{0.1251}=0.6434
\end{align*}

In R, the computations to compute the tangency portfolio are:

```{r}
#| echo: true
#| results: asis
# compute portfolio weights
top = (mu.A - r.f)*sig2.B - (mu.B - r.f)*sig.AB
bot = (mu.A - r.f)*sig2.B + (mu.B - r.f)*sig2.A - (mu.A - r.f + mu.B - r.f)*sig.AB
x.A.tan = top/bot
x.B.tan = 1 - x.A.tan
mu.p.tan = x.A.tan*mu.A + x.B.tan*mu.B
sig2.p.tan = x.A.tan^2 * sig2.A + x.B.tan^2 * sig2.B + 2*x.A.tan*x.B.tan*sig.AB
sig.p.tan = sqrt(sig2.p.tan)
SR.tan = (mu.p.tan - r.f)/sig.p.tan
data.tbl = as.data.frame(t(c(x.A.tan,x.B.tan,c(mu.p.tan,sig.p.tan,SR.tan))))
col.names = c("$x_{A}^{tan}$","$x_{B}^{tan}$", "$\\mu_{tan}$", 
              "$\\sigma_{tan}$","$\\mathrm{SR}_{tan}$" )
tbl.data = kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
cat(tbl.data)
```

\hfill{}$\blacksquare$
:::

### Mutual Fund Separation
\index{Mutual fund}
The efficient portfolios are combinations of the tangency portfolio (risky asset)
and the T-bill. Accordingly, using @eq-muTbillsPlusRisky and
@eq-sigmaTbillsPlusRisky the expected return and standard
deviation of any efficient portfolio are given by: 
$$
\mu_{p}^{e} =r_{f}+x_{T}(\mu_{tan}-r_{f}),
$$ {#eq-mup-efficient}
$$
\sigma_{p}^{e} =x_{T}\sigma_{tan},
$$ {#eq-sigp-efficient}
where $x_{T}$ represents the fraction of wealth invested in the tangency
portfolio, $1-x_{T}$ represents the fraction of wealth invested
in T-Bills, and $\mu_{tan}$ and $\sigma_{tan}$ are the expected return
and standard deviation of the tangency portfolio, respectively. This
important result is known as the *mutual fund separation theorem*.\index{Mutual fund separation theorem}
The tangency portfolio can be considered as a mutual fund (i.e. portfolio) of the two
risky assets, where the shares of the two assets in the mutual fund
are determined by the tangency portfolio weights ($x_{A}^{tan}$ and
$x_{B}^{tan}$ determined from @eq-tangency-port, and the T-bill
can be considered as a mutual fund of risk-free assets. The expected
return-risk trade-off of these portfolios is given by the line connecting
the risk-free rate to the tangency point on the efficient frontier
of risky asset only portfolios. 

The optimal combination of the tangency portfolio and the T-bill an investor will choose depends on the investor's risk preferences. If the investor is very risk averse, then she will choose a portfolio with low volatility which will be a portfolio with
very little weight in the tangency portfolio and a lot of weight in
the T-bill. This will produce a portfolio with an expected return
close to the risk-free rate and a variance that is close to zero.
If the investor can tolerate a large amount of risk, then she would
prefer a portfolio with highest expected return regardless of the
volatility. This portfolio may involve borrowing at the risk-free
rate (leveraging) and investing the proceeds in the tangency portfolio
to achieve a high expected return.

::: {#exm-efficient-portfolios-chosen-by-risk-averse-and-risk-tolerant-investors .example name="Efficient portfolios chosen by risk averse and risk tolerant investors"}

A highly risk averse investor may choose to put $10\%$ of her wealth
in the tangency portfolio and $90\%$ in the T-bill. Then she will
hold ($10\%)\times(46.25\%)=4.625\%$ of her wealth in asset A,
$(10\%)\times(53.75\%)=5.375\%$ of her wealth in asset B, and
$90\%$ of her wealth in the T-bill. The expected return on this portfolio
is: 
$$
\mu_{p}^{e}=r_{f}+0.10(\mu_{tan}-r_{f})=0.03+0.10(0.1105-0.03)=0.03805,
$$
and the standard deviation is, 
$$
\sigma_{p}^{e}=0.10\sigma_{tan}=0.10(0.1251)=0.01251.
$$
In @fig-IntroPortSafeRisky, this efficient portfolio
is labeled "Safe" . A very risk tolerant investor may
actually borrow at the risk-free rate and use these funds to leverage
her investment in the tangency portfolio. For example, suppose the
risk tolerant investor borrows 100\% of her wealth at the risk-free
rate and uses the proceed to purchase 200\% of her wealth in the tangency
portfolio. Then she would hold $(200\%)\times(46.25\%)=92.50\%$ of
her wealth in asset A, $(200\%)\times(53.75\%)=107.5\%$ in asset
B, and she would owe $100\%$ of her wealth to her lender. The
expected return and standard deviation on this portfolio is: 
\begin{align*}
\mu_{p}^{e} & =0.03+2(0.1105-0.03)=0.1910,\\
\sigma_{p}^{e} & =2(0.1251)=0.2501.
\end{align*}
In @fig-IntroPortSafeRisky, this efficient portfolio
is labeled "Risky" .
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-IntroPortSafeRisky
#| echo: false
#| results: hide
#| fig-cap: The efficient portfolio labeled safe has 10\% invested in the tangency portfolio
#|   and 90\% invested in T-Bills; the efficient portfolio labeled risky has 200\% invested
#|   in the tangency portfolio and -100\% invested in T-Bills.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
# safe portfolio: 10% in tangency and 90% in T-bills
mu.safe = r.f + 0.10*(mu.p.tan - r.f)
sig.safe = 0.10*sig.p.tan
# risky portfolio: 110% in tangency and -10% in T-bills
mu.risky = r.f + 2*(mu.p.tan - r.f)
sig.risky = 2*sig.p.tan
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)), cex.lab = cex.val,
     xlab=expression(sigma[p]), ylab=expression(mu[p]))
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
text(x=sig.p.min, y=mu.p.min, labels="Global min", pos=4, cex = cex.val)
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = cex.val)
text(x=sig.p.tan, y=mu.p.tan, labels="Tangency", pos=2, cex = cex.val)
points(x=sig.p.tan, y=mu.p.tan, pch=16, cex=2, col="green")
points(sig.p.tan.tbill, mu.p.tan.tbill, type="b", col="green", 
       pch=16, cex = cex.val)
points(x=sig.safe, y=mu.safe, pch=16, cex=2)
text(x=sig.safe, y=mu.safe, labels="Safe", pos=3, cex = cex.val)
points(x=sig.risky, y=mu.risky, pch=16, cex=2)
text(x=sig.risky, y=mu.risky, labels="Risky", pos=3, cex = cex.val)
```


### Interpreting Efficient Portfolios

As you have seen, efficient portfolios are those portfolios that have
the highest expected return for a given level of risk as measured
by portfolio standard deviation. For portfolios with expected returns
above the T-bill rate, efficient portfolios can also be characterized
as those portfolios that have minimum risk (as measured by portfolio
standard deviation) for a given target expected return. 

::: {#exm-finding-efficient-portfolios-with-target-expected-return-and-target-risk .example name="Finding efficient portfolios with target expected return and target risk"}

To illustrate, consider @fig-IntroPortEfficientB
which shows the portfolio frontier for two risky assets and the efficient
frontier for two risky assets plus T-Bills. Suppose an investor initially
holds all of his wealth in asset B. The expected return on this
portfolio is $\mu_{B}=0.055$, and the standard deviation (risk) is
$\sigma_{B}=0.115$. An efficient portfolio (combinations of the tangency
portfolio and T-bills) that has the same standard deviation (risk)
as asset B is given by the portfolio on the efficient frontier
that is directly above $\sigma_{B}=0.115$. To find the shares in
the tangency portfolio and T-bills in this portfolio recall from @eq-sigp-efficient
that the standard deviation of an efficient portfolio with $x_{T}$
invested in the tangency portfolio and $1-x_{T}$ invested in T-bills
is $\sigma_{p}^{e}=x_{T}\sigma_{tan}$. Since you want to find the efficient
portfolio with $\sigma_{p}^{e}=\sigma_{B}=0.115$, you solve: 
$$
x_{T}=\frac{\sigma_{B}}{\sigma_{tan}}=\frac{0.115}{0.1251}=0.9195,~x_{f}=1-x_{T}=0.08049.
$$
That is, if you invest $91.95\%$ of our wealth in the tangency portfolio
and $8.049\%$ in T-bills you will have a portfolio with the same standard
deviation as asset B. Since this is an efficient portfolio, the expected
return should be higher than the expected return on asset B. Indeed
it is since: 
$$
\mu_{p}^{e}=r_{f}+x_{T}(\mu_{tan}-r_{f})=0.03+0.9195(0.1105-0.03)=0.1040.
$$
Notice that by diversifying our holding into assets A, B and
T-bills you can obtain a portfolio with the same risk as asset B but
with almost twice the expected return! By moving from asset B into
an efficient portfolio you raise the Sharpe ratio of the investment
from $SR_{B}=0.217$ to $SR_{tan}=0.6434$. 

Next, consider finding an efficient portfolio that has the same expected
return as asset B. Visually, this involves finding the combination
of the tangency portfolio and T-bills that corresponds with the intersection
of a horizontal line with intercept $\mu_{B}=0.055$ and the line
representing efficient combinations of T-bills and the tangency portfolio.
To find the shares in the tangency portfolio and T-bills in this portfolio
recall from @eq-mup-efficient that the expected return of
an efficient portfolio with $x_{T}$ invested in the tangency portfolio
and $1-x_{T}$ invested in T-bills has expected return equal to $\mu_{p}^{e}=r_{f}+x_{T}(\mu_{tan}-r_{f})$.
Since you want to find the efficient portfolio with $\mu_{p}^{e}=\mu_{B}=0.055$
you solve: 
$$
x_{T}=\frac{\mu_{p}^{e}-r_{f}}{\mu_{tan}-r_{f}}=\frac{0.055-0.03}{0.1105-0.03}=0.3105,~x_{f}=1-x_{T}=0.6895.
$$
That is, if you invest $31.05\%$ of wealth in the tangency portfolio
and $68.95\%$ of our wealth in T-bills you have a portfolio with the
same expected return as asset B. Since this is an efficient portfolio,
the standard deviation (risk) of this portfolio should be lower than
the standard deviation on asset B. Indeed it is since: 
$$
\sigma_{p}^{e}=x_{T}\sigma_{tan}=0.3105(0.124)=0.03884.
$$
Notice how large the risk reduction is by forming an efficient portfolio.
The standard deviation on the efficient portfolio is almost three
times smaller than the standard deviation of asset B! Again, by
moving from asset B into an efficient portfolio you raise the Sharpe
ratio of the investment from $SR_{B}=0.217$ to $SR_{tan}=0.6434$
\hfill{}$\blacksquare$
:::

The above example illustrates two ways to interpret the benefits from
forming efficient portfolios. Starting from some benchmark portfolio,
you can fix standard deviation (risk) at the value for the benchmark
and then determine the gain in expected return from forming a diversified
portfolio^[The gain in expected return by investing in an efficient portfolio
abstracts from the costs associated with selling the benchmark portfolio
and buying the efficient portfolio.]. The gain in expected return has concrete meaning. Alternatively,
you can fix expected return at the value for the benchmark and then
determine the reduction in standard deviation (risk) from forming
a diversified portfolio. The meaning to an investor of the reduction
in standard deviation is not as clear as the meaning to an investor
of the increase in expected return. It would be helpful if the risk
reduction benefit can be translated into a number that is more interpretable
than the standard deviation. The concept of Value-at-Risk (VaR) provides
such a translation.

```{r}
#| label: fig-IntroPortEfficientB
#| echo: false
#| results: hide
#| fig-cap: The point e1 represents an efficient portfolio with the same standard deviation
#|   as asset B; the point e2 represents an efficient portfolio with the same expected
#|   returns as asset B.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
# find combination of T-bills and tangency that has same SD as asset B
x.t1 = sig.B/sig.p.tan
# x.t1
mu.p1.e = r.f + x.t1*(mu.p.tan - r.f)
# mu.p1.e
sig.p1.e = x.t1*sig.p.tan
# sig.p1.e

# find combination of T-bills and tangency that has same ER as asset B
x.t2 = (mu.B - r.f)/(mu.p.tan - r.f)
# x.t2
mu.p2.e = r.f + x.t2*(mu.p.tan - r.f)
# mu.p2.e
sig.p2.e = x.t2*sig.p.tan
# sig.p2.e

plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)),
     xlab=expression(sigma[p]), ylab=expression(mu[p]))
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
text(x=sig.p.min, y=mu.p.min, labels="Global min", pos=4, cex = cex.val)
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = cex.val)
text(x=sig.p.tan, y=mu.p.tan, labels="Tangency", pos=4, cex = cex.val)
points(x=sig.p.tan, y=mu.p.tan, pch=16, col="green", cex = cex.val)
points(sig.p.tan.tbill, mu.p.tan.tbill, type="b", col="green", pch=16, cex = cex.val)
segments(x0=sig.B, y0=0, x1=sig.B, y1=mu.p1.e, lwd=3, lty="dotted")
segments(x0=sig.B, y0=mu.p1.e, x1=0, y1=mu.p1.e, lwd=3, lty="dotted")
segments(x0=sig.p2.e, y0=0, x1=sig.p2.e, y1=mu.p2.e, lwd=3, lty="dotted")
segments(x0=0, y0=mu.p2.e, x1=sig.B, y1=mu.p2.e, lwd=3, lty="dotted")
text(x=sig.B, y=mu.p1.e, labels="e1", pos=3, cex = cex.val)
text(x=sig.p2.e, y=mu.B, labels="e2", pos=3, cex = cex.val)
```


### Efficient Portfolios and Value-at-Risk

Recall, the VaR of an investment is the (lower bound of) loss in investment
value over a given horizon with a stated probability. For example,
consider an investor who invests $W_{0}=$ $\$100,000$ in asset B
over the next year. Assuming that $R_{B}\sim N(0.055,(0.115)^{2})$
represents the annual simple return on asset B, the 5\% VaR is:
$$
\mathrm{VaR}_{B,0.05}=-q_{0.05}^{R_{B}}W_{0}=(0.055+0.115(-1.645))\cdot\$100,000=\$13,416.
$$
If an investor holds \$100,000 in asset B over the next year, then
there is a 5\% probability that he will lose \$13,416 or more. 


Now suppose the investor chooses to hold an efficient portfolio with
the same expected return as asset B. This portfolio consists of
$31.05\%$ in the tangency portfolio and $68.95\%$ in T-bills and
has a standard deviation equal to $0.03884$. Then $R_{p}\sim N(0.055,0.03884)$
and the 5\% VaR on the portfolio is: 
$$
\mathrm{VaR}_{p,0.05}=-q_{0.05}^{R_{p}}W_{0}=(0.055+0.03884(-1.645))\cdot\$100,000=\$884.
$$
Notice that the 5\% VaR for the efficient portfolio is almost fifteen
times smaller than the 5\% VaR for the investment in asset B. Since
VaR translates risk into a dollar figure, it is more interpretable
than standard deviation as a measure of investment risk.

## Application to Real World Portfolios {#sec-Application-to-Real}

The portfolio theory for two risky assets and a risk-free asset developed
in this chapter is not as limited as you might think if you realize
that a single risky asset can also be a portfolio of many assets,
such as a mutual fund of stocks or bonds, or an exchange traded fund
(ETF) of stocks or bonds. In particular, a very common asset allocation
for retirement portfolios is a simple two asset portfolio of in which
one risky asset is a mutual fund of stocks and the other (less) risky
asset is a mutual fund of U.S. Treasury bonds. 

In the **introcompfinr** package there is a data object called
`VanguardPrices` that contains monthly adjusted closing prices
on six Vanguard mutual funds over the period January, 1995 through
December, 2014.^[Details about these funds are available at (finance.yahoo.com). ]
The fund VFINX is a low cost fund designed to track the performance
of the S\&P 500 index. The S\&P 500 index is a value-weighted index of the 500 largest stocks (by market capitalization) traded on the NYSE.^[ETF alternatives to VFINX include State Street's SPY, Vanguard's VOO, and iShares IVV. See `www.eft.com`.] I will use this as my stock portfolio. The
fund VBLTX is a low cost fund constructed to track the performance
of a portfolio of long term (maturities greater than 10 years) U.S.
Government Bonds.^[ETF alternatives to VBLTX include iShares TLT and GOVT, and State Street SPTL. See `www.eft.com`.] I will use this as my bond portfolio. I will use the U.S. 3-month T-Bill rate as the 
risk-free asset. The following code creates the monthly simple return data for this example:

```{r}
#| echo: true

VanguardPrices = as.xts(VanguardPrices) 
stockBondPrices = VanguardPrices[, c("vfinx", "vbltx")] 
stockBondRetS = na.omit(Return.calculate(stockBondPrices, method="simple")) 
head(stockBondRetS, n=3)
```
I will construct efficient portfolios using the GWN model estimates
of expected returns, variances, and covariances over four sub-periods:
January, 1995 to January 2000; January, 2000 to January, 2005; January,
2005 to January, 2010; January, 2010 to December, 2014. The approximate
annualized T-Bill rates at the end of these periods are 0.05, 0.025,
0.001, and 0.001, respectively.^[Historical data on U.S. T-Bill rates is obtained from the Federal
Reserve Economic Data (FRED) available at the St. Louis Fed. See `fred.stlouisfed.org`]

```{r}
#| echo: true

# Define sub-periods
smpl1 = "1995-1::2000-1" 
smpl2 = "2000-1::2005-1" 
smpl3 = "2005-1::2010-1" 
smpl4 = "2010-1::2014-12" 
# Monthly risk-free rates for sub-periods 
r.f = c(0.05, 0.025, 0.001, 0.001)/12
```

Historical performance of the stock and bond portfolios over the four
sub-periods is illustrated in @fig-IntroPortStockBondPerf,
created using:^[The **PerformanceAnalytics** function `chart.CumReturns()`
is used to compute the equity curves for each asset.]

```{r}
#| label: fig-IntroPortStockBondPerf
#| echo: true
#| fig-cap: Equity curves for stock (vfinx) and bond (vbltx) portfolios over sub-periods.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2)) 
chart.CumReturns(stockBondRetS[smpl1], main="", wealth.index = TRUE, 
                 legend.loc = "topleft") 
chart.CumReturns(stockBondRetS[smpl2], main="", wealth.index = TRUE) 
chart.CumReturns(stockBondRetS[smpl3], main="", wealth.index = TRUE) 
chart.CumReturns(stockBondRetS[smpl4], main="", wealth.index = TRUE) 
```

The performance of stocks and bonds varies substantially over the
four sub-periods. In the first and fourth sub-periods, which cover
the dot-com boom and the recovery from the 2009 financial crisis,
stocks substantially outperformed bonds. However, in the second and
third sub-periods, which cover the dot-com bust and the 2009 financial
crisis, bonds outperformed stocks. 

The GWN model estimates of the monthly expected returns, volatilities,
and Sharpe ratios on the stock and bond portfolios are computed and
annualized using the square-root-of-time rule with the **PerformanceAnalytics** 
function `table.AnnualizedReturns()`.

```{r}
#| echo: true

# smpl1 = "1995-1::2000-1"
table.AnnualizedReturns(stockBondRetS[smpl1], Rf = r.f[1], geometric = FALSE) 
# smpl2 = "2000-1::2005-1"
table.AnnualizedReturns(stockBondRetS[smpl2], Rf = r.f[2], geometric = FALSE) 
# smpl3 = "2005-1::2010-1" 
table.AnnualizedReturns(stockBondRetS[smpl3], Rf = r.f[3], geometric = FALSE) 
# smpl4 = "2010-1::2014-12" 
table.AnnualizedReturns(stockBondRetS[smpl4], Rf = r.f[4], geometric = FALSE)
```

The annualized estimates confirm the relative performance of bonds
and stocks in the four sub-periods. In general, stock returns are about twice as volatile as bond returns. Bonds performed well in all sub-periods, with annualized mean returns between 6\% and 11\%. In contrast, stocks only performed well in sub-periods 1 and 4 and had negative annualized mean returns in periods 2 and 3. In terms of Sharpe ratios, stocks had annualized Sharpe ratios a bit over 1 in sub-periods 1 and 4 and bonds had annualized Sharpe ratios of 0.9 and 0.55 in sub-periods 2 and 3.^[Investments with annualized Sharpe ratios greater than one are typically considered good investments.] 

The sample correlations between the monthly stock and bond returns, by sub-period, are:

```{r}
#| echo: true
# smpl1 = "1995-1::2000-1"
cor(stockBondRetS[smpl1])[1,2] 
# smpl2 = "2000-1::2005-1"
cor(stockBondRetS[smpl2])[1,2] 
# smpl3 = "2005-1::2010-1" 
cor(stockBondRetS[smpl3])[1,2] 
# smpl4 = "2010-1::2014-12" 
cor(stockBondRetS[smpl4])[1,2]
```

Interestingly, the estimated correlation between stocks and bonds
also varies by sub-period. In sub-periods one and three, the estimated
correlations are positive, and, in sub-periods two and four, the estimated
correlations are negative. The observed low to negative correlation between stock and bond returns suggests good risk reduction possibilities through diversification.

```{r}
#| label: fig-IntroPortStockBondPortfolios
#| echo: false
#| fig-cap: Mean-variance efficient portfolios of stocks and bonds by sub-period. Efficient
#|   portfolios are shown in green and inefficient portfolios are shown in red.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))
smplVals = c(smpl1, smpl2, smpl3, smpl4)
gminVals = matrix(0, 4, 5)
tanVals = matrix(0, 4, 5)
rownames(gminVals) = rownames(tanVals) = c("smpl1", "smpl2", "smpl3","smpl4")
colnames(gminVals) = colnames(tanVals) = c("x.stock","x.bond","mu","sd","SR")
for (i in 1:length(smplVals)) {
muhat.vals = 12*apply(stockBondRetS[smplVals[i]], 2, mean) 
sigmahat.vals = sqrt(12)*apply(stockBondRetS[smplVals[i]], 2, sd) 
covhat = 12*cov(stockBondRetS[smplVals[i]])[1,2] 
corhat = covhat/(sigmahat.vals[1]*sigmahat.vals[2])
mu.vfinx = muhat.vals[1] 
mu.vbltx = muhat.vals[2] 
sig2.vfinx = sigmahat.vals[1]^2 
sig2.vbltx = sigmahat.vals[2]^2 
covmat.a = 12*cov(stockBondRetS[smplVals[i]])
x.vfinx = seq(from=0, to=1, by=0.1) 
x.vbltx = 1 - x.vfinx 
mu.p = x.vfinx*mu.vfinx + x.vbltx*mu.vbltx 
sig2.p = x.vfinx^2 * sig2.vfinx + x.vbltx^2 * sig2.vbltx + 2*x.vfinx*x.vbltx*covhat 
sig.p = sqrt(sig2.p)
# plot portfolio frontier 
cex.val = 1 
if (i == 1) {   
colvals = c(rep("red", 2), rep("green", 10)) 
} 
if (i == 2 || i == 3) {   
colvals = c(rep("green", 3), rep("red", 9)) 
} 
if (i == 4) {   colvals = c(rep("red", 3), rep("green", 9)) 
}
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val, main=smplVals[i],     
     ylim=c(-0.05, 0.27), xlim=c(0, 0.25),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab=cex.val,      
     col=colvals) 
abline(h=0)
points(0, 12*r.f[i], pch=16, cex = cex.val, col="green") 
text(0, 12*r.f[i], labels=expression(r[f]), pos=4, cex = cex.val) 
text(x=sqrt(sig2.vfinx), y=mu.vfinx, labels="stocks", pos=4, cex = cex.val) 
text(x=sqrt(sig2.vbltx), y=mu.vbltx, labels="bonds", pos=4, cex = cex.val)
# tan 
tan.port = tangency_portfolio(muhat.vals, covmat.a, 12*r.f[i]) 
SR.tan = (tan.port$er_port - 12*r.f[i])/tan.port$sd_port
tanVals[i, ] = c(tan.port$weight_vec,tan.port$er_port,tan.port$sd_port,SR.tan)
points(tan.port$sd_port, tan.port$er_port, pch=16, col="blue", cex = cex.val) 
text(tan.port$sd_port, tan.port$er_port, labels = "Tangency", pos = 3, cex = cex.val)
abline(a=12*r.f[i], b=SR.tan, col="green", lwd=2) 
# global min
gmin.port = globalmin_portfolio(muhat.vals, covmat.a)
SR.gmin = (gmin.port$er_port - 12*r.f[i])/gmin.port$sd_port
gminVals[i, ] = c(gmin.port$weight_vec,gmin.port$er_port,gmin.port$sd_port,SR.gmin)
}
# par(mfrow=c(1,1))
```

Using the sub-period GWN model estimates, I compute risky asset only
efficient portfolios and efficient portfolios of T-Bills (risk-free
asset) and the tangency portfolios. The resulting portfolios are illustrated
in @fig-IntroPortStockBondPortfolios. In sub-periods
one and four, stocks are on the risky asset only efficient frontier
and, in sub-periods two and three, bonds are on the risky asset only
efficient frontier. The sub-period global minimum variance and tangency
portfolios are listed in Tables @tbl-tableStocksBondsMin and @tbl-tableStocksBondsTan. The global
minimum variance portfolios are similar across sub-periods, with higher
weights in bonds than stocks. The tangency portfolios, however, are
quite different across sub-periods. In sub-periods one and four, stocks
have weights 79\% and 40%, respectively. In sub-periods two and three
bonds essentially have 100\% weights, respectively, and stocks are slighly shorted. 

```{r}
#| echo: false
#| label: tbl-tableStocksBondsMin
#| tbl-cap: "Sub-period global minimum variance portfolios"
data.tbl = as.data.frame(gminVals)
rownames(data.tbl) = smplVals
col.names = c("\\(x_{stock}\\)","\\(x_{bond}\\)","\\(\\mu_{min}\\)",
                       "\\(\\sigma_{min}\\)","\\(SR_{min}\\)")
kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
```



```{r}
#| echo: false
#| label: tbl-tableStocksBondsTan
#| tbl-cap: "Sub-period tangency portfolios"
data.tbl = as.data.frame(tanVals)
rownames(data.tbl) = smplVals
col.names = c("\\(x_{stock}\\)","\\(x_{bond}\\)","\\(\\mu_{tan}\\)",
                       "\\(\\sigma_{tan}\\)","\\(SR_{tan}\\)")
kbl(data.tbl, col.names=col.names) %>%
   kable_styling(full_width=FALSE)
```
This example shows the clear gains that can be made by forming efficient portfolios relative to holding stock or bonds by themselves. The results, however, are a bit misleading because the efficient portfolios formed in each sub-period use all of the data over the sub-period to estimate expected returns, volatilities, and covariances. This means that the computed efficient portfolios cannot be computed at the beginning of each sub-period. A more practical exercise would use the previous sub-period to compute GWN model estimates to be used in the subsequent sub-period. This method is explored in the end-of-chapter exercises.

## Further Reading: Introduction to Portfolio Theory

Markowitz's mean-variance portfolio theory is described in detail
in [@markowitz-1959], [@markowitz-1987]. Most MBA level corporate finance
and investments textbooks have chapters on mean-variance portfolio
theory. Good intermediate level treatments of portfolio theory from
a quantitative perspective are given in [@benninga-2000], [@bkm-2013] and [@egbg-2014]. 

## Appendix: Review of Optimization and Constrained Optimization

Consider the function of a single variable: 
$$
y=f(x)=x^{2}.
$$
Clearly the minimum of this function occurs at the point $x=0$. Using
calculus, you find the minimum by solving: 
$$
\min_{x}~y=x^{2}.
$$
The first order (necessary) condition for a minimum is: 
$$
0=\frac{d}{dx}f(x)=\frac{d}{dx}x^{2}=2x
$$
and solving for $x$ gives $x=0$. The second order condition for
a minimum is, 
$$
0<\frac{d^{2}}{dx}f(x),
$$
and this condition is clearly satisfied for $f(x)=x^{2}$. 


Next, consider the function of two variables: 
$$
y=f(x,z)=x^{2}+z^{2}
$$ {#eq-f-xz}
This function looks like a salad bowl whose bottom is at $x=0$ and
$z=0$. To find the minimum of @eq-f-xz, you solve: 
$$
\min_{x,z}y=x^{2}+z^{2},
$$
and the first order necessary conditions are, 
\begin{align*}
0 & =\frac{\partial y}{\partial x}=2x,\\
0 & =\frac{\partial y}{\partial z}=2z.
\end{align*}
Solving these two linear equations gives $x=0$ and $z=0$. 

Now suppose you want to minimize @eq-f-xz subject to the linear
constraint: 
$$
x+z=1.
$$ {#eq-restriction}
The minimization problem is now a * constrained minimization*:
\begin{align*}
\min_{x,z}y & =x^{2}+z^{2}\textrm{ subject to }(s.t.)\\
x+z & =1.
\end{align*}
Given the constraint $x+z=1$, the function @eq-f-xz is no
longer minimized at the point $(x,z)=(0,0)$ because this point does
not satisfy $x+z=1$. One simple way to solve this problem is to substitute
the restriction @eq-restriction into the function @eq-f-xz
and reduce the problem to a minimization over one variable. To illustrate,
use the restriction @eq-restriction to solve for $z$ as:
$$
z=1-x.
$$ {#eq-z}
Now substitute @eq-z into @eq-f-xz giving: 
$$
y=f(x,z)=f(x,1-x)=x^{2}+(1-x)^{2}.
$$ {#eq-res-fun}
The function @eq-res-fun satisfies the restriction @eq-restriction
by construction. The constrained minimization problem now becomes:
$$
\min_{x}y=x^{2}+(1-x)^{2}.
$$
The first order conditions for a minimum are: 
$$
0=\frac{d}{dx}(x^{2}+(1-x)^{2})=2x-2(1-x)=4x-2,
$$
and solving for $x$ gives $x=1/2$. To solve for $z$, use @eq-z
to give $z=1-(1/2)=1/2$. Hence, the solution to the constrained minimization
problem is $(x,z)=(1/2,1/2)$. 


Another way to solve the constrained minimization is to use the method
of *Lagrange multipliers*. This method augments the function
to be minimized with a linear function of the constraint in homogeneous
form. The constraint @eq-restriction in homogenous form is:
$$
x+z-1=0.
$$
The augmented function to be minimized is called the *Lagrangian*
and is given by: 
$$
L(x,z,\lambda)=x^{2}+z^{2}-\lambda(x+z-1).
$$
The coefficient on the constraint in homogeneous form, $\lambda$,
is called the *Lagrange multiplier*. It measures the cost, or
shadow price, of imposing the constraint relative to the unconstrained
problem. The constrained minimization problem to be solved is now:
$$
\min_{x,z,\lambda}~L(x,z,\lambda)=x^{2}+z^{2}+\lambda(x+z-1).
$$
The first order conditions for a minimum are: 
\begin{align*}
0 & =\frac{\partial L(x,z,\lambda)}{\partial x}=2x+\lambda,\\
0 & =\frac{\partial L(x,z,\lambda)}{\partial z}=2z+\lambda,\\
0 & =\frac{\partial L(x,z,\lambda)}{\partial\lambda}=x+z-1.
\end{align*}
The first order conditions give three linear equations in three unknowns.
Notice that the first order condition with respect to $\lambda$ imposes
the constraint. The first two conditions give: 
$$
2x=2z=-\lambda,
$$
or, 
$$
x=z.
$$
Substituting $x=z$ into the third condition gives: 
$$
2z-1=0
$$
or, 
$$
z=1/2.
$$
The final solution is $(x,y,\lambda)=(1/2,1/2,-1)$. 

The Lagrange multiplier, $\lambda$, measures the marginal cost, in
terms of the value of the objective function, of imposing the constraint.
Here, $\lambda=-1$ which indicates that imposing the constraint $x+z=1$
reduces the objective function. To understand the role of the Lagrange
multiplier better, consider imposing the constraint $x+z=0$. Notice
that the unconstrained minimum achieved at $x=0,z=0$ satisfies this
constraint. Hence, imposing $x+z=0$ does not cost anything and so
the Lagrange multiplier associated with this constraint should be
zero. To confirm this, you solve the problem: 
$$
\min_{x,z,\lambda}~L(x,z,\lambda)=x^{2}+z^{2}+\lambda(x+z-0).
$$
The first order conditions for a minimum are: 
\begin{align*}
0 & =\frac{\partial L(x,z,\lambda)}{\partial x}=2x-\lambda,\\
0 & =\frac{\partial L(x,z,\lambda)}{\partial z}=2z-\lambda,\\
0 & =\frac{\partial L(x,z,\lambda)}{\partial\lambda}=x+z.
\end{align*}
The first two conditions give: 
$$
2x=2z=-\lambda
$$
or, 
$$
x=z.
$$
Substituting $x=z$ into the third condition gives: 
$$
2z=0,
$$
or, 
$$
z=0.
$$
The final solution is $(x,y,\lambda)=(0,0,0)$. Notice that the Lagrange
multiplier, $\lambda$, is equal to zero in this case.

## Exercises

::: {#exr-01 .exercise}
Annual estimates of the GWN model parameters for
Boeing and Microsoft are given in the table below: 

+----------------+----------+-------------+
|                |  Boeing  |  Microsoft  |
+================+==========+=============+
|  $\mu$         |  0.149   |  0.331      |
+----------------+----------+-------------+
|  $\sigma^{2}$  |  0.069   |  0.136      |
+----------------+----------+-------------+
|  $\rho$        |  -0.008  |  -0.008     |
+----------------+----------+-------------+

1.  Create the following portfolios.
    -  Combinations of Boeing and Microsoft (with $x_{\text{{boeing}}}=-1,-0.9,...,2$ and $x_{\text{{msft}}}=1-x_{\text{{boeing}}}$)
    -  Combinations of Boeing and T-bills (with $x_{\text{{boeing}}}=0,0.1,...,2$)
    -  Combinations of Microsoft and T-bills (with $x_{\text{{msft}}}=0,0.1,...,2$)
    -  Use an annual risk-free rate of $3\%$ per year for the T-bill. For each portfolio, compute $E(R_{p})$, var$(R_{p})$ and SD$(R_{p})$ using the appropriate formulas. For each portfolio, plot $E(R_{p})$ vs. SD$(R_{p})$ and put these values on the same graph. Compute the Sharpe's slope for Boeing and Microsoft. Which asset has the highest slope value?

2.  Compute the global minimum variance portfolio using the analytical formula @eq-minvar-port.
    -  Make a bar chart showing the weights of Boeing and Microsoft in global minimum variance portfolio.
    -  Compute $E(R_{p})$, var$(R_{p})$ and SD$(R_{p})$ for the tangency portfolio.
    -  Compute Sharpe's slope for the tangency portfolio.
    -  Indicate the location of the tangency portfolio on the graph you created previously in question 1.

3.  Using a risk-free rate of $3\%$ per year for the T-bill, compute the tangency portfolio using the analytic formula @eq-tangency-port.
    -  Make a bar chart showing the weights of Boeing and Microsoft in the tangency portfolio.
    -  Compute $E(R_{p})$, var$(R_{p})$ and SD$(R_{p})$ for the tangency portfolio.
    -  Compute the Sharpe's slope for the tangency portfolio. 
    -  Indicate the location of the tangency portfolio on the graph you created previously in question 1.

4.  Consider a portfolio that has $10\%$ in the tangency portfolio and $90\%$ in T-bills.
    -  In this portfolio, what is the percent invested in Boeing and what is the percent invested in Microsoft? Give a bar chart showing the percents invested in T-bills, Boeing and Microsoft.
    -  Compute $E(R_{p})$, var$(R_{p})$ and SD$(R_{p})$ for this portfolio.
    -  Compute Sharpe's slope for this portfolio.
    -  Indicate the location of the tangency portfolio on the graph you created previously in question 1.

5.  Find the efficicient portfolio (combination of T-bills and tangency portfolio) that has the same risk (SD) as Microsoft.
    -  In this portfolio, what is the percent invested in Boeing and what is the percent invested in Microsoft? Give a bar chart showing the percent invested in T-bills, Boeing and Microsoft.
    -  Compute $E(R_{p})$, var$(R_{p})$ and SD$(R_{p})$ for this portfolio.
    -  Compute Sharpe's slope for this portfolio.
    -  Indicate the location of the tangency portfolio on the graph you created previoulsy in question 1.
:::

::: {#exr-real-world-portfolios-revisited .exercise name="Real world portfolios revisited"}

* Re-do the real-world portfolio analysis using forward looking estimates: For each 5-year period use the previous 5-year period to estimate GWN model parameters and then do asset allocation. 

:::

