# Portfolio Theory with Matrix Algebra {#sec-Portfolio-Theory-with-Matrix-Algebra}

Updated: September 11, 2026
Copyright © Eric Zivot 2026


In this chapter, I extend the mean-variance portfolio theory introduced
in chapter @sec-Introduction-to-Portfolio-Theory to handle
more than two risky assets. This extension allows the theory to be
applied to the analysis of real world portfolios that consist of many
risky assets as well as a risk-free asset. When working with large
portfolios, the algebra of representing portfolio expected returns
and variances becomes extremely cumbersome. As shown in chapter @sec-Matrix-Algebra-Review,
the use of matrix algebra can greatly simplify many of these computations.
Matrix algebra formulations are also very useful when it comes to do actual computations on the computer. 
The matrix algebra formulas are easy to translate into matrix programming languages like R. Popular
spreadsheet programs like Microsoft Excel, which are the workhorse
programs of many financial houses, can also handle basic matrix calculations but with more limited functionality.
All of this makes it worthwhile to become familiar with matrix techniques
for portfolio calculations.

This chapter is laid out as follows. In @sec-Portfolios-with-N-risky-assets 
I describe portfolios with $N$ risky assets using matrix algebra. I elaborate the 
concept of portfolio risk diversification using calculations for large portfolios.
In @sec-Determining-the-Global-Minimum-Variance-Portfolio-Using-Matrix-Algebra through 
@ I cover the matrix algebra calculations required
for determining mean-variance efficient portfolios. I give Explicit formulas
for the global minimum variance portfolio, a minimum variance
portfolio that achieves a specified target expected return, and the
tangency portfolio. I also present a simple algorithm to easily trace out the risky
asset only efficient frontier. In Section 6 I discuss
some practical problems associated with very large portfolios, and in 
section 7 I summarize the **introcompfinr** functions for portfolio
analysis. In @sec-Real-World-Application-to-Vanguard-Mutual-Funds I give a real-world 
application of the theory to asset allocation among Vanguard mutual funds.


The examples I show in this chapter use the **corrplot**, **introcompfinr**
and **PerformanceAnalytics** packages. Make sure these packages
are installed and loaded in R before replicating the chapter examples.

```{r}
#| include: false
suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```

## Portfolios with $N$ Risky Assets {#sec-Portfolios-with-N-risky-assets}
\index{Porfolios with multiple risky assets}
Consider a portfolio with $N$ risky assets, where $N$ can be a large
number (e.g., $N=1,000)$. Let $R_{i}$ $(i=1,\ldots,N)$ denote the
simple return on asset $i$ (over some fixed time horizon such as
one year) and assume that the constant expected return (GWN) model
holds for all assets: 
\begin{align*}
R_{i} & \sim GWN(\mu_{i},\sigma_{i}^{2}),\\
\mathrm{cov}(R_{i},R_{j}) & =\sigma_{ij}.
\end{align*}

Let $x_{i}$ denote the share of wealth invested in asset $i$ $(i=1,\ldots,N)$,
and assume that all wealth is invested in the $N$ assets so that
$\sum_{i=1}^{N}x_{i}=1.$ Assume that short sales are allowed so that
some values of $x_{i}$ can be negative. The portfolio return, $R_{p,x},$
is the random variable $R_{p,x}=\sum_{i=1}^{N}x_{i}R_{i}.$ The subscript
"$x$" indicates that the portfolio is constructed
using the "x-weights" $x_{1},x_{2},\ldots,x_{N}$. The expected
return on the portfolio is: 
$$
\mu_{p,x}=\sum_{i=1}^{N}x_{i}\mu_{i},
$$
and the variance of the portfolio return is: 
$$
\sigma_{p,x}^{2}=\sum_{i=1}^{N}x_{i}^{2}\sigma_{i}^{2}+2\sum_{i=1}^{N}\sum_{j\neq i}x_{i}x_{j}\sigma_{ij}.
$$
Notice that variance of the portfolio return depends on $N$ variance
terms and $N(N-1)$ covariance terms. Hence, with $N$ assets there
are many more covariance terms than variance terms contributing to
portfolio variance. For example, with $N=100$ there are $100$ variance
terms and $100\times99=9900$ covariance terms. With $N$ assets,
the algebra representing the portfolio return and risk characteristics
is cumbersome, especially for the variance. You can greatly simplify
the portfolio algebra using matrix notation, and this was previewed
in Chapter 3.

### Portfolio return and risk characteristics using matrix notation

Define the following $N\times1$ column vectors containing the asset
returns and portfolio weights: 
$$
\mathbf{R}=\left(\begin{array}{c}
R_{1}\\
\vdots\\
R_{N}
\end{array}\right),~\mathbf{x}=\left(\begin{array}{c}
x_{1}\\
\vdots\\
x_{N}
\end{array}\right).
$$
In matrix notation we can lump multiple returns in a single vector
which we denote by $\mathbf{R}$. Since each of the elements in $\mathbf{R}$
is a random variable we call $\mathbf{R}$ a *random vector*.
The probability distribution of the random vector $\mathbf{R}$ is
simply the joint distribution of the elements of $\mathbf{R}$. In
the GWN model all returns are jointly normally distributed and this
joint distribution is completely characterized by the means, variances
and covariances of the returns. We can easily express these values
using matrix notation as follows. The $N\times1$ vector of portfolio
expected values is: 
$$
E[\mathbf{R}]=E\left[\left(\begin{array}{c}
R_{1}\\
\vdots\\
R_{N}
\end{array}\right)\right]=\left(\begin{array}{c}
E[R_{1}]\\
\vdots\\
E[R_{N}]
\end{array}\right)=\left(\begin{array}{c}
\mu_{1}\\
\vdots\\
\mu_{N}
\end{array}\right)=\boldsymbol{\mu},
$$
and the $N\times N$ covariance matrix of returns is: 

\begin{align*}
\mathrm{var}(\mathbf{R})& =\left(\begin{array}{cccc}
\mathrm{var}(R_{1}) & \mathrm{cov}(R_{1},R_{2}) & \cdots & \mathrm{cov}(R_{1},R_{N})\\
\mathrm{cov}(R_{1},R_{2}) & \mathrm{var}(R_{2}) & \cdots & \mathrm{cov}(R_{2},R_{N})\\
\vdots & \vdots & \ddots & \vdots\\
\mathrm{cov}(R_{1},R_{N}) & \mathrm{cov}(R_{2},R_{N}) & \cdots & \mathrm{var}(R_{N})
\end{array}\right) \\
&=\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \cdots & \sigma_{1N}\\
\sigma_{12} & \sigma_{2}^{2} & \cdots & \sigma_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1N} & \sigma_{2N} & \cdots & \sigma_{N}^{2}
\end{array}\right)=\boldsymbol{\Sigma}.
\end{align*}

Notice that the covariance matrix is symmetric (elements off the diagonal
are equal so that $\boldsymbol{\Sigma}=\boldsymbol{\Sigma}^{\prime}$, where $\boldsymbol{\Sigma}^{\prime}$
denotes the transpose of $\boldsymbol{\Sigma}$) since $\mathrm{cov}(R_{i},R_{j})=\mathrm{cov}(R_{j},R_{i})$
for $i\neq j$. It will be positive definite provided no pair of assets
is perfectly correlated ($|\rho_{ij}|\neq1$ for all $i\neq j$) and
no asset has a constant return ($\sigma_{i}^{2}>0$ for all $i$). 

The return on the portfolio using matrix notation is: 
$$
R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}=(x_{1},\cdots,x_{N})\cdot\left(\begin{array}{c}
R_{1}\\
\vdots\\
R_{N}
\end{array}\right)=x_{1}R_{1}+\cdots+x_{N}R_{N}.
$$
Similarly, the expected return on the portfolio is: 
$$
\mu_{p,x}=E[\mathbf{x}^{\prime}\mathbf{R]=x}^{\prime}E[\mathbf{R}]=\mathbf{x}^{\prime}\boldsymbol{\mu}=(x_{1},\ldots,x_{N})\cdot\left(\begin{array}{c}
\mu_{1}\\
\vdots\\
\mu_{N}
\end{array}\right)=x_{1}\mu_{1}+\cdots+x_{N}\mu_{N}.
$$
The variance of the portfolio is: 
\begin{align*}
\sigma_{p,x}^{2} & =\mathrm{var}(\mathbf{x}^{\prime}\mathbf{R})=\mathbf{x}^{\prime}\boldsymbol{\Sigma}\mathbf{x}=(x_{1},\ldots,x_{N})\cdot\left(\begin{array}{ccc}
\sigma_{1}^{2} & \cdots & \sigma_{1N}\\
\vdots & \ddots & \vdots\\
\sigma_{1N} & \cdots & \sigma_{N}^{2}
\end{array}\right)\left(\begin{array}{c}
x_{1}\\
\vdots\\
x_{N}
\end{array}\right)\\
 & =\sum_{i=1}^{N}x_{i}^{2}\sigma_{i}^{2}+2\sum_{i=1}^{N}\sum_{j\neq i}x_{i}x_{j}\sigma_{ij}.
\end{align*}
The condition that the portfolio weights sum to one can be expressed
as: 
$$
\mathbf{x}^{\prime}\mathbf{1}=(x_{1},\ldots,x_{N})\cdot\left(\begin{array}{c}
1\\
\vdots\\
1
\end{array}\right)=x_{1}+\cdots+x_{N}=1,
$$
where $\mathbf{1}$ is a $N\times1$ vector with each element equal
to 1. 


Consider another portfolio with weights $\mathbf{y}=(y_{1},\ldots,y_{N})^{\prime}\neq\mathbf{x}$.
The return on this portfolio is $R_{p,y}=\mathbf{y}^{\prime}\mathbf{R}$.
Later on you will need to compute the covariance between the return
on portfolio $\mathbf{x}$ and the return on portfolio $\mathbf{y}$,
$\mathrm{cov}(R_{p,x},R_{p,y})$. Using matrix algebra, this covariance
can be computed as: 
$$
\sigma_{xy}=\mathrm{cov}(R_{p,x},R_{p,y})=\mathrm{cov}(\mathbf{x}^{\prime}\mathbf{R},\mathbf{y}^{\prime}\mathbf{R})=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{y}=(x_{1},\ldots,x_{N})\cdot\left(\begin{array}{ccc}
\sigma_{1}^{2} & \cdots & \sigma_{1N}\\
\vdots & \ddots & \vdots\\
\sigma_{1N} & \cdots & \sigma_{N}^{2}
\end{array}\right)\left(\begin{array}{c}
y_{1}\\
\vdots\\
y_{N}
\end{array}\right).
$$

::: {#exm-three-asset-example-data .example name="Three asset example data"}

To illustrate portfolio calculations in R, table @tbl-Table-ThreeAssetExample
gives example values on monthly means, variances and covariances for
the simple returns on Microsoft, Nordstrom and Starbucks stock based
on sample statistics computed over the five-year period January, 1995
through January, 2000. The risk-free asset is the monthly T-Bill with
rate $r_{f}=0.005$. 


+---------+---------+------------+-------------+-------------+-------------+-------------+
|Asset $i$|$\mu_{i}$|$\sigma_{i}$|  Pair (i,j) |             |             |$\sigma_{ij}$|
+=========+=========+============+=============+=============+=============+=============+
| MSFT    |  0.0427 |  0.1000    | (MSFT,NORD) |             |             |  0.0018     |
+---------+---------+------------+-------------+-------------+-------------+-------------+
| NORD    |  0.0015 |  0.1044    | (MSFT,SBUX) |             |             |  0.0011     |
+---------+---------+------------+-------------+-------------+-------------+-------------+
| SBUX    |  0.0285 |  0.1411    | (NORD,SBUX) |             |             |  0.0026     |
+---------+---------+------------+-------------+-------------+-------------+-------------+
| T-Bill  |  0.005  |  0         |(T-Bill,MSFT)|(T-Bill,NORD)|(T-Bill,SBUX)|  0          |
+---------+---------+------------+-------------+-------------+-------------+-------------+
: Three asset example data. {#tbl-Table-ThreeAssetExample}

The example data in matrix notation is: 

\begin{align*}
\boldsymbol{\mu} & =\left(\begin{array}{c}
\mu_{MSFT}\\
\mu_{NORD}\\
\mu_{SBUX}
\end{array}\right)=\left(\begin{array}{c}
0.0427\\
0.0015\\
0.0285
\end{array}\right),\\
\boldsymbol{\Sigma} & =\left(\begin{array}{ccc}
\mathrm{var}(R_{MSFT}) & \mathrm{cov}(R_{MSFT,NORD}) & \mathrm{cov}(R_{MSFT,SBUX})\\
\mathrm{cov}(R_{MSFT,NORD}) & \mathrm{var}(R_{NORD}) & \mathrm{cov}(R_{NORD,SBUX})\\
\mathrm{cov}(R_{MSFT,SBUX}) & \mathrm{cov}(R_{NORD,SBUX}) & \mathrm{var}(R_{SBUX})
\end{array}\right)\\
&=\left(\begin{array}{ccc}
0.0100 & 0.0018 & 0.0011\\
0.0018 & 0.0109 & 0.0026\\
0.0011 & 0.0026 & 0.0199
\end{array}\right).
\end{align*}

In R, the example data is created using:

```{r}
#| echo: true

asset.names <- c("MSFT", "NORD", "SBUX")
mu.vec = c(0.0427, 0.0015, 0.0285)
names(mu.vec) = asset.names
sigma.mat = matrix(c(0.0100, 0.0018, 0.0011,
		   0.0018, 0.0109, 0.0026,
		   0.0011, 0.0026, 0.0199),
		 nrow=3, ncol=3)
dimnames(sigma.mat) = list(asset.names, asset.names)
r.f=0.005
```

The values of $\mu_{i}$ and $\sigma_{i}$ ($i=\mathrm{MSFT},\,\mathrm{NORD},\,\mathrm{SBUX}$)
are shown in @fig-PortMatrixExampleData, created with: 

```{r}
#| label: fig-PortMatrixExampleData
#| echo: true
#| fig-cap: Risk-return characteristics of example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

sd.vec = sqrt(diag(sigma.mat)) 
cex.val = 1.5
plot(sd.vec, mu.vec,  ylim=c(0, 0.06), xlim=c(0, 0.20), 
     ylab=expression(mu[p]), xlab=expression(sigma[p]), 
     pch=16, col="blue", cex=cex.val, cex.lab=1.5)      
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val) 
text(0, r.f, labels=expression(r[f]), pos=4, cex = cex.val)
```


Clearly, Microsoft provides the best risk-return trade-off (i.e.,
has the highest Sharpe ratio) and Nordstorm provides with worst.
\hfill{}$\blacksquare$
:::

::: {#exm-portfolio-computations-in-r .example name="Portfolio computations in R"}

Consider an equally weighted portfolio with $x_{MSFT}=x_{NORD}=x_{SBUX}=1/3$.
This portfolio has return $R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}$
where $\mathbf{x}=(1/3,1/3,1/3)^{\prime}$. Using R, the portfolio
mean and variance are:

```{r}
#| echo: true

x.vec = rep(1,3)/3
names(x.vec) = asset.names
mu.p.x = crossprod(x.vec,mu.vec)
sig2.p.x = t(x.vec)%*%sigma.mat%*%x.vec
sig.p.x = sqrt(sig2.p.x)
c(mu.p.x, sig.p.x)
```

Next, consider another portfolio with weight vector $\mathbf{y}=(y_{MSFT},y_{NORD},y_{SBUX})^{\prime}=(0.8,0.4,-0.2)^{\prime}$
and return $R_{p,y}=\mathbf{y}^{\prime}\mathbf{R}$. The mean and
volatility of this portfolio are: 

```{r}
#| echo: true

y.vec = c(0.8, 0.4, -0.2) 
names(y.vec) = asset.names 
mu.p.y = crossprod(y.vec,mu.vec) 
sig2.p.y = t(y.vec)%*%sigma.mat%*%y.vec 
sig.p.y = sqrt(sig2.p.y)
c(mu.p.y, sig.p.y)
```

The covariance and correlation between $R_{p,x}$ and $R_{p,y}$ are:

```{r}
#| echo: true

sig.xy = t(x.vec)%*%sigma.mat%*%y.vec
rho.xy = sig.xy/(sig.p.x*sig.p.y)
c(sig.xy,rho.xy)
```

The return and risk characteristics of the three assets together with
portfolios **x** and **y** are illustrated in @fig-PortMatrixExampleData2.
Notice that the equally weighted portfolio, portfolio **x**,
has the smallest volatility.

```{r}
#| label: fig-PortMatrixExampleData2
#| echo: false
#| fig-cap: Risk-return characteristics of example data and portfolios with weight vectors
#|   $\mathbf{x}=(0.333,0.333,0.333)^{\prime}$ and $\mathbf{y}=(0.8,0.4,-0.2)^{\prime}$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

cex.val = 1.15 
plot(sd.vec, mu.vec,  ylim=c(0, 0.06), xlim=c(0, 0.20), ylab=expression(mu[p]),
     xlab=expression(sigma[p]), pch=16, col="blue", cex=2.5, cex.lab=1.5)
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val) 
points(sig.p.x, mu.p.x, pch=16, col="black", cex=2.5)
text(sig.p.x, mu.p.x, labels="X", pos=4, cex = cex.val) 
points(sig.p.y, mu.p.y, pch=16, col="black", cex=2.5) 
text(sig.p.y, mu.p.y, labels="Y", pos=4, cex = cex.val) 
text(0, r.f, labels=expression(r[f]), pos=2, cex = cex.val)
```

\hfill{}$\blacksquare$
:::

::: {#exm-risk-return-characteristics-of-random-portfolios .example name="Risk-return characteristics of random portfolios"}

@fig-PortMatrixExampleData2 shows that the risk-return
points associated with portfolios of three assets do not fall nicely
on a bullet-shaped line, as they do with two asset portfolios. In
fact, the set of risk-return points generated by varying the three
portfolio weights is a solid area. To see this, consider creating
a set of 200 randomly generated three-asset portfolios whose weights
sum to one:

```{r}
#| echo: true

set.seed(123) 
x.msft = runif(200, min=-1.5, max=1.5) 
x.nord = runif(200, min=-1.5, max=1.5) 
x.sbux = 1 - x.msft - x.nord 
head(cbind(x.msft, x.nord, x.sbux))
```

@fig-PortMatrixRandomPortfolios shows the means and volatilities,
as grey dots, of these 200 random portfolios of the three assets.
Notice that the grey dots outline a bullet-shaped area (Markowitz
bullet) that has an outer boundary that looks like the line characterizing
the two-asset portfolio frontier. If we let the number of random portfolios
get very large then the grey dots will fill in a solid area. At the
tip of the Markowitz bullet is the minimum variance portfolio. 


Recall the definition of a mean-variance efficient portfolio given
in Chapter @sec-Introduction-to-Portfolio-Theory. For a fixed
volatility $\sigma_{p,0}$ the efficient portfolio is the one with
the highest expected return. From @fig-PortMatrixRandomPortfolios,
we can see that efficient portfolios will lie on the outer boundary
(above the minimum variance portfolio at the tip of the Markowitz
bullet) of the set of feasible portfolios.

```{r}
#| label: fig-PortMatrixRandomPortfolios
#| echo: false
#| fig-cap: Risk-return characteristics of 200 random portfolios of three assets.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sd.vec, mu.vec,  ylim=c(-0.03, 0.09), xlim=c(0, 0.4), ylab=expression(mu[p]),      
     xlab=expression(sigma[p]), type="n", cex.lab=1.5)      
for (i in 1:length(x.msft)) {   
   z.vec = c(x.msft[i], x.nord[i], x.sbux[i])   
   mu.p = crossprod(z.vec,mu.vec)   
   sig.p = sqrt(t(z.vec)%*%sigma.mat%*%z.vec)   
   points(sig.p, mu.p, pch=16, col="grey", cex=1.5) 
} 
points(sd.vec, mu.vec, pch=16, col="blue", cex=2.5, cex.lab=1.75) 
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val) 
```

\hfill{}$\blacksquare$
:::

### Large portfolios and diversification
\index{Large portfolios}
One of the key benefits of forming large portfolios is risk reduction
due to *portfolio diversification*\index{Portfolio diversification}. Generally, a diversified portfolio is one in which wealth is spread across many assets.

To illustrate the impact of diversification on the risk of large portfolios,
consider an equally weighted portfolio of $N$ risky assets where $N$
is a large number (e.g., $N>100)$. Here, $x_{i}=1/N$ and the $N\times1$
portfolio weight vector is $\mathbf{x}=(1/N,1/N,\ldots,1/N)^{\prime}=(1/N)(1,1,\ldots,1)^{\prime}=(1/N)\mathbf{1}$
where $\mathbf{1}$ is an $N\times1$ vector of ones. Let the $N\times1$
return vector $\mathbf{R}$ be described by the GWN model so that $\mathbf{R}\sim N(\boldsymbol{\mu},\boldsymbol{\Sigma})$.
The portfolio return is $R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}=(1/N)\mathbf{1}^{\prime}\mathbf{R}=(1/N)\sum_{i=1}^{N}R_{i}.$
The variance of this portfolio is: 
$$
\mathrm{var}(R_{p,x})=\mathrm{var}(\mathbf{x}^{\prime}\mathbf{R})=\mathrm{var}\left(\frac{1}{N}\mathbf{1}^{\prime}\mathbf{R}\right)=\frac{1}{N^{2}}\mathbf{1}^{\prime}\boldsymbol{\Sigma}\mathbf{1}.
$$
Now, 
\begin{align*}
\mathbf{1}^{\prime}\boldsymbol{\Sigma}\mathbf{1}&=(\begin{array}{cccc}
1 & 1 & \ldots & 1\end{array})^{\prime}\left(\begin{array}{cccc}
\sigma_{1}^{2} & \sigma_{12} & \ldots & \sigma_{1N}\\
\sigma_{12} & \sigma_{2}^{2} & \ldots & \sigma_{2N}\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{1N} & \sigma_{2N} & \ldots & \sigma_{N}^{2}
\end{array}\right)\left(\begin{array}{c}
1\\
1\\
\vdots\\
1
\end{array}\right)\\
&=\text{ sum of all elements in } \boldsymbol{\Sigma}.
\end{align*}
As a result, we can write: 
$$
\mathbf{1}^{\prime}\boldsymbol{\Sigma}\mathbf{1}=\sum_{i=1}^{N}\sigma_{i}^{2}+\sum_{j\neq i}\sigma_{ij},
$$
which is the sum of the $N$ diagonal variance terms plus the $N(N-1)$
off-diagonal covariance terms. Define the average variance as the
average of the diagonal terms of $\boldsymbol{\Sigma}$:
$$
\overline{\mathrm{var}}=\frac{1}{N}\sum_{i=1}^{N}\sigma_{i}^{2},
$$
and the average covariance as the average of the off-diagonal terms
of $\boldsymbol{\Sigma}$:
$$
\overline{\mathrm{cov}}=\frac{1}{N(N-1)}\sum_{j\neq i}\sigma_{ij}.
$$
Then $\mathbf{1}^{\prime}\boldsymbol{\Sigma}\mathbf{1}=N\overline{\mathrm{var}}+N(N-1)\overline{\mathrm{cov}}$
and $\mathrm{var}(R_{p,x})$ reduces to: 
$$
\mathrm{var}(R_{p,x})  = \frac{1}{N^{2}}\mathbf{1}^{\prime}{\Sigma}\mathbf{1}=\frac{1}{N}\overline{\mathrm{var}}+\frac{N(N-1)}{N^{2}}\overline{\mathrm{cov}}  =  \frac{1}{N}\overline{\mathrm{var}}+\left(1-\frac{1}{N}\right)\overline{\mathrm{cov}}.
$${#eq-portMatrixPortVariance} 
If $N$ is reasonably large (e.g. $N>100$) then 
$$
\frac{1}{N}\overline{\mathrm{var}}\approx0\,\mathrm{and}\,\frac{1}{N}\overline{\mathrm{cov}}\approx0,
$$
so that
$$
\mathrm{var}(R_{p,x})\approx\overline{\mathrm{cov}}.
$$
Hence, in a large diversified portfolio the portfolio variance is
approximately equal to the average of the pairwise covariances. So
what matters for portfolio risk is not the individual asset variances
but rather the average of the asset covariances. In particular, portfolios
with highly positively correlated returns will have higher portfolio
variance than portfolios with less correlated returns. 

@eq-portMatrixPortVariance also shows that portfolio
variance should be a decreasing function of the number of assets in
the portfolio, and that this function should level off at $\overline{\mathrm{cov}}$
for large $N$. 


#### Large portfolio with perfectly positively correlated assets

In chapter @sec-Introduction-to-Portfolio-Theory I showed that the volatility of a long-only 
portfolio of two risky assets with perfectly positively correlated returns ($\rho_{12}=1$) is equal to a share
weighted average of the asset volatilities: $\sigma_{p}=x_{1}\sigma_{1}+x_{2}\sigma_{2}.$
Here, we show using matrix algebra that this result generalizes to the 
case of $N$ assets. Let $\mathbf{R}$ be an $N\times1$ vector of
asset returns and assume that all returns are perfectly positively
correlated ($\rho_{ij}=1$ for all $i,j$). The $N\times N$ return correlation matrix is then:
$$
\mathbf{C}=\left(\begin{array}{cccc}
1 & 1 & \ldots & 1\\
1 & 1 & \ldots & 1\\
\vdots & \vdots & \ddots & \vdots\\
1 & 1 & \ldots & 1
\end{array}\right)=\left(\begin{array}{c}
1\\
1\\
\vdots\\
1
\end{array}\right)\left(\begin{array}{cccc}
1 & 1 & \ldots & 1\end{array}\right)=\mathbf{1}\mathbf{1}^{\prime},
$$
and the $N\times N$ return covariance matrix is:
$$
{\Sigma}=\mathbf{D}\mathbf{C}\mathbf{D=D\mathbf{1}\mathbf{1}^{\prime}D},
$$
where $\mathbf{D}=\mathrm{diag}(\sigma_{1},\sigma_{2},\ldots,\sigma_{N})$.
Let $\mathbf{x}=(x_{1},x_{2},\ldots,x_{N})^{\prime}$ be a portfolio
weight vector and let $R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}$ denote
the portfolio return. Then:
$$
\sigma_{p}^{2}=\mathrm{var}(\mathbf{x}^{\prime}\mathbf{R})=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}=\mathbf{x}^{\prime}\mathbf{D}\mathbf{1}\mathbf{1}^{\prime}\mathbf{Dx}=\left(\mathbf{x}^{\prime}\mathbf{D}\mathbf{1}\right)^{2}.
$$
Now,
\begin{eqnarray*}
\mathbf{x}^{\prime}\mathbf{D}\mathbf{1} & = & \left(\begin{array}{cccc}
x_{1} & x_{2} & \ldots & x_{N}\end{array}\right)\left(\begin{array}{cccc}
\sigma_{1} & 0 & \cdots & 0\\
0 & \sigma_{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & \sigma_{N}
\end{array}\right)\left(\begin{array}{c}
1\\
1\\
\vdots\\
1
\end{array}\right)\\
& = &\left(\begin{array}{cccc}
x_{1}\sigma_{1} & 0 & \cdots & 0\\
0 & x_{2}\sigma_{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & x_{N}\sigma_{N}
\end{array}\right)\left(\begin{array}{c}
1\\
1\\
\vdots\\
1
\end{array}\right)\\
 & = & \sum_{i=1}^{N}x_{i}\sigma_{i}.
\end{eqnarray*}
Therefore, 
$$
\sigma_{p}^{2}=\left(\mathbf{x}^{\prime}\mathbf{D}\mathbf{1}\right)^{2}=\left(\sum_{i=1}^{N}x_{i}\sigma_{i}\right)^{2}
$$
and so: 
$$
\sigma_{p}=\sum_{i=1}^{N}x_{i}\sigma_{i}.
$$
When all assets have perfectly positively correlated returns, portfolio
volatility is a share weighted average of individual asset volatilities.
You can think of this as the risk of a (long-only) portfolio when
there are no diversification benefits. When assets are not all perfectly
positively correlated, there are some diversification benefits and
so $\sigma_{p}$ will be less than $\sum_{i=1}^{N}x_{i}\sigma_{i}.$


## Determining the Global Minimum Variance Portfolio Using Matrix Algebra {#sec-Determining-the-Global-Minimum-Variance-Portfolio-Using-Matrix-Algebra}
\index{Global minimum variance portfolio}
The global minimum variance portfolio $\mathbf{m}=(m_{1},\ldots,m_{N})^{\prime}$
for the $N$ asset case solves the constrained minimization problem:

$$
\min_{\mathbf{m}}~\sigma_{p,m}^{2}=\mathbf{m^{\prime}{\Sigma} m}, ~ s.t. ~ \mathbf{m^{\prime}1}=1.
$$ {#eq-PortMatrixGlobalMinObjective}
The Lagrangian for this problem is: 
\begin{align*}
L(\mathbf{m},\lambda) & =\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}+\lambda(\mathbf{m}^{\prime}\mathbf{1}-1),
\end{align*}
and the first order conditions (FOCs) for a minimum are: 
$$
\mathbf{0} =\frac{\partial L}{\partial\mathbf{m}}=\frac{\partial\mathbf{m}^{\prime}\boldsymbol{\Sigma}\mathbf{ m}}{\partial\mathbf{m}}+\frac{\partial\lambda(\mathbf{m}^{\prime}\mathbf{1}-1)}{\partial\mathbf{m}}=2\boldsymbol{\Sigma} \mathbf{m}+\lambda\mathbf{1},
$$ {#eq-PortMatrixGlobalMinFOC1}
$$
0 =\frac{\partial L}{\partial\lambda}=\frac{\partial\mathbf{m^{\prime}\boldsymbol{\Sigma} m}}{\partial\lambda}+\frac{\partial\lambda(\mathbf{m^{\prime}1}-1)}{\partial\lambda}=\mathbf{m^{\prime}1}-1.
$$ {#eq-PortMatrixGlobalMinFOC2}
The FOCs @eq-PortMatrixGlobalMinFOC1 - @eq-PortMatrixGlobalMinFOC2
give $N+1$ linear equations in $N+1$ unknowns which can be solved
to find the global minimum variance portfolio weight vector $\mathbf{m}$
as follows. The $N+1$ linear equations describing the first order
conditions have the matrix representation: 
$$
\left(\begin{array}{cc}
2\boldsymbol{\Sigma} & \mathbf{1}\\
\mathbf{1}^{\prime} & 0
\end{array}\right)\left(\begin{array}{c}
\mathbf{m}\\
\lambda
\end{array}\right)=\left(\begin{array}{c}
\mathbf{0}\\
1
\end{array}\right).
$$ {#eq-PortMatrixGlobalMinFOCmatrix}

The system @eq-PortMatrixGlobalMinFOCmatrix is of the form
$\mathbf{A}_{m}\mathbf{z}_{m}=\mathbf{b}$, where: 
$$
\mathbf{A}_{m}=\left(\begin{array}{cc}
2\boldsymbol{\Sigma} & \mathbf{1}\\
\mathbf{1}^{\prime} & 0
\end{array}\right),~\mathbf{z}_{m}=\left(\begin{array}{c}
\mathbf{m}\\
\lambda
\end{array}\right)\textrm{ and }\mathbf{b}=\left(\begin{array}{c}
\mathbf{0}\\
1
\end{array}\right).
$$
Provided $\mathbf{A}_{m}$ is invertible, the solution for $\mathbf{z}_{m}$
is:^[For $\mathbf{A}_{m}$ to be invertible we need $\boldsymbol{\Sigma}$
to be invertible, which requires $|\rho_{ij}|\neq1$ for $i\neq j$
and $\sigma_{i}^{2}>0$ for all $i$.] 
$$
\mathbf{z}_{m}=\mathbf{A}_{m}^{-1}\mathbf{b}.
$$ {#eq-PortMatrixGlobalMinSolution}
The first $N$ elements of $\mathbf{z}_{m}$ is the portfolio weight
vector $\mathbf{m}$ for the global minimum variance portfolio with
expected return $\mu_{p,m}=\mathbf{m}^{\prime}\boldsymbol{\mu}$ and variance
$\sigma_{p,m}^{2}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}$.

::: {#exm-global-minimum-variance-portfolio-for-example-data .example name="Global minimum variance portfolio for example data"}

Using the data in Table @tbl-Table-ThreeAssetExample, you can
use R to compute the global minimum variance portfolio weights from
@eq-PortMatrixGlobalMinSolution as follows:

```{r}
#| echo: true

top.mat = cbind(2*sigma.mat, rep(1, 3))
bot.vec = c(rep(1, 3), 0)
Am.mat = rbind(top.mat, bot.vec)
b.vec = c(rep(0, 3), 1)
z.m.mat = solve(Am.mat)%*%b.vec
m.vec = z.m.mat[1:3,1]
m.vec
```

Hence, the global minimum variance portfolio has portfolio weights
$m_{\textrm{msft}}=0.441$, $m_{\textrm{nord}}=0.366$ and $m_{\textrm{sbux}}=0.193$,
and is given by the vector $\mathbf{m}=(0.441,0.366,0.193)^{\prime}$.
The expected return on this portfolio, $\mu_{p,m}=\mathbf{m}^{\prime}\boldsymbol{\mu}$,
is:

```{r}
#| echo: true

mu.gmin = as.numeric(crossprod(m.vec, mu.vec))
mu.gmin
```

The portfolio variance, $\sigma_{p,m}^{2}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}$,
and standard deviation, $\sigma_{p,m}$, are:

```{r}
#| echo: true

sig2.gmin = as.numeric(t(m.vec)%*%sigma.mat%*%m.vec)
sig.gmin = sqrt(sig2.gmin)
c(sig2.gmin, sig.gmin)
```

In @fig-PortMatrixGlobalMin, this portfolio is labeled "GMIN" .

```{r}
#| label: fig-PortMatrixGlobalMin
#| echo: false
#| fig-cap: Global minimum variance portfolio from example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sd.vec, mu.vec,  ylim=c(-0.05, 0.09), xlim=c(0, 0.4), ylab=expression(mu[p]),
     xlab=expression(sigma[p]), type="n", cex.lab=1.5)   
text(0, r.f, labels=expression(r[f]), pos=2, cex = cex.val) 
for (i in 1:length(x.msft)) {   
   z.vec = c(x.msft[i], x.nord[i], x.sbux[i])   
   mu.p = crossprod(z.vec,mu.vec)   
   sig.p = sqrt(t(z.vec)%*%sigma.mat%*%z.vec)   
   points(sig.p, mu.p, pch=16, col="grey", cex=1.5) 
} 
points(sd.vec, mu.vec, pch=16, col="blue", cex=2.5, cex.lab=1.75) 
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val) 
points(sig.gmin, mu.gmin, pch=16, cex=2.5, col="green") 
text(sig.gmin, mu.gmin, labels="GMIN", pos=2.5, cex = cex.val)
```

\hfill{}$\blacksquare$
:::

### Alternative derivation of global minimum variance portfolio

We can use the first order conditions @eq-PortMatrixGlobalMinFOC1
- @eq-PortMatrixGlobalMinFOC2 to give an explicit solution
for the global minimum variance portfolio $\mathbf{m}$ as follows.
First, use @eq-PortMatrixGlobalMinFOC1 to solve for $\mathbf{m}$
as a function of $\lambda$: 
$$
\mathbf{m}=-\frac{1}{2}\cdot\lambda\boldsymbol{\Sigma}^{-1}\mathbf{1}.
$$ {#eq-PortMatrixM}
Next, multiply both sides of @eq-PortMatrixM by $\mathbf{1}^{\prime}$
and use @eq-PortMatrixGlobalMinFOC2 to solve for $\lambda$:
\begin{align*}
1 & =\mathbf{1}^{\prime}\mathbf{m}=-\frac{1}{2}\cdot\lambda\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}\\
 & \Rightarrow\lambda=-2\cdot\frac{1}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}}.
\end{align*}
Finally, substitute the value for $\lambda$ back into @eq-PortMatrixM
to solve for $\mathbf{m}$: 
$$
\mathbf{m}=-\frac{1}{2}\left(-2\right)\frac{1}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}}\boldsymbol{\Sigma}^{-1}\mathbf{1}=\frac{\boldsymbol{\Sigma}^{-1}\mathbf{1}}{\mathbf{1}^{\prime}{\boldsymbol{\Sigma}}^{-1}\mathbf{1}}.
$$ {#eq-PortMatrixGlobalMinSolutionM}
Notice that @eq-PortMatrixGlobalMinSolutionM shows that a solution for $\mathbf{m}$ exists as long as $\boldsymbol{\Sigma}$ is invertible.

::: {#exm-finding-global-minimum-variance-portfolio-for-example-data .example name="Finding global minimum variance portfolio for example data"}

Using the data in Table @tbl-Table-ThreeAssetExample, we can
use R to compute the global minimum variance portfolio weights from
@eq-PortMatrixGlobalMinSolutionM as follows:

```{r}
#| echo: true

one.vec = rep(1, 3)
sigma.inv.mat = solve(sigma.mat)
top.mat = sigma.inv.mat%*%one.vec
bot.val = as.numeric((t(one.vec)%*%sigma.inv.mat%*%one.vec))
m.mat = top.mat/bot.val
m.mat[,1]
```

\hfill{}$\blacksquare$
:::

## Determining Mean-Variance Efficient Portfolios Using Matrix Algebra {#sec-Determining-Mean-Variance-Effici}
\index{Mean-Variance efficient portfolios}
The investment opportunity set is the set of portfolio expected return,
$\mu_{p}$, and portfolio standard deviation, $\sigma_{p}$, values
for all possible portfolios whose weights sum to one. As in the two
risky asset case, this set can be described in a graph with $\mu_{p}$
on the vertical axis and $\sigma_{p}$ on the horizontal axis. With
two assets, the investment opportunity set in ($\mu_{p},\sigma_{p}$)-space
lies on a curve (one side of a hyperbola). With three or more assets,
the investment opportunity set in ($\mu_{p},\sigma_{p}$)-space is
described by set of values whose general shape is complicated and
depends crucially on the covariance terms $\sigma_{ij}$.^[This set is partially visualized in  @fig-PortMatrixRandomPortfolios using random portfolios.] However, we do not have to fully characterize the entire investment opportunity set. If we assume that investors choose portfolios to
maximize expected return subject to a target level of risk, or, equivalently,
to minimize risk subject to a target expected return, then we can
simplify the asset allocation problem by only concentrating on the
set of efficient portfolios. These portfolios lie on the boundary
of the investment opportunity set above the global minimum variance
portfolio. This is the framework originally developed by Harry Markowitz,
the father of portfolio theory and winner of the Nobel Prize in economics.


Following Markowitz, we assume that investors wish to find portfolios
that have the best expected return-risk trade-off. We can characterize
these efficient portfolios in two equivalent ways. In the first way,
investors seek to find portfolios that maximize portfolio expected
return for a given level of risk as measured by portfolio variance.
Let $\sigma_{p,0}^{2}$ denote a target level of risk. Then the constrained
maximization problem to find an efficient portfolio is: 
$$
\begin{aligned}
\max_{\mathbf{x}}\mu_{p} & =\mathbf{x}^{\prime}\boldsymbol{\mu} \\
\textrm{ s.t. } \sigma_{p}^{2} & =\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}=\sigma_{p,0}^{2}\textrm{ and }\mathbf{x}^{\prime}\mathbf{1}=1. 
\end{aligned}
$$ {#eq-PortMatrixEfficientPortfolio1}
The investor's problem of maximizing portfolio expected return subject
to a target level of risk has an equivalent dual representation in
which the investor minimizes the risk of the portfolio (as measured
by portfolio variance) subject to a target expected return level.
Let $\mu_{p,0}$ denote a target expected return level. Then the dual
problem is the constrained minimization problem:^[This duality is similar to the duality in economics of profit maximization and cost minimization.]
$$
\begin{aligned}
\min_{\mathbf{x}}~\sigma_{p,x}^{2} & =\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x} \\
\textrm{ s.t.  } \mu_{p} & =\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p,0}\mathbf{,}\textrm{ and }\mathbf{x}^{\prime}\mathbf{1}=1. 
\end{aligned}
$$ {#eq-PortMatrixEfficientPortfolio2}
To find efficient portfolios of risky assets in practice, the dual
problem @eq-PortMatrixEfficientPortfolio2 is most often solved.
This is partially due to computational conveniences and partly due
to investors being more willing to specify target expected returns
rather than target risk levels. The efficient portfolio frontier is
a graph of $\mu_{p}$ versus $\sigma_{p}$ values for the set of efficient
portfolios generated by solving @eq-PortMatrixEfficientPortfolio2
for all possible target expected return levels $\mu_{p,0}$ above
the expected return on the global minimum variance portfolio. Just
as in the two asset case, the resulting efficient frontier will resemble
one side of an hyperbola and is often called the "Markowitz bullet". This frontier 
is illustrated in  @fig-PortMatrixRandomPortfolios
as the boundary of the set generated by random portfolios above the
global minimum variance portfolio. 

To solve the constrained minimization problem @eq-PortMatrixEfficientPortfolio2,
first form the Lagrangian function: 
$$
L(x,\lambda_{1},\lambda_{2})=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}+\lambda_{1}(\mathbf{x}^{\prime}\boldsymbol{\mu}-\mu_{p,0})+\lambda_{2}(\mathbf{x}^{\prime}\mathbf{1}-1).
$$
Because there are two constraints ($\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p,0}$
and $\mathbf{x}^{\prime}\mathbf{1}=1)$ there are two Lagrange multipliers
$\lambda_{1}$ and $\lambda_{2}$. The FOCs for a minimum are the
linear equations: 
$$
\frac{\partial L(\mathbf{x},\lambda_{1},\lambda_{2})}{\partial\mathbf{x}} =2\boldsymbol{\Sigma} \mathbf{x}+\lambda_{1}\boldsymbol{\mu}+\lambda_{2}\mathbf{1}=\mathbf{0,}
$$ {#eq-PortMatrixEfficientFOC1}
$$
\frac{\partial L(\mathbf{x},\lambda_{1},\lambda_{2})}{\partial\lambda_{1}}  =\mathbf{x}^{\prime}\boldsymbol{\mu}-\mu_{p,0}=0,
$$ {#eq-PortMatrixEfficientFOC2}
$$
\frac{\partial L(\mathbf{x},\lambda_{1},\lambda_{2})}{\partial\lambda_{2}} =\mathbf{x}^{\prime}\mathbf{1}-1=0.
$$ {#eq-PortMatrixEfficientFOC3}

These FOCs consist of $N+2$ linear equations in $N+2$ unknowns ($\mathbf{x},\,\lambda_{1},\,\lambda_{2})$.
We can represent the system of linear equations using matrix algebra
as: 
$$
\left(\begin{array}{ccc}
2\boldsymbol{\Sigma} & \boldsymbol{\mu} & \mathbf{1}\\
\boldsymbol{\mu}^{\prime} & 0 & 0\\
\mathbf{1}^{\prime} & 0 & 0
\end{array}\right)\left(\begin{array}{c}
\mathbf{x}\\
\lambda_{1}\\
\lambda_{2}
\end{array}\right)=\left(\begin{array}{c}
\mathbf{0}\\
\mu_{p,0}\\
1
\end{array}\right),
$$
which is of the form $\mathbf{Az}_{x}\mathbf{=b}_{0}$, where 
$$
\mathbf{A}=\left(\begin{array}{ccc}
2\boldsymbol{\Sigma} & \boldsymbol{\mu} & \mathbf{1}\\
\boldsymbol{\mu}^{\prime} & 0 & 0\\
\mathbf{1}^{\prime} & 0 & 0
\end{array}\right),~\mathbf{z}_{x}=\left(\begin{array}{c}
\mathbf{x}\\
\lambda_{1}\\
\lambda_{2}
\end{array}\right)\textrm{ and }\mathbf{b}_{0}=\left(\begin{array}{c}
\mathbf{0}\\
\mu_{p,0}\\
1
\end{array}\right).
$$
The solution for $\mathbf{z}_{x}$ is then: 
$$
\mathbf{z}_{x}=\mathbf{A}^{-1}\mathbf{b}_{0}.
$$ {#eq-PortMatrixEfficientPortfolioSolution}
The first $N$ elements of $\mathbf{z}_{x}$ are the portfolio weights
$\mathbf{x}$ for the minimum variance portfolio with expected return
$\mu_{p,x}=\mu_{p,0}$. If $\mu_{p,0}$ is greater than or equal to
the expected return on the global minimum variance portfolio, then
$\mathbf{x}$ is an efficient (frontier) portfolio. Otherwise, it
is an inefficient (frontier) portfolio. 

::: {#exm-efficient-portfolio-with-the-same-expected-return-as-microsoft .example name="Efficient portfolio with the same expected return as Microsoft"}

Using the data in Table @tbl-Table-ThreeAssetExample, consider
finding a minimum variance portfolio with the same expected return
as Microsoft. This will be an efficient portfolio because $\mu_{\textrm{msft}}=0.043>\mu_{p,m}=0.025$.
Call this portfolio $\mathbf{x}=(x_{\textrm{msft}},x_{\textrm{nord}},x_{\textrm{sbux}})^{\prime}$.
That is, consider solving @eq-PortMatrixEfficientPortfolio2
with target expected return $\mu_{p,0}=\mu_{\textrm{msft}}=0.043$
using @eq-PortMatrixEfficientPortfolioSolution. The R calculations
to create the matrix $\mathbf{A}_{x}$ and the vectors $\mathbf{z}_{x}$
and $\mathbf{b}_{\textrm{msft}}$ are:

```{r}
#| echo: true

top.mat = cbind(2*sigma.mat, mu.vec, rep(1, 3))
mid.vec = c(mu.vec, 0, 0)
bot.vec = c(rep(1, 3), 0, 0)
Ax.mat = rbind(top.mat, mid.vec, bot.vec)
bmsft.vec = c(rep(0, 3), mu.vec["MSFT"], 1)
```

and the R code to solve for $\mathbf{x}$ using @eq-PortMatrixEfficientPortfolioSolution
is:

```{r}
#| echo: true

z.mat = solve(Ax.mat)%*%bmsft.vec
x.vec = z.mat[1:3,]
x.vec
```

The efficient portfolio with the same expected return as Microsoft
has portfolio weights $x_{\textrm{msft}}=0.8275$, $x_{\textrm{nord}}=-0.0907$
and $x_{\textrm{sbux}}=0.2633$, and is given by the vector $\mathbf{x}=(0.8275,-0.0907,0.2633)^{\prime}.$
The expected return on this portfolio, $\mu_{p,x}=\mathbf{x}^{\prime}\boldsymbol{\mu}$,
is equal to the target return $\mu_{\textrm{msft}}$:

```{r}
#| echo: true

mu.px = as.numeric(crossprod(x.vec, mu.vec))
mu.px
```

The portfolio variance, $\sigma_{p,x}^{2}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}$,
and standard deviation, $\sigma_{p,x}$, are:

```{r}
#| echo: true

sig2.px = as.numeric(t(x.vec)%*%sigma.mat%*%x.vec)
sig.px = sqrt(sig2.px)
c(sig2.px, sig.px)
```

and are smaller than the corresponding values for Microsoft (see Table
@tbl-Table-ThreeAssetExample). This efficient portfolio is labeled
"E1" in @fig-PortMatrixEfficientPortfolios.
\hfill{}$\blacksquare$
:::

::: {#exm-efficient-portfolio-with-the-same-expected-return-as-starbucks .example name="Efficient portfolio with the same expected return as Starbucks"}

To find a minimum variance portfolio $\mathbf{y}=(y_{\textrm{msft}},y_{\textrm{nord}},y_{\textrm{sbux}})^{\prime}$
with the same expected return as Starbucks we use @eq-PortMatrixEfficientPortfolioSolution
with $\mathbf{b}_{\textrm{sbux}}=(\mathbf{0}',\mu_{\textrm{sbux}},1)^{\prime}$:

```{r}
#| echo: true

bsbux.vec = c(rep(0, 3), mu.vec["SBUX"], 1)
z.mat = solve(Ax.mat)%*%bsbux.vec
y.vec = z.mat[1:3,]
y.vec
```

The portfolio $\mathbf{y}=(0.519,0.273,0.207)^{\prime}$ is an efficient
portfolio on the outer boundary because $\mu_{\textrm{sbux}}=0.0285>\mu_{p,m}=0.0249$.
The portfolio expected return and standard deviation are:

```{r}
#| echo: true

mu.py = as.numeric(crossprod(y.vec, mu.vec))
sig2.py = as.numeric(t(y.vec)%*%sigma.mat%*%y.vec)
sig.py = sqrt(sig2.py)
c(mu.py,sig.py)
```

This efficient portfolio is labeled "E2" in @fig-PortMatrixEfficientPortfolios. 
It has the same expected return as SBUX but a smaller standard deviation. 
\hfill{}$\blacksquare$
:::

The covariance and correlation values between the portfolio returns
$R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}$ and $R_{p,y}=\mathbf{y}^{\prime}\mathbf{R}$
are given by:

```{r}
#| echo: true

sigma.xy = as.numeric(t(x.vec)%*%sigma.mat%*%y.vec)
rho.xy = sigma.xy/(sig.px*sig.py)
c(sigma.xy, rho.xy)
```

This covariance will be used later on when constructing the entire
frontier of efficient portfolios.

```{r}
#| label: fig-PortMatrixEfficientPortfolios
#| echo: false
#| fig-cap: Minimum variance efficient portfolios from example data. Portfolio E1 has
#|   the same expected return as Microsoft, and portfolio E2 has the same expected returns
#|   as Starbucks.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

sd.vec = sqrt(diag(sigma.mat)) 
plot(sd.vec, mu.vec,  ylim=c(0, 0.06), xlim=c(0, 0.20), ylab=expression(mu[p]),      
     xlab=expression(sigma[p]), pch=16, col="blue", cex=2, cex.lab=1.5) 
points(sig.gmin, mu.gmin, pch=16, cex=2, col="green") 
points(sig.px, mu.px, pch=16, cex=2, col="green") 
points(sig.py, mu.py, pch=16, cex=2, col="green")       
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val) 
text(sig.gmin, mu.gmin, labels="GMIN", pos=2, cex = cex.val)         
text(sig.px, mu.px, labels="E1", pos=2, cex = cex.val)  
text(sig.py, mu.py, labels="E2", pos=2, cex = cex.val)  
text(0, r.f, labels=expression(r[f]), pos=2, cex = cex.val)
```


### Alternative derivation of an efficient portfolio

@eq-PortMatrixEfficientPortfolioSolution does
not give an explicit solution for the minimum variance portfolio $\mathbf{x}$.
As with the global minimum variance portfolio @eq-PortMatrixGlobalMinSolutionM,
an explicit solution for $\mathbf{x}$ can also be found. Consider
the first order conditions @eq-PortMatrixEfficientFOC1 - @eq-PortMatrixEfficientFOC3
from the optimization problem @eq-PortMatrixEfficientPortfolio2.
First, use @eq-PortMatrixEfficientFOC1 to solve for the $N\times1$
vector $\mathbf{x}$: 
$$
\mathbf{x}=-\frac{1}{2}\lambda_{1}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu}-\frac{1}{2}\lambda_{2}\boldsymbol{\Sigma}^{-1}\mathbf{1}.
$$ {#eq-PortMatrixEfficientX}
Define the $N\times2$ matrix $\mathbf{M}=[\boldsymbol{\mu}$ $\vdots$
$\mathbf{1}]$ and the $2\times1$ vector $\boldsymbol{\lambda}=(\lambda_{1},\lambda_{2})^{\prime}$.
Then we can rewrite @eq-PortMatrixEfficientX in matrix form
as: 
$$
\mathbf{x}=-\frac{1}{2}\boldsymbol{\Sigma}^{-1}\mathbf{M}\boldsymbol{\lambda}.
$$ {#eq-PortMatrixEfficientX2}
Next, to find the values for $\lambda_{1}$ and $\lambda_{2}$, pre-multiply
@eq-PortMatrixEfficientX by $\boldsymbol{\mu}^{\prime}$ and use @eq-PortMatrixEfficientFOC2 to give: 
$$ 
\mu_{0} = \boldsymbol{\mu}^{\prime}\mathbf{x}
 = \frac{1}{2}\lambda_{1}\boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu} - \frac{1}{2}\lambda_{2}
 \boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}. 
$$ {#eq-PortMatrixEfficientLambda1}
Similarly, pre-multiply @eq-PortMatrixEfficientX by $\mathbf{1}^{\prime}$
and use @eq-PortMatrixEfficientFOC3 to give: 
$$
1= \mathbf{1}^{\prime}\mathbf{x}= -\frac{1}{2}\lambda_{1}\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu} 
-\frac{1}{2}\lambda_{2}\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}.
$$ {#eq-PortMatrixEfficientLambda2}
Now, we have two linear equations @eq-PortMatrixEfficientLambda1
and @eq-PortMatrixEfficientLambda2 involving $\lambda_{1}$
and $\lambda_{2}$ which we can write in matrix notation as: 
$$
-\frac{1}{2}\left(\begin{array}{cc}
\boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu} & \boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}\\
\boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1} & \mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}
\end{array}\right)\left(\begin{array}{c}
\lambda_{1}\\
\lambda_{2}
\end{array}\right)=\left(\begin{array}{c}
\mu_{0}\\
1
\end{array}\right).
$$ {#eq-PortMatrixEfficientLambdaEqn}
Define, 
\begin{align*}
\left(\begin{array}{cc}
\boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu} & \boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}\\
\boldsymbol{\mu}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1} & \mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{1}
\end{array}\right) & =\mathbf{M}^{\prime}\boldsymbol{\Sigma}^{-1}\mathbf{M=B},\\
\tilde{\boldsymbol{\mu}}_{0} & =\left(\begin{array}{c}
\mu_{0}\\
1
\end{array}\right),
\end{align*}
so that we can rewrite @eq-PortMatrixEfficientLambdaEqn as,
$$
-\frac{1}{2}\mathbf{B}\boldsymbol{\lambda}=\tilde{\boldsymbol{\mu}}_{0}.
$$
Provided $\mathbf{B}$ is invertible, the solution for $\boldsymbol{\lambda}=(\lambda_{1},\lambda_{2})^{\prime}$
is 
$$
\boldsymbol{\lambda}=-2\mathbf{B}^{-1}\tilde{\boldsymbol{\mu}}_{0}.
$$ {#eq-PortMatrixEfficientLambdaSolution}
Substituting @eq-PortMatrixEfficientLambdaSolution back into
@eq-PortMatrixEfficientX2 gives an explicit expression for
the efficient portfolio weight vector $\mathbf{x}:$ 
$$
\mathbf{x}=-\frac{1}{2}\boldsymbol{\Sigma}^{-1}\mathbf{M}\boldsymbol{\lambda}=-\frac{1}{2}\boldsymbol{\Sigma}^{-1}\mathbf{M}\left(-2\mathbf{B}^{-1}\tilde{\mu}_{0}\right)=\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}\tilde{\boldsymbol{\mu}}_{0}.
$$ {#eq-PortMatrixEfficientXsolution}

::: {#exm-alternative-solution-for-efficient-portfolio-with-the-same-expected-return-as-microsoft .example name="Alternative solution for efficient portfolio with the same expected return as Microsoft"}

The R code to compute the efficient portfolio with the same expected
return as Microsoft using @eq-PortMatrixEfficientXsolution
is:

```{r}
#| echo: true

M.mat = cbind(mu.vec, one.vec)
B.mat = t(M.mat)%*%solve(sigma.mat)%*%M.mat
mu.tilde.msft = c(mu.vec["MSFT"], 1)
x.vec.2 = solve(sigma.mat)%*%M.mat%*%solve(B.mat)%*%mu.tilde.msft
t(x.vec.2)
```

\hfill{}$\blacksquare$
:::

## Computing the Mean-Variance Efficient Frontier {#sec-Computing-the-Mean-Variance}
\index{Mean-Variance efficient frontier}
The analytic expression for a minimum variance portfolio @eq-PortMatrixEfficientXsolution
can be used to show that any minimum variance portfolio can be created
as a linear combination of any two minimum variance portfolios with
different target expected returns. If the expected return on the resulting
portfolio is greater than the expected return on the global minimum variance
portfolio, then the portfolio is an efficient frontier portfolio.
Otherwise, the portfolio is an inefficient frontier portfolio. As
a result, to compute the portfolio frontier in $(\mu_{p},\sigma_{p})$
space (Markowitz bullet) we only need to find two efficient portfolios.
The remaining frontier portfolios can then be expressed as linear
combinations of these two portfolios. The following proposition describes the process for the three risky
asset case using matrix algebra.

::: {#prp-creating-a-frontier-portfolio-from-two-minimum-variance-portfolios .proposition name="Creating a frontier portfolio from two minimum variance portfolios"}

Let $\mathbf{x}$ and $\mathbf{y}$ be any two minimum variance portfolios
with different target expected returns $\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p,0}\neq\mathbf{y}^{\prime}\boldsymbol{\mu}=\mu_{p,1}$.
That is, portfolio $\mathbf{x}$ solves: 
\begin{align*}
\min_{\mathbf{x}}~\sigma_{p,x}^{2}&=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\\
\textrm{ s.t. }\mathbf{x}^{\prime}\boldsymbol{\mu}&=\mu_{p,0}\textrm{ and }\mathbf{x}^{\prime}\mathbf{1}=1,
\end{align*}
and portfolio $\mathbf{y}$ solves, 
\begin{align*}
\min_{\mathbf{y}}\sigma_{p,y}^{2}&=\mathbf{y}^{\prime}\boldsymbol{\Sigma} \mathbf{y}\\
\textrm{ s.t. }\mathbf{y}^{\prime}\boldsymbol{\mu}&=\mu_{p,1}\textrm{ and }\mathbf{y}^{\prime}\mathbf{1}=1.
\end{align*}
Let $\alpha$ be any constant and define the portfolio $\mathbf{z}$
as a linear combination of portfolios $\mathbf{x}$ and $\mathbf{y}$:
$$
\begin{aligned}
\mathbf{z} & =\alpha\cdot\mathbf{x}+(1-\alpha)\cdot\mathbf{y}\\
 & =\left(\begin{array}{c}
\alpha x_{1}+(1-\alpha)y_{1}\\
\vdots\\
\alpha x_{N}+(1-\alpha)y_{N}
\end{array}\right).
\end{aligned}
$$ {#eq-PortMatrixConvexCombination}
Then the following results hold:

1.  The portfolio $\mathbf{z}$ is a minimum variance portfolio with expected return and variance given by: 
$$
\mu_{p,z} =\mathbf{z}^{\prime}\boldsymbol{\mu}=\alpha\cdot\mu_{p,x}+(1-\alpha)\cdot\mu_{p,y},
$$ {#eq-PortMatrixMupz} 
$$
\sigma_{p,z}^{2} =\mathbf{z}^{\prime}\boldsymbol{\Sigma} \mathbf{z}=\alpha^{2}\sigma_{p,x}^{2}+(1-\alpha)^{2}\sigma_{p,y}^{2}+2\alpha(1-\alpha)\sigma_{xy},
$$ {#eq-PortMatrixSig2pz}
where $\sigma_{p,x}^{2}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x},\sigma_{p,y}^{2}=\mathbf{y}^{\prime}\boldsymbol{\Sigma} \mathbf{y},\sigma_{xy}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{y}.$ 
2.  If $\mu_{p,z}\geq\mu_{p,m}$, where $\mu_{p,m}$ is the expected return on the global minimum variance portfolio, then portfolio $\mathbf{z}$ is an efficient frontier portfolio. Otherwise, $\mathbf{z}$ is an inefficient frontier portfolio. 
:::

The proof of @creating-a-frontier-portfolio-from-two-minimum-variance-portfolios a follows directly from applying @eq-PortMatrixEfficientXsolution
to portfolios $\mathbf{x}$ and $\mathbf{y}$: 
\begin{align*}
\mathbf{x} & =\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}\tilde{\boldsymbol{\mu}}_{x},\\
\mathbf{y} & =\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}\tilde{\boldsymbol{\mu}}_{y},
\end{align*}
where $\tilde{\boldsymbol{\mu}}_{x}=(\mu_{p,x},1)^{\prime}$ and $\tilde{\boldsymbol{\mu}}_{y}=(\mu_{p,y},1)^{\prime}$.
Then, for portfolio $\mathbf{z}$: 
\begin{align*}
\mathbf{z} & =\alpha\cdot\mathbf{x}+(1-\alpha)\cdot\mathbf{y}\\
 & =\alpha\cdot\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}\tilde{\boldsymbol{\mu}}_{x}+(1-\alpha)\cdot\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}\tilde{\boldsymbol{\mu}}_{y}\\
 & =\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}(\alpha\cdot\tilde{\boldsymbol{\mu}}_{x}+(1-\alpha)\cdot\tilde{\boldsymbol{\mu}}_{y})\\
 & =\boldsymbol{\Sigma}^{-1}\mathbf{MB}^{-1}\tilde{\boldsymbol{\mu}}_{z},
\end{align*}
where $\tilde{\boldsymbol{\mu}}_{z}=\alpha\cdot\tilde{\boldsymbol{\mu}}_{x}+(1-\alpha)\cdot\tilde{\boldsymbol{\mu}}_{y}=(\mu_{p,z},1)^{\prime}$.
Result 2. follows from the definition of an efficient portfolio.

::: {#exm-create-an-arbitrary-efficient-frontier-portfolio-from-two-efficient-minimum-variance-portfolios .example name="Create an arbitrary efficient frontier portfolio from two efficient minimum variance portfolios"}

Consider the data in Table 1 and the previously computed minimum variance
portfolios that have the same expected return as Microsoft and Starbucks,
respectively, and let $\alpha=0.5$. From @eq-PortMatrixConvexCombination,
the frontier portfolio $\mathbf{z}$ is constructed using: 
\begin{align*}
\mathbf{z} & =\alpha\cdot\mathbf{x}+(1-\alpha)\cdot\mathbf{y}\\
 & =0.5\cdot\left(\begin{array}{c}
0.8275\\
-0.0907\\
0.2633
\end{array}\right)+0.5\cdot\left(\begin{array}{c}
0.519\\
0.273\\
0.207
\end{array}\right)\\
 & =\left(\begin{array}{c}
(0.5)(0.8275)\\
(0.5)(-0.0907)\\
(0.5)(0.2633)
\end{array}\right)+\left(\begin{array}{c}
(0.5)(0.519)\\
(0.5)(0.273)\\
(0.5)(0.207)
\end{array}\right)\\
 & =\left(\begin{array}{c}
0.6734\\
0.0912\\
0.2354
\end{array}\right)=\left(\begin{array}{c}
z_{A}\\
z_{B}\\
z_{C}
\end{array}\right).
\end{align*}
In R, the new frontier portfolio is computed using:

```{r}
#| echo: true

a = 0.5
z.vec = a*x.vec + (1-a)*y.vec
z.vec
```

Using $\mu_{p,z}=\mathbf{z}^{\prime}\boldsymbol{\mu}$ and $\sigma_{p,z}^{2}=\mathbf{z}^{\prime}\boldsymbol{\Sigma} \mathbf{z}$,
the expected return, variance and standard deviation of this portfolio
are:

```{r}
#| echo: true

mu.pz = as.numeric(crossprod(z.vec, mu.vec))
sig2.pz = as.numeric(t(z.vec)%*%sigma.mat%*%z.vec)
sig.pz = sqrt(sig2.pz)
c(mu.pz, sig.pz)
```

Equivalently, using $\mu_{p,z}=\alpha\mu_{p,x}+(1-\alpha)\mu_{p,y}$
and $\sigma_{p,z}^{2}=\alpha^{2}\sigma_{p,x}^{2}+(1-\alpha)^{2}\sigma_{p,y}^{2}+2\alpha(1-\alpha)\sigma_{xy}$
the expected return, variance and standard deviation of this portfolio
are:

```{r}
#| echo: true

mu.pz = a*mu.px + (1-a)*mu.py
sig.xy = as.numeric(t(x.vec)%*%sigma.mat%*%y.vec)
sig2.pz = a^2 * sig2.px + (1-a)^2 * sig2.py + 2*a*(1-a)*sig.xy
sig.pz = sqrt(sig2.pz)          
c(mu.pz, sig.pz)
```

Because $\mu_{p,z}=0.0356>\mu_{p,m}=0.0249$ the frontier portfolio
$\mathbf{z}$ is an efficient frontier portfolio. The three efficient
frontier portfolios $\mathbf{x},\,\mathbf{y}$ and $\mathbf{z}$ are
illustrated in @fig-PortMatrixEfficientPortfolios2 and
are labeled "E1" , "E2" and "E3" ,
respectively.
\hfill{}$\blacksquare$
:::

::: {#exm-create-an-efficient-frontier-portfolio-with-a-given-expected-return-from-two-efficient-minimum-variance-portfolios .example name="Create an efficient frontier portfolio with a given expected return from two efficient minimum variance portfolios"}

Given the two minimum variance portfolios with target expected returns
equal to the expected returns on Microsoft and Starbucks, respectively,
consider creating a frontier portfolio with target expected return
equal to $0.05$. To determine the value of $\alpha$ that corresponds
with this portfolio we use the equation 
$$
\mu_{p,z}=\alpha_{.05}\mu_{p,x}+(1-\alpha_{.05})\mu_{p,y}=0.05.
$$
We can then solve for $\alpha_{.05}$: 
$$
\alpha_{.05}=\frac{0.05-\mu_{p,y}}{\mu_{p,x}-\mu_{p,y}}=\frac{0.05-0.0285}{0.0427-0.0285}=1.51.
$$
Given $\alpha_{.05}$ we can solve for the portfolio weights using:
\begin{eqnarray*}
\mathbf{z}_{.05} & = & \alpha_{.05}\mathbf{x}+(1-\alpha_{.05})\mathbf{y}\\
 & = & 1.51\times\left(\begin{array}{c}
0.8275\\
-0.0907\\
0.2633
\end{array}\right)-0.514\times\left(\begin{array}{c}
0.519\\
0.273\\
0.207
\end{array}\right)\\
 & = & \left(\begin{array}{c}
0.986\\
-0.278\\
0.292
\end{array}\right).
\end{eqnarray*}
We can then compute the mean and variance of this portfolio using
$\mu_{p,z_{.05}}=\mathbf{z}_{.05}^{\prime}\boldsymbol{\mu}$ and $\sigma_{p,z_{.05}}^{2}=\mathbf{z}_{.05}^{\prime}\boldsymbol{\Sigma} \mathbf{z}_{.05}.$ 
Using R, the calculations are:

```{r}
#| echo: true

# compute alpha 
a.05 = (0.05 - mu.py)/(mu.px - mu.py) 
a.05
# compute weights 
z.05 = a.05*x.vec + (1 - a.05)*y.vec 
z.05
# compute mean and volatility 
mu.pz.05 = as.numeric(crossprod(z.05,mu.vec))
sig.pz.05 = as.numeric(sqrt(t(z.05)%*%sigma.mat%*%z.05))
c(mu.pz.05,sig.pz.05)
```

This portfolio is labeled "E4" in @fig-PortMatrixEfficientPortfolios2.
\hfill{}$\blacksquare$
:::

::: {#exm-create-an-inefficient-frontier-portfolio-with-a-given-expected-return-from-two-minimum-variance-portfolios .example name="Create an inefficient frontier portfolio with a given expected return from two minimum variance portfolios"}

Given the two minimum variance portfolios with target expected returns
equal to the expected returns on Microsoft and Starbucks, respectively,
consider creating a frontier portfolio with target expected return
equal to the expected return on Nordstrom. Then, 
$$
\mu_{p,z}=\alpha_{nord}\mu_{p,x}+(1-\alpha_{nord})\mu_{p,y}=\mu_{\textrm{nord}}=0.0015,
$$
and we can solve for $\alpha_{nord}$ using: 
$$
\alpha_{nord}=\frac{\mu_{\textrm{nord}}-\mu_{p,y}}{\mu_{p,x}-\mu_{p,y}}=\frac{0.0015-0.0285}{0.0427-0.0285}=-1.901.
$$
The portfolio weights are:
\begin{eqnarray*}
\mathbf{z}_{nord} & = & \alpha_{nord}\mathbf{x}+(1-\alpha_{nord})\mathbf{y}\\
 & = & -1.901\times\left(\begin{array}{c}
0.8275\\
-0.0907\\
0.2633
\end{array}\right)+2.9\times\left(\begin{array}{c}
0.519\\
0.273\\
0.207
\end{array}\right)\\
 & = & \left(\begin{array}{c}
-0.0064\\
0.9651\\
0.1013
\end{array}\right).
\end{eqnarray*}
Using R, the calculations are:

```{r}
#| echo: true

# compute alpha 
a.nord = (mu.vec["NORD"] - mu.py)/(mu.px - mu.py) 
a.nord 
# compute weights 
z.nord = a.nord*x.vec + (1 - a.nord)*y.vec 
z.nord 
# compute mean and volatility 
mu.pz.nord = as.numeric(crossprod(z.nord,mu.vec)) 
sig.pz.nord = as.numeric(sqrt(t(z.nord)%*%sigma.mat%*%z.nord))         
c(mu.pz.nord,sig.pz.nord)
```

Because $\mu_{p,z}=0.0015<\mu_{p,m}=0.02489$ the frontier portfolio
$\mathbf{z}$ is an inefficient frontier portfolio. This portfolio
is labeled "IE1" in  @fig-PortMatrixEfficientPortfolios2.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-PortMatrixEfficientPortfolios2
#| echo: false
#| fig-cap: Minimum variance portfolios created as convex combinations of two minimum
#|   variance portfolios. Portfolios E3, E4 and IE1 are created from portfolios E1 and
#|   E2.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

sd.vals = c(sig.px, sig.py, sig.pz, sig.pz.05) 
mu.vals = c(mu.px, mu.py, mu.pz, mu.pz.05) 
plot(sd.vals, mu.vals,  ylim=c(0, 0.06), xlim=c(0, 0.20), ylab=expression(mu[p]),      
     xlab=expression(sigma[p]), pch=16, col="green", cex=2, cex.lab=1.5)  
points(sig.gmin, mu.gmin, pch=16, cex=2, col="green")      
points(sd.vec, mu.vec, pch=16, col="blue", cex=2) 
points(sig.pz.nord, mu.pz.nord, pch=16, col="red", cex=2) 
text(sd.vals, mu.vals, labels=c("E1","E2","E3","E4"), pos=2, cex = cex.val) 
text(sig.gmin, mu.gmin, labels="GMIN", pos=2, cex = cex.val)         
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val) 
text(sig.pz.nord, mu.pz.nord, labels="IE1", pos=2, cex = cex.val) 
text(0, r.f, labels=expression(r[f]), pos=2, cex = cex.val)
```


### Algorithm for computing efficient frontier

The efficient frontier of portfolios, i.e., those frontier portfolios
with expected return greater than the expected return on the global
minimum variance portfolio, can be conveniently created using @eq-PortMatrixConvexCombination
with two specific efficient portfolios. The first efficient portfolio
is the global minimum variance portfolio @eq-PortMatrixGlobalMinObjective.
The second efficient portfolio is the efficient portfolio whose target
expected return is equal to the highest expected return among all
of the assets under consideration. The choice of these two efficient
portfolios makes it easy to produce a nice plot of the efficient frontier.
The steps for constructing the efficient frontier are: 

1.  Compute the global minimum variance portfolio $\mathbf{m}$ by solving @eq-PortMatrixGlobalMinObjective, and compute $\mu_{p,m}=\mathbf{m}^{\prime}\boldsymbol{\mu}$ and $\sigma_{p,m}^{2}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}$. 
2.  Compute the efficient portfolio $\mathbf{x}$ with target expected return equal to the maximum expected return of the assets under consideration. That is, solve @eq-PortMatrixEfficientPortfolio2 with $\mu_{0}=\max\{\mu_{1},\ldots,\mu_{N}\}$, and compute $\mu_{p,x}=\mathbf{x}^{\prime}\boldsymbol{\mu}$ and $\sigma_{p,x}^{2}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}$. 
3.  Compute $\mathrm{cov}(R_{p,m},R_{p,x})=\sigma_{mx}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{x}$. 
4.  Create an initial grid of $\alpha$ values $\{0,0.1,\ldots0.9,1\}$, and compute the frontier portfolios $\mathbf{z}$ using 
$$
\mathbf{z}=\alpha\times\mathbf{x}+(1-\alpha)\times\mathbf{m},
$$
and compute their expected returns and variances using @eq-PortMatrixConvexCombination, @eq-PortMatrixMupz and @eq-PortMatrixSig2pz, respectively.
5.  Plot $\mu_{p,z}$ against $\sigma_{p,z}$ and adjust the grid of $\alpha$ values appropriately to create a nice plot. Negative values of $\alpha$ will give inefficient frontier portfolios. Values of $\alpha$ greater than one will give efficient frontier portfolios with expected returns greater than $\mu_{0}.$


::: {#exm-compute-and-plot-the-efficient-frontier-of-risky-assets-markowitz-bullet .example name="Compute and plot the efficient frontier of risky assets (Markowitz bullet)"}

To compute the efficient frontier from the three risky assets in Table
@tbl-Table-ThreeAssetExample in R use:

```{r}
#| echo: true

a = seq(from=0, to=1, by=0.1) 
n.a = length(a) 
z.mat = matrix(0, n.a, 3) 
colnames(z.mat) = names(mu.vec) 
mu.z = rep(0, n.a) 
sig2.z = rep(0, n.a) 
sig.mx = t(m.vec)%*%sigma.mat%*%x.vec 
for (i in 1:n.a) {   
   z.mat[i, ] = a[i]*x.vec + (1-a[i])*m.vec   
   mu.z[i] = a[i]*mu.px + (1-a[i])*mu.gmin   
   sig2.z[i] = a[i]^2 * sig2.px + (1-a[i])^2 * sig2.gmin + 
      2*a[i]*(1-a[i])*sig.mx 
}
```

The variables `z.mat`, `mu.z` and `sig2.z` contain
the weights, expected returns and variances, respectively, of the
efficient frontier portfolios for a grid of $\alpha$ values between
0 and 1. The resulting efficient frontier is illustrated in 
@fig-PortMatrixEfficientFrontier created with:

```{r}
#| label: fig-PortMatrixEfficientFrontier
#| echo: true
#| fig-cap: Efficient frontier computed from example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sqrt(sig2.z), mu.z, type="b", ylim=c(0, 0.06), xlim=c(0, 0.16), 
     pch=16, col="green", cex = cex.val, ylab=expression(mu[p]), 
     xlab=expression(sigma[p]),cex.lab=1.5)
points(sd.vec, mu.vec, pch=16, cex=2, lwd=2, col="blue")
points(sig.gmin, mu.gmin, pch=16, col="green", cex=2)
points(sig.px, mu.px, pch=16, col="green", cex=2)
text(sig.gmin, mu.gmin, labels="GLOBAL MIN", pos=2, cex = cex.val)
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val)
text(sig.px, mu.px, labels="E1", pos=2, cex = cex.val)
```


Each point on the efficient frontier is a portfolio of Microsoft,
Nordstrom and Starbucks. It is instructive to visualize the weights
in these portfolios as we move along the frontier from the global
minimum variance portfolio to the efficient portfolio with expected
return equal to the expected return on Microsoft. We can do this easily
using the **PerformanceAnalytics** function `chart.StackedBar()`:

```{r}
#| label: fig-PortMatrixFrontierWeights
#| echo: true
#| fig-cap: Portfolio weights in efficient frontier portfolios. x-axis is portfolio standard
#|   deviation
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

chart.StackedBar(z.mat, xaxis.labels=round(sqrt(sig2.z),digits=3), 
                 legend.loc="upper",                   
                 colorset=rainbow8equal[1:3], ylab="Weights")

```

The resulting plot is shown in  @fig-PortMatrixFrontierWeights.
As we move along the frontier, the allocation to Microsoft increases
and the allocation to Nordstrom decreases whereas the allocation to
Starbucks stays about the same.
\hfill{}$\blacksquare$
:::

## Computing Efficient Portfolios of N risky Assets and a Risk-Free Asset Using Matrix Algebra {#sec-Efficient-portfolios-of}
\index{Efficient portfolio of multiple risky assets and a risk-free asset}
In Chapter @sec-Introduction-to-Portfolio-Theory, we showed
that efficient portfolios of two risky assets and a single risk-free
(T-Bill) asset are portfolios consisting of the highest Sharpe ratio
portfolio (tangency portfolio) and the T-Bill. With three or more
risky assets and a T-Bill the same result holds.

### Computing the tangency portfolio using matrix algebra
\index{Tangency portfolio}
The tangency portfolio is the portfolio of risky assets that has the
highest Sharpe ratio. The tangency portfolio, denoted $\mathbf{t}=(t_{\textrm{1}},\ldots,t_{N})^{\prime}$,
solves the constrained maximization problem: 
$$
\underset{\mathbf{t}}{\max}~\frac{\mathbf{t}^{\prime}\boldsymbol{\mu}-r_{f}}{(\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t})^{{\frac{1}{2}}}}=\frac{\mu_{p,t}-r_{f}}{\sigma_{p,t}}\textrm{ s.t. }\mathbf{t}^{\prime}\mathbf{1}=1,
$$ {#eq-PortMatrixMaxSharpeRatio}
where $\mu_{p,t}=\mathbf{t}^{\prime}\boldsymbol{\mu}$ and $\sigma_{p,t}=(\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t})^{{\frac{1}{2}}}$.
The Lagrangian for this problem is: 
$$
L(\mathbf{t},\lambda)=\left(\mathbf{t}^{\prime}\boldsymbol{\mu}-r_{f}\right)(\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t})^{-{\frac{1}{2}}}+\lambda(\mathbf{t}^{\prime}\mathbf{1}-1).
$$
Using the chain rule, the first order conditions are: 
\begin{align*}
\frac{\partial L(\mathbf{t},\lambda)}{\partial\mathbf{t}} & =\boldsymbol{\mu}(\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t})^{-{\frac{1}{2}}}-\left(\mathbf{t}^{\prime}\boldsymbol{\mu}-r_{f}\right)(\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t})^{-3/2}\boldsymbol{\Sigma} \mathbf{t}+\lambda\mathbf{1}=\mathbf{0},\\
\frac{\partial L(\mathbf{t},\lambda)}{\partial\lambda} & =\mathbf{t}^{\prime}\mathbf{1}-1=0.
\end{align*}
After much tedious algebra, it can be shown that the solution for
$\mathbf{t}$ has a nice simple expression: 
$$
\mathbf{t}=\frac{\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}.
$$ {#eq-PortMatrixTangencySolution}
The formula for the tangency portfolio @eq-PortMatrixTangencySolution
looks similar to the formula for the global minimum variance portfolio
@eq-PortMatrixGlobalMinSolutionM. Both formulas have $\boldsymbol{\Sigma}^{-1}$
in the numerator and $\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}$ in
the denominator.
 
::: {callout-note}
## Remark
The location of the tangency portfolio, and the sign of the Sharpe
ratio, depends on the relationship between the risk-free rate $r_{f}$
and the expected return on the global minimum variance portfolio $\mu_{p,m}$.
If $\mu_{p,m}>r_{f}$, which is the usual case, then the tangency
portfolio will have a positive Sharpe ratio. If $\mu_{p,m}<r_{f}$,
which could occur when stock prices are falling and the economy is
in a recession, then the tangency portfolio will have a negative Sharpe
slope. In this case, efficient portfolios involve shorting the tangency
portfolio and investing the proceeds in T-Bills.^[For a mathematical proof of these results, see Ingersoll (1987).]
:::

::: {#exm-computing-the-tangency-portfolio .example name="Computing the tangency portfolio"}

Suppose $r_{f}=0.005$. To compute the tangency portfolio @eq-PortMatrixTangencySolution
in R for the three risky assets in Table @tbl-Table-ThreeAssetExample
use:

```{r}
#| echo: false

r.f = 0.005
sigma.inv.mat = solve(sigma.mat)
one.vec = rep(1, 3)
mu.minus.rf = mu.vec - r.f*one.vec
top.mat = sigma.inv.mat%*%mu.minus.rf
bot.val = as.numeric(t(one.vec)%*%top.mat)
t.vec = top.mat[,1]/bot.val
t.vec
```

The tangency portfolio has weights $t_{\textrm{msft}}=1.027,$ $t_{\textrm{nord}}=-0.326$
and $t_{\textrm{sbux}}=0.299,$ and is given by the vector $\mathbf{t}=(1.027,-0.326,0.299)^{\prime}.$
Notice that Nordstrom, which has the lowest mean return, is sold short
in the tangency portfolio. The expected return on the tangency portfolio,
$\mu_{p,t}=\mathbf{t}^{\prime}\boldsymbol{\mu}$, is:

```{r}
#| echo: true

mu.t = as.numeric(crossprod(t.vec, mu.vec))
mu.t
```

The portfolio variance, $\sigma_{p,t}^{2}=\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t}$,
and standard deviation, $\sigma_{p,t}$, are:

```{r}
#| echo: true

sig2.t = as.numeric(t(t.vec)%*%sigma.mat%*%t.vec)
sig.t = sqrt(sig2.t)
c(sig2.t, sig.t)
```

Because $r_{f}=0.005<\mu_{p,m}=0.0249$ the tangency portfolio has
a positive Sharpe's ratio/slope given by: 

```{r}
#| echo: true

SR.t = (mu.t - r.f)/sig.t 
SR.t
```

The tangency portfolio is illustrated in @fig-PortMatrixTangencyPortfolio.
It is the portfolio on the efficient frontier of risky assets in which
a straight line drawn from the risk-free rate to the tangency portfolio
(green line) is just tangent to the efficient frontier (blue dots).

```{r}
#| label: fig-PortMatrixTangencyPortfolio
#| echo: false
#| fig-cap: Tangency portfolio from example data.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.t = seq(0, 2, by=0.1) 
mu.pe = r.f + x.t*(mu.t - r.f) 
sig.pe = x.t*sig.t  
slope.t = (mu.t - r.f)/sig.t
a = seq(from=0, to=2, by=0.1) 
n.a = length(a) 
z.mat = matrix(0, n.a, 3) 
colnames(z.mat) = names(mu.vec) 
mu.z = rep(0, n.a) 
sig2.z = rep(0, n.a) 
sig.mx = t(m.vec)%*%sigma.mat%*%x.vec 
for (i in 1:n.a) {   
   z.mat[i, ] = a[i]*x.vec + (1-a[i])*m.vec   
   mu.z[i] = a[i]*mu.px + (1-a[i])*mu.gmin   
   sig2.z[i] = a[i]^2 * sig2.px + (1-a[i])^2 * sig2.gmin + 2*a[i]*(1-a[i])*sig.mx 
}
plot(sqrt(sig2.z), mu.z, type="b", ylim=c(0, 0.08), xlim=c(0, 0.17),       
     pch=16, col="blue", cex=1.15, ylab=expression(mu[p]), xlab=expression(sigma[p]),
     cex.lab=1.5) 
abline(a=r.f, b=slope.t, col="green", lwd=2) 
points(sig.t, mu.t, pch=16, col="green", cex=1.15) 
points(sd.vec, mu.vec, pch=16, cex=2, col="black") 
text(sig.gmin, mu.gmin, labels="GMIN", pos=4, cex=1.15) 
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex=1.15) 
text(sig.t, mu.t, labels="TAN", pos=4,cex=1.15) 
text(0, r.f, labels=expression(r[f]), pos=2, cex=1.15)
```

\hfill{}$\blacksquare$
:::

### Alternative derivation of the tangency portfolio

The derivation of tangency portfolio formula @eq-PortMatrixTangencySolution
from the optimization problem @eq-PortMatrixMaxSharpeRatio
is a very tedious problem. It can be derived in a different way as
follows. Consider forming portfolios of $N$ risky assets with return
vector $\mathbf{R}$ and T-bills (risk-free asset) with constant return
$r_{f}$. Let $\mathbf{x}$ denote the $N\times1$ vector of risky
asset weights and let $x_{f}$ denote the safe asset weight and assume
that $\mathbf{x}^{\prime}\mathbf{1}+x_{f}=1$ so that all wealth is
allocated to these assets. The portfolio return is: 
$$
R_{p,x}=\mathbf{x}^{\prime}\mathbf{R}+x_{f}r_{f}=\mathbf{x}^{\prime}\mathbf{R}+(1-\mathbf{x}^{\prime}\mathbf{1})r_{f}=r_{f}+\mathbf{x}^{\prime}(\mathbf{R}-r_{f}\cdot\mathbf{1}).
$$
The portfolio excess return is: 
$$
R_{p,x}-r_{f}=\mathbf{x}^{\prime}(\mathbf{R}-r_{f}\cdot\mathbf{1)}.
$$ {#eq-port-excess-return}
The expected portfolio excess return (risk premium) and portfolio
variance are: 
$$
\mu_{p,x}-r_{f} =\mathbf{x}^{\prime}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1)},
$$ {#eq-port-risk-premium}
$$
\sigma_{p,x}^{2} =\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}.
$$ {#eq-port-var}
For notational simplicity, define $\mathbf{\tilde{R}}=\mathbf{R}-r_{f}\cdot\mathbf{1}$,
$\tilde{\boldsymbol{\mu}}=\boldsymbol{\mu}-r_{f}\cdot\mathbf{1}$, $\tilde{R}_{p,x}=R_{p,x}-r_{f}$,
and $\tilde{\mu}_{p,x}=\mu_{p,x}-r_{f}$. Then @eq-port-excess-return
and @eq-port-risk-premium can be re-expressed as: 
$$
\tilde{R}_{p,x} =\mathbf{x}^{\prime}\tilde{\mathbf{R}},
$$ {#eq-port-excess-return-2}
$$
\tilde{\mu}_{p,x} =\mathbf{x}^{\prime}\tilde{\boldsymbol{\mu}}.
$$ {#eq-port-risk-premium-2}
To find the minimum variance portfolio of risky assets and a risk
free asset that achieves the target excess return $\tilde{\mu}_{p,0}=\mu_{p,0}-r_{f}$
we solve the minimization problem: 
$$
\min_{\mathbf{x}}~\sigma_{p,x}^{2}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\textrm{ s.t. }\tilde{\mu}_{p,x}=\tilde{\mu}_{p,0}.
$$
Note that $\mathbf{x}^{\prime}\mathbf{1}=1$ is not a constraint because
wealth need not all be allocated to the risky assets; some wealth
may be held in the riskless asset. The Lagrangian is: 
$$
L(\mathbf{x},\lambda)=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}+\lambda(\mathbf{x}^{\prime}\tilde{\boldsymbol{\mu}}-\tilde{\mu}_{p,0}).
$$
The first order conditions for a minimum are: 
$$
\frac{\partial L(\mathbf{x},\lambda)}{\partial\mathbf{x}} =2\boldsymbol{\Sigma} \mathbf{x}+\lambda\tilde{\boldsymbol{\mu}}=\mathbf{0},
$$ {#eq-foc-tan-1}
$$
\frac{\partial L(\mathbf{x},\lambda)}{\partial\lambda}  =\mathbf{x}^{\prime}\tilde{\boldsymbol{\mu}}-\tilde{\mu}_{p,0}=0.
$$ {#eq-foc-tan-2}
Using the @eq-foc-tan-1, we can solve for $\mathbf{x}$
in terms of $\lambda$: 
$$
\mathbf{x}=-\frac{1}{2}\lambda\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}.
$$ {#eq-x-tan-1}
@eq-foc-tan-2 implies that $\mathbf{x}^{\prime}\tilde{\boldsymbol{\mu}}=\tilde{\boldsymbol{\mu}}^{\prime}\mathbf{x}=\tilde{\mu}_{p,0}$.
Then pre-multiplying @eq-x-tan-1 by $\tilde{\boldsymbol{\mu}}^{\prime}$
gives: 
$$
\tilde{\boldsymbol{\mu}}^{\prime}\mathbf{x}=-\frac{1}{2}\lambda\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}=\tilde{\mu}_{p,0},
$$
which we can use to solve for $\lambda$: 
$$
\lambda=-\frac{2\tilde{\mu}_{p,0}}{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}.
$$ {#eq-lambda-tan}
Plugging @eq-lambda-tan into @eq-x-tan-1 then gives
the solution for $\mathbf{x}$: 
$$
\mathbf{x}=-\frac{1}{2}\lambda\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}=-\frac{1}{2}\left(-\frac{2\tilde{\mu}_{p,0}}{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}\right)\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}=\tilde{\mu}_{p,0}\cdot\frac{\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}.
$$ {#eq-x-tan-2}
The solution for $x_{f}$ is then $1-\mathbf{x}^{\prime}1$. 

Now, the tangency portfolio $\mathbf{t}$ is 100\% invested in risky
assets so that $\mathbf{t}^{\prime}\mathbf{1}=\mathbf{1}^{\prime}\mathbf{t}=1$.
Using @eq-x-tan-2, the tangency portfolio satisfies: 
$$
\mathbf{1}^{\prime}\mathbf{t}=\tilde{\mu}_{p,t}\cdot\frac{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}=1,
$$
which implies that, 
$$
\tilde{\mu}_{p,t}=\frac{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}.
$$ {#eq-risk-premium-tan}
Plugging @eq-risk-premium-tan back into @eq-x-tan-2
then gives an explicit solution for $\mathbf{t}$: 
\begin{align*}
\mathbf{t} & \mathbf{=}\left(\frac{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}\right)\frac{\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}{\tilde{\boldsymbol{\mu}}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}=\frac{\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}\tilde{\boldsymbol{\mu}}}\\
 & =\frac{\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})},
\end{align*}
which is the result @eq-PortMatrixTangencySolution we got
from finding the portfolio of risky assets that has the maximum Sharpe
ratio.


### Mutual fund separation theorem again
\index{Mutual fund separation theorem}
When there is a risk-free asset (T-bill) available, the efficient
frontier of T-bills and risky assets consists of portfolios of T-bills
and the tangency portfolio. The expected return and standard deviation
values of any such efficient portfolio are given by: 
$$
\mu_{p}^{e} =r_{f}+x_{t}(\mu_{p,t}-r_{f}),
$$ {#eq-PortMatrixMupEfficient}
$$
\sigma_{p}^{e} =x_{t}\sigma_{p,t},
$$ {#eq-PortMatrixSigpEfficient}
where $x_{t}$ represents the fraction of wealth invested in the tangency
portfolio ($1-x_{t}$ represents the fraction of wealth invested in
T-Bills), and $\mu_{p,t}=\mathbf{t}^{\prime}\boldsymbol{\mu}$ and $\sigma_{p,t}=(\mathbf{t}^{\prime}\boldsymbol{\Sigma} \mathbf{t})^{1/2}$
are the expected return and standard deviation on the tangency portfolio,
respectively. Recall, this result is known as the _mutual fund
separation theorem_. The tangency portfolio can be considered as a
mutual fund of the risky assets, where the shares of the assets in
the mutual fund are determined by the tangency portfolio weights,
and the T-bill can be considered as a mutual fund of risk-free assets.
The expected return-risk trade-off of these portfolios is given by
the line connecting the risk-free rate to the tangency point on the
efficient frontier of risky asset only portfolios. Which combination
of the tangency portfolio and the T-bill an investor will choose depends
on the investor's risk preferences. If the investor is very risk averse
and prefers portfolios with very low volatility, then she will choose
a combination with very little weight in the tangency portfolio and
a lot of weight in the T-bill. This will produce a portfolio with
an expected return close to the risk-free rate and a variance that
is close to zero. If the investor can tolerate a large amount of volatility,
then she will prefer a portfolio with a high expected return regardless
of volatility. This portfolio may involve borrowing at the risk-free
rate (leveraging) and investing the proceeds in the tangency portfolio
to achieve a high expected return.


::: {#exm-efficient-portfolios-of-three-risky-assets-and-t-bills-chosen-by-risk-averse-and-risk-tolerant-investors .example name="Efficient portfolios of three risky assets and T-bills chosen by risk averse and risk tolerant investors"}

Consider the tangency portfolio computed from the example data in
Table @tbl-Table-ThreeAssetExample with $r_{f}=0.005$. This
portfolio is:

```{r}
#| echo: true

ans = c(t.vec, mu.t, sig.t)
names(ans)[4:5] = c("mu.t", "sigma.t")
ans
```

The efficient portfolios of T-Bills and the tangency portfolio is
illustrated in  @fig-PortMatrixEfficientPortfolioExample.

We want to compute an efficient portfolio that would be preferred
by a highly risk averse investor, and a portfolio that would be preferred
by a highly risk tolerant investor. A highly risk averse investor
might have a low volatility (risk) target for his efficient portfolio.
For example, suppose the volatility target is $\sigma_{p}^{e}=0.02$
or $2\%$. Using @eq-PortMatrixSigpEfficient and solving for
$x_{t}$, the weights in the tangency portfolio and the T-Bill are:

```{r}
#| echo: true

x.t.02 = 0.02/sig.t
c(x.t.02, 1-x.t.02)
```

In this efficient portfolio, the weights in the risky assets are proportional
to the weights in the tangency portfolio:

```{r}
#| echo: true

x.t.02*t.vec
```

The expected return and volatility values of this portfolio are:

```{r}
#| echo: true

mu.t.02 = x.t.02*mu.t + (1-x.t.02)*r.f
sig.t.02 = x.t.02*sig.t
c(mu.t.02, sig.t.02)
```

These values are illustrated in  @fig-PortMatrixEfficientPortfolioExample
as the portfolio labeled "E1" . 

A highly risk tolerant investor might have a high expected return
target for his efficient portfolio. For example, suppose the expected
return target is $\mu_{p}^{e}=0.07$ or $7\%$. Using @eq-PortMatrixMupEfficient
and solving for the $x_{t}$, the weights in the tangency portfolio
and the T-Bill are:

```{r}
#| echo: true

x.t.07 = (0.07 - r.f)/(mu.t - r.f)
c(x.t.07, 1-x.t.07)
```

Notice that this portfolio involves borrowing at the T-Bill rate (leveraging)
and investing the proceeds in the tangency portfolio. In this efficient
portfolio, the weights in the risky assets are:

```{r}
#| echo: true

x.t.07*t.vec
```

The expected return and volatility values of this portfolio are:

```{r}
#| echo: true

mu.t.07 = x.t.07*mu.t + (1-x.t.07)*r.f
sig.t.07 = x.t.07*sig.t
c(mu.t.07, sig.t.07)
```

In order to achieve the target expected return of 7\%, the investor
must tolerate a 15.47\% volatility. These values are illustrated in
@fig-PortMatrixEfficientPortfolioExample as the portfolio
labeled "E2" .
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-PortMatrixEfficientPortfolioExample
#| echo: false
#| fig-cap: Tangency portfolio
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sqrt(sig2.z), mu.z, type="b", ylim=c(0, 0.08), xlim=c(0, 0.17),       
     pch=16, col="blue", cex=cex.val, ylab=expression(mu[p]), xlab=expression(sigma[p]), cex.lab=1.5) 
abline(a=r.f, b=slope.t, col="green", lwd=2) 
points(sig.t, mu.t, pch=16, col="green", cex=2) 
points(sig.t.02, mu.t.02, pch=16, col="green", cex=2) 
points(sig.t.07, mu.t.07, pch=16, col="green", cex=2) 
points(sd.vec, mu.vec, pch=16, col="black", cex=cex.val) 
text(sig.gmin, mu.gmin, labels="Global min", pos=4, cex=cex.val) 
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex=cex.val) 
text(sig.t, mu.t, labels="tangency", pos=4, cex=cex.val) 
text(sig.t.02, mu.t.02, labels="E1", pos=3, cex=cex.val) 
text(sig.t.07, mu.t.07, labels="E2", pos=3, cex=cex.val) 
text(0, r.f, labels=expression(r[f]), pos=2, cex=cex.val)
segments(0.02, 0, 0.02, mu.t.02, col="green", lty="dashed", lwd=2) 
text(0.02, 0, labels="0.02", col="black", lwd=2, pos=3) 
segments(0, mu.t.07, sig.t.07, mu.t.07, col="green", lty="dashed", lwd=2) 
text(0, 0.07, labels="0.07", col="black", lwd=2, pos=4)
```


## Computational Problems with Very Large Portfolios {#sec-Computational-Problems-with}
\index{Large portfolios}
In principle, mean-variance portfolio analysis can be applied in situations
in which there is a very large number of risky assets (e.g., $N=5,000)$.
However, there are a number of practical problems that can arise.
First, the computation of efficient portfolios requires inverting
the $N\times N$ asset return covariance matrix $\boldsymbol{\Sigma}$.
When $N$ is very large, inverting $\boldsymbol{\Sigma}$ can be computationally
burdensome. Second, the practical application of the theory requires
the estimation of $\boldsymbol{\Sigma}$. Recall, there are $N$ variance
terms and $N(N-1)/2$ unique covariance terms in $\boldsymbol{\Sigma}$.
When $N=5,000$, there are $12,502,500$ unique elements of $\boldsymbol{\Sigma}$
to estimate. And since each estimated element of $\boldsymbol{\Sigma}$
has estimation error, there is a tremendous amount of estimation error
in the estimate of $\boldsymbol{\Sigma}$. There is an additional problem
with the estimation of $\boldsymbol{\Sigma}$ using the sample covariance
matrix of asset returns when $N$ is very large. If the number of
assets, $N$, is greater than the number of sample observations, $T$,
then the $N\times N$ sample covariance matrix: 
\begin{eqnarray*}
\hat{\boldsymbol{\Sigma}} & = & \frac{1}{T-1}\sum_{t=1}^{T}(\mathbf{R}_{t}-\hat{\boldsymbol{\mu}})(\mathbf{R}_{t}-\hat{\boldsymbol{\mu}})^{\prime},\\
\hat{\boldsymbol{\mu}} & = & \frac{1}{T}\sum_{t=1}^{T}\mathbf{R}_{t},
\end{eqnarray*}
is only positive semi-definite and less than full rank $N$. This
means that $\hat{\boldsymbol{\Sigma}}$ is not invertible and so mean-variance
efficient portfolios cannot be uniquely computed. This problem can
happen often. For example, suppose $N=5,000$. For the sample covariance
matrix to be full rank, you need at least $T=5,000$ sample observations.
For daily data, this means you would need $5,000/250=20$ years of
daily data.^[Recall, there are approximately 250 trading days per year.]
For weekly data, you would need $5000/52=96.2$ years of weekly data.
For monthly data, you would need $5,000/12=417$ years of monthly
data. 

::: {#exm-nonsingular-sample-return-covariance-matrix .example name="Nonsingular sample return covariance matrix"}

To illustrate the rank failure of $\hat{\boldsymbol{\Sigma}}$ that occurs
when the number of assets $N$ is greater than the number of data
observations $T$, consider computing $\hat{\boldsymbol{\Sigma}}$ for
the six Vanguard mutual funds in the **introcompfinr** data object
`VanguardPrices` using only five monthly observations:

```{r}
#| echo: true

colnames(VanguardPrices) 
range(index(VanguardPrices)) 
VanguardRetS = na.omit(Return.calculate(VanguardPrices, 
                                        method="simple")) 
covhat = cov(VanguardRetS[1:5, ])
```

A quick way to determine if $\hat{\boldsymbol{\Sigma}}$ is full rank
(and invertible) is to compute the Cholesky decomposition $\hat{\boldsymbol{\Sigma}}=\hat{\mathbf{C}}\hat{\mathbf{C}}^{\prime}$,
where $\hat{\mathbf{C}}$ is a lower triangular matrix with non-negative
diagonal elements. If all of the diagonal elements of $\hat{\mathbf{C}}$
are positive then $\hat{\boldsymbol{\Sigma}}$ is positive definite, full
rank, and invertible. In R, we compute $\hat{\mathbf{C}}$ using the
function `chol()`:

```{r}
#| echo: true

# chol(covhat) # uncomment this one to see the result
```

Here, `chol()` returns an error that indicates $\hat{\boldsymbol{\Sigma}}$
is not positive definite and less than full rank. If we try to invert
$\hat{\boldsymbol{\Sigma}}$ using `solve()` we will also get an
error indicating $\hat{\boldsymbol{\Sigma}}$ is not invertible:^[The **Matrix** function `rankMatrix()` can be used to
compute the exact rank of a matrix.]

```{r}
#| echo: true

# solve(covhat) # uncomment this one to see the result
```

\hfill{}$\blacksquare$
:::

Due to these practical problems of using the sample covariance matrix
$\hat{\boldsymbol{\Sigma}}$ to compute mean-variance efficient portfolios
when $N$ is large, there is a need for alternative methods for estimating
$\boldsymbol{\Sigma}$ when $N$ is large. One such method based on the
*Single Index Model*\index{Single index model} for returns is presented in Chapter @sec-Single-Index-Model.

## Portfolio Analysis Functions in R {#sec-Portfolio-Analysis-Functions}
\index{Portfolio analysis functions}
The package **IntroCompFinR** contains a few R functions for
computing Markowitz mean-variance efficient portfolios allowing for
short sales using matrix algebra computations. These functions allow
for the easy computation of the global minimum variance portfolio,
an efficient portfolio with a given target expected return, the tangency
portfolio, and the efficient frontier. These functions are summarized
in Table @tbl-TablePorfolioFunctions.

------------------------------------------------------------------------------------------
**Function**                **Description**
------------------------    --------------------------------------------------------------
`get_portfolio`             create portfolio object 
  
`globalmin_portfolio`       compute global minimum variance portfolio 
  
`efficient_portfolio`       compute minimum variance portfolio subject to target return 
  
`tangency_portfolio`        compute tangency portfolio 
  
`efficient_frontier`        compute efficient frontier of risky assets 
------------------------------------------------------------------------------------------

:  **introcompfinr** functions for computing mean-variance efficient portfolios {#tbl-TablePorfolioFunctions}


The following examples illustrate the use of the functions in Table
@tbl-TablePorfolioFunctions using the example data in Table
@tbl-Table-ThreeAssetExample:

```{r}
#| echo: true
asset.names <- c("MSFT", "NORD", "SBUX")
mu.vec = c(0.0427, 0.0015, 0.0285)
sigma.mat = matrix(c(0.0100, 0.0018, 0.0011,
           0.0018, 0.0109, 0.0026,
           0.0011, 0.0026, 0.0199),
         nrow=3, ncol=3)
r.f=0.005
mu.vec
sigma.mat
r.f
```

To specify a `portfolio` object, you need an expected
return vector and covariance matrix for the assets under consideration
as well as a vector of portfolio weights. To create an equally weighted
portfolio use:

```{r}
#| echo: true

ew = rep(1,3)/3
equalWeight.portfolio = get_portfolio(er_vec=mu.vec,cov_mat=sigma.mat,weight_vec=ew)
class(equalWeight.portfolio)
```

Portfolio objects have the following components:

```{r}
#| echo: true

names(equalWeight.portfolio)
```

There are `print()`, `summary()` and `plot()`
methods for portfolio objects. The `print()` method gives:

```{r}
#| echo: true

equalWeight.portfolio
```

The `plot()` method shows a bar chart of the portfolio weights:

```{r}
#| label: fig-PortMatrixPlotMethod
#| echo: true
#| fig-cap: Plot method for objects of class portfolio
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(equalWeight.portfolio, col="cornflowerblue")
```


The resulting plot is shown in @fig-PortMatrixPlotMethod.
The global minimum variance portfolio (allowing for short sales) $\mathbf{m}$
solves the optimization problem @eq-PortMatrixGlobalMinObjective
and is computed using @eq-PortMatrixGlobalMinSolution. To
compute this portfolio use the function `globalMin.portfolio()`:

```{r}
#| echo: true

gmin.port = globalmin_portfolio(mu.vec, sigma.mat)
class(gmin.port)
gmin.port
```

A mean-variance efficient portfolio $\mathbf{x}$ that achieves the
target expected return $\mu_{0}$ solves the optimization problem
@eq-PortMatrixEfficientPortfolio2 and is computed using @eq-PortMatrixEfficientPortfolioSolution.
To compute this portfolio for the target expected return $\mu_{0}=E[R_{\textrm{msft}}]=0.0427$
use the `efficient.portfolio()` function:

```{r}
#| echo: true

target.return = mu.vec[1]
e.port.msft = efficient_portfolio(mu.vec, sigma.mat, target.return)
class(e.port.msft)
e.port.msft
```

The tangency portfolio $\mathbf{t}$ is the portfolio of risky assets
with the highest Sharpe's slope and has solutions given by @eq-PortMatrixTangencySolution.
To compute this portfolio with $r_{f}=0.005$ use the `tangency.portfolio()`
function:

```{r}
#| echo: true

tan.port = tangency_portfolio(mu.vec, sigma.mat, r.f)
class(tan.port)
tan.port
```

The the set of efficient portfolios of risky assets can be computed
as a convex combination of any two efficient portfolios. It is convenient
to use the global minimum variance portfolio as one portfolio and
an efficient portfolio with target expected return equal to the maximum
expected return of the assets under consideration as the other portfolio.
Call these portfolios $\mathbf{m}$ and $\mathbf{x}$, respectively.
For any number $\alpha$, another efficient portfolio can be computed
as: 
$$
\mathbf{z}=\alpha\mathbf{x}+(1-\alpha)\mathbf{m}
$$
The function `efficient.frontier()` constructs the set of efficient
portfolios using this method for a collection of $\alpha$ values
on an equally spaced grid between $\alpha_{min}$ and $\alpha_{max}$.
For example, to compute 20 efficient portfolios for values of $\alpha$
between -2 and 1.5 use:

```{r}
#| echo: true

ef = efficient_frontier(mu.vec, sigma.mat, alpha_min=-0.5, 
                        alpha_max=2, nport=20)
attributes(ef)
ef
```

Use the `summary()` method to show the weights of these portfolios.
Use the `plot()` method to create a simple plot the efficient
frontier:

```{r}
#| label: fig-PortMatrixMarkowitzPlotMethod
#| echo: true
#| fig-cap: Plot method for objects of class Markowitz
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(ef)
```


The resulting plot is shown in @fig-PortMatrixMarkowitzPlotMethod.
To create a more elaborate plot of the efficient frontier showing
the original assets and the tangency portfolio use:

```{r}
#| label: fig-PortMatrixEfficientFrontierPlot
#| echo: true
#| fig-cap: Efficient frontier plot.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(ef, plot.assets=T, col="blue", pch=16) 
points(gmin.port$sd_port, gmin.port$er_port, col="green", pch=16, cex=2) 
text(gmin.port$sd_port, gmin.port$er_port, labels = "Global min", pos = 4) 
points(tan.port$sd_port, tan.port$er_port, col="red", pch=16, cex=2) 
text(tan.port$sd_port, tan.port$er_port, labels = "Tangency", pos = 3) 
sr.tan = (tan.port$er_port - r.f)/tan.port$sd_port 
abline(a=r.f, b=sr.tan, col="green", lwd=2)
```

The resulting plot is shown in @fig-PortMatrixEfficientFrontierPlot.


## Application to Vanguard Mutual Fund

In this section, we use mean-variance portfolio theory to study asset
allocation among a collection of Vanguard mutual Funds. To motivate
this example, consider an investor who is deciding how to allocate
savings in an employee-sponsored retirement plan.^[Common employee-sponsored retirement portfolios in the US are 401K
plans (for employees of private companies) and 403B plans (for employees
of government agencies). ] Typical plans only allow participants to invest in a limited set
of mutual funds. An important limitation of investing in mutual funds
is that they cannot be sold short. 


The **introcompfinr** data object `VanguardPrices` contains
monthly closing prices on six diversified and passively managed Vanguard
mutual funds over the period January, 1995 through December, 2014:

```{r}
#| echo: true

VanguardPrices = as.xts(VanguardPrices) 
colnames(VanguardPrices) 
range(index(VanguardPrices))
```

Table @tbl-PortMatrixTableVanguard gives a brief description
of the six funds.^[More detailed information on these funds is available at (finance.yahoo.com).]


------------------------------------------------------------------------------------------------
Fund Ticker       Fund Name                             Description
-------------     ---------------------------------     ----------------------------------------
vfinx             Vanguard 500 Index                    Tracks the S\&P 500 index

veurx             Vanguard European Stock Index         Tracks the MSCI Europe index

veiex             Vanguard Emerging Markets Index       Tracks the MSCI Emerging Markets index

vbltx             Vanguard Long-Term Bond Index         Tracks a US long-term bond index

vbisx             Vanguard Short-Term Bond Index        Tracks a US short-term bond index

vpacx             Vanguard Pacific Stock Index          Tracks the MSCI Pacific index
------------------------------------------------------------------------------------------------

: Vanguard mutual funds. {#tbl-PortMatrixTableVanguard} 


We construct mean-variance efficient portfolios using annualized GWN
model estimates of the expected return vector and covariance matrix
based on simple returns over the sub-period January, 2010 through
December, 2014. These estimates are computed using:

```{r}
#| echo: true

VanguardRetS = na.omit(Return.calculate(VanguardPrices, 
                                        method="simple")) 
smpl = "2010-1::2014-12"
muhat = 12*colMeans(VanguardRetS[smpl]) 
covhat = 12*cov(VanguardRetS[smpl])
```

The estimated annualized expected returns and standard deviations
are:

```{r}
#| echo: true

muhat
sqrt(diag(covhat))
```

The stock funds have higher average returns and volatilities than
the bond funds. The annual risk-free rate at the end of this period
is approximately 0.1\%. Using this rate, the estimated Sharpe ratios
on the six funds are:

```{r}
#| echo: true

r.f = 0.001 
(muhat - r.f)/sqrt(diag(covhat))
```

Interestingly, the short term US bond fund vbisx has the highest estimated
annualized Sharpe ratio at 1.36 followed by vfinx (S\&P 500 fund)
and vbltx (long-term US bond fund). The Sharpe ratios for the other
stock funds are much smaller than one. 

The estimated return correlation matrix of the six funds is illustrated
in @fig-PortMatrixVanguardCorrelations, created with
the **corrplot** function `corrplot.mixed()`:

```{r}
#| label: fig-PortMatrixVanguardCorrelations
#| echo: true
#| fig-cap: Estimated return correlations between six Vanguard mutual funds
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

corrplot.mixed(cov2cor(covhat), upper = "ellipse")
```


The stock funds are all strongly positively correlated with each other,
whereas the bond funds are either negatively correlated or very weakly
positively correlated with the stock funds. Hence, combining the stock
and bond funds in a portfolio should be beneficial in terms of risk
diversification. 


We use the **introcompfinr** portfolio functions to compute estimates
of the global minimum variance portfolio, the tangency portfolio,
and the efficient frontier of risky assets:

```{r}
#| echo: true

gmin.port = globalmin_portfolio(muhat, covhat)
tan.port = tangency_portfolio(muhat, covhat, r.f)
ef = efficient_frontier(muhat, covhat, alpha_min=-0.5,                          
                        alpha_max=1.5, nport=20)
```

The estimated global minimum variance portfolio is:

```{r}
#| echo: true

gmin.port
```

This portfolio is mostly concentrated in the short-term US bond fund
vbisx (which has the smallest volatility) and has small short positions
in veurx, veiex and vbltx. This portfolio is not feasible, however,
because investors are not allowed to short sell mutual funds. 


The estimated tangency portfolio is:

```{r}
#| echo: true

tan.port
```

This portfolio is also an infeasible long-short portfolio with long
positions in the US funds vfinx, vbltx and vbisx and short positions
in the non-US stock funds veurx, veiex and vpacx. 


The mean-variance efficient frontiers are illustrated in @fig-PortMatrixEfficientVanguard,
created using:

```{r}
#| label: fig-PortMatrixEfficientVanguard
#| echo: true
#| fig-cap: Mean variance efficient portfolios of the Vanguard mutual funds. The risky
#|   asset portfolio frontier is shown in blue. The efficient portfolios of the risk-free
#|   asset and the risky assets is shown in green. The global minimum variance portfolio
#|   is labeled gmin, and the tangency portfolio is labeled tan.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(ef, plot.assets=TRUE, col="blue", pch=16) 
text(gmin.port$sd_port, gmin.port$er_port,labels = "gmin", pos = 2) 
text(tan.port$sd_port, tan.port$er_port,labels = "tan", pos = 3) 
sr.tan = (tan.port$er_port - r.f)/tan.port$sd_port 
text(0, r.f, labels=expression(r[f]), pos = 4)
abline(a=r.f, b=sr.tan, col="green", lwd=2) 
abline(h=0, v=0)
```


The portfolio frontier of risky assets is shown in blue. These portfolios
are infeasible because they all contain short sales in one or more
of the mutual funds:

```{r}
#| echo: true

ef$weight_mat
```

The locations of vbisx and the global minimum variance portfolio are
almost identical, which suggests that vbisx is a good feasible substitute
for the infeasible global minimum variance portfolio. The remaining
funds are well inside the risky asset efficient frontier. The set
of efficient portfolios of risky assets plus the risk-free asset is
shown in green. These are portfolios of the risk-free asset and the
tangency portfolio. Since the tangency portfolio contains short sales,
these portfolios are not feasible. However, the tangency portfolio
is approximately 70\% vbisx and 30\% vfinx so a close substitute for
the efficient portfolios are combinations of the risk-free asset and
the approximate tangency portfolio that is 70\% vbisx and 30\% vfinx.


This example highlights some practical problems associated with applying
mean-variance portfolio theory in a real-world asset allocation context.
In particular, mean-variance efficient portfolios often contain short
sales in some assets. When the underlying assets are mutual funds,
investors are not allowed to directly short these assets and it may
not be possible to exactly implement the theory. The next chapter
shows how mean-variance portfolio theory can be modified to allow
for additional constraints created by short sales restrictions.

## Further Reading: Portfolio Theory with Matrix Algebra

Portfolio theory using matrix algebra is discussed in advanced textbooks
on asset pricing and portfolio theory. Good treatments are given in
[@ingersoll-1987], [@clm-1997], and [@cochrane-2008]. A nice review of portfolio theory with matrix algebra is given
in [@cm-1995]. In R, mean-variance optimized
portfolios can be computed using the packages **fPortfolio**,
**PARMA**, **PortfolioAnalytics**, and **stockPortfolio**.

## Exercises
::: {#exr-solve-for-global-minimum-variance-portfolio .exercise name="Solve for global minimum variance portfolio"}
The global minimum variance portfolio $\mathbf{m}=(m_{1},m_{2},m_{3})^{\prime}$
for the $3$ asset case solves the constrained minimization problem:

\begin{align*}
\min_{m_{1},m_{2},m_{3}}~\sigma_{p,m}^{2}&=m_{1}^{2}\sigma_{1}^{2}+m_{2}^{2}\sigma_{2}^{2}+m_{2}^{2}\sigma_{2}^{2}\\
&+2m_{1}m_{2}\sigma_{12}+2m_{1}m_{3}\sigma_{13}+2m_{2}m_{3}\sigma_{23}\\
&\textrm{s.t. }m_{1}+m_{2}+m_{3}=1.
\end{align*}
The Lagrangian for this problem is: 
\begin{align*}
L(m_{1},m_{2},m_{3},\lambda) & =m_{A}^{2}\sigma_{A}^{2}+m_{B}^{2}\sigma_{B}^{2}+m_{C}^{2}\sigma_{C}^{2}\\
 & +2m_{A}m_{B}\sigma_{AB}+2m_{A}m_{C}\sigma_{AC}+2m_{B}m_{C}\sigma_{BC}\\
 & +\lambda(m_{A}+m_{B}+m_{C}-1)
\end{align*}

(a) Write out the first order conditions 
:::


::: {#exr-solve-for-global-minimum-variance-portfolio .exercise name="Solve for global minimum variance portfolio"}

* Use 2 x 2 matrix inversion rule to explicitly solve for

:::



::: {#exr-solve-for-tangency-portfolio .exercise name="Solve for tangency portfolio"}

* Give explicit formula for tangency portfolio in 2 x 2 case

:::
