# Portfolio Theory with Short Sales Constraints {#sec-Portfolio-Theory-No-Shorts}

Updated: September 12, 2026

Copyright © Eric Zivot 2026

The portfolio theory I described in Chapters @sec-Introduction-to-Portfolio-Theory
and @sec-Portfolio-Theory-with-Matrix-Algebra assumed that
investors can freely short sell all assets. In many situation, however,
investors may not be allowed to short sell some or all assets and/or
there may be restrictions on when short sales may occur. In particular,
certain types of assets (e.g., mutual funds) cannot be sold short.
This has implications for asset allocation in individual retirement
accounts. Some institutions, such as pension funds, are prevented
from directly short selling assets. Government regulations may prevent
certain assets from being sold short, or prevent short sales from
occurring at certain times. For example, during the 2008 financial
crisis the Securities and Exchange Commission temporarily blocked
the short sale of financial stocks. Because of these practical limitations
on short sales, it is important to be able to incorporate short sales
restrictions into mean-variance portfolio theory in practice and in this chapter
I describe how to do it.

This chapter is organized as follows. In @sec-Portfolio-Theory-with I
describe the impact of short sales constraints, on the investment in risky assets,
on efficient portfolios in the simplified two risky asset setting presented
in @sec-Introduction-to-Portfolio-Theory. In @sec-Portfolio-Theory-with-1
I discuss the computation of mean-variance efficient portfolios when
there are short sales constraints in the general case of $N$ risky
assets, as described in @sec-Portfolio-Theory-with-Matrix-Algebra, using quadratic programming.
In @sec-Application-to-Vanguard-Mutual-Funds I give a real-world application of the theory to asset
allocation among Vanguard mutual funds. 


The examples I cover in this chapter use the **introcompfinr, PerformanceAnalytics**,
and **quadprog** packages. Make sure these packages are installed
and loaded in R before replicating the chapter examples.


```{r}
#| include: false

suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(PerformanceAnalytics))
suppressPackageStartupMessages(library(quadprog))
options(digits=3)
```

## Overview of Short Selling

Most investments involve purchases of assets that you believe will
increase in value over time. When you buy an asset you establish a
*long* position in the asset. You profit from the long position
by selling the asset in the future at a price higher than the original
purchase price. However, sometimes you may believe that an asset will
decrease in value over time. To profit on an investment in which the
asset price is lower in the future requires *short selling* the asset.
In general, short selling involves selling an asset you borrow from
someone else today and repurchasing the asset in the future and returning
it to its owner. You profit on the short sale if you repurchase the
asset for a lower price than you initially sold it. Because short
selling involves borrowing it is a type of leveraged investment and
so risk increases with the size of the short position. 


In practice, a short sale of an asset works as follows. Suppose you
have an investment account at some brokerage firm and you are interested
in short selling one share of an individual stock, such as Microsoft,
that you do not currently own. Because the brokerage firm has so many
clients, one of the clients will own Microsoft. The brokerage firm
will allow you to borrow Microsoft, sell it and add the proceeds of
the short sale to your investment account. Doing so establishes a
*short position* in Microsoft which is a liability to return the
borrowed share sometime in the future. You close the short position
in the future by repurchasing one share of Microsoft through the brokerage
firm who returns the share to the original owner. However, borrowing
is not free. You will have to pay interest on the amount borrowed.
You will also need sufficient capital in your account to cover any
potential losses from the short sale. In addition, if the price of
the short sold asset rises substantially during the short sale period
then the brokerage firm may require you to add additional margin capital
to your account. If you cannot do this then the brokerage firm will
force you to close the short position immediately.

## Portfolio Theory with Short Sales Constraints in a Simplified Setting {#sec-Portfolio-Theory-with}
\index{Portfolio with short sales}

In this section, we illustrate the impact of short sales constraints
on risky assets on portfolios in the simplified setting described
in Chapter @sec-Introduction-to-Portfolio-Theory. We first
look at simple portfolios of two risky assets, then consider portfolios
of a risky asset plus a risk-free asset, and finish with portfolios
of two risky assets plus a risk-free asset. 

### Two Risky Assets 
\index{Two risky assets}
Consider the two risky asset example data from Chapter @sec-Introduction-to-Portfolio-Theory:

```{r}
#| echo: true

mu.A = 0.175 
sig.A = 0.258 
sig2.A = sig.A^2 
mu.B = 0.055 
sig.B = 0.115 
sig2.B = sig.B^2 
rho.AB = -0.164 
sig.AB = rho.AB*sig.A*sig.B
assetNames = c("Asset A", "Asset B") 
mu.vals = c(mu.A, mu.B) 
names(mu.vals) = assetNames 
Sigma = matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), 2, 2) 
colnames(Sigma) = rownames(Sigma) = assetNames
```

Asset A is the high average return and high risk asset, and asset
B is the low average return and low risk asset. Consider the following
set of 19 portfolios: 

```{r}
#| echo: true

x.A = seq(from=-0.4, to=1.4, by=0.1) 
x.B = 1 - x.A 
n.port = length(x.A) 
names(x.A) = names(x.B) = paste("p", 1:n.port, sep=".") 
mu.p = x.A*mu.A + x.B*mu.B 
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB 
sig.p = sqrt(sig2.p)
```

Figure @fig-NoShorts2assetFrontier shows the risky asset
only portfolio frontier created using:

```{r}
#| label: fig-NoShorts2assetFrontier
#| echo: true
#| fig-cap: Two-risky asset portfolio frontier. Long-only portfolios are in black (between
#|   Asset B and Asset A), and long-short portfolios are in blue (below Asset B and above
#|   Asset A).
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

cex.val = 1.5 
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,      
ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)),      
xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab=cex.val,      
col=c(rep("lightblue", 4), rep("black", 11), rep("lightblue", 4))) 
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val) 
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
```

The long only (no short sale) portfolios, shown in black dots between
the points labeled "Asset B" and "Asset A", are portfolios
5 to 15. The asset weights on these portfolios are:

```{r}
#| echo: true

longOnly = rbind(x.A[5:15], x.B[5:15]) 
rownames(longOnly) = assetNames 
longOnly
```

The long-short portfolios, shown in blue dots below the point labeled
"Asset B" and above the point labeled "Asset A", are portfolios
1 to 4 and portfolios 16 to 19: 

```{r}
#| echo: true

longShort = rbind(x.A[c(1:4, 16:19)], x.B[c(1:4, 16:19)]) 
rownames(longShort) = assetNames 
longShort
```

Not allowing short sales for asset A eliminates the inefficient portfolios
1 to 4 (blue dots below the label "Asset B" ),
and not allowing short sales for asset B eliminates the efficient
portfolios 16 to 19 (blue dots above the label "Asset
A" ). We make the following remarks:

-  Not allowing short sales in both assets limits the set of feasible portfolios.
-  It is not possible to invest in a portfolio with a higher expected return than asset A.
-  It is not possible to invest in a portfolio with a lower expected return than asset B.
 

#### Short sales constraints and the global minimum variance portfolio 

Recall, the two risky asset global minimum variance portfolio solves
the unconstrained optimization problem:
\begin{align*}
&\underset{m_{A},m_{B}}{\min}\sigma_{p}^{2}  =m_{A}^{2}\sigma_{A}^{2}+m_{B}^{2}\sigma_{B}^{2}+2m_{A}m_{B}\sigma_{AB}\\
&s.t.~m_{A}+m_{B} =1.
\end{align*}

From Chapter @sec-Introduction-to-Portfolio-Theory, the two
risky asset unconstrained global minimum variance portfolio weights
are: 

$$
m_{A}=\frac{\sigma_{B}^{2}-\sigma_{AB}}{\sigma_{A}^{2}+\sigma_{B}^{2}-2\sigma_{AB}},~m_{B}=1-m_{A}.
$$ {#eq-NoShortsMinVarPort}

For the example data, the unconstrained global minimum variance portfolio
weights are both positive:

```{r}
#| echo: true

gmin.port = globalmin_portfolio(mu.vals, Sigma) 
gmin.port
```

This means that if we disallow short-sales in both assets we still
get the same global minimum variance portfolio weights. In this case,
we say that the no short sales constraints are *not binding*
(have no effect) on the solution for the unconstrained optimization
problem @eq-NoShortsMinVarPort. However, it is possible for
one of the weights in the global minimum variance portfolio to be
negative if correlation between the returns on assets A and B is sufficiently
positive. To see this, the numerator in the formula for the global
minimum variance weight for asset A can be re-written as:^[By the Cauchy-Schwarz inequality the denominator for $m_{A}$ in @eq-NoShortsMinVarPort
is always positive. ]

$$
\sigma_{B}^{2}-\sigma_{AB}=\sigma_{B}^{2}-\rho_{AB}\sigma_{A}\sigma_{B}=\sigma_{B}^{2}\left(1-\rho_{AB}\frac{\sigma_{A}}{\sigma_{B}}\right).
$$

Hence, the weight in asset A in the global minimum variance portfolio
will be negative if:

\begin{eqnarray*}
1-\rho_{AB}\frac{\sigma_{A}}{\sigma_{B}} & <0 & \Rightarrow\rho_{AB}>\frac{\sigma_{B}}{\sigma_{A}}.
\end{eqnarray*}

For the example data, this cut-off value for the correlation is:

```{r}
#| echo: true

sig.B/sig.A
```

Figure @fig-NoShorts2assetFrontier2 shows the two risky
asset portfolio frontier computed with $\rho_{AB}=0.8$. Here, the
global minimum variance portfolio is: 

```{r}
#| echo: true

rho.AB = 0.8 
sig.AB = rho.AB*sig.A*sig.B 
Sigma = matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), 2, 2) 
gmin.port = globalmin_portfolio(mu.vals, Sigma) 
gmin.port
```

and contains a negative weight in asset A. This portfolio is shown
as the green dot in Figure @fig-NoShorts2assetFrontier2,
and plots below the point labeled "Asset B" .
Here, the no-short sales constraint on asset A is a *binding
constraint*: it is not possible to invest in the global minimum variance
portfolio without shorting asset A. Visually, we can see that the
feasible long-only portfolio with the smallest variance is 100\% invested
in asset B. This is the *no short sales constrained* global minimum
variance portfolio. Notice that the variance of this portfolio is
larger than the variance of the global minimum variance portfolio.
This is the cost, in terms of variance, of imposing the no-short sales
constraint. 


```{r}
#| label: fig-NoShorts2assetFrontier2
#| echo: false
#| fig-cap: Two risky asset portfolio frontier with short sale in Asset A in the global
#|   minimum variance portfolio.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

rho.AB = 0.8 
sig.AB = rho.AB*sig.A*sig.B 
# create portfolios for frontier plot 
x.A = seq(from=-0.4, to=1.4, by=0.1) 
x.B = 1 - x.A 
n.port = length(x.A) 
names(x.A) = names(x.B) = paste("p", 1:n.port, sep=".") 
mu.p = x.A*mu.A + x.B*mu.B 
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB 
sig.p = sqrt(sig2.p)
# plot frontier
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,      
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab=cex.val,      
     col=c("lightblue", "green", rep("lightblue", 2), rep("black", 11),             
     rep("lightblue", 4))) 
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val) 
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val) 
text(gmin.port$sd, gmin.port$er, labels="Global Min", pos = 4,       
     cex = cex.val)
```


### One risky asset and a risk-free asset 

Now consider portfolios of a risky asset (e.g., asset A) and a risk-free
asset (e.g., U.S. T-Bill) with risk-free rate $r_{f}>0$. Let $x$
denote the portfolio weight in the risky asset. Then $1-x$ gives
the weight in the risk-free asset. From Chapter @sec-Introduction-to-Portfolio-Theory,
the expected return and volatility of portfolios of one risky asset
and one risk-free asset are:
\begin{eqnarray*}
\mu_{p} & = & r_{f}+x(\mu-r_{f}),\\
\sigma_{p} & = & |x|\sigma,
\end{eqnarray*}
where $\mu$ and $\sigma$ are the expected return and volatility
of the risky asset, respectively. For example, Figure @fig-NoShortsRiskyTbill
shows 29 portfolios of asset A and the risk-free asset with $r_{f}=0.03$
created with:


```{r}
#| label: fig-NoShortsRiskyTbill
#| echo: true
#| fig-cap: Return-risk characteristics of portfolios of asset A and a risk-free asset
#|   with $r_{f}=0.03$. Long-only portfolios are in black (between $\mathrm{r}_{f}$ and
#|   Asset A), and long-short portfolios are in blue (below $\mathrm{r}_{f}$ and above
#|   Asset A).
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

r.f = 0.03 
x.A = seq(from=-1.4, to=1.4, by=0.1) 
mu.p.A = r.f + x.A*(mu.A - r.f) 
sig.p.A = abs(x.A)*sig.A 
plot(sig.p.A, mu.p.A, type="b", ylim=c(min(mu.p.A), max(mu.p.A)),      
xlim=c(-0.01, max(sig.p.A)), pch=16, cex = cex.val,      
xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab = cex.val,      
col=c(rep("lightblue", 14), rep("black", 11), rep("lightblue", 4))) 
abline(h=0) 
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val) 
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = cex.val)
```

The portfolios (blue dots) above the point labeled "Asset A" are
short the risk-free asset and long more than 100\% of asset A. The
portfolios (blue dots) below $r_{f}=0.03$ are short the risky asset
and long more than 100\% in the risk-free asset. 


Here, we assume that the short sales constraint only applies to the
risky asset and not the risk-free asset. That is, we assume that we
can borrow and lend at the risk-free rate $r_{f}$ without constraints.
In this situation, the no short sales constraint on the risky asset
eliminates the inefficient portfolios (blue dots) below the risk-free
rate $r_{f}$.

### Two risky assets and risk-free asset

As discussed in Chapter @sec-Introduction-to-Portfolio-Theory,
in the case of two risky assets (with no short sales restrictions)
plus a risk-free asset the set of efficient portfolios is a combination
of the tangency portfolio (i.e., the maximum Sharpe ratio portfolio
of two risky assets) and the risk-free asset. For the example data
with $\rho_{AB}=-0.164$ and $r_{f}=0.03$, the tangency portfolio
is:

```{r}
#| echo: true

rho.AB = -0.164 
sig.AB = rho.AB*sig.A*sig.B 
Sigma = matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), 2, 2) 
r.f = 0.03 
tan.port = tangency_portfolio(mu.vals, Sigma, r.f) 
summary(tan.port, r.f)
```

No asset is sold short in the unconstrained tangency portfolio, and
so this is also the short-sales constrained tangency portfolio. The
set of efficient portfolios is illustrated in Figure @fig-NoShortsTangency,
created with:

```{r}
#| label: fig-NoShortsTangency
#| echo: true
#| fig-cap: 'Efficient portfolios of two risky assets plus a risk-free asset: $\rho_{AB}=-0.164$
#|   and $r_{f}=0.03$. No short sales in unconstrained tangency portfolio.'
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# risky asset only portfolios
x.A = seq(from=0, to=1, by=0.1) 
x.B = 1 - x.A 
mu.p = x.A*mu.A + x.B*mu.B 
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB 
sig.p = sqrt(sig2.p) 
# T-bills plus tangency 
x.tan = seq(from=0, to=2.4, by=0.1) 
mu.p.tan.tbill = r.f + x.tan*(tan.port$er_port - r.f) 
sig.p.tan.tbill = x.tan*tan.port$sd_port
# global minimum variance portfolio
gmin.port = globalmin_portfolio(mu.vals, Sigma)
# plot efficient portfolios 
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,      
     ylim=c(0, 0.20), xlim=c(-0.01, 0.35),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab = cex.val) 
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val) 
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val) 
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = cex.val) 
text(x=tan.port$sd_port, y=tan.port$er_port, labels="Tangency", pos=2, cex = cex.val) 
text(gmin.port$sd_port, gmin.port$er_port, labels="Global Min", pos = 4,       
     cex = cex.val)
points(sig.p.tan.tbill, mu.p.tan.tbill, type="l", col="green",         
       lwd=2, cex = cex.val)
```

Figure @fig-NoShortsTangency shows that when the tangency
portfolio is located between the global minimum variance portfolio
(tip of the Markowitz bullet) and asset A, there will be no short
sales in the tangency portfolio. However, if the correlation between
assets A and B is sufficiently positive (and the risk-free rate is
adjusted so that it is less than the expected return on the global
minimum variance portfolio) then asset B can be sold short in the
unconstrained tangency portfolio. For example, if $\rho_{AB}=0.7$
and $r_{f}=0.01$ the unconstrained tangency portfolio becomes:

```{r}
#| echo: true

rho.AB = 0.7 
sig.AB = rho.AB*sig.A*sig.B 
Sigma = matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), 2, 2) 
r.f = 0.01 
tan.port = tangency_portfolio(mu.vals, Sigma, r.f) 
summary(tan.port, r.f)
```

Now, asset B is sold short in the tangency portfolio. The set of efficient
portfolios in this case is illustrated in Figure @fig-NoShortsTangency2.
Notice that the tangency portfolio is located on the set of risky
asset only portfolios above the point labeled "Asset A", which
indicates that asset B is sold short and more than 100\% is invested
in asset A.^[Also notice that when $\rho_{AB}=0.7$ the expected return on the
global minimum variance drops (frontier of risky assets is shifted
down) below $0.03$ so we need to reduce $r_{f}$ to $0.01$ to ensure
that the tangency portfolio has a positive slope. ]

```{r}
#| label: fig-NoShortsTangency2
#| echo: false
#| fig-cap: 'Efficient portfolios of two risky assets plus a risk free asset: $\rho_{AB}=0.7$
#|   and $r_{f}=0.01$. Asset B sold short in unconstrained tangency portfolio. The short
#|   sale constrained tangency portfolio is the point labeled "Asset A". '
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

x.A = seq(from=-0.4, to=1.8, by=0.1) 
x.B = 1 - x.A 
mu.p = x.A*mu.A + x.B*mu.B 
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB 
sig.p = sqrt(sig2.p)
# T-bills plus tangency 
x.tan = seq(from=0, to=2.4, by=0.1) 
mu.p.tan.tbill = r.f + x.tan*(tan.port$er_port - r.f) 
sig.p.tan.tbill = x.tan*tan.port$sd_port
# global minimum variance portfolio
gmin.port = globalmin_portfolio(mu.vals, Sigma)
# plot portfolios
plot(sig.p, mu.p, type="b", pch=16, cex = cex.val,      
     ylim=c(0, max(mu.p)), xlim=c(-0.01, max(sig.p)),      
     col=c(rep("lightblue",4), rep("black", 11), rep("lightblue", 8)),      
xlab=expression(sigma[p]), ylab=expression(mu[p]), cex.lab = cex.val) 
text(x=sig.A, y=mu.A, labels="Asset A", pos=4, cex = cex.val) 
text(x=sig.B, y=mu.B, labels="Asset B", pos=4, cex = cex.val)
text(gmin.port$sd_port, gmin.port$er_port, labels="Global Min", pos = 4,       
     cex = cex.val)
text(x=0, y=r.f, labels=expression(r[f]), pos=2, cex = cex.val) 
text(x=tan.port$sd_port, y=tan.port$er_port, labels="Tangency", pos=2, cex = cex.val) 
points(sig.p.tan.tbill, mu.p.tan.tbill, type="l", col="green",         
       lwd=2, cex = cex.val)
```


When shorts sales of risky assets are not allowed, the unconstrained
tangency portfolio is infeasible. In this case, the constrained tangency
portfolio is not the tangency portfolio but is 100\% invested in asset
A. This portfolio is located at the tangency point of a straight line
drawn from the risk-free rate to the long-only risky asset frontier.
Here, the cost of the no-short sales constraint is the reduction in
the Sharpe ratio of the tangency portfolio. The Sharpe ratio of the
constrained tangency portfolio is the Sharpe ratio of asset A:

```{r}
#| echo: true

(mu.A - r.f)/sig.A
```

Hence, the cost of the no short sales constraint in this example is
very small.

## Portfolio Theory with Short Sales Constraints in a General Setting {#sec-Portfolio-Theory-with-1}
\index{Portfolio with short sales}
In this section we extend the analysis of the previous section to
the more general setting of $N$ risky assets plus a risk free asset. 

### Multiple risky assets
\index{Multiple risky assets}

::: {#exm-random-portfolios-of-microsoft-nordstrom-and-starbucks-with-no-short-sales .example name="Random portfolios of Microsoft Nordstrom and Starbucks with no-short sales"}

The example data on Microsoft, Nordstrom and Starbucks from Chapter
@sec-Portfolio-Theory-with-Matrix-Algebra is used to illustrate
the computation of mean-variance efficient portfolios with short sales
constraints. These data are created using:

```{r}
#| echo: true

asset.names <- c("MSFT", "NORD", "SBUX")
mu.vec = c(0.0427, 0.0015, 0.0285)
names(mu.vec) = asset.names
sigma.mat = matrix(c(0.0100, 0.0018, 0.0011,
		   0.0018, 0.0109, 0.0026,
		   0.0011, 0.0026, 0.0199),
		 nrow=3, ncol=3)
dimnames(sigma.mat) = list(asset.names, asset.names)
sd.vec = sqrt(diag(sigma.mat))
r.f=0.005
```

The impact of the no short sales restrictions on the risk-return characteristics
of feasible portfolios is nicely illustrated by constructing random
long-only portfolios of these three assets:

```{r}
#| echo: true

set.seed(123) 
x.msft = runif(400, min=0, max=1) 
x.nord = runif(400, min=0, max=1) 
x.sbux = 1 - x.msft - x.nord 
long.only = which(x.sbux > 0) 
x.msft = x.msft[long.only] 
x.nord = x.nord[long.only] 
x.sbux = x.sbux[long.only]
length(long.only)
```

Here we first create 400 random weights for Microsoft and Nordstrom
that lie between zero and one, respectively. Then, we determine the
weight on Starbucks and throw out portfolios for which the weight
on Starbucks is negative. The remaining 191 portfolios are then long-only
portfolios. To be sure, these weights are illustrated in Figure @fig-NoShortsRandomWeights
created using:

```{r}
#| label: fig-NoShortsRandomWeights
#| echo: true
#| fig-cap: Weights on 191 random long-only portfolios of Microsoft, Nordstrom and Starbucks.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

chart.StackedBar(cbind(x.msft, x.nord, x.sbux),                   
                 main="Random long-only portfolio weight vectors",
                 xlab = "portfolio", ylab = "weights",
                 xaxis.labels=as.character(1:length(long.only)))
```


```{r}
#| label: fig-NoShortsRandomPortfolios
#| echo: true
#| fig-cap: Risk-return characteristics of 191 random long-only portfolios of Microsoft,
#|   Nordstrom and Starbucks.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sd.vec, mu.vec,  ylim=c(0, 0.05), xlim=c(0, 0.2), 
     ylab=expression(mu[p]),
     xlab=expression(sigma[p]), type="n", cex.lab=1.5)      
for (i in 1:length(long.only)) {   
   z.vec = c(x.msft[i], x.nord[i], x.sbux[i])   
   mu.p = crossprod(z.vec,mu.vec)   
   sig.p = sqrt(t(z.vec)%*%sigma.mat%*%z.vec)   
   points(sig.p, mu.p, pch=16, col="grey", cex=1.5) 
} 
points(sd.vec, mu.vec, pch=16, col="black", cex=2.5, cex.lab=1.75) 
text(sd.vec, mu.vec, labels=asset.names, pos=4, cex = cex.val)
```

The shape of the risk-return scatter is revealing. No random long-only
portfolio has an expected return higher than the expected return on
Microsoft, which is the asset with the highest expected return. Similarly,
no random portfolio has an expected return lower than the expected
return on Nordstrom - which is the asset with the lowest expected
return. Finally, no random portfolio has a volatility higher than
the volatility on Starbucks, the asset with the highest volatility.
The scatter of points has a convex (to the origin) parabolic-shaped
outer boundary with endpoints at Microsoft and Nordstrom, respectively.
The points inside the boundary taper outward from Starbucks towards
the outer boundary. The entire risk-return scatter resembles an umbrella
tipped on its side with spires extending from the outer boundary to
the individual assets.
\hfill{}$\blacksquare$
:::

### Portfolio optimization imposing no-short sales constraints
\index{Portfolio with no-short sales}
Consider an investment universe with $N$ risky assets whose simple
returns are described by the GWN model with expected return vector
$\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$. The optimization
problem to solve for the global minimum variance portfolio $\mathbf{m}$
with no-short sales constraints on all assets is a straightforward
extension of the problem that allows short sales. We simply add the
$N$ no-short sales inequality constraints $m_{i}\geq0\,(i=1,\ldots,N)$.
The optimization problem is: 
$$
\begin{aligned}
\min_{\mathbf{m}}\sigma_{p,m}^{2} & =\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}\\
\text{s.t. } \mathbf{m}^{\prime}\mathbf{1} & =1,\nonumber \\
m_{i} & \geq0\text{ }(i=1,\ldots,N).\nonumber 
\end{aligned}
$$ {#eq-NoShortsGlobalMin}
Here, there is one equality constraint, $\mathbf{m}^{\prime}\mathbf{1}=1$,
and $N$ inequality constraints, $m_{i}\geq0\text{ }(i=1,\ldots,N)$.


Similarly, the optimization problem to solve for a minimum variance
portfolio $\mathbf{x}$ with target expected return $\mu_{p}^{0}$ and
no short sales allowed for any asset becomes:
$$
\begin{aligned}
\min_{\mathbf{x}}\sigma_{p,x}^{2} & =\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x} \\
\text{s.t. }\mu_{p,x} & =\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p}^{0}, \\
\mathbf{x}^{\prime}\mathbf{1} & =1, \\
x_{i} & \geq0\text{ }(i=1,\ldots,N).
\end{aligned}
$$ {#eq-NoShortsTargetReturn}
Here there are two equality constraints, $\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p}^{0}$
and $\mathbf{x}^{\prime}\mathbf{1}=1$, and $N$ inequality constraints,
$x_{i}\geq0\text{ }(i=1,\ldots,N)$


Remarks:

1.  The optimization problems @eq-NoShortsGlobalMin and @eq-NoShortsTargetReturn, unfortunately, cannot be analytically solved using the method of Lagrange multipliers due to the inequality constraints. They must be solved numerically using an optimization algorithm. Fortunately, the optimization problems @eq-NoShortsGlobalMin and @eq-NoShortsTargetReturn are special cases of a more general *quadratic programming* problem for which there is a specialized optimization algorithm that is fast and efficient and available in R in the package **quadprog**.^[See the CRAN Optimization Task View for additional R packages that can be used to solve QP problems.]
2.  There may not be a feasible solution to @eq-NoShortsTargetReturn. That is, there may not exist a no-short sale portfolio $\mathbf{x}$ that reaches the target return $\mu_{p}^{0}.$ For example, in Figure @fig-NoShortsRandomPortfolios there is no efficient portfolio that has target expected return higher than the expected return on Microsoft. In general, with $N$ assets there is no feasible solution to @eq-NoShortsTargetReturn for target expected returns higher than the maximum expected return among the $N$ assets. If you try to solve @eq-NoShortsTargetReturn for an infeasible target expected return, the numerical optimizer will return an error message that indicates no solution is found. 
3.  The portfolio variances associated with the solutions to @eq-NoShortsGlobalMin and @eq-NoShortsTargetReturn, respectively, must be at least as large as the variances associated with the solutions that do not impose the short sales restrictions. Intuitively this makes sense. Adding additional constraints to a minimization problem will lead to a higher minimum at the solution if the additional constraints are binding. If the constraints are not binding, then the solution with and without the additional constraints are the same. 
4.  The entire frontier of minimum variance portfolios without short sales exists only for target expected returns between the expected return on the global minimum variance portfolio with no short sales that solves @eq-NoShortsGlobalMin and the target expected return equal to the maximum expected return among the $N$ assets. For these target returns, the no-short sales frontier will lie inside and to the right of the frontier that allows short sales if the no-short sales restrictions are binding for some asset.  
5.  The entire frontier of minimum variance portfolios without short sales can no longer be constructed from any two frontier portfolios. It has to be computed by brute force for each portfolio solving @eq-NoShortsTargetReturn with target expected return $\mu_{p}^{0}$ above the expected return on the the global minimum variance portfolio that solves @eq-NoShortsGlobalMin and below the target return equal to the maximum expected return among the $N$ assets.


## Quadratic Programming Problems {#sec-Quadradic-programming}
\index{Quadratic Programming}
In this Section, we show that the inequality constrained portfolio
optimization problems @eq-NoShortsGlobalMin and @eq-NoShortsTargetReturn
are special cases of more general quadratic programming problems and
we show how to use the function `solve.QP()` from the R package
**quadprog** to numerically solve these problems. 


*Quadratic programming*\index{Quadratic programming} (QP) problems are of the form:
$$ 
\min_{\mathbf{x}} ~ \frac{1}{2}\mathbf{x}^{\prime}\mathbf{Dx}-\mathbf{d}^{\prime}\mathbf{x}, ~ \text{s.t. }
$$ {#eq-NoShortsQPobjective}
$$
\mathbf{A}_{eq}^{\prime}\mathbf{x}  \geq\mathbf{b}_{eq}\text{ for }l\text{ equality constraints,}
$$ {#eq-NoShortsQPequality}
$$
\mathbf{A}_{neq}^{\prime}\mathbf{x} =\mathbf{b}_{neq}\text{ for }m\text{ inequality constraints},
$$ {#eq-NoShortsQPinequality}

where $\mathbf{D}$ is an $N\times N$ matrix, $\mathbf{x}$ and $\mathbf{d}$
are $N\times1$ vectors, $\mathbf{A}_{neq}^{\prime}$ is an $m\times N$
matrix, $\mathbf{b}_{neq}$ is an $m\times1$ vector, $\mathbf{A}_{eq}^{\prime}$
is an $l\times N$ matrix, and $\mathbf{b}_{eq}$ is an $l\times1$
vector. 


The `solve.QP()` function from the R package **quadprog**
can be used to solve QP functions of the form @eq-NoShortsQPobjective
- @eq-NoShortsQPinequality. The function `solve.QP()`
takes as input the matrices and vectors $\mathbf{D}$, $\mathbf{d}$,
$\mathbf{A}_{eq}$, $\mathbf{b}_{eq},$ $\mathbf{A}_{neq}$ and $\mathbf{b}_{neq}$:

```{r}
#| echo: true

library(quadprog)
args(solve.QP)
```

where the arguments `Dmat` and `dvec` correspond to
the matrix $\mathbf{D}$ and the vector $\mathbf{d}$, respectively.
The function `solve.QP()`, however, assumes that the equality
and inequality matrices ($\mathbf{A}_{eq},\,\mathbf{A}_{neq}$) and
vectors ($\mathbf{b}_{eq},\,\mathbf{b}_{neq}$) are combined into
a single $(l+m)\times N$ matrix $\mathbf{A}^{\prime}$ and a single
$(l+m)$$\times1$ vector $\mathbf{b}$ of the form:
$$
\mathbf{A}^{\prime}=\left[\begin{array}{c}
\mathbf{A}_{eq}^{\prime}\\
\mathbf{A}_{neq}^{\prime}
\end{array}\right],\,\mathbf{b}=\left(\begin{array}{c}
\mathbf{b}_{eq}\\
\mathbf{b}_{neq}
\end{array}\right).
$$
In `solve.QP()`, the argument `Amat` represents the
matrix $\mathbf{A}$ (not $\mathbf{A}^{\prime})$ and the argument
`bmat` represents the vector $\mathbf{b}$. The argument `meq`
determines the number of linear equality constraints (i.e., the number
of rows $l$ of $\mathbf{A}_{eq}^{\prime}$) so that $\mathbf{A}^{\prime}$
can be separated into $\mathbf{A}_{eq}^{\prime}$ and $\mathbf{A}_{neq}^{\prime}$,
respectively. 

### No short sales global minimum variance portfolio

The portfolio optimization problems @eq-NoShortsGlobalMin
and @eq-NoShortsTargetReturn are special cases of the general
QP problem @eq-NoShortsQPobjective - @eq-NoShortsQPinequality.
Consider first the problem to find the global minimum variance portfolio
@eq-NoShortsGlobalMin. The objective function $\sigma_{p,m}^{2}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}$
can be recovered from @eq-NoShortsQPobjective by setting $\mathbf{x}=\mathbf{m}$,
$\mathbf{D}=2\times\boldsymbol{\Sigma}$ and $\mathbf{d}=(0,\ldots,0)^{\prime}.$
Then,
$$
\frac{1}{2}\mathbf{x}^{\prime}\mathbf{Dx}-\mathbf{d}^{\prime}\mathbf{x}=\mathbf{m}^{\prime}\boldsymbol{\Sigma} \mathbf{m}=\sigma_{p,m}^{2}.
$$
Here, there is one equality constraint, $\mathbf{m}^{\prime}\mathbf{1}=1$,
and $N$ inequality constraints, $m_{i}\geq0\text{ }(i=1,\ldots,N)$,
which can be expressed compactly as $\mathbf{m}\geq\mathbf{0}$. These
constraints can be expressed in the form @eq-NoShortsQPequality
and @eq-NoShortsQPinequality by defining the restriction matrices:
\begin{align*}
\underset{(1\times N)}{\mathbf{A}_{eq}^{\prime}} & =\mathbf{1}^{\prime},\text{ }\underset{(1\times1)}{\mathbf{b}_{eq}}=1\\
\underset{(N\times N)}{\mathbf{A}_{neq}^{\prime}} & =\mathbf{I}_{N},\underset{(N\times1)}{\mathbf{b}_{neq}}=(0,\ldots,0)^{\prime}=\mathbf{0},
\end{align*}
so that 
$$
\mathbf{A}^{\prime}=\left[\begin{array}{c}
\mathbf{1}^{\prime}\\
\mathbf{I}_{N}
\end{array}\right],\,\mathbf{b}=\left(\begin{array}{c}
1\\
\mathbf{0}
\end{array}\right).
$$
Here, 
\begin{eqnarray*}
\mathbf{A}_{eq}^{\prime}\mathbf{m} & = & \mathbf{1^{\prime}m}=1=b_{eq},\\
\mathbf{A}_{neq}^{\prime}\mathbf{m} & = & \mathbf{I}_{N}\mathbf{m}=\mathbf{m}\geq\mathbf{0}=\mathbf{b}_{neq}.
\end{eqnarray*}

::: {#exm-global-minimum-variance-portfolio-with-no-short-sales-for-three-asset-example-data .example name="Global minimum variance portfolio with no short sales for three asset example data"}

The unconstrained global minimum variance portfolio of Microsoft,
Nordstrom and Starbucks is:

```{r}
#| echo: true

gmin.port = globalmin_portfolio(mu.vec, sigma.mat) 
gmin.port
```

This portfolio does not have any negative weights and so the no-short
sales restriction is not binding. Hence, the short sales constrained
global minimum variance is the same as the unconstrained global minimum
variance portfolio. 


The restriction matrices and vectors required by `solve.QP()`
to compute the short sales constrained global minimum variance portfolio
are:

```{r}
#| echo: true

D.mat = 2*sigma.mat 
d.vec = rep(0, 3) 
A.mat = cbind(rep(1,3), diag(3)) 
b.vec = c(1, rep(0,3))
D.mat
d.vec
t(A.mat)
b.vec
```

To find the short sales constrained global minimum variance portfolio,
call `solve.QP()` with the above inputs and set `meq=1`
to indicate one equality constraint:

```{r}
#| echo: true

qp.out = solve.QP(Dmat=D.mat, dvec=d.vec,                   
                  Amat=A.mat, bvec=b.vec, meq=1) 
class(qp.out)
names(qp.out)
```

The returned object, `qp.out`, is a list with the following
components:

```{r}
#| echo: true

qp.out
```

The portfolio weights are in the `solution` component, which
match the global minimum variance weights allowing short sales, and
satisfy the required constraints (weights sum to one and all weights
are positive). The minimized value of the objective function (portfolio
variance) is in the `value` component. See the help file for
`solve.QP()` for explanations of the other components. 


You can also use the **introcompfinr** function `globalmin_portfolio()`
to compute the global minimum variance portfolio subject to short-sales
constraints by specifying the optional argument `shorts=FALSE`:

```{r}
#| echo: true

gmin.port = globalmin_portfolio(mu.vec, sigma.mat,                                 
                                shorts=FALSE)
gmin.port
```

When `shorts=FALSE`, `globalmin_portfolio()` uses `solve.QP()` as described above to do the optimization. 
\hfill{}$\blacksquare$
:::

### No short sales minimum variance portfolio with target expected return

Next, consider the problem @eq-NoShortsTargetReturn to find
a minimum variance portfolio with a given target expected return.
The objective function @eq-NoShortsQPobjective has $\mathbf{D}=2\times\boldsymbol{\Sigma}$
and $\mathbf{d}=(0,\ldots,0)^{\prime}.$ The two linear equality constraints,
$\mathbf{x}^{\prime}\boldsymbol{\mu}=\mu_{p}^{0}$ and $\mathbf{x}^{\prime}\mathbf{1}=1$,
and $N$ inequality constraints $\mathbf{x}\geq\mathbf{0}$ have restriction
matrices and vectors:
\begin{align*}
\underset{(2\times N)}{\mathbf{A}_{eq}^{\prime}} & =\left(\begin{array}{c}
\boldsymbol{\mu}^{\prime}\\
\mathbf{1}^{\prime}
\end{array}\right),\,\underset{(2\times1)}{\mathbf{b}_{eq}}=\left(\begin{array}{c}
\mu_{p}^{0}\\
1
\end{array}\right).\text{ }\\
\underset{(N\times N)}{\mathbf{A}_{neq}^{\prime}} & =\mathbf{I}_{N},\,\underset{(N\times1)}{\mathbf{b}_{neq}}=(0,\ldots,0)^{\prime}.
\end{align*}
so that 
$$
\mathbf{A}^{\prime}=\left(\begin{array}{c}
\boldsymbol{\mu}^{\prime}\\
\mathbf{1}^{\prime}\\
\mathbf{I}_{N}
\end{array}\right),\text{ }\mathbf{b}=\left(\begin{array}{c}
\mu_{p}^{0}\\
1\\
\mathbf{0}
\end{array}\right).
$$
Here, 
\begin{eqnarray*}
\mathbf{A}_{eq}^{\prime}\mathbf{x} & = & \left(\begin{array}{c}
\boldsymbol{\mu}^{\prime}\\
\mathbf{1}^{\prime}
\end{array}\right)\mathbf{x=}\left(\begin{array}{c}
\boldsymbol{\mu}^{\prime}\mathbf{x}\\
\mathbf{1}^{\prime}\mathbf{x}
\end{array}\right)=\left(\begin{array}{c}
\mu_{p}^{0}\\
1
\end{array}\right)=\mathbf{b}_{eq},\\
\mathbf{A}_{neq}^{\prime}\mathbf{x} & = & \mathbf{I}_{N}\mathbf{x}=\mathbf{x}\geq0.
\end{eqnarray*}

::: {#exm-minimum-variance-portfolio-with-target-expected-return-and-no-short-sales-for-three-asset-example-data .example name="Minimum variance portfolio with target expected return and no short sales for three asset example data"}

Now, we consider finding the minimum variance portfolio that has the
same mean as Microsoft but does not allow short sales. First, we find
the minimum variance portfolio allowing for short sales using the
**introcompfinr** function `efficient_portfolio()`:

```{r}
#| echo: true

eMsft.port = efficient_portfolio(mu.vec, sigma.mat, 
                                 target_return = mu.vec["MSFT"])
eMsft.port
```

This portfolio has a short sale (negative weight) in Nordstrom. Hence,
when the short sales restriction is imposed it will be binding. Next,
we set up the restriction matrices and vectors required by `solve.QP()`
to compute the minimum variance portfolio with no-short sales:

```{r}
#| echo: true

D.mat = 2*sigma.mat 
d.vec = rep(0, 3) 
A.mat = cbind(mu.vec, rep(1,3), diag(3)) 
b.vec = c(mu.vec["MSFT"], 1, rep(0,3))
t(A.mat)
b.vec
```

Then we call `solve.QP()` with `meq=2` to indicate two equality constraints:

```{r}
#| echo: true

qp.out = solve.QP(Dmat=D.mat, dvec=d.vec,                   
                  Amat=A.mat, bvec=b.vec, meq=2) 
names(qp.out$solution) = names(mu.vec)
round(qp.out$solution, digits=3)
```

With short sales not allowed, the minimum variance portfolio with
the same mean as Microsoft is 100\% invested in Microsoft. This is
consistent with what we see in Figure @fig-NoShortsRandomPortfolios.
The volatility of this portfolio is slightly higher than the volatility
of the portfolio that allows short sales:

```{r}
#| echo: true

sqrt(qp.out$value)
eMsft.port$sd
```

You can also use the **introcompfinr** function `efficient_portfolio()`
to compute a minimum variance portfolio with target expected return
subject to short-sales constraints by specifying the optional argument
`shorts=FALSE`:

```{r}
#| echo: true

efficient_portfolio(mu.vec, sigma.mat, 
                    target_return = mu.vec["MSFT"], shorts = FALSE)
```

Suppose you try to find a minimum variance portfolio with target return
higher than the mean of Microsoft while imposing no short sales:

```{r}
#| echo: true

# efficient_portfolio(mu.vec, sigma.mat, 
#                     target_return = mu.vec["MSFT"]+0.01, 
#                     shorts = FALSE) 
# uncomment this line to see the results
```

Here, `solve.QP()` reports the error "constraints are inconsistent,
no solution!" This agrees with what we see in Figure @fig-NoShortsRandomPortfolios.
\hfill{}$\blacksquare$
:::

::: {#exm-compute-efficient-frontier-portfolios-from-three-asset-example-data .example name="Compute efficient frontier portfolios from three asset example data"}

In this example we compare the efficient frontier portfolios computed
with and without short sales. This can be easily done using the **IntroCompFinR**
function `efficient.frontier()`. First, we compute the efficient
frontier allowing short sales for target returns between the mean
of the global minimum variance portfolio and the mean of Microsoft:

```{r}
#| echo: true

ef = efficient_frontier(mu.vec, sigma.mat, alpha_min=0,                           
                        alpha_max=1, nport=10) 
ef$weight_mat
```

Here, portfolio 1 is the global minimum variance portfolio and portfolio
10 is the efficient portfolio with the same mean as Microsoft. Notice
that portfolios 9 and 10 have negative weights in Nordstrom. Next,
we compute the efficient frontier not allowing short sales for the
same range of target returns:

```{r}
#| echo: true

ef.ns = efficient_frontier(mu.vec, sigma.mat, alpha_min=0,                           
                           alpha_max=1, nport=10, shorts=FALSE) 
ef.ns$weight_mat
```

Portfolios 1 - 8 have all positive weights and are the same as portfolios
1 - 8 when short sales are allowed. However, for portfolios 9 and
10 the short sales constraint is binding. For portfolio 9, the weight
in Nordstrom is forced to zero and the weight in Starbucks is reduced.
For portfolio 10, the weights on Nordstrom and Starbucks are forced
to zero. The two frontiers are illustrated in Figure @fig-NoShortsCompareFrontiers.
For portfolios 1-8, the two frontiers coincide. For portfolios 9 and
10, the no-shorts frontier lies inside and to the right of the short-sales
frontier. The cost of the short sales constraint is the increase in the
volatility of minimum variance portfolios for target returns near
the expected return on Microsoft.

```{r}
#| label: fig-NoShortsCompareFrontiers
#| echo: false
#| fig-cap: 'Efficient frontier portfolios with and without short sales. The unconstrained
#|   efficient frontier portfolios are in blue, and the short sales constrained efficient
#|   portfolios are in red. The unconstrained portfolios labeled port 9 and port 10 have
#|   short sales in Nordstrom. The constrained portfolio labeled port 9 has zero weight
#|   in Nordstrom, and the constrained portfolio labeled port 10 has zero weights in
#|   Nordstrom and Starbucks. '
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(ef$sd, ef$er, type="b", ylim=c(0.001, 0.05), xlim=c(0.06, 0.15),       
     pch=16, col="lightblue", cex=2, ylab=expression(mu[p]), xlab=expression(sigma[p])) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(sd.vec, mu.vec, labels=asset.names, pos=3, cex=1) 
points(ef.ns$sd, ef.ns$er, type="b", pch=16, col="red") 
text(ef$sd[10], ef$er[10], labels="port 10", col="blue", pos=2) 
text(ef$sd[9], ef$er[9], labels="port 9", col="blue", pos=2) 
text(ef.ns$sd[10], ef.ns$er[10], labels="port 10", col="red", pos=4) 
text(ef.ns$sd[9], ef.ns$er[9], labels="port 9", col="red", pos=4)
```

\hfill{}$\blacksquare$
:::

### No short sales tangency portfolio 

Consider an investment universe with $N$ risky assets and a single
risk-free asset. We assume that short sales constraints only apply
to the $N$ risky assets, and that investors can borrow and lend at
the risk-free rate $r_{f}$. As shown in Chapter @sec-Portfolio-Theory-with-Matrix-Algebra,
with short sales allowed, mean-variance efficient portfolios are combinations
of the risk-free asset and the tangency portfolio. The tangency portfolio
is the maximum Sharpe ratio portfolio of risky assets and is the solution
to the maximization problem:
\begin{eqnarray*}
\underset{\mathbf{t}}{\max}\,\frac{\mu_{t}-r_{f}}{\sigma_{t}} & = & \frac{\mathbf{t}^{\prime}\boldsymbol{\mu}-r_{f}}{(\mathbf{t}^{\prime}\boldsymbol{\Sigma}\mathbf{t})^{1/2}}\,s.t.\\
\mathbf{t}^{\prime}\mathbf{1} & = & 1.
\end{eqnarray*}

::: {#exm-compute-efficient-portfolios-allowing-short-sales-from-three-asset-example-data-when-there-is-a-risk-free-asset .example name="Compute efficient portfolios allowing short sales from three asset example data when there is a risk-free asset"}

Assuming a risk-free rate $r_{f}=0.005$, we can compute the unconstrained
tangency portfolio for the three asset example data using the **IntroCompFinR**
function `tangency.portfolio()`: 

```{r}
#| echo: true

r.f = 0.005 
tan.port = tangency_portfolio(mu.vec, sigma.mat, r.f) 
summary(tan.port, r.f)
```

Here, the unconstrained tangency portfolio has a negative weight in
Nordstrom so that the short sales constraint on the risky assets will
be binding. The unconstrained set of efficient portfolios, that are
combinations of the risk-free asset and the unconstrained tangency
portfolio, is illustrated in Figure @fig-Noshorts-Tangency3. 

```{r}
#| label: fig-Noshorts-Tangency3
#| echo: false
#| fig-cap: Efficient portfolios of three risky assets and a risk-free asset allowing
#|   short sales. Nordstrom is sold short in the unconstrained tangency portfolio.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

ef = efficient_frontier(mu.vec, sigma.mat, alpha_min=-0.5,                          
alpha_max=2, nport=20) 
plot(ef, plot.assets=TRUE, col="black", pch=16) 
points(tan.port$sd_port, tan.port$er_port, col="red", pch=16, cex=2) 
sr.tan = (tan.port$er_port - r.f)/tan.port$sd_port
abline(a=r.f, b=sr.tan, col="green", lwd=2) 
text(tan.port$sd_port, tan.port$er_port, labels="Tangency", pos = 2)
```

Here, we see that the unconstrained tangency portfolio is located
on the frontier of risky asset portfolios above the point labeled
"MSFT". The green line represents portfolios of the risk-free
asset and the unconstrained tangency portfolio. 
\hfill{}$\blacksquare$
:::

Under short sales constraints on the risky assets, the maximum Sharpe
ratio portfolio solves:
\begin{eqnarray*}
\max_{\mathbf{t}}\,\frac{\mu_{t}-r_{f}}{\sigma_{t}} & = & \frac{\mathbf{t}^{\prime}\boldsymbol{\mu}-r_{f}}{(\mathbf{t}^{\prime}\boldsymbol{\Sigma}\mathbf{t})^{1/2}}\,s.t.\\
\mathbf{t}^{\prime}\mathbf{1} & = & 1,\\
t_{i} & \geq & 0,\,i=1,\ldots,N.
\end{eqnarray*}
This optimization problem *cannot* be written as a QP optimization as
expressed in @eq-NoShortsQPobjective - @eq-NoShortsQPinequality.
Hence, it looks like we cannot use the function `solve.QP()`
to find the short sales constrained tangency portfolio. However, it
turns out we can use `solve.QP()` to find the short sales constrained
tangency portfolio. To do this we utilize the alternative derivation
of the tangency portfolio presented in Chapter @sec-Portfolio-Theory-with-Matrix-Algebra.
The alternative way to compute the tangency portfolio is to first
find a minimum variance portfolio of risky assets and a risk-free
asset whose excess expected return, $\tilde{\mu}_{p,x}=\boldsymbol{\mu}^{\prime}\mathbf{x}-r_{f}\mathbf{1}$,
achieves a target excess return $\tilde{\mu}_{p,0}=\mu_{p,0}-r_{f}>0$.
This portfolio solves the minimization problem: 
$$
\min_{\mathbf{x}}~\sigma_{p,x}^{2}=\mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\textrm{ s.t. }\tilde{\mu}_{p,x}=\tilde{\mu}_{p,0},
$$
where the weight vector $\mathbf{x}$ is not constrained to satisfy
$\mathbf{x}^{\prime}\mathbf{1}=\mathbf{1}$. The tangency portfolio
is then determined by normalizing the weight vector $\mathbf{x}$
so that its elements sum to one:^[This normalization ensures that the tangency portfolio is 100\% invested
in the risky assets.]
$$
\mathbf{t}=\frac{\mathbf{x}}{\mathbf{x}^{\prime}\mathbf{1}}=\frac{\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}{\mathbf{1}^{\prime}\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}-r_{f}\cdot\mathbf{1})}.
$$
An interesting feature of this result is that it does not depend on
the value of the target excess return value $\tilde{\mu}_{p,0}=\mu_{p,0}-r_{f}>0$.
That is, every value of $\tilde{\mu}_{p,0}=\mu_{p,0}-r_{f}>0$ gives
the same value for the tangency portfolio.^[If $\tilde{\mu}_{p,0}=0$ then the portfolio of risky assets and the
risk-free asset has expected return equal to $r_{f}$. In this case,
$\mathbf{x}=\mathbf{0}$ and $x_{f}=1$. ] We can utilize this alternative derivation of the tangency portfolio
to find the tangency portfolio subject to short-sales restrictions
on risky assets. First we solve the short-sales constrained minimization
problem: 
$$
\min_{\mathbf{x}}~\sigma_{p,x}^{2} = \mathbf{x}^{\prime}\boldsymbol{\Sigma} \mathbf{x}\textrm{ s.t. }
$$ {#eq-NoshortsTan1}
$$
\tilde{\mu}_{p,x} = \tilde{\mu}_{p,0},
$$ {#eq-NoShortsTan2}
$$
x_{i}  \geq  0,\,i=1,\ldots,N.
$$ {#eq-NoShortsTan3}
Here we can use any value for $\tilde{\mu}_{p,0}$ so for convenience
we use $\tilde{\mu}_{p,0}=1$. This is a QP problem with $\mathbf{D}=2\times\boldsymbol{\Sigma}$
and $\mathbf{d}=(0,\ldots,0)^{\prime}$, one linear equality constraint,
$\tilde{\mu}_{p,x}=\tilde{\mu}_{p,0},$ and $N$ inequality constraints
$\mathbf{x}\geq\mathbf{0}$. The restriction matrices and vectors
are:
\begin{align*}
\underset{(1\times N)}{\mathbf{A}_{eq}^{\prime}} & =(\boldsymbol{\mu}-r_{f}\mathbf{1})^{\prime},\,\underset{(1\times1)}{\mathbf{b}_{eq}}=1,\text{ }\\
\underset{(N\times N)}{\mathbf{A}_{neq}^{\prime}} & =\mathbf{I}_{N},\,\underset{(N\times1)}{\mathbf{b}_{neq}}=(0,\ldots,0)^{\prime}.
\end{align*}
so that 
$$
\mathbf{A}^{\prime}=\left(\begin{array}{c}
(\boldsymbol{\mu}-r_{f}\mathbf{1})^{\prime}\\
\mathbf{I}_{N}
\end{array}\right),\text{ }\mathbf{b}=\left(\begin{array}{c}
1\\
\mathbf{0}
\end{array}\right).
$$
Here, 
\begin{eqnarray*}
\mathbf{A}_{eq}^{\prime}\mathbf{x} & = & (\boldsymbol{\mu}-r_{f}\mathbf{1})^{\prime}\mathbf{x=\tilde{\mu}_{p,x}}=1,\\
\mathbf{A}_{neq}^{\prime}\mathbf{x} & = & \mathbf{I}_{N}\mathbf{x}=\mathbf{x}\geq0.
\end{eqnarray*}
After solving the QP problem for $\mathbf{x}$, we then determine
the short-sales constrained tangency portfolio by normalizing the
weight vector $\mathbf{x}$ so that its elements sum to one:
$$
\mathbf{t}=\frac{\mathbf{x}}{\mathbf{x}^{\prime}\mathbf{1}}.
$$

::: {#exm-compute-efficient-portfolios-not-allowing-short-sales-from-three-asset-example-data-when-there-is-a-risk-free-asset .example name="Compute efficient portfolios not allowing short sales from three asset example data when there is a risk-free asset"}

First, we use `solve.QP()` to find the short sales restricted
tangency portfolio. The restriction matrices for the short sales constrained
optimization @eq-NoshortsTan1 - @eq-NoShortsTan3 are:

```{r}
#| echo: true

D.mat = 2*sigma.mat 
d.vec = rep(0, 3) 
A.mat = cbind(mu.vec - rep(1, 3)*r.f, diag(3)) 
b.vec = c(1, rep(0,3))
```

The un-normalized portfolio $\mathbf{x}$ is found using:

```{r}
#| echo: true

qp.out = solve.QP(Dmat=D.mat, dvec=d.vec,                   
                  Amat=A.mat, bvec=b.vec, meq=1) 
x.ns = qp.out$solution 
names(x.ns) = names(mu.vec) 
round(x.ns, digits=3)
```

The short sales constrained tangency portfolio is then:

```{r}
#| echo: true

t.ns = x.ns/sum(x.ns) 
round(t.ns, digits=3)
```

In this portfolio, the allocation to Nordstrom, which was negative
in the unconstrained tangency portfolio, is set to zero. 


To verify that the derivation of the short sales constrained tangency
portfolio does not depend on $\tilde{\mu}_{p,0}=\mu_{p,0}-r_{f}>0$,
we repeat the calculations with $\tilde{\mu}_{p,0}=0.5$ instead of
$\tilde{\mu}_{p,0}=1$:

```{r}
#| echo: true

b.vec = c(0.5, rep(0,3)) 
qp.out = solve.QP(Dmat=D.mat, dvec=d.vec,                   
                  Amat=A.mat, bvec=b.vec, meq=1) 
x.ns = qp.out$solution 
names(x.ns) = names(mu.vec) 
t.ns = x.ns/sum(x.ns) 
round(t.ns, digits=3)
```

You can compute the short sales restricted tangency portfolio using
the **introcompfinr** function `tangency_portfolio()`
with the optional argument `shorts=FALSE`:

```{r}
#| echo: true

tan.port.ns = tangency_portfolio(mu.vec, sigma.mat, r.f, shorts=FALSE) 
summary(tan.port.ns, r.f)
```

Notice that the Sharpe ratio on the short sales restricted tangency
portfolio is slightly smaller than the Sharpe ratio on the unrestricted
tangency portfolio.


The set of efficient portfolios are combinations of the risk-free
asset and the short sales restricted tangency portfolio. These portfolios
are illustrated in Figure @fig-NoShorts-Tangency4, created using:

```{r}
#| label: fig-NoShorts-Tangency4
#| echo: true
#| fig-cap: Efficient frontier portfolios of three risky assets and a risk-free asset not allowing short sales
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

ef.ns = efficient_frontier(mu.vec, sigma.mat, alpha_min=0,
                            alpha_max=1, nport=10, shorts=FALSE) 
plot(ef.ns, plot.assets=TRUE, pch=16) 
points(tan.port.ns$sd_port, tan.port.ns$er_port, col="red", pch=16, cex=2) 
sr.tan.ns = (tan.port.ns$er_port - r.f)/tan.port.ns$sd_port 
abline(a=r.f, b=sr.tan.ns, col="green", lwd=2) 
text(tan.port$sd_port, tan.port$er_port, labels="Tangency", pos = 2)
```

\hfill{}$\blacksquare$
:::

## Application to Vanguard Mutual Funds {#sec-Application-to-Vanguard-Mutual-Funds}

In this section, we revisit the application of mean-variance portfolio
theory to the problem of asset allocation among six Vanguard mutual
funds that was discussed at the end of Chapter @sec-Portfolio-Theory-with-Matrix-Algebra.
For the sub-sample January, 2010 through December, 2014 it was shown
that estimated mean-variance efficient portfolios had short sales
in many of the Vanguard funds. Since it is not possible to short mutual
funds, we re-estimate the efficient portfolios with the additional
constraints that all portfolio weights must be positive. The monthly
returns and annualized GWN model estimates for the Vanguard funds
are created using:

```{r}
#| echo: true

VanguardPrices = as.xts(VanguardPrices) 
VanguardRetS = na.omit(Return.calculate(VanguardPrices, method="simple")) 
smpl = "2010-1::2014-12" 
r.f = 0.001
muhat = 12*colMeans(VanguardRetS[smpl]) 
covhat = 12*cov(VanguardRetS[smpl])
```

We use the **introcompfinr** portfolio functions to compute unrestricted
and short-sales restricted estimates of the global minimum variance
portfolio, the tangency portfolio, and the efficient frontier of risky
assets:

```{r}
#| echo: true

# unrestricted (short sales allowed) portfolios
gmin.port = globalmin_portfolio(muhat, covhat) 
tan.port = tangency_portfolio(muhat, covhat, r.f) 
ef = efficient_frontier(muhat, covhat, alpha_min=0,                                                   
                        alpha_max=1.5, nport=20)
# short sales restricted portfolios
gmin.port.ns = globalmin_portfolio(muhat, covhat, shorts=FALSE) 
tan.port.ns = tangency_portfolio(muhat, covhat, r.f, shorts=FALSE) 
ef.ns = efficient_frontier(muhat, covhat, alpha_min=0,                                                   
                           alpha_max=1, nport=20, shorts=FALSE)
```

The global minimum variance portfolios are:

```{r}
#| echo: true

# short sales allowed
gmin.port
# short sales restricted
gmin.port.ns
```

In the short sales restricted global minimum variance portfolio there
is 98\% weight on vbisx (low volatility US short-term bond fund) and
2\% weight on vfinx (S\&P 500 fund). The weights on other funds are
pushed to zero. The volatility of the short sales restricted portfolio
is slightly higher than the volatility of the unrestricted portfolio
and so the cost of the short sales restrictions is small. 


The estimated tangency portfolios are:

```{r}
#| echo: true

# short sales allowed
summary(tan.port, r.f)
# short sales restricted
summary(tan.port.ns, r.f)
```

The short sales restricted tangency portfolio has positive weights
on vfinx, vbltx, and vbisx and zero weights for the remaining assets.
The Sharpe ratio of the short sales restricted tangency portfolio,
however, is substantially smaller than the Sharpe ratio of the unrestricted
portfolio. 


Figure @fig-NoShortsVanguardEfficientPortfolios shows the short
sales restricted and unrestricted efficient portfolio frontiers, created
using:

```{r}
#| label: fig-NoShortsVanguardEfficientPortfolios
#| echo: true
#| fig-cap: Unrestricted and short sales restricted efficient portfolio frontiers.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# unrestricted efficient frontiers 
plot(ef, plot.assets=TRUE, col="blue", pch=16) 
text(tan.port$sd_port, tan.port$er_port,labels = "tan", pos = 2) 
sr.tan = (tan.port$er_port - r.f)/tan.port$sd_port 
text(0, r.f, labels=expression(r[f]), pos = 4) 
abline(a=r.f, b=sr.tan, col="green", lwd=2) 
abline(h=0, v=0) 
# short sales restricted frontiers 
sr.tan.ns = (tan.port.ns$er_port - r.f)/tan.port.ns$sd_port
points(ef.ns$sd_vec, ef.ns$er_vec, type="b", pch=16, col="red") 
abline(a=r.f, b=sr.tan.ns, col="green", lwd=2) 
text(tan.port.ns$sd_port, tan.port.ns$er_port,labels = "tan (no shorts)", pos = 4)
```

The short sales restricted risky asset only frontier, shown in red,
lies underneath and to the right of the unrestricted risky asset frontier,
shown in blue. The short sales restricted tangency portfolio has a
similar expected return as the unrestricted tangency portfolio but
has a higher volatility and a lower Sharpe ratio. 


## Further Reading: Portfolio Theory with Short Sales Constraints

-  Mention other quadratic programming packages: ROI etc. See Doug's book
-  Mention Pfaff's R book [@pfaff-2013] ??
-  Mention Grilli's book etc.

To be completed...

## Exercises

To be completed


{MartinSchererYollin}  Martin, R.D., Scherer, B., and Yollin, G. (2016). 
