# Statistical Analysis of Portfolios {#sec-Statistical-Analysis-of-Portfolios}

Updated: September 19, 2026

Copyright © Eric Zivot 2026

In @sec-Introduction-to-Portfolio-Theory I introduced mean-variance portfolio theory in the context of simple portfolios with two risky assets and a safe asset. The theory presented assumed that investors know the true expected returns, variances, and correlations between returns (i.e., the parameters of the GWN model of returns). Of course, the real world implementation of mean-variance portfolio theory requires investors to estimate expected returns, variances, and correlations. In this chapter I explore the implications of estimation errors in the parameters of the GWM for asset returns for the application of mean-variance portfolio theory. The quantification of estimation error in mean-variance portfolio analysis is novel and under appreciated by many practitioners, and provides  insights into some pitfalls in the practical application of mean-variance portfolio theory. Because mean-variance portfolios are functions of GWN model parameters, I use tools for the statistical analysis of functions of GWN model parameters presented in @sec-The-Bootstrap throughout this chapter. 

The R packages used in this chapter are **boot**, **bootstrap**, **car**, **ellipse**, 
**introcompfinr**, **mvtnorm**, and **PerformanceAnalytics**. Have these packages installed and 
loaded in R before replicating the chapter examples.

```{r}
#| include: false
suppressPackageStartupMessages(library(boot))
suppressPackageStartupMessages(library(bootstrap))
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(ellipse))
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(mvtnorm))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```


## Portfolios of Two Risky Assets
\index{Statistical analysis}
Most of the issues involved with the statistical analysis of portfolios
can be illustrated in the simple case of a two asset portfolio. Let
$R_{A}$ and $R_{B}$ denote the simple returns on two risky assets
and assume these returns are characterized by the GWN model: 
$$
\left(\begin{array}{c}
R_{A}\\
R_{B}
\end{array}\right)\sim N\left(\left(\begin{array}{c}
\mu_{A}\\
\mu_{B}
\end{array}\right),\,\left(\begin{array}{cc}
\sigma_{A}^{2} & \sigma_{AB}\\
\sigma_{AB} & \sigma_{B}^{2}
\end{array}\right)\right). 
$$ {#eq-StatisticalGWNmodel}

In matrix notation:
$$
\mathbf{R}\sim N(\boldsymbol{\mu},\,\boldsymbol{\Sigma}),
$$
where 
$$
\mathbf{R}=\left(\begin{array}{c}
R_{A}\\
R_{B}
\end{array}\right),\,\boldsymbol{\mu}=\left(\begin{array}{c}
\mu_{A}\\
\mu_{B}
\end{array}\right),\,\boldsymbol{\Sigma}=\left(\begin{array}{cc}
\sigma_{A}^{2} & \sigma_{AB}\\
\sigma_{AB} & \sigma_{B}^{2}
\end{array}\right).
$$
A portfolio with weight vector $\mathbf{x}=(x_{A},x_{B})^{\prime}$
has return $R_{p}=\mathbf{x}^{\prime}\mathbf{R}=x_{A}R_{A}+x_{B}R_{B}$
which is also described by the GWN model:
\begin{eqnarray*}
R_{p} & \sim & N(\mu_{p},\,\sigma_{p}^{2}),\\
\mu_{p} & = & \mathbf{x}^{\prime}\mu=x_{A}\mu_{A}+x_{B}\mu_{B},\\
\sigma_{p}^{2} & = & \mathbf{x}^{\prime}\mathbf{\varSigma}\mathbf{x}=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}.
\end{eqnarray*}
I assume that $\{\mathbf{r}_{t}\}_{t=1}^{T}$ is the observed sample of asset returns of size $T$, where $\mathbf{r}_{t}=(r_{At},r_{Bt})^{\prime}$, is
generated from the GWN model @eq-StatisticalGWNmodel from
which I create the sample portfolio returns $r_{p,t}=\mathbf{x}^{\prime}\mathbf{r}_{t}=x_{A}r_{At}+x_{B}r_{Bt}$. I assume that the portfolio weight vector $\mathbf{x}$ remains fixed over the observed sample and that the portfolio is rebalanced every period to maintain the fixed weights. 

The true GWN model parameters are unknown in practice and must be
estimated from the observed data. As described in @sec-Estimation-of-The-GWN-Model, the GWN model parameters are estimated using the sample statistics: 
$$
\begin{aligned}
& \hat{\mu}_{i}=\frac{1}{T}\sum_{t=1}^{T}r_{it},\,\hat{\sigma}_{i}^{2}=\frac{1}{T-1}\sum_{t=1}^{T}(r_{it}-\hat{\mu}_{i})^{2}, \nonumber \\
& \hat{\sigma}_{i}=\sqrt{\hat{\sigma}_{i}^{2}},\,\hat{\sigma}_{ij}^{2}=\frac{1}{T-1}\sum_{t=1}^{T}(r_{it}-\hat{\mu}_{i})(r_{jt}-\hat{\mu}_{j}).
\end{aligned}
$$ {#eq-StatisticalGWNestimatesAssets}
These estimates are consistent, asymptotically normally distributed, and have estimation errors which can be quantified by standard error formulas:^[$\hat{\sigma}_{i}$ is asymptotically unbiased and the bias for finite $T$ is typically very small so that it is essentially unbiased for moderately sized $T$.] 

$$
\mathrm{se}(\hat{\mu}_{i})=\frac{\sigma_{i}}{\sqrt{T}},\,\mathrm{se}(\hat{\sigma}_{i}^{2}) \approx \frac{\sigma_{i}^{2}}{\sqrt{T/2}},\,\mathrm{se}(\hat{\sigma}_{i}) \approx \frac{\sigma_{i}}{\sqrt{2T}},\,\mathrm{se}(\hat{\sigma}_{ij}) \approx \frac{\sqrt{\sigma_i^2 \sigma_j^2 + \sigma_{ij}^2}}{\sqrt{T}}.
$$ {#eq-StatisticalGWNseAssets}

### Estimation of GWN Model Parameters for Portfolios

You can estimate the GWN model parameters $\mu_{p}$, $\sigma_{p}^{2}$
and $\sigma_{p}$ for the portfolio return, $R_{p}$, using two equivalent
methods. In the first method, you use the sample portfolio returns
$\{r_{p,t}\}_{t=1}^{T}$ and compute the sample statistics:
$$
\hat{\mu}_{p}=\frac{1}{T}\sum_{t=1}^{T}r_{p,t},\,\hat{\sigma}_{p}^{2}=\frac{1}{T-1}\sum_{t=1}^{T}(r_{p,t}-\hat{\mu}_{p})^{2},\,\hat{\sigma}_{p}=\sqrt{\hat{\sigma}_{p}^{2}}.
$$ {#eq-StatisticalGWNestimatesPortfolio}
These estimates are unbiased and asymptotically normal with standard
errors:
$$
\mathrm{se}(\hat{\mu}_{p})=\frac{\sigma_{p}}{\sqrt{T}},\,\mathrm{se}(\hat{\sigma}_{p}^{2})\approx\frac{\sigma_{p}^{2}}{\sqrt{T/2}},\,\mathrm{se}(\hat{\sigma}_{p})\approx\frac{\sigma_{p}}{\sqrt{2T}}.
$$ {#eq-StatisticalGWNsePortfolio}
In the second method, I compute estimates of $\mu_{p}$ and $\sigma_{p}^{2}$
directly from the individual asset estimates in @eq-StatisticalGWNestimatesAssets:
$$
\hat{\mu}_{p} = \mathbf{x}^{\prime}\hat{\boldsymbol{\mu}}=x_{A}\hat{\mu}_{A}+x_{B}\hat{\mu}_{B},
$$ {#eq-StatisticalGWNestimatesPortfolioMu}
$$
\hat{\sigma}_{p}^{2} = \mathbf{x}^{\prime}\widehat{\boldsymbol{\Sigma}}\mathbf{x}=x_{A}^{2}\hat{\sigma}_{A}^{2}+x_{B}^{2}\hat{\sigma}_{B}^{2}+2x_{A}x_{B}\hat{\sigma}_{AB}.
$$ {#eq-StatisticalGWNestimatesPortfolioVar}
The estimates of $\mu_{p}$ and $\sigma_{p}^{2}$ in @eq-StatisticalGWNestimatesPortfolioMu
and @eq-StatisticalGWNestimatesPortfolioVar, respectively,
are numerically identical to the estimates in @eq-StatisticalGWNestimatesPortfolio.
To see this, consider the calculation of $\hat{\mu}_{p}$ in @eq-StatisticalGWNestimatesPortfolioMu:
\begin{eqnarray*}
\hat{\mu}_{p} & = & x_{A}\hat{\mu}_{A}+x_{B}\hat{\mu}_{B}=x_{A}\left(\frac{1}{T}\sum_{t=1}^{T}r_{At}\right)+x_{B}\left(\frac{1}{T}\sum_{t=1}^{T}r_{Bt}\right)\\
 & = & \frac{1}{T}\sum_{t=1}^{T}(x_{A}r_{At}+x_{B}r_{Bt})=\frac{1}{T}\sum_{t=1}^{T}r_{p,t}.
\end{eqnarray*}
The proof of the equivalence of $\hat{\sigma}_{p}^{2}$ in @eq-StatisticalGWNestimatesPortfolioVar
with $\hat{\sigma}_{p}^{2}$ in @eq-StatisticalGWNestimatesPortfolio
is left as an end-of-chapter exercise. 

The standard errors of $\hat{\mu}_{p}$ and $\hat{\sigma}_{p}^{2}$
can also be directly calculated from @eq-StatisticalGWNestimatesPortfolioMu
and @eq-StatisticalGWNestimatesPortfolioVar. Consider the
calculation for $\mathrm{se}(\hat{\mu}_{p})$:
$$
\mathrm{var}(\hat{\mu}_{p})=\mathrm{var}(x_{A}\hat{\mu}_{A}+x_{B}\hat{\mu}_{B})=x_{A}^{2}\mathrm{var}(\hat{\mu}_{A})+x_{B}^{2}\mathrm{var}(\hat{\mu}_{B})+2x_{A}x_{B}\mathrm{cov}(\hat{\mu}_{A},\hat{\mu}_{B}).
$$ {#eq-StatisticalSEmu}
In chapter @sec-Estimation-of-The-GWN-Model it was shown that
$$
\mathrm{var}(\hat{\mu}_{A})=\frac{\sigma_{A}^{2}}{T},\,\mathrm{var}(\hat{\mu}_{B})=\frac{\sigma_{B}^{2}}{T},\, \mathrm{cov}(\hat{\mu}_{A},\hat{\mu}_{B})=\frac{\sigma_{AB}}{T}. 
$$
Then @eq-StatisticalSEmu can be rewritten as:
\begin{eqnarray*}
\mathrm{var}(\hat{\mu}_{p}) & = & x_{A}\frac{\sigma_{A}^{2}}{T}+x_{B}\frac{\sigma_{B}^{2}}{T}+2x_{A}x_{B}\frac{\sigma_{AB}}{T}\\
 & = & \frac{1}{T}\left(x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}\right)\\
 & = & \frac{\sigma_{p}^{2}}{T}.
\end{eqnarray*}
Hence, 
$$
\mathrm{se}(\hat{\mu}_{p})=\sqrt{\mathrm{var}(\hat{\mu}_{p})}=\frac{\sigma_{p}}{\sqrt{T}},
$$
which is the same formula given in @eq-StatisticalGWNsePortfolio.

Unfortunately, the calculation of $\mathrm{se}(\hat{\sigma}_{p}^{2})$
is a bit horrific and I only sketch out some of the details. Now,
$$
\begin{aligned}
\mathrm{var}(\hat{\sigma}_{p}^{2}) & =  \mathrm{var}(x_{A}^{2}\hat{\sigma}_{A}^{2}+x_{B}^{2}\hat{\sigma}_{B}^{2}+2x_{A}x_{B}\hat{\sigma}_{AB})\\
 & =  x_{A}^{4}\mathrm{var}(\hat{\sigma}_{A}^{2})+x_{B}^{4}\mathrm{var}(\hat{\sigma}_{B}^{2})+4x_{A}^{2}x_{B}^{2}\mathrm{var}(\hat{\sigma}_{AB})\nonumber \\
 &   +2x_{A}^{2}x_{B}^{2}\mathrm{cov}(\hat{\sigma}_{A}^{2},\hat{\sigma}_{B}^{2})+4x_{A}^{3}x_{B}\mathrm{cov}(\hat{\sigma}_{A}^{2},\hat{\sigma}_{AB})\nonumber \\
 &   +4x_{A}x_{B}^{3}\mathrm{cov}(\hat{\sigma}_{B}^{2},\hat{\sigma}_{AB})\nonumber 
\end{aligned}
$$ {#eq-StatisticalHorrificFormula}
Unfortunately, there are no easy formulas for $\mathrm{var}(\hat{\sigma}_{AB})$,
$\mathrm{cov}(\hat{\sigma}_{A}^{2},\hat{\sigma}_{B}^{2})$, $\mathrm{cov}(\hat{\sigma}_{A}^{2},\hat{\sigma}_{AB})$
and $\mathrm{cov}(\hat{\sigma}_{B}^{2},\hat{\sigma}_{AB})$. With
a CLT approximation and much tedious calculation it can be shown that
@eq-StatisticalHorrificFormula is equal to the
square of $\mathrm{se}(\hat{\sigma}_{p}^{2})$ given in @eq-StatisticalGWNsePortfolio.
Obviously, the method of computing standard errors for $\hat{\mu}_{p}$
and $\hat{\sigma}_{p}^{2}$ directly from the GWN model estimates
computed from the sample portfolio returns is much easier than the
indirect calculations based on the individual asset estimates, especially for $\hat{\sigma}_{p}^{2}$. 

::: {#exm-example-data-for-two-risky-asset-portfolio .example name="Example data for two risky asset portfolio"}

To be completed... table example data

@tbl-TableExampleData from @sec-Introduction-to-Portfolio-Theory gives annual return distribution parameters for two hypothetical assets $A$ and $B$. I use this
distribution but re-scale the values to represent a monthly return
distribution. Asset $A$ is the high risk asset with a monthly return
of $\mu_{A}=1.46\%$ and monthly standard deviation of $\sigma_{A}=7.45\%$.
Asset B is a lower risk asset with monthly return $\mu_{B}=0.458\%$
and monthly standard deviation of $\sigma_{B}=3.32\%$. The assets
are assumed to be slightly negatively correlated with correlation
coefficient $\rho_{AB}=-0.164$. Given the standard deviations and
the correlation, the covariance can be determined from $\sigma_{AB}=\rho_{AB}\sigma_{A}\sigma_{B}=-0.0004$.
In R, these GWN model parameters are created using:

```{r}
#| echo: true

mu.A = 0.175/12
sig.A = 0.258/sqrt(12)
sig2.A = sig.A^2
mu.B = 0.055/12
sig.B = 0.115/sqrt(12)
sig2.B = sig.B^2
rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B
sigma.mat = matrix(c(sig2.A, sig.AB, sig.AB, sig2.B),
                   2, 2, byrow=TRUE) 
mu.vec = c(mu.A, mu.B) 
sd.vec = c(sig.A, sig.B) 
names(mu.vec) = names(sd.vec) = c("asset.A", "asset.B") 
dimnames(sigma.mat) = list(names(mu.vec), names(mu.vec))
```

For portfolio analysis, I consider an equally weighted portfolio
of assets A and B. The GWN model parameters for this portfolio are:

```{r}
#| echo: true

x1.A = 0.5 
x1.B = 1 - x1.A 
x1.vec = c(x1.A, x1.B) 
names(x1.vec) = names(mu.vec)
mu.p1 = x1.A*mu.A + x1.B*mu.B 
sig2.p1 = x1.A^2 * sig2.A + x1.B^2 * sig2.B + 2*x1.A*x1.B*sig.AB 
sig.p1 = sqrt(sig2.p1)
cbind(mu.p1, sig2.p1, sig.p1)
```

I use the above GWN model parameters to simulate $T=60$ hypothetical
returns for assets A and B and the equally weighted portfolio:

```{r}
#| echo: true

n.obs = 60 
set.seed(125) 
returns.sim = rmvnorm(n.obs, mean=mu.vec, sigma=sigma.mat) 
colnames(returns.sim) = names(mu.vec)
rp1.sim = returns.sim%*%x1.vec
```

The estimates of the GWN model parameters from the simulated returns
for the individual assets are:

```{r}
#| echo: true

muhat.vals = colMeans(returns.sim) 
sig2hat.vals = apply(returns.sim, 2, var) 
sigmahat.vals = apply(returns.sim, 2, sd) 
covmat.hat = cov(returns.sim) 
covhat = covmat.hat[1,2]
rbind(muhat.vals, sig2hat.vals, sigmahat.vals) 
covhat 
```

The estimated standard errors and 95\% confidence intervals for $\hat{\mu}_{i}$
and $\hat{\sigma}_{i}$ $(i=A,\,B)$ are:

```{r}
#| echo: true

# estimated standard errors
se.muhat = sigmahat.vals/sqrt(n.obs) 
se.sigmahat = sigmahat.vals/sqrt(2*n.obs) 
rbind(se.muhat, se.sigmahat)
# 95% confidence intervals 
lower.mu = muhat.vals - 2*se.muhat 
upper.mu = muhat.vals + 2*se.muhat 
lower.sigma = sigmahat.vals - 2*se.sigmahat 
upper.sigma = sigmahat.vals + 2*se.sigmahat 
cbind(lower.mu, upper.mu) 
cbind(lower.sigma, upper.sigma)
```

Here, you see that the means are not estimated as precisely as the
volatilities and that all of the true values are contained in the
95\% confidence intervals. 

For the equally weighted portfolio, the GWN model estimates computed
directly from the simulated portfolio returns using @eq-StatisticalGWNestimatesPortfolio
are:

```{r}
#| echo: true
muhat.p1 = mean(rp1.sim) 
sig2hat.p1 = as.numeric(var(rp1.sim)) 
sigmahat.p1 = sd(rp1.sim) 
cbind(muhat.p1, sig2hat.p1, sigmahat.p1)
```

The portfolio estimates computed from the asset estimates @eq-StatisticalGWNestimatesPortfolioMu
and @eq-StatisticalGWNestimatesPortfolioVar are numerically
equivalent:

```{r}
#| echo: true
muhat.p1 = as.numeric(x1.A*muhat.vals[1] + x1.B*muhat.vals[2]) 
sig2hat.p1 = as.numeric(x1.A^2 * sig2hat.vals[1] + 
                        x1.B^2 * sig2hat.vals[2] + 
                        2*x1.A*x1.B*covhat) 
sighat.p1 = sqrt(sig2hat.p1) 
cbind(muhat.p1, sig2hat.p1, sigmahat.p1)
```

The estimated standard errors for $\hat{\mu}_{p}$ and $\hat{\sigma}_{p}$,
and 95\% confidence intervals for $\mu_{p}$ and $\sigma_{p}$ computed
using @eq-StatisticalGWNsePortfolio are:

```{r}
#| echo: true

# estimated standard errors
se.muhat.p1 = sigmahat.p1/sqrt(n.obs) 
se.sigmahat.p1 = sigmahat.p1/sqrt(2*n.obs) 
cbind(se.muhat.p1, se.sigmahat.p1)
# 95% confidence intervals 
lower.mu = muhat.p1 - 2*se.muhat.p1 
upper.mu = muhat.p1 + 2*se.muhat.p1 
lower.sigma = sigmahat.p1 - 2*se.sigmahat.p1 
upper.sigma = sigmahat.p1 + 2*se.sigmahat.p1 
ans = rbind(c(lower.mu, upper.mu),             
c(lower.sigma, upper.sigma)) 
colnames(ans) = c("lower", "upper") 
rownames(ans) = c("mu.p", "sigma.p") 
ans
```

As with the individual assets, the portfolio mean is estimated less
precisely than the portfolio volatility and the true values are in
the 95\% confidence intervals. 
\hfill{}$\blacksquare$
:::

### Estimation error in the risk-return diagram

In the analysis of portfolios in @sec-Introduction-to-Portfolio-Theory, the risk return trade-off between portfolios was visualized by plotting the volatility-expected return pairs $(\hat{\sigma}_{p},\,\hat{\mu}_{p})$ of different portfolios. This was done without considering estimation error in $\hat{\sigma}_{p}$ and $\hat{\mu}_{p}$. Estimation error in $\hat{\sigma}_{p}$ and $\hat{\mu}_{p}$
creates uncertainty about the location of the true values $(\sigma_{p},\,\mu_{p})$ in the risk-return diagram. 

::: {#exm-true-and-estimated-risk-return-diagram-for-example-data .example name="True and estimated risk return diagram for example data"}

@fig-StatisticalRiskReturn shows the true values and
sample estimates of the pairs $(\sigma_{A},\mu_{A}),$ $(\sigma_{B},\mu_{B})$,
and $(\sigma_{p},\mu_{p})$ for the equally weighted portfolio created
using:


```{r}
#| label: fig-StatisticalRiskReturn
#| echo: true
#| fig-cap: Risk-return diagram for example data. $\mathrm{A}=(\sigma_{A},\mu_{A})$,
#|   $\mathrm{B}=(\sigma_{B},\mu_{B})$, $\mathrm{\hat{A}}=(\hat{\sigma}_{A},~\hat{\mu}_{A})$,
#|   $\mathrm{\hat{B}}=(\hat{\sigma}_{B},~\hat{\mu}_{B})$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# true values
plot(sd.vec, mu.vec, pch=16, col="black",       
     ylim=c(-0.01, 0.04), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]),      
     cex=2) 
abline(h=0) 
points(sig.p1, mu.p1, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5) 
# estimates 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
points(sigmahat.p1, muhat.p1, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels=expression(hat(A)),       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels=expression(hat(B)),       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.p1, y=muhat.p1, labels=expression(hat(P)),       
     col="blue", pos=4, cex = 1.5)
```

You see that $(\hat{\sigma}_{B},\hat{\mu}_{B})$ is fairly close to
$(\sigma_{B},\mu_{B})$ but that $(\hat{\sigma}_{A},\hat{\mu}_{A})$
is quite far above $(\sigma_{A},\mu_{A})$ and $(\hat{\sigma}_{P},\hat{\mu}_{P})$
is moderately far above $(\sigma_{P},\mu_{P})$. The large positive
estimation errors in $\hat{\mu}_{P}$ and $\hat{\mu}_{A}$ greatly
overstate the risk-return characteristics of the equally weighted
portfolio and asset A. 
\hfill{}$\blacksquare$
:::


To illustrate estimation error in the risk return diagram, the individual
95\% confidence intervals for $\mu_{i}$ and $\sigma_{i}$ could be
superimposed on the plot as rectangles centered at the pairs $(\hat{\sigma}_{i},\hat{\mu}_{i})$.
However, the probability that these rectangles contain the true pairs
$(\sigma_{i},\mu_{i})$ is not equal to 95\% because the confidence
intervals for $\mu_{i}$ are created independently from the confidence
intervals for $\sigma_{i}$. To create a joint confidence set that
cover the pair $(\sigma_{i},\mu_{i})$ with probability 95\% requires
knowing the joint probability distribution of $(\hat{\sigma}_{i},\hat{\mu}_{i})$.

In @sec-Estimation-of-The-GWN-Model, it was shown that
in the GWN model: 
$$
\left(\begin{array}{c}
\hat{\sigma}_{i}\\
\hat{\mu}_{i}
\end{array}\right)\sim N\left(\left(\begin{array}{c}
\sigma_{i}\\
\mu_{i}
\end{array}\right),\left(\begin{array}{cc}
\mathrm{se}(\hat{\sigma}_{i})^{2} & 0\\
0 & \mathrm{se}(\hat{\mu}_{i})^{2}
\end{array}\right)\right), 
$$ {#eq-StatisticalJointPdf}
for large enough $T$. Hence, $\hat{\sigma}_{i}$ and $\hat{\mu}_{i}$
are (asymptotically) jointly normally distributed and $\mathrm{cov}(\hat{\sigma}_{i},\hat{\mu}_{i})=0$
which implies that they are also independent. Let $\theta=(\sigma_{i},\mu_{i})^{\prime},\,$$\hat{\theta}=(\hat{\sigma}_{i},\hat{\mu}_{i})^{\prime}$
and $\mathbf{V}=\mathrm{diag}(\mathrm{se}(\hat{\sigma}_{i})^{2},\mathrm{se}(\hat{\mu}_{i})^{2})$.
Then @eq-StatisticalJointPdf can be expressed as $\hat{\theta}\sim N(\theta,\,\mathbf{V})$.
It follows that the quadratic form $\left(\hat{\theta}-\theta\right)^{\prime}\mathbf{V}^{-1}\left(\hat{\theta}-\theta\right)\sim\chi^{2}(2),$
and so 
$$
Pr\left\{ \left(\hat{\theta}-\theta\right)^{\prime}
\mathbf{V}^{-1}\left(\hat{\theta}-\theta\right)\leq q_{.95}^{\chi^{2}(2)}\right\}  
=0.95
$$ {#eq-StatisticalJointCI}

where $q_{.95}^{\chi^{2}(2)}$ is the 95\% quantile of the $\chi^{2}(2)$
distribution. The equation $\left(\hat{\theta}-\theta\right)^{\prime}\mathbf{V}^{-1}\left(\hat{\theta}-\theta\right)=q_{.95}^{\chi^{2}(2)}$
defines an un-tilted ellipse in $\sigma_{i}-\mu_{i}$ space centered
at $(\hat{\sigma}_{i},\hat{\mu}_{i})$ with axes proportional to $\mathrm{se}(\hat{\sigma}_{i})$
and $\mathrm{se}(\hat{\mu}_{i})$, respectively. This ellipse is the
joint 95\% confidence set for $(\sigma_{i},\mu_{i})$. 


::: {#exm-true-and-estimated-risk-return-diagram-with-confidence-ellipses-for-example-data .example name="True and estimated risk return diagram with confidence ellipses for example data"}

@fig-StatisticalRiskReturnCI repeats @fig-StatisticalRiskReturn
with the addition of the joint confidence sets for $(\sigma_{A},\mu_{A}),$
$(\sigma_{B},\mu_{B})$, and $(\sigma_{p},\mu_{p})$ , created using

```{r}
#| label: fig-StatisticalRiskReturnCI
#| echo: false
#| fig-cap: Risk return tradeoff with 95\% confidence ellipses
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sd.vec, mu.vec, pch=16, col="black",       
     ylim=c(-0.01, 0.06), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]),cex=2) 
abline(h=0) 
points(sig.p1, mu.p1, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5) 
points(sigmahat.vals, muhat.vals, 
       pch=16, col="blue", cex=2)
points(sigmahat.p1, muhat.p1, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels=expression(hat(A)),       
   col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels=expression(hat(B)),       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.p1, y=muhat.p1, labels=expression(hat(P)),   
     col="blue", pos=4, cex = 1.5)
# Create asymptotic variances 
V.A = matrix(c(se.sigmahat[1]^2, 0 ,0, 
               se.muhat[1]^2), 2, 2, byrow=TRUE) 
V.B = matrix(c(se.sigmahat[2]^2, 0 ,0, 
               se.muhat[2]^2), 2, 2, byrow=TRUE) 
V.P = matrix(c(se.sigmahat.p1^2, 0 ,0, 
               se.muhat.p1^2), 2, 2, byrow=TRUE)
# plot confidence ellipses 
lines(ellipse(V.A, centre=c(sigmahat.vals[1], 
                            muhat.vals[1]), level=0.95), 
      col="red" , lwd=2) 
lines(ellipse(V.B, centre=c(sigmahat.vals[2], 
                            muhat.vals[2]), level=0.95),         
      col="red", lwd=2 ) 
lines(ellipse(V.P, centre=c(sigmahat.p1, 
                            muhat.p1), level=0.95),         
      col="red", lwd=2 )
```


The `ellipse()` function from the R package **ellipse**
is used to draw the 95\% confidence sets. The confidence ellipses
are longer in the $\mu_{i}$ direction because $\mathrm{se}(\hat{\mu}_{i})$
is larger than $\mathrm{se}(\hat{\sigma}_{i})$ and indicate more
uncertainty about the true values of $\mu_{i}$ than the true values
of $\sigma_{i}.$ The ellipse for $(\sigma_{A},\mu_{A})$ is much
bigger than the ellipses for $(\sigma_{B},\mu_{B})$ and $(\sigma_{p},\mu_{p})$
and indicates a wide range of possible values for $(\sigma_{A},\mu_{A})$
. For all pairs, the true values are inside the 95\% confidence ellipses.^[Notice, however, that the pair $(\sigma_{p},\mu_{p})$ is inside the
confidence ellipse for $(\sigma_{B},\mu_{B})$ but that $(\sigma_{B},\mu_{B})$
is not in the confidence ellipse for $(\sigma_{p},\mu_{p})$. ] 
\hfill{}$\blacksquare$
:::

The estimation errors in the points $(\hat{\sigma}_{i},\hat{\mu}{}_{i})$
can also be illustrated using the bootstrap. You simply sample with
replacement from the observed returns $B$ times and compute estimates
of $(\sigma_{i},\mu_{i})$ from each bootstrap sample. You can then
plot the $B$ bootstrap pairs $(\hat{\sigma}_{i}^{*},\hat{\mu}_{i}^{*})$
on the risk return diagram instead of the confidence ellipses. 

::: {#exm-true-and-estimated-risk-return-diagram-with-bootstrap-estimates-for-example-data .example name="True and estimated risk return diagram with bootstrap estimates for example data"}

The following R code creates $B=500$ bootstrap estimates of $(\sigma_{i},\mu_{i})$
for $i=A,B,P$:

```{r}
#| echo: true

n.boot = 500 
mu.boot = matrix(0, n.boot, 3) 
sd.boot = matrix(0, n.boot, 3) 
colnames(mu.boot) = colnames(sd.boot) = c("A", "B", "P")
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
   rp1.boot = ret.boot%*%x1.vec   
   ret.boot = cbind(ret.boot, rp1.boot)   
   mu.boot[i, ] = colMeans(ret.boot)   
   sd.boot[i, ] = apply(ret.boot, 2, sd)  
}
```

@fig-StatisticalRiskReturnBoot repeats @fig-StatisticalRiskReturnCI
and adds the bootstrap estimates of $(\sigma_{i},\mu_{i})$ for $i=A,B,P$.
Notice that the bootstrap estimates for each asset produce a scatter
that fills the 95\% confidence ellipses with just a few estimates
lying outside the ellipses. Hence, the bootstrap is an easy and effective
way to illustrate estimation error in the risk-return diagram.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-StatisticalRiskReturnBoot
#| echo: false
#| fig-cap: Risk return trade off with bootstrap estimates.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# show estimates with 95% confidence intervals 
plot(sd.vec, mu.vec, type="n",  ylim=c(-0.01, 0.06), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# bootstrap estimates 
points(sd.boot[, "A"], mu.boot[, "A"], pch=16, col="lightblue") 
points(sd.boot[, "B"], mu.boot[, "B"], pch=16, col="grey") 
points(sd.boot[, "P"], mu.boot[, "P"], pch=16, col="pink") 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
abline(h=0) 
points(sig.p1, mu.p1, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
text(x=sig.p1, y=mu.p1, labels="P", pos=4, cex = 1.5) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
points(sigmahat.p1, muhat.p1, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
    col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
    col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.p1, y=muhat.p1, labels="P Hat",       
     col="blue", pos=4, cex = 1.5) 
# plot confidence ellipses 
lines( ellipse(V.A, centre=c(sigmahat.vals[1], muhat.vals[1]), level=0.95),         
      col="red" , lwd=2) 
lines( ellipse(V.B, centre=c(sigmahat.vals[2], muhat.vals[2]), level=0.95),         
      col="red", lwd=2 ) 
lines( ellipse(V.P, centre=c(sigmahat.p1, muhat.p1), level=0.95),         
     col="red", lwd=2 ) 
```


### Estimation error in the portfolio frontier

In the case of two risky assets, the portfolio frontier is a plot
of the risk-return characteristics of all feasible portfolios. From
this frontier, you can readily determine the set of efficient portfolios
which are those portfolios that have the highest expected return for
a given risk level. If you construct the portfolio frontier from portfolio
risk and return estimates from the GWN model, then the estimation
error in these estimates leads to estimation error in the entire portfolio
frontier. 

::: {#exm-true-and-estimated-portfolio-frontier-for-example-data .example name="True and estimated portfolio frontier for example data"}

Using the example data, you can construct the true unobservable portfolio
frontier as well as an estimate of this frontier from the simulated
returns. For example, consider creating the true and estimated portfolio
frontiers for the following portfolios:

```{r}
#| echo: true

x.A = seq(from=-0.4, to=1.4, by=0.1) 
x.B = 1 - x.A
```

The risk-return characteristics of the portfolios on the true portfolio
frontier are:

```{r}
#| echo: true

mu.p = x.A*mu.A + x.B*mu.B 
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB 
sig.p = sqrt(sig2.p)
```

The risk-return characteristics of the portfolios on the estimated
frontier are:

```{r}
#| echo: true

muhat.p = x.A*muhat.vals["asset.A"] + x.B*muhat.vals["asset.B"] 
sig2hat.p = x.A^2 * sig2hat.vals["asset.A"] + 
   x.B^2 * sig2hat.vals["asset.B"] + 2*x.A*x.B*covhat 
sighat.p = sqrt(sig2hat.p)
```

The true and estimated portfolios are illustrated in @fig-StatisticalFrontier,
created with

```{r}
#| label: fig-StatisticalFrontier
#| echo: true
#| fig-cap: True and estimated portfolio frontiers
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sig.p, mu.p, pch=16, type="b", col="black",       
     ylim=c(-0.01, 0.041), xlim=c(0, 0.125),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex=1) 
abline(h=0) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# estimated assets and frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=2, cex = 1.5)
```

Due to the large estimation error in $\hat{\mu}_{A}$, the estimated
frontier is considerably higher than the true frontier. As a result,
portfolios on the estimated frontier appear to have higher reward-to-risk
properties than they actually do. 
\hfill{}$\blacksquare$
:::

Sampling uncertainty about the estimated frontier can be easily computed
using the bootstrap. The process is the same as in the case of a single
asset. For each bootstrap sample, estimate the expected return and
volatility of each portfolio on the frontier and then plot these bootstrap
pairs on the plot showing the estimated frontier from the sample returns.

::: {#exm-true-and-estimated-risk-portfolio-frontier-with-bootstrap-estimates-for-example-data .example name="True and estimated risk portfolio frontier with bootstrap estimates for example data"}

The R code to create $B=1000$ bootstrap estimates of risk and return
pairs for the frontier portfolios is

```{r}
#| echo: true

# initialize matrices 
n.boot = 1000 
mu.boot = matrix(0, n.boot, length(x.A)) 
sd.boot = matrix(0, n.boot, length(x.A)) 
colnames(mu.boot) = colnames(sd.boot) = 
      paste("P", 1:length(x.A), sep=".") 
# bootstrap loop 
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
# GWN model estimates   
   muhat.boot = colMeans(ret.boot)   
   sig2hat.boot = apply(ret.boot, 2, var)   
   sigmahat.boot = sqrt(sig2hat.boot)   
   covhat.boot = cov(ret.boot)[1,2]   
# portfolio risk return estimates   
   mu.boot[i, ] = x.A*muhat.boot[1] + x.B*muhat.boot[2]   
   sig2.boot = x.A^2 * sig2hat.boot[1] + 
      x.B^2 * sig2hat.boot[2] + 2*x.A*x.B*covhat.boot   
   sd.boot[i, ] = sqrt(sig2.boot) 
}
```

The true and estimated frontier together with the bootstrap risk-return
pairs for each portfolio on the estimated frontier is illustrated
in @fig-StatisticalFrontierBoot, create using

```{r}
#| label: fig-StatisticalFrontierBoot
#| echo: true
#| fig-cap: Estimated frontier with bootstrap estimates.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# set up plot area
plot(sig.p, mu.p, type="n", ylim=c(-0.02, 0.07), xlim=c(0, 0.13),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# plot bootstrap estimates 
for (i in 1:length(x.A)) {   
   points(sd.boot[, i], mu.boot[, i], pch=16, col="lightblue") 
}
# plot true frontier
points(sig.p, mu.p, pch=16, type="b", col="black", cex=1) 
abline(h=0) 
# plot true assets
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# plot estimated frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
# plot estimated assets
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=2, cex = 1.5)
```

The true frontier portfolios are the black dots, the estimated frontier
portfolios are the dark blue dots and the bootstrap risk-return pairs
are the light blue dots. The bootstrap estimates form a light blue
cloud around the estimated frontier. The bootstrap cloud can be interpreted
as approximating a confidence interval around estimated frontier.
From @eq-StatisticalJointCI, this confidence interval can
be thought of the union of all of the confidence ellipses about the
portfolios on the frontier. The true frontier of portfolios (black
dots) is within the blue cloud. 
\hfill{}$\blacksquare$
:::

### Statistical properties of the global minimum variance portfolio

For portfolios of two risky assets, the global minimum variance portfolio
weights satisfy
$$
x_{A}^{\min}=\frac{\sigma_{B}^{2}-\sigma_{AB}}{\sigma_{A}^{2}+\sigma_{B}^{2}-2\sigma_{AB}},~x_{B}^{\min}=1-x_{A}^{\min}=\frac{\sigma_{A}^{2}-\sigma_{AB}}{\sigma_{A}^{2}+\sigma_{B}^{2}-2\sigma_{AB}}.
$$ {#eq-StatisticalGlobalMin}
The expected return and variance of the global minimum variance portfolio
are
\begin{eqnarray*}
\mu_{p,\mathrm{min}} & = & x_{A}^{\mathrm{min}}\mu_{A}+x_{B}^{\mathrm{min}}\mu_{B},\\
\sigma_{p,\mathrm{min}}^{2} & = & \left(x_{A}^{\mathrm{min}}\right)^{2}\sigma_{A}^{2}+\left(x_{B}^{\mathrm{min}}\right)^{2}\sigma_{B}^{2}+2x_{A}^{\mathrm{min}}x_{B}^{\mathrm{min}}\sigma_{AB}.
\end{eqnarray*}

The estimated global minimum variance portfolio weights are then
$$
\hat{x}_{A}^{\min}=\frac{\hat{\sigma}_{B}^{2}-\hat{\sigma}_{AB}}{\hat{\sigma}_{A}^{2}+\hat{\sigma}{}_{B}^{2}-2\hat{\sigma}_{AB}},~x_{B}^{\min}=1-\hat{x}{}_{A}^{\min}=\frac{\hat{\sigma}{}_{A}^{2}-\hat{\sigma}{}_{AB}}{\hat{\sigma}_{A}^{2}+\hat{\sigma}_{B}^{2}-2\hat{\sigma}{}_{AB}}.
$$ {#eq-StatisticalGlobalMinHat}
From @eq-StatisticalGlobalMinHat you see that estimation error
in the global minimum variance weights is related to estimation error
in the asset variances and the covariance in a complicated nonlinear
way. However, the estimation error does not depend on estimation error
in the expected returns. The estimated expected returns and variance
of the global minimum variance portfolio are
$$
\hat{\mu}_{p,\mathrm{min}} = \hat{x}{}_{A}^{\mathrm{min}}\hat{\mu}_{A}+\hat{x}{}_{B}^{\mathrm{min}}\hat{\mu}_{B},
$$ {#eq-StatisticalGlobalMinMuHat}
$$
\hat{\sigma}_{p,\mathrm{min}}^{2} =  \left(\hat{x}{}_{A}^{\mathrm{min}}\right)^{2}\hat{\sigma}_{A}^{2}+\left(\hat{x}{}_{B}^{\mathrm{min}}\right)^{2}\hat{\sigma}_{B}^{2}+2\hat{x}{}_{A}^{\mathrm{min}}\hat{x}_{B}^{\mathrm{min}}\hat{\sigma}_{AB}
$$ {#eq-StatisticalGlobalMinSig2Hat}

Here, estimation error in $\hat{\mu}_{p,\mathrm{min}}$ depends on
two sources: estimation error in the global minimum variance weights
(which depend on estimation error in the asset variances and the covariance),
and estimation error in the asset expected returns. However, estimation
error in $\hat{\sigma}_{p,\mathrm{min}}^{2}$ only depends on estimation
error in the asset variances and the covariance. Because there is
more estimation error in the asset expected returns than the asset
variances $\hat{\mu}_{p,\mathrm{min}}$, will be estimated more imprecisely
than $\hat{\sigma}_{p,\mathrm{min}}^{2}$. 

::: {#exm-true-and-estimated-global-minimum-variance-portfolio-for-example-data .example name="True and estimated global minimum variance portfolio for example data"}

For the example data, the true and estimated global minimum variance
portfolio are:

```{r}
#| echo: true

# True global min variance portfolio weights
xA.min = (sig2.B - sig.AB)/(sig2.A + sig2.B - 2*sig.AB) 
xB.min = 1 - xA.min 
c(xA.min, xB.min)
# Estimated min variance portfolio weights
xA.min.hat = (sig2hat.vals[2] - covhat)/
             (sig2hat.vals[1] + sig2hat.vals[2] - 2*covhat) 
xB.min.hat = 1 - xA.min.hat 
c(xA.min.hat, xB.min.hat)
```

The estimated weights are close to the true weights. The true and
estimated expected return and volatility of the global minimum variance
portfolio are:

```{r}
#| echo: true

# Expected return and volatility of true global minimum variance portfolio 
mu.p.min = xA.min*mu.A + xB.min*mu.B 
sig2.p.min = xA.min^2 * sig2.A + xB.min^2 * sig2.B + 
               2*xA.min*xB.min*sig.AB 
sig.p.min = sqrt(sig2.p.min) 
c(mu.p.min, sig.p.min)
# Expected return and volatility of estimated global 
# minimum variance portfolio 
mu.p.min.hat = as.numeric(xA.min.hat*muhat.vals[1] + 
                             xB.min.hat*muhat.vals[2]) 
sig2.p.min.hat = xA.min.hat^2 * sig2hat.vals[1] + 
                 xB.min.hat^2 * sig2hat.vals[2] + 
                 2*xA.min.hat*xB.min.hat*covhat 
sig.p.min.hat = as.numeric(sqrt(sig2.p.min.hat)) 
c(mu.p.min.hat, sig.p.min.hat)
```

The estimated volatility of the global minimum variance portfolio
is close to the true volatility but the estimated expected return
is much larger than the true expected return. These portfolios are
illustrated in @fig-StatisticalGlobalMin.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-StatisticalGlobalMin
#| echo: false
#| fig-cap: True and estimated global minimum variance portfolios.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sig.p, mu.p, pch=16, type="b", col="black",       
ylim=c(-0.01, 0.041), xlim=c(0, 0.125),      
xlab=expression(sigma[p]), ylab=expression(mu[p]), cex=1) 
abline(h=0) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# estimated assets and frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
col="blue", pos=1, cex = 1.5) 
# true min variance portfolio 
points(sig.p.min, mu.p.min, pch=16, col="green", cex=2) 
text(x=sig.p.min, y=mu.p.min, labels="GMIN", col="green", pos=2, cex=1.5) 
# estimated min variance portfolio 
points(sig.p.min.hat, mu.p.min.hat, pch=16, col="orange", cex=2) 
text(x=sig.p.min.hat, y=mu.p.min.hat, labels="GMIN Hat", col="orange", pos=2, cex=1.5) 
```

The statistical properties of @eq-StatisticalGlobalMinHat,
@eq-StatisticalGlobalMinMuHat and @eq-StatisticalGlobalMinSig2Hat
are difficult to derive analytically. For example, suppose you would
like to evaluate the bias in the global minimum variance weights.
You would need to evaluate 
$$
E\left[\hat{x}_{A}^{\min}\right]=E\left[\frac{\hat{\sigma}_{B}^{2}-\hat{\sigma}_{AB}}{\hat{\sigma}_{A}^{2}+\hat{\sigma}{}_{B}^{2}-2\hat{\sigma}_{AB}}\right].
$$ {#eq-StatisticalGlobalMinBias}
Now, 
$$
E\left[\frac{\hat{\sigma}_{B}^{2}-\hat{\sigma}_{AB}}{\hat{\sigma}_{A}^{2}+\hat{\sigma}{}_{B}^{2}-2\hat{\sigma}_{AB}}\right]\neq\frac{E\left[\hat{\sigma}_{B}^{2}\right]-E\left[\hat{\sigma}_{AB}\right]}{E\left[\hat{\sigma}_{A}^{2}\right]+E\left[\hat{\sigma}{}_{B}^{2}\right]-2E\left[\hat{\sigma}_{AB}\right]}
$$
because $E[g(X)]\neq g(E[X])$ for nonlinear functions $g(\cdot)$.
Hence, @eq-StatisticalGlobalMinBias is extremely difficult
to evaluate analytically. Similarly, suppose you would like to compute
$\mathrm{se}(\hat{x}_{A}^{\min}).$ You would need to compute 
$$
\mathrm{var}\left(\hat{x}_{A}^{\min}\right)=\mathrm{var}\left(\frac{\hat{\sigma}_{B}^{2}-\hat{\sigma}_{AB}}{\hat{\sigma}_{A}^{2}+\hat{\sigma}{}_{B}^{2}-2\hat{\sigma}_{AB}}\right),
$$
which is a difficult and tedious calculation and can only be approximated
based on the CLT. 


The computations are even more difficult for evaluating bias and computing
standard errors for @eq-StatisticalGlobalMinMuHat and @eq-StatisticalGlobalMinSig2Hat.
For example, to evaluate the bias of $\hat{\mu}_{p,\mathrm{min}}$
you would need to calculate 
$$
E\left[\hat{\mu}_{p,\mathrm{min}}\right]=E\left[\hat{x}{}_{A}^{\mathrm{min}}\hat{\mu}_{A}+\hat{x}{}_{B}^{\mathrm{min}}\hat{\mu}_{B}\right]=E\left[\hat{x}{}_{A}^{\mathrm{min}}\hat{\mu}_{A}\right]+E\left[\hat{x}{}_{B}^{\mathrm{min}}\hat{\mu}_{B}\right].
$$ {#eq-StatisticalBiasGminMu}
Without knowing more about the joint distributions of the asset means
and weights you cannot simplify @eq-StatisticalBiasGminMu.
To compute the standard error of $\hat{\mu}_{p,\mathrm{min}}$ you
need to calculate
\begin{align*}
\mathrm{var}\left(\hat{\mu}_{p,\mathrm{min}}\right)&=\mathrm{var}\left(\hat{x}{}_{A}^{\mathrm{min}}\hat{\mu}_{A}+\hat{x}{}_{B}^{\mathrm{min}}\hat{\mu}_{B}\right)\\
&=\mathrm{var}(\hat{x}{}_{A}^{\mathrm{min}}\hat{\mu}_{A})+\mathrm{var}(\hat{x}{}_{B}^{\mathrm{min}}\hat{\mu}_{B})+2\mathrm{cov}\left(\hat{x}{}_{A}^{\mathrm{min}}\hat{\mu}_{A},\,\hat{x}{}_{B}^{\mathrm{min}}\hat{\mu}_{B}\right).
\end{align*}

Again, without knowing more about the joint distributions of the asset
means and estimated weights you cannot simplify @eq-StatisticalBiasGminMu.


Fortunately, statistical properties of @eq-StatisticalGlobalMinHat,
@eq-StatisticalGlobalMinMuHat and @eq-StatisticalGlobalMinSig2Hat
can be easily quantified using the bootstrap. For each bootstrap sample
you calculate these statistics and from the bootstrap distributions
you can then evaluate bias and compute standard errors and confidence
intervals. 

::: {#exm-true-and-estimated-global-minimum-variance-portfolio-with-bootstrap-estimates-for-example-data .example name="True and estimated global minimum variance portfolio with bootstrap estimates for example data"}

To compute $B=1000$ bootstrap estimates of @eq-StatisticalGlobalMinHat,
@eq-StatisticalGlobalMinMuHat and @eq-StatisticalGlobalMinSig2Hat
use:

```{r}
#| echo: true

# initialize matrices 
n.boot = 1000 
weights.boot = matrix(0, n.boot, 2) 
stats.boot = matrix(0, n.boot, 2) 
colnames(weights.boot) = names(mu.vec)  
colnames(stats.boot) = c("mu", "sigma") 
# bootstrap loop 
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
# GWN model estimates   
   muhat.boot = colMeans(ret.boot)   
   sig2hat.boot = apply(ret.boot, 2, var)   
   sigmahat.boot = sqrt(sig2hat.boot)   
   covhat.boot = cov(ret.boot)[1,2]   
# global minimum variance portfolio weights   
   weights.boot[i, 1] = (sig2hat.boot[2] - covhat.boot)/
                        (sig2hat.boot[1] + sig2hat.boot[2] - 
                            2*covhat.boot)   
   weights.boot[i, 2] = 1 - weights.boot[i, 1]   
# portfolio risk return estimates   
   stats.boot[i, "mu"] = weights.boot[i, 1]*muhat.boot[1] + 
                         weights.boot[i, 2]*muhat.boot[2]   
   sig2.boot = (weights.boot[i, 1]^2 * sig2hat.boot[1] + 
                weights.boot[i, 2]^2 * sig2hat.boot[2]                 
                + 2*weights.boot[i, 1]*weights.boot[i, 2]*covhat.boot)   
   stats.boot[i, "sigma"] = sqrt(sig2.boot) 
}
```

The bootstrap bias estimates for the global minimum variance portfolio
weights are:

```{r}
#| echo: true

colMeans(weights.boot) - c(xA.min.hat, xB.min.hat)
```

These values are close to zero suggesting that the estimated weights
are unbiased. Notice that the estimates are identical but opposite
in sign. This arises because the bootstrap estimates are perfectly
negatively correlated:

```{r}
#| echo: true

cor(weights.boot)[1,2]
```

The bootstrap standard error estimates for the weights are:

```{r}
#| echo: true

apply(weights.boot, 2, sd)
```

The standard errors estimates are identical and close to 0.05 which
is not too large. Figure @fig-StatisticalBootstrapWeights shows
histograms and normal QQ-plots for the bootstrap estimates of the
weights. These distributions are centered at the sample estimates
(white vertical lines) and look slightly asymmetric.


```{r}
#| label: fig-StatisticalBootstrapWeights
#| echo: false
#| fig-cap: Empirical distribution of bootstrap estimates of global minimum variance
#|   portfolio weights.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))   
   hist(weights.boot[, 1], main="Asset A", xlab="Asset A weights", col="slateblue1")   
   abline(v=xA.min.hat, col="white", lwd=2)   
   hist(weights.boot[, 2], main="Asset B", xlab="Asset B weights", col="slateblue1")   
   abline(v=xB.min.hat, col="white", lwd=2)   
   qqnorm(weights.boot[, 1])   
   qqline(weights.boot[, 1])   
   qqnorm(weights.boot[, 2])   
   qqline(weights.boot[, 2]) 
par(mfrow=c(1,1))
```


The bootstrap estimates of bias for the estimated expected return
and volatility of the global minimum variance portfolio are:

```{r}
#| echo: true

colMeans(stats.boot) - c(mu.p.min.hat, sig.p.min.hat)
```

These small values indicate that the estimates are roughly unbiased.
The bootstrap standard error estimates are:

```{r}
#| echo: true

apply(stats.boot, 2, sd)
```

Here, the bootstrap standard error estimate for $\hat{\mu}_{p,\mathrm{min}}$
is larger than the estimate for $\hat{\sigma}_{p,\mathrm{min}}$ indicating
more uncertainty about the mean of the global minimum variance portfolio
than the volatility of the portfolio. Interestingly, the bootstrap
standard error estimates are close to the naive analytical standard
error formulas based on the CLT that ignore estimation error in the
weights:

```{r}
#| echo: true

sig.p.min.hat/sqrt(n.obs) 
sig.p.min.hat/sqrt(2*n.obs)
```

That is, 
$$
\mathrm{se_{boot}}(\hat{\mu}_{p,\mathrm{min}})\approx\frac{\hat{\sigma}_{p,\mathrm{min}}}{\sqrt{T}},\,\mathrm{se_{boot}}\left(\hat{\sigma}_{p,\mathrm{min}}\right)\approx\frac{\hat{\sigma}_{p,\mathrm{min}}}{\sqrt{2\cdot T}}.
$$
The histograms and normal QQ-plots of the bootstrap estimates of $\mu_{p,min}$
and $\sigma_{p,min}$ are illustrated in @fig-StatisticalBootstrapGmin,
and look like normal distributions. 


```{r}
#| label: fig-StatisticalBootstrapGmin
#| echo: false
#| fig-cap: Empirical distribution of bootstrap estimates of $\mu_{p,min}$ and $\sigma_{p,min}$.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))   
   hist(stats.boot[, 1], main="mu", xlab="bootstrap estimates of mu", col="slateblue1")   
   abline(v=mu.p.min.hat, col="white", lwd=2)   
   hist(stats.boot[, 2], main="sigma", xlab="bootstrap estimates of sigma", col="slateblue1")   
   abline(v=sig.p.min.hat, col="white", lwd=2)   
   qqnorm(stats.boot[, "mu"])   
   qqline(stats.boot[, "mu"])   
   qqnorm(stats.boot[, "sigma"])   
   qqline(stats.boot[, "sigma"]) 
par(mfrow=c(1,1))
```

The sampling uncertainty in $\hat{\mu}_{p,\mathrm{min}}$ and $\hat{\sigma}_{p,\mathrm{min}}$
can be visualized by plotting the bootstrap estimates of $\mu_{p,min}$
and $\sigma_{p,min}$ on the risk return diagram, as shown in 
@fig-StatisticalBootstrapGminRiskReturn. Clearly there is much
more uncertainty about the location of $\mu_{p,min}$ than the location
of $\sigma_{p,min}$ . 
\hfill{}$\blacksquare$
:::

To sum up, for the global minimum variance portfolio you have roughly
unbiased estimates of the weights, expected return and variance. You
have a fairly precise estimate of volatility but an imprecise estimate
of expected return.


```{r}
#| label: fig-StatisticalBootstrapGminRiskReturn
#| echo: false
#| fig-cap: Estimation error in estimates of the global minimum variance portfolio.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sig.p, mu.p, pch=16, type="n", ylim=c(-0.01, 0.041), xlim=c(0, 0.125),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# bootstrap estimates 
points(stats.boot[, "sigma"], stats.boot[, "mu"], pch=16, col="lightgoldenrod2") 
# true frontier 
points(sig.p, mu.p, pch=16, type="b", col="black", cex=1) 
abline(h=0) 
points(sd.vec, mu.vec, pch=16, col="black", cex=2) 
text(x=sig.A, y=mu.A, labels="A", pos=3, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=3, cex = 1.5) 
# estimated assets and frontier 
points(sighat.p, muhat.p, type="b", pch=16, col="blue", cex=1) 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=3, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=1, cex = 1.5) 
# true min variance portfolio 
points(sig.p.min, mu.p.min, pch=16, col="green", cex=2) 
text(x=sig.p.min, y=mu.p.min, labels="GMIN", col="green", pos=2, cex=1.5) 
# estimated min variance portfolio 
points(sig.p.min.hat, mu.p.min.hat, pch=16, col="orange", cex=2) 
text(x=sig.p.min.hat, y=mu.p.min.hat, labels="GMIN Hat", col="orange", pos=2, cex=1.5) 
```


### Statistical properties of the Sharpe ratio 

Let $r_{f}$ denote the rate of return on a risk free asset with maturity
equal to the investment horizon of the return $R_{i}$ on risky asset
$i$. Recall from @sec-Introduction-to-Portfolio-Theory,
the Shapre ratio of asset $i$ is defined as 
$$
\mathrm{SR}_{i}=\frac{\mu_{i}-r_{f}}{\sigma_{i}},
$$
and is a measure of risk-adjusted expected return. Graphically in
the risk-return diagram, $\mathrm{S}\mathrm{R}_{i}$ is the slope
of a straight line from the risk-free rate that passes through the
point $(\sigma_{i},\mu_{i})$ and represents the risk-return tradeoff
of portfolios of the risk-free asset and the risky asset. The estimated
Sharpe ratio is 
$$
\widehat{\mathrm{SR}}_{i}=\frac{\hat{\mu}_{i}-r_{f}}{\hat{\sigma}_{i}},
$$
which inherits estimation error from $\hat{\mu}_{i}$ and $\hat{\sigma}_{i}$
in a nonlinear way. As a result, analytically computing the bias and
standard error for $\widehat{\mathrm{SR}}_{i}$ is complicated and
typically involves approximations based on the CLT. However, computing
numerical estimates of the bias and the standard error for $\widehat{\mathrm{SR}}_{i}$
using the bootstrap is simple. 


::: {#exm-true-and-estimated-sharpe-ratios-for-example-data .example name="True and estimated Sharpe ratios for example data"}

For the example data, assume a monthly risk-free rate of $r_{f}=0.03/12=0.0025.$
The true Sharpe ratios are:

```{r}
#| echo: true

r.f = 0.03/12
SR.true = (mu.vec - r.f)/sd.vec 
SR.true
```

Here, asset A has a much higher Sharpe ratio than asset B. For the
simulated data, the estimated Sharpe ratios are:

```{r}
#| echo: true

SR.hat = (muhat.vals - r.f)/sigmahat.vals 
SR.hat
```

For both assets the estimated Sharpe ratios are quite different from
the true Sharpe ratios and are highly misleading. The estimated Sharpe
ratio for asset A is more than two times larger than the true Sharpe
ratio and indicates a much higher risk adjusted preformance than is
actually available. For asset B, the estimated Sharpe ratio is slightly
negative when, in fact, the true Sharpe ratio is slightly positive.
The true and estimated Sharpe ratios are illustrated in Figure @fig-StatisticalSharpeRatios.
The slopes of the black lines are the true Sharpe ratios, and the
slopes of the blues lines are the estimated Sharpe ratios. 
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-StatisticalSharpeRatios
#| echo: false
#| fig-cap: True and estimated Sharpe ratios.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

plot(sd.vec, mu.vec, pch=16, col="black",       
     ylim=c(-0.01, 0.04), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p]), cex=2) 
abline(h=0) 
points(0, r.f, pch=16, col="green", cex=2) 
text(0, r.f, labels=expression(r[f]), cex=1.5, pos=2) 
abline(a=r.f, b=SR.true[1], col="black") 
abline(a=r.f, b=SR.true[2], col="black")
text(x=sig.A, y=mu.A, labels="A", pos=4, cex = 1.5) 
text(x=sig.B, y=mu.B, labels="B", pos=4, cex = 1.5) 
# estimates 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
col="blue", pos=4, cex = 1.5) 
abline(a=r.f, b=SR.hat[1], col="blue") 
abline(a=r.f, b=SR.hat[2], col="blue")
```



::: {#exm-delta-method-standard-errors-for-estimated-sharpe-ratio .example name="Delta method standard errors for estimated Sharpe ratio"}

Let $\boldsymbol{\theta} = (\mu, \sigma)'$. Then the Sharpe ratio can be considered as a function of $\boldsymbol{\theta}$: 

$$
\mathrm{SR}(\boldsymbol{\theta}) = \frac{\mu - r_f}{\sigma}
$$

Since $\hat{\boldsymbol{\theta}}$ is asymptotically normally distributed with covariance given by (), the delta method can be used to obtain an analytic standard error for $\widehat{\mathrm{SR}} = \mathrm{SR}(\hat{\boldsymbol{\theta}})$. It is straightforward to show (see end-of-chapter exercises) that the gradient function is: 

$$
\mathbf{g}(\boldsymbol{\theta}) = \frac{\partial \mathrm{SR}(\boldsymbol{\theta})}{\partial \boldsymbol{\theta}} = \left( \begin{array}{c} 
\frac{-1}{\sigma} \\ 
-(\sigma)^{-2}(\mu - r_f)
\end{array} \right).
$$

The delta method variance is then:

$$
\mathbf{g}(\boldsymbol{\theta})'\mathrm{var}(\hat{\boldsymbol{\theta}})\mathbf{g}(\boldsymbol{\theta}) = \frac{1}{T}\left( 1 + \frac{1}{2}\mathrm{SR}^2 \right ) = \mathrm{se}(\widehat{\mathrm{SR}})^2,
$$

The corresponding standard error is:

$$
\mathrm{se}(\widehat{\mathrm{SR}}) = \sqrt{\frac{1}{T}\left( 1 + \frac{1}{2}\mathrm{SR}^2 \right ) }
$$ {#eq-se-Sharpe-ratio}

Notice that $\mathrm{se}(\widehat{\mathrm{SR}})$ increases with $\mathrm{SR}^2$ and decreases with $T$. The practically useful estimated standard error replace the unknown value of $\mathrm{SR}$ with its estimated value $\widehat{\mathrm{SR}}$ and is given by

$$
\mathrm{\widehat{se}}(\widehat{\mathrm{SR}}) = \sqrt{\frac{1}{T}\left( 1 + \frac{1}{2}\widehat{\mathrm{SR}}^2 \right ) } 
$$ {#eq-estimated-se-Sharpe-ratio}
\hfill{}$\blacksquare$
:::


::: {#exm-applying-the-analytic-delta-method-to-the-sharpe-ratio .example name="Applying the analytic delta method to the Sharpe Ratio"}

The estimated monthly SRs for assets A and B (using $r_f = 0.03/12$), their standard errors, and 95% confidence intervals are computed using:

```{r}
seSR.hat = sqrt((1/n.obs)*(1 + 0.5*(SR.hat^2)))
lower.SR = SR.hat - 2*seSR.hat
upper.SR = SR.hat + 2*seSR.hat
width.SR = upper.SR - lower.SR
ans = rbind(SR.hat, seSR.hat, lower.SR, upper.SR, width.SR)
colnames(ans) = c("Asset A", "Asset B")
rownames(ans) = c("Estimate", "Std Error", "Lower", "Upper", "Width")
ans
```
The estimated monthly SRs for assets A and B are close to zero, with a large estimated standard errors and a 95\% confidence interval containing both negative and positive values. Clearly, the SR is not estimated very well.
\hfill{}$\blacksquare$
:::

::: {#exm-applying-the-numerical-delta-method-to-the-sharpe-ratio .example name="Applying the numerical delta method to the Sharpe Ratio"}

I first create the named inputs $\hat{\boldsymbol{\theta}}$ and $\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})$ from the GWN model for simple returns:

```{r}
thetahat = c(muhat.vals[1], sigmahat.vals[1])
names(thetahat) = c("mu", "sigma")
var.thetahat = matrix(c(se.muhat[1]^2,0,0,se.sigmahat[1]^2), 2, 2, byrow=TRUE)
rownames(var.thetahat) = colnames(var.thetahat) = names(thetahat)
thetahat
var.thetahat
```

I use the **car** function `deltaMethod()` to compute the numerical delta method for the estimated Sharpe ratio $\mathrm{SR}(\hat{\boldsymbol{\theta}})$ for asset A as follows:

```{r}
dmSR.A = deltaMethod(thetahat, "(mu-r.f)/sigma", var.thetahat)
dmSR.A
```
The numerical delta method estimated standard error exactly matches the analytic standard error. The 95\% confidence interval returned by the function `deltaMethod()` uses the formula $\hat{\theta} \pm 1.96 \times \hat{\mathrm{se}}(\hat{\theta})$ as so is slightly different than the analytic result given above.  Using similar computations, The numerical delta method for asset B is

```{r}
#| echo: false
thetahat = c(muhat.vals[2], sigmahat.vals[2])
names(thetahat) = c("mu", "sigma")
var.thetahat = matrix(c(se.muhat[2]^2,0,0,se.sigmahat[2]^2), 2, 2, byrow=TRUE)
rownames(var.thetahat) = colnames(var.thetahat) = names(thetahat)
dmSR.B = deltaMethod(thetahat, "(mu-r.f)/sigma", var.thetahat)
dmSR.B
```
\hfill{}$\blacksquare$
:::

::: {#exm-delta-method-for-joint-distribution-of-two-shapre-ratios .example name="Delta method for joint distribution of two Shapre ratios"}

Let $\boldsymbol{\theta}$ = $(\mu_1,\mu_2,\sigma_1,\sigma_2)'$ and define the $2 \times 1$ vector of Sharpe ratios:

$$
\boldsymbol{\eta} = f(\boldsymbol{\theta}) = \begin{pmatrix} \frac{\mu_A - r_f}{\sigma_A} \\
\frac{\mu_B - r_f}{\sigma_B} \end{pmatrix} = \begin{pmatrix} \mathrm{SR}_A(\boldsymbol{\theta}) \\
\mathrm{SR}_B(\boldsymbol{\theta}) \end{pmatrix},
$$

so that $\eta_A = \mathrm{SR}_A(\boldsymbol{\theta})$ and $\eta_B = \mathrm{SR}_B(\boldsymbol{\theta})$ where $r_f$ denotes the constant risk-free rate. Consider the $\boldsymbol{\theta} =(\mu_A,\mu_B, \sigma_A, \sigma_B)'$ and  $\boldsymbol{\eta} = \mathbf{\mathrm{SR}}(\boldsymbol{\theta})=(\mathrm{SR}_A(\boldsymbol{\theta}), \mathrm{SR}_B(\boldsymbol{\theta}))'$ where
$$
\mathrm{SR}_A(\boldsymbol{\theta}) = \frac{\mu_A - r_f}{\sigma_A}, ~ \mathrm{SR}_B(\boldsymbol{\theta}) = \frac{\mu_B - r_f}{\sigma_B}.
$$

Here, I want to use the delta method to get the joint distribution of $\boldsymbol{\eta}$. Now, using the chain-rule:

\begin{align*}
\frac{\partial \mathrm{SR}_A(\boldsymbol{\theta})}{\partial \mu_A} & = \sigma_A^{-1}, ~ \frac{\partial \mathrm{SR}_A(\boldsymbol{\theta})}{\partial \mu_B} = 0, ~ & \frac{\partial \mathrm{SR}_A(\boldsymbol{\theta})}{\partial \sigma_1} = -\sigma_1^{-2}(\mu_1 - r_f), ~ & \frac{\partial \mathrm{SR}_A(\boldsymbol{\theta})}{\partial \sigma_2} = 0, \\
\frac{\partial \mathrm{SR}_B(\boldsymbol{\theta})}{\partial \mu_1} & = 0, ~ \frac{\partial \mathrm{SR}_B(\boldsymbol{\theta})}{\partial \mu_2} = \sigma_2^{-1}, ~ & \frac{\partial \mathrm{SR}_B(\boldsymbol{\theta})}{\partial \sigma_1} = 0, ~ & \frac{\partial \mathrm{SR}_B(\boldsymbol{\theta})}{\partial \sigma_2} = -\sigma_1^{-2}(\mu_1 - r_f).
\end{align*}

Then, using $\mathrm{SR}_{i} = (\mu_i - r_f)/\sigma_i ~ (i=A,B)$, I have:

$$
\mathbf{G}(\boldsymbol{\theta})' = \begin{pmatrix} \sigma_A^{-1} & 0 & -\sigma_A^{-1} \mathrm{SR}_A & 0 \\ 0 & \sigma_B^{-1} & 0 & -\sigma_B^{-1}\mathrm{SR}_B & 0
\end{pmatrix}
$$

Using Proposition 7.10 and the results from the previous example I have:

$$
\mathrm{var}(\hat{\boldsymbol{\theta}}) = \frac{1}{T} \begin{pmatrix} \sigma_A^2 & \sigma_{AB} & 0 & 0 \\
\sigma_{AB} & \sigma_B^2 & 0 & 0 \\
0 & 0 & \frac{1}{2}\sigma_A^2 & \frac{1}{2} \sigma_{AB}^2 \sigma_A^{-1} \sigma_B^{-1} \\
0 & 0 & \frac{1}{2} \sigma_{AB}^2 \sigma_A^{-1} \sigma_B^{-1} & \frac{1}{2}\sigma_B^2 \end{pmatrix}.
$$

Then, after some straightforward matrix algebra calculations (see end-of-chapter exercises)
I get:

$$
\begin{aligned}
\widehat{\mathrm{var}}(\hat{\eta}) & = \mathbf{G}(\hat{\boldsymbol{\theta}})'\widehat{\mathrm{var}}(\hat{\boldsymbol{\theta}})\mathbf{G}(\hat{\boldsymbol{\theta}}) \\
&= \begin{pmatrix} \widehat{\mathrm{var}}(\widehat{\mathrm{SR}}_A) & \widehat{\mathrm{cov}}(\widehat{\mathrm{SR}}_A, \widehat{\mathrm{SR}}_B) \\ \widehat{\mathrm{cov}}(\widehat{\mathrm{SR}}_A, \widehat{\mathrm{SR}}_B) & \widehat{\mathrm{var}}(\widehat{\mathrm{SR}}_B) \end{pmatrix} \\
& = \begin{pmatrix} \frac{1}{T}(1+\frac{1}{2}\widehat{\mathrm{SR}}_A^2) & \frac{\hat{\rho}_{AB}}{T}(1 + \frac{1}{2}\hat{\rho}_{AB}\widehat{\mathrm{SR}}_A \widehat{\mathrm{SR}}_B) \\ \frac{\hat{\rho}_{AB}}{T}(1 + \frac{1}{2}\hat{\rho}_{AB} \widehat{\mathrm{SR}}_A\widehat{\mathrm{SR}}_B) & \frac{1}{T}(1+\frac{1}{2}\widehat{\mathrm{SR}}_B^2) \end{pmatrix}.
\end{aligned}
$$

The diagonal elements match the delta method variances for the estimated Sharpe ratio derived earlier. The off-diagonal covariance term shows that the two estimated Sharpe ratios are correlated. You explore this correlation in more depth in the end-of-chapter exercises. 
\hfill{}$\blacksquare$
:::

::: {#exm-jackknife-standard-error-for-estimated-sharpe-ratio .example name="Jackknife standard error for estimated Sharpe ratio"}

I use the **bootstrap** function `jackknife()` to compute the jacknife estimated standard for the Sharpe ratio for assets A and B:

```{r}
theta.SR = function(x, xdata, r.f) {
  mu.hat = mean(xdata[x, ])
  sigma.hat = sd(xdata[x, ])
  fun.hat = (mu.hat-r.f)/sigma.hat
  fun.hat
}
SR.jack = jackknife(1:n.obs, theta=theta.SR, 
                    returns.sim[,"asset.A",drop=FALSE], r.f)
ans = cbind(SR.jack$jack.se, dmSR.A$SE)
colnames(ans) = c("Jackknife", "Delta Method")
rownames(ans) = "SR.A"
ans
```

```{r}
#| echo: false
theta.SR = function(x, xdata, r.f) {
  mu.hat = mean(xdata[x, ])
  sigma.hat = sd(xdata[x, ])
  fun.hat = (mu.hat-r.f)/sigma.hat
  fun.hat
}
SR.jack = jackknife(1:n.obs, theta=theta.SR, 
                    returns.sim[,"asset.B",drop=FALSE], r.f)
ans = cbind(SR.jack$jack.se, dmSR.B$SE)
colnames(ans) = c("Jackknife", "Delta Method")
rownames(ans) = "SR.B"
ans
```

\hfill{}$\blacksquare$
:::

::: {#exm-bootstraped-sharpe-ratios-for-example-data .example name="Bootstraped Sharpe ratios for example data"}

To illustrate the magnitude of the estimation errors in the estimated
Sharpe ratios, @fig-StatisticalBootstrapSRplot shows
500 bootstrap estimates of the risk-return pairs $(\sigma_{A},\mu_{A})$
and $(\sigma_{B},\mu_{B})$ along with the maximum and minimum bootstrap
Sharpe ratio estimates for each asset. The bootstrap Sharpe ratio
estimates are computed using:

```{r}
#| echo: true

n.boot = 500 
mu.boot = matrix(0, n.boot, 2) 
sd.boot = matrix(0, n.boot, 2) 
SR.boot = matrix(0, n.boot, 2) 
colnames(mu.boot) = colnames(sd.boot) = colnames(SR.boot) = c("Asset.A", "Asset.B")
set.seed(123) 
for (i in 1:n.boot) {   
   boot.idx = sample(n.obs, replace=TRUE)   
   ret.boot = returns.sim[boot.idx, ]    
   mu.boot[i, ] = colMeans(ret.boot)   
   sd.boot[i, ] = apply(ret.boot, 2, sd)    
   SR.boot[i, ] = (mu.boot[i, ] - r.f)/sd.boot[i, ] 
} 
# find index of max and min bootstrap Sharpe ratios
maxSRidx.A = which(SR.boot[, 1] == max(SR.boot[, 1])) 
minSRidx.A = which(SR.boot[, 1] == min(SR.boot[, 1])) 
maxSRidx.B = which(SR.boot[, 2] == max(SR.boot[, 2])) 
minSRidx.B = which(SR.boot[, 2] == min(SR.boot[, 2]))
```

```{r}
#| label: fig-StatisticalBootstrapSRplot
#| echo: false
#| fig-cap: Bootstrap risk-return estimates with maximum and minimum Sharpe ratios.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

# layout plot
plot(sd.vec, mu.vec, type="n",  ylim=c(-0.01, 0.06), xlim=c(0, 0.1),      
     xlab=expression(sigma[p]), ylab=expression(mu[p])) 
# bootstrap estimates 
points(sd.boot[, "Asset.A"], mu.boot[, "Asset.A"], pch=16, col="lightblue") 
points(sd.boot[, "Asset.B"], mu.boot[, "Asset.B"], pch=16, col="grey")
abline(h=0) 
# sample estimates 
points(sigmahat.vals, muhat.vals, pch=16, col="blue", cex=2) 
text(x=sigmahat.vals[1], y=muhat.vals[1], labels="A Hat",       
     col="blue", pos=4, cex = 1.5) 
text(x=sigmahat.vals[2], y=muhat.vals[2], labels="B Hat",       
     col="blue", pos=4, cex = 1.5)
# Min and max Sharpe ratios 
abline(a=r.f, b=SR.boot[maxSRidx.A, "Asset.A"], col="lightblue", lwd=2) 
abline(a=r.f, b=SR.boot[minSRidx.A, "Asset.A"], col="lightblue", lwd=2) 
abline(a=r.f, b=SR.boot[maxSRidx.B, "Asset.B"], col="grey", lwd=2) 
abline(a=r.f, b=SR.boot[minSRidx.B, "Asset.B"], col="grey", lwd=2) 
# Risk free rate 
points(0, r.f, pch=16, col="green", cex=2) 
text(0, r.f, labels=expression(r[f]), cex=1.5, pos=2) 
```

The maximum and minimum bootstrap Sharpe ratio estimates for asset
B are the two grey straight lines from the risk-free rate that have
the highest and lowest slopes, respectively. The maximum and minimum
bootstrap Sharpe ratio estimates for asset A are the two light blue
lines with the highest and lowest slopes, respectively. These maximum
and minimum estimates are:

```{r}
#| echo: true

apply(SR.boot, 2, max) 
apply(SR.boot, 2, min)
```

For each asset the difference in the maximum and minimum slopes is
substantial and indicates a large about of estimation error in the
estimated Sharpe ratios. 


The statistical properties of the estimated Sharpe ratios can be estimated
from the bootstrap samples. The bootstrap estimates of bias are:

```{r}
#| echo: true

# bias
biasSR.boot = colMeans(SR.boot) - c(SR.true) 
biasSR.boot 
# bias relative to true values
biasSR.boot/SR.true
```

Here you see there is a substantial upward bias in $\widehat{\mathrm{SR}}_{A}$
(bias relative to true Sharpe ratio is 141\%) and a substantial downward
bias in $\widehat{\mathrm{SR}}_{B}$ (bias relative to true Sharpe
ratio is -105\%). A crude bias adjusted Sharpe ratio estimate, which
substracts the estimated bias from the sample estimate, gives results
much closer to the true Sharpe ratios:

```{r}
#| echo: true

SR.hat.biasAdjust = SR.hat - biasSR.boot 
SR.hat.biasAdjust
SR.true
```

The bootstrap estimated standard errors for both assets are big and
indicate that the Sharpe ratios are not estimated well:

```{r}
#| echo: true

apply(SR.boot, 2, sd)
```

@fig-StatisticalBootstrapSR shows the histograms and
normal QQ-plots of the bootstrap distributions for the estimated Sharpe
ratios. The histograms for both assets show the wide dispersion of
the bootstrap Sharpe ratio estimates that was illustrated in 
@fig-StatisticalBootstrapSRplot. The histogram and QQ-plot for
asset A shows a positive skewness whereas the histogram and QQ-plot
for asset B show a more symmetric distribution. The quantile-based
95\% confidence intervals for the true Sharpe ratios are:

```{r}
#| echo: true

# 95% CI for SR.A
quantile(SR.boot[, 1], probs=c(0.025, 0.975)) 
# 95% CI for SR.B
quantile(SR.boot[, 2], probs=c(0.025, 0.975))
```

These intervals are quite large and indicate much uncertainty about
the true Sharpe ratio values.
\hfill{}$\blacksquare$
:::

```{r}
#| label: fig-StatisticalBootstrapSR
#| echo: false
#| fig-cap: Bootstrap distributions of estimated Sharpe ratios.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center

par(mfrow=c(2,2))   
  hist(SR.boot[, 1], main="Asset A", xlab="bootstrap estimates of SR", col="slateblue1")   
  abline(v=SR.hat[1], col="white", lwd=2)   
  hist(SR.boot[, 2], main="Asset B", xlab="bootstrap estimates of SR", col="slateblue1")   
  abline(v=SR.hat[2], col="white", lwd=2)   
  qqnorm(SR.boot[, 1])   
  qqline(SR.boot[, 1])   
  qqnorm(SR.boot[, 2])   
  qqline(SR.boot[, 2]) 
par(mfrow=c(1,1))
```


## Hypothesis Testing for Sharpe Ratio
To be completed

## Rolling Analysis of Portfolios
To be completed...

## Further Reading: Statistical Analysis of Portfolios

-  Mention other quadratic programming packages: ROI etc. See Doug's book
-  Mention Pfaff's R book
-  Mention Grilli's book etc.
-  Broadie, M. (1993). "Computing efficient frontiers using estimated parameters", Annals of Operations Research 45, 21-58. Describes the logic behind actual frontiers
- Jorion, P. (1992). "Portfolio Optimization in Practice. Financial Analysts Journal", 48(1), 68–74.
-  Michaud (2008), *Efficient Asset Management: A Practical Guide to Stock Portfolio Optimization and Asset Allocation*, Second Edition. 
To be completed...

## Exercises

::: {#exr-01 .exercise}
Show that $\hat{\sigma}_{p}^{2}$ in @eq-StatisticalGWNestimatesPortfolioVar
with $\hat{\sigma}_{p}^{2}$ in @eq-StatisticalGWNestimatesPortfolio are the same when the sample covariance matrix is used in @eq-StatisticalGWNestimatesPortfolioVar.
:::


{Ruppert} Ruppert, D.A.. \textit{Statistics and Data Analysis for Financial Engineering}, Springer-Verlag. 

{MartinSchererYollin}Martin, R.D., Scherer, B., and Yollin, G. (2016). 



