# Estimation of the Single Index Model {#sec-EstimationSingle-Index-Model}

Updated: September 19, 2026

Copyright © Eric Zivot 2026


## Introduction

To be completed

Notes:

1. For consistency follow the chapter format of the GWN model
2. Covers estimation and hypothesis testing

Outline

1. Estimation using Method of Moments
2. Estimation using Least Squares
3. Statistical properties of method of moments/LS estimates
4. Linear regression in R
5. Examples, examples, examples.
6.  Hypothesis Testing in the Single Index Model
    a.  Tests for coefficients
    b.  Tests for model assumptions: check to see if residual correlation matrix is diagonal!
    c.  Diagnostics for covariance stationarity
    d.  Rolling estimates from the single index model
7. Single Index Model and Portfolio Theory
    a.  portfolio optimization using SI covariance matrix   
        i.  always pd, reduces number of estimated components in matrix.
        ii.  Compare results with sample covariance matrix - in estimation chapter
    b.  Risk analysis of asset and portfolios: factor risk reports


The R packages used in this chapter are **corrplot**, **introcompfinr**, **xts**, and
**PerformanceAnalytics**. 

```{r}
#| include: false

suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(introcompfinr))
suppressPackageStartupMessages(library(xts))
suppressPackageStartupMessages(library(PerformanceAnalytics))
options(digits=3)
```


## Data for examples

I use the example data from \@ref() for the examples in this chapter. The data are the
monthly adjusted closing prices on four Northwest stocks (with tickers
in parentheses): Boeing (BA), Nordstrom (JWN), Microsoft (MSFT), Starbucks
(SBUX), and the S\&P 500 index (SPY, proxy for the market index). These prices are extracted from the **introcompfinr** package as follows:

```{r}
#| echo: true
baPrices = to.monthly(baDailyPrices, OHLC=FALSE) 
jwnPrices = to.monthly(jwnDailyPrices, OHLC=FALSE) 
msftPrices = to.monthly(msftDailyPrices, OHLC=FALSE) 
sbuxPrices = to.monthly(sbuxDailyPrices, OHLC=FALSE) 
sp500Prices = to.monthly(sp500DailyPrices, OHLC=FALSE) 
siPrices = merge(baPrices, jwnPrices, msftPrices, 
                 sbuxPrices, sp500Prices)
```

The data sample is from January, 1998 through May, 2012 and returns
are simple returns:

```{r}
#| echo: true
smpl = "1998-01::2012-05" 
siPrices = siPrices[smpl] 
siRetS = na.omit(Return.calculate(siPrices, method="simple")) 
```

## Estimation of SI Model

Consider the SI model @eq-SI-SIreturn - @eq-SI-errorAssets.
The asset specific parameters to be estimated are $\beta_{i,0}$, $\beta_{i,M}$
and $\sigma_{\varepsilon,i}^{2}$, $(i=1,\ldots,N)$, and the market
parameters to be estimated are $\mu_{M}$ and $\sigma_{M}^{2}$. These
parameters can be estimated using the plug-in principle, linear regression,
and maximum likelihood. All methods give essentially the same estimators
for the SI model parameters.

### Plug-in principle estimates

Let $\{(r_{it},r_{Mt})\}_{t=1}^{T}$ denote a sample of size $T$
of observed returns on asset $i$ and the market return which are
assumed to be generated from the SI model. @eq-SI-SIreturn
- @eq-SI-SIdistribution. Recall, the plug-in principle says
to estimate population model parameters using appropriate sample statistics.
For the market parameters, the plug-in principle estimates are the
same as the GWN model estimates
\begin{eqnarray*}
\hat{\mu}_{M} & = & \frac{1}{T}\sum_{t=1}^{T}r_{Mt},\\
\hat{\sigma}_{M}^{2} & = & \frac{1}{T-1}\sum_{t=1}^{T}(r_{Mt}-\hat{\mu}_{M})^{2}.
\end{eqnarray*}
From @eq-SI-SIintercept - @eq-SI-marketBeta we see
that $\beta_{i,0}$ and $\beta_{i,M}$ are functions of population parameters
\begin{eqnarray*}
\beta_{i,0} & = & \mu_{i}-\beta_{i,M}\mu_{M},\\
\beta_{i,M} & = & \frac{\mathrm{cov}(R_{it},R_{Mt})}{\mathrm{var}(R_{Mt})}=\frac{\sigma_{iM}}{\sigma_{M}^{2}}.
\end{eqnarray*}
The corresponding plug-in principle estimates are then:
$$
\hat{\beta}_{i,0}  =  \hat{\mu}_{i}-\hat{\beta}_{i,M}\hat{\mu}_{M},
$$ {#eq-SI-pluginAlpha}
$$
\hat{\beta}_{i,M} = \frac{\hat{\sigma}_{iM}}{\hat{\sigma}_{M}^{2}},
$$ {#eq-SI-pluginBeta}
where 
\begin{eqnarray*}
\hat{\mu}_{i} & = & \frac{1}{T}\sum_{t=1}^{T}r_{it},\\
\hat{\sigma}_{iM} & = & \frac{1}{T-1}\sum_{t=1}^{T}(r_{it}-\hat{\mu}_{i})(r_{Mt}-\hat{\mu}_{M}).
\end{eqnarray*}

Given the plug-in principle estimates $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$,
the plug-in principle estimate of $\varepsilon_{it}$ is 
$$
\hat{\varepsilon}_{it}=r_{it}-\hat{\beta}_{i,0}-\hat{\beta}_{i,M}r_{Mt},\,t=1,\ldots,T.
$$ {#eq-SI-pluginError}
Using @eq-SI-pluginError, the plug-in principle estimate of
$\sigma_{\varepsilon,i}^{2}$ is the sample variance of $\{\hat{\varepsilon}_{it}\}_{t=1}^{T}$
(adjusted for the number of degrees of freedom):
$$
\hat{\sigma}_{\varepsilon,i}^{2} =\frac{1}{T-2}\sum_{t=1}^{T}\hat{\varepsilon}_{t}^{2}=\frac{1}{T-2}\sum_{t=1}^{T}\left(r_{it}-\hat{\beta}_{i,0}-\hat{\beta}_{i,M}r_{Mt}\right)^{2}.
$$ {#eq-SI-pluginErrorVariance}
Plug-in principle estimates of $R^{2}$ based on @eq-SI-R2
can be computed using
$$
\hat{R}^{2} =\frac{\hat{\beta}_{i}^{2}\hat{\sigma}_{M}^{2}}{\hat{\sigma}_{i}^{2}}=1-\frac{\hat{\sigma}_{\varepsilon,i}^{2}}{\hat{\sigma}_{i}^{2}}.
$$ {#eq-SI-pluginR2}

::: {#exm-computing-plug-in-principle-estimators-for-si-model-parameters .example name="Computing plug-in principle estimators for SI model parameters"}

Consider computing the plug-in principle estimates for $\beta_{i,0},$
$\beta_{i,M}$ and $\sigma_{\varepsilon,i}^{2}$ from the example data
using the formulas @eq-SI-pluginAlpha, @eq-SI-pluginBeta
and @eq-SI-pluginErrorVariance, respectively. First, extract
the sample statistics $\hat{\mu}_{i}$, $\hat{\sigma}_{iM}$, $\hat{\mu}_{M}$,
and $\hat{\sigma}_{M}^{2}$:

```{r}
#| echo: true
assetNames = colnames(siRetS)[1:4] 
muhat = colMeans(siRetS) 
Sigmahat = var(siRetS)
sig2hat = diag(Sigmahat)
covAssetsSp500 = Sigmahat[assetNames, "SP500"]
```

Next, estimate $\hat{\beta}_{i,M}$ using

```{r}
#| echo: true
betaMHat = covAssetsSp500/sig2hat["SP500"]
betaMHat
```

Here, we see that $\hat{\beta}_{BA,M}$ and $\hat{\beta}_{SBUX,M}$ are
very close to one and that $\hat{\beta}_{JWN,M}$ and $\hat{\beta}_{MSFT,M}$
are slightly bigger than one. Using the estimates of $\hat{\beta}_{i,M}$
and the sample statistics $\hat{\mu}_{i}$ and $\hat{\mu}_{M}$ the
estimates for $\hat{\beta}_{i,0}$ are

```{r}
#| echo: true
beta0Hat = muhat[assetNames] - betaMHat*muhat["SP500"] 
beta0Hat
```

All of the estimates of $\hat{\beta}_{i,0}$ are close to zero. 

The estimates of $\sigma_{\varepsilon,i}^{2}$ and $\sigma_{\varepsilon,i}$ can be computed using:

```{r}
#| echo: true
sig2eHat = rep(0, length(assetNames)) 
names(sig2eHat) = assetNames 
for (aName in assetNames) {   
     eHat = siRetS[, aName] - beta0Hat[aName] - betaMHat[aName]*siRetS[, "SP500"]   
     sig2eHat[aName] = crossprod(eHat)/(length(eHat) - 2) 
} 
sigeHat = sqrt(sig2eHat)
ans = rbind(sig2eHat,sigeHat)
ans
```
Here, Nordstrom (JWN) has the largest asset specific volatility and Boeing (BA) has the lowest. 

Lastly, the estimates of $R^{2}$ can be computed using

```{r}
#| echo: true

R2 = 1 - sig2eHat/sig2hat[assetNames] 
R2
```
Microsoft has the largest fraction of returns explained by the market index and Starbucks has the lowest. 

The fitted SI model is visualized in the figure below created using:
```{r}
#| echo: true

```



\hfill{}$\blacksquare$
:::

### Least squares estimates

The SI model representation @eq-SI-SIreturn shows that returns
are a linear function of the market return and an asset specific error
term
$$
R_{it}=\beta_{i,0}+\beta_{i,M}R_{Mt}+\varepsilon_{it},
$$
here $\beta_{i,0}$ is the intercept and $\beta_{i,M}$ is the slope.
*Least squares regression*\index{Least squares regression} is a method for estimating $\beta_{i,0}$ and $\beta_{i,M}$ by finding the "best fitting" line to the scatterplot
of returns where $R_{it}$ is on the vertical axis and $R_{Mt}$ is
on the horizontal axis.

To see how the method of least squares determines the "best fitting"
line, consider the scatterplot of the sample returns on Boeing and
the S\&P 500 index illustrated in Figure xxx. In the figure, the black
line is a fitted line with initial guess $\hat{\beta}_{BA,0}=0$ and
$\hat{\beta}_{BA,M}=0.5.$ The differences between the observed returns
(blue dots) and the values on the fitted line are the estimated errors
$\hat{\varepsilon}_{BA,t}=r_{BA,t}-\hat{\beta}_{BA,0}-\hat{\beta}_{BA,M}r_{Mt}=r_{BA,t}-0-0.5\times R_{Mt}.$
Some estimated errors are big and some are small. The overall fit
of the line can be measured using a statistic based on all $t=1,\ldots,T$
of the estimated errors. A natural choice is the sum of the errors
$\sum_{t=1}^{T}\hat{\varepsilon}_{t}$. However, this choice can be misleading
due to the canceling out of large positive and negative errors. To
avoid this problem, it is better to measure the overall fit using
$\sum_{t=1}^{T}\hat{|\varepsilon}_{t}|$ or $\sum_{t=1}^{T}\hat{\varepsilon}_{t}^{2}.$
Then the best fitting line can be determined by finding the intercept
and slope values that minimize $\sum_{t=1}^{T}\hat{|\varepsilon}_{t}|$
or $\sum_{t=1}^{T}\hat{\varepsilon}_{t}^{2}.$ 


The method of least squares regression defines the "best fitting"
line by finding the intercept and slope values that minimize the sum
of squared errors
$$ 
\mathrm{SSE}(\hat{\beta}_{i,0},\hat{\beta}_{i,M})=\sum_{t=1}^{T}\hat{\varepsilon}_{it}^{2}=\sum_{t=1}^{T}\left(r_{it}-\hat{\beta}_{i,0}-\hat{\beta}_{i,M}r_{Mt}\right)^{2}.
$$ {#eq-SI-SSE}
Because $\mathrm{SSE}(\hat{\beta}_{i,0},\hat{\beta}_{i,M})$ is a continuous
and differential function of $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$,
the minimizing values of $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$
can be determined using simple calculus. The first order conditions
for a minimum are:
$$
0 =\frac{\partial\mathrm{SSE}(\hat{\beta}_{i,0},\hat{\beta}_{i,M})}{\partial\hat{\beta}_{i,0}}=-2\sum_{t=1}^{T}(r_{it}-\hat{\beta}_{i,0}-\hat{\beta}_{i,M}r_{Mt})=-2\sum_{t=1}^{T}\hat{\varepsilon}_{it},
$$ {#eq-SI-normalEqn1}
$$
0 =\frac{\partial\mathrm{SSE}(\hat{\beta}_{i,0},\hat{\beta}_{i,M})}{\partial\hat{\beta}_{i,M}}=-2\sum_{t=1}^{T}(r_{it}-\hat{\beta}_{i,0}-\hat{\beta}_{i,M}r_{Mt})r_{Mt}=-2\sum_{t=1}^{T}\hat{\varepsilon}_{it}r_{Mt}.
$$ {#eq-SI-normalEqn2}
These are two linear equations in two unknowns which can be re-expressed
as
\begin{eqnarray*}
\hat{\beta}_{i,0}T+\hat{\beta}_{i,M}\sum_{t=1}^{T}r_{Mt} & = & \sum_{t=1}^{T}r_{it},\\
\hat{\beta}_{i,0}\sum_{t=1}^{T}r_{Mt}+\hat{\beta}_{i,M}\sum_{t=1}^{T}r_{Mt}^{2} & = & \sum_{t=1}^{T}r_{it}r_{Mt}.
\end{eqnarray*}
Using matrix algebra, we can write these equations as:
$$
\left(\begin{array}{cc}
T & \sum_{t=1}^{T}r_{Mt}\\
\sum_{t=1}^{T}r_{Mt} & \sum_{t=1}^{T}r_{Mt}^{2}
\end{array}\right)\left(\begin{array}{c}
\hat{\beta}_{i,0}\\
\hat{\beta}_{i,M}
\end{array}\right)=\left(\begin{array}{c}
\sum_{t=1}^{T}r_{it}\\
\sum_{t=1}^{T}r_{it}r_{Mt}
\end{array}\right),
$$ {#eq-SI-matrixSystem}
which is of the form $\mathbf{Ax}=\mathbf{b}$ with 
$$
\mathbf{A}=\left(\begin{array}{cc}
T & \sum_{t=1}^{T}r_{Mt}\\
\sum_{t=1}^{T}r_{Mt} & \sum_{t=1}^{T}r_{Mt}^{2}
\end{array}\right),\,\mathbf{x}=\left(\begin{array}{c}
\hat{\beta}_{i,0}\\
\hat{\beta}_{i,M}
\end{array}\right),\,\mathbf{b}=\left(\begin{array}{c}
\sum_{t=1}^{T}r_{it}\\
\sum_{t=1}^{T}r_{it}r_{Mt}
\end{array}\right).
$$
Hence, we can determine $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$
by solving $\mathbf{x}=\mathbf{A}^{-1}\mathbf{b}.$ Now,^[The matrix $\mathbf{A}$ is invertible provided $\mathrm{det}(A)\neq0.$
This requires the sample variance of $R_{Mt}$ to be non-zero. ]
\begin{eqnarray*}
\mathbf{A}^{-1} & = & \frac{1}{\mathrm{det}(\mathbf{A})}\left(\begin{array}{cc}
\sum_{t=1}^{T}r_{Mt}^{2} & -\sum_{t=1}^{T}r_{Mt}\\
-\sum_{t=1}^{T}r_{Mt} & T
\end{array}\right),\\
\mathrm{det}(\mathbf{A}) & = & T\sum_{t=1}^{T}r_{Mt}^{2}-\left(\sum_{t=1}^{T}r_{Mt}\right)^{2}=T\sum_{t=1}^{T}\left(r_{Mt}-\hat{\mu}_{M}\right)^{2},\\
\hat{\mu}_{M} & = & \frac{1}{T}\sum_{t=1}^{T}r_{Mt}.
\end{eqnarray*}
Consequently,
$$
\left(\begin{array}{c}
\hat{\beta}_{i,0}\\
\hat{\beta}_{i,M}
\end{array}\right)=\frac{1}{T\sum_{t=1}^{T}\left(r_{Mt}-\hat{\mu}_{M}\right)^{2}}\left(\begin{array}{cc}
\sum_{t=1}^{T}r_{Mt}^{2} & -\sum_{t=1}^{T}r_{Mt}\\
-\sum_{t=1}^{T}r_{Mt} & T
\end{array}\right)\left(\begin{array}{c}
\sum_{t=1}^{T}r_{it}\\
\sum_{t=1}^{T}r_{it}r_{Mt}
\end{array}\right)
$$ {#eq-SI-ls-solution}
and so
$$
\hat{\beta}_{i,0} = \frac{\sum_{t=1}^{T}r_{Mt}^{2}\sum_{t=1}^{T}r_{it}-\sum_{t=1}^{T}r_{Mt}^{2}\sum_{t=1}^{T}r_{it}r_{Mt}}{T\sum_{t=1}^{T}\left(r_{Mt}-\hat{\mu}_{M}\right)^{2}},
$$ {#eq-SI-ls-alpha}
$$
\hat{\beta}_{i,M} = \frac{T\sum_{t=1}^{T}r_{it}r_{Mt}-\sum_{t=1}^{T}r_{Mt}\sum_{t=1}^{T}r_{it}}{T\sum_{t=1}^{T}\left(r_{Mt}-\hat{\mu}_{M}\right)^{2}}.
$$ {#eq-SI-ls-beta}
After a little bit of algebra (see end-of-chapter exercises) it can
be shown that
\begin{eqnarray*}
\hat{\beta}_{i,0} & = & \hat{\mu}_{i}-\hat{\beta}_{i}\hat{\mu}_{M},\\
\hat{\beta}_{i,M} & = & \frac{\hat{\sigma}_{iM}}{\hat{\sigma}_{M}^{2}},
\end{eqnarray*}
which are plug-in estimates for $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$
determined earlier. Hence, the least squares estimates of $\hat{\beta}_{i,0}$
and $\hat{\beta}_{i,M}$ are identical to the plug-in estimates.

The solution for the least squares estimates in @eq-SI-ls-solution
has an elegant representation using matrix algebra. To see this, define
the $T\times1$ vectors $\mathbf{r}_{i}=(r_{i1},\ldots,r_{iT})^{\prime}$,
$\mathbf{r}_{M}=(r_{M1},\ldots,r_{MT})^{\prime}$ and $\mathbf{1}=(1,\ldots,1)^{\prime}$.
Then we can re-write @eq-SI-matrixSystem as
$$
\left(\begin{array}{cc}
\mathbf{1}^{\prime}\mathbf{1} & \mathbf{1}^{\prime}\mathbf{r}_{M}\\
\mathbf{1}^{\prime}\mathbf{r}_{M} & \mathbf{r}_{M}^{\prime}\mathbf{r}_{M}
\end{array}\right)\left(\begin{array}{c}
\hat{\beta}_{i,0}\\
\hat{\beta}_{i,M}
\end{array}\right)=\left(\begin{array}{c}
\mathbf{1}^{\prime}\mathbf{r}_{i}\\
\mathbf{r}_{M}^{\prime}\mathbf{r}_{i}
\end{array}\right)
$$
or
$$
\mathbf{X}^{\prime}\mathbf{X}\hat{\boldsymbol{\beta}}_{i}=\mathbf{X}^{\prime}\mathbf{r}_{i}
$$ {#eq-SI-normalEquationsMatrix}
where $\mathbf{X}=(\begin{array}{cc}
\mathbf{1} & \mathbf{r}_{M})\end{array}$is a $T\times2$ matrix and $\hat{\boldsymbol{\beta}}_{i}=(\hat{\beta}_{i,0},\hat{\beta}_{i,M})^{\prime}$.
Provided $\mathbf{X}^{\prime}\mathbf{X}$ is invertible, solving @eq-SI-normalEquationsMatrix
for $\hat{\boldsymbol{\beta}}_{i}$ gives the least squares estimates
in matrix form:
$$
\hat{\boldsymbol{\beta}}_{i}=\left(\mathbf{X}^{\prime}\mathbf{X}\right)^{-1}\mathbf{X}^{\prime}\mathbf{r}_{i}.
$$ {#eq-SI-ls-solution-matrix}
The matrix form solution @eq-SI-ls-solution-matrix is especially
convenient for computation in R.

The least squares estimates of $\varepsilon_{t}$, $\sigma_{\varepsilon,i}^{2}$
and $R^{2}$ are the same as the plug-in estimators @eq-SI-pluginError,
@eq-SI-pluginErrorVariance and @eq-SI-pluginR2, respectively.
In the context of least squares estimation, the estimate $\hat{\sigma}_{\varepsilon,i}=\sqrt{\hat{\sigma}_{\varepsilon,i}^{2}}$
is called the *standard error of the regression*\index{Standard error} and measures
the typical magnitude of $\hat{\varepsilon}_{t}$ (difference between
observed return and fitted regression line). 

### Simple linear regression in R

-  don't do regression examples until statistical theory is discussed
-  computing least squares estimates using matrix algebra formulas
-  computing least squares estimates using `lm()`
    -  See discussion from my regression chapter in MFTSR
    -  describe structure of `lm()` function, extractor and method functions
-  Do analysis of example data
To be completed...


## Statistical Properties of SI Model Estimates

To determine the statistical properties of the plug-in principle/least
squares/ML estimators $\hat{\beta}_{i,0}$, $\hat{\beta}_{i,M}$ and
$\hat{\sigma}_{\varepsilon,i}^{2}$ in the SI model, we treat them as
functions of the random variables $\{(R_{i,t},R_{Mt})\}_{t=1}^{T}$
where $R_{t}$ and $R_{Mt}$ are assumed to be generated by the SI
model @eq-SI-SIreturn - @eq-SI-errorAssets. 

### Bias
\index{SI bias}
In the SI model, the estimators $\hat{\beta}_{i,0}$, $\hat{\beta}_{i,M}$
and $\hat{\sigma}_{\varepsilon,i}^{2}$ (with degrees-of-freedom adjustment)
are unbiased:
\begin{eqnarray*}
E[\hat{\beta}_{i,0}] & = & \beta_{i,0},\\
E[\hat{\beta}_{i,M}] & = & \beta_{i,M},\\
E[\hat{\sigma}_{\varepsilon,i}^{2}] & = & \sigma_{\varepsilon,i}^{2}.
\end{eqnarray*}
To shows that $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$ are unbiased,
it is useful to consider the SI model for asset $i$ in matrix form
for $t=1,\ldots,T$:
$$
\mathbf{R}_{i}=\beta_{i,0}\mathbf{1}+\beta_{i,M}\mathbf{R}_{M}+\varepsilon_{i}=\mathbf{X}\boldsymbol{\beta}_{i}+\varepsilon_{i},
$$
where $\mathbf{R}_{i}=(R_{i1},\ldots,R_{iT})^{\prime},$ $\mathbf{R}_{M}=(R_{M1},\ldots,R_{MT})^{\prime}$,
$\varepsilon_{i}=(\varepsilon_{i1},\ldots,\varepsilon_{iT})^{\prime}$,
$\mathbf{1}=(1,\ldots,1)^{\prime}$, $\mathbf{X}=(\begin{array}{cc}
\mathbf{1} & \mathbf{R}_{M})\end{array},$ and $\boldsymbol{\beta}_{i}=(\beta_{i,0},\beta_{i,M})^{\prime}.$ The estimator
for $\boldsymbol{\beta}_{i}$ is
\begin{eqnarray*}
\hat{\boldsymbol{\beta}}_{i} & = & (\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\mathbf{R}_{i}.
\end{eqnarray*}
Pluging in $\mathbf{R}_{i}=\mathbf{X}\boldsymbol{\beta}_{i}+\varepsilon_{i}$
gives 
\begin{eqnarray*}
\hat{\boldsymbol{\beta}}_{i} & = & (\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\left(\mathbf{X}\gamma_{i}+\varepsilon_{i}\right)=(\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\mathbf{X}\gamma_{i}+(\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\varepsilon_{i}\\
 & = & \boldsymbol{\beta}_{i}+(\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\varepsilon_{i}.
\end{eqnarray*}
Then
\begin{eqnarray*}
E[\hat{\boldsymbol{\beta}}_{i}] & = & \boldsymbol{\beta}_{i}+E\left[(\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\varepsilon_{i}\right]\\
 & = & \boldsymbol{\beta}_{i}+E\left[(\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}\right]E[\varepsilon_{i}]\,(\mathrm{because}\,\varepsilon_{it}\,\mathrm{is}\,\mathrm{independent}\,\mathrm{of}\,R_{Mt})\\
 & = & \boldsymbol{\beta}_{i}\,(\mathrm{because}\,E[\varepsilon_{i}]=\mathbf{0}).
\end{eqnarray*}
The derivation of $E[\hat{\sigma}_{\varepsilon,i}^{2}]=\sigma_{\varepsilon,i}^{2}$
is beyond the scope of this book and can be found in graduate econometrics
textbooks such as Hayashi (1980). 

### Precision
\index{SI precision}
Under the assumptions of the SI model, analytic formulas for estimates
of the standard errors for $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$
are given by:
$$
\widehat{\mathrm{se}}(\hat{\beta}_{i,0})  \approx\frac{\hat{\sigma}_{\varepsilon,i}}{\sqrt{T\cdot\hat{\sigma}_{M}^{2}}}\cdot\sqrt{\frac{1}{T}\sum_{t=1}^{T}r_{Mt}^{2}},
$$ {#eq-SI-se-alphahat}
$$
\widehat{\mathrm{se}}(\hat{\beta}_{i,M})  \approx\frac{\hat{\sigma}_{\varepsilon,i}}{\sqrt{T\cdot\hat{\sigma}_{M}^{2}}},
$$ {#eq-SI-se-betahat}
where "$\approx$" denotes an approximation based on the CLT that
gets more accurate the larger the sample size. 

**Remarks**:

-  $\widehat{\mathrm{se}}(\hat{\beta}_{i,0})$ and $\widehat{\mathrm{se}}(\hat{\beta}_{i,M})$ are smaller the smaller is $\hat{\sigma}_{\varepsilon,i}$. That is, the closer are returns to the fitted regression line the smaller are the estimation errors in $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$.
-  $\widehat{\mathrm{se}}(\hat{\beta}_{i,M})$ is smaller the larger is $\hat{\sigma}_{M}^{2}$. That is, the greater the variability in the market return $R_{Mt}$ the smaller is the estimation error in the estimated slope coefficient $\hat{\beta}_{i,M}$. This is illustrated in Figure xxx. The left panel shows a data sample from the SI model with a small value of $\hat{\sigma}_{M}^{2}$ and the right panel shows a sample with a large value of $\hat{\sigma}_{M}^{2}$. The right panel shows that the high variability in $R_{Mt}$ makes it easier to identify the slope of the line. 
-  Both $\widehat{\mathrm{se}}(\hat{\beta}_{i,0})$ and $\widehat{\mathrm{se}}(\hat{\beta}_{i,M})$ go to zero as the sample size, $T$, gets large. Since $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$ are unbiased estimators, this implies that they are also consistent estimators. That is, they converge to the true values $\beta_{i,0}$ and $\beta_{i,M}$, respectively, as $T\rightarrow\infty$.
-  In R, the standard error values @eq-SI-se-alphahat and @eq-SI-se-betahat are computed using the `summary()` function on an "`lm`" object. 


There are no easy formulas for the estimated standard errors for $\hat{\sigma}_{\varepsilon,i}^{2}$,
$\hat{\sigma}_{\varepsilon,i}$ and $\hat{R}^{2}$. Estimated standard
errors for these estimators, however, can be easily computed using
the bootstrap.

::: {#exm-computing-estimated-standard-errors-for-hat-beta-i-0-and-hat-beta-i-m-in-r .example name="Computing estimated standard errors for $\\hat{\\beta}_{i,0}$ and $\\hat{\\beta}_{i,M}$ in R"}

to be completed

\hfill{}$\blacksquare$
:::

### Sampling distribution and confidence intervals.
\index{SI confidence intervals}
Using arguments based on the CLT, it can be shown that for large enough
$T$ the estimators $\hat{\beta}_{i,0}$ and $\hat{\beta}_{i,M}$ are
approximately normally distributed:
\begin{eqnarray*}
\hat{\beta}_{i,0} & \sim & N(\beta_{i,0},\widehat{\mathrm{se}}(\hat{\beta}_{i,0})^{2}),\\
\hat{\beta}_{i,M} & \sim & N(\beta_{i,M},\widehat{\mathrm{se}}(\hat{\beta}_{i,M})^{2}),
\end{eqnarray*}
where $\widehat{\mathrm{se}}(\hat{\beta}_{i,0})$ and $\widehat{\mathrm{se}}(\hat{\beta}_{i,M})$
are given by @eq-SI-se-alphahat and @eq-SI-se-betahat,
respectively.

## Further Reading: Single Index Model

To be completed ...

## Exercicses

To be completed ... 


{Ruppert} Ruppert, D.A.. \textit{Statistics and Data Analysis for Financial Engineering}, Springer-Verlag. 


